### 2019-05-03 emp-notes/emp-notes.tex
Initial commit

\documentclass[12pt,letter]{article}
\usepackage[left=0.8in,right=0.8in,top=1in,bottom=1in]{geometry}
\usepackage{amsmath}
\usepackage{pdflscape}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{multicol}
\usepackage{microtype}
\usepackage{euscript}
\usepackage{epsfig}
\usepackage{epstopdf}
\usepackage{mathrsfs}
\usepackage{tikz}
\newcommand{\hypo}{\mathcal{H}}            
\bibliographystyle{ieeetr}

\usepackage[flushleft]{threeparttable}

%\usepackage[cp1251]{inputenc}
\usepackage[english]{babel} 
\DeclareMathOperator{\rank}{rank}
\newcommand*{\hm}[1]{#1\nobreak\discretionary{}
            {\hbox{$\mathsurround=0pt #1$}}{}}

            \def\onepc{$^{\ast\ast}$} \def\fivepc{$^{\ast}$}
\def\tenpc{$^{\dag}$}
\def\legend{\multicolumn{4}{l}{\footnotesize{Significance levels
:\hspace{1em} $\dag$ : 10\% \hspace{1em}
$\ast$ : 5\% \hspace{1em} $\ast\ast$ : 1\% \normalsize}}}

\newcommand{\bs}[1]{\boldsymbol{#1}}  
\newcommand{\bsA}{\boldsymbol{A}}

%\setstretch{1}                         
\flushbottom                            
\righthyphenmin=2                      
\pagestyle{plain}                       
%\settimeformat{hhmmsstime}  
\widowpenalty=300                   
\clubpenalty=3000                     
\setlength{\parindent}{0em}           
\setlength{\topsep}{0pt}              
\usepackage[pdftex,unicode,colorlinks=true,urlcolor=blue]{hyperref}
\usepackage{bbm}
\renewcommand{\emptyset}{\varnothing}

\setlength{\parskip}{0.5\baselineskip plus2pt minus2pt}

\newcommand{\e}{\varepsilon}
\DeclareMathOperator*{\Argmax}{\mathrm{Argmax}}
\DeclareMathOperator*{\Argmin}{\mathrm{Argmin}}
\DeclareMathOperator*{\argmax}{\mathrm{arg\,max}}
\DeclareMathOperator*{\argmin}{\mathrm{argmin}}

\newcommand{\blp}{\mathrm{BLP}}
\DeclareMathOperator*{\plim}{\mathrm{plim}}
\DeclareMathOperator{\Max}{\mathrm{Max}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\Y}{\mathcal{Y}}
\newcommand{\Z}{\mathcal{Z}}
\renewcommand{\geqslant}{\geq}
\renewcommand{\leqslant}{\leq}
\newcommand{\p}{\bs p}
\newcommand{\y}{\bs y}
\def\dd#1#2{\frac{\partial#1}{\partial#2}}

\renewcommand{\emptyset}{\varnothing}

\DeclareMathOperator{\tr}{\mathrm{tr}}

\newcommand{\bb}{\bs \beta}
\newcommand{\X}{\bs X}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\PP}{\mathbb{P}}
\DeclareMathOperator{\V}{\mathbb{V}}
\DeclareMathOperator{\CM}{\mathbb{C}}
\renewcommand{\C}{\CM}
\DeclareMathOperator{\var}{\mathrm{var}}
\DeclareMathOperator{\cov}{\mathrm{cov}}
\DeclareMathOperator{\corr}{\mathrm{corr}}
\DeclareMathOperator{\MSE}{\mathrm{MSE}}
\DeclareMathOperator{\Bias}{\mathrm{Bias}}
\renewcommand{\P}{\PP}
\newcommand{\dsim}{\stackrel{d}{\sim}}
\newcommand{\hn}{\mathcal{H}_0}
\newcommand{\ha}{\mathcal{H}_a}
\newcommand{\thetab}{\bs \theta}
\newcommand{\pv}{\text{P-value}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\MLE}{\scriptscriptstyle MLE}
\newcommand{\LR}{\mathrm{LR}}
\newcommand{\I}{\mathbb{I}}
\newcommand{\sumin}{\sum\limits_{i=1}^n}
\newcommand{\sumti}{\sum\limits_{t=0}^\infty}
\newcommand{\hbeta}{\hat{\beta}}
\newcommand{\halpha}{\hat{\alpha}}
\newcommand{\hsigma}{\hat{\sigma}}
\newcommand{\hvar}{\widehat{\var}}
\newcommand{\hcov}{\widehat{\cov}}
\newcommand{\Q}{\mathbb{Q}}

\newcommand{\pconv}{\xrightarrow{ \ p \ }}
\newcommand{\dconv}{\xrightarrow{ \ d \ }}
\newcommand{\asconv}{\xrightarrow{ \ a.s. \ }}
\newcommand{\msconv}{\xrightarrow{ \ m.s. \ }}

\newcommand{\pic}[4][h!]{\begin{figure}[#1]

\begin{center}\includegraphics[width=#2cm]{#3}\caption{#4\label{#3}}\end{center}
\end{figure}}

%outtex
\def\onepc{$^{\ast\ast}$} \def\fivepc{$^{\ast}$}
\def\tenpc{$^{\dag}$}
\def\legend{\multicolumn{4}{l}{\footnotesize{Significance levels
:\hspace{1em} $\dag$ : 10\% \hspace{1em}
$\ast$ : 5\% \hspace{1em} $\ast\ast$ : 1\% \normalsize}}}
%end outtex

%\bibliographystyle{ieeetr}

\newcommand{\laseq}{\stackrel{\lambda\text{-a.e.}}{=}}
\renewcommand{\d}{\underline}
\renewcommand{\u}{\overline}
\newcommand{\td}{\underline{\theta}}
\newcommand{\tu}{\overline{\theta}}
%\renewcommand{\theenumi}{\alph{enumi}}
%\renewcommand{\labelenumi}{(\theenumi)}
%\renewcommand{\theenumii}{\roman{enumii}}
%\renewcommand{\labelenumii}{\theenumii.}
%\renewcommand{\theenumiii}{\arabic{enumiii}}
%\renewcommand{\labelenumiii}{\theenumiii.}
%\renewcommand{\epsilon}{\varepsilon}
\newcommand{\hneq}{\stackrel{\hn}{=}}
\newcommand{\deq}{\stackrel{d}{=}}
%\title{416-2 Final Project\\
%Reproductive Technologies, Aging and Fertility Choice (Proposal)}
%\author{Egor Kozlov}
\begin{document}
%\begin{center}\textbf{Economics 416-1} \\ \emph{By Egor Kozlov}\end{center}
%\maketitle

\section{Introduction}
This note studies how relative timing of marriage and fertility affects households' outcomes in the US economy. The key finding is that for general US population having kids before marriage is associated with higher divorce chances in the future, as opposed to the people who marry before having kids. This pattern, however, is not uniform across different population groups. Namely, it is apparent for the households with high income per person or high level of female's education. In contrast, for the groups with lower income and lower education the relation can be even reversed. In the further parts I provide more details on how I establish this finding, show a comparison between different subgroups and some additional data evidence and then try to hypothesise about what can be the origin of these differences.

\section{Empirical strategy}
I use American Community Survey yearly data for the period of 2008--2016. The main reason for this choice is that year of marriage dates are collected only since 2008. The main limitation is that the survey keeps only the year of the most recent marriage, and I am mostly interested in the event of the first marriage and the having the first child. Additional issues are obvious cross-sectional nature of the data and no possibility of tracking kids that do not live with their mothers. However, assuming none of the above drives the results entirely with additional conditioning described below we still have more than a million of relevant observations.

Since upon divorce children usually stay with mothers, I consider only individual records of females here. My sample contains women of ages 18--50, who is marked as household's head or head's spouse. I also keep only observations where ACS  variable ``number of families'' is equal to one, mainly to avoid considering mulitgenerational households. Finally, for the main part I keep only people who married only once. In practice, it means that some cases are excluded from the sample, for instance:
\begin{enumerate}
\item If a couple is divorce and no kids stay with mother, she is excluded from the sample.
\item If a single mother with a child marries someone else than child's father (and she was never married before), the couple will be considered as if they had their kid before marriage. This is a complex issue that is not obvious how to consider, however as a supplemental evidence I try to exclude married couples where the child is marked as ``step-child'' relative to one of the parents and it does not change findings much.
\item Multigenerational or multi-family households are excluded, as they have complicated structure and household's traits may be misassigned. For instance, a divorced women living with her parents and her child will not be considered as household's head if one of the parents work. 
\item If all children have moved out, this household is considered as having no kids and is excluded from the consideration. 
\item If the family had two kids, but one of them has moved out, my methodology will consider this household as having only one kid. If the older kid was born before the marriage and the younger --- after, this will lead to the wrong assignment of relative timing for this household. This cannot be treated well in ACS data, to mitigate this I keep only the records of women on ages 50 and below, so the majority of the sample have kids before the age of 18. 
\end{enumerate}

Full cross-section of women of age 18--50 satisfying the restriction above contains 3.7 million observations. For the considerations of comparing divorce probabilities I have to exclude women who were never married, married more than once or do not have kids in their household's record. Around two thirds of the sample have married exactly once, interestingly this percentage is not drastically different for women of 40--50: share of never married falls but share of re-married increases as well.

\subsection{Timing Variables\label{km}}
The key variable for this study is the time when the first kid was born relative to the time woman was first married. ACS has record of the eldest own child present in the household, and the year of the first marriage. Both variables have some quality issues, so I excluded situations where age of the eldest child is more that age of the mother minus 14, and when the age of first marriage was below 18. Then, for the people married once I construct the following variable:
\[\{ \text{Marriage $\to$ Kids}\}_i = \I(\text{Year eldest kid born} - \text{Year married} \geq 1).\]

Threshold 1 seems like a natural choice, however it slightly misclassifies couples: if the couple got first kid in April, 2015 and got married in December, 2014, then most likely they knew that the wife is pregnant, so from the point of view of household's decisions it should be classified as $\text{K}\to\text{M}$. However, changing this threshold for 2 does not make a lot of difference (see the robustness check part for this).

\subsection{Income Variables}
For the purpose of comparing people with different income levels, I build the variable of household income per capita. To account for presence of children, I use two different definitions:
\[\text{HHI}_{eq} = \frac{\text{Total Household Income}}{1 + 0.7\times \I(\text{husband/partner is present}) + 0.5 \times \{ \text{\# of children}\}}\]
\[\text{HHI}_{pc} = \frac{\text{Total Household Income}}{\{\text{\# of people of any age in the household}\}}\]
The second one income per capita in literal definition, the first one corresponds to OECD equivalence scale and accounts for economies of scale in the household. The first definition is more preferable, though it may be flawed in case of more complicated family structures (say, more than two adults per household), but since I restrict the attention to the households of particular composition described above this does not seem to be an important issue. Note that in ACS a sizeable share of divorced and separated women still have some partner present in the household, and both measures account for this possibility.

Since the quality of self-reported income is questionable, for the income considerations I exclude the households where $\text{HHI}_{eq}$ is below $\$3,000$ (yearly), corresponding to about $\$10,000$ household income for two-adults and three-children household. This for sure excludes population living way below the poverty line, however the it is important not to pool people with low income and people with badly reported income together.

The key component of the work is to construct income groups. In each state, year and age group I compute household income quantiles separately, where age groups are $[18;25]$, $[26;30]$, $[31;35]$, ..., $[46;50]$ based on female's age. So, threshold for being top 30\% at the age of 40 in California in 2009 is different from threshold for being top 30\% in Kentucky at the age of 30 in 2015.

As a result, my income groups are $\{\text{Bottom 30\%},\,\text{Mid 40\%},\,\text{Top 30\%}\}$. Finer division is also possible, however this given three-variable conditioning this may lead to some groups being rather small.

\subsection{Education Variables}
ACS reports education variables for everyone in household. To avoid ambiguity, I classify households based on female education only. I have the following four groups: high school or below, some college, college graduate and more than college. Interestingly, although education and income groups are correlated, about 10\% of highest education group fall within lower 30\% of income.

\subsection{Weights}
In calculations of means and frequencies I use personal weights provided by ACS. The only limitation is that computing income quantiles to classify people by groups is done without weights (based on raw observations). This is a software issue and I am working on fixing it, however it does not feel like something important for the results (as using unweighted data everywhere delivers very similar results).

\section{Key results}
See Table \ref{main-results} for the main findings. Income variables are base on $\text{HHI}_{eq}$ there. The table reads as follows: first column indicates different subgroups of women, the second denotes number of observations of any marital/fertility status. Left part provides breakdown of the shares of people in cross-section who have never married, married once or married more than once (re-married someone after divorce or death of spouse). ``\% divorced'' indicates are how much of these people were divorced at the moment of survey.

Right part of the table focuses on people who have married exactly once and, in addition, have at least one own child in the household. These people are divided on two groups $\text{M}\to\text{K}$ (marriage then kids) and $\text{K}\to\text{M}$ (kids then marriage), as described in section \ref{km}. Group size column gives the proportion of these two groups. Lastly, the table reports shares of divorced people. Column ``All'' corresponds to a share of divorced women among those who have married once and have at least one kid\footnote{This is provided mainly for the purpose of comparing it with share of divorced women in general population. It turns out to be slightly higher yet very close, and that is easily explained by the fact that people who had kids are on average older}. Two last columns are of the main interest: they report shares of divorced people in two groups. Difference between them is the key object I am describing here.

The key findings summarizing the table are:
\begin{enumerate}
\item For the general population of women (satisfying restrictions stated above), having kids before marriage is associated with about one-quarter to one-third larger divorce chances, depending on the ages we consider.
\item The fact that this difference is more pronounced for age group of 40--50 may just reflect the fact that some younger people in $\text{K}\to\text{M}$ groups are at a higher risk of divorce but it has not happened yet. However, this may as well reflect the fact that the effect I document here is more apparent for elder cohorts. This can be partially controlled with different methodology when I compare regression coefficients on dummy rather than raw differences, but this approach is less transparent so I do not do it in this exercise. 
\item The most important finding is that the observed differences in divorce probabilities are mainly driven by medium and high-income groups. The overall divorce rates fall with income for both groups, but the difference in divorce rates increases. For lower 30\% of income, in contrast, having kids before marriage does not really predicts divorce.
\item Divorce rates vary sharply be race, moving in the same direction with percentage of never married. Difference between the two groups I consider is more pronounced for white population rather than for non-whites. However, this is the result of income differences between different racial groups: conditional on being in high income, black people have very similar difference between groups to whites, however their overall divorce rate is still higher. I conclude from that that the race is not the main driving factor, though it drives overall divorce rates quite a lot.
\item Conditioning on education reveals the same pattern as conditioning on income: for the people who finished college having kids before marriage predicts higher likelihood of divorce, and it does not work well for people who have never been to college. However, females who have never been to college but live in high-income households look similar enough to females who have finished the college. In general, it seems like income is still a better predictor than education, but since income variables are more noisy, education story also seems compelling. Some people with high (potential) earnings may be classified as low-income, for instance when the wife is out of labor force because she takes care of a newborn.
\end{enumerate}

In general, it seems like having kids before marriage increases divorce probabilities for educated, middle-income Americans, who typically have small families and high preference for kid's education or high value of time that is required to be spent with the child. Something with investment to children is probably the key dimension of rationalizing observed differences.

\section{Additional Evidence}
This part establishes several pieces of additional evidence that may help understanding the sources of the result.

\subsection{Sample of Recently Divorced\label{rec-div-section}}
In addition to the current marital status ACS has a variable ``divorced last year''. This group is relatively small (1--2\% of female population), however this is still enough for the descriptive purposes, as sample sizes reach few millions.

First, see Figures \ref{age-kid} and \ref{age-mother} for distribution of mother's and child's age at the moment of divorce. The graphs have strikingly different nonuniformity: for couples who have kids after their marriage share of divorced couples where kids have age of 5 is very similar to the share of divorced couples where kids have the age of 16. In contrast, for the couples who have kids before marriage likelihood of divorce grows significantly with the child's age. This comparison is, however, flawed because of the following issue: my construction, there will be very few divorced people with kids of age 2 who made these kids before the marriage as this would mean the couple got married exactly 1 year ago. In contrast, the group of couples whose kids are 2 year old and who made them after they got married contains people who married 3 or more years ago, and this group is definitely larger.

Therefore, additionally, I plot the share of recently divorced by the age of the eldest kid:
\[D(a,M\to K) = \frac{\text{\# Divorced this year, eldest kid of age $a$, $M\to K$}}{\text{\# eldest kid of age $a$, $M\to K$}},\]
\[D(a,K\to M) = \frac{\text{\# Divorced this year, eldest kid of age $a$, $K\to M$}}{\text{\# eldest kid of age $a$, $K\to M$}}.\]
see Figures \ref{rec-div} for the main result and \ref{rec-div-educ} for the breakdown in education.

These pictures are suggestive, though not totally conclusive, that divorces for the people who have kids before marriage are concentrated around the ages 6--12, and this concentration is somewhat more pronounced for the groups with lower education level. For people who have marriage then kids the pattern is, again, more uniform. This evidence is pretty weak due to smaller sample sizes, cross-sectional nature of the data, that makes these shares tricky to interpret as probability, and generally noisy look of the graphs. In the further iterations I may come up with some formal statistical test to verify this claim, however defining rigorously what is meant by being ``more concentrated'' is tricky.

\section{Robustness Checks}
In this part I change some things in methodology to look how it affects the key findings. I replicate the second part of the table for the following cases, corresponding to parts of table \ref{rcheck}:
\begin{enumerate}
\item I compute income percentiles without conditioning on states (so top 30\% of income refers to to 30\% in the whole US population and not in the given state). Things stay the same qualitatively, but now the pattern is more uniform across income groups. This for sure requires additional investigation, but in general I think that income groups should be defined within states to control for things like housing and healthcare costs.
\item I return conditioning on states but use income per capita instead of equivalence scale income. The results hardly change from the baseline. This probably means that equivalence scale (assuming particular family structure) is good enough for this exercise.
\item I redefine group $M\to K$ so that difference in year your kid is born and in year you are married should be above two. Differences are a bit lower (since we mixed up two groups partially), but still look the same.
\item Instead of probabilities of being divorced I consider probabilities of being divorced or separated. In this case, the differences between groups are less apparent (it seems that separation is more common for low-income people). Key finding remains the same.
\item Using alternative divorce variables: I use ``divorced last year'' as in Section \ref{rec-div-section} as an alternative to using ``divorced now''. Key findings are still the same, and again income differences pattern is less apparent (but is still present if we consider the ratio of odds).
\end{enumerate}

In general, key moment remains similar with these robustness checks, though its relation with income is more fragile. However, we can observe the same results with respect to education (that is measured more precisely than income), so I still think the story is valid. It may be beneficial to discard income variables altogether, however the group of college graduates has great dispersion within it. This all requires some additional investigation. However, I think that the main strategy for the further parts is using regression on dummy variables with many controls instead of comparing raw differences: my group division is crude and there is a lot of heterogeneity inside.

\newpage
\begin{landscape}
\begin{table}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c||c|c|c|c|c|}\hline
& \multicolumn{5}{|c||}{\textbf{All sample}}  &  \multicolumn{5}{|c|}{\textbf{Married once and have a kid}}\\\hline
& & \multicolumn{3}{|c|}{Times married} &  & \multicolumn{2}{|c|}{Group size}  & \multicolumn{3}{|c|}{\% divorced}  \\\hline
 &  Obs & \footnotesize Never &  \footnotesize Once &  \footnotesize More & \footnotesize \% divorced  & \footnotesize M $\rightarrow$ K &  \footnotesize K $\rightarrow$ M &  All & \footnotesize M $\rightarrow$ K & \footnotesize K $\rightarrow$ M \\\hline
\emph{All women, 20--50} & \footnotesize 3,798,950 & 19.3\% & 65.7\% & 12.8\% &   9.5\% & 75.5\% & 24.5\% & 10.3\% &  9.5\% & 12.7\% \\
\emph{All women, 40--50} & \footnotesize 1,793,232 & 10.3\% & 67.1\% & 18.6\% &  11.4\% & 82.8\% & 17.2\% & 12.6\% & 11.9\% & 16.3\% \\\hline
\emph{White, non-hispanic, 20--50} &  \footnotesize  2,540,035 & 14.8\% & 67.4\% & 14.8\% &   9.0\% & 81.5\% & 18.5\% &  9.9\% &  9.2\% & 12.7\% \\
\emph{White, non-hispanic, 40--50} &   \footnotesize 1,242,287 &  7.1\% & 66.9\% & 20.9\% &  10.7\% & 89.2\% & 10.8\% & 11.9\% & 11.4\% & 16.1\% \\ \hline
\emph{Hispanic, 20--50} &   \footnotesize 356,923 & 17.1\% & 70.5\% & 11.3\% &   8.6\% & 65.8\% & 34.2\% &  9.3\% &  9.3\% &  9.4\% \\
\emph{Hispanic, 40--50} &   \footnotesize 149,912 &  9.2\% & 72.4\% & 16.2\% &  10.6\% & 73.4\% & 26.6\% & 11.7\% & 11.6\% & 12.0\% \\\hline
\emph{Black, 20--50} &  \footnotesize 398,573 & 43.8\% & 46.7\% &  8.5\% &  17.5\% & 50.0\% & 50.0\% & 19.7\% & 20.8\% & 18.6\% \\
\emph{Black, 40--50} &  \footnotesize 184,861 & 28.6\% & 55.7\% & 13.7\% &  20.3\% & 58.3\% & 41.7\% & 23.7\% & 24.3\% & 22.9\% \\\hline\hline
\emph{Not in MSA, 20--50} &  \footnotesize 515,274 & 14.5\% & 64.5\% & 16.8\% &   9.2\% & 73.3\% & 26.7\% & 11.3\% & 10.5\% & 13.3\% \\
%\emph{Rural, 40--50} &  248,656 &  6.8\% & 62.7\% & 23.4\% &  10.2\% & 84.1\% & 15.9\% & 13.0\% & 12.1\% & 17.4\% \\\hline
\emph{Main city in MSA, 20--50} &  \footnotesize 454,220 & 33.7\% & 57.0\% &  8.3\% &  11.3\% & 70.1\% & 29.9\% & 11.4\% & 10.5\% & 13.5\% \\
%\emph{Urban, 40--50} & 184,873 & 21.7\% & 62.8\% & 13.4\% &  14.8\% & 74.7\% & 25.3\% & 14.8\% & 14.0\% & 17.4\% \\
\hline\hline
\emph{Lower 30\% income, 20--50} &  \footnotesize 1,104,277 & 23.0\% & 61.7\% & 12.8\% &  13.5\% & 65.6\% & 34.4\% & 15.5\% & 15.6\% & 15.3\% \\
\emph{... if white} &  \footnotesize 603,952 & 16.9\% & 62.3\% & 16.7\% &  14.7\% & 72.1\% & 27.9\% & 17.8\% & 18.0\% & 17.6\% \\
\emph{... if black} &  \footnotesize 168,278 & 48.5\% & 43.0\% &  7.6\% &  20.6\% & 43.5\% & 56.5\% & 23.7\% & 27.2\% & 21.0\% \\
\emph{Lower 30\% income, 40--50} &  \footnotesize  524,698 & 13.5\% & 64.2\% & 17.9\% &  15.8\% & 75.2\% & 24.8\% & 18.9\% & 19.0\% & 18.6\% \\\hline
\emph{Mid 40\% income, 20--50} &  \footnotesize 1,468,999 & 15.2\% & 68.7\% & 13.7\% &   8.3\% & 77.2\% & 22.8\% &  8.8\% &  8.3\% & 10.3\% \\
\emph{Mid 40\% income, 40--50} &  \footnotesize 701,740 &  7.5\% & 69.0\% & 19.5\% &  10.3\% & 84.0\% & 16.0\% & 11.4\% & 10.7\% & 14.8\% \\\hline
\emph{High 30\% income, 20--50} &  \footnotesize 1,098,178 & 16.8\% & 69.1\% & 12.2\% &   6.0\% & 88.4\% & 11.6\% &  4.5\% &  4.1\% &  7.6\% \\
\emph{... if white} &  \footnotesize 840,754 & 14.8\% & 69.9\% & 13.1\% &   5.7\% & 90.3\% &  9.7\% &  4.2\% &  3.9\% &  6.7\% \\
\emph{... if black} &  \footnotesize 63,672 & 35.5\% & 53.6\% &  9.8\% &  13.0\% & 67.7\% & 32.3\% & 11.5\% & 11.0\% & 12.7\% \\
\emph{High 30\% income, 40--50} &  \footnotesize 526,694 &  9.2\% & 68.9\% & 18.4\% &   7.6\% & 91.0\% &  9.0\% &  5.9\% &  5.4\% & 10.3\% \\\hline\hline
\emph{High school only, 20--50} &  \footnotesize 1,360,933 & 20.2\% & 62.1\% & 14.7\% &   9.8\% & 64.5\% & 35.5\% & 11.2\% & 11.0\% & 11.5\% \\
\emph{... if white} &  \footnotesize 791,722 & 13.5\% & 62.3\% & 19.3\% &  10.0\% & 70.5\% & 29.5\% & 12.2\% & 12.1\% & 12.6\% \\
\emph{... if black} &  \footnotesize 170,288 & 49.7\% & 41.5\% &  7.7\% &  17.2\% & 41.2\% & 58.8\% & 19.5\% & 22.2\% & 17.6\% \\
\emph{... if bottom 30\% income} &  \footnotesize 628,079 & 22.9\% & 61.6\% & 12.9\% &  12.0\% & 61.3\% & 38.7\% & 13.5\% & 13.8\% & 13.1\% \\
\emph{... if top 30\% income} &  \footnotesize 164,635 & 12.7\% & 63.6\% & 19.3\% &   5.8\% & 75.5\% & 24.5\% &  5.4\% &  5.0\% &  6.5\% \\
\hline
\emph{College graduates, 20--50} &  \footnotesize 915,741 & 17.9\% & 71.6\% &  9.4\% &   7.6\% & 87.6\% & 12.4\% &  7.4\% &  6.8\% & 11.8\% \\
\emph{... if white} &  \footnotesize 684,079 & 16.3\% & 72.7\% &  9.8\% &   7.2\% & 90.6\% &  9.4\% &  6.9\% &  6.5\% & 10.5\% \\
\emph{... if black} &  \footnotesize 62,078 & 36.5\% & 54.4\% &  8.3\% &  15.9\% & 62.8\% & 37.2\% & 17.2\% & 17.0\% & 17.6\% \\
\emph{... if bottom 30\% income} &  \footnotesize 128,296 & 18.5\% & 69.9\% & 10.2\% &  12.5\% & 82.0\% & 18.0\% & 13.6\% & 13.0\% & 16.4\% \\
\emph{... if top 30\% income} &  \footnotesize 410,656 & 18.6\% & 71.2\% &  9.2\% &   5.7\% & 91.6\% &  8.4\% &  4.0\% &  3.7\% &  7.9\% \\
\hline
\end{tabular}
\end{center}
\caption{ACS 2008--2016, Main Results\label{main-results}}
\end{table}
\end{landscape}

\newpage

\begin{figure}
\begin{center}
\includegraphics[scale=0.9]{hist-div-age.eps}
\caption{Age of (eldest) kid at the moment of divorce. \\ Restriction: eldest kid is below 18.\label{age-kid}}
\end{center}
\end{figure}
\begin{figure}
\begin{center}
\includegraphics[scale=0.9]{hist-div-age-mother.eps}
\caption{Age of mother at the moment of divorce.  \\ Restriction: eldest kid is below 18.\label{age-mother}}
\end{center}
\end{figure}

\newpage

\begin{figure}
\begin{center}
\includegraphics[scale=0.9]{rec-div.eps}
\caption{Share of recently divorced by age of the kid.\label{rec-div}}
\end{center}
\end{figure}
\begin{figure}
\begin{center}
\includegraphics[scale=0.9]{rec-div-educ.eps}
\caption{Share of recently divorced by age of the kid and education.\label{rec-div-educ}}
\end{center}
\end{figure}

\newpage

\begin{table}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|}\hline
&   \multicolumn{5}{|c|}{\textbf{Married once and have a kid}}\\\hline
&  \multicolumn{2}{|c|}{Group size}   & \multicolumn{3}{|c|}{\% divorced}  \\\hline
& \footnotesize M $\rightarrow$ K &  \footnotesize K $\rightarrow$ M &  All & \footnotesize M $\rightarrow$ K & \footnotesize K $\rightarrow$ M \\\hline
\emph{All women, 20--50} & 75.5\% & 24.5\% & 10.3\% &  9.5\% & 12.7\% \\\hline\hline
\multicolumn{6}{|c|}{\emph{Baseline (for reference)}} \\\hline\hline
\emph{Lower 30\% income, 20--50} &  75.2\% & 24.8\% & 18.9\% & 19.0\% & 18.6\% \\\hline
\emph{Mid 40\% income, 20--50} & 77.2\% & 22.8\% &  8.8\% &  8.3\% & 10.3\% \\\hline
\emph{High 30\% income, 40--50} & 91.0\% &  9.0\% &  5.9\% &  5.4\% & 10.3\% \\
\emph{... if white} & 90.3\% &  9.7\% &  4.2\% &  3.9\% &  6.7\% \\
\emph{... if black} & 67.7\% & 32.3\% & 11.5\% & 11.0\% & 12.7\% \\\hline\hline
\multicolumn{6}{|c|}{\emph{No conditioning on states when computing income quantiles}} \\\hline\hline
\emph{Lower 30\% income, 20--50} & 73.3\% & 26.7\% & 11.3\% & 10.5\% & 13.3\% \\\hline
\emph{Mid 40\% income, 20--50} &   83.2\% & 16.8\% &  6.5\% &  6.1\% &  8.4\% \\\hline
\emph{High 30\% income, 40--50} & 92.0\% &  8.0\% &  3.5\% &  3.2\% &  6.4\% \\
\emph{... if white} & 92.9\% &  7.1\% &  3.4\% &  3.2\% &  5.7\% \\
\emph{... if black} & 76.3\% & 23.7\% &  9.2\% &  8.7\% & 10.8\% \\\hline\hline
\multicolumn{6}{|c|}{\emph{Income definition: $HHI_{pc}$}} \\\hline\hline
\emph{Lower 30\% income, 20--50} &  65.2\% & 34.8\% & 15.1\% & 15.1\% & 15.1\% \\\hline
\emph{Mid 40\% income, 20--50} & 78.3\% & 21.7\% &  8.4\% &  8.0\% & 10.0\% \\\hline
\emph{High 30\% income, 40--50}& 89.6\% & 10.4\% &  4.5\% &  4.1\% &  7.6\% \\
\emph{... if white} & 91.1\% &  8.9\% &  4.2\% &  4.0\% &  6.9\% \\
\emph{... if black} & 70.3\% & 29.7\% & 11.2\% & 10.7\% & 12.3\% \\\hline\hline
\multicolumn{6}{|c|}{\emph{Redefine:} $\{\text{M $\to$ K}\}_i = \I(\text{Year eldest kid born} - \text{Year married} \geq 2)$} \\\hline\hline
\emph{All women, 20--50} &  63.1\% & 36.9\% & 10.3\% &  9.2\% & 12.2\% \\\hline
\emph{Lower 30\% income, 20--50} &  50.6\% & 49.4\% & 15.5\% & 16.0\% & 14.9\% \\\hline
\emph{Mid 40\% income, 20--50} & 64.9\% & 35.1\% &  8.8\% &  8.1\% & 10.0\% \\\hline
\emph{High 30\% income, 20--50} & 79.8\% & 20.2\% &  4.5\% &  3.9\% &  6.9\% \\\hline\hline
\multicolumn{6}{|c|}{\emph{Divorced or separated instead of just divorced}} \\\hline\hline
\emph{All women, 20--50} &  75.5\% & 24.5\% & 14.1\% & 12.2\% & 19.9\% \\\hline
\emph{Lower 30\% income, 20--50} &  65.6\% & 34.4\% & 22.5\% & 21.0\% & 25.4\% \\\hline
\emph{Mid 40\% income, 20--50} & 77.2\% & 22.8\% & 10.7\% &  9.8\% & 13.8\% \\\hline
\emph{High 30\% income, 20--50} & 88.4\% & 11.6\% &  5.2\% &  4.7\% &  9.3\% \\\hline\hline
\multicolumn{6}{|c|}{\emph{Divorced last year only}} \\\hline\hline
\emph{All women, 20--50} &  75.5\% & 24.5\% &  1.3\% &  1.1\% &  1.8\% \\\hline
\emph{Lower 30\% income, 20--50} &  73.3\% & 26.7\% &  1.5\% &  1.3\% &  2.2\% \\\hline
\emph{Mid 40\% income, 20--50} & 77.2\% & 22.8\% &  1.1\% &  1.0\% &  1.4\% \\\hline
\emph{High 30\% income, 20--50} & 88.4\% & 11.6\% &  0.5\% &  0.5\% &  0.9\% \\\hline\hline
\end{tabular}
\end{center}
\caption{ACS 2008--2016, Robustness checks\label{rcheck}}
\end{table}

\end{document}

### 2019-05-03 main/draft.tex
Initial commit

\documentclass[12pt,letter]{article}
\usepackage[left=0.8in,right=0.8in,top=1in,bottom=1in]{geometry}
\usepackage{amsmath}
\usepackage{pdflscape}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{multicol}
\usepackage{multirow}
\usepackage{microtype}
\usepackage{euscript}
\usepackage{epsfig}
\usepackage{epstopdf}
\usepackage{mathrsfs}
\usepackage{tikz}
\newcommand{\hypo}{\mathcal{H}}            
\bibliographystyle{ieeetr}

\usepackage[flushleft]{threeparttable}

%\usepackage[cp1251]{inputenc}
\usepackage[english]{babel} 
\DeclareMathOperator{\rank}{rank}
\newcommand*{\hm}[1]{#1\nobreak\discretionary{}
            {\hbox{$\mathsurround=0pt #1$}}{}}

            \def\onepc{$^{\ast\ast}$} \def\fivepc{$^{\ast}$}
\def\tenpc{$^{\dag}$}
\def\legend{\multicolumn{4}{l}{\footnotesize{Significance levels
:\hspace{1em} $\dag$ : 10\% \hspace{1em}
$\ast$ : 5\% \hspace{1em} $\ast\ast$ : 1\% \normalsize}}}

\newcommand{\bs}[1]{\boldsymbol{#1}}  
\newcommand{\bsA}{\boldsymbol{A}}

%\setstretch{1}                         
\flushbottom                            
\righthyphenmin=2                      
\pagestyle{plain}                       
%\settimeformat{hhmmsstime}  
\widowpenalty=300                   
\clubpenalty=3000                     
\setlength{\parindent}{0em}           
\setlength{\topsep}{0pt}              
\usepackage[pdftex,unicode,colorlinks=true,urlcolor=blue]{hyperref}
\usepackage{bbm}
\usepackage{tabularx}
\renewcommand{\emptyset}{\varnothing}

\setlength{\parskip}{0.5\baselineskip plus2pt minus2pt}

\newcommand{\e}{\varepsilon}
\DeclareMathOperator*{\Argmax}{\mathrm{Argmax}}
\DeclareMathOperator*{\Argmin}{\mathrm{Argmin}}
\DeclareMathOperator*{\argmax}{\mathrm{arg\,max}}
\DeclareMathOperator*{\argmin}{\mathrm{argmin}}

\newcommand{\blp}{\mathrm{BLP}}
\DeclareMathOperator*{\plim}{\mathrm{plim}}
\DeclareMathOperator{\Max}{\mathrm{Max}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\Y}{\mathcal{Y}}
\newcommand{\Z}{\mathcal{Z}}
\renewcommand{\geqslant}{\geq}
\renewcommand{\leqslant}{\leq}
\newcommand{\p}{\bs p}
\newcommand{\y}{\bs y}
\def\dd#1#2{\frac{\partial#1}{\partial#2}}

\renewcommand{\emptyset}{\varnothing}

\DeclareMathOperator{\tr}{\mathrm{tr}}

\newcommand{\bb}{\bs \beta}
\newcommand{\X}{\bs X}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\PP}{\mathbb{P}}
\DeclareMathOperator{\V}{\mathbb{V}}
\DeclareMathOperator{\CM}{\mathbb{C}}
\renewcommand{\C}{\CM}
\DeclareMathOperator{\var}{\mathrm{var}}
\DeclareMathOperator{\cov}{\mathrm{cov}}
\DeclareMathOperator{\corr}{\mathrm{corr}}
\DeclareMathOperator{\MSE}{\mathrm{MSE}}
\DeclareMathOperator{\Bias}{\mathrm{Bias}}
\renewcommand{\P}{\PP}
\newcommand{\dsim}{\stackrel{d}{\sim}}
\newcommand{\hn}{\mathcal{H}_0}
\newcommand{\ha}{\mathcal{H}_a}
\newcommand{\thetab}{\bs \theta}
\newcommand{\pv}{\text{P-value}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\MLE}{\scriptscriptstyle MLE}
\newcommand{\LR}{\mathrm{LR}}
\newcommand{\I}{\mathbb{I}}
\newcommand{\sumin}{\sum\limits_{i=1}^n}
\newcommand{\sumti}{\sum\limits_{t=0}^\infty}
\newcommand{\hbeta}{\hat{\beta}}
\newcommand{\halpha}{\hat{\alpha}}
\newcommand{\hsigma}{\hat{\sigma}}
\newcommand{\hvar}{\widehat{\var}}
\newcommand{\hcov}{\widehat{\cov}}
\newcommand{\Q}{\mathbb{Q}}

\newcommand{\pconv}{\xrightarrow{ \ p \ }}
\newcommand{\dconv}{\xrightarrow{ \ d \ }}
\newcommand{\asconv}{\xrightarrow{ \ a.s. \ }}
\newcommand{\msconv}{\xrightarrow{ \ m.s. \ }}

\newcommand{\pic}[4][h!]{\begin{figure}[#1]

\begin{center}\includegraphics[width=#2cm]{#3}\caption{#4\label{#3}}\end{center}
\end{figure}}

%outtex
\def\onepc{$^{\ast\ast}$} \def\fivepc{$^{\ast}$}
\def\tenpc{$^{\dag}$}
\def\legend{\multicolumn{4}{l}{\footnotesize{Significance levels
:\hspace{1em} $\dag$ : 10\% \hspace{1em}
$\ast$ : 5\% \hspace{1em} $\ast\ast$ : 1\% \normalsize}}}
%end outtex

%\bibliographystyle{ieeetr}

\newcommand{\laseq}{\stackrel{\lambda\text{-a.e.}}{=}}
\renewcommand{\d}{\underline}
\renewcommand{\u}{\overline}
\newcommand{\td}{\underline{\theta}}
\newcommand{\tu}{\overline{\theta}}
%\renewcommand{\theenumi}{\alph{enumi}}
%\renewcommand{\labelenumi}{(\theenumi)}
%\renewcommand{\theenumii}{\roman{enumii}}
%\renewcommand{\labelenumii}{\theenumii.}
%\renewcommand{\theenumiii}{\arabic{enumiii}}
%\renewcommand{\labelenumiii}{\theenumiii.}
%\renewcommand{\epsilon}{\varepsilon}
\newcommand{\hneq}{\stackrel{\hn}{=}}
\newcommand{\deq}{\stackrel{d}{=}}
%\title{416-2 Final Project\\
%Reproductive Technologies, Aging and Fertility Choice (Proposal)}
%\author{Egor Kozlov}
\begin{document}
%\begin{center}\textbf{Economics 416-1} \\ \emph{By Egor Kozlov}\end{center}
%\maketitle

\begin{abstract}
Unplanned pregnancies often trigger marriage for unmarried cohabitating couples. I show that on average marriages in which a child appeared before or at the year of marriage are characterized with higher divorce probabilities and other outcomes suggesting lower marital quality. This association is robust to different ways of controlling for composition, and it is more pronounced for higher income and more educated couples. I argue that this pattern is consistent with household bargaining models, and the main driver of the result is unequal exposure of spouses to the childcare costs: since being single mother is a bad outcome for an educated female, she is more likely to agree to enter the marriages of lower quality. I estimate a lifecycle model with dynamic bargaining and fertility decisions to show how much of these differences is driven by household bargaining and how childcare costs contribute to observable divorce rates.
\end{abstract}

\section{Introduction}
%Understanding how people make marriage decisions and more generally how household structure is formed is crucial for analyzing labor market outcomes, assets accumulation, various dimensions of inequality and children development outcome. Most of existing studies, even ones that take household structure seriously, typically abstract from the fact that household structure is itself a choice variable. However, existing trends in marriage and divorce rates, increasing share of kids with unmarried parents and in blended families, and, especially, how different these trends are by income and racial groups --- this all suggests economic importance of household issues.

%In particular, economists know pretty little about how people decide on marriage and divorce, despite of the fact that these issues have lots of economic content. Stylized models describing marriage (and, in smaller extent, divorce) emerged in family economics, however most of them describe qualitative issues and do not deliver quantitative predictions that can be applied for analyzing policy or just understanding how different factors contribute to the recent changes. ...]

\subsection{Trends in Marriage And Divorce}

\subsection{Existing Results}

\section{Empirical Evidence}
The key empirical observation is association between marriage timing and divorce probabilities. Statistically, women who have their kids before they are married have significantly higher chance to be divorced. American Community Survey (ACS) data are suitable for establishing this fact only of particular subsample of women: most importantly, those who had no more than one marriage in their life. To re-confirm the patterns I additionally use Survey of Income and Program Participation (SIPP) data: they have smaller sample size but collect exact questions I care about.

\subsection{Defining Variables}

General strategy involves dividing sample in two groups: couples who had kids before their marriage ($K\to M$) and those who had kids after their marriage ($M\to K$). After this division I perform multiple comparisons of these groups: the main comparison is looking at divorce rates in several variations, few additional comparisons involve marriage inequality and outcomes of children.

Although the couples are the primary focus of the paper, the surveys I use look at individuals and households. Therefore I follow females rather than males: upon divorce, children are more likely to stay with mothers (share of single mothers in ACS is ..., single fathers ...). The timing I consider in the  most general form is (for females, given both events occurred): 
\[\Delta T = T^{\text{First child  is born}} - T^{\text{First marriage happened}},\]
where $T$ refers to moments of time. After that I define two groups (given $\Delta T$ is measured in years):
\begin{align*}&M\to K \Leftrightarrow \Delta T \in \{1,2,...\},\\
&K\to M \Leftrightarrow \Delta T \in \{...,-2,-1,0\},\end{align*}

In practice, I exclude observations with $\Delta T < -5$ or $\Delta T > 10$. The reason for the first one is mainly related to step-children: for the subsample when it can be tracked, share of step-children is becomes higher with more negative $\Delta  T$ in ACS, and having step-children that are relatively adult at the moment of marriage is a complex issue that is beyond the scope of the paper. See a note ... in the appendix about this. The reason for the second one is mainly data quality (...).

As the evidence is mainly cross-sectional, I refer to share of divorced people in a group as divorce rates. Unlike fertility, people can change their marriage status all the time, so there is no such thing as completed divorce rate, therefore I just use the share of people who ever had a divorce. More accurate measuring of divorce rates involves conditioning on age, however fixing age restricts the attention to a particular cohort, and pooling people of different ages together allows to think about more general population.

\subsection{Sample Restrictions}
\subsubsection{ACS data}
I take ACS samples of 2008--2016, as only they have variables concerning the duration of current marital status. Namely, people report the year of their most recent marriage. Fertility history is not observed perfectly: the survey reports age of the eldest child residing with individual. However, both of these variables can recover the timing of interest for relatively large subpopulation of females who:
\begin{enumerate}
\item Had exactly one marriage, so they are either married or divorced now (treating legally separated, spouse absent and widowed as married)
\item Have children present in household, with age of the eldest child below 18
\item Most likely reside with their eldest child judging by mother's age (I pick 20--40)
\end{enumerate}
%The last restriction can be relaxed, but it allows to track partners and children more clearly and also insures better quality of self-reported data.

The most important restriction here is being married once --- that means that people who re-marry drop out of sample. For instance, if it take longer to re-marry for the people who had kids before their first marriage, share of divorced people in the group $K\to M$ would be larger without differences in probability of their first marriage to dissolve. This concern can only partially be addressed with ACS data, but this is not an issue in SIPP.

All mean and quantile calculations are performed using person weights.

\subsubsection{SIPP data}

I use wave 1 of 2014 SIPP. It it representative for the whole population, I use person weights in all calculations.

SIPP has much more detailed information on marital history and fertility: it explicitly asks about times of first marriage and first births, it also asks about multi-partner fertility. However it has substantially smaller sample size. For SIPP I relax the restrictions and consider all females who had at least one marriage and are of age 20--40 (treating married with spouses absent, separated and widowed as married), as their marital and fertility history is recovered perfectly.

%When dividing people by income, however, I have to further restrict attention to households that consist of mother + spouse/partner (if reported) + children. This creates some attrition especially for young people, but differences in defining household members are minimal between ACS and SIPP, so it is not likely that it creates an issue.

\subsection{Defining Income}
This part describes how I define income in ACS.

Following development and household literature, when dividing people by income group I use total household income adjusted by OECD equivalence scale. The scale assigns weight 1 to the household head, 0.5 to spouse and 0.3 to each child.

As I use the household income as the main variable of interest, I have to restrict attention to couples that live by themselves and single females with children. I do this both for ACS and SIPP, and as definition of households wary a little, this may drop out slightly different groups of people.

Formally,for income considerations I take a subsample of females who is marked as household's head or head's spouse, and who do not have any people living in household other than their spouse/unmarried partner and their children. This, for instance, excludes multigenerational household and people living with roommates, as total household's income may not be a relevant thing for some types of household composition. I cannot define income equivalence scale for more complicated living arrangements, although there are different scales that allow to do this.

To mitigate selectivity, I compute income percentiles in all sample of females with household composition mentioned above, regardless of their marriage and fertility status. I use total family income as the key variable. To mitigate misreporting I drop observations with total yearly family income below $\$4000$ for ACS (that is about a third of federal the poverty level) or monthly income below $\$300$ for SIPP.

For income percentiles I divide data on cells that consist of the US state, survey year (for ACS) and age group. I define age groups as 20--25, 26--30, 31--35, 36--40, and this classification is based on the age of female. Conditioning for year is done mainly to mitigate consequences of income decline in 2008.

I drop cells with less than 100 observations (for instance, Vermont in ACS 2015 has 31 females in age group 20--25), this constitutes around [2\%] of all observations.

\subsection{Raw Comparisons}
First, I compare divorce probabilities in cross-sections of people between $K\to M$ and $M\to K$ groups. See Table \ref{diff-raw} for this.

\begin{table}[h!]
\begin{center}
\begin{tabular}{|l|c||c|c||c||c|c|}\cline{2-7}
\multicolumn{1}{c|}{} & \multicolumn{3}{c||}{\small ACS: sample of married once} & \multicolumn{3}{c|}{\small SIPP: sample of  ever married} \\\cline{2-7}
\multicolumn{1}{c|}{}  & \multirow{2}{*}{\small \% K$\to$M}& \multicolumn{2}{c||}{\textbf{\% divorced (now)}} & \multirow{2}{*}{\footnotesize \% K$\to$M }& \multicolumn{2}{c|}{\textbf{\% divorced (ever)}} \\\cline{3-4}\cline{6-7}
\multicolumn{1}{c|}{}  &  &\small if M$\to$K &  \small if K$\to$M & & \small if M$\to$K & \small if K$\to$M \\\hline
\textit{All females, 20--40}  & $23.1\%$ & $9.8\%$ & $17.3\%$ & $22.8\%$ & $24.0\%$ & $34.7\%$ \\
\textit{All females, 35--40} & $17.8\%$ & $11.0\%$ & $22.3\%$ & $18.9\%$ & $28.1\%$ & $45.5\%$ \\\hline\hline
\multicolumn{7}{|p{0.9\linewidth}|}{\footnotesize \textbf{Education partition.} Note: ``college graduates'' excludes ``some college'' and ``more than college''.}\\\hline
\textit{High school only, 20--40} & $34.2\%$ & $13.3\%$ & $16.5\%$ & $33.4\%$ & $29.0\%$  & $34.1\%$ \\
\textit{College graduates, 20--40} & $11.5\%$ & $5.7\%$ & $14.3\%$ & $10.5\%$ & $14.9\%$ & $28.1\%$ \\\hline\hline
\multicolumn{7}{|p{0.9\linewidth}|}{\footnotesize \textbf{Income partition:} subsample of household heads/spouses living with children and nobody else.}\\\hline
\textit{Low 30\% income, 20--40} & $33.0\%$ & $17.7\%$ & $22.8\%$ & $29.4\%$ & $34.3\%$ &  $36.0\%$ \\
\textit{Mid 40\% income, 20--40} & $22.8\%$ & $7.6\%$ & $11.8\%$ & $21.9\%$ & $23.8\%$ & $31.1\%$ \\
\textit{Top 30\% income, 20--40} & $11.1\%$ & $4.6\%$ & $13.0\%$ & $12.5\%$ & $11.6\%$  & $29.8\%$ \\\hline
\multicolumn{7}{|p{0.9\linewidth}|}{\footnotesize $\Delta T = \text{Year}^{\text{first child}} - \text{Year}^{\text{first marriage}}$. Restriction: $\Delta T \in \{-5,...,10\}$.}\\
\multicolumn{7}{|p{0.9\linewidth}|}{\footnotesize Definition: K$\to$M if $\Delta T \in \{-5,...,0\}$. M$\to$K if $\Delta T \in \{1,...,10\}$.}\\
\hline
\end{tabular}
\caption{Differences in raw means, ACS and SIPP\label{diff-raw}}
\end{center}
\end{table}

\subsection{Differences in Composition}
This section captures demographic and economic differences of K$\to$M and M$\to$K groups. Many factors could compute to the observable differences, here I quantify them and this justifies use of controls in the next part. I present these differences for ACS, SIPP tables are available in [Appendix ...]. For ACS, since M$\to$K and K$\to$M are defined only for subsample of people who were married once and have children present, I also compare this subsample with more general female population. See Table \ref{diff-contr} for the results.

\begin{table}[h!]
\begin{center}\small
\begin{tabular}{|l|c|c|c|c|}\cline{2-5}
\multicolumn{1}{c|}{} & \multirow{2}{*}{All population} & \multicolumn{3}{c|}{Married once + have kids}\\\cline{3-5}
\multicolumn{1}{c|}{}  & & Everyone & K$\to$M & M$\to$K \\\cline{2-5}\hline
\textit{Age} & 30.2    &  33.1     &       31.7     &       33.6 \\
\textit{Age at first marriage} & 24.2         &   23.3        &    23.6        &    23.2 \\
\textit{Age at first birth}  & 24.0       &     25.4     &       21.7       &     26.5 \\
\textit{Husband's age} & 34.9        &    35.8      &      34.4   &         36.2  \\
\textit{Age difference (W$-$H)} &  2.9       &      2.8      &       3.0      &       2.7  \\\hline\hline
\textit{Eldest child's age} & 8.6         &    7.8    &        10.0     &        7.1 \\
\textit{Number of children} & 1.1       &      2.0      &       2.3     &       2.0 \\\hline\hline
\textit{\% college or + (own)} & 32.1\%   &         39.5\%     &       16.9\%     &       46.3\% \\
\textit{\% college or + (husband)} & 33.3\%     &       36.4\%     &       14.2\%       &     42.7\% \\\hline\hline
\textit{\% In Labor Force} & 75.6\%    &        68.6\%    &        70.0\%        &    68.2\% \\
\textit{Own income (thousands)} & 32.2 &         38.4 &          29.1 &         41.1 \\
\textit{Hours worked} & 36.6       &     36.5    &        36.8        &    36.4\\\hline\hline
\textit{\% white (incl hispanic)} & 69.2\%      &      76.5\%      &      72.2\%      &      77.8\% \\
\textit{\% black} & 14.8\% &        6.7\% &            12.2\%       &       5.0\% \\\hline
\multicolumn{5}{|p{0.7\linewidth}|}{\footnotesize ACS, women, 20--40. Means. Everything is subsamples when applicable, income and hours are for subsample of women who are employed.}\\\hline
\end{tabular}
\caption{Differences in composition, ACS\label{diff-comp}}
\end{center}
\end{table}

\subsection{Statistical Tests and Controlling for Composition}

In this section I formally test differences in shares for different subgroups. I also attempt to control for composition using regressions with many fixed effects. I show that generally observable mean differences are not likely to be attributed to composition, and introducing various controls still leaves substantial part of the difference.

In general, I perform the following exercise: I regress indicator of female being divorced on set of controls $X$ and dummy variable $\I(\text{K$\to$M})_i$. The specification I use is
\[D_i = \Delta \cdot \I(\text{K$\to$M})_i + X_i\gamma + \varepsilon_i.\]
Replacing $X_i$ with a constant corresponds to raw mean comparison described above.

I report three possible estimates of $\Delta$:
\begin{itemize}
\item \textbf{Raw}, corresponding to differences in raw means in the previous part
\item \textbf{Controls}, where the regression includes dummies for ACS year, the US state, number of children, years after marriage, age of the eldest child, race and education group + 4th degree polynomial in woman's age
\item \textbf{Controls+income}, where the regression includes all of the above + dummies for income groups (low 30\%, mid 40\%, top 30\%) on a subsample where income equivalence scale is defined. This subsample is smaller, but has arguably better data quality.
\end{itemize}

See Table \ref{diff-contr} for the results.

\begin{table}
\begin{center}
\begin{tabular}{|l|c||c|c|c|c|c|c|}\cline{2-8}
\multicolumn{1}{c|}{} & \multicolumn{7}{c|}{ACS: sample of married once}\\\cline{2-8}
\multicolumn{1}{c|}{} & \small Share of &\multicolumn{2}{c|}{\small Raw difference} & \multicolumn{2}{c|}{\small with controls} & \multicolumn{2}{c|}{\small controls + income} \\ \cline{3-8}
\multicolumn{1}{c|}{} & \small divorced & $\Delta$ & (s.e.) & $\Delta$ & (s.e.) & $\Delta$ & (s.e.) \\  \hline
\textit{All females, 20--50} & 13.0\% & 7.6\% & (0.1\%) & 5.5\% & (0.2\%) & 5.1\% & (0.2\%) \\
\textit{All females, 35--40} & 16.0\% & 11.3\% & (0.2\%) & 6.5\% & (0.3\%) & 5.9\% & (0.3\%) \\\hline\hline
\multicolumn{8}{|p{0.9\linewidth}|}{\footnotesize \textbf{Education partition.} Note: ``college graduates'' excludes ``some college'' and ``more than college''.}\\\hline
\textit{High school only, 20--50} & 16.2\% & 3.2\% & (0.2\%) & 8.0\% & (0.3\%) & 7.6\% & (0.3\%) \\
\textit{College graduates, 20--50} & 8.3\% & 8.6\% & (0.3\%) & 3.9\% & (0.3\%) & 3.5\% & (0.3\%) \\\hline\hline
\multicolumn{8}{|p{0.9\linewidth}|}{\footnotesize \textbf{Income partition:} subsample of household heads/spouses living with children and nobody else.}\\\hline
\textit{Low 30\% income, 20--40} & 20.3\% & 5.1\% & (0.2\%) & 10.3\% & (0.3\%) & \multicolumn{2}{c|}{---}\\
\textit{Mid 40\% income, 20--40} & 10.1\% & 4.3\% & (0.2\%) & 2.8\% & (0.2\%) & \multicolumn{2}{c|}{---}\\
\textit{Top 30\% income, 20--40} & 9.0\% & 8.3\% & (0.3\%) & 2.7\% & (0.4\%) & \multicolumn{2}{c|}{---}\\
\hline
\end{tabular}
\caption{Differences, controlled for composition, ACS\label{diff-contr}}
\end{center}
\end{table}
\subsection{Additional Comparisons}

[I pick recently divorced people in ACS, I compare marriage durations for step and non-step children. I also check share of divorced in $M\to K$ and $K\to M$ given child's age and given marriage duration.]

\subsection{Intra-Household Allocations}

[I look at earnings, hours and LFP differences] [Grade retention depends on child's age, so need to compare given child's age]
\section{Model}
The model is a mixture between conventional heterogeneous agents savings model (B--A--H framework) and household formation and dynamic bargaining, that is extensively developed by [...].

Agents live for $T$ periods. The agents are males, females and married couples. Couples can have a child together, and if a couple with a child divorces up, the child stays with mother, and this is how singe females can have kids. Single females with kids randomly re-enter marriage market by becoming single females without kids.\footnote{The reason for this assumption is to avoid modeling mixture between step and own children, and the fact that single mothers are not a main focus of the paper.}

Single people without kids randomly meet partners with characteristics similar to them and decide whether to create a couple. Couples agree on some marriage terms, represented by each spouse's weight in couple's objective function. The rationale for marriage includes risk-sharing, returns to scale in consumption, possibility to have a child in the future and additional utility surplus of being married (love shock). Each period after random shocks are realized, couples may decide to renegotiate their marriage terms if one of the spouses is better off being single. If they cannot find marriage terms that are satisfactory for both, divorce happens.

Couples can choose to have a child and can also experience unplanned pregnancies. Unplanned pregnancies cannot be aborted. In addition to couples, single people who met each other may experience an unplanned pregnancy just before they decide whether to create a couple. In this case, upon disagreement female becomes a single mother. This mechanism is a stylized representation of kids appearing in unmarried cohabitating couples: since I do not model cohabitation, those kids appear just before people actually decide to marry each other.\textit{ I discuss this issue additionally in Section ....}

\subsection{Remarks on Notation}
I write $V^{i,j}$ to represent individual value functions in the model. Here $i$ represents gender: $f$ for female or $m$ for male. $j$ represents demographic status: $s$ for singles without kids, $sk$ for singles with kids (females only), $c$ for couples without kids and $ck$ for couples with kids. Couple's joint value functions have only one superscript $j \in \{c,ck\}$. Relation between joint value function $V^{c}$ and each spouse's value functions $V^{m,c}$ and $V^{f,c}$ is described in details in couple's section.

\subsection{Singles Without Kids}
Singles without kids are characterized by age $t$, labor productivity $z$ and assets (savings) $a$. I use $\omega = \{z,a\}$ to individual characteristics other than age.

Each period single individuals meet a partner of the same age and characteristics $\omega^p$ with probability $p^{\text{meet}}_t$. If meeting happens, with probability $p^{\text{preg}}_t$ unplanned pregnancy happens. After realizing new period's shocks and pregnancy status couple tries to negotiate marriage terms. If they found marriage terms that are satisfactory for both ($m^{np} = 1$ in case of no pregnancy or $m^p = 1$ in case of pregnancy), they form couple with characteristics $\Omega^c$ . If satisfactory marriage terms could not be found, the individual stays single, or becomes single mother if she is female and unplanned pregnancy happened.

The value function of male agent therefore is
\begin{align}V^{m,s}_t(\omega) = \max\limits_{c} & \bigg\{ u(c) + \beta \E_t \Big[ (1 - p^{\text{meet}}_t)\cdot V^{m,s}_{t+1}(\omega') + \\ \nonumber
& \hspace{7em} p^{\text{meet}}_t (1-p^{\text{preg}}_t) \big\{ m^{np} \cdot V^{m,c}_{t+1}(\Omega^c) + (1-m^{np})V^{m,s}_t(\omega')\big\} + \\  \nonumber
& \hspace{10em} p^{\text{meet}}_t p^{\text{preg}}_t \big\{ m^{p} \cdot V^{m,ck}_{t+1}(\Omega^c) + (1-m^{p})V^{m,s}_{t+1}(\omega')\big\}  \Big]  \bigg\},\\  \nonumber
 \end{align}\vspace{-3em}
 \begin{align*}
 \text{s.t. \ }  &  a' = R\cdot a  + W^m_t(z) - c  & \text{ (evolution of assets)}\\
 &  z' = z + \varepsilon^{z,m}_t, \ \ \varepsilon^{z,m}_t \sim \mathcal{N}(0;\sigma_{z,m}^2) &  \text{ (evolution of productivity)}\\
  & \log W^m_t = z_t + \text{Trend}^m_t, \ \ \  \text{Trend}^m_t = a^m_0 + a^m_1\cdot t  +  a^m_2 \cdot t^2 &  \text{ (labor income and trend)}\\
  & \Omega^c = \mathcal{M}^m(a',z') &  \text{ (marriage prospectives)}
\end{align*}
where random function $\mathcal{M}^m$, representing characteristics of potential couple is defined in Section [...].

The value function of female agent is
\begin{align}V^{f,s}_t(\omega) = \max\limits_{c} & \bigg\{ u(c) + \beta \E_t \Big[ (1 - p^{\text{meet}}_t)\cdot V^{f,s}_{t+1}(\omega') + \label{single-fem} \\  \nonumber
& \hspace{7em} p^{\text{meet}}_t (1-p^{\text{preg}}_t) \big\{ m^{np} \cdot V^{f,c}_{t+1}(\Omega^c) + (1-m^{np})V^{f,s}_t(\omega')\big\} + \\  \nonumber
& \hspace{10em} p^{\text{meet}}_t p^{\text{preg}}_t \big\{ m^{p} \cdot V^{f,ck}_{t+1}(\Omega^c) + (1-m^{p})V^{f,sk}_{t+1}(\omega')\big\}  \Big]  \bigg\},
\end{align}\vspace{-3em}
\begin{align*}
 \text{s.t. \ }  &  a' = R\cdot a  + W^f_t(z) - c  & \text{ (evolution of assets)}\\
 &  z' = z + \varepsilon^{z,f}_t, \ \ \varepsilon^{z,f}_t \sim \mathcal{N}(0;\sigma_{z,f}^2) &  \text{ (evolution of productivity)}\\
  & \log W^f_t = z_t + \text{Trend}^f_t, \ \ \  \text{Trend}^f_t = a^f_0 + a^f_1\cdot t  +  a^f_2 \cdot t^2 &  \text{ (labor income and trend)}\\
  & \Omega^c = \mathcal{M}^f(a',z') &  \text{ (marriage prospectives)}
\end{align*}

% 
The crucial difference here is the term $V^{f,sk}$: it reflects the fact that upon disagreement woman becomes a single female with a child.

\subsection{Couples Without Kids}
Couple is characterized by age $t$ (assumed the same for both spouses), marriage terms $\theta$, productivities of female and male $z^f$, $z^m$ and additive marriage surplus $\psi$. Couple maximizes weighted expected lifetime utility, and $\theta$ represents share of female in this objective function, therefore $\Omega = \{\theta,\psi,z^f,z^m\}$ represent characteristics of couple.

The key element shaping couple's decisions is participation constraints: each period the expected lifetime utility of staying in couple should be not less that expected lifetime utility of getting a divorce and becoming a single agent. Depending on realization of the shocks, if it is possible to satisfy both participation constraints the next period (so no divorce happens, $d = 0$) state is $\Omega'$. If participation constraints cannot be satisfied, the couple gets a divorce $d = 1$ and its characteristics are dissolved onto $\omega^{df}$ and $\omega^{dm}$.  In more details this is described in Section [...].

Fertility dimension is an important element. Each period couples get fertility shock $p^{\text{preg}}_t$. If it happens, they become couple with a child if they stay together (renegotiation happens after they know realization of the shock), or become single male and single female with a child if they do not. If the shock does not happen, they still may make a decision to get a baby \emph{after} they renegotiate the marriage terms if they stay together. This timing is subtle, but is required for everything to be internally consistent, see Appendix ... for more details on how timing interacts with renegotiation.

After realization of all the shocks and making fertility and divorce decisions in the current period the problem of the couple that stays childless is
\begin{align}& \hspace{5em}  V^{c}_t(\Omega) = \max\limits_{c^f,c^m}  \bigg\{ \theta\cdot u(c^f) + (1-\theta)\cdot u(c^m) + \psi +  \label{vf-c} \\   \nonumber
 &  \beta \E_t \Big[   (1 - p^{\text{preg}}_t)\cdot \left\{ (1-d^{\text{np}})\cdot \max\left\{ V^{ck}_{t+1}(\Omega'),V^{c}_{t+1}(\Omega')\right\} + d^{\text{np}}\cdot [ \theta V_{t+1}^{f,s}(\omega^{df}) + (1-\theta)V_{t+1}^{m,s}(\omega^{dm})]\right\}  +  \\  \nonumber
& \hspace{5em} p^{\text{preg}}_t\cdot \left\{ (1-d^{\text{p}})\cdot V^{ck}_{t+1}(\Omega') + d^{\text{p}}\cdot [ \theta V_{t+1}^{f,sk}(\omega^{df}) + (1-\theta)V_{t+1}^{m,s}(\omega^{dm})]\right\} \Big] \bigg\},
\end{align}\vspace{-2em}
\begin{align*}
\text{s.t. \ }& a' + c = R\cdot a  + W^m_t(z^m) + W^f_t(z^f) & \text{(evolution of joint assets)},\\
				 & c = [(c^f)^{1+\rho_c} + (c^m)^{1+\rho_c}]^{\frac1{1+\rho_c}} & \text{(increasing returns in consumption)},\\
				 &  z^{f\prime} = z^f + \varepsilon^{z,f}_t, \ \ \varepsilon^{z,f}_t \sim \mathcal{N}(0;\sigma_{z,f}^2) &  \text{ (evolution of female productivity)}\\
				 &  z^{m\prime} = z^m + \varepsilon^{z,m}_t, \ \ \varepsilon^{z,m}_t \sim \mathcal{N}(0;\sigma_{z,m}^2) &  \text{ (evolution of male productivity)}\\
                    & \psi' = \psi + \varepsilon^{\psi}_t, \ \ \varepsilon^{\psi}_t \sim \mathcal{N}(0;\sigma_{\psi}^2)  & \text{(evolution of marriage surplus),} \\
                    & (\Omega',\omega^{df},\omega^{dm}) = \mathcal{R}(\theta,a',z^{m\prime},z^{f\prime},\psi') & \text{(renegotiation correspondence)},
\end{align*}

\subsubsection{Defining Individual Values}
The couple's collective value function $V^c$ is the object relevant for making decisions of consumption and fertility. Making marriage and divorce decisions involves individual value functions of spouses living in the couple. I denote them as $V^{f,c}$ and $V^{m,c}$. These value functions are obtained by plugging optimal decisions of couple into intertemporal utilities of each agent and accounting for possible transitions. Note that they cannot be derived from a maximization problem, and common properties of value functions such as envelope theorems do not hold for them.

Namely, given couple's decisions $c^f(\Omega)$, female's value of being in couple is defined recursively as
\begin{align}
& \hspace{5em}  V^{f,c}_t(\Omega) =    u(c^f) + \psi +  \\   \nonumber
 &  \beta \E_t \Big[   (1 - p^{\text{preg}}_t)\cdot \left\{ (1-d^{\text{np}})\cdot \max{}^C \left\{ V^{f,ck}_{t+1}(\Omega'),V^{f,c}_{t+1}(\Omega')\right\} + d^{\text{np}}\cdot [ V_{t+1}^{f,s}(\omega^{df})]\right\}  +  \\  \nonumber
& \hspace{5em} p^{\text{preg}}_t\cdot \left\{ (1-d^{\text{p}})\cdot V^{f,ck}_{t+1}(\Omega') + d^{\text{p}}\cdot [ V_{t+1}^{f,sk}(\omega^{df}) ]\right\} \Big] 
\end{align}
where operator $\max{}^C\{A,B\} \equiv A\cdot \I[V^{ck}_{t+1}(\Omega')\geq V^{c}_{t+1}(\Omega')] + B\cdot \I[V^{ck}_{t+1}(\Omega')< V^{c}_{t+1}(\Omega')]$. It reflects the fact that couple makes joint decision that is not necessary individually optimal.

Another important property of individual values is that
\begin{equation} V^{c}_t(\Omega) \neq \theta\cdot V^{f,c}_t(\Omega) + (1-\theta)\cdot V^{m,c}_t(\Omega),\label{tht_noneq}\end{equation}
the main reason for this is that $\theta$ changes in the future as a result of random shocks, and this change is not likely to be symmetric. This, in particular, causes couple's expected future value function to be discontinuous in $(a,z)$. See Appendix [...] for some additional discussion.

\subsection{Couples With Kids}
State space for couples is the same as for singles, except for additional variable $\xi$ representing child's age, therefore $\Omega = \{\theta,\psi,z^f,z^m,\xi\}$. I consider just two states for $\xi \in \{y,s\}$ representing young and school-age child. At birth the child is young, then each period with probability $p^{s}$ the child becomes of school-age. The difference between young and school-age children are mother's time requirement, namely young children require more time and elder --- more money.

Couples with kids have additional choice variable $q$, that can be interpreted as flow of child quality or child consumption. Child quality is produced using mother's labor $l_f$ and monetary expenditures $x$ according to constant returns to scale production function. Labor input to child quality causes female to lose part of her labor income. The production function depends on age, namely I parametrize it as $q = [\mu_\xi\cdot l_f^{\theta_q} + (1-\mu_\xi)\cdot x^{\theta_q}]^{\frac1{\theta_q}}$. This follows, for instance, [Sommer ...], with the addition of changing labor share $\mu$.

\begin{align}& \hspace{5em}  V^{ck}_t(\Omega) = \max\limits_{c^f,c^m,q,l_f,x}  \bigg\{ \theta\cdot u(c^f,q) + (1-\theta)\cdot u(c^m,q) + \psi + \label{vf_ck} \\  \nonumber
 & \hspace{10em} \beta \E_t \Big[   \left\{ (1-d)\cdot   V^{ck}_{t+1}(\Omega') + d\cdot [ \theta V_{t+1}^{f,sk}(\omega^{df}) + (1-\theta)V_{t+1}^{m,s}(\omega^{dm})]\right\} \Big] \bigg\},
\end{align}\vspace{-2em}
\begin{align*}
\text{s.t. \ } & a' + c + x = R\cdot a  + W^m_t(z^m) + (1-l_f)\cdot W^f_t(z^f) & \text{(evolution of joint assets)},\\
                    & c = [(c^f)^{1+\rho_c} + (c^m)^{1+\rho_c}]^{\frac1{1+\rho_c}} & \text{(increasing returns in consumption)},\\
                    & q = [\mu_\xi\cdot l_f^{\theta_q} + (1-\mu_\xi)\cdot x^{\theta_q}]^{\frac1{\theta_q}} & \text{(production function of child quality)},\\
                    &  z^{f\prime} = z^f + \varepsilon^{z,f}_t, \ \ \varepsilon^{z,f}_t \sim \mathcal{N}(0;\sigma_{z,f}^2) &  \text{ (evolution of female productivity)}\\
				 &  z^{m\prime} = z^m + \varepsilon^{z,m}_t, \ \ \varepsilon^{z,m}_t \sim \mathcal{N}(0;\sigma_{z,m}^2) &  \text{ (evolution of male productivity)}\\
                    & \psi' = \psi + \varepsilon^{\psi}_t, \ \ \varepsilon^{\psi}_t \sim \mathcal{N}(0;\sigma_{\psi}^2)  & \text{(evolution of marriage surplus),} \\
                   &  \P(\xi' = s | \xi = y) = p^s, \ \ \P(\xi' = s | \xi = s) = 1 & \text{(evolution of child's age)},\\
                    & (\Omega',\omega^{df},\omega^{dm}) = \mathcal{R}(\theta,a',z^{m\prime},z^{f\prime},\psi',\xi) & \text{(renegotiation correspondence)},
\end{align*}

\subsection{Single Mothers}
State space for single mothers contains individual productivity and child's age. Single mothers cannot marry, but have chance $p^{\text{out}}$ of recovering their marriage prospectives. In the model this is captured by transition of them to single females. This form is somewhat restrictive, although any other option requires modeling marriage of single mothers separately. This is a complex issue, for instance, males would seek females who already have kids if we do not make difference between step and own children. Since for the purposes of the model option of being single mother is used mainly as a factor affecting bargaining, I introduce additional (constant) utility factor $S$ to flexibly control utility of exiting the state. This can be interpreted as equivalent of utility of kids from past marriages when entering the marriage market again. This form seems flexible enough to capture variation in an option of being a single mother without putting extra computational and modeling burden.

The problem of a female who stays single mother is
\begin{align}V^{f,sk}_t(\omega) = \max\limits_{c} & \bigg\{ u(c,q) + \beta \E_t \Big[ p^{\text{out}} \cdot \left\{ V^{f,s}_{t+1}(\omega') + S\right\} +(1-p^{\text{out}})V^{f,sk}_{t+1}(\omega') \Big]  \bigg\},
\end{align}\vspace{-1em}
\begin{align*}
 \text{s.t. \ }  &  a' = R\cdot a  + (1-l_f)W^f_t(z) - c  & \text{ (evolution of assets)}\\
 & q = [\mu_\xi\cdot l_f^{\theta_q} + (1-\mu_\xi)\cdot x^{\theta_q}]^{\frac1{\theta_q}} & \text{(production function of child quality)},\\
 &  z' = z + \varepsilon^{z,f}_t, \ \ \varepsilon^{z,f}_t \sim \mathcal{N}(0;\sigma_{z,f}^2) &  \text{ (evolution of productivity)}\\
 &  \P(\xi' = s | \xi = y) = p^s, \ \ \P(\xi' = s | \xi = s) = 1 & \text{(evolution of child's age)},\\
\end{align*}

\subsection{Marriage Market}
Single males and females meet potential partners of identical age and characteristics that depend on their assets $a$ (chosen in the previous period) and productivity $z$ (revealed after realization of shock):
\begin{equation}\log a^p = \log a + \varepsilon^{a,p}, \ \ \varepsilon^{a,p} \sim \mathcal{N}(0,\sigma_{a,p}^2),\end{equation}
\begin{equation}	z^p = z + \varepsilon^{z,p}, \ \ \varepsilon^{z,p} \sim \mathcal{N}(0,\sigma_{z,p}^2),\end{equation}
for example, $\sigma_{a,p} = 0.1$ can be read as standard deviation of partner's assets being 10\% around of what agent has. 
Additionally, initial marriage surplus (love shock) is drawn from
\begin{equation} \psi \sim \mathcal{N}(0,\sigma_{\psi,0}^2).\end{equation}

If people agree to be a couple they pool their assets $a^c = a + a^p$. If no unplanned pregnancy happens, people agree to marry if set of mutually satisfactory marriage terms
\begin{equation}\small \Theta^{np}_t = \left\{ \theta : V_t^{f,c}(\theta,\psi,a^c,z^f,z^m) \geq V_t^{f,s}(a^f,z^f), \ \ V_t^{m,c}(\theta,\psi_0,a^c,z^f,z^m) \geq V_t^{m,s}(a^m,z^m)\right\}\end{equation}
is non-empty. If potential couple gets the pregnancy shock, the set changes to
\begin{equation} \small \Theta^{p}_t = \left\{ \theta : V_t^{f,ck}(\theta,\psi,a^c,z^f,z^m,y) \geq V_t^{f,sk}(a^f,z^f,y), \ \ V_t^{m,ck}(\theta,\psi_0,a^c,z^f,z^m,y) \geq V_t^{m,s}(a^m,z^m)\right\} \end{equation}
(at birth child's age is $\xi = y$).

If the respective set is non-empty, initial bargaining power of couple is determined by symmetric Nash Bargaining, for instance:
\small
\begin{align}&\theta^*(a,z,\psi,\epsilon^{a,p},\epsilon^{z,p}) = \label{nbs} \\ \nonumber &= \argmax\limits_{\theta \in \Theta^{np}} \left[V_t^{f,c}(\theta,\psi,a^c,z^f,z^m) - V_t^{f,s}(a^f,z^f)\right] \times \left[V_t^{\vphantom{f}m,c}(\theta,\psi,a^c,z^f,z^m) - V_t^{m,s}(a^m,z^m)\right],\end{align}
and analogously for the case when pregnancy shock happened.

From the prospective of a single agent, this set is determined by realization of $\psi$ and partner's characteristics $\epsilon^{a,p}$ and $\epsilon^{z,p}$. Given them, marriage function $m$ is binary:
\[m^{np}_t(a,z,\psi,\epsilon^{a,p},\epsilon^{z,p}) = \I(\Theta^{np}_t \neq \varnothing),\]
where variables affecting set $\Theta$ are unambiguously recovered from arguments of $(a,z,\psi,\epsilon^{a,p},\epsilon^{z,p})$ as described above.

%Because of good properties of Nash Bargaining, expected future value function is continuous with respect to of $a$ and $z$. See Appendix [...] for additional discussion of this.

Let $\omega = (a,z)$ and $\epsilon = (\psi,\epsilon^{a,p},\epsilon^{z,p})$. This allows to define a random function
\[\Omega^c = \mathcal{M}_t(\omega) = M_t(\omega,\epsilon) = (\theta^*,\psi,a^c,z^m,z^f),\]
so the future characteristics of couple are function of current characteristics of individual and three-dimensional random shock.
\subsection{Renegotiation And Divorce}

Upon divorce, $\kappa\cdot a$ of assets disappears and the rest is split evenly. This generalizes [Voena ...], where I put additional parameter $\kappa$ captures can rationalize extra costs of divorce for more wealthy couples. [Voena ...] argues that this approximates well actual property divisions. So, the splitting rule is
\[a^{df} = 0.5\cdot (1-\kappa)\cdot a,  \ \ a^{dm} = 0.5\cdot (1-\kappa)\cdot a.\]
This splitting rule defines mapping of couple's state $\Omega$ to individual states $\{\omega^{df},\omega^{dm}\}$ in case of divorce.

The option to leave with a share of couple's assets creates pressure for participation constraints. As argues by [...], the household prefers to keep $\theta$ constant. However, after shocks to $z$ and $\psi$ happen, one (or both) participation constraints might be violated. To address this formally, consider couple that already has kids. I define set
\begin{equation}\Theta_t = \left\{ \tilde\theta : V^{f,ck}_t(a,\tilde\theta,\psi,z^f,z^m,\xi) \geq V_t^{f,sk}(a^{df},z^f,\xi) , \ \ V_t^{m,ck}(a,\tilde\theta,\psi,z^f,z^m,\xi)\geq V_t^{m,s}(a^{dm},z^m) \right\} \label{reneg-with-kids}
,\end{equation}
and if couple starts with current value $\theta_{0}$ there are three possible options:
\begin{enumerate}
\item \textit{Status quo.} If $\theta_0 \in \Theta_t$, current marriage terms are satisfactory for both spouses so $\theta_t = \theta_{0}$.
\item \textit{Renegotiation.} If $\theta_0 \not\in \Theta_t$, but $\Theta_t \neq \varnothing$, then spouses can continue to be married if they change their marriage terms. In this case they pick the closest the new value from set $\Theta_t$ such that it is the closest to the old marriage terms: $\theta_t = \min\limits_{\tilde\theta\in\Theta_t} |\tilde\theta - \theta_0|$. 
\item \textit{Divorce.}  If $\Theta_t = \varnothing$, so if there do not exist any marriage terms satisfactory for both spouses, then divorce happens according to the rule described above.
\end{enumerate}
Picking $\theta$ that is the closest to the old one is not a random choice: it is rationalized by models with limited commitment models, where change in $\theta$ represents the value of Lagrange multiplier on binding participation constraint. Intuitively, since couple does not like $\theta$ to be changing and everything is continuous in $\theta$, the couple prefers small changes to larger ones, and this drives the result.

Renegotiation that involves making discrete decisions has few more issues as backward induction has to be used: fertility decisions have to be made after new $\theta$ is set, and these decisions might depend on $\theta$, therefore inequalities defining set $\Theta_t$ need to be modified. This is an artifact of assuming that fertility decisions are made collectively and it can be the case that having a baby benefits couple but hurts wife alone (mainly by worsening her outside option), and such scenario of renegotiation insures that participation constraints hold after the fertility decisions are made. See Appendix \ref{ren-disc} where I elaborate on this.

\subsection{Additional Details}
The model I solve an estimate contains few additional things that are omitted for more clear exposition above.

First, females may exit the state of being single mother immediately. This transition is assumed to happen after divorce or failure to agree to marry, therefore in the value functions and in (re)negotiation I use $V^{f,s?} = p^{\text{out}} [V^{f,s} + S] + (1-p^{\text{out}})V^{f,sk}$ instead of just $V^{f,sk}$. This allows for quicker re-marriage and easier implementation of counterfactual in which females do not have risk of becoming single mother at all. I use notation of $V^{f,s?}$ in Appendix \ref{add-trans} where I show exact value functions.

Second, I assume that having a baby is possible only before certain age $\bar{T}$. After that, value functions and negotiation rules are modified in a straightforward manner.

Finally, to simplify computations I also do not allow marriage status and marriage terms to change after certain $\bar{T}_2$. Therefore I drop participation constraints in the couples, the value functions look the same but do not allow for $\theta$ to change.

\subsection{Functional Forms}
I use standard CES utility functions for individuals and couples with no kids: $u(c) = \frac{c^{1-\sigma}}{1-\sigma}$.  When kids appear, people extract additional utility of child quality, I use the following non-separable utility function:
\[u(c,q) = \frac{\left(e^{\phi_0}\cdot c^{1-\phi_1} \cdot (\phi_2 + q)^{\phi_1}\right)^{1-\sigma}}{1-\sigma},\]
parameter $\phi_0$ represents fixed gain in marginal utility because of having a baby, $\phi_1$ drives relative importance of child quality in utility function and $\phi_2$ is a ``luxury good'' parameter that can rationalize large child expenditures of wealthy couples that are discussed in the estimation section.

\section{Estimation Strategy}
I use techniques of indirect inference, extensively described, for instance, in [...] for estimation of the parameters. Shortly, I select parameters such that features of the real data are similar to the features of the data simulated from model. It is not entirely correct to call the procedure simulated method of moments, as some things that I match are not moments in econometric sense, however it has many similarities with simulated method of moments so I occasionally extend notion of moments to reflect any statistics derived from the data.

\subsection{Overview}
Full list of parameters involves many things:
\begin{itemize}
\item Fundamental parameters of income fluctuation model $(\beta,\sigma,R)$
\item Intra-household returns to scale $\rho_c$
\item Preferences for child's quality: $(\phi_0,\phi_1,\phi_2)$
\item Parameters for child's quality production function: CES power $\theta_k$ and labor shares by age $(\mu_y,\mu_s)$.
\item Transition probabilities:
\begin{itemize}
\item Probability of pregnancy shock by age: $p^{\text{preg}}_t = \lambda^{\text{preg}}_0 - \lambda^{\text{preg}}_1\cdot t$ (2 parameters)
\item Probability of meeting a partner: $p^{\text{meet}}_t:$ piecewise linear function between $(p^{\text{meet}}_{21},p^{\text{meet}}_{27},p^{\text{meet}}_{34},p^{\text{meet}}_{40})$ (4 parameters)
\item Probability of re-entering marriage market and corresponding utility ``compensation'' $(p^{\text{out}},S)$. (2 parameters)
\item Probability that kids grow up (transition from $y$ to $s$): $p_s$ (1 parameter)
\end{itemize}
\item Age profiles of labor income by gender: $(a^f_0,a^f_1,a^f_2)$, $(a^m_0,a^m_1,a^m_2)$ (6 parameters)
\item Variance of shocks (6 parameters):
\begin{itemize}
\item Individual income shocks $\sigma^{z,f}$, $\sigma^{z,m}$.
\item Couple's surplus shocks $\sigma^{\psi}$
\item Couple's initial surplus $\sigma^{\psi_0}$
\item Assets and productivity of potential partner: $(\sigma^{z,p},\sigma^{a,p})$
\end{itemize}
\end{itemize}

Total number of parameters here is 31. Few additional choices involve time horizons, bounds for random walk variables and bounds for assets and $\theta$.

\subsection{Externally Set Parameters}
Length of the period is set to be 1 year, the model assumes people start their life with no assets at the age of $21$ and die deterministically at the age of $81$ (so $T = 30$). However, it assumes that fertility status can change only till age of 40 ($T_f = 19$) and marriage status till age of 50 ($T_m = 29$). It also assumes that after the age of 60 people retire and get $b = 0.4$ [as in Sommers, 2014] of their income at age of 60 (so no transitions happen). I stop simulations at the age of 50.

Following standard approach in literature, I fix $\sigma = 1.5$, $R = 1.03$ and $\beta = 1/R$. Also, following estimates of Voena (2014) I set $\rho_c = 0.23$.

I borrow age profiles from $(a^f_0,a^f_1,a^f_2)$, $(a^m_0,a^m_1,a^m_2)$ and variances of income shocks $\sigma^{z,f}$, $\sigma^{z,m}$ from Low et al 2018 [This will be replaced by my own estimates from SIPP].

I also externally estimate $\sigma^{z,p}$ as variance of (unpredicted part of) log-difference in spouses earnings, see Appendix \ref{partearn} for this. Based on this calculations I set $\sigma^{z,p} = 0.8$. As I both do not have good assets data and do not aim to replicate assets distribution, I set $\sigma^{a,p} = \sigma^{z,p}$.

I also set $p_s = \frac14$ to roughly reflect that children go to school on average at 6 and to pre-school few years earlier and this should correspond to a change in childcare costs structure.

In the part with assets, I assume savings can be between 0 and 20 times average labor income in the beginning of life. This bar in pretty low, but it is enough to create reasonable assets variation at the moment most of marriage decisions are made: say, median assets-to-income ratio at 30 is around 1 [see ...].

\subsection{Estimated Parameters}
Out of 31, 15 parameters are set externally, so I am left with $16$ parameters to estimate: child's parameters $(\phi_0,\phi_1,\phi_2)$, child quality production parameters $(\theta_k,\mu_y,\mu_a)$, variances of initial marriage surplus and its innovations $(\sigma^{\psi_0},\sigma^{\psi})$, meeting probability parameters $(p^{\text{meet}}_{21},p^{\text{meet}}_{27},p^{\text{meet}}_{34},p^{\text{meet}}_{40})$, pregnancy probability parameters $(\lambda^{\text{preg}}_0,\lambda^{\text{preg}}_1)$, transition out of being single mother parameters $p^{\text{out}}$, $S$. Unfortunately, all of them interact and as the model is highly stylized but involves many choices, they cannot be estimated externally.

Conceptually, they can be split onto following groups:
\begin{enumerate}
\item Unplanned pregnancy probabilities are pinned down by share of $K\to M$ females at ages 25 and 35, as well as share of planned births in total population
\item Meeting probabilities and $\sigma^{\psi_0}$ are pinned down by shares of never married females people at 25, 35, share of females with one marriage at 40 (relevant to capture chances of re-marrying) and shares of never married at 30 for high and low income groups.
\item Childcare production function is recovered by ratio of time spent with kids by couple's income (pins down elasticity of substitution), average share of money spent on kids (captures ``average'' $\mu$) and ratio of time spent by child's age (captures difference between $\mu_a$ and $\mu_y$)
\item  Preferences for kids are captured by share of childless females by age of 25, 30, 35 (general scale), share of childless females for low and high income at 30 $\phi_1$. Luxury good parameter $\phi_2$ is pinned down by ratio of amounts of money spent on kids between 75 and 25 income quantiles.
\item Variance of innovation to marriage surplus is mainly pinned down by share of divorced people by age (25, 35, 40) and income groups (at 30).
\item Transition out of single mothers is captured by share of never married single mothers at 25 (picks $S$) and share of divorced women with kids at 30 (picks $p^{\text{out}}$).
\end{enumerate}
On top of it, I use shares of divorced people for $K\to M$ and $M \to K$ at 40 as additional targets that are affected by mixture of parameters described above.

I additionally impose the following constraints: share of rejected marital offers is within $[0.05,0.95]$ and average share of time spent on kids is below $0.8$. [Both of these constraints are non-binding in the solution I found, although turning them off occasionally may lead to implausible solutions.]

\section{Results}
Table ... shows targeted and

\subsection{Counterfactual Simulations}
\begin{itemize}
\item Turn off fertility completely
\item Turn off bargaining
\item No pressure of being single mother
\item Single mother/single father randomly
\end{itemize}

\newpage
\section{Robustness Checks}
\section{Conclusions}

\appendix
\section{Additional Empirical Results}
\subsection{ACS}

\section{Additional Robustness Checks}

\section{Additional Notes on Model}
\subsection{Continuity of Value Functions}
This part discusses violations of continuity caused by the dynamic bargaining structure. The general conclusion is that collective value function does is not necessary continuous in couple's characteristics, and this discontinuity propagates onto all value functions in the model.

To simplify the example, let's consider value function of couple with kids. Similarly to equation \ref{tht_noneq}, we can write
\[ V^{ck}_t(\Omega) = \theta\cdot V^{f,ck}_t(\Omega) + (1-\theta)\cdot V^{m,ck}_t(\Omega) + \Delta(\Omega),\]
and use it to substitute the term under the expectation in equation \ref{vf_ck} 
\begin{align*} & (1-d)\cdot   V^{ck}_{t+1}(\Omega') + d\cdot [ \theta V_{t+1}^{f,sk}(\omega^{df}) + (1-\theta)V_{t+1}^{m,s}(\omega^{dm})] =    V^{ck}_{t+1}(\Omega') + \\
 &  \ \ \ \theta \cdot d(\Omega') \cdot [V_{t+1}^{f,sk}(\omega^{df}) - V_{t+1}^{f,ck}(\Omega')] + (1-\theta) \cdot d(\Omega') \cdot [V_{t+1}^{m,sk}(\omega^{dm}) - V_{t+1}^{m,ck}(\Omega')] - d(\Omega')\cdot \Delta(\Omega'),
\end{align*}
if we consider couple that is close to divorce, so there exist two $\Omega'_1$ and $\Omega'_2$  that are close to each other according to some metric, but $d(\Omega'_2) = 1$ and $d(\Omega'_1) = 0$. Note that around the divorce threshold by [...] \textbf{both} spouses should be indifferent between staying on leaving, therefore
\[ V_{t+1}^{f,sk}(\omega^{df}) - V_{t+1}^{f,ck}(\Omega')\approx 0, \ \ V_{t+1}^{m,sk}(\omega^{dm}) - V_{t+1}^{m,ck}(\Omega') \approx 0,\]
therefore if $\Delta(\Omega')$ is non-zero in these points, this whole expression jumps where value of $d$ jumps. [It is not clear whether it is actually zero or not, but in simulations it does look like it is positive.]

\subsection{Renegotiation With Collective Discrete Decisions\label{ren-disc}}
This section explains how discrete decisions interact with couple's renegotiation. The theory requires participation constraints to be satisfied \textit{after} any discrete choice. In the context of fertility decisions, that means that both spouses can predict future decisions and re-negotiate $\theta$ accounting for this prediction (rather than assuming that fertility status does not change).

With slight abuse of notation, let's suppose that couple gets collective value $V^{ck}(\theta,\Omega)$ if it makes a baby and $V^{c}(\theta,\Omega)$ is it does not. Given $\theta$ total value before the decision is made is \[V^{c?}(\theta,\Omega) = \max\{ V^{ck}(\theta,\Omega), V^{c}(\theta,\Omega)\}\]
after that I define
\[V^{f,c?}(\theta,\Omega) = V^{f,ck}(\theta,\Omega)\cdot \I[V^{ck}(\theta,\Omega) \geq V^{c}(\theta,\Omega)] + V^{f,c}(\theta,\Omega) \cdot \I[V^{ck}(\theta,\Omega) < V^{c}(\theta,\Omega)],\]
note that it is not necessary equal to $\max\{ V^{f,ck},V^{f,c} \}$. Similarly we can define $V^{m,c?}(\theta,\Omega)$.

The main thing to notice is that although $V^{c?}$ is continuous in $\theta$ individual values $V^{f,c?}$ and $V^{m,c?}$ are not, unless both spouses always have agreement on whether to make a baby: $V^{ck} \geq V^{c} \Leftrightarrow V^{f,ck} \geq V^{f,c} \Leftrightarrow V^{m,ck} \geq V^{m,c}$. Therefore set
\[\Omega = \left\{\tilde{\theta} : V^{f,c?}(\tilde\theta,\Omega) \geq V^{f,s}(\omega^{df}), \ \  V^{m,c?}(\tilde\theta,\Omega) \geq V^{m,s}(\omega^{dm}) \right\}\]
may be disjoint, and careful approximation of $V$ with respect to $\Omega$ is required.

The possible solution to it would be to limit fertility decisions to the cases where both spouses agree on having kids, this, however, goes against the logic where couple maximizes collective utility subject to participation constraints.

\subsection{Accounting for Additional Transitions\label{add-trans}}
In value functions I stated above I assumed that females stays single mother for at least one period if it exits a couple with a kid or if she has a baby at the moment of marriage and disagrees to marry. In the actual model I relax this assumption so the female can stop being single mother immediately. This is needed, for instance, for evaluating counterfactuals in which female can exit couple without pressure of being a single mother.

I assume that at the moment of (re)negotiation female is unsure about whether she is going to be a single mother. Therefore, I define expected value of leaving the couple to be
\[V_t^{f,c?}(\omega) = p^{\text{out}}[V_t^{f,s}(\omega) + S] + (1-p^{\text{out}})V_t^{f,sk}(\omega),\]
therefore equaiton \ref{single-fem} becomes
\begin{align}V^{f,s}_t(\omega) = \max\limits_{c} & \bigg\{ u(c) + \beta \E_t \Big[ (1 - p^{\text{meet}}_t)\cdot V^{f,s}_{t+1}(\omega') + \label{single-fem-true} \\  \nonumber
& \hspace{7em} p^{\text{meet}}_t (1-p^{\text{preg}}_t) \big\{ m^{np} \cdot V^{f,c}_{t+1}(\Omega^c) + (1-m^{np})V^{f,s}_t(\omega')\big\} + \\  \nonumber
& \hspace{10em} p^{\text{meet}}_t p^{\text{preg}}_t \big\{ m^{p} \cdot V^{f,ck}_{t+1}(\Omega^c) + (1-m^{p})V^{f,s?}_{t+1}(\omega')\big\}  \Big]  \bigg\},
\end{align}
the set is \ref{reneg-with-kids} becomes
\begin{equation}\Theta_t = \left\{ \tilde\theta : V^{f,ck}_t(a,\tilde\theta,\psi,z^f,z^m,\xi) \geq V_t^{f,s?}(a^{df},z^f,\xi) , \ \ V_t^{m,ck}(a,\tilde\theta,\psi,z^f,z^m,\xi)\geq V_t^{m,s}(a^{dm},z^m) \right\} \label{reneg-with-kids}
,\end{equation}
and the analogous term is introduced for initial negotiation of couples in case of pregnancy shock.

\section{Computational Notes}
This section describes the procedures I use to approximate the solution of the model above. In general, since the model is finite-horizon, value function iteration starting from the last period is used. The value functions and corresponding policy functions are approximated on a grid. Since state space is quite large (it has at least 4 continuous dimensions for couples), I use approximation techniques of [Judd--Maliar--Maliar] to deal with dimensionality. Namely, in the dimension of productivity, marriage terms and marriage surplus the functions are approximated by Smolyak polynomials (that are a variation of multi-dimensional Chebyshev polynomials) on sparse Smolyak grid. However, since the assets dimension is important and non-convex value functions can cause troubles, so I use full grid in the dimension of assets and linearly interpolate between assets points.

An additional issue is integration, required to compute expected future values of the value function accounting for possible (re)negotiation and divorce. This is noted to be the most costly part of value function iteration. I utilize monomial integration rules describe by [JMM] to do it efficiently. The second part of this section describe this.

\subsection{Value Function Approximation}
I describe techniques I used for value functions, the same principle applies for policy functions: they are just byproduct of value function iteration that I use for simulations.

State space of continuous variables for couples is $\Omega = (a,\theta,\psi,z^f,z^m)$. I partition it as $\Omega = (a,O)$. Smolyak method generates $S$ gridpoints $G = \{O_1,...,O_S\}$, where each $O$ represents some combination of $(\theta,\psi,z^f,z^m)$; and polynomials $P = \{p_1(o),...,p_S(o)\}$. Therefore any function $f(o)$ is approximated by
\[f(o) \approx \hat{f}(o) = \sum_{i=1}^S c_{i}\cdot p_i(o),\]
and coefficients $c$ are found from collocation relation
\[\hat{f}(O_i) = f(O_i) \ \forall i = 1,...,S.\]

Since we have to approximate functions for different levels of $a$, I introduce additional grid $A = \{a_1,...,a_J\}$. For each $a_j$ we can find corresponding coefficients $c$, I denote them $c_i(a_j)$. At points $a_j$ the function can be written as
\[\hat{f}(a_j,o) = \sum\limits_{i=1}^S c_i(a_j) p_i(o),\]
for points between grid $a \not A$ I linearly interpolate $f(a_j,o)$. Generically, to interpolate values for arbitrary $a$ and $o$ the  reasonable approach is to first compute values of $f$ on each point of $A$ grid for current $o$, to get $F = \{f(a_1,o),...,f(a_J,o)\}$ and then interpolate $F$ with respect to $a$. However, I use linear interpolation and since the function is bilinear in $c$ and $p$, this is fully equivalent to linear interpolation of coefficients $c_i(a_j)$. Therefore I define $\hat{c}_i(a)$ to be linearly interpolated value of coefficients, and therefore the function at arbitrary point is approximated by
\[\hat{f}(a,o) = \sum\limits_{i=1}^S \bar{c}_i(a) p_i(o).\]
If we use interpolation other than linear with respect to $a$ this equivalence would be violated so the procedure becomes more complicated as it requires evaluating many values of $f$.

\subsection{Monomial Quadratures}
See [JMM] for a detailed discussion. Let $\epsilon$ be $d$-dimensional normal distribution and we are interested in evaluating $\E(f(\epsilon))$. Standard techinque would involve using Gauss--Hermite quadrature with respect to each dimension of $\epsilon$, this, however, quickly becomes infeasible: if we want to have 5 points with respect to each dimension of 4-dimensional vector we would have $5^d = 625$ points, out of which many will have very low weights.

Monomial quadratures generate number of integration node that grows much lower than exponent of $d$. I use the one that is proportional to $d^2$. So, the rule generates $Q$ vectors $\{\epsilon_1,...,\epsilon_q\}$ with weights $w_q$, and the integral can just be written as $\E(f(x))\approx \sum_{q} f(\epsilon_q)$.

As it is noted by [...] that there is a separate issue of using quadratures to approximate integrals of discontinuous functions. However, they also show that the model with deterministic transition decisions can be interpreted as a limiting case of model with stochastic (logit) transitions, for which integration by quadratures is valid. I also believe that this can be addressed by re-interpreting shocks to have a discrete distribution with probability mass $w_q$ rather than true normal distribution (so we do not pretend to recover the integrals under true normal distribution). In general, the math background for this does not seem to be well-developed and I use conventional tools at this stage of the project without asserting its mathematical rigor. In the future, I may try to test the approximations I use against just discretizing all random walks with number of points.

\section{Auxiliary Estimates}
\subsection{Partner's Earnings\label{partearn}}
The model assumes that
\[ z^p = z^{\text{own}} + \varepsilon^{z,p}, \ \ \ \varepsilon^{z,p}\sim\mathcal{N}(0,(\sigma^{z,p})^2),\]
so detrended log-wage of potential partner is normally distributed with the mean of own detrended log-wage and standard deviation of $\sigma^{z,p}$.

To estimate $\sigma^{z,p}$ I regress log-difference in spouses earning for couples in ACS on several predictors, that include ages of spouses, ACS year and state. In this section I present several alternative specifications, including an attempt to correct for selection into marriage, and show what variance they imply.

The main specification is
\[\log \frac{\text{Earnings of wife}}{\text{Earnings of husband}} = \alpha + X'\gamma + \epsilon,\]
where I run regression on subsample of couples in which both spouses work full time and report labor income above 5000 per year.  Controls $X$ include 4th degree polynomials and wife's and husband's age, as well as absolute value of age difference, dummies for states and ACS years. I regress difference in per-hour earnings to the set of controls for couples who married in the year previous to the survey year and who have no children in the household. Relaxing this definition, using total earnings instead of per-hour and even controlling for education lead to slightly different estimates, but all of them are within $[0.66,0.8]$ range.

I pick the upper bound of this interval and set $\sigma^{z,p} = 0.8$: as in the model people are more likely to agree to marry similar partners, I expect similar statistic in the simulations to be slightly lower. [Similar statistic computed on simulated data returns ...]

\newpage
\section*{Very Technical Appendix, Not For Final Version}
\section{Expressions for Value Functions}
\subsection{Notation}
This technical note shows exact expressions for approximating value functions. For the couple's state I use partition $\Omega = (a,O)$, where $O = (\theta,\psi,z^f,z^m)$. For the age variable I put $\xi\in\{y,s\}$ counts as a separate letter, so states for couples with kids are noted like $cky$ and $cks$. Let $p^s$ represents probability that the child ``grows up'' so couple switches from $cky$ to $cks$. For individual's state there are only two variables $\omega = (a,z)$, and child's age for single mother is reported as state $sky$ if $\xi = y$ and $sks$ if $\xi = y$.

Grid consists of all possible combinations of $a$ and $O$ ($a$ and $z$ in single's case), so \[G = \{ \{a_1,...,a_I\} \times \{o_1,...,o_J\}\}\]

Let $i$ refers to assets position and $j$ refers to values of everything else. Finally, indices $q$ refer to quadrature nodes, and $w_q$ to quadrature weights that add up to 1.

\subsection{Single Agents}
Consider females, males are handled analogously. Value function \ref{single-fem} can be written as
\[V^{f,s}_t(a,z) = \max\limits_{c} \left\{ u(c) + \beta \mathcal{V}^{f,s}_{t+1}(s,z)\right\}, \ \ c + s = M(a,z),\]
Here  $\mathcal{V}^{f,s}_{t+1}$ represents expectation of tomorrow's individual value function, accounting for all possible transitions. Argument $s$ refers to savings made in the current period (it is not necessary equal to future assets as future assets change if the partner is met), and $z$ refers to current value of productivity. Approximating $\mathcal{V}_{t+1}(s,z)$ (i.e. integration) is the key challenge. Note that it does not depend on current assets, only on savings $s$ that define future assets position. $M(a,z) = R\cdot a + \exp(z + \text{Trend}_t)$ is the amount of money in the end of the period.

I treat the function $\mathcal{V}^{f,s}_{t+1}(s,z)$ as a collection of functions $\mathcal{V}^{f,s}_{mj,t+1}$ for each value of current productivity $z_j$ and savings $s_m$. I generate quadrature nodes that consist of shock to individual productivity $\epsilon^{z,f}$, shocks to partner's assets and productivity positions
$(\epsilon^{a,p},\epsilon^{z,p})$ and initial couple's surplus $\psi$. I denote these nodes as $x_q = (\epsilon^{z,f}_q,\epsilon^{a,p},\epsilon^{z,p},\psi_q)$, in total monomial rule results in $Q = [...]$ nodes. Weight $w_q$ corresponds to each node. Given savings $s$ and grid point in $z_j$ I generate partner's and couple's characteristics. 
\[a^p_{mq} = s_m\cdot \exp(\epsilon^{a,p}_q), \ \ a^c_{mq} = s_m\cdot [1 + \exp(\epsilon^{a,p}_q)], \ \ z^m_{jq} = z^f_j \exp(\epsilon^{z,p}_q),\]
(in some sense in this setup by saving more individuals have chances of getting more wealthy partners).  Also, individual next period's productivity is generated according to $z^f_{jq} = z^f\cdot \exp(\epsilon^{z,f}_q)$. Note that $z^f_j$ refers to the current value and $z^f_{jq}$ refers to the next period's values;

This allows to generate potential couple's characteristics $(a^c_{mq},\psi_q,z^{f}_{jq},z^{m}_{jq})$, and combining it with value functions for the next period we can define:
\[\Theta^{np}_{mjq,t+1} = \left\{ \tilde\theta : V^{f,c}_{t+1}(a^c_{mjq}, \tilde\theta, \psi_q,z^{f}_{jq},z^{m}_{jq}) \geq V^{f,s}(s_m,z^f_{jq}), \ V^{m,c}_{t+1}(a^c_{mjq}, \tilde\theta, \psi_q,z^{f}_{jq},z^{m}_{jq}) \geq V^{m,s}(a^p_{mq},z^m_{jq}) \right\},\]
if $\Theta^{np}_{mjq,t+1}$ is empty\footnote{I figure this out by defining a uniformly spaced grid for $\theta \in [0.05,0.95]$. Because of Smolyak grid, this can be done very efficiently as $V^{f,c}_{t+1}(a^c_{mjq}, \tilde\theta, \psi_q,z^{f}_{jq},z^{m}_{jq})$ are just a polynomial in $\tilde{\theta}$.}
, $m^{np}_{mjq,t+1} = 0$ (so no marriage happens at node $jq$ if agents made savings of $s_m$). If the set is not empty, then $\theta_{mjq}$ is determined by Nash Bargaining \ref{nbs} and $m^{np}_{mjq,t+1} = 1$. Therefore at node $(s_m,z_j)$ future value function in case the partner is met and pregnancy did not happen can be approximated as
\[\E_t  \big\{ m^{np} \cdot V^{f,c}_{t+1}(\Omega^c) + (1-m^{np})V^{f,s}_t(\omega')\big\} \approx \mathcal{E}^{np}(s_m,z_j),\]
where
\[\mathcal{E}^{np}(s_m,z_j) = \sum\limits_{q=1}^{Q} w_q \cdot \left\{ m^{np}_{mjq,t+1}\cdot V^{f,c}_{t+1}(a^c_{mq},\theta_{mjq},\psi_q,z^{f\prime}_{jq},z^m_{jq}) + (1-m^{np}_{mjq,t+1})\cdot V^{f,s}_{t+1}(s_m,z_j)\right\}.\]
In exactly the same manner we define the set in the case the pregnancy shock did happen:
\[\Theta^{p}_{mjq,t+1} = \left\{ \tilde\theta : V^{f,ck}_{t+1}(a^c_{mjq}, \tilde\theta, \psi_q,z^{f}_j,z^{m}_{jq}) \geq V^{f,s?}(s_m,z^f_{j}), \ V^{m,ck}_{t+1}(a^c_{mjq}, \tilde\theta, \psi_q,z^{f}_{jq},z^{m}_{jq}) \geq V^{m,s}(a^p_{mq},z^m_{jq}) \right\},\]
where function $V^{f,s?} = p^{\text{out}}[V^{f,s} + S] + (1-p^{\text{out}})V^{f,sky}_t$ captures possible immediate transition out of being single mother. In the same way 
\begin{align*}\mathcal{E}^{p}(s_m,z_j) = \sum\limits_{q=1}^{Q} w_q \cdot  \Big\{ m^{p}_{mjq,t+1}\cdot V^{f,cky}_{t+1}(a^c_{mq},\theta_{mjq},\psi_q,z^f_j,z^m_{jq}) + (1-m^{p}_{mjq,t+1})\cdot V^{f,s?}(s_m,z_j) \Big\},
\end{align*}
lastly, the value of not meeting any partner is simply
\begin{align*}\mathcal{E}^{0}(s_m,z_j) = \sum\limits_{q=1}^{Q} w_q \cdot  V^{f,s}_{t+1}(s_m,z^f_{jq}),
\end{align*}
finally, the whole function $\mathcal{V}^{f,s}_{t+1}(s_m,z_j)$ can be approximated by weighted average
\[\mathcal{V}^{f,s}_{t+1}(s_m,z_j) \approx (1-p^{\text{meet}}_t) \cdot \mathcal{E}^{0}(s_m,z_j) +  p^{\text{meet}}_t(1-p^{\text{preg}}_t)  \cdot \mathcal{E}^{np}(s_m,z_j) 
+  p^{\text{meet}}_t p^{\text{preg}}_t  \cdot \mathcal{E}^{p}(s_m,z_j).\]

After that I use linear interpolation in $s$ dimension to approximate $\mathcal{V}^{f,s}_{t+1}(s_m,z_j)$ for arbitrary $s$. Therefore I can get approximation of $\mathcal{V}^{f,s}_{t+1}(s,z_j)$. As a result, value function in each asset point $a_i$, $z_j$ is obtained by numerically solving
\[V^{f,s}_t(a_i,z_j) = \max\limits_{s \in [\underline{A},\overline{A}]} \left\{ u(M(a_i,z_j) - s) + \beta \mathcal{V}^{f,s}_{t+1}(s,z_j)\right\}.\]
I take $\underline{A}$ and $\overline{A}$ the same as upper and lover bound for individual's assets grid (this is important as approximation may be crude). I use Matlab's \textbf{fminbnd} to solve for the maximum point, although I also test that results are very similar when I just use discrete grid $s \in \mathcal{S} = M(a_i,z_j) \cdot\{0,...,0.99\}$ with some spacing, though this variation is much slower.

Choice of solver can generically be an issue as because of concerns described above function $\mathcal{V}^{f,s}_{t+1}(s_m,z_j)$ may be non-smooth or even discontinuous.

\subsection{Couples}
I consider couples that do not have kids yet as they have richer set of possible transitions.  Following [...] I split household's decision problem on intrahousehold, intratemporal and intertemporal levels, where first two can be solved analytically. For intrahousehold, I define
\[U(c,\theta) = \max\limits_{c^f,c^m} \left\{\theta u(c^f)  + (1-\theta)u(c^m) \right\}, \ \ \text{s.t.} \ \ \ c = \left[ (c^f)^{1+\rho_c} + (c^m)^{1+\rho_c}\right]^{\frac1{1+\rho_c}},\]
and given CES assumption this function has simple analytical form.

Intratemporal dimension is trivial, as total consumption is the only intratemporal choice.

For intertemporal I can write the problem \ref{vf-c} as
\[V^{c}_{t+1}(a,\theta,\psi,z^f,z^m) = \max\limits_{c} \left\{ U(c,\theta) + \beta \mathcal{V}_{t+1}(s,\theta,\psi,z^f,z^m) \right\},\]
where $\mathcal{V}_{t+1}(s,\theta,\psi,z^f,z^m)$ consists of two terms: what happens if pregnancy shock arrives and what if it does not.

If the shock arrives, in the next period couple is in state $cky$, and if it decides to divorce female becomes a single mother (and may quit this state immediately), and male becomes just a single male. I approximate $\mathcal{V}_{t+1}(s,\theta,\psi,z^f,z^m)$ here.

I write current state as combination of $(a_i,O_j)$. Function $\mathcal{V}_{t+1}(s,\theta,\psi,z^f,z^m)$ depends on current state $O= (\theta,\psi,z^f,z^m)$ and future savings $s$.  Integration nodes contains $x_q = (\epsilon^{z,f}_q,\epsilon^{z,m}_q,\epsilon^{\psi}_q)$. I take grid with respect to current states, so $O_j= (\theta_j,\psi_j,z^f_j,z^m_j) \in O$ and future savings $s \in A$, so I again approximate expected future value function at each combination of $s_m$ and $O_j$, so I compute values $\mathcal{V}_{mj,t+1} = \mathcal{V}_{t+1}(s_m,\theta_j,\psi_j,z^f_j,z^m_j)$. Given each $j$, I can compute next period values of $(\psi_{jq},z^f_{jq},z^m_{jq})$ by adding shocks given by quadrature notes to the current point. I also can compute divorce options: $\omega^{df}_{mjq} = ( (1-\kappa)\cdot 0.5 \cdot s_m, \ z^f_{mjq})$ and $\omega^{dm}_{mjq} = ( (1-\kappa)\cdot 0.5 \cdot s_m, \ z^m_{mjq})$. This allows to define for each quadrature node a renegotiation set:
\[\Omega^{p}_{mjq,t+1} = \left\{ \tilde\theta \ : \ V^{f,cky}_{t+1}(s_m,\tilde\theta,\psi_{jq},z^f_{jq},z^m_{jq}) \geq V^{f,s?}_{t+1}(\omega^{df}_{mjq}), \ \ V^{m,cky}_{t+1}(s_m,\tilde\theta,\psi_{jq},z^f_{jq},z^m_{jq}) \geq V^{m,s}_{t+1}(\omega^{dm}_{mjq}) \right\},\]
and then if $\Omega^{p}_{mjq,t+1}$ is not empty, I pick from it next period's $\theta_{mjq}$ closest to initial $\theta_j$ and set $d^{p}_{mjq} = 0$, otherwise divorce happens and $d^{p}_{mjq} = 0$.

Therefore expected future value if pregnancy shock arrives is 
{\small
\[\mathcal{E}^{p}(s_m,O_j) = \sum\limits_{q=1}^{Q} w_q \cdot\left\{ d^p_{mjq}\cdot V_{t+1}^{cky}(s_m,\theta_{mjq},\psi_{jq},z^f_{jq},z^m_{jq}) + (1-d^p_{mjq}) [\theta_{j}\cdot V^{f,sky}(\omega^{df}_{mjq}) + (1-\theta_j)V^{m,s}(\omega^{dm}_{mjq})] \right\} \]
}
If pregnancy shock does not arrive, then renegotiation should account for discrete decision of whether to have a kids that happens afterwards, and this decision may depend on $\theta$. Namely, I define
\[k_{mjq}(\tilde\theta) = \I\left[ V^{cky}_{t+1}(s_m,\tilde\theta,\psi_{jq},z^f_{jq},z^m_{jq}) \geq V^{c}_{t+1}(s_m,\tilde\theta,\psi_{jq},z^f_{jq},z^m_{jq})\right],\]
then I define individual value accounting for this choice:
\[V^{f,c?}_{t+1}(s_m,\tilde\theta,\psi_{jq},z^f_{jq},z^m_{jq}) = k_{mjq}(\tilde\theta)\cdot V^{f,cky}_{t+1}(s_m,\tilde\theta,\psi_{jq},z^f_{jq},z^m_{jq}) + (1-k_{mjq}(\tilde\theta))\cdot V^{f,c}_{t+1}(s_m,\tilde\theta,\psi_{jq},z^f_{jq},z^m_{jq}),\]
and the renegotiation set is defined as
\[\Omega^{np}_{mjq,t+1} = \left\{ \tilde\theta \ : \ V^{f,c?}_{t+1}(s_m,\tilde\theta,\psi_{jq},z^f_{jq},z^m_{jq}) \geq V^{f,s}_{t+1}(\omega^{df}_{mjq}), \ \ V^{m,c?}_{t+1}(s_m,\tilde\theta,\psi_{jq},z^f_{jq},z^m_{jq}) \geq V^{m,s}_{t+1}(\omega^{dm}_{mjq}) \right\},\]
and we define $\theta_{mjq}$ and $d^{np}_{mjq}$ analogously.

Therefore we define expected future value if pregnancy shock does not arrive to be
{\small
\[\mathcal{E}^{np}(s_m,O_j) = \sum\limits_{q=1}^{Q} w_q \cdot\left\{ d^{np}_{mjq}\cdot V_{t+1}^{c?}(s_m,\theta_{mjq},\psi_{jq},z^f_{jq},z^m_{jq}) + (1-d^{np}_{mjq}) [\theta_{j}\cdot V^{f,s}(\omega^{df}_{mjq}) + (1-\theta_j)V^{m,s}(\omega^{dm}_{mjq})] \right\} \]
}
so we can finally define value of $\mathcal{V}$ at points $s_m$:
\[\mathcal{V}^{c}_{t+1}(s_m,O_j) = p_t^{\text{preg}}\cdot \mathcal{E}^{p}(s_m,O_j) + (1-p_t^{\text{preg}})\cdot \mathcal{E}^{np}(s_m,O_j),\]
and then values at arbitrary points $s$ are obtained by linear interpolation. Therefore the ultimate problem is
\[V^{c}_t(a_i,O_j) = \max\limits_{s \in [\underline{A},\overline{A}]} \left\{ U(M(a_i,O_j) - s,\theta_j) + \beta \mathcal{V}^{c}_{t+1}(s,O_j)\right\}.\]

\subsection{Couples With Kids} 
Namely, for intrahousehold allocation of consumption I define
\[U(c,q,\theta) = \max\limits_{c^f,c^m} \left\{\theta u(c^f,q)  + (1-\theta)u(c^m,q) \right\}, \ \ \text{s.t.} \ \ \ c = \left[ (c^f)^{1+\rho_c} + (c^m)^{1+\rho_c}\right]^{\frac1{1+\rho_c}},\]
and it still has simple tractable form.

For intratemporal level I have to solve
\[U(m,W^f,\xi,\theta) = \max\limits_{c,q,l_f,x} U(c,q,\theta), \ \ \ \text{s.t.} \ \ \ c + x + l_f\cdot W^f = m, \ \ q =[\mu_\xi\cdot l_f^{\theta_q} + (1-\mu_\xi)\cdot x^{\theta_q}]^{\frac1{\theta_q}}. \]
this function can also be derived analytically given my assumption on functional forms.

After aggregation, the intertemporal problem looks similar to the one for singles:
\[V^{cky}(a,\theta,\psi,z^f,z^m) = \max\limits_{m} \left\{ U(m,W^f_t(z^f),\xi,\theta) + \beta \mathcal{V}^{ck}_{t+1}(s,\theta,\psi,z^f,z^m,\xi)\right\}, \ \ \text{s.t.} \ \ m + s = M(a,z^f,z^m),\]
where definition of $\mathcal{V}$ is analogous.

\end{document}

### 2019-05-03 main/comp-appendix.tex
Re-organized into few files

--- 2019-05-03 main/draft.tex
@@
-\documentclass[12pt,letter]{article}
-\usepackage[left=0.8in,right=0.8in,top=1in,bottom=1in]{geometry}
-\usepackage{amsmath}
-\usepackage{pdflscape}
-\usepackage{amsfonts}
-\usepackage{amssymb}
-\usepackage{graphicx}
-\usepackage{caption}
-\usepackage{multicol}
-\usepackage{multirow}
-\usepackage{microtype}
-\usepackage{euscript}
-\usepackage{epsfig}
-\usepackage{epstopdf}
-\usepackage{mathrsfs}
-\usepackage{tikz}
-\newcommand{\hypo}{\mathcal{H}}            
-\bibliographystyle{ieeetr}
-
-\usepackage[flushleft]{threeparttable}
-
-%\usepackage[cp1251]{inputenc}
-\usepackage[english]{babel} 
-\DeclareMathOperator{\rank}{rank}
-\newcommand*{\hm}[1]{#1\nobreak\discretionary{}
-            {\hbox{$\mathsurround=0pt #1$}}{}}
-
-            \def\onepc{$^{\ast\ast}$} \def\fivepc{$^{\ast}$}
-\def\tenpc{$^{\dag}$}
-\def\legend{\multicolumn{4}{l}{\footnotesize{Significance levels
-:\hspace{1em} $\dag$ : 10\% \hspace{1em}
-$\ast$ : 5\% \hspace{1em} $\ast\ast$ : 1\% \normalsize}}}
-
-\newcommand{\bs}[1]{\boldsymbol{#1}}  
-\newcommand{\bsA}{\boldsymbol{A}}
-
-%\setstretch{1}                         
-\flushbottom                            
-\righthyphenmin=2                      
-\pagestyle{plain}                       
-%\settimeformat{hhmmsstime}  
-\widowpenalty=300                   
-\clubpenalty=3000                     
-\setlength{\parindent}{0em}           
-\setlength{\topsep}{0pt}              
-\usepackage[pdftex,unicode,colorlinks=true,urlcolor=blue]{hyperref}
-\usepackage{bbm}
-\usepackage{tabularx}
-\renewcommand{\emptyset}{\varnothing}
-
-\setlength{\parskip}{0.5\baselineskip plus2pt minus2pt}
-
-\newcommand{\e}{\varepsilon}
-\DeclareMathOperator*{\Argmax}{\mathrm{Argmax}}
-\DeclareMathOperator*{\Argmin}{\mathrm{Argmin}}
-\DeclareMathOperator*{\argmax}{\mathrm{arg\,max}}
-\DeclareMathOperator*{\argmin}{\mathrm{argmin}}
-
-\newcommand{\blp}{\mathrm{BLP}}
-\DeclareMathOperator*{\plim}{\mathrm{plim}}
-\DeclareMathOperator{\Max}{\mathrm{Max}}
-\newcommand{\R}{\mathbb{R}}
-\newcommand{\Y}{\mathcal{Y}}
-\newcommand{\Z}{\mathcal{Z}}
-\renewcommand{\geqslant}{\geq}
-\renewcommand{\leqslant}{\leq}
-\newcommand{\p}{\bs p}
-\newcommand{\y}{\bs y}
-\def\dd#1#2{\frac{\partial#1}{\partial#2}}
-
-\renewcommand{\emptyset}{\varnothing}
-
-\DeclareMathOperator{\tr}{\mathrm{tr}}
-
-\newcommand{\bb}{\bs \beta}
-\newcommand{\X}{\bs X}
-\DeclareMathOperator{\E}{\mathbb{E}}
-\DeclareMathOperator{\PP}{\mathbb{P}}
-\DeclareMathOperator{\V}{\mathbb{V}}
-\DeclareMathOperator{\CM}{\mathbb{C}}
-\renewcommand{\C}{\CM}
-\DeclareMathOperator{\var}{\mathrm{var}}
-\DeclareMathOperator{\cov}{\mathrm{cov}}
-\DeclareMathOperator{\corr}{\mathrm{corr}}
-\DeclareMathOperator{\MSE}{\mathrm{MSE}}
-\DeclareMathOperator{\Bias}{\mathrm{Bias}}
-\renewcommand{\P}{\PP}
-\newcommand{\dsim}{\stackrel{d}{\sim}}
-\newcommand{\hn}{\mathcal{H}_0}
-\newcommand{\ha}{\mathcal{H}_a}
-\newcommand{\thetab}{\bs \theta}
-\newcommand{\pv}{\text{P-value}}
-\newcommand{\N}{\mathcal{N}}
-\newcommand{\MLE}{\scriptscriptstyle MLE}
-\newcommand{\LR}{\mathrm{LR}}
-\newcommand{\I}{\mathbb{I}}
-\newcommand{\sumin}{\sum\limits_{i=1}^n}
-\newcommand{\sumti}{\sum\limits_{t=0}^\infty}
-\newcommand{\hbeta}{\hat{\beta}}
-\newcommand{\halpha}{\hat{\alpha}}
-\newcommand{\hsigma}{\hat{\sigma}}
-\newcommand{\hvar}{\widehat{\var}}
-\newcommand{\hcov}{\widehat{\cov}}
-\newcommand{\Q}{\mathbb{Q}}
-
-\newcommand{\pconv}{\xrightarrow{ \ p \ }}
-\newcommand{\dconv}{\xrightarrow{ \ d \ }}
-\newcommand{\asconv}{\xrightarrow{ \ a.s. \ }}
-\newcommand{\msconv}{\xrightarrow{ \ m.s. \ }}
-
-\newcommand{\pic}[4][h!]{\begin{figure}[#1]
-
-\begin{center}\includegraphics[width=#2cm]{#3}\caption{#4\label{#3}}\end{center}
-\end{figure}}
-
-%outtex
-\def\onepc{$^{\ast\ast}$} \def\fivepc{$^{\ast}$}
-\def\tenpc{$^{\dag}$}
-\def\legend{\multicolumn{4}{l}{\footnotesize{Significance levels
-:\hspace{1em} $\dag$ : 10\% \hspace{1em}
-$\ast$ : 5\% \hspace{1em} $\ast\ast$ : 1\% \normalsize}}}
-%end outtex
-
-%\bibliographystyle{ieeetr}
-
-\newcommand{\laseq}{\stackrel{\lambda\text{-a.e.}}{=}}
-\renewcommand{\d}{\underline}
-\renewcommand{\u}{\overline}
-\newcommand{\td}{\underline{\theta}}
-\newcommand{\tu}{\overline{\theta}}
-%\renewcommand{\theenumi}{\alph{enumi}}
-%\renewcommand{\labelenumi}{(\theenumi)}
-%\renewcommand{\theenumii}{\roman{enumii}}
-%\renewcommand{\labelenumii}{\theenumii.}
-%\renewcommand{\theenumiii}{\arabic{enumiii}}
-%\renewcommand{\labelenumiii}{\theenumiii.}
-%\renewcommand{\epsilon}{\varepsilon}
-\newcommand{\hneq}{\stackrel{\hn}{=}}
-\newcommand{\deq}{\stackrel{d}{=}}
-%\title{416-2 Final Project\\
-%Reproductive Technologies, Aging and Fertility Choice (Proposal)}
-%\author{Egor Kozlov}
-\begin{document}
-%\begin{center}\textbf{Economics 416-1} \\ \emph{By Egor Kozlov}\end{center}
-%\maketitle
-
-\begin{abstract}
-Unplanned pregnancies often trigger marriage for unmarried cohabitating couples.
-I show that on average marriages in which a child appeared before or at the year of marriage are characterized with higher divorce probabilities and other outcomes suggesting lower marital quality.
-This association is robust to different ways of controlling for composition, and it is more pronounced for higher income and more educated couples.
-I argue that this pattern is consistent with household bargaining models, and the main driver of the result is unequal exposure of spouses to the childcare costs: since being single mother is a bad outcome for an educated female, she is more likely to agree to enter the marriages of lower quality.
-I estimate a lifecycle model with dynamic bargaining and fertility decisions to show how much of these differences is driven by household bargaining and how childcare costs contribute to observable divorce rates.
-\end{abstract}
-
-\section{Introduction}
-%Understanding how people make marriage decisions and more generally how household structure is formed is crucial for analyzing labor market outcomes, assets accumulation, various dimensions of inequality and children development outcome.
-Most of existing studies, even ones that take household structure seriously, typically abstract from the fact that household structure is itself a choice variable.
-However, existing trends in marriage and divorce rates, increasing share of kids with unmarried parents and in blended families, and, especially, how different these trends are by income and racial groups --- this all suggests economic importance of household issues.
-
-%In particular, economists know pretty little about how people decide on marriage and divorce, despite of the fact that these issues have lots of economic content.
-Stylized models describing marriage (and, in smaller extent, divorce) emerged in family economics, however most of them describe qualitative issues and do not deliver quantitative predictions that can be applied for analyzing policy or just understanding how different factors contribute to the recent changes.
-...]
-
-\subsection{Trends in Marriage And Divorce}
-
-\subsection{Existing Results}
-
-\section{Empirical Evidence}
-The key empirical observation is association between marriage timing and divorce probabilities.
-Statistically, women who have their kids before they are married have significantly higher chance to be divorced.
-American Community Survey (ACS) data are suitable for establishing this fact only of particular subsample of women: most importantly, those who had no more than one marriage in their life.
-To re-confirm the patterns I additionally use Survey of Income and Program Participation (SIPP) data: they have smaller sample size but collect exact questions I care about.
-
-\subsection{Defining Variables}
-
-General strategy involves dividing sample in two groups: couples who had kids before their marriage ($K\to M$) and those who had kids after their marriage ($M\to K$).
-After this division I perform multiple comparisons of these groups: the main comparison is looking at divorce rates in several variations, few additional comparisons involve marriage inequality and outcomes of children.
-
-Although the couples are the primary focus of the paper, the surveys I use look at individuals and households.
-Therefore I follow females rather than males: upon divorce, children are more likely to stay with mothers (share of single mothers in ACS is ..., single fathers ...).
-The timing I consider in the  most general form is (for females, given both events occurred):
-\[\Delta T = T^{\text{First child  is born}} - T^{\text{First marriage happened}},\]
-where $T$ refers to moments of time. After that I define two groups (given $\Delta T$ is measured in years):
-\begin{align*}&M\to K \Leftrightarrow \Delta T \in \{1,2,...\},\\
-&K\to M \Leftrightarrow \Delta T \in \{...,-2,-1,0\},\end{align*}
-
-In practice, I exclude observations with $\Delta T < -5$ or $\Delta T > 10$.
-The reason for the first one is mainly related to step-children: for the subsample when it can be tracked, share of step-children is becomes higher with more negative $\Delta  T$ in ACS, and having step-children that are relatively adult at the moment of marriage is a complex issue that is beyond the scope of the paper.
-See a note ...
-in the appendix about this.
-The reason for the second one is mainly data quality (...).
-
-As the evidence is mainly cross-sectional, I refer to share of divorced people in a group as divorce rates.
-Unlike fertility, people can change their marriage status all the time, so there is no such thing as completed divorce rate, therefore I just use the share of people who ever had a divorce.
-More accurate measuring of divorce rates involves conditioning on age, however fixing age restricts the attention to a particular cohort, and pooling people of different ages together allows to think about more general population.
-
-\subsection{Sample Restrictions}
-\subsubsection{ACS data}
-I take ACS samples of 2008--2016, as only they have variables concerning the duration of current marital status.
-Namely, people report the year of their most recent marriage.
-Fertility history is not observed perfectly: the survey reports age of the eldest child residing with individual.
-However, both of these variables can recover the timing of interest for relatively large subpopulation of females who:
-\begin{enumerate}
-\item Had exactly one marriage, so they are either married or divorced now (treating legally separated, spouse absent and widowed as married)
-\item Have children present in household, with age of the eldest child below 18
-\item Most likely reside with their eldest child judging by mother's age (I pick 20--40)
-\end{enumerate}
-%The last restriction can be relaxed, but it allows to track partners and children more clearly and also insures better quality of self-reported data.
-
-The most important restriction here is being married once --- that means that people who re-marry drop out of sample.
-For instance, if it take longer to re-marry for the people who had kids before their first marriage, share of divorced people in the group $K\to M$ would be larger without differences in probability of their first marriage to dissolve.
-This concern can only partially be addressed with ACS data, but this is not an issue in SIPP.
-
-All mean and quantile calculations are performed using person weights.
-
-\subsubsection{SIPP data}
-
-I use wave 1 of 2014 SIPP. It it representative for the whole population, I use person weights in all calculations.
-
-SIPP has much more detailed information on marital history and fertility: it explicitly asks about times of first marriage and first births, it also asks about multi-partner fertility.
-However it has substantially smaller sample size.
-For SIPP I relax the restrictions and consider all females who had at least one marriage and are of age 20--40 (treating married with spouses absent, separated and widowed as married), as their marital and fertility history is recovered perfectly.
-
-%When dividing people by income, however, I have to further restrict attention to households that consist of mother + spouse/partner (if reported) + children.
-This creates some attrition especially for young people, but differences in defining household members are minimal between ACS and SIPP, so it is not likely that it creates an issue.
-
-\subsection{Defining Income}
-This part describes how I define income in ACS.
-
-Following development and household literature, when dividing people by income group I use total household income adjusted by OECD equivalence scale.
-The scale assigns weight 1 to the household head, 0.5 to spouse and 0.3 to each child.
-
-As I use the household income as the main variable of interest, I have to restrict attention to couples that live by themselves and single females with children.
-I do this both for ACS and SIPP, and as definition of households wary a little, this may drop out slightly different groups of people.
-
-Formally,for income considerations I take a subsample of females who is marked as household's head or head's spouse, and who do not have any people living in household other than their spouse/unmarried partner and their children.
-This, for instance, excludes multigenerational household and people living with roommates, as total household's income may not be a relevant thing for some types of household composition.
-I cannot define income equivalence scale for more complicated living arrangements, although there are different scales that allow to do this.
-
-To mitigate selectivity, I compute income percentiles in all sample of females with household composition mentioned above, regardless of their marriage and fertility status.
-I use total family income as the key variable.
-To mitigate misreporting I drop observations with total yearly family income below $\$4000$ for ACS (that is about a third of federal the poverty level) or monthly income below $\$300$ for SIPP.
-
-For income percentiles I divide data on cells that consist of the US state, survey year (for ACS) and age group.
-I define age groups as 20--25, 26--30, 31--35, 36--40, and this classification is based on the age of female.
-Conditioning for year is done mainly to mitigate consequences of income decline in 2008.
-
-I drop cells with less than 100 observations (for instance, Vermont in ACS 2015 has 31 females in age group 20--25), this constitutes around [2\%] of all observations.
-
-\subsection{Raw Comparisons}
-First, I compare divorce probabilities in cross-sections of people between $K\to M$ and $M\to K$ groups. See Table \ref{diff-raw} for this.
-
-\begin{table}[h!]
-\begin{center}
-\begin{tabular}{|l|c||c|c||c||c|c|}\cline{2-7}
-\multicolumn{1}{c|}{} & \multicolumn{3}{c||}{\small ACS: sample of married once} & \multicolumn{3}{c|}{\small SIPP: sample of  ever married} \\\cline{2-7}
-\multicolumn{1}{c|}{}  & \multirow{2}{*}{\small \% K$\to$M}& \multicolumn{2}{c||}{\textbf{\% divorced (now)}} & \multirow{2}{*}{\footnotesize \% K$\to$M }& \multicolumn{2}{c|}{\textbf{\% divorced (ever)}} \\\cline{3-4}\cline{6-7}
-\multicolumn{1}{c|}{}  &  &\small if M$\to$K &  \small if K$\to$M & & \small if M$\to$K & \small if K$\to$M \\\hline
-\textit{All females, 20--40}  & $23.1\%$ & $9.8\%$ & $17.3\%$ & $22.8\%$ & $24.0\%$ & $34.7\%$ \\
-\textit{All females, 35--40} & $17.8\%$ & $11.0\%$ & $22.3\%$ & $18.9\%$ & $28.1\%$ & $45.5\%$ \\\hline\hline
-\multicolumn{7}{|p{0.9\linewidth}|}{\footnotesize \textbf{Education partition.} Note: ``college graduates'' excludes ``some college'' and ``more than college''.}\\\hline
-\textit{High school only, 20--40} & $34.2\%$ & $13.3\%$ & $16.5\%$ & $33.4\%$ & $29.0\%$  & $34.1\%$ \\
-\textit{College graduates, 20--40} & $11.5\%$ & $5.7\%$ & $14.3\%$ & $10.5\%$ & $14.9\%$ & $28.1\%$ \\\hline\hline
-\multicolumn{7}{|p{0.9\linewidth}|}{\footnotesize \textbf{Income partition:} subsample of household heads/spouses living with children and nobody else.}\\\hline
-\textit{Low 30\% income, 20--40} & $33.0\%$ & $17.7\%$ & $22.8\%$ & $29.4\%$ & $34.3\%$ &  $36.0\%$ \\
-\textit{Mid 40\% income, 20--40} & $22.8\%$ & $7.6\%$ & $11.8\%$ & $21.9\%$ & $23.8\%$ & $31.1\%$ \\
-\textit{Top 30\% income, 20--40} & $11.1\%$ & $4.6\%$ & $13.0\%$ & $12.5\%$ & $11.6\%$  & $29.8\%$ \\\hline
-\multicolumn{7}{|p{0.9\linewidth}|}{\footnotesize $\Delta T = \text{Year}^{\text{first child}} - \text{Year}^{\text{first marriage}}$. Restriction: $\Delta T \in \{-5,...,10\}$.}\\
-\multicolumn{7}{|p{0.9\linewidth}|}{\footnotesize Definition: K$\to$M if $\Delta T \in \{-5,...,0\}$. M$\to$K if $\Delta T \in \{1,...,10\}$.}\\
-\hline
-\end{tabular}
-\caption{Differences in raw means, ACS and SIPP\label{diff-raw}}
-\end{center}
-\end{table}
-
-\subsection{Differences in Composition}
-This section captures demographic and economic differences of K$\to$M and M$\to$K groups.
-Many factors could compute to the observable differences, here I quantify them and this justifies use of controls in the next part.
-I present these differences for ACS, SIPP tables are available in [Appendix ...].
-For ACS, since M$\to$K and K$\to$M are defined only for subsample of people who were married once and have children present, I also compare this subsample with more general female population.
-See Table \ref{diff-contr} for the results.
-
-\begin{table}[h!]
-\begin{center}\small
-\begin{tabular}{|l|c|c|c|c|}\cline{2-5}
-\multicolumn{1}{c|}{} & \multirow{2}{*}{All population} & \multicolumn{3}{c|}{Married once + have kids}\\\cline{3-5}
-\multicolumn{1}{c|}{}  & & Everyone & K$\to$M & M$\to$K \\\cline{2-5}\hline
-\textit{Age} & 30.2    &  33.1     &       31.7     &       33.6 \\
-\textit{Age at first marriage} & 24.2         &   23.3        &    23.6        &    23.2 \\
-\textit{Age at first birth}  & 24.0       &     25.4     &       21.7       &     26.5 \\
-\textit{Husband's age} & 34.9        &    35.8      &      34.4   &         36.2  \\
-\textit{Age difference (W$-$H)} &  2.9       &      2.8      &       3.0      &       2.7  \\\hline\hline
-\textit{Eldest child's age} & 8.6         &    7.8    &        10.0     &        7.1 \\
-\textit{Number of children} & 1.1       &      2.0      &       2.3     &       2.0 \\\hline\hline
-\textit{\% college or + (own)} & 32.1\%   &         39.5\%     &       16.9\%     &       46.3\% \\
-\textit{\% college or + (husband)} & 33.3\%     &       36.4\%     &       14.2\%       &     42.7\% \\\hline\hline
-\textit{\% In Labor Force} & 75.6\%    &        68.6\%    &        70.0\%        &    68.2\% \\
-\textit{Own income (thousands)} & 32.2 &         38.4 &          29.1 &         41.1 \\
-\textit{Hours worked} & 36.6       &     36.5    &        36.8        &    36.4\\\hline\hline
-\textit{\% white (incl hispanic)} & 69.2\%      &      76.5\%      &      72.2\%      &      77.8\% \\
-\textit{\% black} & 14.8\% &        6.7\% &            12.2\%       &       5.0\% \\\hline
-\multicolumn{5}{|p{0.7\linewidth}|}{\footnotesize ACS, women, 20--40. Means. Everything is subsamples when applicable, income and hours are for subsample of women who are employed.}\\\hline
-\end{tabular}
-\caption{Differences in composition, ACS\label{diff-comp}}
-\end{center}
-\end{table}
-
-\subsection{Statistical Tests and Controlling for Composition}
-
-In this section I formally test differences in shares for different subgroups.
-I also attempt to control for composition using regressions with many fixed effects.
-I show that generally observable mean differences are not likely to be attributed to composition, and introducing various controls still leaves substantial part of the difference.
-
-In general, I perform the following exercise: I regress indicator of female being divorced on set of controls $X$ and dummy variable $\I(\text{K$\to$M})_i$. The specification I use is
-\[D_i = \Delta \cdot \I(\text{K$\to$M})_i + X_i\gamma + \varepsilon_i.\]
-Replacing $X_i$ with a constant corresponds to raw mean comparison described above.
-
-I report three possible estimates of $\Delta$:
-\begin{itemize}
-\item \textbf{Raw}, corresponding to differences in raw means in the previous part
-\item \textbf{Controls}, where the regression includes dummies for ACS year, the US state, number of children, years after marriage, age of the eldest child, race and education group + 4th degree polynomial in woman's age
-\item \textbf{Controls+income}, where the regression includes all of the above + dummies for income groups (low 30\%, mid 40\%, top 30\%) on a subsample where income equivalence scale is defined.
-This subsample is smaller, but has arguably better data quality.
-\end{itemize}
-
-See Table \ref{diff-contr} for the results.
-
-\begin{table}
-\begin{center}
-\begin{tabular}{|l|c||c|c|c|c|c|c|}\cline{2-8}
-\multicolumn{1}{c|}{} & \multicolumn{7}{c|}{ACS: sample of married once}\\\cline{2-8}
-\multicolumn{1}{c|}{} & \small Share of &\multicolumn{2}{c|}{\small Raw difference} & \multicolumn{2}{c|}{\small with controls} & \multicolumn{2}{c|}{\small controls + income} \\ \cline{3-8}
-\multicolumn{1}{c|}{} & \small divorced & $\Delta$ & (s.e.) & $\Delta$ & (s.e.) & $\Delta$ & (s.e.) \\  \hline
-\textit{All females, 20--50} & 13.0\% & 7.6\% & (0.1\%) & 5.5\% & (0.2\%) & 5.1\% & (0.2\%) \\
-\textit{All females, 35--40} & 16.0\% & 11.3\% & (0.2\%) & 6.5\% & (0.3\%) & 5.9\% & (0.3\%) \\\hline\hline
-\multicolumn{8}{|p{0.9\linewidth}|}{\footnotesize \textbf{Education partition.} Note: ``college graduates'' excludes ``some college'' and ``more than college''.}\\\hline
-\textit{High school only, 20--50} & 16.2\% & 3.2\% & (0.2\%) & 8.0\% & (0.3\%) & 7.6\% & (0.3\%) \\
-\textit{College graduates, 20--50} & 8.3\% & 8.6\% & (0.3\%) & 3.9\% & (0.3\%) & 3.5\% & (0.3\%) \\\hline\hline
-\multicolumn{8}{|p{0.9\linewidth}|}{\footnotesize \textbf{Income partition:} subsample of household heads/spouses living with children and nobody else.}\\\hline
-\textit{Low 30\% income, 20--40} & 20.3\% & 5.1\% & (0.2\%) & 10.3\% & (0.3\%) & \multicolumn{2}{c|}{---}\\
-\textit{Mid 40\% income, 20--40} & 10.1\% & 4.3\% & (0.2\%) & 2.8\% & (0.2\%) & \multicolumn{2}{c|}{---}\\
-\textit{Top 30\% income, 20--40} & 9.0\% & 8.3\% & (0.3\%) & 2.7\% & (0.4\%) & \multicolumn{2}{c|}{---}\\
-\hline
-\end{tabular}
-\caption{Differences, controlled for composition, ACS\label{diff-contr}}
-\end{center}
-\end{table}
-\subsection{Additional Comparisons}
-
-[I pick recently divorced people in ACS, I compare marriage durations for step and non-step children.
-I also check share of divorced in $M\to K$ and $K\to M$ given child's age and given marriage duration.]
-
-\subsection{Intra-Household Allocations}
-
-[I look at earnings, hours and LFP differences] [Grade retention depends on child's age, so need to compare given child's age]
-\section{Model}
-The model is a mixture between conventional heterogeneous agents savings model (B--A--H framework) and household formation and dynamic bargaining, that is extensively developed by [...].
-
-Agents live for $T$ periods.
-The agents are males, females and married couples.
-Couples can have a child together, and if a couple with a child divorces up, the child stays with mother, and this is how singe females can have kids.
-Single females with kids randomly re-enter marriage market by becoming single females without kids.\footnote{The reason for this assumption is to avoid modeling mixture between step and own children, and the fact that single mothers are not a main focus of the paper.}
-
-Single people without kids randomly meet partners with characteristics similar to them and decide whether to create a couple.
-Couples agree on some marriage terms, represented by each spouse's weight in couple's objective function.
-The rationale for marriage includes risk-sharing, returns to scale in consumption, possibility to have a child in the future and additional utility surplus of being married (love shock).
-Each period after random shocks are realized, couples may decide to renegotiate their marriage terms if one of the spouses is better off being single.
-If they cannot find marriage terms that are satisfactory for both, divorce happens.
-
-Couples can choose to have a child and can also experience unplanned pregnancies.
-Unplanned pregnancies cannot be aborted.
-In addition to couples, single people who met each other may experience an unplanned pregnancy just before they decide whether to create a couple.
-In this case, upon disagreement female becomes a single mother.
-This mechanism is a stylized representation of kids appearing in unmarried cohabitating couples: since I do not model cohabitation, those kids appear just before people actually decide to marry each other.\textit{ I discuss this issue additionally in Section ....}
-
-\subsection{Remarks on Notation}
-I write $V^{i,j}$ to represent individual value functions in the model.
-Here $i$ represents gender: $f$ for female or $m$ for male.
-$j$ represents demographic status: $s$ for singles without kids, $sk$ for singles with kids (females only), $c$ for couples without kids and $ck$ for couples with kids.
-Couple's joint value functions have only one superscript $j \in \{c,ck\}$.
-Relation between joint value function $V^{c}$ and each spouse's value functions $V^{m,c}$ and $V^{f,c}$ is described in details in couple's section.
-
-\subsection{Singles Without Kids}
-Singles without kids are characterized by age $t$, labor productivity $z$ and assets (savings) $a$. I use $\omega = \{z,a\}$ to individual characteristics other than age.
-
-Each period single individuals meet a partner of the same age and characteristics $\omega^p$ with probability $p^{\text{meet}}_t$.
-If meeting happens, with probability $p^{\text{preg}}_t$ unplanned pregnancy happens.
-After realizing new period's shocks and pregnancy status couple tries to negotiate marriage terms.
-If they found marriage terms that are satisfactory for both ($m^{np} = 1$ in case of no pregnancy or $m^p = 1$ in case of pregnancy), they form couple with characteristics $\Omega^c$ .
-If satisfactory marriage terms could not be found, the individual stays single, or becomes single mother if she is female and unplanned pregnancy happened.
-
-The value function of male agent therefore is
-\begin{align}V^{m,s}_t(\omega) = \max\limits_{c} & \bigg\{ u(c) + \beta \E_t \Big[ (1 - p^{\text{meet}}_t)\cdot V^{m,s}_{t+1}(\omega') + \\ \nonumber
-& \hspace{7em} p^{\text{meet}}_t (1-p^{\text{preg}}_t) \big\{ m^{np} \cdot V^{m,c}_{t+1}(\Omega^c) + (1-m^{np})V^{m,s}_t(\omega')\big\} + \\  \nonumber
-& \hspace{10em} p^{\text{meet}}_t p^{\text{preg}}_t \big\{ m^{p} \cdot V^{m,ck}_{t+1}(\Omega^c) + (1-m^{p})V^{m,s}_{t+1}(\omega')\big\}  \Big]  \bigg\},\\  \nonumber
- \end{align}\vspace{-3em}
- \begin{align*}
- \text{s.t. \ }  &  a' = R\cdot a  + W^m_t(z) - c  & \text{ (evolution of assets)}\\
- &  z' = z + \varepsilon^{z,m}_t, \ \ \varepsilon^{z,m}_t \sim \mathcal{N}(0;\sigma_{z,m}^2) &  \text{ (evolution of productivity)}\\
-  & \log W^m_t = z_t + \text{Trend}^m_t, \ \ \  \text{Trend}^m_t = a^m_0 + a^m_1\cdot t  +  a^m_2 \cdot t^2 &  \text{ (labor income and trend)}\\
-  & \Omega^c = \mathcal{M}^m(a',z') &  \text{ (marriage prospectives)}
-\end{align*}
-where random function $\mathcal{M}^m$, representing characteristics of potential couple is defined in Section [...].
-
-The value function of female agent is
-\begin{align}V^{f,s}_t(\omega) = \max\limits_{c} & \bigg\{ u(c) + \beta \E_t \Big[ (1 - p^{\text{meet}}_t)\cdot V^{f,s}_{t+1}(\omega') + \label{single-fem} \\  \nonumber
-& \hspace{7em} p^{\text{meet}}_t (1-p^{\text{preg}}_t) \big\{ m^{np} \cdot V^{f,c}_{t+1}(\Omega^c) + (1-m^{np})V^{f,s}_t(\omega')\big\} + \\  \nonumber
-& \hspace{10em} p^{\text{meet}}_t p^{\text{preg}}_t \big\{ m^{p} \cdot V^{f,ck}_{t+1}(\Omega^c) + (1-m^{p})V^{f,sk}_{t+1}(\omega')\big\}  \Big]  \bigg\},
-\end{align}\vspace{-3em}
-\begin{align*}
- \text{s.t. \ }  &  a' = R\cdot a  + W^f_t(z) - c  & \text{ (evolution of assets)}\\
- &  z' = z + \varepsilon^{z,f}_t, \ \ \varepsilon^{z,f}_t \sim \mathcal{N}(0;\sigma_{z,f}^2) &  \text{ (evolution of productivity)}\\
-  & \log W^f_t = z_t + \text{Trend}^f_t, \ \ \  \text{Trend}^f_t = a^f_0 + a^f_1\cdot t  +  a^f_2 \cdot t^2 &  \text{ (labor income and trend)}\\
-  & \Omega^c = \mathcal{M}^f(a',z') &  \text{ (marriage prospectives)}
-\end{align*}
-
-% 
-The crucial difference here is the term $V^{f,sk}$: it reflects the fact that upon disagreement woman becomes a single female with a child.
-
-\subsection{Couples Without Kids}
-Couple is characterized by age $t$ (assumed the same for both spouses), marriage terms $\theta$, productivities of female and male $z^f$, $z^m$ and additive marriage surplus $\psi$.
-Couple maximizes weighted expected lifetime utility, and $\theta$ represents share of female in this objective function, therefore $\Omega = \{\theta,\psi,z^f,z^m\}$ represent characteristics of couple.
-
-The key element shaping couple's decisions is participation constraints: each period the expected lifetime utility of staying in couple should be not less that expected lifetime utility of getting a divorce and becoming a single agent.
-Depending on realization of the shocks, if it is possible to satisfy both participation constraints the next period (so no divorce happens, $d = 0$) state is $\Omega'$.
-If participation constraints cannot be satisfied, the couple gets a divorce $d = 1$ and its characteristics are dissolved onto $\omega^{df}$ and $\omega^{dm}$.
-In more details this is described in Section [...].
-
-Fertility dimension is an important element.
-Each period couples get fertility shock $p^{\text{preg}}_t$.
-If it happens, they become couple with a child if they stay together (renegotiation happens after they know realization of the shock), or become single male and single female with a child if they do not.
-If the shock does not happen, they still may make a decision to get a baby \emph{after} they renegotiate the marriage terms if they stay together.
-This timing is subtle, but is required for everything to be internally consistent, see Appendix ...
-for more details on how timing interacts with renegotiation.
-
-After realization of all the shocks and making fertility and divorce decisions in the current period the problem of the couple that stays childless is
-\begin{align}& \hspace{5em}  V^{c}_t(\Omega) = \max\limits_{c^f,c^m}  \bigg\{ \theta\cdot u(c^f) + (1-\theta)\cdot u(c^m) + \psi +  \label{vf-c} \\   \nonumber
- &  \beta \E_t \Big[   (1 - p^{\text{preg}}_t)\cdot \left\{ (1-d^{\text{np}})\cdot \max\left\{ V^{ck}_{t+1}(\Omega'),V^{c}_{t+1}(\Omega')\right\} + d^{\text{np}}\cdot [ \theta V_{t+1}^{f,s}(\omega^{df}) + (1-\theta)V_{t+1}^{m,s}(\omega^{dm})]\right\}  +  \\  \nonumber
-& \hspace{5em} p^{\text{preg}}_t\cdot \left\{ (1-d^{\text{p}})\cdot V^{ck}_{t+1}(\Omega') + d^{\text{p}}\cdot [ \theta V_{t+1}^{f,sk}(\omega^{df}) + (1-\theta)V_{t+1}^{m,s}(\omega^{dm})]\right\} \Big] \bigg\},
-\end{align}\vspace{-2em}
-\begin{align*}
-\text{s.t. \ }& a' + c = R\cdot a  + W^m_t(z^m) + W^f_t(z^f) & \text{(evolution of joint assets)},\\
-				 & c = [(c^f)^{1+\rho_c} + (c^m)^{1+\rho_c}]^{\frac1{1+\rho_c}} & \text{(increasing returns in consumption)},\\
-				 &  z^{f\prime} = z^f + \varepsilon^{z,f}_t, \ \ \varepsilon^{z,f}_t \sim \mathcal{N}(0;\sigma_{z,f}^2) &  \text{ (evolution of female productivity)}\\
-				 &  z^{m\prime} = z^m + \varepsilon^{z,m}_t, \ \ \varepsilon^{z,m}_t \sim \mathcal{N}(0;\sigma_{z,m}^2) &  \text{ (evolution of male productivity)}\\
-                    & \psi' = \psi + \varepsilon^{\psi}_t, \ \ \varepsilon^{\psi}_t \sim \mathcal{N}(0;\sigma_{\psi}^2)  & \text{(evolution of marriage surplus),} \\
-                    & (\Omega',\omega^{df},\omega^{dm}) = \mathcal{R}(\theta,a',z^{m\prime},z^{f\prime},\psi') & \text{(renegotiation correspondence)},
-\end{align*}
-
-\subsubsection{Defining Individual Values}
-The couple's collective value function $V^c$ is the object relevant for making decisions of consumption and fertility.
-Making marriage and divorce decisions involves individual value functions of spouses living in the couple.
-I denote them as $V^{f,c}$ and $V^{m,c}$.
-These value functions are obtained by plugging optimal decisions of couple into intertemporal utilities of each agent and accounting for possible transitions.
-Note that they cannot be derived from a maximization problem, and common properties of value functions such as envelope theorems do not hold for them.
-
-Namely, given couple's decisions $c^f(\Omega)$, female's value of being in couple is defined recursively as
-\begin{align}
-& \hspace{5em}  V^{f,c}_t(\Omega) =    u(c^f) + \psi +  \\   \nonumber
- &  \beta \E_t \Big[   (1 - p^{\text{preg}}_t)\cdot \left\{ (1-d^{\text{np}})\cdot \max{}^C \left\{ V^{f,ck}_{t+1}(\Omega'),V^{f,c}_{t+1}(\Omega')\right\} + d^{\text{np}}\cdot [ V_{t+1}^{f,s}(\omega^{df})]\right\}  +  \\  \nonumber
-& \hspace{5em} p^{\text{preg}}_t\cdot \left\{ (1-d^{\text{p}})\cdot V^{f,ck}_{t+1}(\Omega') + d^{\text{p}}\cdot [ V_{t+1}^{f,sk}(\omega^{df}) ]\right\} \Big] 
-\end{align}
-where operator $\max{}^C\{A,B\} \equiv A\cdot \I[V^{ck}_{t+1}(\Omega')\geq V^{c}_{t+1}(\Omega')] + B\cdot \I[V^{ck}_{t+1}(\Omega')< V^{c}_{t+1}(\Omega')]$.
-It reflects the fact that couple makes joint decision that is not necessary individually optimal.
-
-Another important property of individual values is that
-\begin{equation} V^{c}_t(\Omega) \neq \theta\cdot V^{f,c}_t(\Omega) + (1-\theta)\cdot V^{m,c}_t(\Omega),\label{tht_noneq}\end{equation}
-the main reason for this is that $\theta$ changes in the future as a result of random shocks, and this change is not likely to be symmetric.
-This, in particular, causes couple's expected future value function to be discontinuous in $(a,z)$.
-See Appendix [...] for some additional discussion.
-
-\subsection{Couples With Kids}
-State space for couples is the same as for singles, except for additional variable $\xi$ representing child's age, therefore $\Omega = \{\theta,\psi,z^f,z^m,\xi\}$.
-I consider just two states for $\xi \in \{y,s\}$ representing young and school-age child.
-At birth the child is young, then each period with probability $p^{s}$ the child becomes of school-age.
-The difference between young and school-age children are mother's time requirement, namely young children require more time and elder --- more money.
-
-Couples with kids have additional choice variable $q$, that can be interpreted as flow of child quality or child consumption.
-Child quality is produced using mother's labor $l_f$ and monetary expenditures $x$ according to constant returns to scale production function.
-Labor input to child quality causes female to lose part of her labor income.
-The production function depends on age, namely I parametrize it as $q = [\mu_\xi\cdot l_f^{\theta_q} + (1-\mu_\xi)\cdot x^{\theta_q}]^{\frac1{\theta_q}}$.
-This follows, for instance, [Sommer ...], with the addition of changing labor share $\mu$.
-
-\begin{align}& \hspace{5em}  V^{ck}_t(\Omega) = \max\limits_{c^f,c^m,q,l_f,x}  \bigg\{ \theta\cdot u(c^f,q) + (1-\theta)\cdot u(c^m,q) + \psi + \label{vf_ck} \\  \nonumber
- & \hspace{10em} \beta \E_t \Big[   \left\{ (1-d)\cdot   V^{ck}_{t+1}(\Omega') + d\cdot [ \theta V_{t+1}^{f,sk}(\omega^{df}) + (1-\theta)V_{t+1}^{m,s}(\omega^{dm})]\right\} \Big] \bigg\},
-\end{align}\vspace{-2em}
-\begin{align*}
-\text{s.t. \ } & a' + c + x = R\cdot a  + W^m_t(z^m) + (1-l_f)\cdot W^f_t(z^f) & \text{(evolution of joint assets)},\\
-                    & c = [(c^f)^{1+\rho_c} + (c^m)^{1+\rho_c}]^{\frac1{1+\rho_c}} & \text{(increasing returns in consumption)},\\
-                    & q = [\mu_\xi\cdot l_f^{\theta_q} + (1-\mu_\xi)\cdot x^{\theta_q}]^{\frac1{\theta_q}} & \text{(production function of child quality)},\\
-                    &  z^{f\prime} = z^f + \varepsilon^{z,f}_t, \ \ \varepsilon^{z,f}_t \sim \mathcal{N}(0;\sigma_{z,f}^2) &  \text{ (evolution of female productivity)}\\
-				 &  z^{m\prime} = z^m + \varepsilon^{z,m}_t, \ \ \varepsilon^{z,m}_t \sim \mathcal{N}(0;\sigma_{z,m}^2) &  \text{ (evolution of male productivity)}\\
-                    & \psi' = \psi + \varepsilon^{\psi}_t, \ \ \varepsilon^{\psi}_t \sim \mathcal{N}(0;\sigma_{\psi}^2)  & \text{(evolution of marriage surplus),} \\
-                   &  \P(\xi' = s | \xi = y) = p^s, \ \ \P(\xi' = s | \xi = s) = 1 & \text{(evolution of child's age)},\\
-                    & (\Omega',\omega^{df},\omega^{dm}) = \mathcal{R}(\theta,a',z^{m\prime},z^{f\prime},\psi',\xi) & \text{(renegotiation correspondence)},
-\end{align*}
-
-\subsection{Single Mothers}
-State space for single mothers contains individual productivity and child's age.
-Single mothers cannot marry, but have chance $p^{\text{out}}$ of recovering their marriage prospectives.
-In the model this is captured by transition of them to single females.
-This form is somewhat restrictive, although any other option requires modeling marriage of single mothers separately.
-This is a complex issue, for instance, males would seek females who already have kids if we do not make difference between step and own children.
-Since for the purposes of the model option of being single mother is used mainly as a factor affecting bargaining, I introduce additional (constant) utility factor $S$ to flexibly control utility of exiting the state.
-This can be interpreted as equivalent of utility of kids from past marriages when entering the marriage market again.
-This form seems flexible enough to capture variation in an option of being a single mother without putting extra computational and modeling burden.
-
-The problem of a female who stays single mother is
-\begin{align}V^{f,sk}_t(\omega) = \max\limits_{c} & \bigg\{ u(c,q) + \beta \E_t \Big[ p^{\text{out}} \cdot \left\{ V^{f,s}_{t+1}(\omega') + S\right\} +(1-p^{\text{out}})V^{f,sk}_{t+1}(\omega') \Big]  \bigg\},
-\end{align}\vspace{-1em}
-\begin{align*}
- \text{s.t. \ }  &  a' = R\cdot a  + (1-l_f)W^f_t(z) - c  & \text{ (evolution of assets)}\\
- & q = [\mu_\xi\cdot l_f^{\theta_q} + (1-\mu_\xi)\cdot x^{\theta_q}]^{\frac1{\theta_q}} & \text{(production function of child quality)},\\
- &  z' = z + \varepsilon^{z,f}_t, \ \ \varepsilon^{z,f}_t \sim \mathcal{N}(0;\sigma_{z,f}^2) &  \text{ (evolution of productivity)}\\
- &  \P(\xi' = s | \xi = y) = p^s, \ \ \P(\xi' = s | \xi = s) = 1 & \text{(evolution of child's age)},\\
-\end{align*}
-
-\subsection{Marriage Market}
-Single males and females meet potential partners of identical age and characteristics that depend on their assets $a$ (chosen in the previous period) and productivity $z$ (revealed after realization of shock):
-\begin{equation}\log a^p = \log a + \varepsilon^{a,p}, \ \ \varepsilon^{a,p} \sim \mathcal{N}(0,\sigma_{a,p}^2),\end{equation}
-\begin{equation}	z^p = z + \varepsilon^{z,p}, \ \ \varepsilon^{z,p} \sim \mathcal{N}(0,\sigma_{z,p}^2),\end{equation}
-for example, $\sigma_{a,p} = 0.1$ can be read as standard deviation of partner's assets being 10\% around of what agent has. 
-Additionally, initial marriage surplus (love shock) is drawn from
-\begin{equation} \psi \sim \mathcal{N}(0,\sigma_{\psi,0}^2).\end{equation}
-
-If people agree to be a couple they pool their assets $a^c = a + a^p$. If no unplanned pregnancy happens, people agree to marry if set of mutually satisfactory marriage terms
-\begin{equation}\small \Theta^{np}_t = \left\{ \theta : V_t^{f,c}(\theta,\psi,a^c,z^f,z^m) \geq V_t^{f,s}(a^f,z^f), \ \ V_t^{m,c}(\theta,\psi_0,a^c,z^f,z^m) \geq V_t^{m,s}(a^m,z^m)\right\}\end{equation}
-is non-empty. If potential couple gets the pregnancy shock, the set changes to
-\begin{equation} \small \Theta^{p}_t = \left\{ \theta : V_t^{f,ck}(\theta,\psi,a^c,z^f,z^m,y) \geq V_t^{f,sk}(a^f,z^f,y), \ \ V_t^{m,ck}(\theta,\psi_0,a^c,z^f,z^m,y) \geq V_t^{m,s}(a^m,z^m)\right\} \end{equation}
-(at birth child's age is $\xi = y$).
-
-If the respective set is non-empty, initial bargaining power of couple is determined by symmetric Nash Bargaining, for instance:
-\small
-\begin{align}&\theta^*(a,z,\psi,\epsilon^{a,p},\epsilon^{z,p}) = \label{nbs} \\ \nonumber &= \argmax\limits_{\theta \in \Theta^{np}} \left[V_t^{f,c}(\theta,\psi,a^c,z^f,z^m) - V_t^{f,s}(a^f,z^f)\right] \times \left[V_t^{\vphantom{f}m,c}(\theta,\psi,a^c,z^f,z^m) - V_t^{m,s}(a^m,z^m)\right],\end{align}
-and analogously for the case when pregnancy shock happened.
-
-From the prospective of a single agent, this set is determined by realization of $\psi$ and partner's characteristics $\epsilon^{a,p}$ and $\epsilon^{z,p}$.
-Given them, marriage function $m$ is binary:
-\[m^{np}_t(a,z,\psi,\epsilon^{a,p},\epsilon^{z,p}) = \I(\Theta^{np}_t \neq \varnothing),\]
-where variables affecting set $\Theta$ are unambiguously recovered from arguments of $(a,z,\psi,\epsilon^{a,p},\epsilon^{z,p})$ as described above.
-
-%Because of good properties of Nash Bargaining, expected future value function is continuous with respect to of $a$ and $z$. See Appendix [...] for additional discussion of this.
-
-Let $\omega = (a,z)$ and $\epsilon = (\psi,\epsilon^{a,p},\epsilon^{z,p})$. This allows to define a random function
-\[\Omega^c = \mathcal{M}_t(\omega) = M_t(\omega,\epsilon) = (\theta^*,\psi,a^c,z^m,z^f),\]
-so the future characteristics of couple are function of current characteristics of individual and three-dimensional random shock.
-\subsection{Renegotiation And Divorce}
-
-Upon divorce, $\kappa\cdot a$ of assets disappears and the rest is split evenly.
-This generalizes [Voena ...], where I put additional parameter $\kappa$ captures can rationalize extra costs of divorce for more wealthy couples.
-[Voena ...] argues that this approximates well actual property divisions.
-So, the splitting rule is
-\[a^{df} = 0.5\cdot (1-\kappa)\cdot a,  \ \ a^{dm} = 0.5\cdot (1-\kappa)\cdot a.\]
-This splitting rule defines mapping of couple's state $\Omega$ to individual states $\{\omega^{df},\omega^{dm}\}$ in case of divorce.
-
-The option to leave with a share of couple's assets creates pressure for participation constraints.
-As argues by [...], the household prefers to keep $\theta$ constant.
-However, after shocks to $z$ and $\psi$ happen, one (or both) participation constraints might be violated.
-To address this formally, consider couple that already has kids.
-I define set
-\begin{equation}\Theta_t = \left\{ \tilde\theta : V^{f,ck}_t(a,\tilde\theta,\psi,z^f,z^m,\xi) \geq V_t^{f,sk}(a^{df},z^f,\xi) , \ \ V_t^{m,ck}(a,\tilde\theta,\psi,z^f,z^m,\xi)\geq V_t^{m,s}(a^{dm},z^m) \right\} \label{reneg-with-kids}
-,\end{equation}
-and if couple starts with current value $\theta_{0}$ there are three possible options:
-\begin{enumerate}
-\item \textit{Status quo.} If $\theta_0 \in \Theta_t$, current marriage terms are satisfactory for both spouses so $\theta_t = \theta_{0}$.
-\item \textit{Renegotiation.} If $\theta_0 \not\in \Theta_t$, but $\Theta_t \neq \varnothing$, then spouses can continue to be married if they change their marriage terms.
-In this case they pick the closest the new value from set $\Theta_t$ such that it is the closest to the old marriage terms: $\theta_t = \min\limits_{\tilde\theta\in\Theta_t} |\tilde\theta - \theta_0|$.
-\item \textit{Divorce.}  If $\Theta_t = \varnothing$, so if there do not exist any marriage terms satisfactory for both spouses, then divorce happens according to the rule described above.
-\end{enumerate}
-Picking $\theta$ that is the closest to the old one is not a random choice: it is rationalized by models with limited commitment models, where change in $\theta$ represents the value of Lagrange multiplier on binding participation constraint.
-Intuitively, since couple does not like $\theta$ to be changing and everything is continuous in $\theta$, the couple prefers small changes to larger ones, and this drives the result.
-
-Renegotiation that involves making discrete decisions has few more issues as backward induction has to be used: fertility decisions have to be made after new $\theta$ is set, and these decisions might depend on $\theta$, therefore inequalities defining set $\Theta_t$ need to be modified.
-This is an artifact of assuming that fertility decisions are made collectively and it can be the case that having a baby benefits couple but hurts wife alone (mainly by worsening her outside option), and such scenario of renegotiation insures that participation constraints hold after the fertility decisions are made.
-See Appendix \ref{ren-disc} where I elaborate on this.
-
-\subsection{Additional Details}
-The model I solve an estimate contains few additional things that are omitted for more clear exposition above.
-
-First, females may exit the state of being single mother immediately.
-This transition is assumed to happen after divorce or failure to agree to marry, therefore in the value functions and in (re)negotiation I use $V^{f,s?} = p^{\text{out}} [V^{f,s} + S] + (1-p^{\text{out}})V^{f,sk}$ instead of just $V^{f,sk}$.
-This allows for quicker re-marriage and easier implementation of counterfactual in which females do not have risk of becoming single mother at all.
-I use notation of $V^{f,s?}$ in Appendix \ref{add-trans} where I show exact value functions.
-
-Second, I assume that having a baby is possible only before certain age $\bar{T}$. After that, value functions and negotiation rules are modified in a straightforward manner.
-
-Finally, to simplify computations I also do not allow marriage status and marriage terms to change after certain $\bar{T}_2$.
-Therefore I drop participation constraints in the couples, the value functions look the same but do not allow for $\theta$ to change.
-
-\subsection{Functional Forms}
-I use standard CES utility functions for individuals and couples with no kids: $u(c) = \frac{c^{1-\sigma}}{1-\sigma}$.
-When kids appear, people extract additional utility of child quality, I use the following non-separable utility function:
-\[u(c,q) = \frac{\left(e^{\phi_0}\cdot c^{1-\phi_1} \cdot (\phi_2 + q)^{\phi_1}\right)^{1-\sigma}}{1-\sigma},\]
-parameter $\phi_0$ represents fixed gain in marginal utility because of having a baby, $\phi_1$ drives relative importance of child quality in utility function and $\phi_2$ is a ``luxury good'' parameter that can rationalize large child expenditures of wealthy couples that are discussed in the estimation section.
-
-\section{Estimation Strategy}
-I use techniques of indirect inference, extensively described, for instance, in [...] for estimation of the parameters.
-Shortly, I select parameters such that features of the real data are similar to the features of the data simulated from model.
-It is not entirely correct to call the procedure simulated method of moments, as some things that I match are not moments in econometric sense, however it has many similarities with simulated method of moments so I occasionally extend notion of moments to reflect any statistics derived from the data.
-
-\subsection{Overview}
-Full list of parameters involves many things:
-\begin{itemize}
-\item Fundamental parameters of income fluctuation model $(\beta,\sigma,R)$
-\item Intra-household returns to scale $\rho_c$
-\item Preferences for child's quality: $(\phi_0,\phi_1,\phi_2)$
-\item Parameters for child's quality production function: CES power $\theta_k$ and labor shares by age $(\mu_y,\mu_s)$.
-\item Transition probabilities:
-\begin{itemize}
-\item Probability of pregnancy shock by age: $p^{\text{preg}}_t = \lambda^{\text{preg}}_0 - \lambda^{\text{preg}}_1\cdot t$ (2 parameters)
-\item Probability of meeting a partner: $p^{\text{meet}}_t:$ piecewise linear function between $(p^{\text{meet}}_{21},p^{\text{meet}}_{27},p^{\text{meet}}_{34},p^{\text{meet}}_{40})$ (4 parameters)
-\item Probability of re-entering marriage market and corresponding utility ``compensation'' $(p^{\text{out}},S)$. (2 parameters)
-\item Probability that kids grow up (transition from $y$ to $s$): $p_s$ (1 parameter)
-\end{itemize}
-\item Age profiles of labor income by gender: $(a^f_0,a^f_1,a^f_2)$, $(a^m_0,a^m_1,a^m_2)$ (6 parameters)
-\item Variance of shocks (6 parameters):
-\begin{itemize}
-\item Individual income shocks $\sigma^{z,f}$, $\sigma^{z,m}$.
-\item Couple's surplus shocks $\sigma^{\psi}$
-\item Couple's initial surplus $\sigma^{\psi_0}$
-\item Assets and productivity of potential partner: $(\sigma^{z,p},\sigma^{a,p})$
-\end{itemize}
-\end{itemize}
-
-Total number of parameters here is 31. Few additional choices involve time horizons, bounds for random walk variables and bounds for assets and $\theta$.
-
-\subsection{Externally Set Parameters}
-Length of the period is set to be 1 year, the model assumes people start their life with no assets at the age of $21$ and die deterministically at the age of $81$ (so $T = 30$).
-However, it assumes that fertility status can change only till age of 40 ($T_f = 19$) and marriage status till age of 50 ($T_m = 29$).
-It also assumes that after the age of 60 people retire and get $b = 0.4$ [as in Sommers, 2014] of their income at age of 60 (so no transitions happen).
-I stop simulations at the age of 50.
-
-Following standard approach in literature, I fix $\sigma = 1.5$, $R = 1.03$ and $\beta = 1/R$. Also, following estimates of Voena (2014) I set $\rho_c = 0.23$.
-
-I borrow age profiles from $(a^f_0,a^f_1,a^f_2)$, $(a^m_0,a^m_1,a^m_2)$ and variances of income shocks $\sigma^{z,f}$, $\sigma^{z,m}$ from Low et al 2018 [This will be replaced by my own estimates from SIPP].
-
-I also externally estimate $\sigma^{z,p}$ as variance of (unpredicted part of) log-difference in spouses earnings, see Appendix \ref{partearn} for this.
-Based on this calculations I set $\sigma^{z,p} = 0.8$.
-As I both do not have good assets data and do not aim to replicate assets distribution, I set $\sigma^{a,p} = \sigma^{z,p}$.
-
-I also set $p_s = \frac14$ to roughly reflect that children go to school on average at 6 and to pre-school few years earlier and this should correspond to a change in childcare costs structure.
-
-In the part with assets, I assume savings can be between 0 and 20 times average labor income in the beginning of life.
-This bar in pretty low, but it is enough to create reasonable assets variation at the moment most of marriage decisions are made: say, median assets-to-income ratio at 30 is around 1 [see ...].
-
-\subsection{Estimated Parameters}
-Out of 31, 15 parameters are set externally, so I am left with $16$ parameters to estimate: child's parameters $(\phi_0,\phi_1,\phi_2)$, child quality production parameters $(\theta_k,\mu_y,\mu_a)$, variances of initial marriage surplus and its innovations $(\sigma^{\psi_0},\sigma^{\psi})$, meeting probability parameters $(p^{\text{meet}}_{21},p^{\text{meet}}_{27},p^{\text{meet}}_{34},p^{\text{meet}}_{40})$, pregnancy probability parameters $(\lambda^{\text{preg}}_0,\lambda^{\text{preg}}_1)$, transition out of being single mother parameters $p^{\text{out}}$, $S$.
-Unfortunately, all of them interact and as the model is highly stylized but involves many choices, they cannot be estimated externally.
-
-Conceptually, they can be split onto following groups:
-\begin{enumerate}
-\item Unplanned pregnancy probabilities are pinned down by share of $K\to M$ females at ages 25 and 35, as well as share of planned births in total population
-\item Meeting probabilities and $\sigma^{\psi_0}$ are pinned down by shares of never married females people at 25, 35, share of females with one marriage at 40 (relevant to capture chances of re-marrying) and shares of never married at 30 for high and low income groups.
-\item Childcare production function is recovered by ratio of time spent with kids by couple's income (pins down elasticity of substitution), average share of money spent on kids (captures ``average'' $\mu$) and ratio of time spent by child's age (captures difference between $\mu_a$ and $\mu_y$)
-\item  Preferences for kids are captured by share of childless females by age of 25, 30, 35 (general scale), share of childless females for low and high income at 30 $\phi_1$.
-Luxury good parameter $\phi_2$ is pinned down by ratio of amounts of money spent on kids between 75 and 25 income quantiles.
-\item Variance of innovation to marriage surplus is mainly pinned down by share of divorced people by age (25, 35, 40) and income groups (at 30).
-\item Transition out of single mothers is captured by share of never married single mothers at 25 (picks $S$) and share of divorced women with kids at 30 (picks $p^{\text{out}}$).
-\end{enumerate}
-On top of it, I use shares of divorced people for $K\to M$ and $M \to K$ at 40 as additional targets that are affected by mixture of parameters described above.
-
-I additionally impose the following constraints: share of rejected marital offers is within $[0.05,0.95]$ and average share of time spent on kids is below $0.8$.
-[Both of these constraints are non-binding in the solution I found, although turning them off occasionally may lead to implausible solutions.]
-
-\section{Results}
-Table ... shows targeted and
-
-\subsection{Counterfactual Simulations}
-\begin{itemize}
-\item Turn off fertility completely
-\item Turn off bargaining
-\item No pressure of being single mother
-\item Single mother/single father randomly
-\end{itemize}
-
-\newpage
-\section{Robustness Checks}
-\section{Conclusions}
-
-\appendix
-\section{Additional Empirical Results}
-\subsection{ACS}
-
-\section{Additional Robustness Checks}
-
-\section{Additional Notes on Model}
-\subsection{Continuity of Value Functions}
-This part discusses violations of continuity caused by the dynamic bargaining structure.
-The general conclusion is that collective value function does is not necessary continuous in couple's characteristics, and this discontinuity propagates onto all value functions in the model.
-
-To simplify the example, let's consider value function of couple with kids. Similarly to equation \ref{tht_noneq}, we can write
-\[ V^{ck}_t(\Omega) = \theta\cdot V^{f,ck}_t(\Omega) + (1-\theta)\cdot V^{m,ck}_t(\Omega) + \Delta(\Omega),\]
-and use it to substitute the term under the expectation in equation \ref{vf_ck} 
-\begin{align*} & (1-d)\cdot   V^{ck}_{t+1}(\Omega') + d\cdot [ \theta V_{t+1}^{f,sk}(\omega^{df}) + (1-\theta)V_{t+1}^{m,s}(\omega^{dm})] =    V^{ck}_{t+1}(\Omega') + \\
- &  \ \ \ \theta \cdot d(\Omega') \cdot [V_{t+1}^{f,sk}(\omega^{df}) - V_{t+1}^{f,ck}(\Omega')] + (1-\theta) \cdot d(\Omega') \cdot [V_{t+1}^{m,sk}(\omega^{dm}) - V_{t+1}^{m,ck}(\Omega')] - d(\Omega')\cdot \Delta(\Omega'),
-\end{align*}
-if we consider couple that is close to divorce, so there exist two $\Omega'_1$ and $\Omega'_2$  that are close to each other according to some metric, but $d(\Omega'_2) = 1$ and $d(\Omega'_1) = 0$.
-Note that around the divorce threshold by [...] \textbf{both} spouses should be indifferent between staying on leaving, therefore
-\[ V_{t+1}^{f,sk}(\omega^{df}) - V_{t+1}^{f,ck}(\Omega')\approx 0, \ \ V_{t+1}^{m,sk}(\omega^{dm}) - V_{t+1}^{m,ck}(\Omega') \approx 0,\]
-therefore if $\Delta(\Omega')$ is non-zero in these points, this whole expression jumps where value of $d$ jumps.
-[It is not clear whether it is actually zero or not, but in simulations it does look like it is positive.]
-
-\subsection{Renegotiation With Collective Discrete Decisions\label{ren-disc}}
-This section explains how discrete decisions interact with couple's renegotiation.
-The theory requires participation constraints to be satisfied \textit{after} any discrete choice.
-In the context of fertility decisions, that means that both spouses can predict future decisions and re-negotiate $\theta$ accounting for this prediction (rather than assuming that fertility status does not change).
-
-With slight abuse of notation, let's suppose that couple gets collective value $V^{ck}(\theta,\Omega)$ if it makes a baby and $V^{c}(\theta,\Omega)$ is it does not.
-Given $\theta$ total value before the decision is made is \[V^{c?}(\theta,\Omega) = \max\{ V^{ck}(\theta,\Omega), V^{c}(\theta,\Omega)\}\]
-after that I define
-\[V^{f,c?}(\theta,\Omega) = V^{f,ck}(\theta,\Omega)\cdot \I[V^{ck}(\theta,\Omega) \geq V^{c}(\theta,\Omega)] + V^{f,c}(\theta,\Omega) \cdot \I[V^{ck}(\theta,\Omega) < V^{c}(\theta,\Omega)],\]
-note that it is not necessary equal to $\max\{ V^{f,ck},V^{f,c} \}$. Similarly we can define $V^{m,c?}(\theta,\Omega)$.
-
-The main thing to notice is that although $V^{c?}$ is continuous in $\theta$ individual values $V^{f,c?}$ and $V^{m,c?}$ are not, unless both spouses always have agreement on whether to make a baby: $V^{ck} \geq V^{c} \Leftrightarrow V^{f,ck} \geq V^{f,c} \Leftrightarrow V^{m,ck} \geq V^{m,c}$.
-Therefore set
-\[\Omega = \left\{\tilde{\theta} : V^{f,c?}(\tilde\theta,\Omega) \geq V^{f,s}(\omega^{df}), \ \  V^{m,c?}(\tilde\theta,\Omega) \geq V^{m,s}(\omega^{dm}) \right\}\]
-may be disjoint, and careful approximation of $V$ with respect to $\Omega$ is required.
-
-The possible solution to it would be to limit fertility decisions to the cases where both spouses agree on having kids, this, however, goes against the logic where couple maximizes collective utility subject to participation constraints.
-
-\subsection{Accounting for Additional Transitions\label{add-trans}}
-In value functions I stated above I assumed that females stays single mother for at least one period if it exits a couple with a kid or if she has a baby at the moment of marriage and disagrees to marry.
-In the actual model I relax this assumption so the female can stop being single mother immediately.
-This is needed, for instance, for evaluating counterfactuals in which female can exit couple without pressure of being a single mother.
-
-I assume that at the moment of (re)negotiation female is unsure about whether she is going to be a single mother. Therefore, I define expected value of leaving the couple to be
-\[V_t^{f,c?}(\omega) = p^{\text{out}}[V_t^{f,s}(\omega) + S] + (1-p^{\text{out}})V_t^{f,sk}(\omega),\]
-therefore equaiton \ref{single-fem} becomes
-\begin{align}V^{f,s}_t(\omega) = \max\limits_{c} & \bigg\{ u(c) + \beta \E_t \Big[ (1 - p^{\text{meet}}_t)\cdot V^{f,s}_{t+1}(\omega') + \label{single-fem-true} \\  \nonumber
-& \hspace{7em} p^{\text{meet}}_t (1-p^{\text{preg}}_t) \big\{ m^{np} \cdot V^{f,c}_{t+1}(\Omega^c) + (1-m^{np})V^{f,s}_t(\omega')\big\} + \\  \nonumber
-& \hspace{10em} p^{\text{meet}}_t p^{\text{preg}}_t \big\{ m^{p} \cdot V^{f,ck}_{t+1}(\Omega^c) + (1-m^{p})V^{f,s?}_{t+1}(\omega')\big\}  \Big]  \bigg\},
-\end{align}
-the set is \ref{reneg-with-kids} becomes
-\begin{equation}\Theta_t = \left\{ \tilde\theta : V^{f,ck}_t(a,\tilde\theta,\psi,z^f,z^m,\xi) \geq V_t^{f,s?}(a^{df},z^f,\xi) , \ \ V_t^{m,ck}(a,\tilde\theta,\psi,z^f,z^m,\xi)\geq V_t^{m,s}(a^{dm},z^m) \right\} \label{reneg-with-kids}
-,\end{equation}
-and the analogous term is introduced for initial negotiation of couples in case of pregnancy shock.
-
 \section{Computational Notes}
@@
 [Similar statistic computed on simulated data returns ...]
-
-\newpage
-\section*{Very Technical Appendix, Not For Final Version}
-\section{Expressions for Value Functions}
-\subsection{Notation}
-This technical note shows exact expressions for approximating value functions.
-For the couple's state I use partition $\Omega = (a,O)$, where $O = (\theta,\psi,z^f,z^m)$.
-For the age variable I put $\xi\in\{y,s\}$ counts as a separate letter, so states for couples with kids are noted like $cky$ and $cks$.
-Let $p^s$ represents probability that the child ``grows up'' so couple switches from $cky$ to $cks$.
-For individual's state there are only two variables $\omega = (a,z)$, and child's age for single mother is reported as state $sky$ if $\xi = y$ and $sks$ if $\xi = y$.
-
-Grid consists of all possible combinations of $a$ and $O$ ($a$ and $z$ in single's case), so \[G = \{ \{a_1,...,a_I\} \times \{o_1,...,o_J\}\}\]
-
-Let $i$ refers to assets position and $j$ refers to values of everything else. Finally, indices $q$ refer to quadrature nodes, and $w_q$ to quadrature weights that add up to 1.
-
-\subsection{Single Agents}
-Consider females, males are handled analogously. Value function \ref{single-fem} can be written as
-\[V^{f,s}_t(a,z) = \max\limits_{c} \left\{ u(c) + \beta \mathcal{V}^{f,s}_{t+1}(s,z)\right\}, \ \ c + s = M(a,z),\]
-Here  $\mathcal{V}^{f,s}_{t+1}$ represents expectation of tomorrow's individual value function, accounting for all possible transitions.
-Argument $s$ refers to savings made in the current period (it is not necessary equal to future assets as future assets change if the partner is met), and $z$ refers to current value of productivity.
-Approximating $\mathcal{V}_{t+1}(s,z)$ (i.e.
-integration) is the key challenge.
-Note that it does not depend on current assets, only on savings $s$ that define future assets position.
-$M(a,z) = R\cdot a + \exp(z + \text{Trend}_t)$ is the amount of money in the end of the period.
-
-I treat the function $\mathcal{V}^{f,s}_{t+1}(s,z)$ as a collection of functions $\mathcal{V}^{f,s}_{mj,t+1}$ for each value of current productivity $z_j$ and savings $s_m$.
-I generate quadrature nodes that consist of shock to individual productivity $\epsilon^{z,f}$, shocks to partner's assets and productivity positions
-$(\epsilon^{a,p},\epsilon^{z,p})$ and initial couple's surplus $\psi$.
-I denote these nodes as $x_q = (\epsilon^{z,f}_q,\epsilon^{a,p},\epsilon^{z,p},\psi_q)$, in total monomial rule results in $Q = [...]$ nodes.
-Weight $w_q$ corresponds to each node.
-Given savings $s$ and grid point in $z_j$ I generate partner's and couple's characteristics.
-\[a^p_{mq} = s_m\cdot \exp(\epsilon^{a,p}_q), \ \ a^c_{mq} = s_m\cdot [1 + \exp(\epsilon^{a,p}_q)], \ \ z^m_{jq} = z^f_j \exp(\epsilon^{z,p}_q),\]
-(in some sense in this setup by saving more individuals have chances of getting more wealthy partners).
-Also, individual next period's productivity is generated according to $z^f_{jq} = z^f\cdot \exp(\epsilon^{z,f}_q)$.
-Note that $z^f_j$ refers to the current value and $z^f_{jq}$ refers to the next period's values;
-
-This allows to generate potential couple's characteristics $(a^c_{mq},\psi_q,z^{f}_{jq},z^{m}_{jq})$, and combining it with value functions for the next period we can define:
-\[\Theta^{np}_{mjq,t+1} = \left\{ \tilde\theta : V^{f,c}_{t+1}(a^c_{mjq}, \tilde\theta, \psi_q,z^{f}_{jq},z^{m}_{jq}) \geq V^{f,s}(s_m,z^f_{jq}), \ V^{m,c}_{t+1}(a^c_{mjq}, \tilde\theta, \psi_q,z^{f}_{jq},z^{m}_{jq}) \geq V^{m,s}(a^p_{mq},z^m_{jq}) \right\},\]
-if $\Theta^{np}_{mjq,t+1}$ is empty\footnote{I figure this out by defining a uniformly spaced grid for $\theta \in [0.05,0.95]$.
-Because of Smolyak grid, this can be done very efficiently as $V^{f,c}_{t+1}(a^c_{mjq}, \tilde\theta, \psi_q,z^{f}_{jq},z^{m}_{jq})$ are just a polynomial in $\tilde{\theta}$.}
-, $m^{np}_{mjq,t+1} = 0$ (so no marriage happens at node $jq$ if agents made savings of $s_m$).
-If the set is not empty, then $\theta_{mjq}$ is determined by Nash Bargaining \ref{nbs} and $m^{np}_{mjq,t+1} = 1$.
-Therefore at node $(s_m,z_j)$ future value function in case the partner is met and pregnancy did not happen can be approximated as
-\[\E_t  \big\{ m^{np} \cdot V^{f,c}_{t+1}(\Omega^c) + (1-m^{np})V^{f,s}_t(\omega')\big\} \approx \mathcal{E}^{np}(s_m,z_j),\]
-where
-\[\mathcal{E}^{np}(s_m,z_j) = \sum\limits_{q=1}^{Q} w_q \cdot \left\{ m^{np}_{mjq,t+1}\cdot V^{f,c}_{t+1}(a^c_{mq},\theta_{mjq},\psi_q,z^{f\prime}_{jq},z^m_{jq}) + (1-m^{np}_{mjq,t+1})\cdot V^{f,s}_{t+1}(s_m,z_j)\right\}.\]
-In exactly the same manner we define the set in the case the pregnancy shock did happen:
-\[\Theta^{p}_{mjq,t+1} = \left\{ \tilde\theta : V^{f,ck}_{t+1}(a^c_{mjq}, \tilde\theta, \psi_q,z^{f}_j,z^{m}_{jq}) \geq V^{f,s?}(s_m,z^f_{j}), \ V^{m,ck}_{t+1}(a^c_{mjq}, \tilde\theta, \psi_q,z^{f}_{jq},z^{m}_{jq}) \geq V^{m,s}(a^p_{mq},z^m_{jq}) \right\},\]
-where function $V^{f,s?} = p^{\text{out}}[V^{f,s} + S] + (1-p^{\text{out}})V^{f,sky}_t$ captures possible immediate transition out of being single mother. In the same way 
-\begin{align*}\mathcal{E}^{p}(s_m,z_j) = \sum\limits_{q=1}^{Q} w_q \cdot  \Big\{ m^{p}_{mjq,t+1}\cdot V^{f,cky}_{t+1}(a^c_{mq},\theta_{mjq},\psi_q,z^f_j,z^m_{jq}) + (1-m^{p}_{mjq,t+1})\cdot V^{f,s?}(s_m,z_j) \Big\},
-\end{align*}
-lastly, the value of not meeting any partner is simply
-\begin{align*}\mathcal{E}^{0}(s_m,z_j) = \sum\limits_{q=1}^{Q} w_q \cdot  V^{f,s}_{t+1}(s_m,z^f_{jq}),
-\end{align*}
-finally, the whole function $\mathcal{V}^{f,s}_{t+1}(s_m,z_j)$ can be approximated by weighted average
-\[\mathcal{V}^{f,s}_{t+1}(s_m,z_j) \approx (1-p^{\text{meet}}_t) \cdot \mathcal{E}^{0}(s_m,z_j) +  p^{\text{meet}}_t(1-p^{\text{preg}}_t)  \cdot \mathcal{E}^{np}(s_m,z_j) 
-+  p^{\text{meet}}_t p^{\text{preg}}_t  \cdot \mathcal{E}^{p}(s_m,z_j).\]
-
-After that I use linear interpolation in $s$ dimension to approximate $\mathcal{V}^{f,s}_{t+1}(s_m,z_j)$ for arbitrary $s$.
-Therefore I can get approximation of $\mathcal{V}^{f,s}_{t+1}(s,z_j)$.
-As a result, value function in each asset point $a_i$, $z_j$ is obtained by numerically solving
-\[V^{f,s}_t(a_i,z_j) = \max\limits_{s \in [\underline{A},\overline{A}]} \left\{ u(M(a_i,z_j) - s) + \beta \mathcal{V}^{f,s}_{t+1}(s,z_j)\right\}.\]
-I take $\underline{A}$ and $\overline{A}$ the same as upper and lover bound for individual's assets grid (this is important as approximation may be crude).
-I use Matlab's \textbf{fminbnd} to solve for the maximum point, although I also test that results are very similar when I just use discrete grid $s \in \mathcal{S} = M(a_i,z_j) \cdot\{0,...,0.99\}$ with some spacing, though this variation is much slower.
-
-Choice of solver can generically be an issue as because of concerns described above function $\mathcal{V}^{f,s}_{t+1}(s_m,z_j)$ may be non-smooth or even discontinuous.
-
-\subsection{Couples}
-I consider couples that do not have kids yet as they have richer set of possible transitions.
-Following [...] I split household's decision problem on intrahousehold, intratemporal and intertemporal levels, where first two can be solved analytically.
-For intrahousehold, I define
-\[U(c,\theta) = \max\limits_{c^f,c^m} \left\{\theta u(c^f)  + (1-\theta)u(c^m) \right\}, \ \ \text{s.t.} \ \ \ c = \left[ (c^f)^{1+\rho_c} + (c^m)^{1+\rho_c}\right]^{\frac1{1+\rho_c}},\]
-and given CES assumption this function has simple analytical form.
-
-Intratemporal dimension is trivial, as total consumption is the only intratemporal choice.
-
-For intertemporal I can write the problem \ref{vf-c} as
-\[V^{c}_{t+1}(a,\theta,\psi,z^f,z^m) = \max\limits_{c} \left\{ U(c,\theta) + \beta \mathcal{V}_{t+1}(s,\theta,\psi,z^f,z^m) \right\},\]
-where $\mathcal{V}_{t+1}(s,\theta,\psi,z^f,z^m)$ consists of two terms: what happens if pregnancy shock arrives and what if it does not.
-
-If the shock arrives, in the next period couple is in state $cky$, and if it decides to divorce female becomes a single mother (and may quit this state immediately), and male becomes just a single male.
-I approximate $\mathcal{V}_{t+1}(s,\theta,\psi,z^f,z^m)$ here.
-
-I write current state as combination of $(a_i,O_j)$.
-Function $\mathcal{V}_{t+1}(s,\theta,\psi,z^f,z^m)$ depends on current state $O= (\theta,\psi,z^f,z^m)$ and future savings $s$.
-Integration nodes contains $x_q = (\epsilon^{z,f}_q,\epsilon^{z,m}_q,\epsilon^{\psi}_q)$.
-I take grid with respect to current states, so $O_j= (\theta_j,\psi_j,z^f_j,z^m_j) \in O$ and future savings $s \in A$, so I again approximate expected future value function at each combination of $s_m$ and $O_j$, so I compute values $\mathcal{V}_{mj,t+1} = \mathcal{V}_{t+1}(s_m,\theta_j,\psi_j,z^f_j,z^m_j)$.
-Given each $j$, I can compute next period values of $(\psi_{jq},z^f_{jq},z^m_{jq})$ by adding shocks given by quadrature notes to the current point.
-I also can compute divorce options: $\omega^{df}_{mjq} = ( (1-\kappa)\cdot 0.5 \cdot s_m, \ z^f_{mjq})$ and $\omega^{dm}_{mjq} = ( (1-\kappa)\cdot 0.5 \cdot s_m, \ z^m_{mjq})$.
-This allows to define for each quadrature node a renegotiation set:
-\[\Omega^{p}_{mjq,t+1} = \left\{ \tilde\theta \ : \ V^{f,cky}_{t+1}(s_m,\tilde\theta,\psi_{jq},z^f_{jq},z^m_{jq}) \geq V^{f,s?}_{t+1}(\omega^{df}_{mjq}), \ \ V^{m,cky}_{t+1}(s_m,\tilde\theta,\psi_{jq},z^f_{jq},z^m_{jq}) \geq V^{m,s}_{t+1}(\omega^{dm}_{mjq}) \right\},\]
-and then if $\Omega^{p}_{mjq,t+1}$ is not empty, I pick from it next period's $\theta_{mjq}$ closest to initial $\theta_j$ and set $d^{p}_{mjq} = 0$, otherwise divorce happens and $d^{p}_{mjq} = 0$.
-
-Therefore expected future value if pregnancy shock arrives is 
-{\small
-\[\mathcal{E}^{p}(s_m,O_j) = \sum\limits_{q=1}^{Q} w_q \cdot\left\{ d^p_{mjq}\cdot V_{t+1}^{cky}(s_m,\theta_{mjq},\psi_{jq},z^f_{jq},z^m_{jq}) + (1-d^p_{mjq}) [\theta_{j}\cdot V^{f,sky}(\omega^{df}_{mjq}) + (1-\theta_j)V^{m,s}(\omega^{dm}_{mjq})] \right\} \]
-}
-If pregnancy shock does not arrive, then renegotiation should account for discrete decision of whether to have a kids that happens afterwards, and this decision may depend on $\theta$.
-Namely, I define
-\[k_{mjq}(\tilde\theta) = \I\left[ V^{cky}_{t+1}(s_m,\tilde\theta,\psi_{jq},z^f_{jq},z^m_{jq}) \geq V^{c}_{t+1}(s_m,\tilde\theta,\psi_{jq},z^f_{jq},z^m_{jq})\right],\]
-then I define individual value accounting for this choice:
-\[V^{f,c?}_{t+1}(s_m,\tilde\theta,\psi_{jq},z^f_{jq},z^m_{jq}) = k_{mjq}(\tilde\theta)\cdot V^{f,cky}_{t+1}(s_m,\tilde\theta,\psi_{jq},z^f_{jq},z^m_{jq}) + (1-k_{mjq}(\tilde\theta))\cdot V^{f,c}_{t+1}(s_m,\tilde\theta,\psi_{jq},z^f_{jq},z^m_{jq}),\]
-and the renegotiation set is defined as
-\[\Omega^{np}_{mjq,t+1} = \left\{ \tilde\theta \ : \ V^{f,c?}_{t+1}(s_m,\tilde\theta,\psi_{jq},z^f_{jq},z^m_{jq}) \geq V^{f,s}_{t+1}(\omega^{df}_{mjq}), \ \ V^{m,c?}_{t+1}(s_m,\tilde\theta,\psi_{jq},z^f_{jq},z^m_{jq}) \geq V^{m,s}_{t+1}(\omega^{dm}_{mjq}) \right\},\]
-and we define $\theta_{mjq}$ and $d^{np}_{mjq}$ analogously.
-
-Therefore we define expected future value if pregnancy shock does not arrive to be
-{\small
-\[\mathcal{E}^{np}(s_m,O_j) = \sum\limits_{q=1}^{Q} w_q \cdot\left\{ d^{np}_{mjq}\cdot V_{t+1}^{c?}(s_m,\theta_{mjq},\psi_{jq},z^f_{jq},z^m_{jq}) + (1-d^{np}_{mjq}) [\theta_{j}\cdot V^{f,s}(\omega^{df}_{mjq}) + (1-\theta_j)V^{m,s}(\omega^{dm}_{mjq})] \right\} \]
-}
-so we can finally define value of $\mathcal{V}$ at points $s_m$:
-\[\mathcal{V}^{c}_{t+1}(s_m,O_j) = p_t^{\text{preg}}\cdot \mathcal{E}^{p}(s_m,O_j) + (1-p_t^{\text{preg}})\cdot \mathcal{E}^{np}(s_m,O_j),\]
-and then values at arbitrary points $s$ are obtained by linear interpolation. Therefore the ultimate problem is
-\[V^{c}_t(a_i,O_j) = \max\limits_{s \in [\underline{A},\overline{A}]} \left\{ U(M(a_i,O_j) - s,\theta_j) + \beta \mathcal{V}^{c}_{t+1}(s,O_j)\right\}.\]
-
-\subsection{Couples With Kids} 
-Namely, for intrahousehold allocation of consumption I define
-\[U(c,q,\theta) = \max\limits_{c^f,c^m} \left\{\theta u(c^f,q)  + (1-\theta)u(c^m,q) \right\}, \ \ \text{s.t.} \ \ \ c = \left[ (c^f)^{1+\rho_c} + (c^m)^{1+\rho_c}\right]^{\frac1{1+\rho_c}},\]
-and it still has simple tractable form.
-
-For intratemporal level I have to solve
-\[U(m,W^f,\xi,\theta) = \max\limits_{c,q,l_f,x} U(c,q,\theta), \ \ \ \text{s.t.} \ \ \ c + x + l_f\cdot W^f = m, \ \ q =[\mu_\xi\cdot l_f^{\theta_q} + (1-\mu_\xi)\cdot x^{\theta_q}]^{\frac1{\theta_q}}.
-\]
-this function can also be derived analytically given my assumption on functional forms.
-
-After aggregation, the intertemporal problem looks similar to the one for singles:
-\[V^{cky}(a,\theta,\psi,z^f,z^m) = \max\limits_{m} \left\{ U(m,W^f_t(z^f),\xi,\theta) + \beta \mathcal{V}^{ck}_{t+1}(s,\theta,\psi,z^f,z^m,\xi)\right\}, \ \ \text{s.t.} \ \ m + s = M(a,z^f,z^m),\]
-where definition of $\mathcal{V}$ is analogous.
-
-\end{document}

### 2019-05-03 main/draft.tex
Re-organized into few files

@@
 \item Have children present in household, with age of the eldest child below 18
-\item Most likely reside with their eldest child judging by mother's age (I pick 20--40)
+\item Most likely reside with their eldest child judging by mother's age (I pick 21--40)
 \end{enumerate}
@@
 \subsection{Defining Income}
-This part describes how I define income in ACS.
-
-Following development and household literature, when dividing people by income group I use total household income adjusted by OECD equivalence scale.
-The scale assigns weight 1 to the household head, 0.5 to spouse and 0.3 to each child.
-
-As I use the household income as the main variable of interest, I have to restrict attention to couples that live by themselves and single females with children.
-I do this both for ACS and SIPP, and as definition of households wary a little, this may drop out slightly different groups of people.
-
-Formally,for income considerations I take a subsample of females who is marked as household's head or head's spouse, and who do not have any people living in household other than their spouse/unmarried partner and their children.
-This, for instance, excludes multigenerational household and people living with roommates, as total household's income may not be a relevant thing for some types of household composition.
-I cannot define income equivalence scale for more complicated living arrangements, although there are different scales that allow to do this.
-
-To mitigate selectivity, I compute income percentiles in all sample of females with household composition mentioned above, regardless of their marriage and fertility status.
-I use total family income as the key variable.
-To mitigate misreporting I drop observations with total yearly family income below $\$4000$ for ACS (that is about a third of federal the poverty level) or monthly income below $\$300$ for SIPP.
+This part describes how I define income partitions in ACS.
+Shortly, I define income groups based on per hour income of working females, by excluding those who does not work from income considerations.
+In the appendix I try several
 
@@
 \section{Model}
-The model is a mixture between conventional heterogeneous agents savings model (B--A--H framework) and household formation and dynamic bargaining, that is extensively developed by [...].
 
-Agents live for $T$ periods.
-The agents are males, females and married couples.
-Couples can have a child together, and if a couple with a child divorces up, the child stays with mother, and this is how singe females can have kids.
-Single females with kids randomly re-enter marriage market by becoming single females without kids.\footnote{The reason for this assumption is to avoid modeling mixture between step and own children, and the fact that single mothers are not a main focus of the paper.}
-
-Single people without kids randomly meet partners with characteristics similar to them and decide whether to create a couple.
-Couples agree on some marriage terms, represented by each spouse's weight in couple's objective function.
-The rationale for marriage includes risk-sharing, returns to scale in consumption, possibility to have a child in the future and additional utility surplus of being married (love shock).
-Each period after random shocks are realized, couples may decide to renegotiate their marriage terms if one of the spouses is better off being single.
-If they cannot find marriage terms that are satisfactory for both, divorce happens.
-
-Couples can choose to have a child and can also experience unplanned pregnancies.
-Unplanned pregnancies cannot be aborted.
-In addition to couples, single people who met each other may experience an unplanned pregnancy just before they decide whether to create a couple.
-In this case, upon disagreement female becomes a single mother.
-This mechanism is a stylized representation of kids appearing in unmarried cohabitating couples: since I do not model cohabitation, those kids appear just before people actually decide to marry each other.\textit{ I discuss this issue additionally in Section ....}
-
-\subsection{Remarks on Notation}
-I write $V^{i,j}$ to represent individual value functions in the model.
-Here $i$ represents gender: $f$ for female or $m$ for male.
-$j$ represents demographic status: $s$ for singles without kids, $sk$ for singles with kids (females only), $c$ for couples without kids and $ck$ for couples with kids.
-Couple's joint value functions have only one superscript $j \in \{c,ck\}$.
-Relation between joint value function $V^{c}$ and each spouse's value functions $V^{m,c}$ and $V^{f,c}$ is described in details in couple's section.
-
-\subsection{Singles Without Kids}
-Singles without kids are characterized by age $t$, labor productivity $z$ and assets (savings) $a$. I use $\omega = \{z,a\}$ to individual characteristics other than age.
-
-Each period single individuals meet a partner of the same age and characteristics $\omega^p$ with probability $p^{\text{meet}}_t$.
-If meeting happens, with probability $p^{\text{preg}}_t$ unplanned pregnancy happens.
-After realizing new period's shocks and pregnancy status couple tries to negotiate marriage terms.
-If they found marriage terms that are satisfactory for both ($m^{np} = 1$ in case of no pregnancy or $m^p = 1$ in case of pregnancy), they form couple with characteristics $\Omega^c$ .
-If satisfactory marriage terms could not be found, the individual stays single, or becomes single mother if she is female and unplanned pregnancy happened.
-
-The value function of male agent therefore is
-\begin{align}V^{m,s}_t(\omega) = \max\limits_{c} & \bigg\{ u(c) + \beta \E_t \Big[ (1 - p^{\text{meet}}_t)\cdot V^{m,s}_{t+1}(\omega') + \\ \nonumber
-& \hspace{7em} p^{\text{meet}}_t (1-p^{\text{preg}}_t) \big\{ m^{np} \cdot V^{m,c}_{t+1}(\Omega^c) + (1-m^{np})V^{m,s}_t(\omega')\big\} + \\  \nonumber
-& \hspace{10em} p^{\text{meet}}_t p^{\text{preg}}_t \big\{ m^{p} \cdot V^{m,ck}_{t+1}(\Omega^c) + (1-m^{p})V^{m,s}_{t+1}(\omega')\big\}  \Big]  \bigg\},\\  \nonumber
- \end{align}\vspace{-3em}
- \begin{align*}
- \text{s.t. \ }  &  a' = R\cdot a  + W^m_t(z) - c  & \text{ (evolution of assets)}\\
- &  z' = z + \varepsilon^{z,m}_t, \ \ \varepsilon^{z,m}_t \sim \mathcal{N}(0;\sigma_{z,m}^2) &  \text{ (evolution of productivity)}\\
-  & \log W^m_t = z_t + \text{Trend}^m_t, \ \ \  \text{Trend}^m_t = a^m_0 + a^m_1\cdot t  +  a^m_2 \cdot t^2 &  \text{ (labor income and trend)}\\
-  & \Omega^c = \mathcal{M}^m(a',z') &  \text{ (marriage prospectives)}
-\end{align*}
-where random function $\mathcal{M}^m$, representing characteristics of potential couple is defined in Section [...].
-
-The value function of female agent is
-\begin{align}V^{f,s}_t(\omega) = \max\limits_{c} & \bigg\{ u(c) + \beta \E_t \Big[ (1 - p^{\text{meet}}_t)\cdot V^{f,s}_{t+1}(\omega') + \label{single-fem} \\  \nonumber
-& \hspace{7em} p^{\text{meet}}_t (1-p^{\text{preg}}_t) \big\{ m^{np} \cdot V^{f,c}_{t+1}(\Omega^c) + (1-m^{np})V^{f,s}_t(\omega')\big\} + \\  \nonumber
-& \hspace{10em} p^{\text{meet}}_t p^{\text{preg}}_t \big\{ m^{p} \cdot V^{f,ck}_{t+1}(\Omega^c) + (1-m^{p})V^{f,sk}_{t+1}(\omega')\big\}  \Big]  \bigg\},
-\end{align}\vspace{-3em}
-\begin{align*}
- \text{s.t. \ }  &  a' = R\cdot a  + W^f_t(z) - c  & \text{ (evolution of assets)}\\
- &  z' = z + \varepsilon^{z,f}_t, \ \ \varepsilon^{z,f}_t \sim \mathcal{N}(0;\sigma_{z,f}^2) &  \text{ (evolution of productivity)}\\
-  & \log W^f_t = z_t + \text{Trend}^f_t, \ \ \  \text{Trend}^f_t = a^f_0 + a^f_1\cdot t  +  a^f_2 \cdot t^2 &  \text{ (labor income and trend)}\\
-  & \Omega^c = \mathcal{M}^f(a',z') &  \text{ (marriage prospectives)}
-\end{align*}
-
-% 
-The crucial difference here is the term $V^{f,sk}$: it reflects the fact that upon disagreement woman becomes a single female with a child.
-
-\subsection{Couples Without Kids}
-Couple is characterized by age $t$ (assumed the same for both spouses), marriage terms $\theta$, productivities of female and male $z^f$, $z^m$ and additive marriage surplus $\psi$.
-Couple maximizes weighted expected lifetime utility, and $\theta$ represents share of female in this objective function, therefore $\Omega = \{\theta,\psi,z^f,z^m\}$ represent characteristics of couple.
-
-The key element shaping couple's decisions is participation constraints: each period the expected lifetime utility of staying in couple should be not less that expected lifetime utility of getting a divorce and becoming a single agent.
-Depending on realization of the shocks, if it is possible to satisfy both participation constraints the next period (so no divorce happens, $d = 0$) state is $\Omega'$.
-If participation constraints cannot be satisfied, the couple gets a divorce $d = 1$ and its characteristics are dissolved onto $\omega^{df}$ and $\omega^{dm}$.
-In more details this is described in Section [...].
-
-Fertility dimension is an important element.
-Each period couples get fertility shock $p^{\text{preg}}_t$.
-If it happens, they become couple with a child if they stay together (renegotiation happens after they know realization of the shock), or become single male and single female with a child if they do not.
-If the shock does not happen, they still may make a decision to get a baby \emph{after} they renegotiate the marriage terms if they stay together.
-This timing is subtle, but is required for everything to be internally consistent, see Appendix ...
-for more details on how timing interacts with renegotiation.
-
-After realization of all the shocks and making fertility and divorce decisions in the current period the problem of the couple that stays childless is
-\begin{align}& \hspace{5em}  V^{c}_t(\Omega) = \max\limits_{c^f,c^m}  \bigg\{ \theta\cdot u(c^f) + (1-\theta)\cdot u(c^m) + \psi +  \label{vf-c} \\   \nonumber
- &  \beta \E_t \Big[   (1 - p^{\text{preg}}_t)\cdot \left\{ (1-d^{\text{np}})\cdot \max\left\{ V^{ck}_{t+1}(\Omega'),V^{c}_{t+1}(\Omega')\right\} + d^{\text{np}}\cdot [ \theta V_{t+1}^{f,s}(\omega^{df}) + (1-\theta)V_{t+1}^{m,s}(\omega^{dm})]\right\}  +  \\  \nonumber
-& \hspace{5em} p^{\text{preg}}_t\cdot \left\{ (1-d^{\text{p}})\cdot V^{ck}_{t+1}(\Omega') + d^{\text{p}}\cdot [ \theta V_{t+1}^{f,sk}(\omega^{df}) + (1-\theta)V_{t+1}^{m,s}(\omega^{dm})]\right\} \Big] \bigg\},
-\end{align}\vspace{-2em}
-\begin{align*}
-\text{s.t. \ }& a' + c = R\cdot a  + W^m_t(z^m) + W^f_t(z^f) & \text{(evolution of joint assets)},\\
-				 & c = [(c^f)^{1+\rho_c} + (c^m)^{1+\rho_c}]^{\frac1{1+\rho_c}} & \text{(increasing returns in consumption)},\\
-				 &  z^{f\prime} = z^f + \varepsilon^{z,f}_t, \ \ \varepsilon^{z,f}_t \sim \mathcal{N}(0;\sigma_{z,f}^2) &  \text{ (evolution of female productivity)}\\
-				 &  z^{m\prime} = z^m + \varepsilon^{z,m}_t, \ \ \varepsilon^{z,m}_t \sim \mathcal{N}(0;\sigma_{z,m}^2) &  \text{ (evolution of male productivity)}\\
-                    & \psi' = \psi + \varepsilon^{\psi}_t, \ \ \varepsilon^{\psi}_t \sim \mathcal{N}(0;\sigma_{\psi}^2)  & \text{(evolution of marriage surplus),} \\
-                    & (\Omega',\omega^{df},\omega^{dm}) = \mathcal{R}(\theta,a',z^{m\prime},z^{f\prime},\psi') & \text{(renegotiation correspondence)},
-\end{align*}
-
-\subsubsection{Defining Individual Values}
-The couple's collective value function $V^c$ is the object relevant for making decisions of consumption and fertility.
-Making marriage and divorce decisions involves individual value functions of spouses living in the couple.
-I denote them as $V^{f,c}$ and $V^{m,c}$.
-These value functions are obtained by plugging optimal decisions of couple into intertemporal utilities of each agent and accounting for possible transitions.
-Note that they cannot be derived from a maximization problem, and common properties of value functions such as envelope theorems do not hold for them.
-
-Namely, given couple's decisions $c^f(\Omega)$, female's value of being in couple is defined recursively as
-\begin{align}
-& \hspace{5em}  V^{f,c}_t(\Omega) =    u(c^f) + \psi +  \\   \nonumber
- &  \beta \E_t \Big[   (1 - p^{\text{preg}}_t)\cdot \left\{ (1-d^{\text{np}})\cdot \max{}^C \left\{ V^{f,ck}_{t+1}(\Omega'),V^{f,c}_{t+1}(\Omega')\right\} + d^{\text{np}}\cdot [ V_{t+1}^{f,s}(\omega^{df})]\right\}  +  \\  \nonumber
-& \hspace{5em} p^{\text{preg}}_t\cdot \left\{ (1-d^{\text{p}})\cdot V^{f,ck}_{t+1}(\Omega') + d^{\text{p}}\cdot [ V_{t+1}^{f,sk}(\omega^{df}) ]\right\} \Big] 
-\end{align}
-where operator $\max{}^C\{A,B\} \equiv A\cdot \I[V^{ck}_{t+1}(\Omega')\geq V^{c}_{t+1}(\Omega')] + B\cdot \I[V^{ck}_{t+1}(\Omega')< V^{c}_{t+1}(\Omega')]$.
-It reflects the fact that couple makes joint decision that is not necessary individually optimal.
-
-Another important property of individual values is that
-\begin{equation} V^{c}_t(\Omega) \neq \theta\cdot V^{f,c}_t(\Omega) + (1-\theta)\cdot V^{m,c}_t(\Omega),\label{tht_noneq}\end{equation}
-the main reason for this is that $\theta$ changes in the future as a result of random shocks, and this change is not likely to be symmetric.
-This, in particular, causes couple's expected future value function to be discontinuous in $(a,z)$.
-See Appendix [...] for some additional discussion.
-
-\subsection{Couples With Kids}
-State space for couples is the same as for singles, except for additional variable $\xi$ representing child's age, therefore $\Omega = \{\theta,\psi,z^f,z^m,\xi\}$.
-I consider just two states for $\xi \in \{y,s\}$ representing young and school-age child.
-At birth the child is young, then each period with probability $p^{s}$ the child becomes of school-age.
-The difference between young and school-age children are mother's time requirement, namely young children require more time and elder --- more money.
-
-Couples with kids have additional choice variable $q$, that can be interpreted as flow of child quality or child consumption.
-Child quality is produced using mother's labor $l_f$ and monetary expenditures $x$ according to constant returns to scale production function.
-Labor input to child quality causes female to lose part of her labor income.
-The production function depends on age, namely I parametrize it as $q = [\mu_\xi\cdot l_f^{\theta_q} + (1-\mu_\xi)\cdot x^{\theta_q}]^{\frac1{\theta_q}}$.
-This follows, for instance, [Sommer ...], with the addition of changing labor share $\mu$.
-
-\begin{align}& \hspace{5em}  V^{ck}_t(\Omega) = \max\limits_{c^f,c^m,q,l_f,x}  \bigg\{ \theta\cdot u(c^f,q) + (1-\theta)\cdot u(c^m,q) + \psi + \label{vf_ck} \\  \nonumber
- & \hspace{10em} \beta \E_t \Big[   \left\{ (1-d)\cdot   V^{ck}_{t+1}(\Omega') + d\cdot [ \theta V_{t+1}^{f,sk}(\omega^{df}) + (1-\theta)V_{t+1}^{m,s}(\omega^{dm})]\right\} \Big] \bigg\},
-\end{align}\vspace{-2em}
-\begin{align*}
-\text{s.t. \ } & a' + c + x = R\cdot a  + W^m_t(z^m) + (1-l_f)\cdot W^f_t(z^f) & \text{(evolution of joint assets)},\\
-                    & c = [(c^f)^{1+\rho_c} + (c^m)^{1+\rho_c}]^{\frac1{1+\rho_c}} & \text{(increasing returns in consumption)},\\
-                    & q = [\mu_\xi\cdot l_f^{\theta_q} + (1-\mu_\xi)\cdot x^{\theta_q}]^{\frac1{\theta_q}} & \text{(production function of child quality)},\\
-                    &  z^{f\prime} = z^f + \varepsilon^{z,f}_t, \ \ \varepsilon^{z,f}_t \sim \mathcal{N}(0;\sigma_{z,f}^2) &  \text{ (evolution of female productivity)}\\
-				 &  z^{m\prime} = z^m + \varepsilon^{z,m}_t, \ \ \varepsilon^{z,m}_t \sim \mathcal{N}(0;\sigma_{z,m}^2) &  \text{ (evolution of male productivity)}\\
-                    & \psi' = \psi + \varepsilon^{\psi}_t, \ \ \varepsilon^{\psi}_t \sim \mathcal{N}(0;\sigma_{\psi}^2)  & \text{(evolution of marriage surplus),} \\
-                   &  \P(\xi' = s | \xi = y) = p^s, \ \ \P(\xi' = s | \xi = s) = 1 & \text{(evolution of child's age)},\\
-                    & (\Omega',\omega^{df},\omega^{dm}) = \mathcal{R}(\theta,a',z^{m\prime},z^{f\prime},\psi',\xi) & \text{(renegotiation correspondence)},
-\end{align*}
-
-\subsection{Single Mothers}
-State space for single mothers contains individual productivity and child's age.
-Single mothers cannot marry, but have chance $p^{\text{out}}$ of recovering their marriage prospectives.
-In the model this is captured by transition of them to single females.
-This form is somewhat restrictive, although any other option requires modeling marriage of single mothers separately.
-This is a complex issue, for instance, males would seek females who already have kids if we do not make difference between step and own children.
-Since for the purposes of the model option of being single mother is used mainly as a factor affecting bargaining, I introduce additional (constant) utility factor $S$ to flexibly control utility of exiting the state.
-This can be interpreted as equivalent of utility of kids from past marriages when entering the marriage market again.
-This form seems flexible enough to capture variation in an option of being a single mother without putting extra computational and modeling burden.
-
-The problem of a female who stays single mother is
-\begin{align}V^{f,sk}_t(\omega) = \max\limits_{c} & \bigg\{ u(c,q) + \beta \E_t \Big[ p^{\text{out}} \cdot \left\{ V^{f,s}_{t+1}(\omega') + S\right\} +(1-p^{\text{out}})V^{f,sk}_{t+1}(\omega') \Big]  \bigg\},
-\end{align}\vspace{-1em}
-\begin{align*}
- \text{s.t. \ }  &  a' = R\cdot a  + (1-l_f)W^f_t(z) - c  & \text{ (evolution of assets)}\\
- & q = [\mu_\xi\cdot l_f^{\theta_q} + (1-\mu_\xi)\cdot x^{\theta_q}]^{\frac1{\theta_q}} & \text{(production function of child quality)},\\
- &  z' = z + \varepsilon^{z,f}_t, \ \ \varepsilon^{z,f}_t \sim \mathcal{N}(0;\sigma_{z,f}^2) &  \text{ (evolution of productivity)}\\
- &  \P(\xi' = s | \xi = y) = p^s, \ \ \P(\xi' = s | \xi = s) = 1 & \text{(evolution of child's age)},\\
-\end{align*}
-
-\subsection{Marriage Market}
-Single males and females meet potential partners of identical age and characteristics that depend on their assets $a$ (chosen in the previous period) and productivity $z$ (revealed after realization of shock):
-\begin{equation}\log a^p = \log a + \varepsilon^{a,p}, \ \ \varepsilon^{a,p} \sim \mathcal{N}(0,\sigma_{a,p}^2),\end{equation}
-\begin{equation}	z^p = z + \varepsilon^{z,p}, \ \ \varepsilon^{z,p} \sim \mathcal{N}(0,\sigma_{z,p}^2),\end{equation}
-for example, $\sigma_{a,p} = 0.1$ can be read as standard deviation of partner's assets being 10\% around of what agent has. 
-Additionally, initial marriage surplus (love shock) is drawn from
-\begin{equation} \psi \sim \mathcal{N}(0,\sigma_{\psi,0}^2).\end{equation}
-
-If people agree to be a couple they pool their assets $a^c = a + a^p$. If no unplanned pregnancy happens, people agree to marry if set of mutually satisfactory marriage terms
-\begin{equation}\small \Theta^{np}_t = \left\{ \theta : V_t^{f,c}(\theta,\psi,a^c,z^f,z^m) \geq V_t^{f,s}(a^f,z^f), \ \ V_t^{m,c}(\theta,\psi_0,a^c,z^f,z^m) \geq V_t^{m,s}(a^m,z^m)\right\}\end{equation}
-is non-empty. If potential couple gets the pregnancy shock, the set changes to
-\begin{equation} \small \Theta^{p}_t = \left\{ \theta : V_t^{f,ck}(\theta,\psi,a^c,z^f,z^m,y) \geq V_t^{f,sk}(a^f,z^f,y), \ \ V_t^{m,ck}(\theta,\psi_0,a^c,z^f,z^m,y) \geq V_t^{m,s}(a^m,z^m)\right\} \end{equation}
-(at birth child's age is $\xi = y$).
-
-If the respective set is non-empty, initial bargaining power of couple is determined by symmetric Nash Bargaining, for instance:
-\small
-\begin{align}&\theta^*(a,z,\psi,\epsilon^{a,p},\epsilon^{z,p}) = \label{nbs} \\ \nonumber &= \argmax\limits_{\theta \in \Theta^{np}} \left[V_t^{f,c}(\theta,\psi,a^c,z^f,z^m) - V_t^{f,s}(a^f,z^f)\right] \times \left[V_t^{\vphantom{f}m,c}(\theta,\psi,a^c,z^f,z^m) - V_t^{m,s}(a^m,z^m)\right],\end{align}
-and analogously for the case when pregnancy shock happened.
-
-From the prospective of a single agent, this set is determined by realization of $\psi$ and partner's characteristics $\epsilon^{a,p}$ and $\epsilon^{z,p}$.
-Given them, marriage function $m$ is binary:
-\[m^{np}_t(a,z,\psi,\epsilon^{a,p},\epsilon^{z,p}) = \I(\Theta^{np}_t \neq \varnothing),\]
-where variables affecting set $\Theta$ are unambiguously recovered from arguments of $(a,z,\psi,\epsilon^{a,p},\epsilon^{z,p})$ as described above.
-
-%Because of good properties of Nash Bargaining, expected future value function is continuous with respect to of $a$ and $z$. See Appendix [...] for additional discussion of this.
-
-Let $\omega = (a,z)$ and $\epsilon = (\psi,\epsilon^{a,p},\epsilon^{z,p})$. This allows to define a random function
-\[\Omega^c = \mathcal{M}_t(\omega) = M_t(\omega,\epsilon) = (\theta^*,\psi,a^c,z^m,z^f),\]
-so the future characteristics of couple are function of current characteristics of individual and three-dimensional random shock.
-\subsection{Renegotiation And Divorce}
-
-Upon divorce, $\kappa\cdot a$ of assets disappears and the rest is split evenly.
-This generalizes [Voena ...], where I put additional parameter $\kappa$ captures can rationalize extra costs of divorce for more wealthy couples.
-[Voena ...] argues that this approximates well actual property divisions.
-So, the splitting rule is
-\[a^{df} = 0.5\cdot (1-\kappa)\cdot a,  \ \ a^{dm} = 0.5\cdot (1-\kappa)\cdot a.\]
-This splitting rule defines mapping of couple's state $\Omega$ to individual states $\{\omega^{df},\omega^{dm}\}$ in case of divorce.
-
-The option to leave with a share of couple's assets creates pressure for participation constraints.
-As argues by [...], the household prefers to keep $\theta$ constant.
-However, after shocks to $z$ and $\psi$ happen, one (or both) participation constraints might be violated.
-To address this formally, consider couple that already has kids.
-I define set
-\begin{equation}\Theta_t = \left\{ \tilde\theta : V^{f,ck}_t(a,\tilde\theta,\psi,z^f,z^m,\xi) \geq V_t^{f,sk}(a^{df},z^f,\xi) , \ \ V_t^{m,ck}(a,\tilde\theta,\psi,z^f,z^m,\xi)\geq V_t^{m,s}(a^{dm},z^m) \right\} \label{reneg-with-kids}
-,\end{equation}
-and if couple starts with current value $\theta_{0}$ there are three possible options:
-\begin{enumerate}
-\item \textit{Status quo.} If $\theta_0 \in \Theta_t$, current marriage terms are satisfactory for both spouses so $\theta_t = \theta_{0}$.
-\item \textit{Renegotiation.} If $\theta_0 \not\in \Theta_t$, but $\Theta_t \neq \varnothing$, then spouses can continue to be married if they change their marriage terms.
-In this case they pick the closest the new value from set $\Theta_t$ such that it is the closest to the old marriage terms: $\theta_t = \min\limits_{\tilde\theta\in\Theta_t} |\tilde\theta - \theta_0|$.
-\item \textit{Divorce.}  If $\Theta_t = \varnothing$, so if there do not exist any marriage terms satisfactory for both spouses, then divorce happens according to the rule described above.
-\end{enumerate}
-Picking $\theta$ that is the closest to the old one is not a random choice: it is rationalized by models with limited commitment models, where change in $\theta$ represents the value of Lagrange multiplier on binding participation constraint.
-Intuitively, since couple does not like $\theta$ to be changing and everything is continuous in $\theta$, the couple prefers small changes to larger ones, and this drives the result.
-
-Renegotiation that involves making discrete decisions has few more issues as backward induction has to be used: fertility decisions have to be made after new $\theta$ is set, and these decisions might depend on $\theta$, therefore inequalities defining set $\Theta_t$ need to be modified.
-This is an artifact of assuming that fertility decisions are made collectively and it can be the case that having a baby benefits couple but hurts wife alone (mainly by worsening her outside option), and such scenario of renegotiation insures that participation constraints hold after the fertility decisions are made.
-See Appendix \ref{ren-disc} where I elaborate on this.
-
-\subsection{Additional Details}
-The model I solve an estimate contains few additional things that are omitted for more clear exposition above.
-
-First, females may exit the state of being single mother immediately.
-This transition is assumed to happen after divorce or failure to agree to marry, therefore in the value functions and in (re)negotiation I use $V^{f,s?} = p^{\text{out}} [V^{f,s} + S] + (1-p^{\text{out}})V^{f,sk}$ instead of just $V^{f,sk}$.
-This allows for quicker re-marriage and easier implementation of counterfactual in which females do not have risk of becoming single mother at all.
-I use notation of $V^{f,s?}$ in Appendix \ref{add-trans} where I show exact value functions.
-
-Second, I assume that having a baby is possible only before certain age $\bar{T}$. After that, value functions and negotiation rules are modified in a straightforward manner.
-
-Finally, to simplify computations I also do not allow marriage status and marriage terms to change after certain $\bar{T}_2$.
-Therefore I drop participation constraints in the couples, the value functions look the same but do not allow for $\theta$ to change.
-
-\subsection{Functional Forms}
-I use standard CES utility functions for individuals and couples with no kids: $u(c) = \frac{c^{1-\sigma}}{1-\sigma}$.
-When kids appear, people extract additional utility of child quality, I use the following non-separable utility function:
-\[u(c,q) = \frac{\left(e^{\phi_0}\cdot c^{1-\phi_1} \cdot (\phi_2 + q)^{\phi_1}\right)^{1-\sigma}}{1-\sigma},\]
-parameter $\phi_0$ represents fixed gain in marginal utility because of having a baby, $\phi_1$ drives relative importance of child quality in utility function and $\phi_2$ is a ``luxury good'' parameter that can rationalize large child expenditures of wealthy couples that are discussed in the estimation section.
+\input{model}
 
@@
 \section{Additional Empirical Results}
-\subsection{ACS}
+\subsection{Alternative Income Partitions\label{inc-part}}
+
+Defining income groups in cross-sectional comparisons is no trivial, as income is endogenous to household's composition that is the main object of study here.
+I in this section I build several altrenative income partitions and argue that qualitatively pattern I care about holds for variety of definitions, although exact numbers may differ quite a lot.
+
+\subsubsection{Equivalence Scale}
+
+Following development and household literature, when dividing people by income group I use total household income adjusted by OECD equivalence scale.
+The scale assigns weight 1 to the household head, 0.5 to spouse and 0.3 to each child.
+
+As I use the household income as the main variable of interest, I have to restrict attention to couples that live by themselves and single females with children.
+I do this both for ACS and SIPP, and as definition of households wary a little, this may drop out slightly different groups of people.
+
+Formally,for income considerations I take a subsample of females who is marked as household's head or head's spouse, and who do not have any people living in household other than their spouse/unmarried partner and their children.
+This, for instance, excludes multigenerational household and people living with roommates, as total household's income may not be a relevant thing for some types of household composition.
+I cannot define income equivalence scale for more complicated living arrangements, although there are different scales that allow to do this.
+
+To mitigate selectivity, I compute income percentiles in all sample of females with household composition mentioned above, regardless of their marriage and fertility status.
+I use total family income as the key variable.
+To mitigate misreporting I drop observations with total yearly family income below $\$4000$ for ACS (that is about a third of federal the poverty level) or monthly income below $\$300$ for SIPP.
 
@@
 
-\section{Additional Notes on Model}
-\subsection{Continuity of Value Functions}
-This part discusses violations of continuity caused by the dynamic bargaining structure.
-The general conclusion is that collective value function does is not necessary continuous in couple's characteristics, and this discontinuity propagates onto all value functions in the model.
-
-To simplify the example, let's consider value function of couple with kids. Similarly to equation \ref{tht_noneq}, we can write
-\[ V^{ck}_t(\Omega) = \theta\cdot V^{f,ck}_t(\Omega) + (1-\theta)\cdot V^{m,ck}_t(\Omega) + \Delta(\Omega),\]
-and use it to substitute the term under the expectation in equation \ref{vf_ck} 
-\begin{align*} & (1-d)\cdot   V^{ck}_{t+1}(\Omega') + d\cdot [ \theta V_{t+1}^{f,sk}(\omega^{df}) + (1-\theta)V_{t+1}^{m,s}(\omega^{dm})] =    V^{ck}_{t+1}(\Omega') + \\
- &  \ \ \ \theta \cdot d(\Omega') \cdot [V_{t+1}^{f,sk}(\omega^{df}) - V_{t+1}^{f,ck}(\Omega')] + (1-\theta) \cdot d(\Omega') \cdot [V_{t+1}^{m,sk}(\omega^{dm}) - V_{t+1}^{m,ck}(\Omega')] - d(\Omega')\cdot \Delta(\Omega'),
-\end{align*}
-if we consider couple that is close to divorce, so there exist two $\Omega'_1$ and $\Omega'_2$  that are close to each other according to some metric, but $d(\Omega'_2) = 1$ and $d(\Omega'_1) = 0$.
-Note that around the divorce threshold by [...] \textbf{both} spouses should be indifferent between staying on leaving, therefore
-\[ V_{t+1}^{f,sk}(\omega^{df}) - V_{t+1}^{f,ck}(\Omega')\approx 0, \ \ V_{t+1}^{m,sk}(\omega^{dm}) - V_{t+1}^{m,ck}(\Omega') \approx 0,\]
-therefore if $\Delta(\Omega')$ is non-zero in these points, this whole expression jumps where value of $d$ jumps.
-[It is not clear whether it is actually zero or not, but in simulations it does look like it is positive.]
-
-\subsection{Renegotiation With Collective Discrete Decisions\label{ren-disc}}
-This section explains how discrete decisions interact with couple's renegotiation.
-The theory requires participation constraints to be satisfied \textit{after} any discrete choice.
-In the context of fertility decisions, that means that both spouses can predict future decisions and re-negotiate $\theta$ accounting for this prediction (rather than assuming that fertility status does not change).
-
-With slight abuse of notation, let's suppose that couple gets collective value $V^{ck}(\theta,\Omega)$ if it makes a baby and $V^{c}(\theta,\Omega)$ is it does not.
-Given $\theta$ total value before the decision is made is \[V^{c?}(\theta,\Omega) = \max\{ V^{ck}(\theta,\Omega), V^{c}(\theta,\Omega)\}\]
-after that I define
-\[V^{f,c?}(\theta,\Omega) = V^{f,ck}(\theta,\Omega)\cdot \I[V^{ck}(\theta,\Omega) \geq V^{c}(\theta,\Omega)] + V^{f,c}(\theta,\Omega) \cdot \I[V^{ck}(\theta,\Omega) < V^{c}(\theta,\Omega)],\]
-note that it is not necessary equal to $\max\{ V^{f,ck},V^{f,c} \}$. Similarly we can define $V^{m,c?}(\theta,\Omega)$.
-
-The main thing to notice is that although $V^{c?}$ is continuous in $\theta$ individual values $V^{f,c?}$ and $V^{m,c?}$ are not, unless both spouses always have agreement on whether to make a baby: $V^{ck} \geq V^{c} \Leftrightarrow V^{f,ck} \geq V^{f,c} \Leftrightarrow V^{m,ck} \geq V^{m,c}$.
-Therefore set
-\[\Omega = \left\{\tilde{\theta} : V^{f,c?}(\tilde\theta,\Omega) \geq V^{f,s}(\omega^{df}), \ \  V^{m,c?}(\tilde\theta,\Omega) \geq V^{m,s}(\omega^{dm}) \right\}\]
-may be disjoint, and careful approximation of $V$ with respect to $\Omega$ is required.
-
-The possible solution to it would be to limit fertility decisions to the cases where both spouses agree on having kids, this, however, goes against the logic where couple maximizes collective utility subject to participation constraints.
-
-\subsection{Accounting for Additional Transitions\label{add-trans}}
-In value functions I stated above I assumed that females stays single mother for at least one period if it exits a couple with a kid or if she has a baby at the moment of marriage and disagrees to marry.
-In the actual model I relax this assumption so the female can stop being single mother immediately.
-This is needed, for instance, for evaluating counterfactuals in which female can exit couple without pressure of being a single mother.
-
-I assume that at the moment of (re)negotiation female is unsure about whether she is going to be a single mother. Therefore, I define expected value of leaving the couple to be
-\[V_t^{f,c?}(\omega) = p^{\text{out}}[V_t^{f,s}(\omega) + S] + (1-p^{\text{out}})V_t^{f,sk}(\omega),\]
-therefore equaiton \ref{single-fem} becomes
-\begin{align}V^{f,s}_t(\omega) = \max\limits_{c} & \bigg\{ u(c) + \beta \E_t \Big[ (1 - p^{\text{meet}}_t)\cdot V^{f,s}_{t+1}(\omega') + \label{single-fem-true} \\  \nonumber
-& \hspace{7em} p^{\text{meet}}_t (1-p^{\text{preg}}_t) \big\{ m^{np} \cdot V^{f,c}_{t+1}(\Omega^c) + (1-m^{np})V^{f,s}_t(\omega')\big\} + \\  \nonumber
-& \hspace{10em} p^{\text{meet}}_t p^{\text{preg}}_t \big\{ m^{p} \cdot V^{f,ck}_{t+1}(\Omega^c) + (1-m^{p})V^{f,s?}_{t+1}(\omega')\big\}  \Big]  \bigg\},
-\end{align}
-the set is \ref{reneg-with-kids} becomes
-\begin{equation}\Theta_t = \left\{ \tilde\theta : V^{f,ck}_t(a,\tilde\theta,\psi,z^f,z^m,\xi) \geq V_t^{f,s?}(a^{df},z^f,\xi) , \ \ V_t^{m,ck}(a,\tilde\theta,\psi,z^f,z^m,\xi)\geq V_t^{m,s}(a^{dm},z^m) \right\} \label{reneg-with-kids}
-,\end{equation}
-and the analogous term is introduced for initial negotiation of couples in case of pregnancy shock.
-
-\section{Computational Notes}
-This section describes the procedures I use to approximate the solution of the model above.
-In general, since the model is finite-horizon, value function iteration starting from the last period is used.
-The value functions and corresponding policy functions are approximated on a grid.
-Since state space is quite large (it has at least 4 continuous dimensions for couples), I use approximation techniques of [Judd--Maliar--Maliar] to deal with dimensionality.
-Namely, in the dimension of productivity, marriage terms and marriage surplus the functions are approximated by Smolyak polynomials (that are a variation of multi-dimensional Chebyshev polynomials) on sparse Smolyak grid.
-However, since the assets dimension is important and non-convex value functions can cause troubles, so I use full grid in the dimension of assets and linearly interpolate between assets points.
-
-An additional issue is integration, required to compute expected future values of the value function accounting for possible (re)negotiation and divorce.
-This is noted to be the most costly part of value function iteration.
-I utilize monomial integration rules describe by [JMM] to do it efficiently.
-The second part of this section describe this.
-
-\subsection{Value Function Approximation}
-I describe techniques I used for value functions, the same principle applies for policy functions: they are just byproduct of value function iteration that I use for simulations.
-
-State space of continuous variables for couples is $\Omega = (a,\theta,\psi,z^f,z^m)$.
-I partition it as $\Omega = (a,O)$.
-Smolyak method generates $S$ gridpoints $G = \{O_1,...,O_S\}$, where each $O$ represents some combination of $(\theta,\psi,z^f,z^m)$; and polynomials $P = \{p_1(o),...,p_S(o)\}$.
-Therefore any function $f(o)$ is approximated by
-\[f(o) \approx \hat{f}(o) = \sum_{i=1}^S c_{i}\cdot p_i(o),\]
-and coefficients $c$ are found from collocation relation
-\[\hat{f}(O_i) = f(O_i) \ \forall i = 1,...,S.\]
-
-Since we have to approximate functions for different levels of $a$, I introduce additional grid $A = \{a_1,...,a_J\}$.
-For each $a_j$ we can find corresponding coefficients $c$, I denote them $c_i(a_j)$.
-At points $a_j$ the function can be written as
-\[\hat{f}(a_j,o) = \sum\limits_{i=1}^S c_i(a_j) p_i(o),\]
-for points between grid $a \not A$ I linearly interpolate $f(a_j,o)$.
-Generically, to interpolate values for arbitrary $a$ and $o$ the  reasonable approach is to first compute values of $f$ on each point of $A$ grid for current $o$, to get $F = \{f(a_1,o),...,f(a_J,o)\}$ and then interpolate $F$ with respect to $a$.
-However, I use linear interpolation and since the function is bilinear in $c$ and $p$, this is fully equivalent to linear interpolation of coefficients $c_i(a_j)$.
-Therefore I define $\hat{c}_i(a)$ to be linearly interpolated value of coefficients, and therefore the function at arbitrary point is approximated by
-\[\hat{f}(a,o) = \sum\limits_{i=1}^S \bar{c}_i(a) p_i(o).\]
-If we use interpolation other than linear with respect to $a$ this equivalence would be violated so the procedure becomes more complicated as it requires evaluating many values of $f$.
-
-\subsection{Monomial Quadratures}
-See [JMM] for a detailed discussion.
-Let $\epsilon$ be $d$-dimensional normal distribution and we are interested in evaluating $\E(f(\epsilon))$.
-Standard techinque would involve using Gauss--Hermite quadrature with respect to each dimension of $\epsilon$, this, however, quickly becomes infeasible: if we want to have 5 points with respect to each dimension of 4-dimensional vector we would have $5^d = 625$ points, out of which many will have very low weights.
-
-Monomial quadratures generate number of integration node that grows much lower than exponent of $d$.
-I use the one that is proportional to $d^2$.
-So, the rule generates $Q$ vectors $\{\epsilon_1,...,\epsilon_q\}$ with weights $w_q$, and the integral can just be written as $\E(f(x))\approx \sum_{q} f(\epsilon_q)$.
-
-As it is noted by [...] that there is a separate issue of using quadratures to approximate integrals of discontinuous functions.
-However, they also show that the model with deterministic transition decisions can be interpreted as a limiting case of model with stochastic (logit) transitions, for which integration by quadratures is valid.
-I also believe that this can be addressed by re-interpreting shocks to have a discrete distribution with probability mass $w_q$ rather than true normal distribution (so we do not pretend to recover the integrals under true normal distribution).
-In general, the math background for this does not seem to be well-developed and I use conventional tools at this stage of the project without asserting its mathematical rigor.
-In the future, I may try to test the approximations I use against just discretizing all random walks with number of points.
-
-\section{Auxiliary Estimates}
-\subsection{Partner's Earnings\label{partearn}}
-The model assumes that
-\[ z^p = z^{\text{own}} + \varepsilon^{z,p}, \ \ \ \varepsilon^{z,p}\sim\mathcal{N}(0,(\sigma^{z,p})^2),\]
-so detrended log-wage of potential partner is normally distributed with the mean of own detrended log-wage and standard deviation of $\sigma^{z,p}$.
-
-To estimate $\sigma^{z,p}$ I regress log-difference in spouses earning for couples in ACS on several predictors, that include ages of spouses, ACS year and state.
-In this section I present several alternative specifications, including an attempt to correct for selection into marriage, and show what variance they imply.
-
-The main specification is
-\[\log \frac{\text{Earnings of wife}}{\text{Earnings of husband}} = \alpha + X'\gamma + \epsilon,\]
-where I run regression on subsample of couples in which both spouses work full time and report labor income above 5000 per year.
-Controls $X$ include 4th degree polynomials and wife's and husband's age, as well as absolute value of age difference, dummies for states and ACS years.
-I regress difference in per-hour earnings to the set of controls for couples who married in the year previous to the survey year and who have no children in the household.
-Relaxing this definition, using total earnings instead of per-hour and even controlling for education lead to slightly different estimates, but all of them are within $[0.66,0.8]$ range.
-
-I pick the upper bound of this interval and set $\sigma^{z,p} = 0.8$: as in the model people are more likely to agree to marry similar partners, I expect similar statistic in the simulations to be slightly lower.
-[Similar statistic computed on simulated data returns ...]
-
-\newpage
-\section*{Very Technical Appendix, Not For Final Version}
-\section{Expressions for Value Functions}
-\subsection{Notation}
-This technical note shows exact expressions for approximating value functions.
-For the couple's state I use partition $\Omega = (a,O)$, where $O = (\theta,\psi,z^f,z^m)$.
-For the age variable I put $\xi\in\{y,s\}$ counts as a separate letter, so states for couples with kids are noted like $cky$ and $cks$.
-Let $p^s$ represents probability that the child ``grows up'' so couple switches from $cky$ to $cks$.
-For individual's state there are only two variables $\omega = (a,z)$, and child's age for single mother is reported as state $sky$ if $\xi = y$ and $sks$ if $\xi = y$.
-
-Grid consists of all possible combinations of $a$ and $O$ ($a$ and $z$ in single's case), so \[G = \{ \{a_1,...,a_I\} \times \{o_1,...,o_J\}\}\]
-
-Let $i$ refers to assets position and $j$ refers to values of everything else. Finally, indices $q$ refer to quadrature nodes, and $w_q$ to quadrature weights that add up to 1.
-
-\subsection{Single Agents}
-Consider females, males are handled analogously. Value function \ref{single-fem} can be written as
-\[V^{f,s}_t(a,z) = \max\limits_{c} \left\{ u(c) + \beta \mathcal{V}^{f,s}_{t+1}(s,z)\right\}, \ \ c + s = M(a,z),\]
-Here  $\mathcal{V}^{f,s}_{t+1}$ represents expectation of tomorrow's individual value function, accounting for all possible transitions.
-Argument $s$ refers to savings made in the current period (it is not necessary equal to future assets as future assets change if the partner is met), and $z$ refers to current value of productivity.
-Approximating $\mathcal{V}_{t+1}(s,z)$ (i.e.
-integration) is the key challenge.
-Note that it does not depend on current assets, only on savings $s$ that define future assets position.
-$M(a,z) = R\cdot a + \exp(z + \text{Trend}_t)$ is the amount of money in the end of the period.
-
-I treat the function $\mathcal{V}^{f,s}_{t+1}(s,z)$ as a collection of functions $\mathcal{V}^{f,s}_{mj,t+1}$ for each value of current productivity $z_j$ and savings $s_m$.
-I generate quadrature nodes that consist of shock to individual productivity $\epsilon^{z,f}$, shocks to partner's assets and productivity positions
-$(\epsilon^{a,p},\epsilon^{z,p})$ and initial couple's surplus $\psi$.
-I denote these nodes as $x_q = (\epsilon^{z,f}_q,\epsilon^{a,p},\epsilon^{z,p},\psi_q)$, in total monomial rule results in $Q = [...]$ nodes.
-Weight $w_q$ corresponds to each node.
-Given savings $s$ and grid point in $z_j$ I generate partner's and couple's characteristics.
-\[a^p_{mq} = s_m\cdot \exp(\epsilon^{a,p}_q), \ \ a^c_{mq} = s_m\cdot [1 + \exp(\epsilon^{a,p}_q)], \ \ z^m_{jq} = z^f_j \exp(\epsilon^{z,p}_q),\]
-(in some sense in this setup by saving more individuals have chances of getting more wealthy partners).
-Also, individual next period's productivity is generated according to $z^f_{jq} = z^f\cdot \exp(\epsilon^{z,f}_q)$.
-Note that $z^f_j$ refers to the current value and $z^f_{jq}$ refers to the next period's values;
-
-This allows to generate potential couple's characteristics $(a^c_{mq},\psi_q,z^{f}_{jq},z^{m}_{jq})$, and combining it with value functions for the next period we can define:
-\[\Theta^{np}_{mjq,t+1} = \left\{ \tilde\theta : V^{f,c}_{t+1}(a^c_{mjq}, \tilde\theta, \psi_q,z^{f}_{jq},z^{m}_{jq}) \geq V^{f,s}(s_m,z^f_{jq}), \ V^{m,c}_{t+1}(a^c_{mjq}, \tilde\theta, \psi_q,z^{f}_{jq},z^{m}_{jq}) \geq V^{m,s}(a^p_{mq},z^m_{jq}) \right\},\]
-if $\Theta^{np}_{mjq,t+1}$ is empty\footnote{I figure this out by defining a uniformly spaced grid for $\theta \in [0.05,0.95]$.
-Because of Smolyak grid, this can be done very efficiently as $V^{f,c}_{t+1}(a^c_{mjq}, \tilde\theta, \psi_q,z^{f}_{jq},z^{m}_{jq})$ are just a polynomial in $\tilde{\theta}$.}
-, $m^{np}_{mjq,t+1} = 0$ (so no marriage happens at node $jq$ if agents made savings of $s_m$).
-If the set is not empty, then $\theta_{mjq}$ is determined by Nash Bargaining \ref{nbs} and $m^{np}_{mjq,t+1} = 1$.
-Therefore at node $(s_m,z_j)$ future value function in case the partner is met and pregnancy did not happen can be approximated as
-\[\E_t  \big\{ m^{np} \cdot V^{f,c}_{t+1}(\Omega^c) + (1-m^{np})V^{f,s}_t(\omega')\big\} \approx \mathcal{E}^{np}(s_m,z_j),\]
-where
-\[\mathcal{E}^{np}(s_m,z_j) = \sum\limits_{q=1}^{Q} w_q \cdot \left\{ m^{np}_{mjq,t+1}\cdot V^{f,c}_{t+1}(a^c_{mq},\theta_{mjq},\psi_q,z^{f\prime}_{jq},z^m_{jq}) + (1-m^{np}_{mjq,t+1})\cdot V^{f,s}_{t+1}(s_m,z_j)\right\}.\]
-In exactly the same manner we define the set in the case the pregnancy shock did happen:
-\[\Theta^{p}_{mjq,t+1} = \left\{ \tilde\theta : V^{f,ck}_{t+1}(a^c_{mjq}, \tilde\theta, \psi_q,z^{f}_j,z^{m}_{jq}) \geq V^{f,s?}(s_m,z^f_{j}), \ V^{m,ck}_{t+1}(a^c_{mjq}, \tilde\theta, \psi_q,z^{f}_{jq},z^{m}_{jq}) \geq V^{m,s}(a^p_{mq},z^m_{jq}) \right\},\]
-where function $V^{f,s?} = p^{\text{out}}[V^{f,s} + S] + (1-p^{\text{out}})V^{f,sky}_t$ captures possible immediate transition out of being single mother. In the same way 
-\begin{align*}\mathcal{E}^{p}(s_m,z_j) = \sum\limits_{q=1}^{Q} w_q \cdot  \Big\{ m^{p}_{mjq,t+1}\cdot V^{f,cky}_{t+1}(a^c_{mq},\theta_{mjq},\psi_q,z^f_j,z^m_{jq}) + (1-m^{p}_{mjq,t+1})\cdot V^{f,s?}(s_m,z_j) \Big\},
-\end{align*}
-lastly, the value of not meeting any partner is simply
-\begin{align*}\mathcal{E}^{0}(s_m,z_j) = \sum\limits_{q=1}^{Q} w_q \cdot  V^{f,s}_{t+1}(s_m,z^f_{jq}),
-\end{align*}
-finally, the whole function $\mathcal{V}^{f,s}_{t+1}(s_m,z_j)$ can be approximated by weighted average
-\[\mathcal{V}^{f,s}_{t+1}(s_m,z_j) \approx (1-p^{\text{meet}}_t) \cdot \mathcal{E}^{0}(s_m,z_j) +  p^{\text{meet}}_t(1-p^{\text{preg}}_t)  \cdot \mathcal{E}^{np}(s_m,z_j) 
-+  p^{\text{meet}}_t p^{\text{preg}}_t  \cdot \mathcal{E}^{p}(s_m,z_j).\]
-
-After that I use linear interpolation in $s$ dimension to approximate $\mathcal{V}^{f,s}_{t+1}(s_m,z_j)$ for arbitrary $s$.
-Therefore I can get approximation of $\mathcal{V}^{f,s}_{t+1}(s,z_j)$.
-As a result, value function in each asset point $a_i$, $z_j$ is obtained by numerically solving
-\[V^{f,s}_t(a_i,z_j) = \max\limits_{s \in [\underline{A},\overline{A}]} \left\{ u(M(a_i,z_j) - s) + \beta \mathcal{V}^{f,s}_{t+1}(s,z_j)\right\}.\]
-I take $\underline{A}$ and $\overline{A}$ the same as upper and lover bound for individual's assets grid (this is important as approximation may be crude).
-I use Matlab's \textbf{fminbnd} to solve for the maximum point, although I also test that results are very similar when I just use discrete grid $s \in \mathcal{S} = M(a_i,z_j) \cdot\{0,...,0.99\}$ with some spacing, though this variation is much slower.
-
-Choice of solver can generically be an issue as because of concerns described above function $\mathcal{V}^{f,s}_{t+1}(s_m,z_j)$ may be non-smooth or even discontinuous.
-
-\subsection{Couples}
-I consider couples that do not have kids yet as they have richer set of possible transitions.
-Following [...] I split household's decision problem on intrahousehold, intratemporal and intertemporal levels, where first two can be solved analytically.
-For intrahousehold, I define
-\[U(c,\theta) = \max\limits_{c^f,c^m} \left\{\theta u(c^f)  + (1-\theta)u(c^m) \right\}, \ \ \text{s.t.} \ \ \ c = \left[ (c^f)^{1+\rho_c} + (c^m)^{1+\rho_c}\right]^{\frac1{1+\rho_c}},\]
-and given CES assumption this function has simple analytical form.
-
-Intratemporal dimension is trivial, as total consumption is the only intratemporal choice.
-
-For intertemporal I can write the problem \ref{vf-c} as
-\[V^{c}_{t+1}(a,\theta,\psi,z^f,z^m) = \max\limits_{c} \left\{ U(c,\theta) + \beta \mathcal{V}_{t+1}(s,\theta,\psi,z^f,z^m) \right\},\]
-where $\mathcal{V}_{t+1}(s,\theta,\psi,z^f,z^m)$ consists of two terms: what happens if pregnancy shock arrives and what if it does not.
-
-If the shock arrives, in the next period couple is in state $cky$, and if it decides to divorce female becomes a single mother (and may quit this state immediately), and male becomes just a single male.
-I approximate $\mathcal{V}_{t+1}(s,\theta,\psi,z^f,z^m)$ here.
-
-I write current state as combination of $(a_i,O_j)$.
-Function $\mathcal{V}_{t+1}(s,\theta,\psi,z^f,z^m)$ depends on current state $O= (\theta,\psi,z^f,z^m)$ and future savings $s$.
-Integration nodes contains $x_q = (\epsilon^{z,f}_q,\epsilon^{z,m}_q,\epsilon^{\psi}_q)$.
-I take grid with respect to current states, so $O_j= (\theta_j,\psi_j,z^f_j,z^m_j) \in O$ and future savings $s \in A$, so I again approximate expected future value function at each combination of $s_m$ and $O_j$, so I compute values $\mathcal{V}_{mj,t+1} = \mathcal{V}_{t+1}(s_m,\theta_j,\psi_j,z^f_j,z^m_j)$.
-Given each $j$, I can compute next period values of $(\psi_{jq},z^f_{jq},z^m_{jq})$ by adding shocks given by quadrature notes to the current point.
-I also can compute divorce options: $\omega^{df}_{mjq} = ( (1-\kappa)\cdot 0.5 \cdot s_m, \ z^f_{mjq})$ and $\omega^{dm}_{mjq} = ( (1-\kappa)\cdot 0.5 \cdot s_m, \ z^m_{mjq})$.
-This allows to define for each quadrature node a renegotiation set:
-\[\Omega^{p}_{mjq,t+1} = \left\{ \tilde\theta \ : \ V^{f,cky}_{t+1}(s_m,\tilde\theta,\psi_{jq},z^f_{jq},z^m_{jq}) \geq V^{f,s?}_{t+1}(\omega^{df}_{mjq}), \ \ V^{m,cky}_{t+1}(s_m,\tilde\theta,\psi_{jq},z^f_{jq},z^m_{jq}) \geq V^{m,s}_{t+1}(\omega^{dm}_{mjq}) \right\},\]
-and then if $\Omega^{p}_{mjq,t+1}$ is not empty, I pick from it next period's $\theta_{mjq}$ closest to initial $\theta_j$ and set $d^{p}_{mjq} = 0$, otherwise divorce happens and $d^{p}_{mjq} = 0$.
-
-Therefore expected future value if pregnancy shock arrives is 
-{\small
-\[\mathcal{E}^{p}(s_m,O_j) = \sum\limits_{q=1}^{Q} w_q \cdot\left\{ d^p_{mjq}\cdot V_{t+1}^{cky}(s_m,\theta_{mjq},\psi_{jq},z^f_{jq},z^m_{jq}) + (1-d^p_{mjq}) [\theta_{j}\cdot V^{f,sky}(\omega^{df}_{mjq}) + (1-\theta_j)V^{m,s}(\omega^{dm}_{mjq})] \right\} \]
-}
-If pregnancy shock does not arrive, then renegotiation should account for discrete decision of whether to have a kids that happens afterwards, and this decision may depend on $\theta$.
-Namely, I define
-\[k_{mjq}(\tilde\theta) = \I\left[ V^{cky}_{t+1}(s_m,\tilde\theta,\psi_{jq},z^f_{jq},z^m_{jq}) \geq V^{c}_{t+1}(s_m,\tilde\theta,\psi_{jq},z^f_{jq},z^m_{jq})\right],\]
-then I define individual value accounting for this choice:
-\[V^{f,c?}_{t+1}(s_m,\tilde\theta,\psi_{jq},z^f_{jq},z^m_{jq}) = k_{mjq}(\tilde\theta)\cdot V^{f,cky}_{t+1}(s_m,\tilde\theta,\psi_{jq},z^f_{jq},z^m_{jq}) + (1-k_{mjq}(\tilde\theta))\cdot V^{f,c}_{t+1}(s_m,\tilde\theta,\psi_{jq},z^f_{jq},z^m_{jq}),\]
-and the renegotiation set is defined as
-\[\Omega^{np}_{mjq,t+1} = \left\{ \tilde\theta \ : \ V^{f,c?}_{t+1}(s_m,\tilde\theta,\psi_{jq},z^f_{jq},z^m_{jq}) \geq V^{f,s}_{t+1}(\omega^{df}_{mjq}), \ \ V^{m,c?}_{t+1}(s_m,\tilde\theta,\psi_{jq},z^f_{jq},z^m_{jq}) \geq V^{m,s}_{t+1}(\omega^{dm}_{mjq}) \right\},\]
-and we define $\theta_{mjq}$ and $d^{np}_{mjq}$ analogously.
-
-Therefore we define expected future value if pregnancy shock does not arrive to be
-{\small
-\[\mathcal{E}^{np}(s_m,O_j) = \sum\limits_{q=1}^{Q} w_q \cdot\left\{ d^{np}_{mjq}\cdot V_{t+1}^{c?}(s_m,\theta_{mjq},\psi_{jq},z^f_{jq},z^m_{jq}) + (1-d^{np}_{mjq}) [\theta_{j}\cdot V^{f,s}(\omega^{df}_{mjq}) + (1-\theta_j)V^{m,s}(\omega^{dm}_{mjq})] \right\} \]
-}
-so we can finally define value of $\mathcal{V}$ at points $s_m$:
-\[\mathcal{V}^{c}_{t+1}(s_m,O_j) = p_t^{\text{preg}}\cdot \mathcal{E}^{p}(s_m,O_j) + (1-p_t^{\text{preg}})\cdot \mathcal{E}^{np}(s_m,O_j),\]
-and then values at arbitrary points $s$ are obtained by linear interpolation. Therefore the ultimate problem is
-\[V^{c}_t(a_i,O_j) = \max\limits_{s \in [\underline{A},\overline{A}]} \left\{ U(M(a_i,O_j) - s,\theta_j) + \beta \mathcal{V}^{c}_{t+1}(s,O_j)\right\}.\]
-
-\subsection{Couples With Kids} 
-Namely, for intrahousehold allocation of consumption I define
-\[U(c,q,\theta) = \max\limits_{c^f,c^m} \left\{\theta u(c^f,q)  + (1-\theta)u(c^m,q) \right\}, \ \ \text{s.t.} \ \ \ c = \left[ (c^f)^{1+\rho_c} + (c^m)^{1+\rho_c}\right]^{\frac1{1+\rho_c}},\]
-and it still has simple tractable form.
-
-For intratemporal level I have to solve
-\[U(m,W^f,\xi,\theta) = \max\limits_{c,q,l_f,x} U(c,q,\theta), \ \ \ \text{s.t.} \ \ \ c + x + l_f\cdot W^f = m, \ \ q =[\mu_\xi\cdot l_f^{\theta_q} + (1-\mu_\xi)\cdot x^{\theta_q}]^{\frac1{\theta_q}}.
-\]
-this function can also be derived analytically given my assumption on functional forms.
-
-After aggregation, the intertemporal problem looks similar to the one for singles:
-\[V^{cky}(a,\theta,\psi,z^f,z^m) = \max\limits_{m} \left\{ U(m,W^f_t(z^f),\xi,\theta) + \beta \mathcal{V}^{ck}_{t+1}(s,\theta,\psi,z^f,z^m,\xi)\right\}, \ \ \text{s.t.} \ \ m + s = M(a,z^f,z^m),\]
-where definition of $\mathcal{V}$ is analogous.
+\input{theory-appendix} % notes on model
+\input{comp-appendix} % notes on computations
+\input{tech-appendix} % notes on solution process
 

### 2019-05-03 main/model.tex
Re-organized into few files

% This is part of draft.tex containig the model
% \section{Model}
The model is a mixture between conventional heterogeneous agents savings model (B--A--H framework) and household formation and dynamic bargaining, that is extensively developed by [...].

Agents live for $T$ periods. The agents are males, females and married couples. Couples can have a child together, and if a couple with a child divorces up, the child stays with mother, and this is how singe females can have kids. Single females with kids randomly re-enter marriage market by becoming single females without kids.\footnote{The reason for this assumption is to avoid modeling mixture between step and own children, and the fact that single mothers are not a main focus of the paper.}

Single people without kids randomly meet partners with characteristics similar to them and decide whether to create a couple. Couples agree on some marriage terms, represented by each spouse's weight in couple's objective function. The rationale for marriage includes risk-sharing, returns to scale in consumption, possibility to have a child in the future and additional utility surplus of being married (love shock). Each period after random shocks are realized, couples may decide to renegotiate their marriage terms if one of the spouses is better off being single. If they cannot find marriage terms that are satisfactory for both, divorce happens.

Couples can choose to have a child and can also experience unplanned pregnancies. Unplanned pregnancies cannot be aborted. In addition to couples, single people who met each other may experience an unplanned pregnancy just before they decide whether to create a couple. In this case, upon disagreement female becomes a single mother. This mechanism is a stylized representation of kids appearing in unmarried cohabitating couples: since I do not model cohabitation, those kids appear just before people actually decide to marry each other.\textit{ I discuss this issue additionally in Section ....}

\subsection{Remarks on Notation}
I write $V^{i,j}$ to represent individual value functions in the model. Here $i$ represents gender: $f$ for female or $m$ for male. $j$ represents demographic status: $s$ for singles without kids, $sk$ for singles with kids (females only), $c$ for couples without kids and $ck$ for couples with kids. Couple's joint value functions have only one superscript $j \in \{c,ck\}$. Relation between joint value function $V^{c}$ and each spouse's value functions $V^{m,c}$ and $V^{f,c}$ is described in details in couple's section.

\subsection{Singles Without Kids}
Singles without kids are characterized by age $t$, labor productivity $z$ and assets (savings) $a$. I use $\omega = \{z,a\}$ to individual characteristics other than age.

Each period single individuals meet a partner of the same age and characteristics $\omega^p$ with probability $p^{\text{meet}}_t$. If meeting happens, with probability $p^{\text{preg}}_t$ unplanned pregnancy happens. After realizing new period's shocks and pregnancy status couple tries to negotiate marriage terms. If they found marriage terms that are satisfactory for both ($m^{np} = 1$ in case of no pregnancy or $m^p = 1$ in case of pregnancy), they form couple with characteristics $\Omega^c$ . If satisfactory marriage terms could not be found, the individual stays single, or becomes single mother if she is female and unplanned pregnancy happened.

The value function of male agent therefore is
\begin{align}V^{m,s}_t(\omega) = \max\limits_{c} & \bigg\{ u(c) + \beta \E_t \Big[ (1 - p^{\text{meet}}_t)\cdot V^{m,s}_{t+1}(\omega') + \\ \nonumber
& \hspace{7em} p^{\text{meet}}_t (1-p^{\text{preg}}_t) \big\{ m^{np} \cdot V^{m,c}_{t+1}(\Omega^c) + (1-m^{np})V^{m,s}_t(\omega')\big\} + \\  \nonumber
& \hspace{10em} p^{\text{meet}}_t p^{\text{preg}}_t \big\{ m^{p} \cdot V^{m,ck}_{t+1}(\Omega^c) + (1-m^{p})V^{m,s}_{t+1}(\omega')\big\}  \Big]  \bigg\},\\  \nonumber
 \end{align}\vspace{-3em}
 \begin{align*}
 \text{s.t. \ }  &  a' = R\cdot a  + W^m_t(z) - c  & \text{ (evolution of assets)}\\
 &  z' = z + \varepsilon^{z,m}_t, \ \ \varepsilon^{z,m}_t \sim \mathcal{N}(0;\sigma_{z,m}^2) &  \text{ (evolution of productivity)}\\
  & \log W^m_t = z_t + \text{Trend}^m_t, \ \ \  \text{Trend}^m_t = a^m_0 + a^m_1\cdot t  +  a^m_2 \cdot t^2 &  \text{ (labor income and trend)}\\
  & \Omega^c = \mathcal{M}^m(a',z') &  \text{ (marriage prospectives)}
\end{align*}
where random function $\mathcal{M}^m$, representing characteristics of potential couple is defined in Section [...].

The value function of female agent is
\begin{align}V^{f,s}_t(\omega) = \max\limits_{c} & \bigg\{ u(c) + \beta \E_t \Big[ (1 - p^{\text{meet}}_t)\cdot V^{f,s}_{t+1}(\omega') + \label{single-fem} \\  \nonumber
& \hspace{7em} p^{\text{meet}}_t (1-p^{\text{preg}}_t) \big\{ m^{np} \cdot V^{f,c}_{t+1}(\Omega^c) + (1-m^{np})V^{f,s}_t(\omega')\big\} + \\  \nonumber
& \hspace{10em} p^{\text{meet}}_t p^{\text{preg}}_t \big\{ m^{p} \cdot V^{f,ck}_{t+1}(\Omega^c) + (1-m^{p})V^{f,sk}_{t+1}(\omega')\big\}  \Big]  \bigg\},
\end{align}\vspace{-3em}
\begin{align*}
 \text{s.t. \ }  &  a' = R\cdot a  + W^f_t(z) - c  & \text{ (evolution of assets)}\\
 &  z' = z + \varepsilon^{z,f}_t, \ \ \varepsilon^{z,f}_t \sim \mathcal{N}(0;\sigma_{z,f}^2) &  \text{ (evolution of productivity)}\\
  & \log W^f_t = z_t + \text{Trend}^f_t, \ \ \  \text{Trend}^f_t = a^f_0 + a^f_1\cdot t  +  a^f_2 \cdot t^2 &  \text{ (labor income and trend)}\\
  & \Omega^c = \mathcal{M}^f(a',z') &  \text{ (marriage prospectives)}
\end{align*}

% 
The crucial difference here is the term $V^{f,sk}$: it reflects the fact that upon disagreement woman becomes a single female with a child.

\subsection{Couples Without Kids}
Couple is characterized by age $t$ (assumed the same for both spouses), marriage terms $\theta$, productivities of female and male $z^f$, $z^m$ and additive marriage surplus $\psi$. Couple maximizes weighted expected lifetime utility, and $\theta$ represents share of female in this objective function, therefore $\Omega = \{\theta,\psi,z^f,z^m\}$ represent characteristics of couple.

The key element shaping couple's decisions is participation constraints: each period the expected lifetime utility of staying in couple should be not less that expected lifetime utility of getting a divorce and becoming a single agent. Depending on realization of the shocks, if it is possible to satisfy both participation constraints the next period (so no divorce happens, $d = 0$) state is $\Omega'$. If participation constraints cannot be satisfied, the couple gets a divorce $d = 1$ and its characteristics are dissolved onto $\omega^{df}$ and $\omega^{dm}$.  In more details this is described in Section [...].

Fertility dimension is an important element. Each period couples get fertility shock $p^{\text{preg}}_t$. If it happens, they become couple with a child if they stay together (renegotiation happens after they know realization of the shock), or become single male and single female with a child if they do not. If the shock does not happen, they still may make a decision to get a baby \emph{after} they renegotiate the marriage terms if they stay together. This timing is subtle, but is required for everything to be internally consistent, see Appendix ... for more details on how timing interacts with renegotiation.

After realization of all the shocks and making fertility and divorce decisions in the current period the problem of the couple that stays childless is
\begin{align}& \hspace{5em}  V^{c}_t(\Omega) = \max\limits_{c^f,c^m}  \bigg\{ \theta\cdot u(c^f) + (1-\theta)\cdot u(c^m) + \psi +  \label{vf-c} \\   \nonumber
 &  \beta \E_t \Big[   (1 - p^{\text{preg}}_t)\cdot \left\{ (1-d^{\text{np}})\cdot \max\left\{ V^{ck}_{t+1}(\Omega'),V^{c}_{t+1}(\Omega')\right\} + d^{\text{np}}\cdot [ \theta V_{t+1}^{f,s}(\omega^{df}) + (1-\theta)V_{t+1}^{m,s}(\omega^{dm})]\right\}  +  \\  \nonumber
& \hspace{5em} p^{\text{preg}}_t\cdot \left\{ (1-d^{\text{p}})\cdot V^{ck}_{t+1}(\Omega') + d^{\text{p}}\cdot [ \theta V_{t+1}^{f,sk}(\omega^{df}) + (1-\theta)V_{t+1}^{m,s}(\omega^{dm})]\right\} \Big] \bigg\},
\end{align}\vspace{-2em}
\begin{align*}
\text{s.t. \ }& a' + c = R\cdot a  + W^m_t(z^m) + W^f_t(z^f) & \text{(evolution of joint assets)},\\
				 & c = [(c^f)^{1+\rho_c} + (c^m)^{1+\rho_c}]^{\frac1{1+\rho_c}} & \text{(increasing returns in consumption)},\\
				 &  z^{f\prime} = z^f + \varepsilon^{z,f}_t, \ \ \varepsilon^{z,f}_t \sim \mathcal{N}(0;\sigma_{z,f}^2) &  \text{ (evolution of female productivity)}\\
				 &  z^{m\prime} = z^m + \varepsilon^{z,m}_t, \ \ \varepsilon^{z,m}_t \sim \mathcal{N}(0;\sigma_{z,m}^2) &  \text{ (evolution of male productivity)}\\
                    & \psi' = \psi + \varepsilon^{\psi}_t, \ \ \varepsilon^{\psi}_t \sim \mathcal{N}(0;\sigma_{\psi}^2)  & \text{(evolution of marriage surplus),} \\
                    & (\Omega',\omega^{df},\omega^{dm}) = \mathcal{R}(\theta,a',z^{m\prime},z^{f\prime},\psi') & \text{(renegotiation correspondence)},
\end{align*}

\subsubsection{Defining Individual Values}
The couple's collective value function $V^c$ is the object relevant for making decisions of consumption and fertility. Making marriage and divorce decisions involves individual value functions of spouses living in the couple. I denote them as $V^{f,c}$ and $V^{m,c}$. These value functions are obtained by plugging optimal decisions of couple into intertemporal utilities of each agent and accounting for possible transitions. Note that they cannot be derived from a maximization problem, and common properties of value functions such as envelope theorems do not hold for them.

Namely, given couple's decisions $c^f(\Omega)$, female's value of being in couple is defined recursively as
\begin{align}
& \hspace{5em}  V^{f,c}_t(\Omega) =    u(c^f) + \psi +  \\   \nonumber
 &  \beta \E_t \Big[   (1 - p^{\text{preg}}_t)\cdot \left\{ (1-d^{\text{np}})\cdot \max{}^C \left\{ V^{f,ck}_{t+1}(\Omega'),V^{f,c}_{t+1}(\Omega')\right\} + d^{\text{np}}\cdot [ V_{t+1}^{f,s}(\omega^{df})]\right\}  +  \\  \nonumber
& \hspace{5em} p^{\text{preg}}_t\cdot \left\{ (1-d^{\text{p}})\cdot V^{f,ck}_{t+1}(\Omega') + d^{\text{p}}\cdot [ V_{t+1}^{f,sk}(\omega^{df}) ]\right\} \Big] 
\end{align}
where operator $\max{}^C\{A,B\} \equiv A\cdot \I[V^{ck}_{t+1}(\Omega')\geq V^{c}_{t+1}(\Omega')] + B\cdot \I[V^{ck}_{t+1}(\Omega')< V^{c}_{t+1}(\Omega')]$. It reflects the fact that couple makes joint decision that is not necessary individually optimal.

Another important property of individual values is that
\begin{equation} V^{c}_t(\Omega) \neq \theta\cdot V^{f,c}_t(\Omega) + (1-\theta)\cdot V^{m,c}_t(\Omega),\label{tht_noneq}\end{equation}
the main reason for this is that $\theta$ changes in the future as a result of random shocks, and this change is not likely to be symmetric. This, in particular, causes couple's expected future value function to be discontinuous in $(a,z)$. See Appendix [...] for some additional discussion.

\subsection{Couples With Kids}
State space for couples is the same as for singles, except for additional variable $\xi$ representing child's age, therefore $\Omega = \{\theta,\psi,z^f,z^m,\xi\}$. I consider just two states for $\xi \in \{y,s\}$ representing young and school-age child. At birth the child is young, then each period with probability $p^{s}$ the child becomes of school-age. The difference between young and school-age children are mother's time requirement, namely young children require more time and elder --- more money.

Couples with kids have additional choice variable $q$, that can be interpreted as flow of child quality or child consumption. Child quality is produced using mother's labor $l_f$ and monetary expenditures $x$ according to constant returns to scale production function. Labor input to child quality causes female to lose part of her labor income. The production function depends on age, namely I parametrize it as $q = [\mu_\xi\cdot l_f^{\theta_q} + (1-\mu_\xi)\cdot x^{\theta_q}]^{\frac1{\theta_q}}$. This follows, for instance, [Sommer ...], with the addition of changing labor share $\mu$.

\begin{align}& \hspace{5em}  V^{ck}_t(\Omega) = \max\limits_{c^f,c^m,q,l_f,x}  \bigg\{ \theta\cdot u(c^f,q) + (1-\theta)\cdot u(c^m,q) + \psi + \label{vf_ck} \\  \nonumber
 & \hspace{10em} \beta \E_t \Big[   \left\{ (1-d)\cdot   V^{ck}_{t+1}(\Omega') + d\cdot [ \theta V_{t+1}^{f,sk}(\omega^{df}) + (1-\theta)V_{t+1}^{m,s}(\omega^{dm})]\right\} \Big] \bigg\},
\end{align}\vspace{-2em}
\begin{align*}
\text{s.t. \ } & a' + c + x = R\cdot a  + W^m_t(z^m) + (1-l_f)\cdot W^f_t(z^f) & \text{(evolution of joint assets)},\\
                    & c = [(c^f)^{1+\rho_c} + (c^m)^{1+\rho_c}]^{\frac1{1+\rho_c}} & \text{(increasing returns in consumption)},\\
                    & q = [\mu_\xi\cdot l_f^{\theta_q} + (1-\mu_\xi)\cdot x^{\theta_q}]^{\frac1{\theta_q}} & \text{(production function of child quality)},\\
                    &  z^{f\prime} = z^f + \varepsilon^{z,f}_t, \ \ \varepsilon^{z,f}_t \sim \mathcal{N}(0;\sigma_{z,f}^2) &  \text{ (evolution of female productivity)}\\
				 &  z^{m\prime} = z^m + \varepsilon^{z,m}_t, \ \ \varepsilon^{z,m}_t \sim \mathcal{N}(0;\sigma_{z,m}^2) &  \text{ (evolution of male productivity)}\\
                    & \psi' = \psi + \varepsilon^{\psi}_t, \ \ \varepsilon^{\psi}_t \sim \mathcal{N}(0;\sigma_{\psi}^2)  & \text{(evolution of marriage surplus),} \\
                   &  \P(\xi' = s | \xi = y) = p^s, \ \ \P(\xi' = s | \xi = s) = 1 & \text{(evolution of child's age)},\\
                    & (\Omega',\omega^{df},\omega^{dm}) = \mathcal{R}(\theta,a',z^{m\prime},z^{f\prime},\psi',\xi) & \text{(renegotiation correspondence)},
\end{align*}

\subsection{Single Mothers}
State space for single mothers contains individual productivity and child's age. Single mothers cannot marry, but have chance $p^{\text{out}}$ of recovering their marriage prospectives. In the model this is captured by transition of them to single females. This form is somewhat restrictive, although any other option requires modeling marriage of single mothers separately. This is a complex issue, for instance, males would seek females who already have kids if we do not make difference between step and own children. Since for the purposes of the model option of being single mother is used mainly as a factor affecting bargaining, I introduce additional (constant) utility factor $S$ to flexibly control utility of exiting the state. This can be interpreted as equivalent of utility of kids from past marriages when entering the marriage market again. This form seems flexible enough to capture variation in an option of being a single mother without putting extra computational and modeling burden.

The problem of a female who stays single mother is
\begin{align}V^{f,sk}_t(\omega) = \max\limits_{c} & \bigg\{ u(c,q) + \beta \E_t \Big[ p^{\text{out}} \cdot \left\{ V^{f,s}_{t+1}(\omega') + S\right\} +(1-p^{\text{out}})V^{f,sk}_{t+1}(\omega') \Big]  \bigg\},
\end{align}\vspace{-1em}
\begin{align*}
 \text{s.t. \ }  &  a' = R\cdot a  + (1-l_f)W^f_t(z) - c  & \text{ (evolution of assets)}\\
 & q = [\mu_\xi\cdot l_f^{\theta_q} + (1-\mu_\xi)\cdot x^{\theta_q}]^{\frac1{\theta_q}} & \text{(production function of child quality)},\\
 &  z' = z + \varepsilon^{z,f}_t, \ \ \varepsilon^{z,f}_t \sim \mathcal{N}(0;\sigma_{z,f}^2) &  \text{ (evolution of productivity)}\\
 &  \P(\xi' = s | \xi = y) = p^s, \ \ \P(\xi' = s | \xi = s) = 1 & \text{(evolution of child's age)},\\
\end{align*}

\subsection{Marriage Market}
Single males and females meet potential partners of identical age and characteristics that depend on their assets $a$ (chosen in the previous period) and productivity $z$ (revealed after realization of shock):
\begin{equation}\log a^p = \log a + \varepsilon^{a,p}, \ \ \varepsilon^{a,p} \sim \mathcal{N}(0,\sigma_{a,p}^2),\end{equation}
\begin{equation}	z^p = z + \varepsilon^{z,p}, \ \ \varepsilon^{z,p} \sim \mathcal{N}(0,\sigma_{z,p}^2),\end{equation}
for example, $\sigma_{a,p} = 0.1$ can be read as standard deviation of partner's assets being 10\% around of what agent has. 
Additionally, initial marriage surplus (love shock) is drawn from
\begin{equation} \psi \sim \mathcal{N}(0,\sigma_{\psi,0}^2).\end{equation}

If people agree to be a couple they pool their assets $a^c = a + a^p$. If no unplanned pregnancy happens, people agree to marry if set of mutually satisfactory marriage terms
\begin{equation}\small \Theta^{np}_t = \left\{ \theta : V_t^{f,c}(\theta,\psi,a^c,z^f,z^m) \geq V_t^{f,s}(a^f,z^f), \ \ V_t^{m,c}(\theta,\psi_0,a^c,z^f,z^m) \geq V_t^{m,s}(a^m,z^m)\right\}\end{equation}
is non-empty. If potential couple gets the pregnancy shock, the set changes to
\begin{equation} \small \Theta^{p}_t = \left\{ \theta : V_t^{f,ck}(\theta,\psi,a^c,z^f,z^m,y) \geq V_t^{f,sk}(a^f,z^f,y), \ \ V_t^{m,ck}(\theta,\psi_0,a^c,z^f,z^m,y) \geq V_t^{m,s}(a^m,z^m)\right\} \end{equation}
(at birth child's age is $\xi = y$).

If the respective set is non-empty, initial bargaining power of couple is determined by symmetric Nash Bargaining, for instance:
\small
\begin{align}&\theta^*(a,z,\psi,\epsilon^{a,p},\epsilon^{z,p}) = \label{nbs} \\ \nonumber &= \argmax\limits_{\theta \in \Theta^{np}} \left[V_t^{f,c}(\theta,\psi,a^c,z^f,z^m) - V_t^{f,s}(a^f,z^f)\right] \times \left[V_t^{\vphantom{f}m,c}(\theta,\psi,a^c,z^f,z^m) - V_t^{m,s}(a^m,z^m)\right],\end{align}
and analogously for the case when pregnancy shock happened.

From the prospective of a single agent, this set is determined by realization of $\psi$ and partner's characteristics $\epsilon^{a,p}$ and $\epsilon^{z,p}$. Given them, marriage function $m$ is binary:
\[m^{np}_t(a,z,\psi,\epsilon^{a,p},\epsilon^{z,p}) = \I(\Theta^{np}_t \neq \varnothing),\]
where variables affecting set $\Theta$ are unambiguously recovered from arguments of $(a,z,\psi,\epsilon^{a,p},\epsilon^{z,p})$ as described above.

%Because of good properties of Nash Bargaining, expected future value function is continuous with respect to of $a$ and $z$. See Appendix [...] for additional discussion of this.

Let $\omega = (a,z)$ and $\epsilon = (\psi,\epsilon^{a,p},\epsilon^{z,p})$. This allows to define a random function
\[\Omega^c = \mathcal{M}_t(\omega) = M_t(\omega,\epsilon) = (\theta^*,\psi,a^c,z^m,z^f),\]
so the future characteristics of couple are function of current characteristics of individual and three-dimensional random shock.
\subsection{Renegotiation And Divorce}

Upon divorce, $\kappa\cdot a$ of assets disappears and the rest is split evenly. This generalizes [Voena ...], where I put additional parameter $\kappa$ captures can rationalize extra costs of divorce for more wealthy couples. [Voena ...] argues that this approximates well actual property divisions. So, the splitting rule is
\[a^{df} = 0.5\cdot (1-\kappa)\cdot a,  \ \ a^{dm} = 0.5\cdot (1-\kappa)\cdot a.\]
This splitting rule defines mapping of couple's state $\Omega$ to individual states $\{\omega^{df},\omega^{dm}\}$ in case of divorce.

The option to leave with a share of couple's assets creates pressure for participation constraints. As argues by [...], the household prefers to keep $\theta$ constant. However, after shocks to $z$ and $\psi$ happen, one (or both) participation constraints might be violated. To address this formally, consider couple that already has kids. I define set
\begin{equation}\Theta_t = \left\{ \tilde\theta : V^{f,ck}_t(a,\tilde\theta,\psi,z^f,z^m,\xi) \geq V_t^{f,sk}(a^{df},z^f,\xi) , \ \ V_t^{m,ck}(a,\tilde\theta,\psi,z^f,z^m,\xi)\geq V_t^{m,s}(a^{dm},z^m) \right\} \label{reneg-with-kids}
,\end{equation}
and if couple starts with current value $\theta_{0}$ there are three possible options:
\begin{enumerate}
\item \textit{Status quo.} If $\theta_0 \in \Theta_t$, current marriage terms are satisfactory for both spouses so $\theta_t = \theta_{0}$.
\item \textit{Renegotiation.} If $\theta_0 \not\in \Theta_t$, but $\Theta_t \neq \varnothing$, then spouses can continue to be married if they change their marriage terms. In this case they pick the closest the new value from set $\Theta_t$ such that it is the closest to the old marriage terms: $\theta_t = \min\limits_{\tilde\theta\in\Theta_t} |\tilde\theta - \theta_0|$. 
\item \textit{Divorce.}  If $\Theta_t = \varnothing$, so if there do not exist any marriage terms satisfactory for both spouses, then divorce happens according to the rule described above.
\end{enumerate}
Picking $\theta$ that is the closest to the old one is not a random choice: it is rationalized by models with limited commitment models, where change in $\theta$ represents the value of Lagrange multiplier on binding participation constraint. Intuitively, since couple does not like $\theta$ to be changing and everything is continuous in $\theta$, the couple prefers small changes to larger ones, and this drives the result.

Renegotiation that involves making discrete decisions has few more issues as backward induction has to be used: fertility decisions have to be made after new $\theta$ is set, and these decisions might depend on $\theta$, therefore inequalities defining set $\Theta_t$ need to be modified. This is an artifact of assuming that fertility decisions are made collectively and it can be the case that having a baby benefits couple but hurts wife alone (mainly by worsening her outside option), and such scenario of renegotiation insures that participation constraints hold after the fertility decisions are made. See Appendix \ref{ren-disc} where I elaborate on this.

\subsection{Additional Details}
The model I solve an estimate contains few additional things that are omitted for more clear exposition above.

First, females may exit the state of being single mother immediately. This transition is assumed to happen after divorce or failure to agree to marry, therefore in the value functions and in (re)negotiation I use $V^{f,s?} = p^{\text{out}} [V^{f,s} + S] + (1-p^{\text{out}})V^{f,sk}$ instead of just $V^{f,sk}$. This allows for quicker re-marriage and easier implementation of counterfactual in which females do not have risk of becoming single mother at all. I use notation of $V^{f,s?}$ in Appendix \ref{add-trans} where I show exact value functions.

Second, I assume that having a baby is possible only before certain age $\bar{T}$. After that, value functions and negotiation rules are modified in a straightforward manner.

Finally, to simplify computations I also do not allow marriage status and marriage terms to change after certain $\bar{T}_2$. Therefore I drop participation constraints in the couples, the value functions look the same but do not allow for $\theta$ to change.

\subsection{Functional Forms}
I use standard CES utility functions for individuals and couples with no kids: $u(c) = \frac{c^{1-\sigma}}{1-\sigma}$.  When kids appear, people extract additional utility of child quality, I use the following non-separable utility function:
\[u(c,q) = \frac{\left(e^{\phi_0}\cdot c^{1-\phi_1} \cdot (\phi_2 + q)^{\phi_1}\right)^{1-\sigma}}{1-\sigma},\]
parameter $\phi_0$ represents fixed gain in marginal utility because of having a baby, $\phi_1$ drives relative importance of child quality in utility function and $\phi_2$ is a ``luxury good'' parameter that can rationalize large child expenditures of wealthy couples that are discussed in the estimation section.

### 2019-05-03 main/tech-appendix.tex
Re-organized into few files

\newpage
\section*{Very Technical Appendix, Not For Final Version}

\section{Expressions for Value Functions}
\subsection{Notation}
This technical note shows exact expressions for approximating value functions. For the couple's state I use partition $\Omega = (a,O)$, where $O = (\theta,\psi,z^f,z^m)$. For the age variable I put $\xi\in\{y,s\}$ counts as a separate letter, so states for couples with kids are noted like $cky$ and $cks$. Let $p^s$ represents probability that the child ``grows up'' so couple switches from $cky$ to $cks$. For individual's state there are only two variables $\omega = (a,z)$, and child's age for single mother is reported as state $sky$ if $\xi = y$ and $sks$ if $\xi = y$.

Grid consists of all possible combinations of $a$ and $O$ ($a$ and $z$ in single's case), so \[G = \{ \{a_1,...,a_I\} \times \{o_1,...,o_J\}\}\]

Let $i$ refers to assets position and $j$ refers to values of everything else. Finally, indices $q$ refer to quadrature nodes, and $w_q$ to quadrature weights that add up to 1.

\subsection{Single Agents}
Consider females, males are handled analogously. Value function \ref{single-fem} can be written as
\[V^{f,s}_t(a,z) = \max\limits_{c} \left\{ u(c) + \beta \mathcal{V}^{f,s}_{t+1}(s,z)\right\}, \ \ c + s = M(a,z),\]
Here  $\mathcal{V}^{f,s}_{t+1}$ represents expectation of tomorrow's individual value function, accounting for all possible transitions. Argument $s$ refers to savings made in the current period (it is not necessary equal to future assets as future assets change if the partner is met), and $z$ refers to current value of productivity. Approximating $\mathcal{V}_{t+1}(s,z)$ (i.e. integration) is the key challenge. Note that it does not depend on current assets, only on savings $s$ that define future assets position. $M(a,z) = R\cdot a + \exp(z + \text{Trend}_t)$ is the amount of money in the end of the period.

I treat the function $\mathcal{V}^{f,s}_{t+1}(s,z)$ as a collection of functions $\mathcal{V}^{f,s}_{mj,t+1}$ for each value of current productivity $z_j$ and savings $s_m$. I generate quadrature nodes that consist of shock to individual productivity $\epsilon^{z,f}$, shocks to partner's assets and productivity positions
$(\epsilon^{a,p},\epsilon^{z,p})$ and initial couple's surplus $\psi$. I denote these nodes as $x_q = (\epsilon^{z,f}_q,\epsilon^{a,p},\epsilon^{z,p},\psi_q)$, in total monomial rule results in $Q = [...]$ nodes. Weight $w_q$ corresponds to each node. Given savings $s$ and grid point in $z_j$ I generate partner's and couple's characteristics. 
\[a^p_{mq} = s_m\cdot \exp(\epsilon^{a,p}_q), \ \ a^c_{mq} = s_m\cdot [1 + \exp(\epsilon^{a,p}_q)], \ \ z^m_{jq} = z^f_j \exp(\epsilon^{z,p}_q),\]
(in some sense in this setup by saving more individuals have chances of getting more wealthy partners).  Also, individual next period's productivity is generated according to $z^f_{jq} = z^f\cdot \exp(\epsilon^{z,f}_q)$. Note that $z^f_j$ refers to the current value and $z^f_{jq}$ refers to the next period's values;

This allows to generate potential couple's characteristics $(a^c_{mq},\psi_q,z^{f}_{jq},z^{m}_{jq})$, and combining it with value functions for the next period we can define:
\[\Theta^{np}_{mjq,t+1} = \left\{ \tilde\theta : V^{f,c}_{t+1}(a^c_{mjq}, \tilde\theta, \psi_q,z^{f}_{jq},z^{m}_{jq}) \geq V^{f,s}(s_m,z^f_{jq}), \ V^{m,c}_{t+1}(a^c_{mjq}, \tilde\theta, \psi_q,z^{f}_{jq},z^{m}_{jq}) \geq V^{m,s}(a^p_{mq},z^m_{jq}) \right\},\]
if $\Theta^{np}_{mjq,t+1}$ is empty\footnote{I figure this out by defining a uniformly spaced grid for $\theta \in [0.05,0.95]$. Because of Smolyak grid, this can be done very efficiently as $V^{f,c}_{t+1}(a^c_{mjq}, \tilde\theta, \psi_q,z^{f}_{jq},z^{m}_{jq})$ are just a polynomial in $\tilde{\theta}$.}
, $m^{np}_{mjq,t+1} = 0$ (so no marriage happens at node $jq$ if agents made savings of $s_m$). If the set is not empty, then $\theta_{mjq}$ is determined by Nash Bargaining \ref{nbs} and $m^{np}_{mjq,t+1} = 1$. Therefore at node $(s_m,z_j)$ future value function in case the partner is met and pregnancy did not happen can be approximated as
\[\E_t  \big\{ m^{np} \cdot V^{f,c}_{t+1}(\Omega^c) + (1-m^{np})V^{f,s}_t(\omega')\big\} \approx \mathcal{E}^{np}(s_m,z_j),\]
where
\[\mathcal{E}^{np}(s_m,z_j) = \sum\limits_{q=1}^{Q} w_q \cdot \left\{ m^{np}_{mjq,t+1}\cdot V^{f,c}_{t+1}(a^c_{mq},\theta_{mjq},\psi_q,z^{f\prime}_{jq},z^m_{jq}) + (1-m^{np}_{mjq,t+1})\cdot V^{f,s}_{t+1}(s_m,z_j)\right\}.\]
In exactly the same manner we define the set in the case the pregnancy shock did happen:
\[\Theta^{p}_{mjq,t+1} = \left\{ \tilde\theta : V^{f,ck}_{t+1}(a^c_{mjq}, \tilde\theta, \psi_q,z^{f}_j,z^{m}_{jq}) \geq V^{f,s?}(s_m,z^f_{j}), \ V^{m,ck}_{t+1}(a^c_{mjq}, \tilde\theta, \psi_q,z^{f}_{jq},z^{m}_{jq}) \geq V^{m,s}(a^p_{mq},z^m_{jq}) \right\},\]
where function $V^{f,s?} = p^{\text{out}}[V^{f,s} + S] + (1-p^{\text{out}})V^{f,sky}_t$ captures possible immediate transition out of being single mother. In the same way 
\begin{align*}\mathcal{E}^{p}(s_m,z_j) = \sum\limits_{q=1}^{Q} w_q \cdot  \Big\{ m^{p}_{mjq,t+1}\cdot V^{f,cky}_{t+1}(a^c_{mq},\theta_{mjq},\psi_q,z^f_j,z^m_{jq}) + (1-m^{p}_{mjq,t+1})\cdot V^{f,s?}(s_m,z_j) \Big\},
\end{align*}
lastly, the value of not meeting any partner is simply
\begin{align*}\mathcal{E}^{0}(s_m,z_j) = \sum\limits_{q=1}^{Q} w_q \cdot  V^{f,s}_{t+1}(s_m,z^f_{jq}),
\end{align*}
finally, the whole function $\mathcal{V}^{f,s}_{t+1}(s_m,z_j)$ can be approximated by weighted average
\[\mathcal{V}^{f,s}_{t+1}(s_m,z_j) \approx (1-p^{\text{meet}}_t) \cdot \mathcal{E}^{0}(s_m,z_j) +  p^{\text{meet}}_t(1-p^{\text{preg}}_t)  \cdot \mathcal{E}^{np}(s_m,z_j) 
+  p^{\text{meet}}_t p^{\text{preg}}_t  \cdot \mathcal{E}^{p}(s_m,z_j).\]

After that I use linear interpolation in $s$ dimension to approximate $\mathcal{V}^{f,s}_{t+1}(s_m,z_j)$ for arbitrary $s$. Therefore I can get approximation of $\mathcal{V}^{f,s}_{t+1}(s,z_j)$. As a result, value function in each asset point $a_i$, $z_j$ is obtained by numerically solving
\[V^{f,s}_t(a_i,z_j) = \max\limits_{s \in [\underline{A},\overline{A}]} \left\{ u(M(a_i,z_j) - s) + \beta \mathcal{V}^{f,s}_{t+1}(s,z_j)\right\}.\]
I take $\underline{A}$ and $\overline{A}$ the same as upper and lover bound for individual's assets grid (this is important as approximation may be crude). I use Matlab's \textbf{fminbnd} to solve for the maximum point, although I also test that results are very similar when I just use discrete grid $s \in \mathcal{S} = M(a_i,z_j) \cdot\{0,...,0.99\}$ with some spacing, though this variation is much slower.

Choice of solver can generically be an issue as because of concerns described above function $\mathcal{V}^{f,s}_{t+1}(s_m,z_j)$ may be non-smooth or even discontinuous.

\subsection{Couples}
I consider couples that do not have kids yet as they have richer set of possible transitions.  Following [...] I split household's decision problem on intrahousehold, intratemporal and intertemporal levels, where first two can be solved analytically. For intrahousehold, I define
\[U(c,\theta) = \max\limits_{c^f,c^m} \left\{\theta u(c^f)  + (1-\theta)u(c^m) \right\}, \ \ \text{s.t.} \ \ \ c = \left[ (c^f)^{1+\rho_c} + (c^m)^{1+\rho_c}\right]^{\frac1{1+\rho_c}},\]
and given CES assumption this function has simple analytical form.

Intratemporal dimension is trivial, as total consumption is the only intratemporal choice.

For intertemporal I can write the problem \ref{vf-c} as
\[V^{c}_{t+1}(a,\theta,\psi,z^f,z^m) = \max\limits_{c} \left\{ U(c,\theta) + \beta \mathcal{V}_{t+1}(s,\theta,\psi,z^f,z^m) \right\},\]
where $\mathcal{V}_{t+1}(s,\theta,\psi,z^f,z^m)$ consists of two terms: what happens if pregnancy shock arrives and what if it does not.

If the shock arrives, in the next period couple is in state $cky$, and if it decides to divorce female becomes a single mother (and may quit this state immediately), and male becomes just a single male. I approximate $\mathcal{V}_{t+1}(s,\theta,\psi,z^f,z^m)$ here.

I write current state as combination of $(a_i,O_j)$. Function $\mathcal{V}_{t+1}(s,\theta,\psi,z^f,z^m)$ depends on current state $O= (\theta,\psi,z^f,z^m)$ and future savings $s$.  Integration nodes contains $x_q = (\epsilon^{z,f}_q,\epsilon^{z,m}_q,\epsilon^{\psi}_q)$. I take grid with respect to current states, so $O_j= (\theta_j,\psi_j,z^f_j,z^m_j) \in O$ and future savings $s \in A$, so I again approximate expected future value function at each combination of $s_m$ and $O_j$, so I compute values $\mathcal{V}_{mj,t+1} = \mathcal{V}_{t+1}(s_m,\theta_j,\psi_j,z^f_j,z^m_j)$. Given each $j$, I can compute next period values of $(\psi_{jq},z^f_{jq},z^m_{jq})$ by adding shocks given by quadrature notes to the current point. I also can compute divorce options: $\omega^{df}_{mjq} = ( (1-\kappa)\cdot 0.5 \cdot s_m, \ z^f_{mjq})$ and $\omega^{dm}_{mjq} = ( (1-\kappa)\cdot 0.5 \cdot s_m, \ z^m_{mjq})$. This allows to define for each quadrature node a renegotiation set:
\[\Omega^{p}_{mjq,t+1} = \left\{ \tilde\theta \ : \ V^{f,cky}_{t+1}(s_m,\tilde\theta,\psi_{jq},z^f_{jq},z^m_{jq}) \geq V^{f,s?}_{t+1}(\omega^{df}_{mjq}), \ \ V^{m,cky}_{t+1}(s_m,\tilde\theta,\psi_{jq},z^f_{jq},z^m_{jq}) \geq V^{m,s}_{t+1}(\omega^{dm}_{mjq}) \right\},\]
and then if $\Omega^{p}_{mjq,t+1}$ is not empty, I pick from it next period's $\theta_{mjq}$ closest to initial $\theta_j$ and set $d^{p}_{mjq} = 0$, otherwise divorce happens and $d^{p}_{mjq} = 0$.

Therefore expected future value if pregnancy shock arrives is 
{\small
\[\mathcal{E}^{p}(s_m,O_j) = \sum\limits_{q=1}^{Q} w_q \cdot\left\{ d^p_{mjq}\cdot V_{t+1}^{cky}(s_m,\theta_{mjq},\psi_{jq},z^f_{jq},z^m_{jq}) + (1-d^p_{mjq}) [\theta_{j}\cdot V^{f,sky}(\omega^{df}_{mjq}) + (1-\theta_j)V^{m,s}(\omega^{dm}_{mjq})] \right\} \]
}
If pregnancy shock does not arrive, then renegotiation should account for discrete decision of whether to have a kids that happens afterwards, and this decision may depend on $\theta$. Namely, I define
\[k_{mjq}(\tilde\theta) = \I\left[ V^{cky}_{t+1}(s_m,\tilde\theta,\psi_{jq},z^f_{jq},z^m_{jq}) \geq V^{c}_{t+1}(s_m,\tilde\theta,\psi_{jq},z^f_{jq},z^m_{jq})\right],\]
then I define individual value accounting for this choice:
\[V^{f,c?}_{t+1}(s_m,\tilde\theta,\psi_{jq},z^f_{jq},z^m_{jq}) = k_{mjq}(\tilde\theta)\cdot V^{f,cky}_{t+1}(s_m,\tilde\theta,\psi_{jq},z^f_{jq},z^m_{jq}) + (1-k_{mjq}(\tilde\theta))\cdot V^{f,c}_{t+1}(s_m,\tilde\theta,\psi_{jq},z^f_{jq},z^m_{jq}),\]
and the renegotiation set is defined as
\[\Omega^{np}_{mjq,t+1} = \left\{ \tilde\theta \ : \ V^{f,c?}_{t+1}(s_m,\tilde\theta,\psi_{jq},z^f_{jq},z^m_{jq}) \geq V^{f,s}_{t+1}(\omega^{df}_{mjq}), \ \ V^{m,c?}_{t+1}(s_m,\tilde\theta,\psi_{jq},z^f_{jq},z^m_{jq}) \geq V^{m,s}_{t+1}(\omega^{dm}_{mjq}) \right\},\]
and we define $\theta_{mjq}$ and $d^{np}_{mjq}$ analogously.

Therefore we define expected future value if pregnancy shock does not arrive to be
{\small
\[\mathcal{E}^{np}(s_m,O_j) = \sum\limits_{q=1}^{Q} w_q \cdot\left\{ d^{np}_{mjq}\cdot V_{t+1}^{c?}(s_m,\theta_{mjq},\psi_{jq},z^f_{jq},z^m_{jq}) + (1-d^{np}_{mjq}) [\theta_{j}\cdot V^{f,s}(\omega^{df}_{mjq}) + (1-\theta_j)V^{m,s}(\omega^{dm}_{mjq})] \right\} \]
}
so we can finally define value of $\mathcal{V}$ at points $s_m$:
\[\mathcal{V}^{c}_{t+1}(s_m,O_j) = p_t^{\text{preg}}\cdot \mathcal{E}^{p}(s_m,O_j) + (1-p_t^{\text{preg}})\cdot \mathcal{E}^{np}(s_m,O_j),\]
and then values at arbitrary points $s$ are obtained by linear interpolation. Therefore the ultimate problem is
\[V^{c}_t(a_i,O_j) = \max\limits_{s \in [\underline{A},\overline{A}]} \left\{ U(M(a_i,O_j) - s,\theta_j) + \beta \mathcal{V}^{c}_{t+1}(s,O_j)\right\}.\]

\subsection{Couples With Kids} 
Namely, for intrahousehold allocation of consumption I define
\[U(c,q,\theta) = \max\limits_{c^f,c^m} \left\{\theta u(c^f,q)  + (1-\theta)u(c^m,q) \right\}, \ \ \text{s.t.} \ \ \ c = \left[ (c^f)^{1+\rho_c} + (c^m)^{1+\rho_c}\right]^{\frac1{1+\rho_c}},\]
and it still has simple tractable form.

For intratemporal level I have to solve
\[U(m,W^f,\xi,\theta) = \max\limits_{c,q,l_f,x} U(c,q,\theta), \ \ \ \text{s.t.} \ \ \ c + x + l_f\cdot W^f = m, \ \ q =[\mu_\xi\cdot l_f^{\theta_q} + (1-\mu_\xi)\cdot x^{\theta_q}]^{\frac1{\theta_q}}. \]
this function can also be derived analytically given my assumption on functional forms.

After aggregation, the intertemporal problem looks similar to the one for singles:
\[V^{cky}(a,\theta,\psi,z^f,z^m) = \max\limits_{m} \left\{ U(m,W^f_t(z^f),\xi,\theta) + \beta \mathcal{V}^{ck}_{t+1}(s,\theta,\psi,z^f,z^m,\xi)\right\}, \ \ \text{s.t.} \ \ m + s = M(a,z^f,z^m),\]
where definition of $\mathcal{V}$ is analogous.

### 2019-05-03 main/theory-appendix.tex
Re-organized into few files

\section{Additional Notes on Model}
\subsection{Continuity of Value Functions}
This part discusses violations of continuity caused by the dynamic bargaining structure. The general conclusion is that collective value function does is not necessary continuous in couple's characteristics, and this discontinuity propagates onto all value functions in the model.

To simplify the example, let's consider value function of couple with kids. Similarly to equation \ref{tht_noneq}, we can write
\[ V^{ck}_t(\Omega) = \theta\cdot V^{f,ck}_t(\Omega) + (1-\theta)\cdot V^{m,ck}_t(\Omega) + \Delta(\Omega),\]
and use it to substitute the term under the expectation in equation \ref{vf_ck} 
\begin{align*} & (1-d)\cdot   V^{ck}_{t+1}(\Omega') + d\cdot [ \theta V_{t+1}^{f,sk}(\omega^{df}) + (1-\theta)V_{t+1}^{m,s}(\omega^{dm})] =    V^{ck}_{t+1}(\Omega') + \\
 &  \ \ \ \theta \cdot d(\Omega') \cdot [V_{t+1}^{f,sk}(\omega^{df}) - V_{t+1}^{f,ck}(\Omega')] + (1-\theta) \cdot d(\Omega') \cdot [V_{t+1}^{m,sk}(\omega^{dm}) - V_{t+1}^{m,ck}(\Omega')] - d(\Omega')\cdot \Delta(\Omega'),
\end{align*}
if we consider couple that is close to divorce, so there exist two $\Omega'_1$ and $\Omega'_2$  that are close to each other according to some metric, but $d(\Omega'_2) = 1$ and $d(\Omega'_1) = 0$. Note that around the divorce threshold by [...] \textbf{both} spouses should be indifferent between staying on leaving, therefore
\[ V_{t+1}^{f,sk}(\omega^{df}) - V_{t+1}^{f,ck}(\Omega')\approx 0, \ \ V_{t+1}^{m,sk}(\omega^{dm}) - V_{t+1}^{m,ck}(\Omega') \approx 0,\]
therefore if $\Delta(\Omega')$ is non-zero in these points, this whole expression jumps where value of $d$ jumps. [It is not clear whether it is actually zero or not, but in simulations it does look like it is positive.]

\subsection{Renegotiation With Collective Discrete Decisions\label{ren-disc}}
This section explains how discrete decisions interact with couple's renegotiation. The theory requires participation constraints to be satisfied \textit{after} any discrete choice. In the context of fertility decisions, that means that both spouses can predict future decisions and re-negotiate $\theta$ accounting for this prediction (rather than assuming that fertility status does not change).

With slight abuse of notation, let's suppose that couple gets collective value $V^{ck}(\theta,\Omega)$ if it makes a baby and $V^{c}(\theta,\Omega)$ is it does not. Given $\theta$ total value before the decision is made is \[V^{c?}(\theta,\Omega) = \max\{ V^{ck}(\theta,\Omega), V^{c}(\theta,\Omega)\}\]
after that I define
\[V^{f,c?}(\theta,\Omega) = V^{f,ck}(\theta,\Omega)\cdot \I[V^{ck}(\theta,\Omega) \geq V^{c}(\theta,\Omega)] + V^{f,c}(\theta,\Omega) \cdot \I[V^{ck}(\theta,\Omega) < V^{c}(\theta,\Omega)],\]
note that it is not necessary equal to $\max\{ V^{f,ck},V^{f,c} \}$. Similarly we can define $V^{m,c?}(\theta,\Omega)$.

The main thing to notice is that although $V^{c?}$ is continuous in $\theta$ individual values $V^{f,c?}$ and $V^{m,c?}$ are not, unless both spouses always have agreement on whether to make a baby: $V^{ck} \geq V^{c} \Leftrightarrow V^{f,ck} \geq V^{f,c} \Leftrightarrow V^{m,ck} \geq V^{m,c}$. Therefore set
\[\Omega = \left\{\tilde{\theta} : V^{f,c?}(\tilde\theta,\Omega) \geq V^{f,s}(\omega^{df}), \ \  V^{m,c?}(\tilde\theta,\Omega) \geq V^{m,s}(\omega^{dm}) \right\}\]
may be disjoint, and careful approximation of $V$ with respect to $\Omega$ is required.

The possible solution to it would be to limit fertility decisions to the cases where both spouses agree on having kids, this, however, goes against the logic where couple maximizes collective utility subject to participation constraints.

\subsection{Accounting for Additional Transitions\label{add-trans}}
In value functions I stated above I assumed that females stays single mother for at least one period if it exits a couple with a kid or if she has a baby at the moment of marriage and disagrees to marry. In the actual model I relax this assumption so the female can stop being single mother immediately. This is needed, for instance, for evaluating counterfactuals in which female can exit couple without pressure of being a single mother.

I assume that at the moment of (re)negotiation female is unsure about whether she is going to be a single mother. Therefore, I define expected value of leaving the couple to be
\[V_t^{f,c?}(\omega) = p^{\text{out}}[V_t^{f,s}(\omega) + S] + (1-p^{\text{out}})V_t^{f,sk}(\omega),\]
therefore equaiton \ref{single-fem} becomes
\begin{align}V^{f,s}_t(\omega) = \max\limits_{c} & \bigg\{ u(c) + \beta \E_t \Big[ (1 - p^{\text{meet}}_t)\cdot V^{f,s}_{t+1}(\omega') + \label{single-fem-true} \\  \nonumber
& \hspace{7em} p^{\text{meet}}_t (1-p^{\text{preg}}_t) \big\{ m^{np} \cdot V^{f,c}_{t+1}(\Omega^c) + (1-m^{np})V^{f,s}_t(\omega')\big\} + \\  \nonumber
& \hspace{10em} p^{\text{meet}}_t p^{\text{preg}}_t \big\{ m^{p} \cdot V^{f,ck}_{t+1}(\Omega^c) + (1-m^{p})V^{f,s?}_{t+1}(\omega')\big\}  \Big]  \bigg\},
\end{align}
the set is \ref{reneg-with-kids} becomes
\begin{equation}\Theta_t = \left\{ \tilde\theta : V^{f,ck}_t(a,\tilde\theta,\psi,z^f,z^m,\xi) \geq V_t^{f,s?}(a^{df},z^f,\xi) , \ \ V_t^{m,ck}(a,\tilde\theta,\psi,z^f,z^m,\xi)\geq V_t^{m,s}(a^{dm},z^m) \right\} \label{reneg-with-kids}
,\end{equation}
and the analogous term is introduced for initial negotiation of couples in case of pregnancy shock.

### 2019-05-22 main/draft.tex
Preparing literature part

@@
 \newcommand{\deq}{\stackrel{d}{=}}
-%\title{416-2 Final Project\\
-%Reproductive Technologies, Aging and Fertility Choice (Proposal)}
-%\author{Egor Kozlov}
+\title{Kids First or Marriage First?}
+\author{Egor Kozlov}
 \begin{document}
 %\begin{center}\textbf{Economics 416-1} \\ \emph{By Egor Kozlov}\end{center}
-%\maketitle
+\maketitle
 
@@
 I argue that this pattern is consistent with household bargaining models, and the main driver of the result is unequal exposure of spouses to the childcare costs: since being single mother is a bad outcome for an educated female, she is more likely to agree to enter the marriages of lower quality.
-I estimate a lifecycle model with dynamic bargaining and fertility decisions to show how much of these differences is driven by household bargaining and how childcare costs contribute to observable divorce rates.
+I estimate a lifecycle model with dynamic bargaining and fertility decisions to show how much of these differences is driven by household bargaining and how unplanned pregnancies creating unexpected childcare costs contribute to observable divorce rates.
 \end{abstract}
@@
 \section{Introduction}
+As of 2018, [40\%] of new births occurred by unmarried mothers.
+This number is rising for decades, but, more importantly, it is rather heterogeneous, reaching three-quarters for mothers who are African American, [...] under 25 and [...] for high school or less in education.
+But even for educated and older women this share is substantial [...].
+An increase in non-marital cohabitation is also a documented trend, despite of this, persistent non-marital cohabitation is still not the most common thing: according to SIPP 2014, only [...] of women 21--40 cohabitate for longer than 5 years without being married.
+A substantial part of children by unmarried parents are born in such cohabitating unions (SIPP [...]).
+
+These numbers have important economic implications: children born by unmarried women, as documented by, for instance [...], display significantly worse future outcomes in terms of education and income, and this pattern survives controlling in many dimensions.
+Although it is well understood that children of never married single mothers or children of blended families (when a woman marries someone else than child's father) have generally worse outcomes, even when parents marry each other after they made a baby their marriage is statistically less stable, as I argue in this paper, and the larger chances of marriage dissolution do affect future prospects of the children.
+
+This pairs well with common perception that some couples marry at or shortly after the moment they make a baby: long-lasting marital cohabitations with children are even less common (among ...).
+If cohabitation is viewed as uncommitted relationship, unexpected fertility may serve as a trigger to obtain more commitment.
+Narrowly, we can interpret this as shotgun marriage story, however in the times with large abortion access many unplanned pregnancies are aborted, and cohabitating couples seem to \emph{choose} to keep an unplanned baby.
+The whole role of marriage changes as it is more socially acceptable to cohabitate, though it is much more conventional to do it before having kids.
+As highlighted by, for instance, [...], despite of general decrease of role of marriage in the society, securing investment to children is one of the most important forces that makes people marry each other.
+
+This paper highlights one important aspect of this: I argue that when marriage decision is driven by having an unexpected child, bargaining friction makes marital outcome worse.
+The source of this friction is the large childcare costs and unequal relative exposure to them: as women bear relatively more childcare costs, they are more likely to enter relatively lower quality marriages as a commitment to share a part of these costs.
+This can rationalize observable differences in marital outcomes between couples with and without out-of-wedlock births.
+
+Namely, I document that females who had their first birth before their first marriage have higher chances of divorce in the future, this is mainly driven by women married to the parents of their children, and this is robust and relatively more pronounced for high education or high income groups.
+I then provide a model incorporating the bargaining mechanism that can rationalize these differences.
+I estimate the model parameters to match observable patterns in the data and show what it implies about the role of unequal exposure to childcare costs in observable marriage selection and divorce patterns.
+Shortly, the model suggests if people could perfectly control timing of their births, general share of divorced people would decrease by ..., especially in high-income groups.
+
+Section 2 of this paper discuss existing results and general trends that are already described by the literature.
+Section [...] discuss how I construct the dataset and timing variables, and Section 3 establishes main observations about differences in groups depending on their marriage timing, together with a number of additional considerations useful to understand the content of these differences.
+After that I explain theoretical model in Section 4, the process of setting its parameters in Section 5.
+Section 6 provides estimation results and counterfactual simulations implied by these results.
+Section 7 provides more general discussion and concludes.
+
 %Understanding how people make marriage decisions and more generally how household structure is formed is crucial for analyzing labor market outcomes, assets accumulation, various dimensions of inequality and children development outcome.
@@
 
-\subsection{Trends in Marriage And Divorce}
+\section{Existing Results}
 
-\subsection{Existing Results}
+My work builds on few branches of literature, following long tradition of modeling fertility and marriage markets in family economics.
+I am not aware of any studies that cover precisely the narrow issue I focus on, but there are few recently established result that cover related aspects and compliment what I do.
+Three main building blocks that I utilize are literature on marriage selection and dynamic bargaining, few detailed lifecycle models of fertility and unplanned pregnancies and number of descriptive studies of microdata describing reduced form effects of out-of-wedlock births, trends in divorce rates and heterogeneity in time use of parents.
 
-\section{Empirical Evidence}
+The main conceptual framework that I utilize are models of household's dynamic bargaining and limited commitment, extensively reviewed by
+
+\section{Empirical Strategy}
 The key empirical observation is association between marriage timing and divorce probabilities.
-Statistically, women who have their kids before they are married have significantly higher chance to be divorced.
+The goal is to argue that statistically women who have their kids before they are married have significantly higher chance to be divorced.
 American Community Survey (ACS) data are suitable for establishing this fact only of particular subsample of women: most importantly, those who had no more than one marriage in their life.
@@
 \subsection{Defining Income}
-This part describes how I define income partitions in ACS.
+This part describes how I define income partitions in ACS and SIPP.
 Shortly, I define income groups based on per hour income of working females, by excluding those who does not work from income considerations.
-In the appendix I try several
+In the appendix \ref{inc-part} I try several alternative partitions, shortly they do not affect the conclusions although they do affect magnitudes in different groups.
 
-For income percentiles I divide data on cells that consist of the US state, survey year (for ACS) and age group.
-I define age groups as 20--25, 26--30, 31--35, 36--40, and this classification is based on the age of female.
-Conditioning for year is done mainly to mitigate consequences of income decline in 2008.
+To define per-hour income, I restrict attention to working women only.
+I define working as those who are employed now, report to work more than 25 weeks in the last year, report non-missing labor income above or equal to $\$1000$ per year, report non-missing usual working hours above or equal to $5$ hours per week.
 
-I drop cells with less than 100 observations (for instance, Vermont in ACS 2015 has 31 females in age group 20--25), this constitutes around [2\%] of all observations.
+As weeks worked last year are defined only on intervals for ACS, I impute them using the midpoints of respective intervals, then define yearly hours as a product of usual hours and imputed weeks worked.
+This allows to define log per-hour income as log-difference of yearly log wage and salary income with yearly log hours.
+I residualize these income data by regressing it on 4th degree polynomial in age, dummies and linear trends in age for each US state and dummies for each ACS survey year.
 
+In SIPP 2014, I perform the similar operations.
+I exclude females who had less than $1000/12$ monthly income or had average weekly hours less than 5 during the last month.
+Then I divide monthly income on four times average weekly hours to get estimate for per-hour earnings.
+As before, I residualize logs of these income by regressing them on the same factors, except that instead of year controls (I only use 2014) I use month controls that allow to capture variation in number of work days per month.
+
+\section{Reduced Form Evidence}
 \subsection{Raw Comparisons}
@@
 \multicolumn{1}{c|}{}  &  &\small if M$\to$K &  \small if K$\to$M & & \small if M$\to$K & \small if K$\to$M \\\hline
-\textit{All females, 20--40}  & $23.1\%$ & $9.8\%$ & $17.3\%$ & $22.8\%$ & $24.0\%$ & $34.7\%$ \\
-\textit{All females, 35--40} & $17.8\%$ & $11.0\%$ & $22.3\%$ & $18.9\%$ & $28.1\%$ & $45.5\%$ \\\hline\hline
-\multicolumn{7}{|p{0.9\linewidth}|}{\footnotesize \textbf{Education partition.} Note: ``college graduates'' excludes ``some college'' and ``more than college''.}\\\hline
-\textit{High school only, 20--40} & $34.2\%$ & $13.3\%$ & $16.5\%$ & $33.4\%$ & $29.0\%$  & $34.1\%$ \\
-\textit{College graduates, 20--40} & $11.5\%$ & $5.7\%$ & $14.3\%$ & $10.5\%$ & $14.9\%$ & $28.1\%$ \\\hline\hline
-\multicolumn{7}{|p{0.9\linewidth}|}{\footnotesize \textbf{Income partition:} subsample of household heads/spouses living with children and nobody else.}\\\hline
-\textit{Low 30\% income, 20--40} & $33.0\%$ & $17.7\%$ & $22.8\%$ & $29.4\%$ & $34.3\%$ &  $36.0\%$ \\
-\textit{Mid 40\% income, 20--40} & $22.8\%$ & $7.6\%$ & $11.8\%$ & $21.9\%$ & $23.8\%$ & $31.1\%$ \\
-\textit{Top 30\% income, 20--40} & $11.1\%$ & $4.6\%$ & $13.0\%$ & $12.5\%$ & $11.6\%$  & $29.8\%$ \\\hline
+\textit{All females, 21--40}  & $23.2\%$ & $10.1\%$ & $18.0\%$ & $22.8\%$ & $24.1\%$ & $34.8\%$ \\
+\textit{All females, 35--40} & $18.0\%$ & $11.3\%$ & $23.0\%$ & $18.9\%$ & $28.1\%$ & $45.5\%$ \\\hline\hline
+\multicolumn{7}{|p{0.9\linewidth}|}{\footnotesize \textbf{Education partition.}}\\\hline
+\textit{High school only, 21--40} & $34.2\%$ & $13.9\%$ & $17.2\%$ & $33.4\%$ & $29.2\%$  & $34.2\%$ \\
+\textit{College graduates, 21--40} & $9.9\%$ & $5.3\%$ & $14.7\%$ & $10.5\%$ & $12.9\%$ & $27.5\%$ \\\hline\hline
+\multicolumn{7}{|p{0.9\linewidth}|}{\footnotesize \textbf{Income partition:} per-hour income for subsample of working women.}\\\hline
+\textit{Low 30\% income, 21--40} & $30.8\%$ & $17.6\%$ & $24.1\%$ & $34.6\%$ & $30.5\%$ &  $42.5\%$ \\
+\textit{Mid 40\% income, 21--40} & $25.5\%$ & $14.2\%$ & $21.5\%$ & $24.1\%$ & $28.3\%$ & $45.8\%$ \\
+\textit{Top 30\% income, 21--40} & $14.1\%$ & $7.7\%$ & $18.1\%$ & $12.4\%$ & $22.9\%$  & $32.8\%$ \\\hline
 \multicolumn{7}{|p{0.9\linewidth}|}{\footnotesize $\Delta T = \text{Year}^{\text{first child}} - \text{Year}^{\text{first marriage}}$. Restriction: $\Delta T \in \{-5,...,10\}$.}\\
@@
 
-\subsection{Differences in Composition}
+Three key observations from the table are:
+\begin{enumerate}
+\item About one-fifth of females have their first kid before or at the year they marry, this share quickly decreases with income and education
+\item Share of divorced people among those who had their first kid before on at the year of their first marriage is larger.
+\item The share of divorced people is about 1.5 higher for low income/education females and more than twice higher for high income/education.
+\item The ACS result is not likely to be driven by picking only people with one marriage, as the pattern is exactly the same if I use SIPP and count for remarriages
+\end{enumerate}
+
+This alone is not enough to draw any causal conclusions: composition of these groups is drastically different.
+For instance, as suggested by the table, people with lower income are more likely to get divorced, so even conditional on income/education group K$\to$M couples may be poorer with each subgroup.
+Additionally, many other factors may drive the differences, including different ages of children (people in K$\to$M group by construction have older babies after any given number of years after marriage), regional and racial differences in attitudes to divorce, age composition, possible presence of step-children in K$\to$M couples and so on.
+I attempt to control for few of these things and argue that they explain these differences only partially, that suggests the presence of causal mechanism that I describe.
+
+\subsubsection{Duration Considerations}
+One obvious concern about comparing two groups is the temporal nature of the data: the variable I use captures fact that divorce happened in the past, and the exact timing may be different between the two groups.
+Due to cumulative nature of the divorce indicator, females who had marriage long time ago are more likely to be divorced even if their divorce chances in any given period are the same.
+Therefore, I it is useful to look at divorced females conditional on number of years passed from marriage, see the left panel of Figure \ref{pic-aftmar} for this.
+\begin{figure}[h!]
+\includegraphics[width=0.45\linewidth]{div_by_aftmar.eps}
+\includegraphics[width=0.45\linewidth]{aftmar_by_mk.eps}
+\caption{Divorced by years after marriage and distribution of years after marriage\label{pic-aftmar}}
+\end{figure}
+
+First, the pattern about different divorce probabilities hold almost uniformly by the ``age of marriage''.
+Group that has kids first both have higher share of divorced and faster growth of it as marriage progresses.
+Combining these two graphs allows to see that M$\to$K group are on average married for longer time at the moment of survey, but overall composition is not dramatically different.
+That would mean that comparing raw shares of divorced actually \emph{underestimates} the difference relative to controlling for number of years after marriage.
+This is confirmed by regression results that I skip, as in the following section I present the version where I control for years after marriage and few other factors.
+
+\subsection{Adjusting for Composition}
+
+This part documents large differences in variables that may affect divorce rates between the two groups.
+After that I try to flexibly capture these factors to see what part of the difference is attributed to the fact of being in K$\to$M group.
+These exercises use ACS data as they have more observations so the tests are less noisy.
+Shortly, after adding many controls, observable difference decreases about twice, but still remains sizable and significant.
+
+\subsection{Differences in Observables}
 This section captures demographic and economic differences of K$\to$M and M$\to$K groups.
@@
 
-\subsection{Statistical Tests and Controlling for Composition}
+Shortly, differences in composition are pretty drastic for income and education dimension (as suggested by different shares of K$\to$M in subgroups described above).
+Most of these things are intuitive, and this suggests importance of flexibly controlling for composition.
+Similar patterns are observed if I compare composition within each subgroup that I discuss above, so partitions should matter.
 
-In this section I formally test differences in shares for different subgroups.
-I also attempt to control for composition using regressions with many fixed effects.
-I show that generally observable mean differences are not likely to be attributed to composition, and introducing various controls still leaves substantial part of the difference.
+\subsubsection{Statistical Tests and Controls}
+
+In this section I formally test differences in shares for different subgroups and attempt to control for composition using regressions with many fixed effects.
 
@@
 \item \textbf{Raw}, corresponding to differences in raw means in the previous part
-\item \textbf{Controls}, where the regression includes dummies for ACS year, the US state, number of children, years after marriage, age of the eldest child, race and education group + 4th degree polynomial in woman's age
-\item \textbf{Controls+income}, where the regression includes all of the above + dummies for income groups (low 30\%, mid 40\%, top 30\%) on a subsample where income equivalence scale is defined.
+\item \textbf{Controls}, where the regression includes dummies for ACS year, the US state, number of children, number of years after marriage, age of the eldest child, race and education group + 4th degree polynomial in woman's age
+\item \textbf{Controls+income}, where the regression includes all of the above + 3rd degree polynomial in log female per-hour income (residualized as described above) for subsample of working women.
 This subsample is smaller, but has arguably better data quality.
@@
 
+%The most important controls are age of the eldest child and number of years after marriage. Basically, being in K$\to$M group is defined by their interaction
+
 See Table \ref{diff-contr} for the results.
+Generally, controlling for composition decreases observable difference, but this decrease is less than half in magnitude: even controlling for many observable factors, many of which may itself be results of unplanned fertility, we cannot assign all observable difference to composition of the groups.
+This also implicitly suggests that the income partition I choose makes sense: conditional on income group, additional controlling for income does not change the observable coefficient much.
 
@@
 \multicolumn{1}{c|}{} & \small divorced & $\Delta$ & (s.e.) & $\Delta$ & (s.e.) & $\Delta$ & (s.e.) \\  \hline
-\textit{All females, 20--50} & 13.0\% & 7.6\% & (0.1\%) & 5.5\% & (0.2\%) & 5.1\% & (0.2\%) \\
-\textit{All females, 35--40} & 16.0\% & 11.3\% & (0.2\%) & 6.5\% & (0.3\%) & 5.9\% & (0.3\%) \\\hline\hline
-\multicolumn{8}{|p{0.9\linewidth}|}{\footnotesize \textbf{Education partition.} Note: ``college graduates'' excludes ``some college'' and ``more than college''.}\\\hline
-\textit{High school only, 20--50} & 16.2\% & 3.2\% & (0.2\%) & 8.0\% & (0.3\%) & 7.6\% & (0.3\%) \\
-\textit{College graduates, 20--50} & 8.3\% & 8.6\% & (0.3\%) & 3.9\% & (0.3\%) & 3.5\% & (0.3\%) \\\hline\hline
-\multicolumn{8}{|p{0.9\linewidth}|}{\footnotesize \textbf{Income partition:} subsample of household heads/spouses living with children and nobody else.}\\\hline
-\textit{Low 30\% income, 20--40} & 20.3\% & 5.1\% & (0.2\%) & 10.3\% & (0.3\%) & \multicolumn{2}{c|}{---}\\
-\textit{Mid 40\% income, 20--40} & 10.1\% & 4.3\% & (0.2\%) & 2.8\% & (0.2\%) & \multicolumn{2}{c|}{---}\\
-\textit{Top 30\% income, 20--40} & 9.0\% & 8.3\% & (0.3\%) & 2.7\% & (0.4\%) & \multicolumn{2}{c|}{---}\\
+\textit{All females, 20--50} & 14.3\% &  7.9\% &   (0.1\%) &    5.7\%  &    (0.2\%) &     4.6\%   &   0.2\% \\
+\textit{All females, 35--40} &  17.3\% &  11.7\%  &  (0.2\%) &   6.7\% &  (0.3\%) &     6.2\%  &   (0.4\%) \\\hline\hline
+\multicolumn{8}{|p{0.9\linewidth}|}{\footnotesize \textbf{Education partition.}}\\\hline
+\textit{High school only, 20--50} & 18.1\% &   3.3\%  &    (0.2\%) &   8.4\% &   (0.3\%)   &  7.8\%   &  (0.5\%) \\
+\textit{College graduates, 20--50} & 8.5\% &   9.4\% &   (0.2\%) &   3.9\% &   (0.3\%) &   3.2\%   &  (0.4\%) \\\hline\hline
+\multicolumn{8}{|p{0.9\linewidth}|}{\footnotesize \textbf{Income partition:} working women only.}\\\hline
+\textit{Low 30\% income, 20--40} &  21.6\% &  6.5\% &    (0.3\%) &    7.8\% &  (0.5\%) &   7.8\%   &  (0.5\%) \\
+\textit{Mid 40\% income, 20--40} &  17.1\% &  7.3\% &   (0.3\%) &    3.6\%   & (0.4\%) &   3.6\% &    (0.4\%) \\
+\textit{Top 30\% income, 20--40} & 11.1\% &  10.4\% &  (0.3\%) &    3.5\% &   (0.4\%) & 3.5\% &  (0.4\%) \\
 \hline
@@
 
-[I pick recently divorced people in ACS, I compare marriage durations for step and non-step children.
-I also check share of divorced in $M\to K$ and $K\to M$ given child's age and given marriage duration.]
+Here I show few additional results.
+First, I present several graphs that suggest relation of divorce chances to child's age.
+Second, I use alternative variable and consider only people who divorced in the year of survey, as for them duration of marriage and age of the first child is (almost) perfectly observed.
+The main point of these calculations is to show that divorces seem to be more concentrated around the moment where children get older.
+
+\subsubsection{Effects by Child's Age}
+First, I show how share of K$\to$M people varies with age of the eldest child.
+I focus on subsample children are 18 or younger, as very few people have children above 18 living in the household.
+This exercise highlights a nuance in the construction of the groups: almost no one in K$\to$M group has children of age less than a year, because that would mean the same year of marriage, first birth and ACS survey.
+Additionally, since the population is trimmed at the age of 40, having kids around age of 18 means fairly low age of the first birth, that is typically associated with being in M$\to$K group.
+Left panel of Figure \ref{pics-raw} clearly depicts both issues, and this suggests the importance of using controls.
+In the specifications above I control both for number of years after marriage and the age of the eldest child, so this should help mitigating these issues.
+
+Right panel of Figure \ref{pics-raw} shows evolution of divorce probabilities for the two groups.
+The increasing pattern is a result of cumulative nature of processes: people with elder children have more periods to get divorced.
+But the difference between two probabilities is illustrative: between ages 5 and 10 size of K$\to$M group changes a little, but difference becomes very pronounced.
+
+\begin{figure}[h!]
+\includegraphics[width=0.45\linewidth]{km_by_age.eps}
+\includegraphics[width=0.45\linewidth]{div_by_age_raw.eps}
+\caption{Share of K$\to$M couples and share of divorced, no controls\label{pics-raw}}
+\end{figure}
+
+I focus on studying difference more carefully then: first, I simply run a regression
+\[D_i = \sum_{a=0}^{18} \beta_a \cdot \I(\text{Age of eldest child}_i = a) +  \sum_{a=0}^{18} \Delta_a \cdot \I(\text{Age of eldest child} = a)\cdot \I(\text{K$\to$M})_i + \varepsilon_i,\]
+and plot $\Delta_a$.
+
+Second, for a subsample of working women I run a regression with controls that were used for Table \ref{diff-contr} above (including polynomials in income),
+\[D_i = \sum_{a=0}^{18} \Delta_a \cdot \I(\text{Age of eldest child} = a)\cdot \I(\text{K$\to$M})_i + X_i\gamma + \varepsilon_i,\]
+note that these controls include dummies for the age of the eldest child.
+
+See Figure \ref{pics-diff} for the results.
+The left panel essentially replicates the differences on the previous graph.
+The right panel is of the main focus: if the set of controls captures the issues of construction of the groups, then the differences associated with age follow the conjecture above: they are small for younger children and start to hike at ages 5--10 (I believe that first few points are mostly caused by composition).
+In general, these graphs are kind of consistent with the idea that divorces are concentrated around the moments where kids get older and childcare costs structure changes as they arguably require less time, though they do not provide a formal test.
+\begin{figure}[h!]
+\includegraphics[width=0.45\linewidth]{diff_raw_by_age.eps}
+\includegraphics[width=0.45\linewidth]{diff_cont_by_age.eps}
+\caption{Differences in divorce probabilities by child's age\label{pics-diff}}
+\end{figure}
+
+\subsubsection{Sample of Recently Divorced}
+In this exercise I utilize an alternative variable available in ACS --- indicator whether a particular person got a divorce in the year of survey.
+It can be interpreted as \textit{marginal} divorce probability as opposed to the previous value being \emph{cumulative}.
+Therefore, for instance, it does not have an upward trend in number of years married, see Figure \ref{pic-now-aftmar}.
+\begin{figure}[h]
+\centering
+\includegraphics[width=0.45\linewidth]{divnow_by_aftmar.eps}
+\caption{Probabilities of getting divorced in the year of survey\label{pic-now-aftmar}}
+\end{figure}
+
+Table \ref{diff-contr-marginal} replicates the results for marginal divorce probabilities.
+The key conclusions and proportions remain very similar, that implicitly suggests validity of the approach used above.
+
+\begin{table}
+\begin{center}
+\begin{tabular}{|l|c||c|c|c|c|c|c|}\cline{2-8}
+\multicolumn{1}{c|}{} & \multicolumn{7}{c|}{ACS: sample of married once}\\\cline{2-8}
+\multicolumn{1}{c|}{} & \small Share of &\multicolumn{2}{c|}{\small Raw difference} & \multicolumn{2}{c|}{\small with controls} & \multicolumn{2}{c|}{\small controls + income} \\ \cline{3-8}
+\multicolumn{1}{c|}{} & \small divorced & $\Delta$ & (s.e.) & $\Delta$ & (s.e.) & $\Delta$ & (s.e.) \\  \hline
+\textit{All females, 20--50} & 1.94\% &  1.37\% &  (0.05\%) &  0.82\% &  (0.08\%) &  0.79\%  &  (0.11\%) \\
+\textit{All females, 35--40} &   1.59\% &  1.04\% &  (0.08\%) &   0.77\% &  (0.11\%)  &  0.74\%   &  (0.15\%) \\\hline\hline
+\multicolumn{8}{|p{0.9\linewidth}|}{\footnotesize \textbf{Education partition.}}\\\hline
+\textit{High school only, 20--50} & 2.31\% &  0.86\% &  (0.08\%) &   0.82\% &  (0.12\%)  &  0.82\%  &   (0.20\%) \\
+\textit{College graduates, 20--50} & 1.19\% &  1.24\% & (0.10\%) &   0.59\% &   (0.13\%)  &  0.50\% &    (0.16\%) \\\hline\hline
+\multicolumn{8}{|p{0.9\linewidth}|}{\footnotesize \textbf{Income partition:} working women only.}\\\hline
+\textit{Low 30\% income, 20--40} &  3.02\% &  1.31\% &  (0.14\%) &   0.90\% &  (0.21\%) &   0.91\%  &   (0.21\%) \\
+\textit{Mid 40\% income, 20--40} &  2.54\% &  1.34\% &  (0.12\%) &   0.54\%  & (0.18\%) &   0.54\% &    (0.18\%) \\
+\textit{Top 30\% income, 20--40} & 1.69\% &  1.69\% &   (0.15\%) &   0.90\%  & (0.20\%) &   0.90\% &    (0.20\%) \\
+\hline
+\end{tabular}
+\caption{Differences in marginal probablities, controlled for composition, ACS\label{diff-contr-marginal}}
+\end{center}
+\end{table}
 
@@
 [I look at earnings, hours and LFP differences] [Grade retention depends on child's age, so need to compare given child's age]
+
+[THIS IS AN OLD VERSION OF TABLE]
+
+\begin{center}
+\begin{tabular}{|l||c|c||c|c|}\hline
+& \multicolumn{2}{|c||}{\footnotesize \textit{All couples}}  & \multicolumn{2}{|c|}{\footnotesize \textit{Top 40\% of income} } \\\hline
+&                       \footnotesize  M$\to$K & \footnotesize  K$\to$M &  \footnotesize  M$\to$K & \footnotesize  K$\to$M \\\hline
+\footnotesize \textit{Female in Labor Force}        & $70.3\%$  & $71.4\%$  & $79.3\%$ & $85.3\%$  \\\hline
+\footnotesize \textit{Earninigs Ratio}    &  $95.2\%$ & $93.9\%$  &  $95.8\%$ &  $96.4\%$ \\\hline
+\footnotesize \textit{Kids' Grade Retention}  & $4.6\%$    & $7.6\%$ &   $3.3\%$   & $5.9\%$ \\\hline
+\end{tabular}
+\end{center}
+
 \section{Model}
@@
 Length of the period is set to be 1 year, the model assumes people start their life with no assets at the age of $21$ and die deterministically at the age of $81$ (so $T = 30$).
-However, it assumes that fertility status can change only till age of 40 ($T_f = 19$) and marriage status till age of 50 ($T_m = 29$).
+However, it assumes that fertility status can change only till age of 40 ($T_f = 20$) and marriage status till age of 45 ($T_m = 25$).
 It also assumes that after the age of 60 people retire and get $b = 0.4$ [as in Sommers, 2014] of their income at age of 60 (so no transitions happen).
-I stop simulations at the age of 50.
+I only simulate the data for age 21--40, therefore these choices are not likely to play a crucial role.
 
-Following standard approach in literature, I fix $\sigma = 1.5$, $R = 1.03$ and $\beta = 1/R$. Also, following estimates of Voena (2014) I set $\rho_c = 0.23$.
+Following standard approach in literature, I fix $\sigma = 1.5$, $R = 1.03$ and $\beta = 1/R$. %Also, following estimates of Voena (2014) I set $\rho_c = 0.23$.
 
-I borrow age profiles from $(a^f_0,a^f_1,a^f_2)$, $(a^m_0,a^m_1,a^m_2)$ and variances of income shocks $\sigma^{z,f}$, $\sigma^{z,m}$ from Low et al 2018 [This will be replaced by my own estimates from SIPP].
+I take age-income profiles  $(a^f_0,a^f_1,a^f_2)$, $(a^m_0,a^m_1,a^m_2)$ to match coefficients in per-hour labor earnings in ACS (see Appendix [...] for details).
+I borrow variances of income shocks $\sigma^{z,f}$, $\sigma^{z,m}$ from Low et al 2018 as they cannot be estimated in cross-sectional data [This could be fixed in later versions as their variances are for subsample of non-college graduates].
 
@@
 \item Unplanned pregnancy probabilities are pinned down by share of $K\to M$ females at ages 25 and 35, as well as share of planned births in total population
-\item Meeting probabilities and $\sigma^{\psi_0}$ are pinned down by shares of never married females people at 25, 35, share of females with one marriage at 40 (relevant to capture chances of re-marrying) and shares of never married at 30 for high and low income groups.
+\item Meeting probabilities and $\sigma^{\psi_0}$ are pinned down by shares of never married females people at 25, 35, 40 and share of females with one marriage at 40 (relevant to capture chances of re-marrying).
+\item  Preferences for kids are captured by share of childless females by age of 25, 30, 35 (identifies general scale $\phi_0$), share of childless females for low and high income at 30 $\phi_1$.
+Luxury good parameter $\phi_2$ is pinned down by ratio of amounts of money spent on kids between 75 and 25 income quantiles.
 \item Childcare production function is recovered by ratio of time spent with kids by couple's income (pins down elasticity of substitution), average share of money spent on kids (captures ``average'' $\mu$) and ratio of time spent by child's age (captures difference between $\mu_a$ and $\mu_y$)
-\item  Preferences for kids are captured by share of childless females by age of 25, 30, 35 (general scale), share of childless females for low and high income at 30 $\phi_1$.
-Luxury good parameter $\phi_2$ is pinned down by ratio of amounts of money spent on kids between 75 and 25 income quantiles.
-\item Variance of innovation to marriage surplus is mainly pinned down by share of divorced people by age (25, 35, 40) and income groups (at 30).
+\item Variance of innovation to marriage surplus is mainly pinned down by share of divorced people by age (25, 35, 40).
 \item Transition out of single mothers is captured by share of never married single mothers at 25 (picks $S$) and share of divorced women with kids at 30 (picks $p^{\text{out}}$).
+\item Returns to scale in marriage $\rho_c$ are pinned down by comparing divorce probabilities between high and low income females.
 \end{enumerate}
-On top of it, I use shares of divorced people for $K\to M$ and $M \to K$ at 40 as additional targets that are affected by mixture of parameters described above.
+On top of it, I use shares of divorced people for $K\to M$ and $M \to K$ at 40 and their ratio by income groups as additional targets that are affected by mixture of parameters described above.
 
-I additionally impose the following constraints: share of rejected marital offers is within $[0.05,0.95]$ and average share of time spent on kids is below $0.8$.
-[Both of these constraints are non-binding in the solution I found, although turning them off occasionally may lead to implausible solutions.]
+For income considerations I trim females/couples with $l$ larger than its 77\% percentile (corresponding to $77\%$ of ACS females of 21--40 who work at least 5 hours, imputing $l = 0$ for those who does not have kids) in simulated data, and treat their log per-hour income as missing.
+I then perform similar detrending and divide individuals by group using the same regression procedure.
+I prefer detrending to simply picking $z$ as choices of $l$ are not independent on age.
 
 \section{Results}
-Table ... shows targeted and
+Table \ref{targets-main} shows the targets in the model and in the data.
+[Marriage dimension is somewhat off, this is not crucial as the study mainly contains married couples, though I have few ideas how to fix it in finite time.]
+\begin{table}
+\begin{center}
+\begin{tabular}{|l|c|c|}\hline
+Target & Model & Data \\\hline
+\multicolumn{3}{|c|}{Marriage Dimension} \\\hline
+Never married by 25  & $32.8\%$ & $68.3\%$  \\
+Never married by 30 & $12.6\%$ & $25.6\%$  \\
+Never married by 40 & $6.3\%$ & $18.8\%$  \\\hline
+One marriage by 40 & $66.3\%$ & $64.7\%$ \\\hline
+\multicolumn{3}{|c|}{Fertility Dimension} \\\hline
+No kids at 25 & $59.0\%$ & $67.7\%$ \\
+No kids at 30 (low $30\%$ income) & $46.7\%$  &   $44.0\%$ \\
+No kids at 30 (top $30\%$ income) & $52.1\%$  &   $57.8\%$ \\
+No kids at 35  & $32.9\%$  &   $28.1\%$ \\
+No kids at 40 & $17.6\%$ & $25.6\%$ \\\hline
+\multicolumn{3}{|c|}{Marriage/Fertility Interactions} \\\hline
+Never married with kids at 30 & $7.0\%$  & $12.3\%$ \\
+Divorced with kids at 30 & $5.7\%$ & $3.8\%$ \\\hline
+\multicolumn{3}{|c|}{Divorce Dimension (Divorced = Divorced and not remarried)} \\\hline
+Divorced by 25 & 6.6\%   &  10.7\% \\
+Divorced by 30 (low $30\%$ income)  & 14.5\% &     20.8\% \\
+Divorced by 30 (top $30\%$ income)  & 14.0\% &     8.1\% \\
+Divorced by 35  & 15.8\% &    15.9\% \\
+Divorced by 40  & 15.5\% &    19.1\% \\\hline
+\multicolumn{3}{|c|}{K$\to$M --- M$\to$K partition} \\\hline
+Divorced by 40 if K$\to$M & 22.1\% &   23.5\% \\
+Divorced by 40 if M$\to$K & 13.8\% &      13.0\% \\
+Divorced by 40 if K$\to$M and low 30\% & 34.0\% & 28.5\% \\
+Divorced by 40 if K$\to$M and top 30\% & 28.9\% & 23.3\% \\
+Divorced by 40 if M$\to$K and low 30\% & 34.0\% & 28.5\% \\
+Divorced by 40 if M$\to$K and top 30\% & 19.3\% & 11.3\% \\
+Divorced by 40 if M$\to$K and low 30\% & 28.7\% & 18.1\% \\\hline
+\multicolumn{3}{|c|}{Childcare} \\\hline
+Money for childcare / labor income & 25.7\% & 40.0\% \\
+Time spent 0--5 / Time spent 6--12 & 1.92 & 2 \\
+Time spent top 30\% / Time spent low 30\% & 0.92 & 1.19 \\
+Money spent top 30\% / Money spent low 30\% & 2.09 & 2.67 \\\hline
+
+\end{tabular}
+\caption{Targets in the data and in the model\label{targets-main}}
+\end{center}
+\end{table}
 
 \subsection{Counterfactual Simulations}
+[TBD]
 \begin{itemize}
@@
 \end{itemize}
+
+\subsubsection{Altering Fertility Shocks}
+I consider two counterfactual scenarios: when unexpected fertility never happens at the moment people meet and when it never happens at all (but people still can make kids by choice).
+I compare key targets in Table \ref{cf-shocks}.
+In general, it looks like contribution of fertility shocks to the divorce rates is quite substantial, and major part of it is driven by fertility shocks before the marriage happens.
+Especially reducing fertility shocks affects divorces in early stages.
+\begin{table}
+\begin{center}
+\begin{tabular}{|l|c|c|c|}\hline
+Target & Baseline & No shocks at meeting & No shocks at all \\\hline
+\multicolumn{4}{|c|}{Marriage Dimension} \\\hline
+Never married by 25  & $32.8\%$ & $30.6\%$ & $31.1\%$  \\
+Never married by 30 & $12.6\%$ & $10.8\%$ & $11.2\%$  \\
+Never married by 40 & $6.3\%$ & $5.7\%$ & $6.1\%$  \\\hline
+\multicolumn{4}{|c|}{Fertility Dimension} \\\hline
+No kids at 25 & $59.0\%$ & $66.3\%$ & $68.3\%$ \\
+No kids at 30 (low $30\%$ income) & $46.7\%$  &   $53.5\%$ & $54.8\%$ \\
+No kids at 30 (top $30\%$ income) & $52.1\%$  &   $61.9\%$ & $62.7\%$ \\
+No kids at 35  & $32.9\%$  &   $36.2\%$ & $36.6\%$ \\\hline
+\multicolumn{4}{|c|}{Divorce Dimension} \\\hline
+Divorced by 25 & 6.6\%   &  2.6\% & 1.4\% \\
+Divorced by 30 (low $30\%$ income)  & 14.5\% &     11.0\% & 8.6\% \\
+Divorced by 30 (top $30\%$ income)  & 14.0\% &     9.0\% & 7.5\% \\
+Divorced by 35  & 15.8\% &    13.6\% & 11.9\% \\
+Divorced by 40  & 15.5\% &    14.3\% & 13.0\% \\\hline
+\end{tabular}
+\caption{Contribution of fertility shocks \label{cf-shocks}}
+\end{center}
+\end{table}
 
@@
 
+[TBD]
+
 \section{Additional Robustness Checks}

### 2019-05-22 main/draft.tex
Introduction version 1

@@
 But even for educated and older women this share is substantial [...].
-An increase in non-marital cohabitation is also a documented trend, despite of this, persistent non-marital cohabitation is still not the most common thing: according to SIPP 2014, only [...] of women 21--40 cohabitate for longer than 5 years without being married.
-A substantial part of children by unmarried parents are born in such cohabitating unions (SIPP [...]).
+An increase in non-marital cohabitation is also a documented trend, despite of this, persistent non-marital cohabitation is still not the most common thing: according to National Survey of Family And Growth, median duration of cohabitation is 22 months, and median duration of cohabitation if it ends by marriage is 21.\footnote{Source: Copen et al, \cite{copen}}.
+Unmarried unions with children are still pretty rare: according to Payne, 2013\nocite{payne}, only about 3\% of children live in unions with unmarried cohabitating parents (as opposed to 21\% with single mothers).
 
-These numbers have important economic implications: children born by unmarried women, as documented by, for instance [...], display significantly worse future outcomes in terms of education and income, and this pattern survives controlling in many dimensions.
+These numbers have important economic implications: children born by unmarried women, as documented by, for instance Kearney, Levine, 2017, display significantly worse future outcomes in terms of education and income, and this pattern survives controlling in many dimensions.
 Although it is well understood that children of never married single mothers or children of blended families (when a woman marries someone else than child's father) have generally worse outcomes, even when parents marry each other after they made a baby their marriage is statistically less stable, as I argue in this paper, and the larger chances of marriage dissolution do affect future prospects of the children.
 
-This pairs well with common perception that some couples marry at or shortly after the moment they make a baby: long-lasting marital cohabitations with children are even less common (among ...).
+This pairs well with common perception that some couples marry at or shortly after the moment they make a baby.
 If cohabitation is viewed as uncommitted relationship, unexpected fertility may serve as a trigger to obtain more commitment.
@@
 The whole role of marriage changes as it is more socially acceptable to cohabitate, though it is much more conventional to do it before having kids.
-As highlighted by, for instance, [...], despite of general decrease of role of marriage in the society, securing investment to children is one of the most important forces that makes people marry each other.
+As highlighted by, for instance, Lafortune, 2018, despite of general decrease of role of marriage in the society, securing investment to children is one of the most important forces that makes people marry each other.
 
@@
 I estimate the model parameters to match observable patterns in the data and show what it implies about the role of unequal exposure to childcare costs in observable marriage selection and divorce patterns.
-Shortly, the model suggests if people could perfectly control timing of their births, general share of divorced people would decrease by ..., especially in high-income groups.
+Shortly, the model suggests if people could perfectly control timing of their births, general share of divorced people would decrease by [...], especially in high-income groups.
 
 Section 2 of this paper discuss existing results and general trends that are already described by the literature.
-Section [...] discuss how I construct the dataset and timing variables, and Section 3 establishes main observations about differences in groups depending on their marriage timing, together with a number of additional considerations useful to understand the content of these differences.
-After that I explain theoretical model in Section 4, the process of setting its parameters in Section 5.
-Section 6 provides estimation results and counterfactual simulations implied by these results.
-Section 7 provides more general discussion and concludes.
+Section 3 discuss how I construct the dataset and timing variables, and Section 5 establishes main observations about differences in groups depending on their marriage timing, together with a number of additional considerations useful to understand the content of these differences.
+After that I explain theoretical model in Section 5 and the process of setting its parameters in Section 6.
+Section 7 describes estimation results and counterfactual simulations implied by these results.
+Section 8 provides more general discussion and concludes.
 
-%Understanding how people make marriage decisions and more generally how household structure is formed is crucial for analyzing labor market outcomes, assets accumulation, various dimensions of inequality and children development outcome.
-Most of existing studies, even ones that take household structure seriously, typically abstract from the fact that household structure is itself a choice variable.
-However, existing trends in marriage and divorce rates, increasing share of kids with unmarried parents and in blended families, and, especially, how different these trends are by income and racial groups --- this all suggests economic importance of household issues.
-
-%In particular, economists know pretty little about how people decide on marriage and divorce, despite of the fact that these issues have lots of economic content.
-Stylized models describing marriage (and, in smaller extent, divorce) emerged in family economics, however most of them describe qualitative issues and do not deliver quantitative predictions that can be applied for analyzing policy or just understanding how different factors contribute to the recent changes.
-...]
-
-\section{Existing Results}
+\section{Related Literature}
 
@@
 
-The main conceptual framework that I utilize are models of household's dynamic bargaining and limited commitment, extensively reviewed by
+The main conceptual framework that I utilize are models of household's dynamic bargaining and limited commitment, extensively reviewed by Chiaporri, Mazzocco 2015.\nocite{chiappori-review} These models were used to answer various marriage and divorce related questions.
+Theoretical part of my model is mostly based on Voena, 2015\nocite{voena-1}, that uses dynamic bargaining to consider impact of changes in divorce laws, and Low et al, 2018\nocite{low-1}, focusing on marriage and divorce selection related to welfare utilitzation.
+There are few more recent papers utilitizing similar approaches, for instance, Shephard, 2019\nocite{shephard} studying gender and age pay gaps through equilibrium on marriage markets.
+
+To model fertility choices and shocks, I rely on recent lifecycle models.
+I make many choices similar to Sommer, 2015\nocite{sommer} who studies how changes in reproductive horizon affect couple's fertility decisions with in the environment with income shocks, paying detailed attention to time and money structure of childcare.
+More detailed treatment of unintended pregnancies is done by Ejrnæs, Jørgensen, 2018\nocite{ejrnaes}, they extend the model to allow for non-random abortion decisions though do not model childcare costs in details.
+
+Commitment in marriage and its interaction with fertility decisions attract recent attention as possible drivers of trends in marriage and divorce.
+Work by Lafortune, Low 2018\nocite{lafortune-1} uses ACS data and housing prices, as well as marriage market model, to argue that married people buy houses as a device to commit not to divorce, that allows them to secure investment in their future children.
+This is parallel and complimentary with my results, that mainly focus on consequences of not having enough commitment and how they relate to fertility timing.
+
+Impact of non-marital childbearing and parents divorce are well-documented empirically from children prospective.
+For example, Kearney, Levine 2017 use PSID to document how being born out of marriage statistically child's outcome in school years and in early adult life, and the heterogeneity of this by parent's education and age groups.
+Shortly, they find presence of ``marriage premium'' for children at each education level, though for less educated and younger mothers benefits of having married parents are more pronounced in early life, and for elder and more educated mothers --- in adult life.
+A classic study by Akerlof, 1996\nocite{akerlof} highlight the mechanism why unplanned pregnancies may not be aborted in couples so they are still present even with large abortion access.
+Finally, Ginther, Pollak, 2004 show that outcomes of children in blended families (i.e.
+having step-parents) are similar to outcomes of children of single parents.
+
+This paper fills an important gap in the literature considering impact of out-of-marriage births for women.
+As fertility, marriage and divorce involve many individual decisions affected by many unobservable factors, it is not possible to ask casual questions by using only the data, and existing empirical studies only provide descriptive evidence.
+This study contributes to the growing literature of quantitative models that both use the tools of family economics to endogenize timing of fertility, marriage, divorce and their interactions, and, on the other hand, provide numbers and patterns consistent with microdata, so they can be used to answer policy questions.
+
+As noted by Ginther and Pollak, causal questions in the context of marriage and divorce should be taken with caution: to answer, for instance, how parents' divorce affect child's future we have to be precise what counterfactual we compare it with (either parents never divorced, parents stayed together for few more years, parents never married).
+The same issues concern impact of unplanned pregnancies in cohabitating couples, therefore, instead of trying to ask causal quesitons I the main goal of using the model is to deliver counterfactual outcomes, that still facilitate understanding of some causality.
 
@@
 
-\newpage
 \section{Robustness Checks}
 \section{Conclusions}
+\clearpage
+\section*{Literature}
+\bibliography{bib}
 
+\newpage
 \appendix

### 2019-05-31 main/draft.tex
Received fixes from Martí

@@
 \begin{abstract}
-Unplanned pregnancies often trigger marriage for unmarried cohabitating couples.
-I show that on average marriages in which a child appeared before or at the year of marriage are characterized with higher divorce probabilities and other outcomes suggesting lower marital quality.
+Unplanned pregnancies often trigger marriage for unmarried couples.
+I show that, on average, marriages in which a child appeared before or at the year of marriage are characterized with higher divorce probabilities and other outcomes suggesting lower marital quality.
 This association is robust to different ways of controlling for composition, and it is more pronounced for higher income and more educated couples.
@@
 \subsection{Raw Comparisons}
-First, I compare divorce probabilities in cross-sections of people between $K\to M$ and $M\to K$ groups. See Table \ref{diff-raw} for this.
+The main result I discuss is summarized by Figure \ref{shrs}. Table \ref{diff-raw} presents these results in numbers.
 
-\begin{table}[h!]
+\begin{figure}[h!]
+\includegraphics[width=0.9\linewidth]{shrs.eps}
+\caption{Share of K$\to$M couples and share of divorced\label{shrs}}
+\end{figure}
+
+%First, I compare divorce probabilities in cross-sections of people between $K\to M$ and $M\to K$ groups. See Table \ref{diff-raw} for this.
+
+\begin{table}[p]
 \begin{center}
@@
 Many factors could compute to the observable differences, here I quantify them and this justifies use of controls in the next part.
-I present these differences for ACS, SIPP tables are available in [Appendix ...].
+I present these differences for ACS, SIPP data are not very different except smaller sample size.
 For ACS, since M$\to$K and K$\to$M are defined only for subsample of people who were married once and have children present, I also compare this subsample with more general female population.
@@
 \item Fundamental parameters of income fluctuation model $(\beta,\sigma,R)$
-\item Intra-household returns to scale $\rho_c$
-\item Preferences for child's quality: $(\phi_0,\phi_1,\phi_2)$
-\item Parameters for child's quality production function: CES power $\theta_k$ and labor shares by age $(\mu_y,\mu_s)$.
+\item Preferences for child's quality: $(\phi_0,\phi_1,\phi_2,\phi_3)$
+\item Parameters for child's quality production function: CES power $\theta_k$ and labor shares by age $(\mu_y,\mu_s)$
+\item Unavoidable childcare costs $\underline{q}$
 \item Transition probabilities:
@@
 \item Probability of meeting a partner: $p^{\text{meet}}_t:$ piecewise linear function between $(p^{\text{meet}}_{21},p^{\text{meet}}_{27},p^{\text{meet}}_{34},p^{\text{meet}}_{40})$ (4 parameters)
-\item Probability of re-entering marriage market and corresponding utility ``compensation'' $(p^{\text{out}},S)$. (2 parameters)
+\item Probability of re-entering marriage market and corresponding utility ``compensation'' $(p^{\text{out}},S)$. I set $S = 0$ in current calibration (so 1 parameter).
 \item Probability that kids grow up (transition from $y$ to $s$): $p_s$ (1 parameter)
@@
 
-Total number of parameters here is 31. Few additional choices involve time horizons, bounds for random walk variables and bounds for assets and $\theta$.
+Few additional choices involve time horizons, bounds for random walk variables and bounds for assets and $\theta$.
 
 \subsection{Externally Set Parameters}
-Length of the period is set to be 1 year, the model assumes people start their life with no assets at the age of $21$ and die deterministically at the age of $81$ (so $T = 30$).
-However, it assumes that fertility status can change only till age of 40 ($T_f = 20$) and marriage status till age of 45 ($T_m = 25$).
-It also assumes that after the age of 60 people retire and get $b = 0.4$ [as in Sommers, 2014] of their income at age of 60 (so no transitions happen).
+Length of the period is set to be 1 year, the model assumes people start their life with no assets at the age of $21$ and die deterministically at the age of $81$ (so $T = 60$).
+However, it assumes that fertility status can change only till age of 37 ($T_f = 17$) and marriage status till age of 45 ($T_m = 25$).
+It also assumes that after the age of 60 people retire and get $b = 0.4$ as in Sommer, 2014 of their income at age of 60 (so no transitions happen after $T = 40$).
 I only simulate the data for age 21--40, therefore these choices are not likely to play a crucial role.
 
-Following standard approach in literature, I fix $\sigma = 1.5$, $R = 1.03$ and $\beta = 1/R$. %Also, following estimates of Voena (2014) I set $\rho_c = 0.23$.
+Following standard approach in the macro literature, I fix $\sigma = 2$, $R = 1.03$ and $\beta = 1/R$. Also, following estimates of Voena, 2014 I set $\rho_c = 0.23$.
 
 I take age-income profiles  $(a^f_0,a^f_1,a^f_2)$, $(a^m_0,a^m_1,a^m_2)$ to match coefficients in per-hour labor earnings in ACS (see Appendix [...] for details).
-I borrow variances of income shocks $\sigma^{z,f}$, $\sigma^{z,m}$ from Low et al 2018 as they cannot be estimated in cross-sectional data [This could be fixed in later versions as their variances are for subsample of non-college graduates].
+I borrow variances of income shocks $\sigma^{z,f}$, $\sigma^{z,m}$ from Low et al 2018 as they cannot be estimated in cross-sectional data, and although they were estimated for subsample of non college graduates in SIPP, in the model they produce IQR at the age of 30 that is very similar to the full sample of ACS data.
 
@@
 In the part with assets, I assume savings can be between 0 and 20 times average labor income in the beginning of life.
-This bar in pretty low, but it is enough to create reasonable assets variation at the moment most of marriage decisions are made: say, median assets-to-income ratio at 30 is around 1 [see ...].
+This bar in pretty low, but it is enough to create reasonable assets variation at the moment most of marriage decisions are made: say, median assets-to-income ratio at 30 is around 1.
 
@@
 \item Meeting probabilities and $\sigma^{\psi_0}$ are pinned down by shares of never married females people at 25, 35, 40 and share of females with one marriage at 40 (relevant to capture chances of re-marrying).
-\item  Preferences for kids are captured by share of childless females by age of 25, 30, 35 (identifies general scale $\phi_0$), share of childless females for low and high income at 30 $\phi_1$.
-Luxury good parameter $\phi_2$ is pinned down by ratio of amounts of money spent on kids between 75 and 25 income quantiles.
+\item  Preferences for kids are captured by share of childless females by age of 25, 30, 35 (identifies general scale $\phi_0$ and $\phi_3$), share of childless females for low and high income at 30 (captures $\phi_0$ and $\phi_1$).
+Luxury good parameter $\phi_2$ is mainly pinned down by ratio of amounts of money spent on kids between 75 and 25 income quantiles.
 \item Childcare production function is recovered by ratio of time spent with kids by couple's income (pins down elasticity of substitution), average share of money spent on kids (captures ``average'' $\mu$) and ratio of time spent by child's age (captures difference between $\mu_a$ and $\mu_y$)
-\item Variance of innovation to marriage surplus is mainly pinned down by share of divorced people by age (25, 35, 40).
-\item Transition out of single mothers is captured by share of never married single mothers at 25 (picks $S$) and share of divorced women with kids at 30 (picks $p^{\text{out}}$).
-\item Returns to scale in marriage $\rho_c$ are pinned down by comparing divorce probabilities between high and low income females.
+\item Variance of innovation to marriage surplus is mainly pinned down by share of divorced people by age (25, 35, 40) and divorce rates by income groups.
+\item Transition out of single mothers is captured by share of never married single mothers at 25 and share of divorced women with kids at 30 (picks $p^{\text{out}}$).
 \end{enumerate}
-On top of it, I use shares of divorced people for $K\to M$ and $M \to K$ at 40 and their ratio by income groups as additional targets that are affected by mixture of parameters described above.
+On top of it, I use shares of divorced people for $K\to M$ and $M \to K$ at 40 per se and by income groups as additional targets that are affected by mixture of parameters described above.
 
@@
 \section{Results}
-Table \ref{targets-main} shows the targets in the model and in the data.
-[Marriage dimension is somewhat off, this is not crucial as the study mainly contains married couples, though I have few ideas how to fix it in finite time.]
+Table \ref{targets-main} shows the targets in the model and in the data. [Income partition is somewhat off, but things get better over time.]
+
 \begin{table}
@@
 \multicolumn{3}{|c|}{Marriage Dimension} \\\hline
-Never married by 25  & $32.8\%$ & $68.3\%$  \\
-Never married by 30 & $12.6\%$ & $25.6\%$  \\
-Never married by 40 & $6.3\%$ & $18.8\%$  \\\hline
-One marriage by 40 & $66.3\%$ & $64.7\%$ \\\hline
+Never married by 25  & $67.8\%$ & $68.3\%$ \\
+Never married by 30 & $25.2\%$ & $25.6\%$  \\
+Never married by 40 & $20.6\%$ & $18.8\%$  \\\hline
+One marriage by 40  & $68.9\%$ & $64.7\%$ \\\hline
 \multicolumn{3}{|c|}{Fertility Dimension} \\\hline
-No kids at 25 & $59.0\%$ & $67.7\%$ \\
-No kids at 30 (low $30\%$ income) & $46.7\%$  &   $44.0\%$ \\
-No kids at 30 (top $30\%$ income) & $52.1\%$  &   $57.8\%$ \\
-No kids at 35  & $32.9\%$  &   $28.1\%$ \\
-No kids at 40 & $17.6\%$ & $25.6\%$ \\\hline
+No kids at 25 & $71.7\%$ & $67.7\%$ \\
+No kids at 30 (low $30\%$ income) & $58.6\%$  &   $44.0\%$ \\
+No kids at 30 (top $30\%$ income) & $56.6\%$  &   $57.8\%$ \\
+No kids at 35  & $22.4\%$  &   $28.1\%$ \\
+No kids at 40 & $19.4\%$ & $25.6\%$ \\\hline
 \multicolumn{3}{|c|}{Marriage/Fertility Interactions} \\\hline
-Never married with kids at 30 & $7.0\%$  & $12.3\%$ \\
-Divorced with kids at 30 & $5.7\%$ & $3.8\%$ \\\hline
-\multicolumn{3}{|c|}{Divorce Dimension (Divorced = Divorced and not remarried)} \\\hline
-Divorced by 25 & 6.6\%   &  10.7\% \\
-Divorced by 30 (low $30\%$ income)  & 14.5\% &     20.8\% \\
-Divorced by 30 (top $30\%$ income)  & 14.0\% &     8.1\% \\
-Divorced by 35  & 15.8\% &    15.9\% \\
-Divorced by 40  & 15.5\% &    19.1\% \\\hline
+Never married with kids at 30 & $12.4\%$  & $12.3\%$ \\
+Divorced with kids at 30 & $3.2\%$ & $3.8\%$ \\\hline
+\multicolumn{3}{|c|}{Divorce Dimension (Divorced $=$ Divorced and not remarried)} \\\hline
+Divorced by 25 & $4.9\%$   & $10.7\%$ \\
+Divorced by 30 (low $30\%$ income)  & 15.6\% &     20.8\% \\
+Divorced by 30 (top $30\%$ income)  & 9.9\% &     8.1\% \\
+Divorced by 35  & 8.4\% &    15.9\% \\
+Divorced by 40  & 10.6\% &    19.1\% \\\hline
 \multicolumn{3}{|c|}{K$\to$M --- M$\to$K partition} \\\hline
-Divorced by 40 if K$\to$M & 22.1\% &   23.5\% \\
-Divorced by 40 if M$\to$K & 13.8\% &      13.0\% \\
+Divorced by 40 if K$\to$M & 20.2\% &   23.5\% \\
+Divorced by 40 if M$\to$K & 8.7\% &      13.0\% \\
 Divorced by 40 if K$\to$M and low 30\% & 34.0\% & 28.5\% \\
-Divorced by 40 if K$\to$M and top 30\% & 28.9\% & 23.3\% \\
-Divorced by 40 if M$\to$K and low 30\% & 34.0\% & 28.5\% \\
-Divorced by 40 if M$\to$K and top 30\% & 19.3\% & 11.3\% \\
-Divorced by 40 if M$\to$K and low 30\% & 28.7\% & 18.1\% \\\hline
+Divorced by 40 if K$\to$M and top 30\% & 26.6\% & 23.3\% \\
+Divorced by 40 if M$\to$K and low 30\% & 15.3\% & 18.0\% \\
+Divorced by 40 if M$\to$K and top 30\% & 9.3\% & 11.3\% \\\hline
 \multicolumn{3}{|c|}{Childcare} \\\hline
-Money for childcare / labor income & 25.7\% & 40.0\% \\
-Time spent 0--5 / Time spent 6--12 & 1.92 & 2 \\
-Time spent top 30\% / Time spent low 30\% & 0.92 & 1.19 \\
-Money spent top 30\% / Money spent low 30\% & 2.09 & 2.67 \\\hline
-
+Money for childcare / labor income & 40.1\% & 40.0\% \\
+Time spent 0--5 / Time spent 6--12 & 1.91 & 2 \\
+Time spent top 30\% / Time spent low 30\% & 0.80 & 1.19 \\
+Money spent top 30\% / Money spent low 30\% & 1.42 & 2.67 \\\hline
 \end{tabular}
@@
 \subsection{Counterfactual Simulations}
-[TBD]
-\begin{itemize}
-\item Turn off fertility completely
-\item Turn off bargaining
-\item No pressure of being single mother
-\item Single mother/single father randomly
-\end{itemize}
 
-\subsubsection{Altering Fertility Shocks}
+This part does two things.
+First, to understand the anatomy of result, I simulate individual impact of fertility shocks, comparing several counterfactual scenarios for simulated group of individuals.
+Second, I look at how changes in parameters driving fertility shocks and bargaining affect the observable (aggregate) moments.
+
+\subsubsection{Individual Impact}
+
+This part looks at counterfactual impact of the fertility shocks.
+To do this, I generate two sequences of shocks from the estimated distribution, $\mathcal{S}_{m,p}$ and $\mathcal{S}_{m,np}$.
+These sequences are identical in terms of all shocks and in both of them all females meet the first partner at $\bar{t}$, but in $\mathcal{S}_{m,p}$ they get an unexpected fertility shock at $\bar{t}$ and in $\mathcal{S}_{m,np}$ they do not.
+After $\bar{t}$ the sequences are identical (including chances of meeting other partners if the first marriage does not happen/breaks up).
+For an auxiliary considerations I also generate sequence of shocks $\mathcal{S}_{nm}$, where at $\bar{t}$ (and before) the agents did not meet anyone.
+
+First, one can look at average impact of the shock, presented at Figure \ref{individual-av}.
+I take $\bar{t} = 25$.
+On the top panel, I present probabilities of being married for those who agreed to marry to the partner they met at 25 (so at 25 this is one).
+As we can see, In the middle, there are unconditional shares of married for two groups.
+In the bottom I present average utility surplus (love shock) of married couples.
+
+There are two important observations.
+First, conditional on having satisfactory partner women who did receive fertility shocks have around higher chances of not being married (to anyone) in 10 years.
+Second, threshold for satisfactory partners falls upon receiving fertility shock: more people marry if the shock happens and average surplus of married couples is lower.
+This indicates the presence of females who married \emph{because} the pregnancy shock happened (meaning that they would reject the partner if they were not pregnant).
+Then we can notice that after a number of years surviving couples become very identical in the amount of ``love shock'', meaning that couples with low $\psi$ break up.
+Third, in the model the largest impact of fertility shock is hurting future marriage prospective for those who stayed a single mother.
+This is curious although pretty speculative, as how the lifecycle of single mothers is modeled in pretty stylized way.
+
+In the second part, I decompose the observable impact on three groups: those who would marry the partners they met with and without fertility shocks (always-marry), those who marry only if fertility shocks happen (compliers) and those who never marry their matches (never-marry).
+In principle, there could exist a group of those who marry only without the shock (defiers), but it is empty in my simulations.
+Figure \ref{ind-decomp} shows this decomposition.
+Two dashed lines correspond to the top panel of Figure \ref{individual-av}.
+The average impact of shock (dashed line with crosses) is a weighted average of two groups (presented by two solid lines).
+We can see that mainly the divorce probability is driven by small (5.5\%) group of compliers who have dramatically lower probabilities to stay married.
+Effect on fertility shocks on those who would've married anyway still persists, but is around 2-3\% (this is, however, substantial if we compare it to average probability of not-staying-married without shock that is slightly above 10\%.
+
+In terms of characteristics, the group of compliers have about 14\% higher median savings and 1.7\% higher median labor income relative to the general population.
+What makes them compliers is mainly the value of love shock $\psi$: for this group, it is lower than the shock for general population of new couples by about 2/3 of initial standard deviation (i.e.
+$\sigma(\psi_0)$).
+This sort of highlights that  the compliers are generally more picky group (that is roughly consistent with evidence on education/income partition) who got a pregnancy shock together with unlucky draw of $\psi$.
+This is however partially hardcoded in the model, as I target different divorce probabilities for higher-income women.
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=0.85\linewidth]{individual-av.eps}
+\caption{Average impact of fertility shock at $t=25$\label{individual-av}}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=0.85\linewidth]{ind-decomp.eps}
+\vspace{-4em}
+\caption{Decomposition of the average impact\label{ind-decomp}}
+\end{figure}
+
+\subsubsection{Aggregate Impact}
 I consider two counterfactual scenarios: when unexpected fertility never happens at the moment people meet and when it never happens at all (but people still can make kids by choice).
 I compare key targets in Table \ref{cf-shocks}.
-In general, it looks like contribution of fertility shocks to the divorce rates is quite substantial, and major part of it is driven by fertility shocks before the marriage happens.
-Especially reducing fertility shocks affects divorces in early stages.
+First, absence of fertility shocks dramatically reduces early fertility (in the model every couple before 37 can have a baby by choice).
+Then, fertility shocks at meeting are large driver of observable divorce rates, especially for higher-income couples.
+Finally, fertility shocks in couples (i.e.
+mistimed births) are also large contributor to the divorce rate,disabling them eliminates early divorces almost completely (though is not so important for later divorces relative to shocks at births).
 \begin{table}
@@
 \multicolumn{4}{|c|}{Marriage Dimension} \\\hline
-Never married by 25  & $32.8\%$ & $30.6\%$ & $31.1\%$  \\
-Never married by 30 & $12.6\%$ & $10.8\%$ & $11.2\%$  \\
-Never married by 40 & $6.3\%$ & $5.7\%$ & $6.1\%$  \\\hline
+Never married by 25  & $67.8\%$ & $68.2\%$ & $66.4\%$  \\
+Never married by 30 & $25.2\%$ & $21.8\%$ & $21.4\%$  \\
+Never married by 40 & $20.6\%$ & $18.1\%$ & $17.5\%$  \\\hline
 \multicolumn{4}{|c|}{Fertility Dimension} \\\hline
-No kids at 25 & $59.0\%$ & $66.3\%$ & $68.3\%$ \\
-No kids at 30 (low $30\%$ income) & $46.7\%$  &   $53.5\%$ & $54.8\%$ \\
-No kids at 30 (top $30\%$ income) & $52.1\%$  &   $61.9\%$ & $62.7\%$ \\
-No kids at 35  & $32.9\%$  &   $36.2\%$ & $36.6\%$ \\\hline
+No kids at 25 & $59.0\%$ & $85.8\%$ & $99.8\%$ \\
+No kids at 30 (low $30\%$ income) & $76.0\%$  &   $53.5\%$ & $92.8\%$ \\
+No kids at 30 (top $30\%$ income) & $70.5\%$  &   $61.9\%$ & $89.5\%$ \\
+No kids at 35  & $32.9\%$  &   $26.8\%$ & $29.0\%$ \\
+No kids at 40 & $19.4\%$   &  $23.2\%$  & $25.1\%$ \\\hline
 \multicolumn{4}{|c|}{Divorce Dimension} \\\hline
-Divorced by 25 & 6.6\%   &  2.6\% & 1.4\% \\
-Divorced by 30 (low $30\%$ income)  & 14.5\% &     11.0\% & 8.6\% \\
-Divorced by 30 (top $30\%$ income)  & 14.0\% &     9.0\% & 7.5\% \\
-Divorced by 35  & 15.8\% &    13.6\% & 11.9\% \\
-Divorced by 40  & 15.5\% &    14.3\% & 13.0\% \\\hline
+Divorced by 25 & 6.6\%   &  2.9\% & 0.0\% \\
+Divorced by 30 (low $30\%$ income)  & 14.5\% &     11.4\% & 1.4\% \\
+Divorced by 30 (top $30\%$ income)  & 14.0\% &     6.5\%  & 0.0\% \\
+Divorced by 35  & 15.8\% &    6.3\% & 4.5\% \\
+Divorced by 40  & 15.5\% &    8.7\% & 7.4\% \\\hline
 \end{tabular}
@@
 
-\section{Robustness Checks}
-\section{Conclusions}
+%\section{Robustness Checks}
+\section{Discussion And Conclusions}
+
+[TBD]
+
 \clearpage
+
 \section*{Literature}

### 2019-05-31 main/model.tex
Received fixes from Martí

@@
 The production function depends on age, namely I parametrize it as $q = [\mu_\xi\cdot l_f^{\theta_q} + (1-\mu_\xi)\cdot x^{\theta_q}]^{\frac1{\theta_q}}$.
-This follows, for instance, [Sommer ...], with the addition of changing labor share $\mu$.
+This follows Sommer, 2014, with the addition of changing labor share $\mu$.
+Also, following the same paper, I introduce the lower bound for possible choices of $q$ (that may rationalize additional distress in case of unplanned pregnancies).
 
@@
                     & q = [\mu_\xi\cdot l_f^{\theta_q} + (1-\mu_\xi)\cdot x^{\theta_q}]^{\frac1{\theta_q}} & \text{(production function of child quality)},\\
+                    & q \geq \underline{q} & \text{(required childcare costs)},\\
                     &  z^{f\prime} = z^f + \varepsilon^{z,f}_t, \ \ \varepsilon^{z,f}_t \sim \mathcal{N}(0;\sigma_{z,f}^2) &  \text{ (evolution of female productivity)}\\
@@
  & q = [\mu_\xi\cdot l_f^{\theta_q} + (1-\mu_\xi)\cdot x^{\theta_q}]^{\frac1{\theta_q}} & \text{(production function of child quality)},\\
+& q \geq \underline{q} & \text{(required childcare costs)},\\
  &  z' = z + \varepsilon^{z,f}_t, \ \ \varepsilon^{z,f}_t \sim \mathcal{N}(0;\sigma_{z,f}^2) &  \text{ (evolution of productivity)}\\
@@
 When kids appear, people extract additional utility of child quality, I use the following non-separable utility function:
-\[u(c,q) = \frac{\left(e^{\phi_0}\cdot c^{1-\phi_1} \cdot (\phi_2 + q)^{\phi_1}\right)^{1-\sigma}}{1-\sigma},\]
-parameter $\phi_0$ represents fixed gain in marginal utility because of having a baby, $\phi_1$ drives relative importance of child quality in utility function and $\phi_2$ is a ``luxury good'' parameter that can rationalize large child expenditures of wealthy couples that are discussed in the estimation section.
+\[u(c,q) = \frac{\left(e^{\phi_0}\cdot c^{1-\phi_1} \cdot (\phi_2 + q)^{\phi_1}\right)^{1-\sigma}}{1-\sigma} + \psi_3,\]
+parameter $\phi_0$ represents fixed gain in marginal utility because of having a baby, $\phi_1$ drives relative importance of child quality in utility function and $\phi_2$ is a ``luxury good'' parameter that can rationalize large child expenditures of wealthy couples that are discussed in the estimation section, $\phi_3$ is an additional utility shifter.

### 2019-06-02 main/draft.tex
Many edits, preparing estimation section

@@
 \usepackage{tikz}
+\usetikzlibrary{shapes,arrows}
+\usetikzlibrary{positioning}
+\tikzstyle{block} = [rectangle, draw, rounded corners]
+\tikzstyle{line} = [draw, -latex']
 \newcommand{\hypo}{\mathcal{H}}            
@@
 Unplanned pregnancies often trigger marriage for unmarried couples.
-I show that, on average, marriages in which a child appeared before or at the year of marriage are characterized with higher divorce probabilities and other outcomes suggesting lower marital quality.
-This association is robust to different ways of controlling for composition, and it is more pronounced for higher income and more educated couples.
-I argue that this pattern is consistent with household bargaining models, and the main driver of the result is unequal exposure of spouses to the childcare costs: since being single mother is a bad outcome for an educated female, she is more likely to agree to enter the marriages of lower quality.
-I estimate a lifecycle model with dynamic bargaining and fertility decisions to show how much of these differences is driven by household bargaining and how unplanned pregnancies creating unexpected childcare costs contribute to observable divorce rates.
+I document two robust empirical facts.
+First, women who have their first child before or at the year of marriage are statistically more likely to be divorced in the future than those who marry before they have a child.
+Second, this difference is more pronounced for women with higher income and educational attainment, although it is less common for them to have their kids before marriage.
+I estimate a lifecycle model with dynamic bargaining, marriage selection and unplanned pregnancy risks.
+I show that the first fact can be attributed to household bargaining friction, as women agree to enter lower quality marriages to avoid being a single mother, and  the second fact is consistent with childcare requiring both time and money, as this suggests that the costs of being a single mother are higher for more productive women.
+The model suggests large causal effect of unplanned pregnancies on observable divorce rates.
 \end{abstract}
@@
 
-\section{Empirical Strategy}
-The key empirical observation is association between marriage timing and divorce probabilities.
-The goal is to argue that statistically women who have their kids before they are married have significantly higher chance to be divorced.
-American Community Survey (ACS) data are suitable for establishing this fact only of particular subsample of women: most importantly, those who had no more than one marriage in their life.
-To re-confirm the patterns I additionally use Survey of Income and Program Participation (SIPP) data: they have smaller sample size but collect exact questions I care about.
+\section{Empirical Facts}
 
-\subsection{Defining Variables}
+This section presents key empirical facts and several exercises on testing their validity. In summary:
+\begin{enumerate}
+\item Women who had their first child before their first marriage (K$\to$M) are more likely to be divorced than those who had their first child in their first marriage (M$\to$K).
+\item Relative difference in the share of divorced between K$\to$M and M$\to$K is increasing with income and education
+\item These results are statistically significant and robust to various methods of controlling for observable composition of the groups
+\end{enumerate}
+
+To establish these results, I mainly use American Community Survey (ACS), that has large sample sizes and many relevant observable characteristics of households.
+Its only limitation is that it does not recover marriage and fertility history perfectly.
+Because of this, I re-confirm the results with the data from Survey of Income and Program Participation (SIPP 2014), that has fewer observations but more detailed marriage and fertility history.
+
+I begin with describing empirical strategy, then Section \ref{keycomp} presents the key results for ACS and SIPP, the following sections present several validity checks.
+
+\subsection{Defining Groups}
 
@@
 
-\subsection{Defining Income}
-This part describes how I define income partitions in ACS and SIPP.
-Shortly, I define income groups based on per hour income of working females, by excluding those who does not work from income considerations.
-In the appendix \ref{inc-part} I try several alternative partitions, shortly they do not affect the conclusions although they do affect magnitudes in different groups.
-
-To define per-hour income, I restrict attention to working women only.
-I define working as those who are employed now, report to work more than 25 weeks in the last year, report non-missing labor income above or equal to $\$1000$ per year, report non-missing usual working hours above or equal to $5$ hours per week.
+\subsection{Income Partition}
+I define income groups based on per hour income of working females, by excluding those who does not work or work too little from income considerations.
+Precisely, I define working as those who are employed now, report to work more than 25 weeks in the last year, report non-missing labor income above or equal to $\$1000$ per year, report non-missing usual working hours above or equal to $5$ hours per week.
 
@@
 Then I divide monthly income on four times average weekly hours to get estimate for per-hour earnings.
-As before, I residualize logs of these income by regressing them on the same factors, except that instead of year controls (I only use 2014) I use month controls that allow to capture variation in number of work days per month.
+As before, I residualize logs of these income by regressing them on the same factors, except that instead of year controls (I only use 2014) I use month controls.
 
-\section{Reduced Form Evidence}
-\subsection{Raw Comparisons}
-The main result I discuss is summarized by Figure \ref{shrs}. Table \ref{diff-raw} presents these results in numbers.
+\subsection{Raw Comparisons\label{keycomp}}
+The main results I discuss is summarized by Figure \ref{shrs} for ACS and Figure \ref{shrs-sipp} for SIPP. Table \ref{diff-raw} in presents these results in numbers.
 
 \begin{figure}[h!]
-\includegraphics[width=0.9\linewidth]{shrs.eps}
-\caption{Share of K$\to$M couples and share of divorced\label{shrs}}
+\centering
+\includegraphics[width=0.8\linewidth]{shrs.eps}
+\begin{tabular}{|p{0.65\linewidth}|}
+\hline
+\footnotesize $\Delta T = \text{Year}^{\text{first child}} - \text{Year}^{\text{first marriage}}$. Restriction: $\Delta T \in \{-5,...,10\}$\\
+\footnotesize Definition: K$\to$M if $\Delta T \in \{-5,...,0\}$. M$\to$K if $\Delta T \in \{1,...,10\}$.\\\hline
+\end{tabular}
+\caption{Share of K$\to$M couples and share of divorced, ACS\label{shrs}}
 \end{figure}
 
-%First, I compare divorce probabilities in cross-sections of people between $K\to M$ and $M\to K$ groups. See Table \ref{diff-raw} for this.
+\begin{figure}[h!]
+\centering
+\includegraphics[width=0.8\linewidth]{shrs-sipp.eps}
+\caption{Share of K$\to$M couples and share of ever divorced, SIPP\label{shrs-sipp}}
+\end{figure}
 
@@
 \end{tabular}
-\caption{Differences in raw means, ACS and SIPP\label{diff-raw}}
+\caption{Differences in raw means, ACS and SIPP (repeats Figures \ref{shrs} and \ref{shrs-sipp}) \label{diff-raw}}
 \end{center}
@@
 
-This alone is not enough to draw any causal conclusions: composition of these groups is drastically different.
+This alone is not enough to draw strong conclusions: composition of these groups is drastically different.
 For instance, as suggested by the table, people with lower income are more likely to get divorced, so even conditional on income/education group K$\to$M couples may be poorer with each subgroup.
@@
 
-\subsubsection{Duration Considerations}
-One obvious concern about comparing two groups is the temporal nature of the data: the variable I use captures fact that divorce happened in the past, and the exact timing may be different between the two groups.
-Due to cumulative nature of the divorce indicator, females who had marriage long time ago are more likely to be divorced even if their divorce chances in any given period are the same.
-Therefore, I it is useful to look at divorced females conditional on number of years passed from marriage, see the left panel of Figure \ref{pic-aftmar} for this.
-\begin{figure}[h!]
-\includegraphics[width=0.45\linewidth]{div_by_aftmar.eps}
-\includegraphics[width=0.45\linewidth]{aftmar_by_mk.eps}
-\caption{Divorced by years after marriage and distribution of years after marriage\label{pic-aftmar}}
-\end{figure}
-
-First, the pattern about different divorce probabilities hold almost uniformly by the ``age of marriage''.
-Group that has kids first both have higher share of divorced and faster growth of it as marriage progresses.
-Combining these two graphs allows to see that M$\to$K group are on average married for longer time at the moment of survey, but overall composition is not dramatically different.
-That would mean that comparing raw shares of divorced actually \emph{underestimates} the difference relative to controlling for number of years after marriage.
-This is confirmed by regression results that I skip, as in the following section I present the version where I control for years after marriage and few other factors.
-
 \subsection{Adjusting for Composition}
-
-This part documents large differences in variables that may affect divorce rates between the two groups.
-After that I try to flexibly capture these factors to see what part of the difference is attributed to the fact of being in K$\to$M group.
-These exercises use ACS data as they have more observations so the tests are less noisy.
-Shortly, after adding many controls, observable difference decreases about twice, but still remains sizable and significant.
-
-\subsection{Differences in Observables}
-This section captures demographic and economic differences of K$\to$M and M$\to$K groups.
-Many factors could compute to the observable differences, here I quantify them and this justifies use of controls in the next part.
-I present these differences for ACS, SIPP data are not very different except smaller sample size.
-For ACS, since M$\to$K and K$\to$M are defined only for subsample of people who were married once and have children present, I also compare this subsample with more general female population.
-See Table \ref{diff-contr} for the results.
-
-\begin{table}[h!]
-\begin{center}\small
-\begin{tabular}{|l|c|c|c|c|}\cline{2-5}
-\multicolumn{1}{c|}{} & \multirow{2}{*}{All population} & \multicolumn{3}{c|}{Married once + have kids}\\\cline{3-5}
-\multicolumn{1}{c|}{}  & & Everyone & K$\to$M & M$\to$K \\\cline{2-5}\hline
-\textit{Age} & 30.2    &  33.1     &       31.7     &       33.6 \\
-\textit{Age at first marriage} & 24.2         &   23.3        &    23.6        &    23.2 \\
-\textit{Age at first birth}  & 24.0       &     25.4     &       21.7       &     26.5 \\
-\textit{Husband's age} & 34.9        &    35.8      &      34.4   &         36.2  \\
-\textit{Age difference (W$-$H)} &  2.9       &      2.8      &       3.0      &       2.7  \\\hline\hline
-\textit{Eldest child's age} & 8.6         &    7.8    &        10.0     &        7.1 \\
-\textit{Number of children} & 1.1       &      2.0      &       2.3     &       2.0 \\\hline\hline
-\textit{\% college or + (own)} & 32.1\%   &         39.5\%     &       16.9\%     &       46.3\% \\
-\textit{\% college or + (husband)} & 33.3\%     &       36.4\%     &       14.2\%       &     42.7\% \\\hline\hline
-\textit{\% In Labor Force} & 75.6\%    &        68.6\%    &        70.0\%        &    68.2\% \\
-\textit{Own income (thousands)} & 32.2 &         38.4 &          29.1 &         41.1 \\
-\textit{Hours worked} & 36.6       &     36.5    &        36.8        &    36.4\\\hline\hline
-\textit{\% white (incl hispanic)} & 69.2\%      &      76.5\%      &      72.2\%      &      77.8\% \\
-\textit{\% black} & 14.8\% &        6.7\% &            12.2\%       &       5.0\% \\\hline
-\multicolumn{5}{|p{0.7\linewidth}|}{\footnotesize ACS, women, 20--40. Means. Everything is subsamples when applicable, income and hours are for subsample of women who are employed.}\\\hline
-\end{tabular}
-\caption{Differences in composition, ACS\label{diff-comp}}
-\end{center}
-\end{table}
-
-Shortly, differences in composition are pretty drastic for income and education dimension (as suggested by different shares of K$\to$M in subgroups described above).
-Most of these things are intuitive, and this suggests importance of flexibly controlling for composition.
-Similar patterns are observed if I compare composition within each subgroup that I discuss above, so partitions should matter.
-
-\subsubsection{Statistical Tests and Controls}
-
-In this section I formally test differences in shares for different subgroups and attempt to control for composition using regressions with many fixed effects.
+As people in K$\to$M and M$\to$K groups obviously have different characteristics, as in particular suggested by Figure \ref{keycomp}, I use regression techniques to see what part of difference is attributed to this composition.
+Appendix \ref{comp-diff-appendix} quantifies these differences, shortly, people in K$\to$M group are younger, less educated, have kids earlier, marry slightly earlier and have more unequal marriages.
+In this part I control for difference in observable characteristics to see what part of the difference is attributed to the fact of being in K$\to$M group.
+Shortly, controlling for observables mitigate some part of the difference but it still remains sizable and significant.
 
@@
 \end{itemize}
-
-%The most important controls are age of the eldest child and number of years after marriage. Basically, being in K$\to$M group is defined by their interaction
 
@@
 \end{table}
+
 \subsection{Additional Comparisons}
 
-Here I show few additional results.
-First, I present several graphs that suggest relation of divorce chances to child's age.
-Second, I use alternative variable and consider only people who divorced in the year of survey, as for them duration of marriage and age of the first child is (almost) perfectly observed.
-The main point of these calculations is to show that divorces seem to be more concentrated around the moment where children get older.
+Appendix \ref{extra-comparisons} shows few additional graphs and exercise. I summarize their results here:
+\begin{enumerate}
+\item In the sample I use M$\to$K women are married for longer at the moment of survey\footnote{This does not contradict the fact that K$\to$M couples marry slightly earlier as I use the sample of married women.} (so they had \emph{more} years to break up), comparing marriages of the same duration only exacerbates the difference.
+\item Children of K$\to$M women are on average older (as they are born in earlier ages).
+Divergence between K$\to$M and M$\to$K observable divorce probabilities happens around ages 5--10.
+After controlling for composition, however, I show that K$\to$M couples divorce more often even conditional on child's age.
+\item To further address concerns about duration, I recompute the numbers using getting the divorce at the year of survey, corresponding to marginal divorce probability as opposed to cumulative divorce probability (being divorced at by the moment survey).
+The magnitudes and patterns are very similar, though the numbers are harder to interpret.
+This suggests that
+\item I show that the difference is not likely to be driven by presence of stepchildren in K$\to$M families.
+In fact, as I restrict $\Delta T \geq -5$, share of stepchildren is relatively small, and surviving couples with stepchildren do not differ in observable marriage duration with couples without stepchildren.
+\end{enumerate}
 
-\subsubsection{Effects by Child's Age}
-First, I show how share of K$\to$M people varies with age of the eldest child.
-I focus on subsample children are 18 or younger, as very few people have children above 18 living in the household.
-This exercise highlights a nuance in the construction of the groups: almost no one in K$\to$M group has children of age less than a year, because that would mean the same year of marriage, first birth and ACS survey.
-Additionally, since the population is trimmed at the age of 40, having kids around age of 18 means fairly low age of the first birth, that is typically associated with being in M$\to$K group.
-Left panel of Figure \ref{pics-raw} clearly depicts both issues, and this suggests the importance of using controls.
-In the specifications above I control both for number of years after marriage and the age of the eldest child, so this should help mitigating these issues.
+\subsection{Intra-Household Allocations [TBD]}
+I document few additional facts describing the difference between K$\to$M and M$\to$K besides being divorced.
+Following Lafortune et al, 2018, I construct measure of children's grade retention based on ACS (that is an indicator if any school-age child in the household is in grade that is lower than prescribed by her/his age).
+I show this together with few other intra-houeshold characteristics: measure of inequality and labor force participation.
 
-Right panel of Figure \ref{pics-raw} shows evolution of divorce probabilities for the two groups.
-The increasing pattern is a result of cumulative nature of processes: people with elder children have more periods to get divorced.
-But the difference between two probabilities is illustrative: between ages 5 and 10 size of K$\to$M group changes a little, but difference becomes very pronounced.
-
-\begin{figure}[h!]
-\includegraphics[width=0.45\linewidth]{km_by_age.eps}
-\includegraphics[width=0.45\linewidth]{div_by_age_raw.eps}
-\caption{Share of K$\to$M couples and share of divorced, no controls\label{pics-raw}}
-\end{figure}
-
-I focus on studying difference more carefully then: first, I simply run a regression
-\[D_i = \sum_{a=0}^{18} \beta_a \cdot \I(\text{Age of eldest child}_i = a) +  \sum_{a=0}^{18} \Delta_a \cdot \I(\text{Age of eldest child} = a)\cdot \I(\text{K$\to$M})_i + \varepsilon_i,\]
-and plot $\Delta_a$.
-
-Second, for a subsample of working women I run a regression with controls that were used for Table \ref{diff-contr} above (including polynomials in income),
-\[D_i = \sum_{a=0}^{18} \Delta_a \cdot \I(\text{Age of eldest child} = a)\cdot \I(\text{K$\to$M})_i + X_i\gamma + \varepsilon_i,\]
-note that these controls include dummies for the age of the eldest child.
-
-See Figure \ref{pics-diff} for the results.
-The left panel essentially replicates the differences on the previous graph.
-The right panel is of the main focus: if the set of controls captures the issues of construction of the groups, then the differences associated with age follow the conjecture above: they are small for younger children and start to hike at ages 5--10 (I believe that first few points are mostly caused by composition).
-In general, these graphs are kind of consistent with the idea that divorces are concentrated around the moments where kids get older and childcare costs structure changes as they arguably require less time, though they do not provide a formal test.
-\begin{figure}[h!]
-\includegraphics[width=0.45\linewidth]{diff_raw_by_age.eps}
-\includegraphics[width=0.45\linewidth]{diff_cont_by_age.eps}
-\caption{Differences in divorce probabilities by child's age\label{pics-diff}}
-\end{figure}
-
-\subsubsection{Sample of Recently Divorced}
-In this exercise I utilize an alternative variable available in ACS --- indicator whether a particular person got a divorce in the year of survey.
-It can be interpreted as \textit{marginal} divorce probability as opposed to the previous value being \emph{cumulative}.
-Therefore, for instance, it does not have an upward trend in number of years married, see Figure \ref{pic-now-aftmar}.
-\begin{figure}[h]
-\centering
-\includegraphics[width=0.45\linewidth]{divnow_by_aftmar.eps}
-\caption{Probabilities of getting divorced in the year of survey\label{pic-now-aftmar}}
-\end{figure}
-
-Table \ref{diff-contr-marginal} replicates the results for marginal divorce probabilities.
-The key conclusions and proportions remain very similar, that implicitly suggests validity of the approach used above.
+See Table \ref{intra} for the numbers.
+The exercise suggests that children on K$\to$M women repeat grades significantly more often, and this happens even in higher-income households.
+[This table is based on old sample so it is slightly inconsistent with what I do above, I will redo this in one of the future iterations.]
 
 \begin{table}
-\begin{center}
-\begin{tabular}{|l|c||c|c|c|c|c|c|}\cline{2-8}
-\multicolumn{1}{c|}{} & \multicolumn{7}{c|}{ACS: sample of married once}\\\cline{2-8}
-\multicolumn{1}{c|}{} & \small Share of &\multicolumn{2}{c|}{\small Raw difference} & \multicolumn{2}{c|}{\small with controls} & \multicolumn{2}{c|}{\small controls + income} \\ \cline{3-8}
-\multicolumn{1}{c|}{} & \small divorced & $\Delta$ & (s.e.) & $\Delta$ & (s.e.) & $\Delta$ & (s.e.) \\  \hline
-\textit{All females, 20--50} & 1.94\% &  1.37\% &  (0.05\%) &  0.82\% &  (0.08\%) &  0.79\%  &  (0.11\%) \\
-\textit{All females, 35--40} &   1.59\% &  1.04\% &  (0.08\%) &   0.77\% &  (0.11\%)  &  0.74\%   &  (0.15\%) \\\hline\hline
-\multicolumn{8}{|p{0.9\linewidth}|}{\footnotesize \textbf{Education partition.}}\\\hline
-\textit{High school only, 20--50} & 2.31\% &  0.86\% &  (0.08\%) &   0.82\% &  (0.12\%)  &  0.82\%  &   (0.20\%) \\
-\textit{College graduates, 20--50} & 1.19\% &  1.24\% & (0.10\%) &   0.59\% &   (0.13\%)  &  0.50\% &    (0.16\%) \\\hline\hline
-\multicolumn{8}{|p{0.9\linewidth}|}{\footnotesize \textbf{Income partition:} working women only.}\\\hline
-\textit{Low 30\% income, 20--40} &  3.02\% &  1.31\% &  (0.14\%) &   0.90\% &  (0.21\%) &   0.91\%  &   (0.21\%) \\
-\textit{Mid 40\% income, 20--40} &  2.54\% &  1.34\% &  (0.12\%) &   0.54\%  & (0.18\%) &   0.54\% &    (0.18\%) \\
-\textit{Top 30\% income, 20--40} & 1.69\% &  1.69\% &   (0.15\%) &   0.90\%  & (0.20\%) &   0.90\% &    (0.20\%) \\
-\hline
-\end{tabular}
-\caption{Differences in marginal probablities, controlled for composition, ACS\label{diff-contr-marginal}}
-\end{center}
-\end{table}
-
-\subsection{Intra-Household Allocations}
-
-[I look at earnings, hours and LFP differences] [Grade retention depends on child's age, so need to compare given child's age]
-
-[THIS IS AN OLD VERSION OF TABLE]
-
-\begin{center}
+\centering
 \begin{tabular}{|l||c|c||c|c|}\hline
@@
 \end{tabular}
-\end{center}
+\caption{Child's outcomes and intra-household allocations [to be fixed later]\label{intra}}
+\end{table}
+
+%\section{Mechanisms}
+%This part describes the mechanism I am trying to capture in the model that is presented in the following section.
+Suppose single agents with characteristics $\omega^m$ and $\omega^f$ can meet each other and form a couple.
+On top of individual characteristics, let couples be characterized by idiosyncratic utility gain $\psi$ from being together, that can be interpreted as a love shock.
+Let also there exists a variable $\theta$ that describes terms of marriage (that can be interpreted as decision weight of wife in a collective household).
+Let for $i \in \{f,m\}$ functions $V^{i,c}$ and $V^{i,s}$ denote values of being in a couple and being single, respectively.
+People agree to marry if there exist marriage terms $\theta$ such that gains of both spouses from marriage is positive, that is
+%\[\Delta^f = V^{f,c}(\omega^f,\omega^m,\theta,\psi) - V^{f,s}(\omega^f) \geq 0, \ \ \Delta^m = V^{m,c}(\omega^f,\omega^m,\theta,\psi) -  V^{m,s}(\omega^m) \geq 0.\]
+%
+%Assume that utilities of potential spouses increase in $\psi$.
+Assume also that wife benefits from higher $\theta$ and husband benefits from lower $\theta$.
+The following illustration captures the intuition for the model:
 
@@
 
-\section{Estimation Strategy}
-I use techniques of indirect inference, extensively described, for instance, in [...] for estimation of the parameters.
-Shortly, I select parameters such that features of the real data are similar to the features of the data simulated from model.
-It is not entirely correct to call the procedure simulated method of moments, as some things that I match are not moments in econometric sense, however it has many similarities with simulated method of moments so I occasionally extend notion of moments to reflect any statistics derived from the data.
+\section{Quantification}
+I use techniques of indirect inference to fit the model to the data.
+Namely, given the parameter values I solve for decision rules and simulate $5{,}000$ agents for the first 20 periods of their life corresponding to ages of (21--40 used in the data part).
+During estimation procedure I look for the parameter values such that observable differences between between the simulated and actual data, measured by the targets I describe below, is minimal.
 
-\subsection{Overview}
-Full list of parameters involves many things:
-\begin{itemize}
-\item Fundamental parameters of income fluctuation model $(\beta,\sigma,R)$
-\item Preferences for child's quality: $(\phi_0,\phi_1,\phi_2,\phi_3)$
-\item Parameters for child's quality production function: CES power $\theta_k$ and labor shares by age $(\mu_y,\mu_s)$
-\item Unavoidable childcare costs $\underline{q}$
-\item Transition probabilities:
-\begin{itemize}
-\item Probability of pregnancy shock by age: $p^{\text{preg}}_t = \lambda^{\text{preg}}_0 - \lambda^{\text{preg}}_1\cdot t$ (2 parameters)
-\item Probability of meeting a partner: $p^{\text{meet}}_t:$ piecewise linear function between $(p^{\text{meet}}_{21},p^{\text{meet}}_{27},p^{\text{meet}}_{34},p^{\text{meet}}_{40})$ (4 parameters)
-\item Probability of re-entering marriage market and corresponding utility ``compensation'' $(p^{\text{out}},S)$. I set $S = 0$ in current calibration (so 1 parameter).
-\item Probability that kids grow up (transition from $y$ to $s$): $p_s$ (1 parameter)
-\end{itemize}
-\item Age profiles of labor income by gender: $(a^f_0,a^f_1,a^f_2)$, $(a^m_0,a^m_1,a^m_2)$ (6 parameters)
-\item Variance of shocks (6 parameters):
-\begin{itemize}
-\item Individual income shocks $\sigma^{z,f}$, $\sigma^{z,m}$.
-\item Couple's surplus shocks $\sigma^{\psi}$
-\item Couple's initial surplus $\sigma^{\psi_0}$
-\item Assets and productivity of potential partner: $(\sigma^{z,p},\sigma^{a,p})$
-\end{itemize}
-\end{itemize}
+%\subsection{Overview}
+%Full list of parameters involves many things:
+%\begin{itemize}
+%\item Fundamental parameters of income fluctuation model $(\beta,\sigma,R)$
+%\item Preferences for child's quality: $(\phi_0,\phi_1,\phi_2,\phi_3)$
+%\item Parameters for child's quality production function: CES power $\theta_k$ and labor shares by age $(\mu_y,\mu_s)$
+%\item Unavoidable childcare costs $\underline{q}$
+%\item Transition probabilities:
+%\begin{itemize}
+%\item Probability of pregnancy shock by age: $p^{\text{preg}}_t = \lambda^{\text{preg}}_0 - \lambda^{\text{preg}}_1\cdot t$ (2 parameters)
+%\item Probability of meeting a partner: $p^{\text{meet}}_t:$ piecewise linear function between $(p^{\text{meet}}_{21},p^{\text{meet}}_{27},p^{\text{meet}}_{34},p^{\text{meet}}_{40})$ (4 parameters)
+%\item Probability of re-entering marriage market and corresponding utility ``compensation'' $(p^{\text{out}},S)$. I set $S = 0$ in current calibration (so 1 parameter).
+%\item Probability that kids grow up (transition from $y$ to $s$): $p_s$ (1 parameter)
+%\end{itemize}
+%\item Age profiles of labor income by gender: $(a^f_0,a^f_1,a^f_2)$, $(a^m_0,a^m_1,a^m_2)$ (6 parameters)
+%\item Variance of shocks (6 parameters):
+%\begin{itemize}
+%\item Individual income shocks $\sigma^{z,f}$, $\sigma^{z,m}$.
+%\item Couple's surplus shocks $\sigma^{\psi}$
+%\item Couple's initial surplus $\sigma^{\psi_0}$
+%\item Assets and productivity of potential partner: $(\sigma^{z,p},\sigma^{a,p})$
+%\end{itemize}
+%\end{itemize}
 
@@
 I also externally estimate $\sigma^{z,p}$ as variance of (unpredicted part of) log-difference in spouses earnings, see Appendix \ref{partearn} for this.
-Based on this calculations I set $\sigma^{z,p} = 0.8$.
+Based on this calculations I set $\sigma^{z,p} = 0.6$ (this reproduces $\sigma = 0.7$ in the simulated data because income shocks are uncorrelated in the model as opposed to the data).
 As I both do not have good assets data and do not aim to replicate assets distribution, I set $\sigma^{a,p} = \sigma^{z,p}$.
@@
 In the part with assets, I assume savings can be between 0 and 20 times average labor income in the beginning of life.
-This bar in pretty low, but it is enough to create reasonable assets variation at the moment most of marriage decisions are made: say, median assets-to-income ratio at 30 is around 1.
+This bar in pretty low, but it is enough to create reasonable assets variation at the moment most of marriage decisions are made, creating median assets-to-income ratio at 30 being slightly above 1.
 
-\subsection{Estimated Parameters}
-Out of 31, 15 parameters are set externally, so I am left with $16$ parameters to estimate: child's parameters $(\phi_0,\phi_1,\phi_2)$, child quality production parameters $(\theta_k,\mu_y,\mu_a)$, variances of initial marriage surplus and its innovations $(\sigma^{\psi_0},\sigma^{\psi})$, meeting probability parameters $(p^{\text{meet}}_{21},p^{\text{meet}}_{27},p^{\text{meet}}_{34},p^{\text{meet}}_{40})$, pregnancy probability parameters $(\lambda^{\text{preg}}_0,\lambda^{\text{preg}}_1)$, transition out of being single mother parameters $p^{\text{out}}$, $S$.
+%\subsection{Estimated Parameters}
+%Out of 31, 15 parameters are set externally, so I am left with $16$ parameters to estimate: child's parameters $(\phi_0,\phi_1,\phi_2)$, child quality production parameters $(\theta_k,\mu_y,\mu_a)$, variances of initial marriage surplus and its innovations $(\sigma^{\psi_0},\sigma^{\psi})$, meeting probability parameters $(p^{\text{meet}}_{21},p^{\text{meet}}_{27},p^{\text{meet}}_{34},p^{\text{meet}}_{40})$, pregnancy probability parameters $(\lambda^{\text{preg}}_0,\lambda^{\text{preg}}_1)$, transition out of being single mother parameters $p^{\text{out}}$, $S$.
 Unfortunately, all of them interact and as the model is highly stylized but involves many choices, they cannot be estimated externally.
 
-Conceptually, they can be split onto following groups:
-\begin{enumerate}
-\item Unplanned pregnancy probabilities are pinned down by share of $K\to M$ females at ages 25 and 35, as well as share of planned births in total population
-\item Meeting probabilities and $\sigma^{\psi_0}$ are pinned down by shares of never married females people at 25, 35, 40 and share of females with one marriage at 40 (relevant to capture chances of re-marrying).
-\item  Preferences for kids are captured by share of childless females by age of 25, 30, 35 (identifies general scale $\phi_0$ and $\phi_3$), share of childless females for low and high income at 30 (captures $\phi_0$ and $\phi_1$).
+%Conceptually, they can be split onto following groups:
+%\begin{enumerate}
+%\item Unplanned pregnancy probabilities are pinned down by share of $K\to M$ females at ages 25 and 35, as well as share of planned births in total population
+%\item Meeting probabilities and $\sigma^{\psi_0}$ are pinned down by shares of never married females people at 25, 35, 40 and share of females with one marriage at 40 (relevant to capture chances of re-marrying).
+%\item  Preferences for kids are captured by share of childless females by age of 25, 30, 35 (identifies general scale $\phi_0$ and $\phi_3$), share of childless females for low and high income at 30 (captures $\phi_0$ and $\phi_1$).
 Luxury good parameter $\phi_2$ is mainly pinned down by ratio of amounts of money spent on kids between 75 and 25 income quantiles.
-\item Childcare production function is recovered by ratio of time spent with kids by couple's income (pins down elasticity of substitution), average share of money spent on kids (captures ``average'' $\mu$) and ratio of time spent by child's age (captures difference between $\mu_a$ and $\mu_y$)
-\item Variance of innovation to marriage surplus is mainly pinned down by share of divorced people by age (25, 35, 40) and divorce rates by income groups.
-\item Transition out of single mothers is captured by share of never married single mothers at 25 and share of divorced women with kids at 30 (picks $p^{\text{out}}$).
-\end{enumerate}
-On top of it, I use shares of divorced people for $K\to M$ and $M \to K$ at 40 per se and by income groups as additional targets that are affected by mixture of parameters described above.
-
-For income considerations I trim females/couples with $l$ larger than its 77\% percentile (corresponding to $77\%$ of ACS females of 21--40 who work at least 5 hours, imputing $l = 0$ for those who does not have kids) in simulated data, and treat their log per-hour income as missing.
+%\item Childcare production function is recovered by ratio of time spent with kids by couple's income (pins down elasticity of substitution), average share of money spent on kids (captures ``average'' $\mu$) and ratio of time spent by child's age (captures difference between $\mu_a$ and $\mu_y$)
+%\item Variance of innovation to marriage surplus is mainly pinned down by share of divorced people by age (25, 35, 40).
+%\item Transition out of single mothers is captured by share of never married single mothers at 25 and share of divorced women with kids at 30 (picks $p^{\text{out}}$).
+%\end{enumerate}
+%On top of it, I use shares of divorced people for $K\to M$ and $M \to K$ at 40 per se and by income groups as additional targets that are affected by mixture of parameters described above.
+%
+For income considerations I trim females/couples with $l$ larger than its 77\% percentile\footnote{This corresponds to $77\%$ of ACS females of 21--40 who are employed and at least 5 hours.} (imputing $l = 0$ for those who do not have children) in simulated data, and treat their log per-hour income as missing.
 I then perform similar detrending and divide individuals by group using the same regression procedure.
-I prefer detrending to simply picking $z$ as choices of $l$ are not independent on age.
 
 \section{Results}
-Table \ref{targets-main} shows the targets in the model and in the data. [Income partition is somewhat off, but things get better over time.]
+Table \ref{targets-main} shows the targets in the model and in the data.
 
@@
 \appendix
-\section{Additional Empirical Results}
-\subsection{Alternative Income Partitions\label{inc-part}}
+\section{Additional Empirical Results\label{extra-comparisons}}
+This part shows few additional comparisons supplemental to the numbers I present in the main part.
 
-Defining income groups in cross-sectional comparisons is no trivial, as income is endogenous to household's composition that is the main object of study here.
-I in this section I build several altrenative income partitions and argue that qualitatively pattern I care about holds for variety of definitions, although exact numbers may differ quite a lot.
+\subsection{Differences in Observables\label{comp-diff-appendix}}
+This section captures demographic and economic differences of K$\to$M and M$\to$K groups.
+Many factors could compute to the observable differences, here I quantify them and this justifies use of controls in the next part.
+I present these differences for ACS, SIPP data are not very different except smaller sample size.
+For ACS, since M$\to$K and K$\to$M are defined only for subsample of people who were married once and have children present, I also compare this subsample with more general female population.
+See Table \ref{diff-contr} for the results.
 
-\subsubsection{Equivalence Scale}
+\begin{table}[h!]
+\begin{center}\small
+\begin{tabular}{|l|c|c|c|c|}\cline{2-5}
+\multicolumn{1}{c|}{} & \multirow{2}{*}{All population} & \multicolumn{3}{c|}{Married once + have kids}\\\cline{3-5}
+\multicolumn{1}{c|}{}  & & Everyone & K$\to$M & M$\to$K \\\cline{2-5}\hline
+\textit{Age} & 30.2    &  33.1     &       31.7     &       33.6 \\
+\textit{Age at first marriage} & 24.2         &   23.3        &    23.6        &    23.2 \\
+\textit{Age at first birth}  & 24.0       &     25.4     &       21.7       &     26.5 \\
+\textit{Husband's age} & 34.9        &    35.8      &      34.4   &         36.2  \\
+\textit{Age difference (W$-$H)} &  2.9       &      2.8      &       3.0      &       2.7  \\\hline\hline
+\textit{Eldest child's age} & 8.6         &    7.8    &        10.0     &        7.1 \\
+\textit{Number of children} & 1.1       &      2.0      &       2.3     &       2.0 \\\hline\hline
+\textit{\% college or + (own)} & 32.1\%   &         39.5\%     &       16.9\%     &       46.3\% \\
+\textit{\% college or + (husband)} & 33.3\%     &       36.4\%     &       14.2\%       &     42.7\% \\\hline\hline
+\textit{\% In Labor Force} & 75.6\%    &        68.6\%    &        70.0\%        &    68.2\% \\
+\textit{Own income (thousands)} & 32.2 &         38.4 &          29.1 &         41.1 \\
+\textit{Hours worked} & 36.6       &     36.5    &        36.8        &    36.4\\\hline\hline
+\textit{\% white (incl hispanic)} & 69.2\%      &      76.5\%      &      72.2\%      &      77.8\% \\
+\textit{\% black} & 14.8\% &        6.7\% &            12.2\%       &       5.0\% \\\hline
+\multicolumn{5}{|p{0.7\linewidth}|}{\footnotesize ACS, women, 20--40. Means. Everything is subsamples when applicable, income and hours are for subsample of women who are employed.}\\\hline
+\end{tabular}
+\caption{Differences in composition, ACS\label{diff-comp}}
+\end{center}
+\end{table}
 
-Following development and household literature, when dividing people by income group I use total household income adjusted by OECD equivalence scale.
-The scale assigns weight 1 to the household head, 0.5 to spouse and 0.3 to each child.
+Shortly, differences in composition are pretty drastic for income and education dimension (as suggested by different shares of K$\to$M in subgroups described above).
+Most of these things are intuitive, and this suggests importance of flexibly controlling for composition.
+Similar patterns are observed if I compare composition within each subgroup that I discuss above, so partitions should matter.
 
-As I use the household income as the main variable of interest, I have to restrict attention to couples that live by themselves and single females with children.
-I do this both for ACS and SIPP, and as definition of households wary a little, this may drop out slightly different groups of people.
+\subsection{Duration Considerations}
+One obvious concern about comparing two groups is the temporal nature of the data: the variable I use captures fact that divorce happened in the past, and the exact timing may be different between the two groups.
+Due to cumulative nature of the divorce indicator, females who had marriage long time ago are more likely to be divorced even if their divorce chances in any given period are the same.
+Therefore, I it is useful to look at divorced females conditional on number of years passed from marriage, see the left panel of Figure \ref{pic-aftmar} for this.
+\begin{figure}[h!]
+\includegraphics[width=0.45\linewidth]{div_by_aftmar.eps}
+\includegraphics[width=0.45\linewidth]{aftmar_by_mk.eps}
+\caption{Divorced by years after marriage and distribution of years after marriage\label{pic-aftmar}}
+\end{figure}
 
-Formally,for income considerations I take a subsample of females who is marked as household's head or head's spouse, and who do not have any people living in household other than their spouse/unmarried partner and their children.
-This, for instance, excludes multigenerational household and people living with roommates, as total household's income may not be a relevant thing for some types of household composition.
-I cannot define income equivalence scale for more complicated living arrangements, although there are different scales that allow to do this.
+First, the pattern about different divorce probabilities hold almost uniformly by the ``age of marriage''.
+Group that has kids first both have higher share of divorced and faster growth of it as marriage progresses.
+Combining these two graphs allows to see that M$\to$K group are on average married for longer time at the moment of survey, but overall composition is not dramatically different.
+That would mean that comparing raw shares of divorced actually \emph{underestimates} the difference relative to controlling for number of years after marriage.
+This is confirmed by regression results that I skip, as in the following section I present the version where I control for years after marriage and few other factors.
 
-To mitigate selectivity, I compute income percentiles in all sample of females with household composition mentioned above, regardless of their marriage and fertility status.
-I use total family income as the key variable.
-To mitigate misreporting I drop observations with total yearly family income below $\$4000$ for ACS (that is about a third of federal the poverty level) or monthly income below $\$300$ for SIPP.
+\subsection{Effects by Child's Age}
+First, I show how share of K$\to$M people varies with age of the eldest child.
+I focus on subsample children are 18 or younger, as very few people have children above 18 living in the household.
+This exercise highlights a nuance in the construction of the groups: almost no one in K$\to$M group has children of age less than a year, because that would mean the same year of marriage, first birth and ACS survey.
+Additionally, since the population is trimmed at the age of 40, having kids around age of 18 means fairly low age of the first birth, that is typically associated with being in M$\to$K group.
+Left panel of Figure \ref{pics-raw} clearly depicts both issues, and this suggests the importance of using controls.
+In the specifications above I control both for number of years after marriage and the age of the eldest child, so this should help mitigating these issues.
 
-[TBD]
+Right panel of Figure \ref{pics-raw} shows evolution of divorce probabilities for the two groups.
+The increasing pattern is a result of cumulative nature of processes: people with elder children have more periods to get divorced.
+But the difference between two probabilities is illustrative: between ages 5 and 10 size of K$\to$M group changes a little, but difference becomes very pronounced.
 
-\section{Additional Robustness Checks}
+\begin{figure}[h!]
+\includegraphics[width=0.45\linewidth]{km_by_age.eps}
+\includegraphics[width=0.45\linewidth]{div_by_age_raw.eps}
+\caption{Share of K$\to$M couples and share of divorced, no controls\label{pics-raw}}
+\end{figure}
+
+I focus on studying difference more carefully then: first, I simply run a regression
+\[D_i = \sum_{a=0}^{18} \beta_a \cdot \I(\text{Age of eldest child}_i = a) +  \sum_{a=0}^{18} \Delta_a \cdot \I(\text{Age of eldest child} = a)\cdot \I(\text{K$\to$M})_i + \varepsilon_i,\]
+and plot $\Delta_a$.
+
+Second, for a subsample of working women I run a regression with controls that were used for Table \ref{diff-contr} above (including polynomials in income),
+\[D_i = \sum_{a=0}^{18} \Delta_a \cdot \I(\text{Age of eldest child} = a)\cdot \I(\text{K$\to$M})_i + X_i\gamma + \varepsilon_i,\]
+note that these controls include dummies for the age of the eldest child.
+
+See Figure \ref{pics-diff} for the results.
+The left panel essentially replicates the differences on the previous graph.
+The right panel is of the main focus: if the set of controls captures the issues of construction of the groups, then the differences associated with age follow the conjecture above: they are small for younger children and start to hike at ages 5--10 (I believe that first few points are mostly caused by composition).
+In general, these graphs are kind of consistent with the idea that divorces are concentrated around the moments where kids get older and childcare costs structure changes as they arguably require less time, though they do not provide a formal test.
+\begin{figure}[h!]
+\includegraphics[width=0.45\linewidth]{diff_raw_by_age.eps}
+\includegraphics[width=0.45\linewidth]{diff_cont_by_age.eps}
+\caption{Differences in divorce probabilities by child's age\label{pics-diff}}
+\end{figure}
+
+\subsection{Sample of Recently Divorced}
+In this exercise I utilize an alternative variable available in ACS --- indicator whether a particular person got a divorce in the year of survey.
+It can be interpreted as \textit{marginal} divorce probability as opposed to the previous value being \emph{cumulative}.
+Therefore, for instance, it does not have an upward trend in number of years married, see Figure \ref{pic-now-aftmar}.
+\begin{figure}[h]
+\centering
+\includegraphics[width=0.45\linewidth]{divnow_by_aftmar.eps}
+\caption{Probabilities of getting divorced in the year of survey\label{pic-now-aftmar}}
+\end{figure}
+
+Table \ref{diff-contr-marginal} replicates the results for marginal divorce probabilities.
+The key conclusions and proportions remain very similar, that implicitly suggests validity of the approach used above.
+
+\begin{table}
+\begin{center}
+\begin{tabular}{|l|c||c|c|c|c|c|c|}\cline{2-8}
+\multicolumn{1}{c|}{} & \multicolumn{7}{c|}{ACS: sample of married once}\\\cline{2-8}
+\multicolumn{1}{c|}{} & \small Share of &\multicolumn{2}{c|}{\small Raw difference} & \multicolumn{2}{c|}{\small with controls} & \multicolumn{2}{c|}{\small controls + income} \\ \cline{3-8}
+\multicolumn{1}{c|}{} & \small divorced & $\Delta$ & (s.e.) & $\Delta$ & (s.e.) & $\Delta$ & (s.e.) \\  \hline
+\textit{All females, 20--50} & 1.94\% &  1.37\% &  (0.05\%) &  0.82\% &  (0.08\%) &  0.79\%  &  (0.11\%) \\
+\textit{All females, 35--40} &   1.59\% &  1.04\% &  (0.08\%) &   0.77\% &  (0.11\%)  &  0.74\%   &  (0.15\%) \\\hline\hline
+\multicolumn{8}{|p{0.9\linewidth}|}{\footnotesize \textbf{Education partition.}}\\\hline
+\textit{High school only, 20--50} & 2.31\% &  0.86\% &  (0.08\%) &   0.82\% &  (0.12\%)  &  0.82\%  &   (0.20\%) \\
+\textit{College graduates, 20--50} & 1.19\% &  1.24\% & (0.10\%) &   0.59\% &   (0.13\%)  &  0.50\% &    (0.16\%) \\\hline\hline
+\multicolumn{8}{|p{0.9\linewidth}|}{\footnotesize \textbf{Income partition:} working women only.}\\\hline
+\textit{Low 30\% income, 20--40} &  3.02\% &  1.31\% &  (0.14\%) &   0.90\% &  (0.21\%) &   0.91\%  &   (0.21\%) \\
+\textit{Mid 40\% income, 20--40} &  2.54\% &  1.34\% &  (0.12\%) &   0.54\%  & (0.18\%) &   0.54\% &    (0.18\%) \\
+\textit{Top 30\% income, 20--40} & 1.69\% &  1.69\% &   (0.15\%) &   0.90\%  & (0.20\%) &   0.90\% &    (0.20\%) \\
+\hline
+\end{tabular}
+\caption{Differences in marginal probablities, controlled for composition, ACS\label{diff-contr-marginal}}
+\end{center}
+\end{table}
+
+\subsection{Step or Own\label{step-part}}
+People with step-children ex-ante have different marriage characteristics.
+However, as I restrict the sample to $\Delta T \geq 0$, the share of stepchildren in the observable couples is about $14\%$.
+However, this number is not perfectly observed, as ACS reports only relationships of children to the household's head, so whether the children are step can only be seen for existing couples where the head is male.
+Table \ref{step-1} presents total counts of households by values of $\Delta T$ who report that their eldest child is step, as you can see, it sharply declines with $\Delta T$ approaching zero and total share is slightly above one-eights.
+It also shows average years of being married at the moment of survey, and if couples with step children were breaking up more quickly this duration would have been lower for step group, which is not true in the data.
+\begin{table}
+\centering
+\begin{tabular}{|r|r|r||r|r|}\hline
+& \multicolumn{2}{c||}{Total count} & \multicolumn{2}{c|}{Years married} \\\hline
+ $\Delta T$ &  Own & Step & Own & Step\\\hline
+        $-5$ &     6,676  &     2,776 & 6.07  & 5.94\\
+        $-4$ &     7,952  &    2,907 & 6.48 &  6.49  \\
+        $-3$ &     9,513  &    2,638 & 6.85 & 6.87\\
+        $-2$ &    11,905 &     2,227 & 7.13   & 7.24\\
+        $-1$ &     15,344 &    1,390 &  7.52  & 7.81 \\
+         $0$  &  28,657  &      715  & 8.27  & 8.81 \\\hline
+     All $\leq 0$ &    80,047 &    12,653 \\
+     (shares)        & ($86.4\%$)  & ($13.6\%$)\\\cline{1-3}
+\end{tabular}
+\caption{Counts and marriage durations (at the moment of survey) of K$\to$M households with step and own children, ACS.\label{step-1}}
+\end{table}
 

### 2019-06-02 main/model.tex
Many edits, preparing estimation section

@@
 % \section{Model}
-The model is a mixture between conventional heterogeneous agents savings model (B--A--H framework) and household formation and dynamic bargaining, that is extensively developed by [...].
+The model is a mixture between conventional heterogeneous agents savings model (Bewley--Huggett--Aiyagari framework) and household formation and dynamic bargaining, that is extensively surveyed by Chiaporri, Mazzocco 2015.
 
@@
 Couples can choose to have a child and can also experience unplanned pregnancies.
-Unplanned pregnancies cannot be aborted.
+Unplanned pregnancies cannot be aborted, but the arrival rate of unplanned pregnancies depend on woman's productivity as well as on age.
 In addition to couples, single people who met each other may experience an unplanned pregnancy just before they decide whether to create a couple.
-In this case, upon disagreement female becomes a single mother.
-This mechanism is a stylized representation of kids appearing in unmarried cohabitating couples: since I do not model cohabitation, those kids appear just before people actually decide to marry each other.\textit{ I discuss this issue additionally in Section ....}
+In this case, upon disagreement the female becomes a single mother.
+This mechanism is a stylized representation of kids appearing in unmarried cohabitating couples: since I do not model cohabitation, those kids appear just before people actually decide to marry each other.
+
+Figure \ref{transitions} summarizes the discussed transitions.
+
+\begin{figure}
+\begin{center}
+\begin{tikzpicture}[every text node part/.style={align=center}]
+   % Place nodes
+   \node [block] (1) {Single Female\\[-0.5ex]  No Kids};
+   \node [block, above right = 1cm and 1cm of 1] (2) {Married Couple \\[-0.5ex] No Kids};
+   \node [block, right = of 2] (3) {Married Couple \\[-0.5ex] With Kids};
+   \node [block, above = 3cm of 1] (4) {Single Male  \\[-0.5ex]  No Kids};
+   \node [block, below right = 1cm and 3 cm of 1]  (5) {Single Female\\[-0.5ex]  With Kids};
+
+   \draw[->, color = red] (1) to [bend left = 10] node [left] {\scriptsize Met and agreed \\[-1ex] \scriptsize (no shock)} (2);
+   \draw[->, color = brown, dashed] (1) to [bend right = 30] node [below left] {\scriptsize Disagreed \\[-1ex] \scriptsize (shock)} (5);
+   %\draw[->, color = red] (1.210) to [bend right=90]  node [below] {\scriptsize Did't meet or \\[-1ex]  \scriptsize disagreed (no shock) } (1.330);
+    \draw[->, color = brown, dashed] (1.350) to [bend right=10]  node [right] {\scriptsize Met and agreed \\[-1ex]  \scriptsize (shock) } (3.225);
+    \draw[->, color = red] (4) to [bend right = 30] node [left] {\scriptsize Met and agreed \\[-1ex] \scriptsize (no shock)} (2);
+     \draw[->, color = brown, dashed] (4.30) to [bend left = 40] node [right] {\scriptsize Met and agreed \\[-1ex] \scriptsize (shock)} (3.60);
+    %\draw[->, color = red] (4.150) to [bend left=90]  node [above] {\scriptsize Did't meet or disagreed} (4.30);
+    \draw[->, color = blue, dashed] (2.90) to [bend left=30] node  {\scriptsize Made baby by choice \\[-1ex] \scriptsize or shock} (3.90);
+    \draw[->, color = gray] (2.120) to [bend right=10] node {\scriptsize Divorced} (4.0);
+    % \draw[->, color = gray, dashed] (2.270) to [bend left=20] node [above right] {\scriptsize Divorced  \\[-1ex] \scriptsize + became pregnant } (5.90);
+    \draw[->, color = gray] (2.300) to [bend left=10] node [below right] {\scriptsize Divorced} (1.0);
+  \draw[->, color = gray] (3.75) to [bend right=25] node  {\scriptsize \ Divorced} (4.15);
+  \draw[->, color = gray] (3.270) to [bend left=10] node [right] {\scriptsize \ Divorced} (5.90);
+    \draw[->, color = blue] (5.180) to [bend left=10] node [above] {\scriptsize Kids randomly \\[-1ex] \scriptsize disappeared} (1.345);
+\end{tikzpicture}
+\caption{ Summary of important transitions \label{transitions} }
+\end{center}
+\end{figure}
 
@@
 This is a complex issue, for instance, males would seek females who already have kids if we do not make difference between step and own children.
-Since for the purposes of the model option of being single mother is used mainly as a factor affecting bargaining, I introduce additional (constant) utility factor $S$ to flexibly control utility of exiting the state.
-This can be interpreted as equivalent of utility of kids from past marriages when entering the marriage market again.
+Since for the purposes of the model option of being single mother is used mainly as a factor affecting bargaining, I introduce additional marginal utility shifter in single mother's preferences (see \ref{prefs}), that can be interpreted as equivalent of utility/costs of children from past marriages before entering the labor market again.
 This form seems flexible enough to capture variation in an option of being a single mother without putting extra computational and modeling burden.
@@
 The problem of a female who stays single mother is
-\begin{align}V^{f,sk}_t(\omega) = \max\limits_{c} & \bigg\{ u(c,q) + \beta \E_t \Big[ p^{\text{out}} \cdot \left\{ V^{f,s}_{t+1}(\omega') + S\right\} +(1-p^{\text{out}})V^{f,sk}_{t+1}(\omega') \Big]  \bigg\},
+\begin{align}V^{f,sk}_t(\omega) = \max\limits_{c} & \bigg\{ u(c,q) + \beta \E_t \Big[ p^{\text{out}} \cdot  V^{f,s}_{t+1}(\omega') +(1-p^{\text{out}})V^{f,sk}_{t+1}(\omega') \Big]  \bigg\},
 \end{align}\vspace{-1em}
@@
 First, females may exit the state of being single mother immediately.
-This transition is assumed to happen after divorce or failure to agree to marry, therefore in the value functions and in (re)negotiation I use $V^{f,s?} = p^{\text{out}} [V^{f,s} + S] + (1-p^{\text{out}})V^{f,sk}$ instead of just $V^{f,sk}$.
+This transition is assumed to happen after divorce or failure to agree to marry, therefore in the value functions and in (re)negotiation I use $V^{f,s?} = p^{\text{out}} V^{f,s} + (1-p^{\text{out}})V^{f,sk}$ instead of just $V^{f,sk}$.
 This allows for quicker re-marriage and easier implementation of counterfactual in which females do not have risk of becoming single mother at all.
@@
 
-\subsection{Functional Forms}
+\subsection{Specifying Preferences\label{prefs}}
 I use standard CES utility functions for individuals and couples with no kids: $u(c) = \frac{c^{1-\sigma}}{1-\sigma}$.
-When kids appear, people extract additional utility of child quality, I use the following non-separable utility function:
-\[u(c,q) = \frac{\left(e^{\phi_0}\cdot c^{1-\phi_1} \cdot (\phi_2 + q)^{\phi_1}\right)^{1-\sigma}}{1-\sigma} + \psi_3,\]
-parameter $\phi_0$ represents fixed gain in marginal utility because of having a baby, $\phi_1$ drives relative importance of child quality in utility function and $\phi_2$ is a ``luxury good'' parameter that can rationalize large child expenditures of wealthy couples that are discussed in the estimation section, $\phi_3$ is an additional utility shifter.
+When kids appear, people extract additional utility of child quality, so I use the following non-separable utility function:
+\[u(c,q) = \frac{\left(e^{\phi_{0j}}\cdot c^{1-\phi_1} \cdot (\phi_2 + q)^{\phi_1}\right)^{1-\sigma}}{1-\sigma} + \phi_3,\]
+parameter $\phi_{0j}$, $j \in \{c,s\}$ represents gain in marginal utility because of having a baby (where $c$ refers to couple and $s$ to single mothers), $\phi_1$ drives relative importance of child quality in utility function and $\phi_2$ is a ``luxury good'' parameter that can rationalize large child expenditures of wealthy couples that are discussed in the estimation section, $\phi_3$ is an additional utility shifter analogous to marriage surplus in couple.
+
+This non-separability is important as I use average share of expenditures on children as one of my targets.
+In the case of $\phi_2 = 0$ and non-binding lower bar ($q > \bar{q}$) and no labor input in childcare $\mu = 0$ the share is exactly constant at the level of $\phi_1$.
+As childcare requires labor and the stylized fact is that higher-income households spend relatively more time and money on childcare, parameter $\phi_2$ is introduced.
+The role of different utility shifters $\phi_{0,j}$ for having children in couple as opposed by single mother is mainly to mitigate ad-hoc assumptions for lifecycle of single mothers.
+
+\subsection{Pregnancy Shocks}
+I use flexible bilinear form for probability that unplanned pregnancy happens:
+\[p_t(z) = \min \left\{ \max\{ p^{\text{preg}}_0 +  p^{\text{preg}}_{1} \cdot t + p^{\text{preg}}_{2} \cdot z + p^{\text{preg}}_{3} \cdot t\cdot z, 0\}, 1\right\},\]
+where $z$ stands for productivity of the woman.
+This form is a shortcut to avoid modeling variation in abortion and contraception decisions and different attitudes towards pregnancy planning by population group, though still produces substantial randomness.

### 2019-06-25 main/draft.tex
Before redoing the introduction

@@
 I show that the first fact can be attributed to household bargaining friction, as women agree to enter lower quality marriages to avoid being a single mother, and  the second fact is consistent with childcare requiring both time and money, as this suggests that the costs of being a single mother are higher for more productive women.
-The model suggests large causal effect of unplanned pregnancies on observable divorce rates.
+Qualitatively, this sheds some light on household formation and dissolution and importance of accounting for heterogeneous time costs.
+Qualitatively, the model suggests sizable causal effect of unplanned pregnancies on observable divorce rates.
 \end{abstract}
@@
 But even for educated and older women this share is substantial [...].
-An increase in non-marital cohabitation is also a documented trend, despite of this, persistent non-marital cohabitation is still not the most common thing: according to National Survey of Family And Growth, median duration of cohabitation is 22 months, and median duration of cohabitation if it ends by marriage is 21.\footnote{Source: Copen et al, \cite{copen}}.
+An increase in non-marital cohabitation is also a documented trend, despite of this, persistent non-marital cohabitation is still not commong: according to National Survey of Family And Growth, the median duration of cohabitation is 22 months, and median duration of cohabitation if it ends by marriage is 21.\footnote{Source: Copen et al, \cite{copen}}.
 Unmarried unions with children are still pretty rare: according to Payne, 2013\nocite{payne}, only about 3\% of children live in unions with unmarried cohabitating parents (as opposed to 21\% with single mothers).
 
-These numbers have important economic implications: children born by unmarried women, as documented by, for instance Kearney, Levine, 2017, display significantly worse future outcomes in terms of education and income, and this pattern survives controlling in many dimensions.
+These numbers have important economic implications: children born by unmarried women, as documented by, for instance Kearney, Levine, 2017, display significantly worse future outcomes in terms of education and income, and this pattern survives controlling for many dimensions.
 Although it is well understood that children of never married single mothers or children of blended families (when a woman marries someone else than child's father) have generally worse outcomes, even when parents marry each other after they made a baby their marriage is statistically less stable, as I argue in this paper, and the larger chances of marriage dissolution do affect future prospects of the children.
@@
 
-Namely, I document that females who had their first birth before their first marriage have higher chances of divorce in the future, this is mainly driven by women married to the parents of their children, and this is robust and relatively more pronounced for high education or high income groups.
+I document that females who had their first birth before their first marriage have higher chances of divorce in the future, this is mainly driven by women married to the parents of their children, and this is robust and relatively more pronounced for high education or high income groups.
 I then provide a model incorporating the bargaining mechanism that can rationalize these differences.
@@
 
-\subsection{Income Partition}
+\subsection{Income Partition\label{inc-part}}
 I define income groups based on per hour income of working females, by excluding those who does not work or work too little from income considerations.
@@
 
+\subsection{Externally Set Parameters}
+Length of the period is set to be 1 year, the model assumes people start their life with no assets at the age of $21$ and die deterministically at the age of $81$ (so $T = 60$).
+However, it assumes that fertility status can change only till age of 37 ($T_f = 17$) and marriage status till age of 45 ($T_m = 25$).
+It also assumes that after the age of 60 people retire and get $b = 0.4$ as in Sommer, 2014 of their income at age of 60 (so no transitions happen after $T = 40$).
+I only simulate the data for age 21--40, therefore these choices are not likely to play a crucial role.
+
+Following standard approach in the macro literature, I choose $\sigma = 1.5$, $R = 1.03$ and $\beta = 1/R$.
+Also, following estimates of Voena, 2014 I set returns to scale in couple's consumption to be $\rho_c = 0.23$.
+
+I take age-income profiles  $(a^f_0,a^f_1,a^f_2)$, $(a^m_0,a^m_1,a^m_2)$ to match coefficients in per-hour labor earnings in cross-sectional ACS regressions.
+I borrow variances of income shocks $\sigma^{z,f}$, $\sigma^{z,m}$ from Low et al 2018 as they cannot be estimated correctly in cross-sectional data, and although they were estimated for subsample of non college graduates in SIPP, in the model they produce inter-quaritle ratio at the age of 30 that is very similar to the full sample of ACS data.
+
+I also externally estimate $\sigma^{z,p}$ as variance of (unpredicted part of) log-difference in spouses earnings, see Appendix \ref{partearn} for this.
+Based on this calculations I set $\sigma^{z,p} = 0.6$ (this reproduces $\sigma = 0.7$ in the simulated data because income shocks are uncorrelated in the model so they are likely to diverge over time).
+As I both do not have good assets data and do not aim to replicate the assets distribution, I set $\sigma^{a,p} = \sigma^{z,p}$.
+
+I also set $p_s = \frac14$ to roughly reflect that children go to school on average at 6 and to pre-school few years earlier and this should correspond to a change in childcare costs structure.
+
+In the part with assets, I assume savings can be between 0 and 20 times average labor income in the beginning of life.
+This bar in pretty low, but it is enough to create reasonable assets variation at the moment most of marriage decisions are made, creating median assets-to-income ratio at 30 being slightly above 1.
+
+$9\%$ of individuals are assumed to be married at the age of $21$ ($t = 1$) as suggested by ACS, no one is assumed to have any children at $t = 1$ for simplicity.
+People start their life with zero assets and with productivity distribution as in Low et al 2018 (variances of initial $z$ are $0.15$ for women and $0.18$ for men).
+
+Random walk variables are limited to stay within $\pm 20$ one-period standard deviations around zero.
+Marriage terms $\theta$ are forced to be within $[0.1,0.9]$ as wider bounds occasionally create numerical issues.
+
+\subsection{Estimated Parameters}
+This section describes parameters estimated jointly in simulated method of moments, as well as targets that are relevant for capturing each parameter.
+Shares of divorced for K$\to$M and M$\to$K groups are used in addition to these targets, it is not obvious which parameters they identify directly, but they help to overcome possibly insufficient identifying power of the ones listed below.
+
+\begin{table}[h!!]
+\begin{tabular}{|c|p{0.25\linewidth}|c|p{0.25\linewidth}|}\hline
+Parameters & Description & Values & Identifying variation \\\hline
+($\sigma_{\psi,0},\sigma_{\psi}$) & { \footnotesize
+ Marriage surplus initial distribution \& shocks, standard deviations} &$(0.76,1.73)$ & {\footnotesize
+ \% of divorced at $(25,35,40)$, \% of women with one marriage at 40} \\\hline\hline
+$p^{\text{meet}}_{\{t = 21, 30, 37, 45\}}$ & {\footnotesize Meeting probabilities by age (linearly interpolated in between)} & 
+ $(0.13,0.10,0.07,0.12)$ & {\footnotesize \% of never married by age , \% of women with one marriage at 40} \\\hline
+$(\phi_{0,c},\phi_3)$ & {\footnotesize Scale parameters for utility of child's quality} & $(-1.89,2.58)$ & {\footnotesize \% of women with no children at 25, 35, and at 30 by income groups}\\\hline
+$(\phi_1,\phi_2,\theta)$ & {\footnotesize Child quality preference structure and substitutability} & $(0.44,1.07,-1.02)$  & {\footnotesize Average share of money spent on children, ratio of money and time spent by income groups} \\\hline
+$(\mu_y,\mu_a)$ & {\footnotesize Time structure of childcare by child's age} &  $(0.35,0.01)$ & {\footnotesize Time spent with children 0--5 relative to 6--12}  \\\hline
+$\bar{q}$ & {\footnotesize Childcare quality requirement } & $2.12$ & {\footnotesize Income of never married with kids and divorced with kids relative to general population} \\\hline
+$(p^{\text{out}},\phi_{0,s})$ & {\footnotesize Single mothers: probability of re-entering marriage market and utility shifter} & $(0.10,-3.34)$ & {\footnotesize \% of divorced with kids and never married with kids at $30$} \\\hline\hline
+$(p^{\text{preg}}_0,p^{\text{preg}}_1,p^{\text{preg}}_2,p^{\text{preg}}_3)$  & {\footnotesize Arrival probabilities of unplanned pregnancies by age and $z$} & $(0.67,0.13,-0.50,-0.005)$ & {\footnotesize Share of K$\to$M at 25, 35; share of K$\to$M at 30 and 40 by income groups} \\\hline
+\end{tabular}
+\caption{Estimated parameters and relevant variation.}
+\end{table}
+
 %\subsection{Overview}
@@
 
-Few additional choices involve time horizons, bounds for random walk variables and bounds for assets and $\theta$.
-
-\subsection{Externally Set Parameters}
-Length of the period is set to be 1 year, the model assumes people start their life with no assets at the age of $21$ and die deterministically at the age of $81$ (so $T = 60$).
-However, it assumes that fertility status can change only till age of 37 ($T_f = 17$) and marriage status till age of 45 ($T_m = 25$).
-It also assumes that after the age of 60 people retire and get $b = 0.4$ as in Sommer, 2014 of their income at age of 60 (so no transitions happen after $T = 40$).
-I only simulate the data for age 21--40, therefore these choices are not likely to play a crucial role.
-
-Following standard approach in the macro literature, I fix $\sigma = 2$, $R = 1.03$ and $\beta = 1/R$. Also, following estimates of Voena, 2014 I set $\rho_c = 0.23$.
-
-I take age-income profiles  $(a^f_0,a^f_1,a^f_2)$, $(a^m_0,a^m_1,a^m_2)$ to match coefficients in per-hour labor earnings in ACS (see Appendix [...] for details).
-I borrow variances of income shocks $\sigma^{z,f}$, $\sigma^{z,m}$ from Low et al 2018 as they cannot be estimated in cross-sectional data, and although they were estimated for subsample of non college graduates in SIPP, in the model they produce IQR at the age of 30 that is very similar to the full sample of ACS data.
-
-I also externally estimate $\sigma^{z,p}$ as variance of (unpredicted part of) log-difference in spouses earnings, see Appendix \ref{partearn} for this.
-Based on this calculations I set $\sigma^{z,p} = 0.6$ (this reproduces $\sigma = 0.7$ in the simulated data because income shocks are uncorrelated in the model as opposed to the data).
-As I both do not have good assets data and do not aim to replicate assets distribution, I set $\sigma^{a,p} = \sigma^{z,p}$.
-
-I also set $p_s = \frac14$ to roughly reflect that children go to school on average at 6 and to pre-school few years earlier and this should correspond to a change in childcare costs structure.
-
-In the part with assets, I assume savings can be between 0 and 20 times average labor income in the beginning of life.
-This bar in pretty low, but it is enough to create reasonable assets variation at the moment most of marriage decisions are made, creating median assets-to-income ratio at 30 being slightly above 1.
+%Few additional choices involve time horizons, bounds for random walk variables and bounds for assets and $\theta$.
 
@@
 %
-For income considerations I trim females/couples with $l$ larger than its 77\% percentile\footnote{This corresponds to $77\%$ of ACS females of 21--40 who are employed and at least 5 hours.} (imputing $l = 0$ for those who do not have children) in simulated data, and treat their log per-hour income as missing.
-I then perform similar detrending and divide individuals by group using the same regression procedure.
+
+\subsection{Simulating Attrition\label{income-attr}}
+For income considerations in the simulated data I put missing values of per-hour income for females/couples who have children and choose $l$ larger than its 58\% percentile.
+This corresponds to $58\%$ of ACS females of 21--40 who are employed and at least 5 hours and have children.
+I then perform similar detrending and divide individuals by group using the same regression procedure as described in \ref{inc-part}.
+This aims to replicate similar attrition in the data (as I focus on female's income many women with young children are missing from considerations), without explicitly modeling labor force participation margin.
+
+In the same way, while classifying on M$\to$K and K$\to$M I exclude women who married more than once.
+I also exclude never married single mothers, even if they re-entered the marriage market and married someone.
+Therefore, in the model K$\to$M refers only to couples who got a pregnancy shock at the meeting ($\Delta T = 0$).
+A subtle consequence of this is that as I use K$\to$M share to discipline the pregnancy shock and assume that in existing couples probabilities of unplanned pregnancies are the same as for people who just met, inside-couple fertility may not be modeled correctly.
+In the current calibration, however, majority of couple's pregnancies are by choice rather that by fertility shock\footnote{This is consistent, with, for instance, \href{https://www.guttmacher.org/fact-sheet/unintended-pregnancy-united-states}{data from Guttmacher Institute} reporting that about 4 out of 10 births are unplanned.}, and this does not seem to affect the numbers significantly, so I currently avoid introducing another parameter that adjusts intensity of pregnancy shocks for married and unmarried women.
 
 \section{Results}
-Table \ref{targets-main} shows the targets in the model and in the data.
+Table \ref{targets-main} shows the targets in the model and in the data. The model fits reasonably well and replicated key magnitudes. [I keep updating the solution occasionally].
 
@@
 Target & Model & Data \\\hline
-\multicolumn{3}{|c|}{Marriage Dimension} \\\hline
-Never married by 25  & $67.8\%$ & $68.3\%$ \\
-Never married by 30 & $25.2\%$ & $25.6\%$  \\
-Never married by 40 & $20.6\%$ & $18.8\%$  \\\hline
-One marriage by 40  & $68.9\%$ & $64.7\%$ \\\hline
-\multicolumn{3}{|c|}{Fertility Dimension} \\\hline
-No kids at 25 & $71.7\%$ & $67.7\%$ \\
-No kids at 30 (low $30\%$ income) & $58.6\%$  &   $44.0\%$ \\
-No kids at 30 (top $30\%$ income) & $56.6\%$  &   $57.8\%$ \\
-No kids at 35  & $22.4\%$  &   $28.1\%$ \\
-No kids at 40 & $19.4\%$ & $25.6\%$ \\\hline
-\multicolumn{3}{|c|}{Marriage/Fertility Interactions} \\\hline
-Never married with kids at 30 & $12.4\%$  & $12.3\%$ \\
-Divorced with kids at 30 & $3.2\%$ & $3.8\%$ \\\hline
-\multicolumn{3}{|c|}{Divorce Dimension (Divorced $=$ Divorced and not remarried)} \\\hline
-Divorced by 25 & $4.9\%$   & $10.7\%$ \\
-Divorced by 30 (low $30\%$ income)  & 15.6\% &     20.8\% \\
-Divorced by 30 (top $30\%$ income)  & 9.9\% &     8.1\% \\
-Divorced by 35  & 8.4\% &    15.9\% \\
-Divorced by 40  & 10.6\% &    19.1\% \\\hline
-\multicolumn{3}{|c|}{K$\to$M --- M$\to$K partition} \\\hline
-Divorced by 40 if K$\to$M & 20.2\% &   23.5\% \\
-Divorced by 40 if M$\to$K & 8.7\% &      13.0\% \\
-Divorced by 40 if K$\to$M and low 30\% & 34.0\% & 28.5\% \\
-Divorced by 40 if K$\to$M and top 30\% & 26.6\% & 23.3\% \\
-Divorced by 40 if M$\to$K and low 30\% & 15.3\% & 18.0\% \\
-Divorced by 40 if M$\to$K and top 30\% & 9.3\% & 11.3\% \\\hline
-\multicolumn{3}{|c|}{Childcare} \\\hline
-Money for childcare / labor income & 40.1\% & 40.0\% \\
-Time spent 0--5 / Time spent 6--12 & 1.91 & 2 \\
-Time spent top 30\% / Time spent low 30\% & 0.80 & 1.19 \\
-Money spent top 30\% / Money spent low 30\% & 1.42 & 2.67 \\\hline
+\multicolumn{3}{|c|}{Marriage Dimension (source: my calculations on ACS)} \\\hline
+Never married by 25  & $60.6\%$ & $68.3\%$ \\
+Never married by 35 & $24.6\%$ & $25.6\%$  \\
+Never married by 40 & $19.0\%$ & $18.8\%$  \\\hline
+One marriage by 40  & $69.6\%$ & $64.7\%$ \\\hline
+\multicolumn{3}{|c|}{Fertility Dimension (source: my calculations on ACS)} \\\hline
+No kids at 25 & $65.0\%$  & $67.7\%$ \\
+No kids at 30 (low $30\%$) & $52.1\%$  &   $44.0\%$ \\
+No kids at 30 (top $30\%$) & $56.8\%$ &   $57.8\%$ \\
+No kids at 35  &  $27.2\%$ &   $28.1\%$ \\
+No kids at 40 & $23.8\%$ & $25.6\%$ \\\hline
+\multicolumn{3}{|c|}{Marriage/Fertility Interactions (source: my calculations on ACS)} \\\hline
+Never married with kids at 30 & $10.4\%$  & $14.1\%$ \\
+Divorced with kids at 30 & $4.2\%$ & $4.0\%$ \\
+{\footnotesize Income(Never mar with kids at 30) / Income(Everyone at 30)} & $86.5\%$ & $70.7\%$\\
+{\footnotesize Income(Divorced with kids at 30) / Income(Everyone at 30)} & $82.2\%$ & $87.0\%$\\\hline
+\multicolumn{3}{|c|}{Divorce Dimension } \\\hline
+Divorced by 25 &  $4.5\%$  & $10.7\%$ \\
+Divorced by 30 (low $30\%$)  & $17.2\%$ &     20.8\% \\
+Divorced by 30 (top $30\%$)  & $9.0\%$ &     8.1\% \\
+Divorced by 35  & $10.8\%$ &    15.9\% \\
+Divorced by 40  & $16.3\%$  &    19.1\% \\\hline
+\multicolumn{3}{|c|}{K$\to$M --- M$\to$K partition (source: my calculations on ACS)} \\\hline
+Divorced by 40 if K$\to$M & 25.1\% &   23.5\% \\
+Divorced by 40 if M$\to$K & 14.8\% &      13.0\% \\
+Divorced by 40 if K$\to$M (low 30\%) & $32.5\%$ & 28.5\% \\
+Divorced by 40 if K$\to$M (top 30\%) & $26.1\%$ & 23.3\% \\
+Divorced by 40 if M$\to$K (low 30\%) & $23.0\%$ & 18.0\% \\
+Divorced by 40 if M$\to$K (top 30\%) & $15.2\%$ & 11.3\% \\\hline
+\multicolumn{3}{|c|}{Childcare (data: CEX and ATUS, sources in parenthesis)} \\\hline
+Money for childcare /Total expenditures &$43.4\%$ & 40.0\% \cite{sommer} \\
+Time spent 0--5 / Time spent 6--12 & $2.07$ & 2 [\href{https://www.bls.gov/charts/american-time-use/activity-by-parent.htm}{BLS}]\\
+Time spent top 30\% / Time spent low 30\% & 1.11 & 1.19  \cite{schneider} \\
+Money spent top 30\% / Money spent low 30\% & 2.56  & 2.67 \cite{schneider} \\\hline
+\multicolumn{3}{|p{0.7\linewidth}|}{\footnotesize Note 1: shares of divorced are computed for subsample of married exactly once in both data and model.}\\
+\multicolumn{3}{|p{0.7\linewidth}|}{\footnotesize Note 2: low 30\% -- mid 40\% -- top 30\% refers to woman's per hour income.
+The income partition is not full as income can be missing, see \ref{inc-part} and \ref{income-attr}.}\\\hline
 \end{tabular}
@@
 Third, in the model the largest impact of fertility shock is hurting future marriage prospective for those who stayed a single mother.
-This is curious although pretty speculative, as how the lifecycle of single mothers is modeled in pretty stylized way.
+This is curious although pretty speculative, as the lifecycle of single mothers is modeled in a pretty stylized way.
 
-In the second part, I decompose the observable impact on three groups: those who would marry the partners they met with and without fertility shocks (always-marry), those who marry only if fertility shocks happen (compliers) and those who never marry their matches (never-marry).
+In the second part, I decompose the observable impact on three groups: those who would marry the partners they met with and without fertility shocks (always-marry), those who marry only if fertility shocks happen (compliers) and those who never marry their current matches (never-marry).
 In principle, there could exist a group of those who marry only without the shock (defiers), but it is empty in my simulations.
@@
 The average impact of shock (dashed line with crosses) is a weighted average of two groups (presented by two solid lines).
-We can see that mainly the divorce probability is driven by small (5.5\%) group of compliers who have dramatically lower probabilities to stay married.
-Effect on fertility shocks on those who would've married anyway still persists, but is around 2-3\% (this is, however, substantial if we compare it to average probability of not-staying-married without shock that is slightly above 10\%.
+We can see that mainly the divorce probability is driven by small (8.7\%) group of compliers who have dramatically lower probabilities to stay married.
+Effect on fertility shocks on those who would've married anyway still persists, but is around 1\%.
+%(this is, however, substantial if we compare it to average probability of not-staying-married without shock that is slightly above 10\%.
 
-In terms of characteristics, the group of compliers have about 14\% higher median savings and 1.7\% higher median labor income relative to the general population.
+%In terms of characteristics, the group of compliers have about 14\% higher median savings and 1.7\% higher median labor income relative to the general population.
 What makes them compliers is mainly the value of love shock $\psi$: for this group, it is lower than the shock for general population of new couples by about 2/3 of initial standard deviation (i.e.
@@
 \includegraphics[width=0.85\linewidth]{individual-av.eps}
+\vspace{-3em}
 \caption{Average impact of fertility shock at $t=25$\label{individual-av}}
@@
 \includegraphics[width=0.85\linewidth]{ind-decomp.eps}
-\vspace{-4em}
+\vspace{-0.5em}
 \caption{Decomposition of the average impact\label{ind-decomp}}
@@
 \subsubsection{Aggregate Impact}
-I consider two counterfactual scenarios: when unexpected fertility never happens at the moment people meet and when it never happens at all (but people still can make kids by choice).
-I compare key targets in Table \ref{cf-shocks}.
-First, absence of fertility shocks dramatically reduces early fertility (in the model every couple before 37 can have a baby by choice).
-Then, fertility shocks at meeting are large driver of observable divorce rates, especially for higher-income couples.
+I consider two counterfactual scenarios: when unexpected fertility never happens at the moment people meet and when it never happens at all (but people still can make kids by choice) and when there are no fertility completely.
+I compare key targets in Table \ref{cf-shocks}.The key insight is that fertility shocks do look as a sizable contributor to observable divorce rates, especially at early ages and for lower income couples.
+This kind of aligns with intuitive understanding of how unplanned pregnancies affect households, and the whole magnitude is roughly in line with vairation of share of K$\to$M couples.
+
+%First, absence of fertility shocks reduces early fertility (in the model every couple before 37 can have a baby by choice).
+Then, fertility shocks at meeting are sizable driver of observable divorce rates, especially for higher-income couples and early divorces.
 Finally, fertility shocks in couples (i.e.
@@
 \multicolumn{4}{|c|}{Marriage Dimension} \\\hline
-Never married by 25  & $67.8\%$ & $68.2\%$ & $66.4\%$  \\
-Never married by 30 & $25.2\%$ & $21.8\%$ & $21.4\%$  \\
-Never married by 40 & $20.6\%$ & $18.1\%$ & $17.5\%$  \\\hline
+Never married by 25  & $60.6\%$ & $60.7\%$ & $60.7\%$  \\
+Never married by 35 & $24.6\%$ & $24.3\%$ & $24.3\%$  \\
+Never married by 40 & $19.0\%$ & $18.8\%$ & $18.8\%$  \\\hline
 \multicolumn{4}{|c|}{Fertility Dimension} \\\hline
-No kids at 25 & $59.0\%$ & $85.8\%$ & $99.8\%$ \\
-No kids at 30 (low $30\%$ income) & $76.0\%$  &   $53.5\%$ & $92.8\%$ \\
-No kids at 30 (top $30\%$ income) & $70.5\%$  &   $61.9\%$ & $89.5\%$ \\
-No kids at 35  & $32.9\%$  &   $26.8\%$ & $29.0\%$ \\
-No kids at 40 & $19.4\%$   &  $23.2\%$  & $25.1\%$ \\\hline
+No kids at 25 & $65.0\%$ & $68.4\%$ & $68.9\%$ \\
+No kids at 30 (low $30\%$ income) & $52.1\%$  &   $57.5\%$ & $59.3\%$ \\
+No kids at 30 (top $30\%$ income) & $56.8\%$  &   $57.7\%$ & $58.0\%$ \\
+No kids at 35  & $27.2\%$ &   $28.8\%$ & $29.3\%$ \\
+No kids at 40 & $23.8\%$  &  $25.0\%$  & $25.4\%$ \\\hline
 \multicolumn{4}{|c|}{Divorce Dimension} \\\hline
-Divorced by 25 & 6.6\%   &  2.9\% & 0.0\% \\
-Divorced by 30 (low $30\%$ income)  & 14.5\% &     11.4\% & 1.4\% \\
-Divorced by 30 (top $30\%$ income)  & 14.0\% &     6.5\%  & 0.0\% \\
-Divorced by 35  & 15.8\% &    6.3\% & 4.5\% \\
-Divorced by 40  & 15.5\% &    8.7\% & 7.4\% \\\hline
+Divorced by 25 & $4.5\%$   &  $3.2\%$ & $2.3\%$ \\
+Divorced by 30 (low $30\%$ income)  & $17.2\%$  &     $14.6\%$ & $11.9\%$ \\
+Divorced by 30 (top $30\%$ income)  & $9.0\%$ &     $8.3\%$  & $7.4\%$ \\
+Divorced by 35  & $10.8\%$  &   $9.9\%$ & $9.2\%$  \\
+Divorced by 40  & $16.3\%$ &  $15.4\%$ & $14.8\%$ \\\hline
 \end{tabular}
@@
 
-%\section{Robustness Checks}
-\section{Discussion And Conclusions}
+\section{Conclusions [TBD, these are just some thoughts for the future section]}
+There are several qualitative conclusions to be drawn.
+First, the story of K$\to$M couples experiencing unexpected fertility shocks (rather than choosing to have kids while being unmarried) aligns with the data and economic theory reasonably well.
+The data presents them as having lower quality marriages, and the theory incorporating bargaining provides a good rationale for why this happens.
+In this sense, dynamic bargaining theories get another validation by rationalizing this fact.
 
-[TBD]
-
+%Second, aggregate impact of pregnancy shocks is relatively larger for younger and lower income couples, while impact of selection into marriage is more pronounced for richer households. 
+Second, the data highlight the importance of time-money substitutability: for instance, in counerfactual simulations when the kids are born being adult, K$\to$M couples still divorce more because unexpected pregnancies create distress, but the gradient in terms of income cannot be seen.
+In fact, this highlights the idea that less productive couples can mitigate the distress caused by shock by substituting money with time, where higher-income couples do not have this option.
+This emphasizes importance of time use in modeling childcare.
 \clearpage

### 2019-06-25 main/model.tex
Before redoing the introduction

@@
 As childcare requires labor and the stylized fact is that higher-income households spend relatively more time and money on childcare, parameter $\phi_2$ is introduced.
-The role of different utility shifters $\phi_{0,j}$ for having children in couple as opposed by single mother is mainly to mitigate ad-hoc assumptions for lifecycle of single mothers.
+The role of different utility shifters $\phi_{0j}$ for having children in couple as opposed by single mother is mainly to mitigate ad-hoc assumptions for lifecycle of single mothers.
 

### 2019-06-29 main/draft.tex
First try

@@
 \newcommand{\deq}{\stackrel{d}{=}}
-\title{Kids First or Marriage First?}
+\title{Title TBD}
 \author{Egor Kozlov}
@@
 \begin{abstract}
-Unplanned pregnancies often trigger marriage for unmarried couples.
-I document two robust empirical facts.
-First, women who have their first child before or at the year of marriage are statistically more likely to be divorced in the future than those who marry before they have a child.
-Second, this difference is more pronounced for women with higher income and educational attainment, although it is less common for them to have their kids before marriage.
-I estimate a lifecycle model with dynamic bargaining, marriage selection and unplanned pregnancy risks.
-I show that the first fact can be attributed to household bargaining friction, as women agree to enter lower quality marriages to avoid being a single mother, and  the second fact is consistent with childcare requiring both time and money, as this suggests that the costs of being a single mother are higher for more productive women.
-Qualitatively, this sheds some light on household formation and dissolution and importance of accounting for heterogeneous time costs.
-Qualitatively, the model suggests sizable causal effect of unplanned pregnancies on observable divorce rates.
+[WIP] Many couples marry shortly after woman gets pregnant. 
 \end{abstract}
@@
 \section{Introduction}
-As of 2018, [40\%] of new births occurred by unmarried mothers.
-This number is rising for decades, but, more importantly, it is rather heterogeneous, reaching three-quarters for mothers who are African American, [...] under 25 and [...] for high school or less in education.
-But even for educated and older women this share is substantial [...].
-An increase in non-marital cohabitation is also a documented trend, despite of this, persistent non-marital cohabitation is still not commong: according to National Survey of Family And Growth, the median duration of cohabitation is 22 months, and median duration of cohabitation if it ends by marriage is 21.\footnote{Source: Copen et al, \cite{copen}}.
-Unmarried unions with children are still pretty rare: according to Payne, 2013\nocite{payne}, only about 3\% of children live in unions with unmarried cohabitating parents (as opposed to 21\% with single mothers).
 
-These numbers have important economic implications: children born by unmarried women, as documented by, for instance Kearney, Levine, 2017, display significantly worse future outcomes in terms of education and income, and this pattern survives controlling for many dimensions.
-Although it is well understood that children of never married single mothers or children of blended families (when a woman marries someone else than child's father) have generally worse outcomes, even when parents marry each other after they made a baby their marriage is statistically less stable, as I argue in this paper, and the larger chances of marriage dissolution do affect future prospects of the children.
+[This section is not supposed to be read now :)]
 
-This pairs well with common perception that some couples marry at or shortly after the moment they make a baby.
-If cohabitation is viewed as uncommitted relationship, unexpected fertility may serve as a trigger to obtain more commitment.
-Narrowly, we can interpret this as shotgun marriage story, however in the times with large abortion access many unplanned pregnancies are aborted, and cohabitating couples seem to \emph{choose} to keep an unplanned baby.
-The whole role of marriage changes as it is more socially acceptable to cohabitate, though it is much more conventional to do it before having kids.
-As highlighted by, for instance, Lafortune, 2018, despite of general decrease of role of marriage in the society, securing investment to children is one of the most important forces that makes people marry each other.
+%[Paragraph 1: is shotgun marriage a thing]
+Despite growing access to contraception and abortion technologies, situations when couples get married shortly after making a baby are still common.
+In this paper I study economics of how unexpectedly having a baby affects choices to marry.
+One can call this economics of shotgun marriage: because of number of reasons, both economic and social/cultural, many couples marry shortly after women gets pregnant, and my conjecture is that pregnancy, often unexpected, triggers their marriage.
+%The data suggest that at least 
+%This can be widely seen in the data, for instance: [large share of birth of unmarried women, peak at $dT=0$, large share of unplanned pregnancies, etc]
+[What data should I show here]
 
-This paper highlights one important aspect of this: I argue that when marriage decision is driven by having an unexpected child, bargaining friction makes marital outcome worse.
-The source of this friction is the large childcare costs and unequal relative exposure to them: as women bear relatively more childcare costs, they are more likely to enter relatively lower quality marriages as a commitment to share a part of these costs.
-This can rationalize observable differences in marital outcomes between couples with and without out-of-wedlock births.
+%[Paragraph 1.5: divorce]
+The main finding I present here is that, based on American Community Survey data, couples who experience shotgun marriage are characterized with substantially worse marital outcomes, that I refer to as having lower marriage quality.
+The main outcome I focus is the chance of future divorce, though I also document persistent differences in observable school performance of children and women labor market choices.
+The differences are economically large and survive controlling for many observable characteristics.
+After documenting the differences I attempt to study sources of them, and to do this I develop and estimate a structural model that can rationalize selection into lower quality marriages and explain different components of the observable differences, focusing mainly on divorce rates.
 
-I document that females who had their first birth before their first marriage have higher chances of divorce in the future, this is mainly driven by women married to the parents of their children, and this is robust and relatively more pronounced for high education or high income groups.
-I then provide a model incorporating the bargaining mechanism that can rationalize these differences.
-I estimate the model parameters to match observable patterns in the data and show what it implies about the role of unequal exposure to childcare costs in observable marriage selection and divorce patterns.
-Shortly, the model suggests if people could perfectly control timing of their births, general share of divorced people would decrease by [...], especially in high-income groups.
+%[Paragraph 2: why is it important? Child investment]
+Understanding divorce rates of couples with young children is way beyond just being a question of family economics, mainly because of persistent differences in outcomes of divorced parents' children.
+%Although is widely accepted that having unmarried parents is correlated with lower early-life and adult-life outcomes, and, more recently, children in blended families (having step-parents) are documented to suffer similar problems;
+Having unmarried parents or growing up in a blended family (with step-parents) are accepted as obvious contributors in worse early and adult life outcomes of children.
+In this work I show that even children who have both of their parents married may face similar risks if they are born in a shotgun marriage, as marriages are not created equal.
+These risks happen in particular, but not only because of chances of their parents future divorce.
+Understanding channels driving divorce rates is important too: in particular, I show that early-life childcare costs are important contributor in generating differences in marriage quality, and the policies focusing on childcare may mitigate selection into lower-quality marriages of those who experienced unplanned pregnancy.
+
+In additional to impact of marital stability on children outcomes, many more dimensions of these issues seem to be important.
+First, being divorced with a child early in life substantially affects women labor market choices and outcomes.
+Second, such divorces are presumably [source?] related with lowering re-marriage prospectives and possibly future fertility.
+Finally, general understanding of interaction between pregnancies and marital selection sheds some lights on internal structure of how households make decisions, provides some insights about how one should treat intra-household structure and bargaining and serves as a validity test to existing household models.
+
+The first thing I do in this paper is documenting persistent differences between two groups of people: those who had their first child before or at the year of their first marriage (kids then marriage, K$\to$M) and those who did this at least one year after (M$\to$K).
+I show that in both raw comparisons and controlling for many dimensions of composition K$\to$M group has larger share of divorced women.
+Additionally, although share of K$\to$M is not uniform across population, it is economically large for women with high income or college degrees, and the difference in divorce rates in the largest for them.
+Loosely, it seems that this timing is relatively more important for wealthier population groups.
+%[Paragraph 4: what I do: documenting the main fact]
+
+These difference, however, can hardly have any causal interpretation because of large degree of selection into the groups and many choices people make based on the things we cannot track.
+This is the main justification to use a structural model to dissemble these choices and draw some sort of causal conclusions.
+Data provide enough facts to discipline the model, suggesting that childcare costs and their heterogeneity are a major factor in these divorces, if we believe that more productive couples choose or have to spend more resources on raising their children.
+%[Paragraph 5: what I do: checks and additional facts]
+
+%[Paragraph 6: what I do: setting up the model]
+I use dynamic lifecycle model with income heterogeneity, savings, endogenous marriage, intrahousehold bargaining and divorce, fertility shocks and choices and endogenous margin of childcare costs.
+Although the structure is complicated, all these elements are important in capturing relevant margins in the data, and combining together the model can replicate divorce, marriage and fertility timing observable in the ACS data pretty realistically.
+Combining ACS with different supplemental data sources including time use and consumer expenditure surveys allow to obtain stable identification of crucial parameters and reproduce key numbers in the data almost perfectly, and then to use the model to draw conclusions about what contributes to the divorce rates that we observe.
+
+%[Paragraph 7: model: the key decomposition]
+The main insight from the model is that causal effect, which definition is non-trivial and discussed later, actually has the magnitude similar to the observable cross-sectional difference.
+According to the model, the cross-sectional effect can be decomposed onto effects of age composition, marriage length, productivity, assets and lower marriage quality, and the last one stands for the major part of the effect.
+[This is speculative and to be done].
+As additional decomposition, the causal effect itself is majorly driven by lower marriage quality of those couples who marry because they had kids and would not marry otherwise,  who correspond to the concept of compliers in empirical microeconomics literature.
+
+On more conceptual level, the model implies that a major part of dissolving marriages is drive by the threat of being single mother, and this state is particularly bad for more productive women because childcare is time-intensive activity in early child's life.
+Estimated process for unplanned pregnancies creates sizeable risk to young women, and eliminating chances of unplanned pregnancies dramatically reduces divorce rates in early life.
+[...
+...
+...
+...]
+%[Paragraph 8: model: counterfactuals that implies]
+
+%[Paragraph 9: general conclusions: lower marriage quality]
+In general, the main result that I present is the  robust association between unplanned pregnancies of unmarried women and lower quality of their marriage.
+The economics of this is based on large and heterogeneous costs of having children, that cause women to prefer entering lower quality marriages to avoid bearing childcare costs by themselves.
+The data supports this pattern: as wealthier households spend both more time and more money on their children, the high complimentarity of these activities suggests that more productive women suffer relaitvely larger losses from being single mothers (or, equivalently, gain more in their marriages relative to their outside options).
+This paper, therefore, suggests that lack of proper insurance against becoming single mother is the major factor creating bad quality marriages.
+
+%[Paragraph 10: general conclusions: implications for hh bargaining]
+From the point of view of the theory of household bargaining, this suggests that non-randomness of marriage selection is important: if people were creating couples randomly, no persistence difference in marriage quality would have been observed.
+[...
+is this even necessary]
+%it is important to take into account people's outside options when modeling household behavior, and
 

### 2019-07-02 main/draft.tex
New version

--- 2019-06-25 main/draft.tex
@@
 \begin{abstract}
-Unplanned pregnancies often trigger marriage for unmarried couples.
-I document two robust empirical facts.
-First, women who have their first child before or at the year of marriage are statistically more likely to be divorced in the future than those who marry before they have a child.
-Second, this difference is more pronounced for women with higher income and educational attainment, although it is less common for them to have their kids before marriage.
-I estimate a lifecycle model with dynamic bargaining, marriage selection and unplanned pregnancy risks.
-I show that the first fact can be attributed to household bargaining friction, as women agree to enter lower quality marriages to avoid being a single mother, and  the second fact is consistent with childcare requiring both time and money, as this suggests that the costs of being a single mother are higher for more productive women.
-Qualitatively, this sheds some light on household formation and dissolution and importance of accounting for heterogeneous time costs.
-Qualitatively, the model suggests sizable causal effect of unplanned pregnancies on observable divorce rates.
+[WIP] Many couples marry shortly after woman gets pregnant, that is often referred as a shotgun marriage.
+I show that in American Community Survey data couples who experienced the shotgun marriage divorce more often in the future.
+This feature is robust in different subpopulations, being relatively more pronounced for wealthier people, although the share of shotgun marriages is lower among them.
+I argue the difference in divorce rates can be plausibly explained by endogenous selection into marriage: women who experienced unplanned pregnancy agree to enter marriages of lower quality, and in the data having kids before marriage is a plausible proxy for having lower-quality marriage.
+I argue that these patterns in selection can be rationalized by the structure of childcare costs, where kids require both money and time, and disproportional exposure of women to these costs.
+I estimate a structural model endogenizing marriage selection and use it to argue that marriage quality, as opposed to different composition of the groups, drives the major part of the observable difference in divorce probabilities.
 \end{abstract}
@@
 \section{Introduction}
-As of 2018, [40\%] of new births occurred by unmarried mothers.
-This number is rising for decades, but, more importantly, it is rather heterogeneous, reaching three-quarters for mothers who are African American, [...] under 25 and [...] for high school or less in education.
-But even for educated and older women this share is substantial [...].
-An increase in non-marital cohabitation is also a documented trend, despite of this, persistent non-marital cohabitation is still not commong: according to National Survey of Family And Growth, the median duration of cohabitation is 22 months, and median duration of cohabitation if it ends by marriage is 21.\footnote{Source: Copen et al, \cite{copen}}.
-Unmarried unions with children are still pretty rare: according to Payne, 2013\nocite{payne}, only about 3\% of children live in unions with unmarried cohabitating parents (as opposed to 21\% with single mothers).
 
-These numbers have important economic implications: children born by unmarried women, as documented by, for instance Kearney, Levine, 2017, display significantly worse future outcomes in terms of education and income, and this pattern survives controlling for many dimensions.
-Although it is well understood that children of never married single mothers or children of blended families (when a woman marries someone else than child's father) have generally worse outcomes, even when parents marry each other after they made a baby their marriage is statistically less stable, as I argue in this paper, and the larger chances of marriage dissolution do affect future prospects of the children.
+Term ``shotgun marriage'' is traditionally used to describe situations when having a baby triggers marriage of a couple.
+This practice does not solely refer back to more conservative times: in the modern US data  around one-quarter of married women of age 21--40 had their first birth before or at the year of their first marriage\footnote{SIPP 2014 sample}.
+And, although it seems that abortion and contraception access would make give couples more choice, in the data such couples still look like something forced them to marry each other.
+In particular, their marriages appear less stable: despite of being less than quarter of the population such couples account for more than one-third of divorces\footnote{ACS 2009--2017}.
 
-This pairs well with common perception that some couples marry at or shortly after the moment they make a baby.
-If cohabitation is viewed as uncommitted relationship, unexpected fertility may serve as a trigger to obtain more commitment.
-Narrowly, we can interpret this as shotgun marriage story, however in the times with large abortion access many unplanned pregnancies are aborted, and cohabitating couples seem to \emph{choose} to keep an unplanned baby.
-The whole role of marriage changes as it is more socially acceptable to cohabitate, though it is much more conventional to do it before having kids.
-As highlighted by, for instance, Lafortune, 2018, despite of general decrease of role of marriage in the society, securing investment to children is one of the most important forces that makes people marry each other.
+The families who experienced shotgun marriages are often overlooked by the current literature focusing on child development.
+Growing up with a single mother or in a blended family (i.e.
+with step-parents) if conventionally considered as a predictor of poor early and adult-life outcomes of the children, and those who have two married biological parents are typically not considered to be at risk.
+However, not all marriages are created equal, and growing up in unstable marriage can be a predictor of ending up with a single mother or in a different household, as well as factor of risk by its own.
 
-This paper highlights one important aspect of this: I argue that when marriage decision is driven by having an unexpected child, bargaining friction makes marital outcome worse.
-The source of this friction is the large childcare costs and unequal relative exposure to them: as women bear relatively more childcare costs, they are more likely to enter relatively lower quality marriages as a commitment to share a part of these costs.
-This can rationalize observable differences in marital outcomes between couples with and without out-of-wedlock births.
+In addition to child development consequences of unstable marriages, divorces have aggregate consequences on their own.
+For instance, they appear as important driver of labor market choices and outcomes of women.
+Additionally, being divorced with young children is considered as lowering re-marriage prospective, and this can affect total fertility rates.
+Finally, the question of marriage selection and modeling divorces is curious by its own from prospectives of economic theory: if shotgun marriage predicts lower marriage quality and higher likelihood of divorce, we can use it as identifying variation for studying intra-household questions, that is relevant for developing realistic household models and understanding things that are not easy to observe directly.
 
-I document that females who had their first birth before their first marriage have higher chances of divorce in the future, this is mainly driven by women married to the parents of their children, and this is robust and relatively more pronounced for high education or high income groups.
-I then provide a model incorporating the bargaining mechanism that can rationalize these differences.
-I estimate the model parameters to match observable patterns in the data and show what it implies about the role of unequal exposure to childcare costs in observable marriage selection and divorce patterns.
-Shortly, the model suggests if people could perfectly control timing of their births, general share of divorced people would decrease by [...], especially in high-income groups.
+My main empirical contribution is describing association between having shotgun marriages and having a divorce.
+Although it is far from being causal, it is strong, robust to many ways of controlling for composition and is present among all subpopulations.
+In particular, [...].
+The heterogeneity of this pattern can be informative of the channels and serves as a setup for the next model part of the paper.
+
+As additional findings, I establish few other results. [... ... ...]
+
+To deal with obvious endogeneity and selection affecting the comparisons above I set up a structural model that is capable of understanding of anatomy of the result observed in the data.
+The model is based on growing branch of literature in dynamic household bargaining, and endogenizes savings, marriage and divorce decisions, fertility decisions and fertility shocks, together with the structure of childcare costs.
+The last element is crucial in generating heterogeneous response in terms of income or education: as time of wealthier parents is more expensive, unexpected childcare costs create relatively more distress for them.
+The key driving shock is that right before the moment people decide on their marriage they may unexpectedly get pregnant, and combination of it with differential costs of childcare generates desirable patterns in marriage selection.
+The model is estimated to match the key patterns of the data coming from American Community Survey, as well as supplementary sources, including ATUS and CEX for disciplining childcare costs part, and the structure I use fits the data extremely well.
+
+The key conclusion of the structural part is that in estimated model couples who experienced an unplanned pregnancy shock agree to enter the marriages of the lower quality, and this is the major driver of the number that we observe in the data.
+he model delivers various decomposition, splitting observable cross-sectional difference in divorce rates onto effects of age, marriage length, income, wealth and marriage quality.
+As it turns out, the last factor is the most important contributor to the difference, and its contribution is [off the similar magnitude than the observable difference].
+Decomposing the result further, the model predicts that the major part of the difference in divorce rates is driven by couples who would not have married if they did not have a baby (i.e.
+who complied with the shock), although there is some additional distress of having a baby on those who would have married anyway.
+
+Additionally, the model delivers few nontrivial counterfactual predictions.
+First, in the model major part of divorces that happen early in life are driven by unexpected pregnancies, though divorces happening later are not.
+Second, general impact of pressure of becoming single mother on marriage decisions is substantial, and eliminating unexpected pregnancies happening before the marriage decision takes place lowers amount of people married by 25 by [...].
+[Lastly, the model predicts decline in probabilities of unexpected pregnancies since 1980: the model matching fertility / marriage / divorce profiles from Census of 1980 predicts larger chances of unplanned pregnancies among all income and age groups.]
+
+As a general conclusion, the model teaches that conditional on enough observables, having kids before or at the year of marriage is a good proxy for an inferior marriage quality, and therefore the difference between two groups is not purely based on demography and economic outcomes.
+Therefore observing similar people with slight difference in fertility timing may be informative about the questions of how marital stability affects people's economic choices and future outcomes of them and their children.
+
+The causal lesson is that treat of being single mother is a major threat affecting how people make their marriage decisions, and large and unequally distributed costs of raising young children can be a driver in creation of poor quality marriages.
+This channel can be of major importance for studying welfare policies subsidizing childcare, young families and single mothers, as well as for enforcing alimony payments and child support in general.
+This also emphasizes the link between contraception and abortion access and marriages, although this impact is limited as even the wealthiest couples do experience sizable share of shotgun marriages, though this may be social rather than economic issue.
 

### 2019-07-03 main/draft.tex
Some improvements

@@
 \begin{abstract}
-[WIP] Many couples marry shortly after woman gets pregnant, that is often referred as a shotgun marriage.
-I show that in American Community Survey data couples who experienced the shotgun marriage divorce more often in the future.
+Many couples marry shortly after woman gets pregnant, that is often referred as a shotgun marriage.
+I show that in American Community Survey data couples who experienced a shotgun marriage, i.e.
+had their first child before or at the year of their marriage, divorce more often in the future.
 This feature is robust in different subpopulations, being relatively more pronounced for wealthier people, although the share of shotgun marriages is lower among them.
-I argue the difference in divorce rates can be plausibly explained by endogenous selection into marriage: women who experienced unplanned pregnancy agree to enter marriages of lower quality, and in the data having kids before marriage is a plausible proxy for having lower-quality marriage.
-I argue that these patterns in selection can be rationalized by the structure of childcare costs, where kids require both money and time, and disproportional exposure of women to these costs.
+I argue the difference in divorce rates can be explained by endogenous selection into marriage: women who experienced unplanned pregnancy agree to enter marriages of lower quality, and in the data having kids before marriage is a plausible proxy for having lower-quality marriage.
+These patterns in selection can be rationalized by the observable structure of childcare costs, where kids require both money and time, and disproportional exposure of women to these costs.
 I estimate a structural model endogenizing marriage selection and use it to argue that marriage quality, as opposed to different composition of the groups, drives the major part of the observable difference in divorce probabilities.
+Qualitatively, this all suggests that the main reason for poor performance of shotgun marriages is large, unexpected and heterogeneous childcare costs.
+Quantitatively, the model suggests sizable effects of risks of unplanned pregnancy on observable divorce rates, especially in for young women.
 \end{abstract}
@@
 
+%[Paragraph 1: is shotgun marriage still a thing?]
 Term ``shotgun marriage'' is traditionally used to describe situations when having a baby triggers marriage of a couple.
-This practice does not solely refer back to more conservative times: in the modern US data  around one-quarter of married women of age 21--40 had their first birth before or at the year of their first marriage\footnote{SIPP 2014 sample}.
+This practice does not solely refer back to more conservative times: in the modern US data  around one-quarter of married women of age 21--40 had their first birth before or at the year of their first marriage\footnote{SIPP 2014}.
 And, although it seems that abortion and contraception access would make give couples more choice, in the data such couples still look like something forced them to marry each other.
 In particular, their marriages appear less stable: despite of being less than quarter of the population such couples account for more than one-third of divorces\footnote{ACS 2009--2017}.
+%In this paper I argue that this pattern is rationalized by partners agreeing to enter marriages of lower quality as a response to unplanned pregnancy shock.
 
+%[Paragraph 2: general importance? Child development]
 The families who experienced shotgun marriages are often overlooked by the current literature focusing on child development.
@@
 with step-parents) if conventionally considered as a predictor of poor early and adult-life outcomes of the children, and those who have two married biological parents are typically not considered to be at risk.
-However, not all marriages are created equal, and growing up in unstable marriage can be a predictor of ending up with a single mother or in a different household, as well as factor of risk by its own.
+However, not all marriages are created equal, and treating all marriages equally may understate inequality: growing up in unstable marriage can be a predictor of ending up with a single mother or in a different household, as well as factor of risk by its own.
 
+%[Paragraph 3: general importance? Other ideas]
 In addition to child development consequences of unstable marriages, divorces have aggregate consequences on their own.
@@
 
+%[Paragraph 4: general importance? We need to understand trends]
+On more aggregate level, this paper can broaden understanding of recent trends in household formation.
+First, around 40\% of new births in 2018 occurred by unmarried mothers.
+Second, increase in non-marital cohabitation is a documented trend, yet the median duration of cohabitation is 22 months\footnote{Source: Copen at al,  \cite{copen}}.
+Despite of this, unmarried unions with children are pretty rare: according to Payne, 2013\nocite{payne}, only about 3\% of children live in unions with unmarried cohabitating parents, as opposed to 21\% living with single mothers.
+Even mechanically this should suggest that large chunk of unmarried women transition into marriage shortly after they give births.
+
+%[Paragraph 5: empirical findings --- main difference and subpopulations]
 My main empirical contribution is describing association between having shotgun marriages and having a divorce.
@@
 
+%[Paragraph 6: empirical findings --- duration, step-children, lf and child development outcomes]
 As additional findings, I establish few other results. [... ... ...]
 
+%[Paragraph 7: model --- setup and why do we need a model]
 To deal with obvious endogeneity and selection affecting the comparisons above I set up a structural model that is capable of understanding of anatomy of the result observed in the data.
@@
 
+%[Paragraph 8: model --- what I get from it, mechanics and anatomy]
 The key conclusion of the structural part is that in estimated model couples who experienced an unplanned pregnancy shock agree to enter the marriages of the lower quality, and this is the major driver of the number that we observe in the data.
@@
 
+%[Paragraph 9: model --- counterfactuals and speculative things]
 Additionally, the model delivers few nontrivial counterfactual predictions.
@@
 
+%[Paragraph 10: what I learned qualitatively]
 As a general conclusion, the model teaches that conditional on enough observables, having kids before or at the year of marriage is a good proxy for an inferior marriage quality, and therefore the difference between two groups is not purely based on demography and economic outcomes.
@@
 
+%[Paragraph 11: what I learned about channels]
 The causal lesson is that treat of being single mother is a major threat affecting how people make their marriage decisions, and large and unequally distributed costs of raising young children can be a driver in creation of poor quality marriages.
@@
 
+%[Paragraph 12: not a paragraph, just table of contents]
 Section 2 of this paper discuss existing results and general trends that are already described by the literature.

### 2019-07-04 main/draft.tex
More paragraphs!

@@
 had their first child before or at the year of their marriage, divorce more often in the future.
-This feature is robust in different subpopulations, being relatively more pronounced for wealthier people, although the share of shotgun marriages is lower among them.
-I argue the difference in divorce rates can be explained by endogenous selection into marriage: women who experienced unplanned pregnancy agree to enter marriages of lower quality, and in the data having kids before marriage is a plausible proxy for having lower-quality marriage.
-These patterns in selection can be rationalized by the observable structure of childcare costs, where kids require both money and time, and disproportional exposure of women to these costs.
-I estimate a structural model endogenizing marriage selection and use it to argue that marriage quality, as opposed to different composition of the groups, drives the major part of the observable difference in divorce probabilities.
-Qualitatively, this all suggests that the main reason for poor performance of shotgun marriages is large, unexpected and heterogeneous childcare costs.
-Quantitatively, the model suggests sizable effects of risks of unplanned pregnancy on observable divorce rates, especially in for young women.
+This feature is robust in different subpopulations, being relatively more pronounced for wealthier people, although the share of such marriages is lower among them.
+I argue this difference in divorce rates can be explained by endogenous selection into marriage: women who experienced unplanned pregnancy agree to enter marriages of lower quality.
+This implies that in the data having kids before marriage is a plausible proxy for having lower-quality marriage.
+Shortly, these patterns in selection can be rationalized by the structure of childcare costs, where kids require both money and time, and disproportional exposure of women to these costs.
+I estimate a structural model endogenizing marriage selection and use it to argue that differences in marriage quality, as opposed to different composition of the groups, drives the major part of the observable difference in divorce chances.
+Qualitatively, this suggests that the main reason for poor performance of shotgun marriages is large, unexpected and heterogeneous childcare costs.
+Quantitatively, the model suggests sizable effects of risks of unplanned pregnancy on observable divorce rates, especially for young women.
 \end{abstract}
@@
 This practice does not solely refer back to more conservative times: in the modern US data around one-quarter of married women of age 21--40 had their first birth before or at the year of their first marriage\footnote{SIPP 2014}.
-And, although it seems that abortion and contraception access would make give couples more choice, in the data such couples still look like something forced them to marry each other.
-In particular, their marriages appear less stable: despite of being less than quarter of the population such couples account for more than one-third of divorces\footnote{ACS 2009--2017}.
+And, although it seems that abortion and contraception access would provide people with more choice, in the data such couples still look like something pushed them to marry each other.
+In particular, their marriages appear less stable over time: despite of being less than quarter of the population shotgun marriage couples account for more than one-third of divorces\footnote{ACS 2009--2017}.
 %In this paper I argue that this pattern is rationalized by partners agreeing to enter marriages of lower quality as a response to unplanned pregnancy shock.
@@
 %[Paragraph 2: general importance? Child development]
+Understanding the sources of these difference is not just a household behavior question.
 The families who experienced shotgun marriages are often overlooked by the current literature focusing on child development.
 Growing up with a single mother or in a blended family (i.e.
-with step-parents) if conventionally considered as a predictor of poor early and adult-life outcomes of the children, and those who have two married biological parents are typically not considered to be at risk.
+with step-parents) if conventionally considered\footnote{See Kearney, Levine 2017\nocite{kearney} for an extensive review of these issues} as a predictor of poor early and adult-life outcomes of the children, and those who have two married biological parents are typically not considered to be at risk.
 However, not all marriages are created equal, and treating all marriages equally may understate inequality: growing up in unstable marriage can be a predictor of ending up with a single mother or in a different household, as well as factor of risk by its own.
@@
 %[Paragraph 4: general importance? We need to understand trends]
-On more aggregate level, this paper can broaden understanding of recent trends in household formation.
+On more aggregate level, this paper can broaden the understanding of recent trends in household formation.
 First, around 40\% of new births in 2018 occurred by unmarried mothers.
@@
 Even mechanically this should suggest that large chunk of unmarried women transition into marriage shortly after they give births.
+Finally, this research can contribute to understanding differential behavior of marriage rates among racial and income groups, as the model identifies parameters of preferences and potential partners distribution, although it does not explicitly model cohabitation that seems to become a substitute for married unions.
 
 %[Paragraph 5: empirical findings --- main difference and subpopulations]
+I use American Community survey as the main dataset, as it has large sample sizes, uniform coverage and reports wide range of household and individual characteristics.
+One important drawback of it is that its marriage history is not reported perfectly, as it only has records of the year of the most recent marriage.
+I overcome this by conditioning on women who was married exactly once.
+I then use separate dataset of Survey of Income and Program Participation to re-establish the key findings without this restriction and show that this attrition does not introduce any important distortions besides changing magnitudes of what I refer to as divorce rates.
+Although SIPP provides superior coverage of the issues I study, the current work still mainly uses ACS: SIPP sample sizes do not allow to condition on multiple dimensions like child's age.
+
 My main empirical contribution is describing association between having shotgun marriages and having a divorce.
-Although it is far from being causal, it is strong, robust to many ways of controlling for composition and is present among all subpopulations.
-In particular, [...].
-The heterogeneity of this pattern can be informative of the channels and serves as a setup for the next model part of the paper.
+Namely, I partition population of women who were married and had kids on two groups:  those who gave their first birth at least at the next year of their marriage (marriage then kids, M$\to$K) and those who gave their first birth before or at the year of their marriage (kids then marriage, K$\to$M).
+I refer to the latter group as those who experienced a shotgun marriage.
+I additionally exclude from the comparisons women who had their birth more than five years before marriage, as they are increasingly likely to form a family with someone else than child's father.
+
+%Although it is far from being causal, it is strong, robust to many ways of controlling for composition and is present among all subpopulations.
+
+The empirical result is substantially larger share of divorce people in K$\to$M group.
+Although this contrast is far from being causal, it is strong, robust to many ways of controlling for composition and is present among all subpopulations.
+Namely, chances of being divorced at the moment of survey are around twice larger for those who experienced a shotgun marriage as opposed to for those who did not ($18\%$ against $10\%$).
+This ratio reaches three for those who have college degrees, and is only around 1.25 for those who only finished a high school.
+When controlled for many confounding factors that include age, income, education and geography, as well as marriage duration and child's age, the difference in shares vanishes around a half, but remains large and statistically significant.
+%Similar heterogeneity holds if I use woman's income instead of education, although defining and measuring income can be subtle in case of women with young children.
 
 %[Paragraph 6: empirical findings --- duration, step-children, lf and child development outcomes]
-As additional findings, I establish few other results. [... ... ...]
+Investigating the data further leads to few additional insights.
+First, different timing of marriage and marriage duration at the moment of survey between the two groups does not seem to be an important contributor to the difference.
+Second, the divorces are concentrated around the time the child starts going to school.
+Third, for the cases when we can recover whether the child is step or own, this status does not affect the marriage duration, moreover, less than fifteen percent of the couples I observe seem to have step-children as I exclude those women who had their births long ago after their marriage.
+[SIPP multipartner fertility data confirm these findings?]
+%As additional findings, I establish few other results. [... ... ...]
 
 %[Paragraph 7: model --- setup and why do we need a model]
-To deal with obvious endogeneity and selection affecting the comparisons above I set up a structural model that is capable of understanding of anatomy of the result observed in the data.
+These comparisons, however, contain lots of selection issues and unobservable heterogeneity, and to deal with it I set up a structural model that is capable of understanding of anatomy of the result.
 The model is based on growing branch of literature in dynamic household bargaining, and endogenizes savings, marriage and divorce decisions, fertility decisions and fertility shocks, together with the structure of childcare costs.
@@
 
+The main driving force in the model is an unplanned pregnancy shock.
+Each period single people meet potential partners with some probability, and if the partner is met, couple decides whether to marry each other based on a simple bargaining procedure.
+The potential matches are heterogeneous in partner's characteristics and marriage quality, the latter is modeled as idiosyncratic joint utility gain of couple from being together, that can be interpreted as ``love shock''.
+Before the bargaining decision takes place, partners may or may not get an unplanned pregnancy shock, that proxies for unplanned children appearing in cohabitating unions as I do not model cohabitation explicitly.
+The shock interfers with bargaining: if the couple made a baby, woman risks to stay a single mother if the couple does not agree to marry.
+The conjecture is that in matches that experienced the shock women agree to enter the marriages of lower quality, the importance and even presence of this channel, however, depends on parameters of individual preferences and childcare costs that I estimate structurally.
+
 %[Paragraph 8: model --- what I get from it, mechanics and anatomy]
-The key conclusion of the structural part is that in estimated model couples who experienced an unplanned pregnancy shock agree to enter the marriages of the lower quality, and this is the major driver of the number that we observe in the data.
-he model delivers various decomposition, splitting observable cross-sectional difference in divorce rates onto effects of age, marriage length, income, wealth and marriage quality.
-As it turns out, the last factor is the most important contributor to the difference, and its contribution is [off the similar magnitude than the observable difference].
+The key conclusion of the estimation part is that couples who experienced an unplanned pregnancy shock indeed agree to enter the marriages of the lower quality, and this is the major driver of the number that we observe in the data.
+The model is capable of delivering various decompositions that split observable cross-sectional difference in divorce rates onto effects of age, marriage length, income, wealth and marriage quality.
+As it turns out, the last factor is the most important contributor to the difference, and its contribution is [of the similar magnitude than the observable difference].
 Decomposing the result further, the model predicts that the major part of the difference in divorce rates is driven by couples who would not have married if they did not have a baby (i.e.
@@
 %[Paragraph 9: model --- counterfactuals and speculative things]
-Additionally, the model delivers few nontrivial counterfactual predictions.
-First, in the model major part of divorces that happen early in life are driven by unexpected pregnancies, though divorces happening later are not.
+Additionally, the estimated model delivers few nontrivial counterfactual predictions.
+First, major part of divorces that happen early in life are seen as driven by unexpected pregnancies, though divorces happening later are not.
 Second, general impact of pressure of becoming single mother on marriage decisions is substantial, and eliminating unexpected pregnancies happening before the marriage decision takes place lowers amount of people married by 25 by [...].
-[Lastly, the model predicts decline in probabilities of unexpected pregnancies since 1980: the model matching fertility / marriage / divorce profiles from Census of 1980 predicts larger chances of unplanned pregnancies among all income and age groups.]
+[Last, the model predicts decline in probabilities of unexpected pregnancies since 1980: the model matching fertility / marriage / divorce profiles from Census of 1980 predicts larger chances of unplanned pregnancies among all income and age groups.]
 
@@
 As a general conclusion, the model teaches that conditional on enough observables, having kids before or at the year of marriage is a good proxy for an inferior marriage quality, and therefore the difference between two groups is not purely based on demography and economic outcomes.
-Therefore observing similar people with slight difference in fertility timing may be informative about the questions of how marital stability affects people's economic choices and future outcomes of them and their children.
+Therefore observing similar people with slight difference in fertility timing may be informative about the questions of how marital stability affects people's economic choices and future outcomes of them and their children, and, although this suggestion is model-based, many things in the actual data are consistent with it.
 
 %[Paragraph 11: what I learned about channels]
-The causal lesson is that treat of being single mother is a major threat affecting how people make their marriage decisions, and large and unequally distributed costs of raising young children can be a driver in creation of poor quality marriages.
+The causal lesson is that being single mother is a major threat affecting how people make their marriage decisions, and large and unequally distributed costs of raising young children can be a driver in creation of poor quality marriages.
 This channel can be of major importance for studying welfare policies subsidizing childcare, young families and single mothers, as well as for enforcing alimony payments and child support in general.
@@
 %[Paragraph 12: not a paragraph, just table of contents]
+I elaborate on these points in the following order.
 Section 2 of this paper discuss existing results and general trends that are already described by the literature.

### 2019-07-05 main/draft.tex
Added numbers

@@
 
-%[Paragraph 1: is shotgun marriage still a thing?]
+%[Paragraph 1: is shotgun marriage still a thing? Abstract - 2]
 Term ``shotgun marriage'' is traditionally used to describe situations when having a baby triggers marriage of a couple.
-This practice does not solely refer back to more conservative times: in the modern US data around one-quarter of married women of age 21--40 had their first birth before or at the year of their first marriage\footnote{SIPP 2014}.
-And, although it seems that abortion and contraception access would provide people with more choice, in the data such couples still look like something pushed them to marry each other.
-In particular, their marriages appear less stable over time: despite of being less than quarter of the population shotgun marriage couples account for more than one-third of divorces\footnote{ACS 2009--2017}.
+This practice does not solely refer back to more conservative times: in the modern US data around one-quarter of married women of age 21--40 had their first birth before or at the year of their first marriage.
+And, although it seems that abortion and contraception access would provide people with more choice, in this paper I show that in the data such couples still look like something pushed them to marry each other.
+In particular, their marriages appear less stable over time: despite of being less than quarter of the population shotgun marriage couples account for more than one-third of divorces.
+This paper have two main goals.
+First, I document and quantify the persistent difference in divorce rates of couples who experienced shotgun marriages and how it is pronounced for different population groups.
+Second, based on these findings I estimate a structural model that is capable of identifying importance of different channels that drive this difference.
 %In this paper I argue that this pattern is rationalized by partners agreeing to enter marriages of lower quality as a response to unplanned pregnancy shock.
@@
 %[Paragraph 2: general importance? Child development]
-Understanding the sources of these difference is not just a household behavior question.
+Understanding the sources of variation in divorce rates is not just a household behavior question, for instance, the divorces have large major impact on how people raise their children.
 The families who experienced shotgun marriages are often overlooked by the current literature focusing on child development.
@@
 
-%[Paragraph 5: empirical findings --- main difference and subpopulations]
+%[Paragraph 5: empirics --- what data do I use]
 I use American Community survey as the main dataset, as it has large sample sizes, uniform coverage and reports wide range of household and individual characteristics.
@@
 
-My main empirical contribution is describing association between having shotgun marriages and having a divorce.
-Namely, I partition population of women who were married and had kids on two groups:  those who gave their first birth at least at the next year of their marriage (marriage then kids, M$\to$K) and those who gave their first birth before or at the year of their marriage (kids then marriage, K$\to$M).
+%[Paragraph 6: empirics --- methodological issues shortly]
+My main empirical contribution is describing association between having shotgun marriages and having a divorce, that I have not found to be explicitly discussed in the existing literature.
+Methodologically, I partition population of women who were married and had kids on two groups:  those who gave their first birth at least at the next year of their marriage (marriage then kids, M$\to$K) and those who gave their first birth before or at the year of their marriage (kids then marriage, K$\to$M).
 I refer to the latter group as those who experienced a shotgun marriage.
@@
 
-%Although it is far from being causal, it is strong, robust to many ways of controlling for composition and is present among all subpopulations.
-
+%[Paragraph 7: empirica, findings --- main results]
 The empirical result is substantially larger share of divorce people in K$\to$M group.
@@
 
-%[Paragraph 6: empirical findings --- duration, step-children, lf and child development outcomes]
+%[Paragraph 8: empirical findings --- duration, step-children, lf and child development outcomes]
 Investigating the data further leads to few additional insights.
 First, different timing of marriage and marriage duration at the moment of survey between the two groups does not seem to be an important contributor to the difference.
-Second, the divorces are concentrated around the time the child starts going to school.
-Third, for the cases when we can recover whether the child is step or own, this status does not affect the marriage duration, moreover, less than fifteen percent of the couples I observe seem to have step-children as I exclude those women who had their births long ago after their marriage.
-[SIPP multipartner fertility data confirm these findings?]
-%As additional findings, I establish few other results. [... ... ...]
+Second, the divorces we observe are concentrated around the time the child starts going to school.
+Third, for the cases when one can recover whether the child is step or own (it is not perfectly transparent in ACS samples), this status does not affect the marriage duration, moreover, less than fifteen percent of the couples I for which this is observed seem to have step-children as I exclude those women who had their births long ago after their marriage.
+[SIPP multipartner fertility data confirms these findings?]
 
@@
 
+%[Paragraph 8: model --- what generates the result in theory]
 The main driving force in the model is an unplanned pregnancy shock.
@@
 
-%[Paragraph 8: model --- what I get from it, mechanics and anatomy]
+%[Paragraph 9: model --- what I get from it, mechanics and anatomy]
 The key conclusion of the estimation part is that couples who experienced an unplanned pregnancy shock indeed agree to enter the marriages of the lower quality, and this is the major driver of the number that we observe in the data.
@@
 
-%[Paragraph 9: model --- counterfactuals and speculative things]
-Additionally, the estimated model delivers few nontrivial counterfactual predictions.
+%[Paragraph 10: model --- counterfactuals and speculative things]
+Additionally, the estimated model delivers few nontrivial counterfactual predictions exploiting interaction between unplanned pregnancies and marriage quality.
 First, major part of divorces that happen early in life are seen as driven by unexpected pregnancies, though divorces happening later are not.
@@
 
-%[Paragraph 10: what I learned qualitatively]
-As a general conclusion, the model teaches that conditional on enough observables, having kids before or at the year of marriage is a good proxy for an inferior marriage quality, and therefore the difference between two groups is not purely based on demography and economic outcomes.
-Therefore observing similar people with slight difference in fertility timing may be informative about the questions of how marital stability affects people's economic choices and future outcomes of them and their children, and, although this suggestion is model-based, many things in the actual data are consistent with it.
+%[Paragraph 11: what I learned qualitatively]
+As a general conclusion, the model teaches that conditional on enough observables, having kids before or at the year of marriage is a fairly good proxy for an inferior marriage quality, and therefore the difference in divorces between two groups is not purely explained by demography and economic outcomes.
+Therefore, observing similar people with slight difference in fertility timing may be informative about the questions of how marital stability affects people's economic choices and future outcomes of them and their children.
+Although this suggestion is model-based, many features of the actual data are consistent with it.
+This does not completely rule out preference and social arguments about different attitudes to divorce and childbearing, but suggests that large part of what we see can have solid economic reasoning, and, in particular, can respond on economic policy.
 
-%[Paragraph 11: what I learned about channels]
+%[Paragraph 12: what I learned about channels]
 The causal lesson is that being single mother is a major threat affecting how people make their marriage decisions, and large and unequally distributed costs of raising young children can be a driver in creation of poor quality marriages.
 This channel can be of major importance for studying welfare policies subsidizing childcare, young families and single mothers, as well as for enforcing alimony payments and child support in general.
-This also emphasizes the link between contraception and abortion access and marriages, although this impact is limited as even the wealthiest couples do experience sizable share of shotgun marriages, though this may be social rather than economic issue.
+This also emphasizes the link between contraception and abortion access and marriages, although this impact is limited as even the wealthiest couples do experience sizable share of shotgun marriages, though this may be cultural rather than economic issue.
 
-%[Paragraph 12: not a paragraph, just table of contents]
+%[Paragraph 13: not a paragraph, just table of contents]
 I elaborate on these points in the following order.
@@
 ($\sigma_{\psi,0},\sigma_{\psi}$) & { \footnotesize
- Marriage surplus initial distribution \& shocks, standard deviations} &$(0.76,1.73)$ & {\footnotesize
+ Marriage surplus initial distribution \& shocks, standard deviations} &{\small$(0.76,1.73)$ }& {\footnotesize
  \% of divorced at $(25,35,40)$, \% of women with one marriage at 40} \\\hline\hline
-$p^{\text{meet}}_{\{t = 21, 30, 37, 45\}}$ & {\footnotesize Meeting probabilities by age (linearly interpolated in between)} & 
- $(0.13,0.10,0.07,0.12)$ & {\footnotesize \% of never married by age , \% of women with one marriage at 40} \\\hline
-$(\phi_{0,c},\phi_3)$ & {\footnotesize Scale parameters for utility of child's quality} & $(-1.89,2.58)$ & {\footnotesize \% of women with no children at 25, 35, and at 30 by income groups}\\\hline
-$(\phi_1,\phi_2,\theta)$ & {\footnotesize Child quality preference structure and substitutability} & $(0.44,1.07,-1.02)$  & {\footnotesize Average share of money spent on children, ratio of money and time spent by income groups} \\\hline
-$(\mu_y,\mu_a)$ & {\footnotesize Time structure of childcare by child's age} &  $(0.35,0.01)$ & {\footnotesize Time spent with children 0--5 relative to 6--12}  \\\hline
-$\bar{q}$ & {\footnotesize Childcare quality requirement } & $2.12$ & {\footnotesize Income of never married with kids and divorced with kids relative to general population} \\\hline
-$(p^{\text{out}},\phi_{0,s})$ & {\footnotesize Single mothers: probability of re-entering marriage market and utility shifter} & $(0.10,-3.34)$ & {\footnotesize \% of divorced with kids and never married with kids at $30$} \\\hline\hline
-$(p^{\text{preg}}_0,p^{\text{preg}}_1,p^{\text{preg}}_2,p^{\text{preg}}_3)$  & {\footnotesize Arrival probabilities of unplanned pregnancies by age and $z$} & $(0.67,0.13,-0.50,-0.005)$ & {\footnotesize Share of K$\to$M at 25, 35; share of K$\to$M at 30 and 40 by income groups} \\\hline
+$p^{\text{meet}}_{\{t = 21, 30, 37, 45\}}$ & {\footnotesize Meeting probabilities by age (linearly interpolated in between)} & {\small
+ $(0.13,0.10,0.07,0.12)$ }& {\footnotesize \% of never married by age , \% of women with one marriage at 40} \\\hline
+$(\phi_{0,c},\phi_3)$ & {\footnotesize Scale parameters for utility of child's quality} & {\small$(-1.89,2.58)$} & {\footnotesize \% of women with no children at 25, 35, and at 30 by income groups}\\\hline
+$(\phi_1,\phi_2,\theta)$ & {\footnotesize Child quality preference structure and substitutability} & {\small$(0.44,1.07,-1.02)$}  & {\footnotesize Average share of money spent on children, ratio of money and time spent by income groups} \\\hline
+$(\mu_y,\mu_a)$ & {\footnotesize Time structure of childcare by child's age} &  {\small$(0.35,0.01)$} & {\footnotesize Time spent with children 0--5 relative to 6--12}  \\\hline
+$\bar{q}$ & {\footnotesize Childcare quality requirement } & {\small$2.12$} & {\footnotesize Income of never married with kids and divorced with kids relative to general population} \\\hline
+$(p^{\text{out}},\phi_{0,s})$ & {\footnotesize Single mothers: probability of re-entering marriage market and utility shifter} & {\small$(0.10,-3.34)$} & {\footnotesize \% of divorced with kids and never married with kids at $30$} \\\hline\hline
+$p^{\text{preg}}_{0,1,2,3}$  & {\footnotesize Arrival probabilities of unplanned pregnancies by age and $z$} & {\small$(0.67,0.13,-0.50,-0.005)$} & {\footnotesize Share of K$\to$M at 25, 35; share of K$\to$M at 30 and 40 by income groups} \\\hline
 \end{tabular}

### 2019-08-22 main/draft.tex
Making tables look AER-style, wrote thanks and tittle, shorter abstact

@@
 \usepackage{tikz}
+\usepackage{verbatim}
 \usetikzlibrary{shapes,arrows}
@@
 \newcommand{\deq}{\stackrel{d}{=}}
-\title{Title TBD}
-\author{Egor Kozlov}
+\title{Economics of Shotgun Marriage}
+\author{Egor Kozlov\thanks{Economics Department, Northwestern University, egorkozlov2020@u.northwestern.edu.
+I thank my advisors Matthias Doepke, Martí Mestieri and David Berger for outstanding guidance and proofreading as part of my dissertation committee at Northwestern.
+I am grateful to Alessandra Voena for fantastic amount of help and support.
+I thank my colleagues Jane Olmstead-Rumsey, Kristina Manysheva, Bence Bardoczy, Stephanie Johnson and Gabriela Cugat, as well as all the participants of Macro Lunch at Northwestern for lots of helpful comments.}\\
+{\small Northwestern University}}
 \begin{document}
@@
 had their first child before or at the year of their marriage, divorce more often in the future.
-This feature is robust in different subpopulations, being relatively more pronounced for wealthier people, although the share of such marriages is lower among them.
-I argue this difference in divorce rates can be explained by endogenous selection into marriage: women who experienced unplanned pregnancy agree to enter marriages of lower quality.
+%This feature is robust in different subpopulations, being relatively more pronounced for wealthier people, although the share of such marriages is lower among them.
+I argue that this difference in divorce rates can be explained by endogenous selection into marriage: women who experienced unplanned pregnancy agree to enter marriages of lower quality. 
 This implies that in the data having kids before marriage is a plausible proxy for having lower-quality marriage.
-Shortly, these patterns in selection can be rationalized by the structure of childcare costs, where kids require both money and time, and disproportional exposure of women to these costs.
-I estimate a structural model endogenizing marriage selection and use it to argue that differences in marriage quality, as opposed to different composition of the groups, drives the major part of the observable difference in divorce chances.
+Then I show that these patterns in selection can be rationalized by the structure of childcare costs, where kids require both money and time, and disproportional exposure of women to these costs.
+I estimate a structural model endogenizing marriage selection and use it to argue that causal impact of unplanned fertility shock,
+%differences in marriage quality,
+as opposed to different composition of the groups, drives the major part of the observable difference in divorce chances.
 Qualitatively, this suggests that the main reason for poor performance of shotgun marriages is large, unexpected and heterogeneous childcare costs.
@@
 
-\section{Introduction}
+\newpage
 
@@
 In particular, their marriages appear less stable over time: despite of being less than quarter of the population shotgun marriage couples account for more than one-third of divorces.
-This paper have two main goals.
+This paper has two main goals.
 First, I document and quantify the persistent difference in divorce rates of couples who experienced shotgun marriages and how it is pronounced for different population groups.
@@
 %[Paragraph 2: general importance? Child development]
-Understanding the sources of variation in divorce rates is not just a household behavior question, for instance, the divorces have large major impact on how people raise their children.
+Understanding the sources of variation in divorce rates is not just a household behavior question.
+For example, the divorces have large major impact on how people raise their children.
 The families who experienced shotgun marriages are often overlooked by the current literature focusing on child development.
@@
 %[Paragraph 3: general importance? Other ideas]
-In addition to child development consequences of unstable marriages, divorces have aggregate consequences on their own.
-For instance, they appear as important driver of labor market choices and outcomes of women.
-Additionally, being divorced with young children is considered as lowering re-marriage prospective, and this can affect total fertility rates.
+In addition to child development impact of unstable marriages, divorces have other important consequences.
+For instance, they appear as an important driver of labor market choices and outcomes of women.
+Additionally, being divorced with young children is considered as lowering re-marriage prospective, and this can affect total fertility and marriage rates.
 Finally, the question of marriage selection and modeling divorces is curious by its own from prospectives of economic theory: if shotgun marriage predicts lower marriage quality and higher likelihood of divorce, we can use it as identifying variation for studying intra-household questions, that is relevant for developing realistic household models and understanding things that are not easy to observe directly.
@@
 Even mechanically this should suggest that large chunk of unmarried women transition into marriage shortly after they give births.
-Finally, this research can contribute to understanding differential behavior of marriage rates among racial and income groups, as the model identifies parameters of preferences and potential partners distribution, although it does not explicitly model cohabitation that seems to become a substitute for married unions.
+Finally, this research can contribute to understanding differential behavior of marriage rates among racial and income groups, as the model identifies parameters of preferences and potential partners distribution, although it does not explicitly consider cohabitation that seems to become a substitute for married unions.
 
 %[Paragraph 5: empirics --- what data do I use]
-I use American Community survey as the main dataset, as it has large sample sizes, uniform coverage and reports wide range of household and individual characteristics.
-One important drawback of it is that its marriage history is not reported perfectly, as it only has records of the year of the most recent marriage.
+I use American Community Survey as the main dataset, as it has large sample sizes, uniform coverage and reports wide range of household and individual characteristics.
+One important drawback of it is that the marriage history in ACS is not reported perfectly, as it only has records of the year of the most recent marriage.
 I overcome this by conditioning on women who was married exactly once.
@@
 My main empirical contribution is describing association between having shotgun marriages and having a divorce, that I have not found to be explicitly discussed in the existing literature.
-Methodologically, I partition population of women who were married and had kids on two groups:  those who gave their first birth at least at the next year of their marriage (marriage then kids, M$\to$K) and those who gave their first birth before or at the year of their marriage (kids then marriage, K$\to$M).
+Methodologically, the crucial comparison involves partitioning population of women who were married and had kids on two groups:  those who gave their first birth at least at the next year of their marriage (``marriage first'') and those who gave their first birth before or at the year of their marriage (``kids first'').
 I refer to the latter group as those who experienced a shotgun marriage.
-I additionally exclude from the comparisons women who had their birth more than five years before marriage, as they are increasingly likely to form a family with someone else than child's father.
+I additionally exclude from the comparisons women who had their birth more than five years before marriage, as they are increasingly likely to form a family with someone else than the child's father.
 
 %[Paragraph 7: empirica, findings --- main results]
-The empirical result is substantially larger share of divorce people in K$\to$M group.
-Although this contrast is far from being causal, it is strong, robust to many ways of controlling for composition and is present among all subpopulations.
+The most important empirical result I show is substantially larger share of divorced people in ``kids first'' group.
+Although this difference is far from being causal, it is strong, robust to many ways of controlling for composition and is present among all subpopulations.
 Namely, chances of being divorced at the moment of survey are around twice larger for those who experienced a shotgun marriage as opposed to for those who did not ($18\%$ against $10\%$).
@@
 First, different timing of marriage and marriage duration at the moment of survey between the two groups does not seem to be an important contributor to the difference.
-Second, the divorces we observe are concentrated around the time the child starts going to school.
+Second, the ``extra'' divorces we observe are concentrated around the time the child starts going to school.
 Third, for the cases when one can recover whether the child is step or own (it is not perfectly transparent in ACS samples), this status does not affect the marriage duration, moreover, less than fifteen percent of the couples I for which this is observed seem to have step-children as I exclude those women who had their births long ago after their marriage.
-[SIPP multipartner fertility data confirms these findings?]
+[SIPP has better data to confirm these findings?]
 
@@
 The last element is crucial in generating heterogeneous response in terms of income or education: as time of wealthier parents is more expensive, unexpected childcare costs create relatively more distress for them.
-The key driving shock is that right before the moment people decide on their marriage they may unexpectedly get pregnant, and combination of it with differential costs of childcare generates desirable patterns in marriage selection.
-The model is estimated to match the key patterns of the data coming from American Community Survey, as well as supplementary sources, including ATUS and CEX for disciplining childcare costs part, and the structure I use fits the data extremely well.
+The model is estimated to match the key patterns of the data coming from American Community Survey, as well as supplementary sources, including ATUS and CEX for disciplining childcare costs part, and the structure I use fits the data well.
 
@@
 Each period single people meet potential partners with some probability, and if the partner is met, couple decides whether to marry each other based on a simple bargaining procedure.
-The potential matches are heterogeneous in partner's characteristics and marriage quality, the latter is modeled as idiosyncratic joint utility gain of couple from being together, that can be interpreted as ``love shock''.
+The potential matches are heterogeneous in partner's characteristics and marriage quality, the latter is modeled as idiosyncratic joint utility gain of couple from being together, that can be interpreted as a ``love shock''.
 Before the bargaining decision takes place, partners may or may not get an unplanned pregnancy shock, that proxies for unplanned children appearing in cohabitating unions as I do not model cohabitation explicitly.
 The shock interfers with bargaining: if the couple made a baby, woman risks to stay a single mother if the couple does not agree to marry.
-The conjecture is that in matches that experienced the shock women agree to enter the marriages of lower quality, the importance and even presence of this channel, however, depends on parameters of individual preferences and childcare costs that I estimate structurally.
+The conjecture is that in matches that experienced the shock women agree to enter the marriages of lower quality.
+Importance and even presence of this channel, however, depend on parameters of individual preferences and childcare costs that I estimate structurally.
 
@@
 Decomposing the result further, the model predicts that the major part of the difference in divorce rates is driven by couples who would not have married if they did not have a baby (i.e.
-who complied with the shock), although there is some additional distress of having a baby on those who would have married anyway.
+who complied with the shock), although there is some additional distress of having an unplanned baby on those who would have married anyway.
 
@@
 First, major part of divorces that happen early in life are seen as driven by unexpected pregnancies, though divorces happening later are not.
-Second, general impact of pressure of becoming single mother on marriage decisions is substantial, and eliminating unexpected pregnancies happening before the marriage decision takes place lowers amount of people married by 25 by [...].
+Second, general impact of pressure of becoming a single mother on marriage decisions is substantial, and eliminating unexpected pregnancies happening before the marriage decision takes place lowers amount of people married by 25 by [...].
 [Last, the model predicts decline in probabilities of unexpected pregnancies since 1980: the model matching fertility / marriage / divorce profiles from Census of 1980 predicts larger chances of unplanned pregnancies among all income and age groups.]
@@
 Although this suggestion is model-based, many features of the actual data are consistent with it.
-This does not completely rule out preference and social arguments about different attitudes to divorce and childbearing, but suggests that large part of what we see can have solid economic reasoning, and, in particular, can respond on economic policy.
+This does not completely rule out preference and social arguments about different attitudes to divorce and childbearing, but suggests that large part of what we see can have solid economic reasoning, and, in particular, can respond to economic policy.
 
 %[Paragraph 12: what I learned about channels]
-The causal lesson is that being single mother is a major threat affecting how people make their marriage decisions, and large and unequally distributed costs of raising young children can be a driver in creation of poor quality marriages.
+The main causal insight is that being single mother is a major threat affecting how people make their marriage decisions, and large and unequally distributed costs of raising young children can be a driver in creation of poor quality marriages.
 This channel can be of major importance for studying welfare policies subsidizing childcare, young families and single mothers, as well as for enforcing alimony payments and child support in general.
@@
 I elaborate on these points in the following order.
-Section 2 of this paper discuss existing results and general trends that are already described by the literature.
-Section 3 discuss how I construct the dataset and timing variables, and Section 5 establishes main observations about differences in groups depending on their marriage timing, together with a number of additional considerations useful to understand the content of these differences.
-After that I explain theoretical model in Section 5 and the process of setting its parameters in Section 6.
-Section 7 describes estimation results and counterfactual simulations implied by these results.
-Section 8 provides more general discussion and concludes.
+Section 1 of this paper discuss existing results and general trends that are already described by the literature.
+Section 2 discuss how I construct the dataset and timing variables, and Section 3 establishes main observations about differences in groups depending on their marriage timing, together with a number of additional considerations useful to understand the content of these differences.
+After that I explain theoretical model in Section 4 and the process of setting its parameters in Section 5.
+Section 6 describes estimation results and counterfactual simulations implied by these results.
+Section 7 provides more general discussion and concludes.
 
@@
 \begin{center}
-\begin{tabular}{|l|c||c|c||c||c|c|}\cline{2-7}
-\multicolumn{1}{c|}{} & \multicolumn{3}{c||}{\small ACS: sample of married once} & \multicolumn{3}{c|}{\small SIPP: sample of  ever married} \\\cline{2-7}
-\multicolumn{1}{c|}{}  & \multirow{2}{*}{\small \% K$\to$M}& \multicolumn{2}{c||}{\textbf{\% divorced (now)}} & \multirow{2}{*}{\footnotesize \% K$\to$M }& \multicolumn{2}{c|}{\textbf{\% divorced (ever)}} \\\cline{3-4}\cline{6-7}
-\multicolumn{1}{c|}{}  &  &\small if M$\to$K &  \small if K$\to$M & & \small if M$\to$K & \small if K$\to$M \\\hline
-\textit{All females, 21--40}  & $23.2\%$ & $10.1\%$ & $18.0\%$ & $22.8\%$ & $24.1\%$ & $34.8\%$ \\
-\textit{All females, 35--40} & $18.0\%$ & $11.3\%$ & $23.0\%$ & $18.9\%$ & $28.1\%$ & $45.5\%$ \\\hline\hline
-\multicolumn{7}{|p{0.9\linewidth}|}{\footnotesize \textbf{Education partition.}}\\\hline
-\textit{High school only, 21--40} & $34.2\%$ & $13.9\%$ & $17.2\%$ & $33.4\%$ & $29.2\%$  & $34.2\%$ \\
-\textit{College graduates, 21--40} & $9.9\%$ & $5.3\%$ & $14.7\%$ & $10.5\%$ & $12.9\%$ & $27.5\%$ \\\hline\hline
-\multicolumn{7}{|p{0.9\linewidth}|}{\footnotesize \textbf{Income partition:} per-hour income for subsample of working women.}\\\hline
-\textit{Low 30\% income, 21--40} & $30.8\%$ & $17.6\%$ & $24.1\%$ & $34.6\%$ & $30.5\%$ &  $42.5\%$ \\
-\textit{Mid 40\% income, 21--40} & $25.5\%$ & $14.2\%$ & $21.5\%$ & $24.1\%$ & $28.3\%$ & $45.8\%$ \\
-\textit{Top 30\% income, 21--40} & $14.1\%$ & $7.7\%$ & $18.1\%$ & $12.4\%$ & $22.9\%$  & $32.8\%$ \\\hline
-\multicolumn{7}{|p{0.9\linewidth}|}{\footnotesize $\Delta T = \text{Year}^{\text{first child}} - \text{Year}^{\text{first marriage}}$. Restriction: $\Delta T \in \{-5,...,10\}$.}\\
-\multicolumn{7}{|p{0.9\linewidth}|}{\footnotesize Definition: K$\to$M if $\Delta T \in \{-5,...,0\}$. M$\to$K if $\Delta T \in \{1,...,10\}$.}\\
-\hline
+\begin{tabular}{l c  c c  c  c c}\cline{2-7}
+\multicolumn{1}{c }{} & \multicolumn{3}{c  }{\small ACS: sample of married once} & \multicolumn{3}{c }{\small SIPP: sample of ever married} \\\cline{2-7}
+\multicolumn{1}{c }{}  & \small{\textbf{share of}} & \multicolumn{2}{c  }{\textbf{divorced (now)}} & \small{\textbf{share of}} & \multicolumn{2}{c }{\textbf{divorced (ever)}} \\\cline{3-4}\cline{6-7}
+\multicolumn{1}{c }{}  & \small{\textbf{kids-first}}  &\small marriage-first &  \small kids-first & \small{\textbf{kids-first}} & \small marriage-first & \small kids-first \\\hline\hline
+\textit{All females, 21--40}  & $23.2$ & $10.1$ & $18.0$ & $22.8$ & $24.1$ & $34.8$ \\
+\textit{All females, 35--40} & $18.0$ & $11.3$ & $23.0$ & $18.9$ & $28.1$ & $45.5$ \\\hline\hline
+\multicolumn{7}{p{0.9\linewidth}}{ \textbf{Education partition}}\\\hline
+\textit{High school only, 21--40} & $34.2$ & $13.9$ & $17.2$ & $33.4$ & $29.2$  & $34.2$ \\
+\textit{College graduates, 21--40} & $\phantom{0}9.9$ & $\phantom{0}5.3$ & $14.7$ & $10.5$ & $12.9$ & $27.5$ \\\hline\hline
+\multicolumn{7}{p{0.9\linewidth}}{ \textbf{Income partition:} subsample of working women}\\\hline
+\textit{Low 30\% income, 21--40} & $30.8$ & $17.6$ & $24.1$ & $34.6$ & $30.5$ &  $42.5$ \\
+\textit{Mid 40\% income, 21--40} & $25.5$ & $14.2$ & $21.5$ & $24.1$ & $28.3$ & $45.8$ \\
+\textit{Top 30\% income, 21--40} & $14.1$ & $\phantom{0}7.7$ & $18.1$ & $12.4$ & $22.9$  & $32.8$ \\\hline\hline
+\multicolumn{7}{p{\linewidth}}{ \footnotesize \textit{Notes:} Kids-first refers to women whose year of marriage is 0 to 5 years after their year of the first birth.
+Marriage-first refers to women whose year of marriage is 1 to 10 years after their year of the first birth.
+Women outside of these two groups are excluded.
+Income here refers to residualized per-hour income for subsample of working women, as described in Section \ref{inc-part}.
+ACS is American Community Survey, 2008--2016.
+SIPP is Survey of Income and Program Participation, 2014.}\\\hline
 \end{tabular}
-\caption{Differences in raw means, ACS and SIPP (repeats Figures \ref{shrs} and \ref{shrs-sipp}) \label{diff-raw}}
+\caption{Percentages of kids-first group and percentages of divorced in each group by subpopulations. ACS and SIPP (repeats Figures \ref{shrs} and \ref{shrs-sipp}) \label{diff-raw}}
 \end{center}
@@
 
-\subsection{Adjusting for Composition}
+\subsection{Adjusting for Composition\label{adcomp}}
 As people in K$\to$M and M$\to$K groups obviously have different characteristics, as in particular suggested by Figure \ref{keycomp}, I use regression techniques to see what part of difference is attributed to this composition.
@@
 \begin{center}
-\begin{tabular}{|l|c||c|c|c|c|c|c|}\cline{2-8}
-\multicolumn{1}{c|}{} & \multicolumn{7}{c|}{ACS: sample of married once}\\\cline{2-8}
-\multicolumn{1}{c|}{} & \small Share of &\multicolumn{2}{c|}{\small Raw difference} & \multicolumn{2}{c|}{\small with controls} & \multicolumn{2}{c|}{\small controls + income} \\ \cline{3-8}
-\multicolumn{1}{c|}{} & \small divorced & $\Delta$ & (s.e.) & $\Delta$ & (s.e.) & $\Delta$ & (s.e.) \\  \hline
-\textit{All females, 20--50} & 14.3\% &  7.9\% &   (0.1\%) &    5.7\%  &    (0.2\%) &     4.6\%   &   0.2\% \\
-\textit{All females, 35--40} &  17.3\% &  11.7\%  &  (0.2\%) &   6.7\% &  (0.3\%) &     6.2\%  &   (0.4\%) \\\hline\hline
-\multicolumn{8}{|p{0.9\linewidth}|}{\footnotesize \textbf{Education partition.}}\\\hline
-\textit{High school only, 20--50} & 18.1\% &   3.3\%  &    (0.2\%) &   8.4\% &   (0.3\%)   &  7.8\%   &  (0.5\%) \\
-\textit{College graduates, 20--50} & 8.5\% &   9.4\% &   (0.2\%) &   3.9\% &   (0.3\%) &   3.2\%   &  (0.4\%) \\\hline\hline
-\multicolumn{8}{|p{0.9\linewidth}|}{\footnotesize \textbf{Income partition:} working women only.}\\\hline
-\textit{Low 30\% income, 20--40} &  21.6\% &  6.5\% &    (0.3\%) &    7.8\% &  (0.5\%) &   7.8\%   &  (0.5\%) \\
-\textit{Mid 40\% income, 20--40} &  17.1\% &  7.3\% &   (0.3\%) &    3.6\%   & (0.4\%) &   3.6\% &    (0.4\%) \\
-\textit{Top 30\% income, 20--40} & 11.1\% &  10.4\% &  (0.3\%) &    3.5\% &   (0.4\%) & 3.5\% &  (0.4\%) \\
+\begin{tabular}{l c  c c c c c c}\cline{2-8}
+\multicolumn{1}{c}{} & \multicolumn{7}{c}{ACS: sample of married once}\\\cline{2-8}
+\multicolumn{1}{c}{} & \small Share of &\multicolumn{2}{c}{\small raw difference} & \multicolumn{2}{c}{\small with controls \,\,\, \,\,\,} & \multicolumn{2}{c}{\small controls + income} \\ \cline{3-8}
+\multicolumn{1}{c}{} & \small divorced & $\Delta$ & (s.e.) & $\Delta$ & (s.e.) & $\Delta$ & (s.e.) \\\hline\hline
+\textit{All females, 20--50} & 14.3 &  \phantom{0}7.9 &   (0.1) &    5.7  &    (0.2) &     4.6   &   0.2 \\
+\textit{All females, 35--40} &  17.3 &  11.7  &  (0.2) &   6.7 &  (0.3) &     6.2  &   (0.4) \\\hline\hline
+\multicolumn{8}{p{0.6\linewidth}}{\textbf{Education partition}}\\\hline
+\textit{High school only, 20--50} & 18.1 &   \phantom{0}3.3  &    (0.2) &   8.4 &   (0.3)   &  7.8   &  (0.5) \\
+\textit{College graduates, 20--50} & \phantom{0}8.5 &   \phantom{0}9.4 &   (0.2) &   3.9 &   (0.3) &   3.2   &  (0.4) \\\hline\hline
+\multicolumn{8}{p{0.6\linewidth}}{\textbf{Income partition:} subsample of working women.}\\\hline
+\textit{Low 30\% income, 20--40} &  21.6 &  \phantom{0}6.5 &    (0.3) &    7.8 &  (0.5) &   7.8   &  (0.5) \\
+\textit{Mid 40\% income, 20--40} &  17.1 &  \phantom{0}7.3 &   (0.3) &    3.6   & (0.4) &   3.6 &    (0.4) \\
+\textit{Top 30\% income, 20--40} & 11.1 &  10.4 &  (0.3) &    3.5 &   (0.4) & 3.5 &  (0.4) \\
 \hline
+\multicolumn{8}{p{0.9\linewidth}}{ \footnotesize \textit{Notes:} here $\Delta$ is defined as coefficient in cross-sectional regression
+$\text{Divorced}_i = \Delta \cdot \text{Kids-First}_i + \gamma X_i + \varepsilon_i$,
+where ``Raw difference'' refers to regression with no $X_i$ that corresponds to the difference in means in Table \ref{diff-raw}.
+Controls are described in Section \ref{adcomp}.
+Coefficients and standard errors are multiplied by 100 so they refer to percents.}\\\hline
 \end{tabular}
-\caption{Differences, controlled for composition, ACS\label{diff-contr}}
+\caption{Percentage difference in shares of divorced, raw and adjusted for composition via regression, ACS data.\label{diff-contr}}
 \end{center}
@@
 \centering
-\begin{tabular}{|l||c|c||c|c|}\hline
-& \multicolumn{2}{|c||}{\footnotesize \textit{All couples}}  & \multicolumn{2}{|c|}{\footnotesize \textit{Top 40\% of income} } \\\hline
-&                       \footnotesize  M$\to$K & \footnotesize  K$\to$M &  \footnotesize  M$\to$K & \footnotesize  K$\to$M \\\hline
-\footnotesize \textit{Female in Labor Force}        & $70.3\%$  & $71.4\%$  & $79.3\%$ & $85.3\%$  \\\hline
-\footnotesize \textit{Earninigs Ratio}    &  $95.2\%$ & $93.9\%$  &  $95.8\%$ &  $96.4\%$ \\\hline
-\footnotesize \textit{Kids' Grade Retention}  & $4.6\%$    & $7.6\%$ &   $3.3\%$   & $5.9\%$ \\\hline
+\begin{tabular}{l  c c  c c }
+\cline{2-5}
+& \multicolumn{2}{ c  }{\footnotesize \textit{All couples}}  & \multicolumn{2}{ c }{\footnotesize \textit{Top 40\% of income} } \\\hline
+&                       \footnotesize  Marriage-first & \footnotesize Kids-first &  \footnotesize  Marriage-first & \footnotesize  Kids-first \\\hline
+\footnotesize \textit{Female in Labor Force}        & $70.3$  & $71.4$  & $79.3$ & $85.3$  \\\hline
+\footnotesize \textit{Earninigs Ratio}    &  $95.2$ & $93.9$  &  $95.8$ &  $96.4$ \\\hline
+\footnotesize \textit{Kids' Grade Retention}  & $\phantom{0}4.6$    & $\phantom{0}7.6$ &   $\phantom{0}3.3$   & $\phantom{0}5.9$ \\\hline
 \end{tabular}
-\caption{Child's outcomes and intra-household allocations [to be fixed later]\label{intra}}
+\caption{Child's outcomes and intra-household allocations, all numbers are in percents.\label{intra}} %[This is borrowed from an old draft so its methodology is slightly inconsistent, to be fixed later]\label{intra}}
 \end{table}
@@
 \begin{table}[h!!]
-\begin{tabular}{|c|p{0.25\linewidth}|c|p{0.25\linewidth}|}\hline
+\begin{tabular}{ c p{0.25\linewidth} c p{0.25\linewidth}}\hline\hline
 Parameters & Description & Values & Identifying variation \\\hline
@@
 \end{tabular}
-\caption{Estimated parameters and relevant variation.}
+\caption{Results of method of simulated moments: estimated parameters and variation relevant for them.}
 \end{table}
@@
 \begin{center}
-\begin{tabular}{|l|c|c|}\hline
+\begin{tabular}{ l c c }\hline\hline
 Target & Model & Data \\\hline
-\multicolumn{3}{|c|}{Marriage Dimension (source: my calculations on ACS)} \\\hline
+\multicolumn{3}{ c }{Marriage Dimension (source: my calculations on ACS)} \\\hline
 Never married by 25  & $60.6\%$ & $68.3\%$ \\
@@
 One marriage by 40  & $69.6\%$ & $64.7\%$ \\\hline
-\multicolumn{3}{|c|}{Fertility Dimension (source: my calculations on ACS)} \\\hline
+\multicolumn{3}{ c }{Fertility Dimension (source: my calculations on ACS)} \\\hline
 No kids at 25 & $65.0\%$  & $67.7\%$ \\
@@
 No kids at 40 & $23.8\%$ & $25.6\%$ \\\hline
-\multicolumn{3}{|c|}{Marriage/Fertility Interactions (source: my calculations on ACS)} \\\hline
+\multicolumn{3}{ c }{Marriage/Fertility Interactions (source: my calculations on ACS)} \\\hline
 Never married with kids at 30 & $10.4\%$  & $14.1\%$ \\
@@
 {\footnotesize Income(Divorced with kids at 30) / Income(Everyone at 30)} & $82.2\%$ & $87.0\%$\\\hline
-\multicolumn{3}{|c|}{Divorce Dimension } \\\hline
+\multicolumn{3}{ c }{Divorce Dimension } \\\hline
 Divorced by 25 &  $4.5\%$  & $10.7\%$ \\
@@
 Divorced by 40  & $16.3\%$  &    19.1\% \\\hline
-\multicolumn{3}{|c|}{K$\to$M --- M$\to$K partition (source: my calculations on ACS)} \\\hline
+\multicolumn{3}{ c }{K$\to$M --- M$\to$K partition (source: my calculations on ACS)} \\\hline
 Divorced by 40 if K$\to$M & 25.1\% &   23.5\% \\
@@
 Divorced by 40 if M$\to$K (top 30\%) & $15.2\%$ & 11.3\% \\\hline
-\multicolumn{3}{|c|}{Childcare (data: CEX and ATUS, sources in parenthesis)} \\\hline
+\multicolumn{3}{ c }{Childcare (data: CEX and ATUS, sources in parenthesis)} \\\hline
 Money for childcare /Total expenditures &$43.4\%$ & 40.0\% \cite{sommer} \\
@@
 Money spent top 30\% / Money spent low 30\% & 2.56  & 2.67 \cite{schneider} \\\hline
-\multicolumn{3}{|p{0.7\linewidth}|}{\footnotesize Note 1: shares of divorced are computed for subsample of married exactly once in both data and model.}\\
-\multicolumn{3}{|p{0.7\linewidth}|}{\footnotesize Note 2: low 30\% -- mid 40\% -- top 30\% refers to woman's per hour income.
+\multicolumn{3}{ p{0.7\linewidth} }{\footnotesize \textit{Notes:} }\\
+\multicolumn{3}{ p{0.7\linewidth} }{\footnotesize (a) shares of divorced are computed for subsample of married exactly once in both data and model.}\\
+\multicolumn{3}{ p{0.7\linewidth} }{\footnotesize (b) low 30\% -- mid 40\% -- top 30\% refers to woman's per hour income.
 The income partition is not full as income can be missing, see \ref{inc-part} and \ref{income-attr}.}\\\hline
@@
 \begin{center}
-\begin{tabular}{|l|c|c|c|}\hline
+\begin{tabular}{ l c c c }\hline\hline
 Target & Baseline & No shocks at meeting & No shocks at all \\\hline
-\multicolumn{4}{|c|}{Marriage Dimension} \\\hline
+\multicolumn{4}{ c }{Marriage Dimension} \\\hline
 Never married by 25  & $60.6\%$ & $60.7\%$ & $60.7\%$  \\
@@
 Never married by 40 & $19.0\%$ & $18.8\%$ & $18.8\%$  \\\hline
-\multicolumn{4}{|c|}{Fertility Dimension} \\\hline
+\multicolumn{4}{ c }{Fertility Dimension} \\\hline
 No kids at 25 & $65.0\%$ & $68.4\%$ & $68.9\%$ \\
@@
 No kids at 40 & $23.8\%$  &  $25.0\%$  & $25.4\%$ \\\hline
-\multicolumn{4}{|c|}{Divorce Dimension} \\\hline
+\multicolumn{4}{ c }{Divorce Dimension} \\\hline
 Divorced by 25 & $4.5\%$   &  $3.2\%$ & $2.3\%$ \\
@@
 \begin{center}\small
-\begin{tabular}{|l|c|c|c|c|}\cline{2-5}
-\multicolumn{1}{c|}{} & \multirow{2}{*}{All population} & \multicolumn{3}{c|}{Married once + have kids}\\\cline{3-5}
-\multicolumn{1}{c|}{}  & & Everyone & K$\to$M & M$\to$K \\\cline{2-5}\hline
+\begin{tabular}{ l c c c c }\cline{2-5}
+\multicolumn{1}{c}{} & \multirow{2}{*}{Everyone} & \multicolumn{3}{c}{Married once + have kids}\\\cline{3-5}
+\multicolumn{1}{c}{}  & & Everyone & Kids-first & Marriage-first \\\cline{2-5}\hline
 \textit{Age} & 30.2    &  33.1     &       31.7     &       33.6 \\
@@
 \textit{Husband's age} & 34.9        &    35.8      &      34.4   &         36.2  \\
-\textit{Age difference (W$-$H)} &  2.9       &      2.8      &       3.0      &       2.7  \\\hline\hline
+\textit{Age difference (W$-$H)} &  2.9       &      2.8      &       3.0      &       2.7  \\\hline
 \textit{Eldest child's age} & 8.6         &    7.8    &        10.0     &        7.1 \\
-\textit{Number of children} & 1.1       &      2.0      &       2.3     &       2.0 \\\hline\hline
-\textit{\% college or + (own)} & 32.1\%   &         39.5\%     &       16.9\%     &       46.3\% \\
-\textit{\% college or + (husband)} & 33.3\%     &       36.4\%     &       14.2\%       &     42.7\% \\\hline\hline
-\textit{\% In Labor Force} & 75.6\%    &        68.6\%    &        70.0\%        &    68.2\% \\
-\textit{Own income (thousands)} & 32.2 &         38.4 &          29.1 &         41.1 \\
-\textit{Hours worked} & 36.6       &     36.5    &        36.8        &    36.4\\\hline\hline
-\textit{\% white (incl hispanic)} & 69.2\%      &      76.5\%      &      72.2\%      &      77.8\% \\
-\textit{\% black} & 14.8\% &        6.7\% &            12.2\%       &       5.0\% \\\hline
-\multicolumn{5}{|p{0.7\linewidth}|}{\footnotesize ACS, women, 20--40. Means. Everything is subsamples when applicable, income and hours are for subsample of women who are employed.}\\\hline
+\textit{Number of children} & 1.1       &      2.0      &       2.3     &       2.0 \\\hline
+\textit{College or + (own), percent} & 32.1   &         39.5     &       16.9     &       46.3 \\
+\textit{College or + (husband), percent} & 33.3     &       36.4     &       14.2       &     42.7 \\\hline
+\textit{In Labor Force, percent} & 75.6    &        68.6    &        70.0        &    68.2 \\
+\textit{Own income, in \$1000} & 32.2 &         38.4 &          29.1 &         41.1 \\
+\textit{Hours worked} & 36.6       &     36.5    &        36.8        &    36.4\\\hline
+\textit{White, percent} & 69.2      &      76.5      &      72.2     &      77.8 \\
+\textit{Black, percent} & 14.8 &        6.7 &            12.2       &       5.0 \\\hline\hline
+\multicolumn{5}{p{0.8\linewidth}}{\footnotesize ACS, women, 20--40.
+Means.
+Everything is for subsamples when the characteristic is well-defined, for instance, income and hours are for subsample of women who are employed.}\\\hline
 \end{tabular}
-\caption{Differences in composition, ACS\label{diff-comp}}
+\caption{Comparison of characteristics between kids-first and marriage-first groups\label{diff-comp}}
 \end{center}
@@
 \begin{center}
-\begin{tabular}{|l|c||c|c|c|c|c|c|}\cline{2-8}
-\multicolumn{1}{c|}{} & \multicolumn{7}{c|}{ACS: sample of married once}\\\cline{2-8}
-\multicolumn{1}{c|}{} & \small Share of &\multicolumn{2}{c|}{\small Raw difference} & \multicolumn{2}{c|}{\small with controls} & \multicolumn{2}{c|}{\small controls + income} \\ \cline{3-8}
-\multicolumn{1}{c|}{} & \small divorced & $\Delta$ & (s.e.) & $\Delta$ & (s.e.) & $\Delta$ & (s.e.) \\  \hline
-\textit{All females, 20--50} & 1.94\% &  1.37\% &  (0.05\%) &  0.82\% &  (0.08\%) &  0.79\%  &  (0.11\%) \\
-\textit{All females, 35--40} &   1.59\% &  1.04\% &  (0.08\%) &   0.77\% &  (0.11\%)  &  0.74\%   &  (0.15\%) \\\hline\hline
-\multicolumn{8}{|p{0.9\linewidth}|}{\footnotesize \textbf{Education partition.}}\\\hline
-\textit{High school only, 20--50} & 2.31\% &  0.86\% &  (0.08\%) &   0.82\% &  (0.12\%)  &  0.82\%  &   (0.20\%) \\
-\textit{College graduates, 20--50} & 1.19\% &  1.24\% & (0.10\%) &   0.59\% &   (0.13\%)  &  0.50\% &    (0.16\%) \\\hline\hline
-\multicolumn{8}{|p{0.9\linewidth}|}{\footnotesize \textbf{Income partition:} working women only.}\\\hline
-\textit{Low 30\% income, 20--40} &  3.02\% &  1.31\% &  (0.14\%) &   0.90\% &  (0.21\%) &   0.91\%  &   (0.21\%) \\
-\textit{Mid 40\% income, 20--40} &  2.54\% &  1.34\% &  (0.12\%) &   0.54\%  & (0.18\%) &   0.54\% &    (0.18\%) \\
-\textit{Top 30\% income, 20--40} & 1.69\% &  1.69\% &   (0.15\%) &   0.90\%  & (0.20\%) &   0.90\% &    (0.20\%) \\
+\begin{tabular}{ l c  c c c c c c }\cline{2-8}
+\multicolumn{1}{c }{} & \multicolumn{7}{c|}{ACS: sample of married once}\\\cline{2-8}
+\multicolumn{1}{c }{} & \small Share of &\multicolumn{2}{c }{\small Raw difference} & \multicolumn{2}{c }{\small with controls} & \multicolumn{2}{c }{\small controls + income} \\ \cline{3-8}
+\multicolumn{1}{c }{} & \small divorced & $\Delta$ & (s.e.) & $\Delta$ & (s.e.) & $\Delta$ & (s.e.) \\  \hline
+\textit{All females, 20--50} & 1.94 &  1.37 &  (0.05) &  0.82 &  (0.08) &  0.79  &  (0.11) \\
+\textit{All females, 35--40} &   1.59 &  1.04 &  (0.08) &   0.77 &  (0.11)  &  0.74  &  (0.15) \\\hline\hline
+\multicolumn{8}{ p{0.9\linewidth} }{\textbf{Education partition}}\\\hline
+\textit{High school only, 20--50} & 2.31 &  0.86 &  (0.08) &   0.82 &  (0.12)  &  0.82  &   (0.20) \\
+\textit{College graduates, 20--50} & 1.19 &  1.24 & (0.10) &   0.59 &   (0.13)  &  0.50 &    (0.16) \\\hline\hline
+\multicolumn{8}{ p{0.9\linewidth} }{ \textbf{Income partition:} subsample of working women.}\\\hline
+\textit{Low 30\% income, 20--40} &  3.02 &  1.31 &  (0.14) &   0.90 &  (0.21) &   0.91  &   (0.21) \\
+\textit{Mid 40\% income, 20--40} &  2.54 &  1.34 &  (0.12) &   0.54  & (0.18) &   0.54 &    (0.18) \\
+\textit{Top 30\% income, 20--40} & 1.69 &  1.69 &   (0.15) &   0.90  & (0.20) &   0.90 &    (0.20) \\
 \hline
+\multicolumn{8}{p{0.9\linewidth}}{ \footnotesize \textit{Notes:} this is analogous to Table \ref{diff-comp}, so same definitions apply.
+Instead of being divorced by the moment of survey, this uses fact of getting a divorce \textit{at the year of survey} as an outcome.
+All numbers are in percents.}\\\hline\hline
 \end{tabular}
@@
 \centering
-\begin{tabular}{|r|r|r||r|r|}\hline
-& \multicolumn{2}{c||}{Total count} & \multicolumn{2}{c|}{Years married} \\\hline
+\begin{tabular}{ r r r  r r }\hline
+& \multicolumn{2}{c  }{Total count} & \multicolumn{2}{c }{Years married} \\\hline
  $\Delta T$ &  Own & Step & Own & Step\\\hline
@@
      All $\leq 0$ &    80,047 &    12,653 \\
-     (shares)        & ($86.4\%$)  & ($13.6\%$)\\\cline{1-3}
+     (percent)        & ($86.4$)  & ($13.6$)\\\cline{1-3}
 \end{tabular}

### 2020-02-08 main/draft.tex
Fixing the August stage

@@
 \newcommand{\deq}{\stackrel{d}{=}}
-\title{Economics of Shotgun Marriage}
+\title{The Economics of Shotgun Marriage}
 \author{Egor Kozlov\thanks{Economics Department, Northwestern University, egorkozlov2020@u.northwestern.edu.
@@
 {\small Northwestern University}}
+\date{August, 2019}
 \begin{document}
@@
 \begin{abstract}
-Many couples marry shortly after woman gets pregnant, that is often referred as a shotgun marriage.
+Many couples marry shortly after woman gets pregnant, which is often referred to as a shotgun marriage.
 I show that in American Community Survey data couples who experienced a shotgun marriage, i.e.
@@
 %[Paragraph 1: is shotgun marriage still a thing? Abstract - 2]
-Term ``shotgun marriage'' is traditionally used to describe situations when having a baby triggers marriage of a couple.
+The term ``shotgun marriage'' is traditionally used to describe situations when having a baby triggers marriage of a couple.
 This practice does not solely refer back to more conservative times: in the modern US data around one-quarter of married women of age 21--40 had their first birth before or at the year of their first marriage.
 And, although it seems that abortion and contraception access would provide people with more choice, in this paper I show that in the data such couples still look like something pushed them to marry each other.
-In particular, their marriages appear less stable over time: despite of being less than quarter of the population shotgun marriage couples account for more than one-third of divorces.
+In particular, their marriages appear less stable over time: despite of being less than a quarter of the population shotgun marriage couples account for more than one-third of divorces.
+
 This paper has two main goals.
@@
 %[Paragraph 2: general importance? Child development]
-Understanding the sources of variation in divorce rates is not just a household behavior question.
-For example, the divorces have large major impact on how people raise their children.
+Understanding the sources of a variation in divorce rates is not just a household behavior question.
+For example, divorces have major impact on how people raise their children.
 The families who experienced shotgun marriages are often overlooked by the current literature focusing on child development.
@@
 %[Paragraph 3: general importance? Other ideas]
-In addition to child development impact of unstable marriages, divorces have other important consequences.
+In addition to the child development impact of unstable marriages, divorces have other important consequences.
 For instance, they appear as an important driver of labor market choices and outcomes of women.
-Additionally, being divorced with young children is considered as lowering re-marriage prospective, and this can affect total fertility and marriage rates.
-Finally, the question of marriage selection and modeling divorces is curious by its own from prospectives of economic theory: if shotgun marriage predicts lower marriage quality and higher likelihood of divorce, we can use it as identifying variation for studying intra-household questions, that is relevant for developing realistic household models and understanding things that are not easy to observe directly.
+Additionally, being divorced with young children is considered as lowering re-marriage prospectives, and this can affect total fertility and marriage rates.
+Finally, the question of marriage selection and modeling divorces is interesting on its own from perspectives of economic theory: if shotgun marriage predicts lower marriage quality and higher likelihood of divorce, we can use it as identifying variation for studying intra-household decisions and to develop realistic household models.
 
 %[Paragraph 4: general importance? We need to understand trends]
-On more aggregate level, this paper can broaden the understanding of recent trends in household formation.
+On a more aggregate level, this paper can broaden the understanding of recent trends in household formation.
 First, around 40\% of new births in 2018 occurred by unmarried mothers.
@@
 %[Paragraph 5: empirics --- what data do I use]
-I use American Community Survey as the main dataset, as it has large sample sizes, uniform coverage and reports wide range of household and individual characteristics.
-One important drawback of it is that the marriage history in ACS is not reported perfectly, as it only has records of the year of the most recent marriage.
+I use the American Community Survey as the main dataset, as it has large sample sizes, uniform coverage and reports wide range of household and individual characteristics.
+One important drawback of it is that the marriage history in the ACS is not reported perfectly, as it only has records of the year of the most recent marriage.
 I overcome this by conditioning on women who was married exactly once.
-I then use separate dataset of Survey of Income and Program Participation to re-establish the key findings without this restriction and show that this attrition does not introduce any important distortions besides changing magnitudes of what I refer to as divorce rates.
-Although SIPP provides superior coverage of the issues I study, the current work still mainly uses ACS: SIPP sample sizes do not allow to condition on multiple dimensions like child's age.
+I then use the separate dataset of Survey of Income and Program Participation to re-establish the key findings without this restriction and show that this attrition does not introduce any important distortions besides changing magnitudes of divorce rates.
+Although SIPP provides superior coverage of the issues I study, the current work mainly uses ACS: SIPP sample sizes do not allow to condition on multiple dimensions like child's age.
 
 %[Paragraph 6: empirics --- methodological issues shortly]
-My main empirical contribution is describing association between having shotgun marriages and having a divorce, that I have not found to be explicitly discussed in the existing literature.
-Methodologically, the crucial comparison involves partitioning population of women who were married and had kids on two groups:  those who gave their first birth at least at the next year of their marriage (``marriage first'') and those who gave their first birth before or at the year of their marriage (``kids first'').
+My main empirical contribution is describing the association between having shotgun marriages and having a divorce.
+Methodologically, the crucial comparison involves partitioning population of women who were married and had kids on two groups:  those who gave their first birth the earliest during the year after their marriage (``marriage first'') and those who gave their first birth before or at the year of their marriage (``kids first'').
 I refer to the latter group as those who experienced a shotgun marriage.
-I additionally exclude from the comparisons women who had their birth more than five years before marriage, as they are increasingly likely to form a family with someone else than the child's father.
+I additionally exclude from the comparisons women who had their birth more than five years before marriage, as they are increasingly likely to form a family with someone other than the child's father.
 
 %[Paragraph 7: empirica, findings --- main results]
-The most important empirical result I show is substantially larger share of divorced people in ``kids first'' group.
-Although this difference is far from being causal, it is strong, robust to many ways of controlling for composition and is present among all subpopulations.
+The most important empirical result I show is substantially larger share of divorced people in the ``kids first'' group.
+Although this difference is not a causal, it is strong, robust to many ways of controlling for composition and is present among all subpopulations.
 Namely, chances of being divorced at the moment of survey are around twice larger for those who experienced a shotgun marriage as opposed to for those who did not ($18\%$ against $10\%$).
 This ratio reaches three for those who have college degrees, and is only around 1.25 for those who only finished a high school.
-When controlled for many confounding factors that include age, income, education and geography, as well as marriage duration and child's age, the difference in shares vanishes around a half, but remains large and statistically significant.
+When controlling for many confounding factors that include age, income, education and geography, as well as marriage duration and child's age, the difference in shares declines by about 50\%, but remains large and statistically significant.
 %Similar heterogeneity holds if I use woman's income instead of education, although defining and measuring income can be subtle in case of women with young children.
@@
 %[Paragraph 8: empirical findings --- duration, step-children, lf and child development outcomes]
-Investigating the data further leads to few additional insights.
+Investigating the data further leads to additional insights.
 First, different timing of marriage and marriage duration at the moment of survey between the two groups does not seem to be an important contributor to the difference.
 Second, the ``extra'' divorces we observe are concentrated around the time the child starts going to school.
-Third, for the cases when one can recover whether the child is step or own (it is not perfectly transparent in ACS samples), this status does not affect the marriage duration, moreover, less than fifteen percent of the couples I for which this is observed seem to have step-children as I exclude those women who had their births long ago after their marriage.
-[SIPP has better data to confirm these findings?]
+Third, for the cases when one can recover whether the child is step or own (it is not perfectly transparent in ACS samples), this status does not affect the marriage duration, moreover, less than fifteen percent of the couples for which this is observed seem to have step-children as I exclude those women who had their births long ago after their marriage.
+%[SIPP has better data to confirm these findings?]
 
@@
 These comparisons, however, contain lots of selection issues and unobservable heterogeneity, and to deal with it I set up a structural model that is capable of understanding of anatomy of the result.
-The model is based on growing branch of literature in dynamic household bargaining, and endogenizes savings, marriage and divorce decisions, fertility decisions and fertility shocks, together with the structure of childcare costs.
+The model is based on a growing branch of literature in dynamic household bargaining, and endogenizes savings, marriage and divorce decisions, fertility decisions and fertility shocks, together with the structure of childcare costs.
 The last element is crucial in generating heterogeneous response in terms of income or education: as time of wealthier parents is more expensive, unexpected childcare costs create relatively more distress for them.
@@
 The main driving force in the model is an unplanned pregnancy shock.
-Each period single people meet potential partners with some probability, and if the partner is met, couple decides whether to marry each other based on a simple bargaining procedure.
+Each period single people meet potential partners with some probability, and if the partner is met, the couple decides whether to marry each other based on a simple bargaining procedure.
 The potential matches are heterogeneous in partner's characteristics and marriage quality, the latter is modeled as idiosyncratic joint utility gain of couple from being together, that can be interpreted as a ``love shock''.
 Before the bargaining decision takes place, partners may or may not get an unplanned pregnancy shock, that proxies for unplanned children appearing in cohabitating unions as I do not model cohabitation explicitly.
-The shock interfers with bargaining: if the couple made a baby, woman risks to stay a single mother if the couple does not agree to marry.
+The shock interfers with bargaining: if the couple made a baby, the woman risks to stay a single mother if the couple does not agree to marry.
 The conjecture is that in matches that experienced the shock women agree to enter the marriages of lower quality.
@@
 %[Paragraph 10: model --- counterfactuals and speculative things]
-Additionally, the estimated model delivers few nontrivial counterfactual predictions exploiting interaction between unplanned pregnancies and marriage quality.
-First, major part of divorces that happen early in life are seen as driven by unexpected pregnancies, though divorces happening later are not.
-Second, general impact of pressure of becoming a single mother on marriage decisions is substantial, and eliminating unexpected pregnancies happening before the marriage decision takes place lowers amount of people married by 25 by [...].
+Additionally, the estimated model delivers predictions exploiting interaction between unplanned pregnancies and marriage quality.
+First, a major part of divorces that happen early in life are seen as driven by unexpected pregnancies, though divorces happening later are not.
+Second, the general impact of pressure of becoming a single mother on marriage decisions is substantial, and eliminating unexpected pregnancies happening before the marriage decision takes place lowers amount of people married by 25 by [...].
 [Last, the model predicts decline in probabilities of unexpected pregnancies since 1980: the model matching fertility / marriage / divorce profiles from Census of 1980 predicts larger chances of unplanned pregnancies among all income and age groups.]
@@
 %[Paragraph 12: what I learned about channels]
-The main causal insight is that being single mother is a major threat affecting how people make their marriage decisions, and large and unequally distributed costs of raising young children can be a driver in creation of poor quality marriages.
+The main causal insight is that being a single mother is a major threat affecting how people make their marriage decisions, and large and unequally distributed costs of raising young children can be a driver in the creation of poor quality marriages.
 This channel can be of major importance for studying welfare policies subsidizing childcare, young families and single mothers, as well as for enforcing alimony payments and child support in general.
@@
 I elaborate on these points in the following order.
-Section 1 of this paper discuss existing results and general trends that are already described by the literature.
-Section 2 discuss how I construct the dataset and timing variables, and Section 3 establishes main observations about differences in groups depending on their marriage timing, together with a number of additional considerations useful to understand the content of these differences.
-After that I explain theoretical model in Section 4 and the process of setting its parameters in Section 5.
+Section 1 of this paper discusses the existing literature.
+Section 2 discusses how I construct the dataset and timing variables, and Section 3 establishes main observations about differences in groups depending on their marriage timing, together with a number of additional considerations useful to understand the content of these differences.
+After that I explain the theoretical model in Section 4 and the process of setting its parameters in Section 5.
 Section 6 describes estimation results and counterfactual simulations implied by these results.
-Section 7 provides more general discussion and concludes.
+Section 7 provides more discussion and concludes.
 
@@
 \begin{enumerate}
-\item Women who had their first child before their first marriage (K$\to$M) are more likely to be divorced than those who had their first child in their first marriage (M$\to$K).
-\item Relative difference in the share of divorced between K$\to$M and M$\to$K is increasing with income and education
+\item Women who had their first child before their first marriage (kids-first) are more likely to be divorced than those who had their first child in their first marriage (marriage-first).
+\item Relative difference in the share of divorced between kids-first and marriage-first is increasing with income and education
 \item These results are statistically significant and robust to various methods of controlling for observable composition of the groups
@@
 
-General strategy involves dividing sample in two groups: couples who had kids before their marriage ($K\to M$) and those who had kids after their marriage ($M\to K$).
+General strategy involves dividing sample in two groups: couples who had kids before their marriage (kids-first) and those who had kids after their marriage (marriage-first).
 After this division I perform multiple comparisons of these groups: the main comparison is looking at divorce rates in several variations, few additional comparisons involve marriage inequality and outcomes of children.
@@
 where $T$ refers to moments of time. After that I define two groups (given $\Delta T$ is measured in years):
-\begin{align*}&M\to K \Leftrightarrow \Delta T \in \{1,2,...\},\\
-&K\to M \Leftrightarrow \Delta T \in \{...,-2,-1,0\},\end{align*}
+\begin{align*}&\text{Kids-first} \Leftrightarrow \Delta T \in \{1,2,...\},\\
+&\text{Marriage-first} \Leftrightarrow \Delta T \in \{...,-2,-1,0\},\end{align*}
 
@@
 
-All mean and quantile calculations are performed using person weights.
+All mean and quantile calculations are performed using person weights provided by the surveys.
 
@@
 \includegraphics[width=0.8\linewidth]{shrs.eps}
-\begin{tabular}{|p{0.65\linewidth}|}
-\hline
-\footnotesize $\Delta T = \text{Year}^{\text{first child}} - \text{Year}^{\text{first marriage}}$. Restriction: $\Delta T \in \{-5,...,10\}$\\
-\footnotesize Definition: K$\to$M if $\Delta T \in \{-5,...,0\}$. M$\to$K if $\Delta T \in \{1,...,10\}$.\\\hline
+\begin{tabular}{ p{\linewidth} }
+%\hline
+\footnotesize \emph{Notes}:``K then M'' is kids-then-marriage and refers to women whose year of marriage is 0 to 5 years after their year of the first birth  (kids-first).
+``M then K'' is marriage-then-kids and refers to women whose year of marriage is 1 to 10 years after their year of the first birth (marriage-first).
+Women outside of these two groups are excluded.
+Only women who have exactly one marriage are considered.
+Income refers to residualized per-hour income for subsample of working women, as described in Section \ref{inc-part}.
+Right panel shows share of divorced women \emph{conditional} on being in each group.
+ACS is American Community Survey, 2008--2016.\\
+%\hline
 \end{tabular}
-\caption{Share of K$\to$M couples and share of divorced, ACS\label{shrs}}
+\caption{Share of kids-first and marriage-first women and share of divorced (and not remarried) in each group, ACS data\label{shrs}}
 \end{figure}
@@
 \includegraphics[width=0.8\linewidth]{shrs-sipp.eps}
-\caption{Share of K$\to$M couples and share of ever divorced, SIPP\label{shrs-sipp}}
+\begin{tabular}{ p{\linewidth} }
+%\hline
+\footnotesize \emph{Notes}: Same definitions as Figure \ref{shrs} apply, except that all women who have at least one marriage are considered.
+SIPP is Survey of Income And Program Participation, 2014 (wave 1).\\
+%\hline
+\end{tabular}
+\caption{Share of kids-first and marriage-first women and share of ever divorced in each group, SIPP\label{shrs-sipp}}
 \end{figure}
@@
 For instance, as suggested by the table, people with lower income are more likely to get divorced, so even conditional on income/education group K$\to$M couples may be poorer with each subgroup.
-Additionally, many other factors may drive the differences, including different ages of children (people in K$\to$M group by construction have older babies after any given number of years after marriage), regional and racial differences in attitudes to divorce, age composition, possible presence of step-children in K$\to$M couples and so on.
+Additionally, many other factors may drive the differences, including different ages of children (people in kids-first group by construction have older babies after any given number of years after marriage), regional and racial differences in attitudes to divorce, age composition, possible presence of step-children in kids-first couples and so on.
 I attempt to control for few of these things and argue that they explain these differences only partially, that suggests the presence of causal mechanism that I describe.
@@
 \subsection{Adjusting for Composition\label{adcomp}}
-As people in K$\to$M and M$\to$K groups obviously have different characteristics, as in particular suggested by Figure \ref{keycomp}, I use regression techniques to see what part of difference is attributed to this composition.
-Appendix \ref{comp-diff-appendix} quantifies these differences, shortly, people in K$\to$M group are younger, less educated, have kids earlier, marry slightly earlier and have more unequal marriages.
-In this part I control for difference in observable characteristics to see what part of the difference is attributed to the fact of being in K$\to$M group.
+As people in kids-first and marriage-first groups obviously have different characteristics, as in particular suggested by Figure \ref{keycomp}, I use regression techniques to see what part of difference is attributed to this composition.
+Appendix \ref{comp-diff-appendix} quantifies these differences, shortly, people in kids-first group are younger, less educated, have kids earlier, marry slightly earlier and have more unequal marriages.
+In this part I control for difference in observable characteristics to see what part of the difference is attributed to the fact of being in kids-first group.
 Shortly, controlling for observables mitigate some part of the difference but it still remains sizable and significant.
@@
 In general, I perform the following exercise: I regress indicator of female being divorced on set of controls $X$ and dummy variable $\I(\text{K$\to$M})_i$. The specification I use is
-\[D_i = \Delta \cdot \I(\text{K$\to$M})_i + X_i\gamma + \varepsilon_i.\]
+\[\text{Divorced}_i = \Delta \cdot \text{kids-first}_i + X_i\gamma + \varepsilon_i.\]
 Replacing $X_i$ with a constant corresponds to raw mean comparison described above.
@@
 \multicolumn{8}{p{0.9\linewidth}}{ \footnotesize \textit{Notes:} here $\Delta$ is defined as coefficient in cross-sectional regression
-$\text{Divorced}_i = \Delta \cdot \text{Kids-First}_i + \gamma X_i + \varepsilon_i$,
+$\text{Divorced}_i = \Delta \cdot \text{kids-first}_i + \gamma X_i + \varepsilon_i$,
 where ``Raw difference'' refers to regression with no $X_i$ that corresponds to the difference in means in Table \ref{diff-raw}.
@@
 \begin{enumerate}
-\item In the sample I use M$\to$K women are married for longer at the moment of survey\footnote{This does not contradict the fact that K$\to$M couples marry slightly earlier as I use the sample of married women.} (so they had \emph{more} years to break up), comparing marriages of the same duration only exacerbates the difference.
-\item Children of K$\to$M women are on average older (as they are born in earlier ages).
-Divergence between K$\to$M and M$\to$K observable divorce probabilities happens around ages 5--10.
-After controlling for composition, however, I show that K$\to$M couples divorce more often even conditional on child's age.
+\item In the sample I use marriage-first women are married for longer at the moment of survey\footnote{This does not contradict the fact that kids-first couples marry slightly earlier as I use the sample of married women.} (so they had \emph{more} years to break up), comparing marriages of the same duration only exacerbates the difference.
+\item Children of kids-first women are on average older (as they are born in earlier ages).
+Divergence between kids-first and marriage-first observable divorce probabilities happens around ages 5--10.
+After controlling for composition, however, I show that kids-first couples divorce more often even conditional on child's age.
 \item To further address concerns about duration, I recompute the numbers using getting the divorce at the year of survey, corresponding to marginal divorce probability as opposed to cumulative divorce probability (being divorced at by the moment survey).
@@
 This suggests that
-\item I show that the difference is not likely to be driven by presence of stepchildren in K$\to$M families.
+\item I show that the difference is not likely to be driven by presence of stepchildren in kids-first families.
 In fact, as I restrict $\Delta T \geq -5$, share of stepchildren is relatively small, and surviving couples with stepchildren do not differ in observable marriage duration with couples without stepchildren.
@@
 \subsection{Intra-Household Allocations [TBD]}
-I document few additional facts describing the difference between K$\to$M and M$\to$K besides being divorced.
+I document few additional facts describing the difference between kids-first and marriage-first besides being divorced.
 Following Lafortune et al, 2018, I construct measure of children's grade retention based on ACS (that is an indicator if any school-age child in the household is in grade that is lower than prescribed by her/his age).
@@
 See Table \ref{intra} for the numbers.
-The exercise suggests that children on K$\to$M women repeat grades significantly more often, and this happens even in higher-income households.
+The exercise suggests that children of kids-first women repeat grades significantly more often, and this happens even in higher-income households.
 [This table is based on old sample so it is slightly inconsistent with what I do above, I will redo this in one of the future iterations.]
@@
 \footnotesize \textit{Earninigs Ratio}    &  $95.2$ & $93.9$  &  $95.8$ &  $96.4$ \\\hline
-\footnotesize \textit{Kids' Grade Retention}  & $\phantom{0}4.6$    & $\phantom{0}7.6$ &   $\phantom{0}3.3$   & $\phantom{0}5.9$ \\\hline
+\footnotesize \textit{Kids' Grade Retention}  & $\phantom{0}4.6$    & $\phantom{0}7.6$ &   $\phantom{0}3.3$   & $\phantom{0}5.9$ \\\hline\hline
 \end{tabular}
@@
 \subsection{Estimated Parameters}
-This section describes parameters estimated jointly in simulated method of moments, as well as targets that are relevant for capturing each parameter.
-Shares of divorced for K$\to$M and M$\to$K groups are used in addition to these targets, it is not obvious which parameters they identify directly, but they help to overcome possibly insufficient identifying power of the ones listed below.
+Table \ref{estimated-parameters} describes parameters estimated jointly in simulated method of moments, as well as targets that are relevant for capturing each parameter.
+Shares of divorced for kids-first and marriage-first groups are used in addition to these targets, it is not obvious which parameters they identify directly, but they help to overcome possibly insufficient identifying power of the ones listed below.
 
@@
 $(p^{\text{out}},\phi_{0,s})$ & {\footnotesize Single mothers: probability of re-entering marriage market and utility shifter} & {\small$(0.10,-3.34)$} & {\footnotesize \% of divorced with kids and never married with kids at $30$} \\\hline\hline
-$p^{\text{preg}}_{0,1,2,3}$  & {\footnotesize Arrival probabilities of unplanned pregnancies by age and $z$} & {\small$(0.67,0.13,-0.50,-0.005)$} & {\footnotesize Share of K$\to$M at 25, 35; share of K$\to$M at 30 and 40 by income groups} \\\hline
+$p^{\text{preg}}_{0,1,2,3}$  & {\footnotesize Arrival probabilities of unplanned pregnancies by age and $z$} & {\small$(0.67,0.13,-0.50,-0.005)$} & {\footnotesize Share of kids-first at 25, 35; share of kids-first at 30 and 40 by income groups} \\\hline
 \end{tabular}
-\caption{Results of method of simulated moments: estimated parameters and variation relevant for them.}
+\begin{tabular}{ p{\linewidth} }
+%\hline
+\footnotesize \emph{Notes}: parameters are estimated using simulated method of moments.
+For each combination of parameters the model is solved to obtain value and policy functions, then lifecycle of 5{,}000 female agents is simulated for the first $T = 20$ periods and the statistics is computed based on the simulated data.
+Particle swarm algorithm implemented in Matlab is used to find the combination of parameters minimizing weighted distance between model and data targets.
+\\
+\hline\hline
+\end{tabular}
+\caption{Results of method of simulated moments: estimated parameters and variation relevant for them.\label{estimated-parameters}}
 \end{table}
@@
 
-In the same way, while classifying on M$\to$K and K$\to$M I exclude women who married more than once.
+In the same way, while classifying on marriage-first and kids-first I exclude women who married more than once.
 I also exclude never married single mothers, even if they re-entered the marriage market and married someone.
-Therefore, in the model K$\to$M refers only to couples who got a pregnancy shock at the meeting ($\Delta T = 0$).
-A subtle consequence of this is that as I use K$\to$M share to discipline the pregnancy shock and assume that in existing couples probabilities of unplanned pregnancies are the same as for people who just met, inside-couple fertility may not be modeled correctly.
+Therefore, in the model kids-first refers only to couples who got a pregnancy shock at the meeting ($\Delta T = 0$).
+A subtle consequence of this is that as I use kids-first share to discipline the pregnancy shock and assume that in existing couples probabilities of unplanned pregnancies are the same as for people who just met, inside-couple fertility may not be modeled correctly.
 In the current calibration, however, majority of couple's pregnancies are by choice rather that by fertility shock\footnote{This is consistent, with, for instance, \href{https://www.guttmacher.org/fact-sheet/unintended-pregnancy-united-states}{data from Guttmacher Institute} reporting that about 4 out of 10 births are unplanned.}, and this does not seem to affect the numbers significantly, so I currently avoid introducing another parameter that adjusts intensity of pregnancy shocks for married and unmarried women.
@@
 \multicolumn{3}{ c }{Marriage Dimension (source: my calculations on ACS)} \\\hline
-Never married by 25  & $60.6\%$ & $68.3\%$ \\
-Never married by 35 & $24.6\%$ & $25.6\%$  \\
-Never married by 40 & $19.0\%$ & $18.8\%$  \\\hline
-One marriage by 40  & $69.6\%$ & $64.7\%$ \\\hline
+Never married by 25  & $60.6$ & $68.3$ \\
+Never married by 35 & $24.6$ & $25.6$  \\
+Never married by 40 & $19.0$ & $18.8$  \\\hline
+One marriage by 40  & $69.6$ & $64.7$ \\\hline
 \multicolumn{3}{ c }{Fertility Dimension (source: my calculations on ACS)} \\\hline
-No kids at 25 & $65.0\%$  & $67.7\%$ \\
-No kids at 30 (low $30\%$) & $52.1\%$  &   $44.0\%$ \\
-No kids at 30 (top $30\%$) & $56.8\%$ &   $57.8\%$ \\
-No kids at 35  &  $27.2\%$ &   $28.1\%$ \\
-No kids at 40 & $23.8\%$ & $25.6\%$ \\\hline
+No kids at 25 & $65.0$  & $67.7$ \\
+No kids at 30 (low $30\%$) & $52.1$  &   $44.0$ \\
+No kids at 30 (top $30\%$) & $56.8$ &   $57.8$ \\
+No kids at 35  &  $27.2$ &   $28.1$ \\
+No kids at 40 & $23.8$ & $25.6$ \\\hline
 \multicolumn{3}{ c }{Marriage/Fertility Interactions (source: my calculations on ACS)} \\\hline
-Never married with kids at 30 & $10.4\%$  & $14.1\%$ \\
-Divorced with kids at 30 & $4.2\%$ & $4.0\%$ \\
-{\footnotesize Income(Never mar with kids at 30) / Income(Everyone at 30)} & $86.5\%$ & $70.7\%$\\
-{\footnotesize Income(Divorced with kids at 30) / Income(Everyone at 30)} & $82.2\%$ & $87.0\%$\\\hline
-\multicolumn{3}{ c }{Divorce Dimension } \\\hline
-Divorced by 25 &  $4.5\%$  & $10.7\%$ \\
-Divorced by 30 (low $30\%$)  & $17.2\%$ &     20.8\% \\
-Divorced by 30 (top $30\%$)  & $9.0\%$ &     8.1\% \\
-Divorced by 35  & $10.8\%$ &    15.9\% \\
-Divorced by 40  & $16.3\%$  &    19.1\% \\\hline
-\multicolumn{3}{ c }{K$\to$M --- M$\to$K partition (source: my calculations on ACS)} \\\hline
-Divorced by 40 if K$\to$M & 25.1\% &   23.5\% \\
-Divorced by 40 if M$\to$K & 14.8\% &      13.0\% \\
-Divorced by 40 if K$\to$M (low 30\%) & $32.5\%$ & 28.5\% \\
-Divorced by 40 if K$\to$M (top 30\%) & $26.1\%$ & 23.3\% \\
-Divorced by 40 if M$\to$K (low 30\%) & $23.0\%$ & 18.0\% \\
-Divorced by 40 if M$\to$K (top 30\%) & $15.2\%$ & 11.3\% \\\hline
+Never married with kids at 30 & $10.4$  & $14.1$ \\
+Divorced with kids at 30 & $4.2$ & $4.0$ \\
+{\footnotesize Income(Never mar with kids at 30) / Income(Everyone at 30)} & $86.5$ & $70.7$\\
+{\footnotesize Income(Divorced with kids at 30) / Income(Everyone at 30)} & $82.2$ & $87.0$\\\hline
+\multicolumn{3}{ c }{Divorce Dimension} \\\hline
+Divorced by 25 &  $4.5$  & $10.7$ \\
+Divorced by 30 (low $30\%$)  & $17.2$ &     20.8 \\
+Divorced by 30 (top $30\%$)  & $9.0$ &   8.1 \\
+Divorced by 35  & $10.8$ &    15.9 \\
+Divorced by 40  & $16.3$  &    19.1 \\\hline
+\multicolumn{3}{ c }{kids-first --- marriage-first partition (source: my calculations on ACS)} \\\hline
+Divorced by 40 if kids-first & 25.1 &   23.5 \\
+Divorced by 40 if marriage-first & 14.8 &      13.0 \\
+Divorced by 40 if kids-first (low 30\%) & $32.5$ & 28.5 \\
+Divorced by 40 if kids-first (top 30\%) & $26.1$ & 23.3 \\
+Divorced by 40 if marriage-first (low 30\%) & $23.0$ & 18.0 \\
+Divorced by 40 if marriage-first (top 30\%) & $15.2$ & 11.3 \\\hline
 \multicolumn{3}{ c }{Childcare (data: CEX and ATUS, sources in parenthesis)} \\\hline
-Money for childcare /Total expenditures &$43.4\%$ & 40.0\% \cite{sommer} \\
-Time spent 0--5 / Time spent 6--12 & $2.07$ & 2 [\href{https://www.bls.gov/charts/american-time-use/activity-by-parent.htm}{BLS}]\\
-Time spent top 30\% / Time spent low 30\% & 1.11 & 1.19  \cite{schneider} \\
-Money spent top 30\% / Money spent low 30\% & 2.56  & 2.67 \cite{schneider} \\\hline
-\multicolumn{3}{ p{0.7\linewidth} }{\footnotesize \textit{Notes:} }\\
-\multicolumn{3}{ p{0.7\linewidth} }{\footnotesize (a) shares of divorced are computed for subsample of married exactly once in both data and model.}\\
-\multicolumn{3}{ p{0.7\linewidth} }{\footnotesize (b) low 30\% -- mid 40\% -- top 30\% refers to woman's per hour income.
+Money for childcare /Total expenditures & $43.4$ & 40.0 \cite{sommer} \\
+Time spent 0--5 / Time spent 6--12 (ratio) & $2.07$ & 2 [\href{https://www.bls.gov/charts/american-time-use/activity-by-parent.htm}{BLS}]\\
+Time spent top 30\% / Time spent low 30\% (ratio) & 1.11 & 1.19  \cite{schneider} \\
+Money spent top 30\% / Money spent low 30\% (ratio) & 2.56  & 2.67 \cite{schneider} \\\hline
+\multicolumn{3}{ p{0.8\linewidth} }{\footnotesize \textit{Notes:} }\\
+\multicolumn{3}{ p{0.8\linewidth} }{\footnotesize (a) All numbers are in percents except last three ratios}\\
+\multicolumn{3}{ p{0.8\linewidth} }{\footnotesize (b) Shares of divorced are computed for subsample of married exactly once data and model}\\
+\multicolumn{3}{ p{0.8\linewidth} }{\footnotesize (c) Low 30\% -- mid 40\% -- top 30\% refers to residualized woman's per hour income.
 The income partition is not full as income can be missing, see \ref{inc-part} and \ref{income-attr}.}\\\hline
@@
 \includegraphics[width=0.85\linewidth]{individual-av.eps}
-\vspace{-3em}
-\caption{Average impact of fertility shock at $t=25$\label{individual-av}}
+%\vspace{-3em}
+\begin{tabular}{ p{\linewidth} }
+\footnotesize \emph{Note}: I generate 5{,}000 female agents with characteristics from unconditional distribution implied by the model.
+After that, I alter their sequences of meeting shocks such that they do not meet any partners before 25 and meet a partner at 25 and all of them either became pregnant (``pregnant'' line) or did not (``not pregnant'' line).
+After 25, they draw meeting shocks from the usual distribution.
+Top panel shows evolution of marital status of this whole population.
+Middle and bottom panel explore those who married after the shock.
+Effects for ages other than 25 look similar.
+\end{tabular}
+\caption{Average impact of fertility shock at $t=25$, simulations\label{individual-av}}
 \end{figure}
@@
 \vspace{-0.5em}
+\begin{tabular}{ p{\linewidth} }
+\footnotesize \emph{Note}: This further explores the middle panel of Figure \ref{individual-av} by splitting the group of those who agreed to marry onto those who marry regardless of pregnancy status (``marry anyway'') and those who marry their particular partner only because pregnancy shock happened (``compliers'').
+``Everyone'' line is the weighted average of these tho lines, where proportions of the groups are given in the legend.
+\end{tabular}
 \caption{Decomposition of the average impact\label{ind-decomp}}
@@
 I compare key targets in Table \ref{cf-shocks}.The key insight is that fertility shocks do look as a sizable contributor to observable divorce rates, especially at early ages and for lower income couples.
-This kind of aligns with intuitive understanding of how unplanned pregnancies affect households, and the whole magnitude is roughly in line with vairation of share of K$\to$M couples.
+This kind of aligns with intuitive understanding of how unplanned pregnancies affect households, and the whole magnitude is roughly in line with vairation of share of kids-first couples.
 
@@
 There are several qualitative conclusions to be drawn.
-First, the story of K$\to$M couples experiencing unexpected fertility shocks (rather than choosing to have kids while being unmarried) aligns with the data and economic theory reasonably well.
+First, the story of kids-first couples experiencing unexpected fertility shocks (rather than choosing to have kids while being unmarried) aligns with the data and economic theory reasonably well.
 The data presents them as having lower quality marriages, and the theory incorporating bargaining provides a good rationale for why this happens.
@@
 %Second, aggregate impact of pregnancy shocks is relatively larger for younger and lower income couples, while impact of selection into marriage is more pronounced for richer households. 
-Second, the data highlight the importance of time-money substitutability: for instance, in counerfactual simulations when the kids are born being adult, K$\to$M couples still divorce more because unexpected pregnancies create distress, but the gradient in terms of income cannot be seen.
+Second, the data highlight the importance of time-money substitutability: for instance, in counerfactual simulations when the kids are born being adult, kids-first couples still divorce more because unexpected pregnancies create distress, but the gradient in terms of income cannot be seen.
 In fact, this highlights the idea that less productive couples can mitigate the distress caused by shock by substituting money with time, where higher-income couples do not have this option.
@@
 \subsection{Differences in Observables\label{comp-diff-appendix}}
-This section captures demographic and economic differences of K$\to$M and M$\to$K groups.
+This section captures demographic and economic differences of kids-first and marriage-first groups.
 Many factors could compute to the observable differences, here I quantify them and this justifies use of controls in the next part.
 I present these differences for ACS, SIPP data are not very different except smaller sample size.
-For ACS, since M$\to$K and K$\to$M are defined only for subsample of people who were married once and have children present, I also compare this subsample with more general female population.
+For ACS, since marriage-first and kids-first are defined only for subsample of women who were married once and have children present, I also compare this subsample with more general female population.
 See Table \ref{diff-contr} for the results.
@@
 
-Shortly, differences in composition are pretty drastic for income and education dimension (as suggested by different shares of K$\to$M in subgroups described above).
+Shortly, differences in composition are pretty drastic for income and education dimension (as suggested by different shares of kids-first in subgroups described above).
 Most of these things are intuitive, and this suggests importance of flexibly controlling for composition.
@@
 Group that has kids first both have higher share of divorced and faster growth of it as marriage progresses.
-Combining these two graphs allows to see that M$\to$K group are on average married for longer time at the moment of survey, but overall composition is not dramatically different.
+Combining these two graphs allows to see that marriage-first group are on average married for longer time at the moment of survey, but overall composition is not dramatically different.
 That would mean that comparing raw shares of divorced actually \emph{underestimates} the difference relative to controlling for number of years after marriage.
@@
 \subsection{Effects by Child's Age}
-First, I show how share of K$\to$M people varies with age of the eldest child.
+First, I show how share of kids-first people varies with age of the eldest child.
 I focus on subsample children are 18 or younger, as very few people have children above 18 living in the household.
-This exercise highlights a nuance in the construction of the groups: almost no one in K$\to$M group has children of age less than a year, because that would mean the same year of marriage, first birth and ACS survey.
-Additionally, since the population is trimmed at the age of 40, having kids around age of 18 means fairly low age of the first birth, that is typically associated with being in M$\to$K group.
+This exercise highlights a nuance in the construction of the groups: almost no one in kids-first group has children of age less than a year, because that would mean the same year of marriage, first birth and ACS survey.
+Additionally, since the population is trimmed at the age of 40, having kids around age of 18 means fairly low age of the first birth, that is typically associated with being in marriage-first group.
 Left panel of Figure \ref{pics-raw} clearly depicts both issues, and this suggests the importance of using controls.
@@
 The increasing pattern is a result of cumulative nature of processes: people with elder children have more periods to get divorced.
-But the difference between two probabilities is illustrative: between ages 5 and 10 size of K$\to$M group changes a little, but difference becomes very pronounced.
+But the difference between two probabilities is illustrative: between ages 5 and 10 size of kids-first group changes a little, but difference becomes very pronounced.
 
@@
 \includegraphics[width=0.45\linewidth]{div_by_age_raw.eps}
-\caption{Share of K$\to$M couples and share of divorced, no controls\label{pics-raw}}
+\caption{Share of kids-first couples and share of divorced, no controls\label{pics-raw}}
 \end{figure}
@@
 I focus on studying difference more carefully then: first, I simply run a regression
-\[D_i = \sum_{a=0}^{18} \beta_a \cdot \I(\text{Age of eldest child}_i = a) +  \sum_{a=0}^{18} \Delta_a \cdot \I(\text{Age of eldest child} = a)\cdot \I(\text{K$\to$M})_i + \varepsilon_i,\]
+\[\text{Divorced}_i = \sum_{a=0}^{18} \beta_a \cdot \I(\text{Age of eldest child}_i = a) +  \sum_{a=0}^{18} \Delta_a \cdot \I(\text{Age of eldest child} = a)\cdot \I(\text{kids-first})_i + \varepsilon_i,\]
 and plot $\Delta_a$.
@@
 Second, for a subsample of working women I run a regression with controls that were used for Table \ref{diff-contr} above (including polynomials in income),
-\[D_i = \sum_{a=0}^{18} \Delta_a \cdot \I(\text{Age of eldest child} = a)\cdot \I(\text{K$\to$M})_i + X_i\gamma + \varepsilon_i,\]
+\[\text{Divorced}_i = \sum_{a=0}^{18} \Delta_a \cdot \I(\text{Age of eldest child} = a)\cdot \I(\text{kids-first})_i + X_i\gamma + \varepsilon_i,\]
 note that these controls include dummies for the age of the eldest child.
@@
 \end{tabular}
-\caption{Counts and marriage durations (at the moment of survey) of K$\to$M households with step and own children, ACS.\label{step-1}}
+\caption{Counts and marriage durations (at the moment of survey) of kids-first households with step and own children, ACS.\label{step-1}}
 \end{table}

### 2020-02-08 quest-proposal/proposal-2019.tex
Fixing the August stage

\documentclass[12pt,letter]{article}
\usepackage[left=0.8in,right=0.8in,top=1in,bottom=1in]{geometry}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{multicol}
\usepackage{microtype}
\usepackage{euscript}
\usepackage{epsfig}
\usepackage{epstopdf}
\usepackage{mathrsfs}
\usepackage{tikz}
\newcommand{\hypo}{\mathcal{H}}            
\bibliographystyle{ieeetr}

\usepackage[flushleft]{threeparttable}

%\usepackage[cp1251]{inputenc}
\usepackage[english]{babel} 
\DeclareMathOperator{\rank}{rank}
\newcommand*{\hm}[1]{#1\nobreak\discretionary{}
            {\hbox{$\mathsurround=0pt #1$}}{}}

            \def\onepc{$^{\ast\ast}$} \def\fivepc{$^{\ast}$}
\def\tenpc{$^{\dag}$}
\def\legend{\multicolumn{4}{l}{\footnotesize{Significance levels
:\hspace{1em} $\dag$ : 10\% \hspace{1em}
$\ast$ : 5\% \hspace{1em} $\ast\ast$ : 1\% \normalsize}}}

\newcommand{\bs}[1]{\boldsymbol{#1}}  
\newcommand{\bsA}{\boldsymbol{A}}

%\setstretch{1}                         
\flushbottom                            
\righthyphenmin=2                      
\pagestyle{plain}                       
%\settimeformat{hhmmsstime}  
\widowpenalty=300                   
\clubpenalty=3000                     
\setlength{\parindent}{0em}           
\setlength{\topsep}{0pt}              
\usepackage[pdftex,unicode,colorlinks=true,urlcolor=blue]{hyperref}
\usepackage{bbm}
\renewcommand{\emptyset}{\varnothing}

\setlength{\parskip}{0.5\baselineskip plus2pt minus2pt}

\newcommand{\e}{\varepsilon}
\DeclareMathOperator*{\Argmax}{\mathrm{Argmax}}
\DeclareMathOperator*{\Argmin}{\mathrm{Argmin}}
\DeclareMathOperator*{\argmax}{\mathrm{arg\,max}}
\DeclareMathOperator*{\argmin}{\mathrm{argmin}}

\newcommand{\blp}{\mathrm{BLP}}
\DeclareMathOperator*{\plim}{\mathrm{plim}}
\DeclareMathOperator{\Max}{\mathrm{Max}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\Y}{\mathcal{Y}}
\newcommand{\Z}{\mathcal{Z}}
\renewcommand{\geqslant}{\geq}
\renewcommand{\leqslant}{\leq}
\newcommand{\p}{\bs p}
\newcommand{\y}{\bs y}
\def\dd#1#2{\frac{\partial#1}{\partial#2}}

\renewcommand{\emptyset}{\varnothing}

\DeclareMathOperator{\tr}{\mathrm{tr}}

\newcommand{\bb}{\bs \beta}
\newcommand{\X}{\bs X}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\PP}{\mathbb{P}}
\DeclareMathOperator{\V}{\mathbb{V}}
\DeclareMathOperator{\CM}{\mathbb{C}}
\renewcommand{\C}{\CM}
\DeclareMathOperator{\var}{\mathrm{var}}
\DeclareMathOperator{\cov}{\mathrm{cov}}
\DeclareMathOperator{\corr}{\mathrm{corr}}
\DeclareMathOperator{\MSE}{\mathrm{MSE}}
\DeclareMathOperator{\Bias}{\mathrm{Bias}}
\renewcommand{\P}{\PP}
\newcommand{\dsim}{\stackrel{d}{\sim}}
\newcommand{\hn}{\mathcal{H}_0}
\newcommand{\ha}{\mathcal{H}_a}
\newcommand{\thetab}{\bs \theta}
\newcommand{\pv}{\text{P-value}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\MLE}{\scriptscriptstyle MLE}
\newcommand{\LR}{\mathrm{LR}}
\newcommand{\I}{\mathbb{I}}
\newcommand{\sumin}{\sum\limits_{i=1}^n}
\newcommand{\sumti}{\sum\limits_{t=0}^\infty}
\newcommand{\hbeta}{\hat{\beta}}
\newcommand{\halpha}{\hat{\alpha}}
\newcommand{\hsigma}{\hat{\sigma}}
\newcommand{\hvar}{\widehat{\var}}
\newcommand{\hcov}{\widehat{\cov}}
\newcommand{\Q}{\mathbb{Q}}

\newcommand{\pconv}{\xrightarrow{ \ p \ }}
\newcommand{\dconv}{\xrightarrow{ \ d \ }}
\newcommand{\asconv}{\xrightarrow{ \ a.s. \ }}
\newcommand{\msconv}{\xrightarrow{ \ m.s. \ }}

\newcommand{\pic}[4][h!]{\begin{figure}[#1]

\begin{center}\includegraphics[width=#2cm]{#3}\caption{#4\label{#3}}\end{center}
\end{figure}}

%outtex
\def\onepc{$^{\ast\ast}$} \def\fivepc{$^{\ast}$}
\def\tenpc{$^{\dag}$}
\def\legend{\multicolumn{4}{l}{\footnotesize{Significance levels
:\hspace{1em} $\dag$ : 10\% \hspace{1em}
$\ast$ : 5\% \hspace{1em} $\ast\ast$ : 1\% \normalsize}}}
%end outtex

%\bibliographystyle{ieeetr}

\newcommand{\laseq}{\stackrel{\lambda\text{-a.e.}}{=}}
\renewcommand{\d}{\underline}
\renewcommand{\u}{\overline}
\newcommand{\td}{\underline{\theta}}
\newcommand{\tu}{\overline{\theta}}
%\renewcommand{\theenumi}{\alph{enumi}}
%\renewcommand{\labelenumi}{(\theenumi)}
%\renewcommand{\theenumii}{\roman{enumii}}
%\renewcommand{\labelenumii}{\theenumii.}
%\renewcommand{\theenumiii}{\arabic{enumiii}}
%\renewcommand{\labelenumiii}{\theenumiii.}
%\renewcommand{\epsilon}{\varepsilon}
\newcommand{\hneq}{\stackrel{\hn}{=}}
\newcommand{\deq}{\stackrel{d}{=}}
%\title{416-2 Final Project\\
%Reproductive Technologies, Aging and Fertility Choice (Proposal)}
%\author{Egor Kozlov}
\begin{document}

\title{Three Computational Projects on Marriage and Fertility Choice (Quest Proposal)\footnote{Advisor: Matthias Doepke.}}
\author{Egor Kozlov}
\date{\today}
\maketitle

\begin{abstract}
This briefly describes three related research projects I am requesting Quest renewal for. My main focus for the coming year is a project about economics of shotgun marriage, that is going to be the leading chapter of my dissertation. In this work I use a structural lifecycle model that aims to understand why couples who marry after they had a child divorce more often then those, who marry and have children later. The second project, that I and my colleague recently started, attempts to rationalize recent trends about increasing popularity of long-term unmarried cohabitations as opposed to marriage, that is especially  more pronounced among non-college graduates. Namely, we explore how changes in divorce laws from bilateral and unilateral and changes in assets division rules affected people's choices about whether to marry each other. Finally, I continue doing some work on exploring the effects of reproductive technologies (e.g. In Vitro Fertilization) that allow to postpone having children for several years on labor market behavior and outcomes of high-skilled women. All three project require solving large dynamic optimization models and working with large datasets, and their tasks are typically parallelisable.

%The first one that I originally requested the allocation for, explores the effects of reproductive technologies (e.g. In Vitro Fertilization) that allow to postpone having children for several years on labor market behavior and outcomes of high-skilled women. The second one, using the similar data and modeling techniques, considers how childcare costs distort marriage behavior of women in cases of unplanned pregnancies by incentivizing them to join marriages that are more likely to end up with divorce. Both of these project involve solving large dynamic optimization models and working with large datasets that is demanding both in terms of computational power and storage space.
\end{abstract}

\section*{Brief summary}

\subsection*{The Economics Of Shotgun Marriage}

This is the main focus for the current academic year, that is most likely going to be the leading chapter of my dissertation and my job market paper. Its goal is to understand why couples who marry shortly after woman gets pregnant divorce divorce significantly more often. I show this can be rationalized by endogenous selection into marriage: women who experienced unplanned pregnancy agree to enter marriages of lower quality. In addition to the divorce rates, patterns in the data including how women in shotgun marriages use their time and how often do their children repeat grades suggest the validity of the marriage quality explanation. This patterns in marriage selection can be rationalized by the structure of childcare costs, where kids require both money and time, and disproportional exposure of women to these costs, so the women agree to lower quality marriages in a response to a threat of becoming a single mother. I estimate a structural model endogenizing marriage selection and use it to argue that causal impact of unplanned fertility shock, as opposed to different composition of the groups, drives the major part of the observable difference in divorce chances. Qualitatively, this suggests that the main reason for poor performance of shotgun marriages is large, unexpected and heterogeneous childcare costs. Quantitatively, the model suggests sizable effects of risks of unplanned pregnancy on observable divorce rates, especially for young women.

This year I will be revising the current draft by adding more evidence and building more flexible model. I also present the current version on Midwest Macroeconomics Meetings of Fall 2019 at Michigan State University, November 1-3. At this stage I prepare a new iteration of the project that uses Python instead of Matlab, utilizes more precise and modern methods of approximating value functions and integration routines and using GPUs to perform optimization of possibly non-concave functions efficiently.

\subsection*{Divorce Laws and The Rise of Cohabitation}

This is a joint project with Fabio Blasutto from UC Louvain. We explore how changes in divorce laws and property division rules affected the choice of couples between marriage and unmarital cohabitation. The data part exploits data from different US states that had changed their divorce laws from mutual consent to unilateral divorce during 1970s -- 2000s, as well as had differences between title based and equitable property division. The model part heavily reuses the core things from the first project I show, adding more options for choices of couples and considering marriage and cohabitation as unions with different degree of commitment, that is mainly generated by how costly is it divorce relative to break up in unmarried couple and how asymmetric the costs of divorce are, where the asymmetry is caused by the rules of property divison. This is an early stage research and we expect to have first presentations by the end of December. Computationally it uses very similar tools to the previous project

\subsection*{Reproductive Technologies and Labor Market}
This is an ongoing project that will be a chapter of my thesis. It had three iterations in Fall 2017 and Spring 2018 and Summer 2019. It focuses on how an option to postpone fertility and mitigate infertility shocks affects the lifecycle of high-skilled women. It turns out that a model that matches key patterns in the US data does not predict substantial effects on general population, but the gains increase a lot with the level of education/skills, hence the impact for women who got advanced degrees as MD, JD, PhD can be huge. First, the reproductive technologies allow both to mitigate substantial part of earning within each career losses that are caused by almost inevitable work in career in case of having a baby (by shifting interruption later in life where it appears to be less costly). Second, that is perhaps more important, the technologies change the relative attractiveness of demanding careers, and this, for instance, can attract more women into getting PhDs, as it relaxes the pressure from biological factors for those who put high emphasis on having family. At the coming iteration, I plan to re-calibrate the model to make it focused on small sub-population of the highest skilled women, that I recently was able to get detailed data for, and this will allow to get some concrete figures about how much of recent demographic changes is already driven by  existing state in reproductive technologies and what can we expect in the future when the efficiency of them will be improving (as by now they still have large degree of uncertainty).

\section*{Computational Plans}

I use Quest for both solving theoretical models and working with data. On the modeling part, I numerically solve large finite-horizon dynamic programming models describing interaction of individuals' decision of savings, fertility, marriage and divorce. To adequately capture the phenomena I described these models have to have many dimensions of state space (i.e. several dimensions individual characteristics, like productivity and wealth), and this requires substantial amount of computational resources. Before that I was mainly using Matlab, but this year I transition my projects to Python, as it has some packages and routines not readily available for Matlab. It also allows using Numba and CuPy to efficiently use GPUs that are very suitable for most of the problems I solve. At the current stage, it takes around 40 seconds to compute my main model using Python with GPU as opposed to 2--3 minutes using Matlab with C compiled routines on CPU. Estimating the models to get appropriate parameter values, however, requires solving it hundreds of times, especially given non-smoothness of the fit function that prevents using gradient methods. For Matlab I used Particle Swarm Optimization algorithm to fit the models to the data that is easily parallelizable but pretty inefficient and using 55 cores was delivering appropriate results in 2--3 hours.

In addition to the modeling part I use the US data to find complimentary empirical evidence and to establish key stylized features that I use as goals for calibrating the models. Two main datasets I use is cross-sectional American Community Survey (from the US Census Bureau) and longitudinal National Longitudinal Survey of Youth. They can contain up to several millions of observations with thousands variables, raw file sizes sometimes exceed 10 Gb. I use Stata/MP for the most of the tasks, typical computational task takes several minutes but uses amounts of memory and disk space that I do not have on my computer.

\section*{Links}

The most recent draft of the shotgun marriage project is available \href{https://drive.google.com/open?id=18fwM9vFx5Hq0Ghr9-vG9ETaQbWT7_uXq}{here}.

A presentation on the third project is available \href{https://www.dropbox.com/s/r752m1oljk250os/springLunch.pdf}{here}.

My presentation on dissertation proposal is \href{https://www.dropbox.com/s/656nkkfvhjwzy4w/proposal.pdf}{here}.

Drafts of other papers are available upon request.

Thank you for your consideration!
\end{document}

### 2020-10-02 calibration-note/calib-note.tex
Current stage

\documentclass[12pt,letter]{article}
\usepackage[left=0.8in,right=0.8in,top=1in,bottom=1in]{geometry}
\usepackage{amsmath}
\usepackage{pdflscape}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{multicol}
\usepackage{microtype}
\usepackage{euscript}
\usepackage{epsfig}
\usepackage{epstopdf}
\usepackage{mathrsfs}
\usepackage{tikz}
\newcommand{\hypo}{\mathcal{H}}            
\bibliographystyle{ieeetr}

\usepackage[flushleft]{threeparttable}

%\usepackage[cp1251]{inputenc}
\usepackage[english]{babel} 
\DeclareMathOperator{\rank}{rank}
\newcommand*{\hm}[1]{#1\nobreak\discretionary{}
            {\hbox{$\mathsurround=0pt #1$}}{}}

            \def\onepc{$^{\ast\ast}$} \def\fivepc{$^{\ast}$}
\def\tenpc{$^{\dag}$}
\def\legend{\multicolumn{4}{l}{\footnotesize{Significance levels
:\hspace{1em} $\dag$ : 10\% \hspace{1em}
$\ast$ : 5\% \hspace{1em} $\ast\ast$ : 1\% \normalsize}}}

\newcommand{\bs}[1]{\boldsymbol{#1}}  
\newcommand{\bsA}{\boldsymbol{A}}

%\setstretch{1}                         
\flushbottom                            
\righthyphenmin=2                      
\pagestyle{plain}                       
%\settimeformat{hhmmsstime}  
\widowpenalty=300                   
\clubpenalty=3000                     
\setlength{\parindent}{0em}           
\setlength{\topsep}{0pt}              
\usepackage[pdftex,unicode,colorlinks=true,urlcolor=blue]{hyperref}
\usepackage{bbm}
\renewcommand{\emptyset}{\varnothing}

\setlength{\parskip}{0.5\baselineskip plus2pt minus2pt}

\newcommand{\e}{\varepsilon}
\DeclareMathOperator*{\Argmax}{\mathrm{Argmax}}
\DeclareMathOperator*{\Argmin}{\mathrm{Argmin}}
\DeclareMathOperator*{\argmax}{\mathrm{arg\,max}}
\DeclareMathOperator*{\argmin}{\mathrm{argmin}}

\newcommand{\blp}{\mathrm{BLP}}
\DeclareMathOperator*{\plim}{\mathrm{plim}}
\DeclareMathOperator{\Max}{\mathrm{Max}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\Y}{\mathcal{Y}}
\newcommand{\Z}{\mathcal{Z}}
\renewcommand{\geqslant}{\geq}
\renewcommand{\leqslant}{\leq}
\newcommand{\p}{\bs p}
\newcommand{\y}{\bs y}
\def\dd#1#2{\frac{\partial#1}{\partial#2}}

\renewcommand{\emptyset}{\varnothing}

\DeclareMathOperator{\tr}{\mathrm{tr}}

\newcommand{\bb}{\bs \beta}
\newcommand{\X}{\bs X}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\PP}{\mathbb{P}}
\DeclareMathOperator{\V}{\mathbb{V}}
\DeclareMathOperator{\CM}{\mathbb{C}}
\renewcommand{\C}{\CM}
\DeclareMathOperator{\var}{\mathrm{var}}
\DeclareMathOperator{\cov}{\mathrm{cov}}
\DeclareMathOperator{\corr}{\mathrm{corr}}
\DeclareMathOperator{\MSE}{\mathrm{MSE}}
\DeclareMathOperator{\Bias}{\mathrm{Bias}}
\renewcommand{\P}{\PP}
\newcommand{\dsim}{\stackrel{d}{\sim}}
\newcommand{\hn}{\mathcal{H}_0}
\newcommand{\ha}{\mathcal{H}_a}
\newcommand{\thetab}{\bs \theta}
\newcommand{\pv}{\text{P-value}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\MLE}{\scriptscriptstyle MLE}
\newcommand{\LR}{\mathrm{LR}}
\newcommand{\I}{\mathbb{I}}
\newcommand{\sumin}{\sum\limits_{i=1}^n}
\newcommand{\sumti}{\sum\limits_{t=0}^\infty}
\newcommand{\hbeta}{\hat{\beta}}
\newcommand{\halpha}{\hat{\alpha}}
\newcommand{\hsigma}{\hat{\sigma}}
\newcommand{\hvar}{\widehat{\var}}
\newcommand{\hcov}{\widehat{\cov}}
\newcommand{\Q}{\mathbb{Q}}

\newcommand{\pconv}{\xrightarrow{ \ p \ }}
\newcommand{\dconv}{\xrightarrow{ \ d \ }}
\newcommand{\asconv}{\xrightarrow{ \ a.s. \ }}
\newcommand{\msconv}{\xrightarrow{ \ m.s. \ }}

\newcommand{\pic}[4][h!]{\begin{figure}[#1]

\begin{center}\includegraphics[width=#2cm]{#3}\caption{#4\label{#3}}\end{center}
\end{figure}}

%outtex
\def\onepc{$^{\ast\ast}$} \def\fivepc{$^{\ast}$}
\def\tenpc{$^{\dag}$}
\def\legend{\multicolumn{4}{l}{\footnotesize{Significance levels
:\hspace{1em} $\dag$ : 10\% \hspace{1em}
$\ast$ : 5\% \hspace{1em} $\ast\ast$ : 1\% \normalsize}}}
%end outtex

%\bibliographystyle{ieeetr}

\newcommand{\laseq}{\stackrel{\lambda\text{-a.e.}}{=}}
\renewcommand{\d}{\underline}
\renewcommand{\u}{\overline}
\newcommand{\td}{\underline{\theta}}
\newcommand{\tu}{\overline{\theta}}
%\renewcommand{\theenumi}{\alph{enumi}}
%\renewcommand{\labelenumi}{(\theenumi)}
%\renewcommand{\theenumii}{\roman{enumii}}
%\renewcommand{\labelenumii}{\theenumii.}
%\renewcommand{\theenumiii}{\arabic{enumiii}}
%\renewcommand{\labelenumiii}{\theenumiii.}
%\renewcommand{\epsilon}{\varepsilon}
\newcommand{\hneq}{\stackrel{\hn}{=}}
\newcommand{\deq}{\stackrel{d}{=}}
%\title{416-2 Final Project\\
%Reproductive Technologies, Aging and Fertility Choice (Proposal)}
%\author{Egor Kozlov}
\begin{document}
%\begin{center}\textbf{Economics 416-1} \\ \emph{By Egor Kozlov}\end{center}
%\maketitle

%\section{Introduction}
%Not all unplanned pregnancies happen at teenage age. Among adults quite substantial share of people have kids before or at the year they marry, and this translates is associated with a great increase of share of divorced. In fact, even among college graduates 30 years old kids first females are 3.6\% as opposed to unmarried women with kids being 4.6\%. Counting only  ever married women 21--40 with kids, these women account for more than 10\% in college population and more than 25\% in general population. So the incidence of shotgun marriage is substantial.

\section{Functions}

Instantaneous utility:
\[U^{\text{no children}} = \theta^m \cdot \frac{c_m^{1-\sigma}}{1-\sigma} + \theta^f \cdot \frac{c_f^{1-\xi}}{1-\xi}.\]
\[U^{\text{with children}} = \theta^m \cdot \frac{c_m^{1-\sigma}}{1-\sigma} + \theta^f \cdot \frac{c_f^{1-\sigma}}{1-\sigma} + \phi_0 + \alpha\cdot \frac{Q^{1-\sigma}}{1-\sigma}.\]

Budget constraint (the most general) is:
\[(c_f^{1+\rho} + c_m^{1+\rho})^{\frac1{1+\rho}} + x + a' = R\cdot a + W^m_t(z_m) + l_f\cdot W^f_t(z_f).\]
\[W^i_t = \exp( z_i + \text{Trend}^i_t), \ \ \text{Trend}^i_t = \tau^i_0 + \tau^i_1 \cdot t + \tau^i_2 \cdot t^2 + \tau^i_3 \cdot t^3.\]
\[z_m' = z_m + \varepsilon^m_t, \ \ \varepsilon^m_t \sim \mathcal{N}(0,\sigma^{2}_{\varepsilon,m}), \ \ z_0 \sim \mathcal{N}(0,\sigma^{2}_{z_0,m}).\]
\[z_f' = z_f - \delta(l_f) + \varepsilon^f_t, \ \ \varepsilon^f_t \sim \mathcal{N}(0,\sigma^{2}_{\varepsilon,f}), \ \ z_0 \sim \mathcal{N}(0,\sigma^{2}_{z_0,f}).\]
\[\text{Child quality production function:} \ \ Q = \left[ x^{\lambda}  + \kappa \cdot (1-l_f)^{\lambda}\right]^{\frac1{\lambda}}.\]
\newpage
\section{Parameters}
\begin{center}
\begin{tabular}{|c|p{0.35\linewidth}|c|p{0.3\linewidth}|}\hline
 & Meaning  & Value & Identification \\\hline
\multicolumn{4}{|c|}{Utility parameters:}\\\hline
$\phi_0$ & Additive utility from having a child & 1.80 & {\footnotesize Total \% with kids} \\
$\alpha$ & Weight of child quality in utility & 0.61 & {\footnotesize steepness of age profile } \\
$\kappa$ & Relative productivity of labor & 0.84 & {\footnotesize female labor supply} \\
$\lambda$ & CES power in prod fun &  $0.7^{*}$ & {\footnotesize \textit{External:} Sommer (2016) } \\
$\xi$ & CRRA power of $Q$ in utility &  $1.5^{*}$ & {\footnotesize \textit{External:} as $\sigma$} \\
$\rho$ & RTS in partrens' consumption & $0.23^{*}$ & {\footnotesize \textit{External:} Voena (2015) } \\\hline
\multicolumn{4}{|c|}{$p^{\text{meet}}_t = p_0 + p_1\cdot (\text{Age}-30) + p_2 \cdot (\text{Age}-30)^2$}\\\hline
$p_{21}$ & Meeting probability at 21 &  $0.16$ & \\
$p_{28}$ & Meeting probability at 28 &  $0.37$ & {\footnotesize marriage rate by age} \\
$p_{35}$ & Meeting probability at 35 &  $0.39$ &  \\\hline
\multicolumn{4}{|c|}{$p^{\text{pregnant}}_t = p_0 + p_1\cdot (\text{Age}-30) + p_2\cdot (\text{Age}-30)^2$ }\\\hline
$p_{21}$ & Pregnancy probability at 21 &  $0.028$ & \\
$p_{28}$ & Pregnancy probability at 28 &  $0.031$ &  {\footnotesize shares of KF and MF couples} \\
$p_{35}$ & Pregnancy probability at 35 &  $0.009$ & \\\hline
\multicolumn{4}{|c|}{Love shock: $\psi' = \psi + \varepsilon$}\\\hline
$\sigma_\psi$ & Variance of innovation &  0.58 & {\footnotesize \% divorced by years after marriages}   \\
$\sigma_{\psi_0}$ & Relative variance of initial & $5.16\cdot \sigma_\psi$ & {\footnotesize total divorce rates} \\\hline
\multicolumn{4}{|c|}{Utility shifters:}\\\hline
$\phi_m$ & {\footnotesize Male utility lost at bargaining from if marry a single mother} &  $11.0\cdot \phi_0$ & {\footnotesize stock of divorced w and w/o kids} \\
$\phi_s$ &{\footnotesize  Utility loss at bargaining from not entering a shotgun marriage (both)} & $0.83\cdot \sigma_{\psi_0}$ & {\footnotesize KF and MF divorce difference} \\\hline
\multicolumn{4}{|c|}{${}^*$ indicates externally fixed values, all other values are SMM estimates }\\\hline
\end{tabular}
\end{center}

\section{Utility Shifters in Bargaining}
{\footnotesize
\begin{tabular}{|c||c|c||c|c||c|}\hline
 & \multicolumn{2}{|c||}{Male gets:} & \multicolumn{2}{|c||}{Female gets:} &  \\\hline
Situation & Agree & Disagree & Agree & Disagree &  NBS \\\hline
Regular match & $V^{m,c}_t(\theta)$ & $V^{m,s}_t$ & $V^{f,c}_t(\theta)$ & $V^{f,s}_t$ & $\scriptstyle \left[ V^{m,c}_t(\theta) - V^{m,s}_t\right]\times \left[ V^{f,c}_t(\theta) - V^{f,s}_t\right]$\\
Unplanned pregnancy & $V^{m,ck}_t(\theta)$ & $V^{m,s}_t - \phi_s$ & $V^{f,ck}_t(\theta)$ & $V^{f,s}_t - \phi_s$ & $\scriptstyle  \left[ V^{m,ck}_t(\theta) - V^{m,s}_t + \phi_s\right]\times \left[ V^{f,ck}_t(\theta) - V^{f,s}_t + \phi_s\right]$\\
Meeting a single mother & $V^{m,ck}_t(\theta) - \phi_m$ & $V^{m,s}_t$ & $V^{f,ck}_t(\theta)$ & $V^{f,s}_t$ & $\scriptstyle  \left[ V^{m,ck}_t(\theta) - V^{m,s}_t - \phi_r\right]\times \left[ V^{f,ck}_t(\theta) - V^{f,s}_t\right]$\\\hline
\end{tabular}
}

\newpage
\section{Moments and Fit (College Graduates)}

{
\begin{center}
\begin{tabular}{|l|l|l|}\hline
Target & Model & Data \\\hline
\multicolumn{3}{|c|}{\textbf{Hazards:}} \\\hline
Hazard of marriage by age, 22--35 & \multicolumn{2}{|c|}{see fit graphs}\\
Hazard of the first child by age, 23--35 & \multicolumn{2}{|c|}{see fit graphs}\\\hline
\multicolumn{3}{|c|}{\textbf{Transitions in couples:}} \\\hline
Divorced by years after marriage, 1--10 & \multicolumn{2}{|c|}{see fit graphs}\\
Have children by years after marriage, 1--10 & \multicolumn{2}{|c|}{see fit graphs}\\\hline
\multicolumn{3}{|c|}{\textbf{Group shares:}} \\\hline
\% Kids First in population by age, 23--35 & \multicolumn{2}{|c|}{see fit graphs}\\
\% Marriage First in population by age, 23--35 & \multicolumn{2}{|c|}{see fit graphs}\\
\% Kids First (relative), by age, 23--35 & \multicolumn{2}{|c|}{see fit graphs}\\\hline
\multicolumn{3}{|c|}{\textbf{Total stocks:}}\\\hline
Divorced at 30 if one marriage & 6.2\% & 6.9\% \\
Divorced + with kids at 30  & 2.6\% & 2.7\% \\
Never married + with kids at 30  & 2.3\% & 4.4\% \\
$>1$ marriage at 40 & 11.0\% & 12.0\% \\
\textbf{Divorced if Kids First (21--40) }&14.7\% &14.8\% \\
\textbf{Divorced if Marriage First (21--40)} &5.5\% & 5.4\% \\\hline
\multicolumn{3}{|c|}{\textbf{Household Technology:}}\\\hline
In Labor Force (30, married, kids) & 74.1\% & 74.0\% \\
Child Expenditures / Labor Earnings & 41.1\% & 40.0\% \\
\hline
\end{tabular}
\end{center}

\hspace{-1cm}\begin{tabular}{|c|c|c|}\hline
\multicolumn{3}{|c|}{\textbf{Fit Graphs:}}\\\hline
\includegraphics[scale=0.4]{haz_mar.pdf} & \includegraphics[scale=0.4]{share_kfmf.pdf} & \includegraphics[scale=0.4]{kids_yaftmar.pdf} \\\hline
\includegraphics[scale=0.4]{haz_child.pdf} & \includegraphics[scale=0.4]{share_kfrel.pdf}   & \includegraphics[scale=0.4]{div_yaftmar.pdf}  \\\hline
\end{tabular}

}

\newpage

\section{Less Nice Things (Non-Targeted Moments)}
\begin{center}
\includegraphics[scale=0.6]{men_kids_ratio.pdf} \\
\includegraphics[scale=0.6]{men_mar_ratio.pdf} \\
\includegraphics[scale=0.6]{div_kfmf.pdf} 
\end{center}

\newpage
\section{Removing The Social Stigma}
\begin{center}
\textbf{Lots of couples divorce immediately...}

\includegraphics[scale=1.0]{div_kfmf_ref.pdf} 
\end{center}

\newpage

\section{Identification Concerns}
Identification of scale of $\psi$ is not obvious.

Simple setup: 3-period model, $t \in \{0,1,2\}$ start as couple. Initial love shock is $\psi_0 \sim \mathcal{N}(0,\sigma^2_0)$, shock is $\delta \sim \mathcal{N}(0,\sigma^2_\delta)$.  Suppose $\beta = 1$ and all utility flows are fixed.

\begin{enumerate}
\item Normalize all utilities to $0$, and let males and females are perfectly symmetric. Suppose $10\%$ are divorced in period 1 and $5\%$ are divorced in period 2.

Couple divorces before entering period $2$ if $\psi_0 + \delta_1 + \delta_2 < 0$

Couple divorces before entering period $1$ if $\psi_0 + \delta_1 + \E(\psi_0 + \delta_1 + \delta_2 | \psi_0 + \delta_1 + \delta_2>0) < 0$.

Replacing $\psi_0 \to 5\cdot\psi_0$ and $\delta \to 5\cdot \delta$ do not change the inequalities, so the scale is not identifiable.

\item We need some interaction between $\psi$ and utilities to estimate size of $\psi$. Utility structure is
\[u = \theta^f\cdot \frac{c_f^{1-\sigma}}{1-\sigma} + (1-\theta^f)\cdot \frac{c_m^{1-\sigma}}{1-\sigma} + \phi_0 + \alpha \cdot \frac{\left[ x^\lambda + \kappa\cdot(1-l_f)^{\lambda}\right]^{\frac{1-\sigma}{\lambda}}}{1-\sigma} + \psi.\]

$\frac{x}{c+x + s} \approx 0.4$ identifies $\alpha$. Labor force participation identifies $\kappa$ (in fact, $\alpha$, $\kappa$. $\phi_0$ are very stable). I did quite a lot of estimations that slightly adjust the targets or add more targets without changing major things. $\sigma_{\psi_0} / \sigma_\psi$ is also always around $5$, but $\sigma_\psi$ itself is not.
\end{enumerate}

\newpage
\section{Income Process}
\begin{itemize}
\item Income data: per-hour ACS labor earnings for people who:
\begin{itemize}
\item Work full year ($\geq 48$ weeks)
\item Report ``usual hours worked'' $\geq 10$
\item Report labor income more than $\$ 3000$ per year
\item Drop top and bottom 5\% of resulted per-hour earnings
\end{itemize}
\item Trend $\tau$: cross-sectional regression of per-hour earnings on age polynomial
\item Male variances: $\sigma_{\varepsilon,m}$, $\sigma_{z_0,m}$ --- matching variance of residuals from this regression at 24 and 30. Exact match.
\item Female variances: do the same thing, get $\tilde\sigma_{\varepsilon,f}$, $\tilde\sigma_{z_0,f}$. Then set $\sigma_{z_0,f} = \tilde\sigma_{z_0,f}$ and rescale $\sigma_{z,f} = 0.9\times \sigma_{z,f}$. Result --- very close match.
\begin{center}
\begin{tabular}{|l|l|l|}\hline
\multicolumn{3}{|c|}{\textbf{Fit for the income variances}} \\
\multicolumn{3}{|c|}{ACS, college graduates, per-hour earnings} \\
\multicolumn{3}{|c|}{$W_i = \exp\{z_i + \text{Trend}_{i}\}, \ \ \ z_i' = z_i + \sigma_z\cdot \varepsilon_i$} \\\hline\hline
Target & Data & Model \\
Std of Labor Earnings at 24, male  & 0.445 & 0.420 \\
Std of Labor Earnings at 30, male & 0.452 & 0.422 \\\hline
\multicolumn{3}{|c|}{$\Rightarrow$ implied $\sigma_{z_0,m} = 0.441$, $\sigma_{z,m} = 0.033$} \\\hline\hline
Std of Labor Earnings at 24, female & 0.413 & 0.417\\
Std of Labor Earnings at 30, female & 0.425 & 0.417 \\\hline
\multicolumn{3}{|c|}{$\Rightarrow$ implied $\sigma_{z_0,f} = 0.406$, $\sigma_{z,f} = 0.9 \times 0.042$} \\\hline
\end{tabular}
\end{center}

\item \textbf{Better approach:} correct for selection.
\end{itemize}

\newpage
\section{Marriage Market}
\begin{itemize}
\item Marriage market is reduced form: every period with probability $p^{\text{meet}}_t$ single people meet single people with random but similar characteristics.
\item You only meet partners of your own age.
\item College graduates meet college graduates, high school graduates meet high school graduates.
\item Productivity $z^{\text{partner}} = \mu^i + \alpha^{i} \cdot z^{\text{own}} + \sigma^{i} \cdot \varepsilon$,  $\varepsilon \sim \mathcal{N}(0,1)$.
\item Cross-sectional regressions for females suggest $\mu^f\approx 0$, $\mu^m \approx 0.25$, $\sigma^f \approx 0.5$. Regress age-detrended per-hour wages of females of spouses for recently married couples. 
\item Assets: $\log(a^{\text{partner}}) = \mu_a + \log(a^{\text{own}} +  \sigma^{i} \cdot \varepsilon$
\item $\mu^f = 0.2$ and $\mu^m = -0.2$ \emph{so on average males have 20\% more wealth},  $\sigma^{i} = 0.1$ creates small spread around these values.
\item $\psi_0 \sim \mathcal{N}(0,\sigma^2_{\psi_0})$ is standard.
\item All random shocks are uncorrelated. Resulting wages and asset levels are correlated.
\end{itemize}

\textbf{Question:} how to simulate better approach? May use some less parametric estimates, but how to organize them... Especially assets/savings data. This does not seem crucial but I want some reasonable and simple structure.

\newpage

\section{On Dynamic Bargaining}
Imagine the bargaining problem with uncertainty ($a$, $b$ are future states)
\[V_0 = \theta^f_0u(c^f_0) + \theta^m_0 u(c^m_0) +  p_a \beta \left\{ \theta^f_{1a} u(c^f_{1a}) + \theta^m_{1a} u(c^m_{1a})\right\} +  p_b\beta \left\{ \theta^f_{1b} u(c^f_{1b}) + \theta^m_{1b} u(c^m_{1b})\right\}.\]
\[ \text{participation constraints:} \ \ u(c^f_{1a}) \geq \bar{u}^f_a, \ \ u(c^m_{1a}) \geq \bar{u}^m_a, \ \ u(c^f_{1b}) \geq \bar{u}^f_b, \ \ u(c^m_{1b}) \geq \bar{u}^m_b.\]
People rebargain after they learn if the state is $a$ or $b$.

If nothing is binding, $\theta^m$ and $\theta^f$ are unchanged, solution is $\bar{c}$ and the continuation value is
\[\bar V_0 = \theta^f_0u(\bar c^f_0) + \theta^m_0 u(\bar c^m_0) +  p_a \beta \left\{ \theta^f_{0} u(\bar c^f_{1a}) + \theta^m_{0} u(\bar c^m_{1a})\right\} +  p_b\beta \left\{ \theta^f_{0} u(\bar c^f_{1b}) + \theta^m_{1} u(\bar c^m_{1b})\right\}.\]

Suppose in state $a$ for $f$ is binding (so $\theta_a^f$ is adjusted by $\mu^f_a$) and in state $b$ constraint $m$ is binding (so $\theta_b^m$ is adjusted by $\mu^m_b$). Now
\[\tilde{V}_0 = \theta^f_0u(c^f_0) + \theta^m_0 u(c^m_0) +  p_a \beta \left\{ (\theta^f_{0} + \mu^f_a) u(c^f_{1a}) + \theta^m_{0} u(c^m_{1a})\right\} +  p_b\beta \left\{ \theta^f_{0} u(c^f_{1b}) + (\theta^m_{1}+\mu^m_b) u(c^m_{1b})\right\}.\]
So $c^i_j$ are chosen to maximize $\tilde{V}_0$.

\textit{Does this mean that binding participation constraints increase continuation value?} \textbf{No!}

$\mu$ are just Lagrange multipliers! True problem is 
\begin{align*}L_0 = \theta^f_0u(c^f_0) + \theta^m_0 u(c^m_0) +  p_a \beta \left\{ \theta^f_{0} u(c^f_{1a}) + \theta^m_{0} u(c^m_{1a})\right\} +  p_b\beta \left\{ \theta^f_{0} u(c^f_{1b}) + \theta^m_{1} u(c^m_{1b})\right\} + \\ +\beta \mu^f_a [u^f(c_{1a}) - \bar{u}^f_a] + \beta \mu^m_b [u^m(c_{1b}) - \bar{u}^m_b] .\end{align*}

Now given a sequence of consumption $\tilde{c}$ that satisfies the constraints the value of the Lagrangian is
\[\tilde{V}_0 =  \theta^f_0u(\tilde c^f_0) + \theta^m_0 u(\tilde c^m_0) +  p_a \beta \left\{ \theta^f_{0} u(\tilde c^f_{1a}) + \theta^m_{0} u(\tilde c^m_{1a})\right\} +  p_b\beta \left\{ \theta^f_{0} u(\tilde c^f_{1b}) + \theta^m_{0} u(\tilde c^m_{0b})\right\}.\]

So even though we expect $\theta$ to change in the future the continuation values should not account for this! So all future values should be computed using the same $\theta$!

If in case $b$ the constraints cannot be satisfied, we instead get

\[\hat{V}_0 =  \theta^f_0u(\hat c^f_0) + \theta^m_0 u(\hat c^m_0) +  p_a \beta \left\{ \theta^f_{0} u(\hat c^f_{1a}) + \theta^m_{0} u(\hat c^m_{1a})\right\} +  p_b\beta \left\{ \theta^f_{0} \bar{u}^f_b + \theta^m_{0} \bar{u}^m_b \right\}.\]

So the weights are $\theta^f_0$ everywhere! It also holds that
\[\hat{V}_0 =  \theta^f_0\hat{V}^f_0 + \theta^m_0\hat{V}^m_0.\]
\end{document}

### 2020-10-02 jmp-draft/new-2.tex
Current stage

\documentclass[12pt,letter]{article}
 \linespread{1.25}
\usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
\usepackage{amsmath}
\usepackage{pdflscape}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{multicol}
\usepackage{microtype}
\usepackage{euscript}
\usepackage{epsfig}
\usepackage{verbatim}
\usepackage{epstopdf}
\usepackage{mathrsfs}
\usepackage{tikz}
\newcommand{\hypo}{\mathcal{H}}            
\bibliographystyle{ieeetr}

\usepackage[flushleft]{threeparttable}

%\usepackage[cp1251]{inputenc}
\usepackage[english]{babel} 
\DeclareMathOperator{\rank}{rank}
\newcommand*{\hm}[1]{#1\nobreak\discretionary{}
            {\hbox{$\mathsurround=0pt #1$}}{}}

            \def\onepc{$^{\ast\ast}$} \def\fivepc{$^{\ast}$}
\def\tenpc{$^{\dag}$}
\def\legend{\multicolumn{4}{l}{\footnotesize{Significance levels
:\hspace{1em} $\dag$ : 10\% \hspace{1em}
$\ast$ : 5\% \hspace{1em} $\ast\ast$ : 1\% \normalsize}}}

\newcommand{\bs}[1]{\boldsymbol{#1}}  
\newcommand{\bsA}{\boldsymbol{A}}

%\setstretch{1}                         
\flushbottom                            
\righthyphenmin=2                      
\pagestyle{plain}                       
%\settimeformat{hhmmsstime}  
\widowpenalty=300                   
\clubpenalty=3000                     
\setlength{\parindent}{0em}           
\setlength{\topsep}{0pt}              
\usepackage[pdftex,unicode,colorlinks=true,urlcolor=blue]{hyperref}
\usepackage{bbm}
\renewcommand{\emptyset}{\varnothing}

\setlength{\parskip}{0.5\baselineskip plus2pt minus2pt}

\newcommand{\e}{\varepsilon}
\DeclareMathOperator*{\Argmax}{\mathrm{Argmax}}
\DeclareMathOperator*{\Argmin}{\mathrm{Argmin}}
\DeclareMathOperator*{\argmax}{\mathrm{arg\,max}}
\DeclareMathOperator*{\argmin}{\mathrm{argmin}}

\newcommand{\blp}{\mathrm{BLP}}
\DeclareMathOperator*{\plim}{\mathrm{plim}}
\DeclareMathOperator{\Max}{\mathrm{Max}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\Y}{\mathcal{Y}}
\newcommand{\Z}{\mathcal{Z}}
\renewcommand{\geqslant}{\geq}
\renewcommand{\leqslant}{\leq}
\newcommand{\p}{\bs p}
\newcommand{\y}{\bs y}
\def\dd#1#2{\frac{\partial#1}{\partial#2}}

\renewcommand{\emptyset}{\varnothing}

\DeclareMathOperator{\tr}{\mathrm{tr}}

\newcommand{\bb}{\bs \beta}
\newcommand{\X}{\bs X}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\PP}{\mathbb{P}}
\DeclareMathOperator{\V}{\mathbb{V}}
\DeclareMathOperator{\CM}{\mathbb{C}}
\renewcommand{\C}{\CM}
\DeclareMathOperator{\var}{\mathrm{var}}
\DeclareMathOperator{\cov}{\mathrm{cov}}
\DeclareMathOperator{\corr}{\mathrm{corr}}
\DeclareMathOperator{\MSE}{\mathrm{MSE}}
\DeclareMathOperator{\Bias}{\mathrm{Bias}}
\renewcommand{\P}{\PP}
\newcommand{\dsim}{\stackrel{d}{\sim}}
\newcommand{\hn}{\mathcal{H}_0}
\newcommand{\ha}{\mathcal{H}_a}
\newcommand{\thetab}{\bs \theta}
\newcommand{\pv}{\text{P-value}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\MLE}{\scriptscriptstyle MLE}
\newcommand{\LR}{\mathrm{LR}}
\newcommand{\I}{\mathbb{I}}
\newcommand{\sumin}{\sum\limits_{i=1}^n}
\newcommand{\sumti}{\sum\limits_{t=0}^\infty}
\newcommand{\hbeta}{\hat{\beta}}
\newcommand{\halpha}{\hat{\alpha}}
\newcommand{\hsigma}{\hat{\sigma}}
\newcommand{\hvar}{\widehat{\var}}
\newcommand{\hcov}{\widehat{\cov}}
\newcommand{\Q}{\mathbb{Q}}

\newcommand{\pconv}{\xrightarrow{ \ p \ }}
\newcommand{\dconv}{\xrightarrow{ \ d \ }}
\newcommand{\asconv}{\xrightarrow{ \ a.s. \ }}
\newcommand{\msconv}{\xrightarrow{ \ m.s. \ }}

\newcommand{\pic}[4][h!]{\begin{figure}[#1]

\begin{center}\includegraphics[width=#2cm]{#3}\caption{#4\label{#3}}\end{center}
\end{figure}}

%outtex
\def\onepc{$^{\ast\ast}$} \def\fivepc{$^{\ast}$}
\def\tenpc{$^{\dag}$}
\def\legend{\multicolumn{4}{l}{\footnotesize{Significance levels
:\hspace{1em} $\dag$ : 10\% \hspace{1em}
$\ast$ : 5\% \hspace{1em} $\ast\ast$ : 1\% \normalsize}}}
%end outtex

%\bibliographystyle{ieeetr}

\newcommand{\laseq}{\stackrel{\lambda\text{-a.e.}}{=}}
\renewcommand{\d}{\underline}
\renewcommand{\u}{\overline}
\newcommand{\td}{\underline{\theta}}
\newcommand{\tu}{\overline{\theta}}
%\renewcommand{\theenumi}{\alph{enumi}}
%\renewcommand{\labelenumi}{(\theenumi)}
%\renewcommand{\theenumii}{\roman{enumii}}
%\renewcommand{\labelenumii}{\theenumii.}
%\renewcommand{\theenumiii}{\arabic{enumiii}}
%\renewcommand{\labelenumiii}{\theenumiii.}
%\renewcommand{\epsilon}{\varepsilon}
\newcommand{\hneq}{\stackrel{\hn}{=}}
\newcommand{\deq}{\stackrel{d}{=}}
%\title{416-2 Final Project\\
%Reproductive Technologies, Aging and Fertility Choice (Proposal)}
%\author{Egor Kozlov}

\clubpenalty=10000
\widowpenalty=10000
\begin{document}

\section{Introduction}

Having children before or in the year of marriage substantially increases the risk of divorce, and this phenomenon cannot be easily explained by observable confounders. This indicates important bargaining friction: when pregnancy occurs, a woman may want to marry a partner who otherwise would be rejected. The resulting risky marriages, which I refer to as shotgun marriages, are therefore the marginal ones in terms of responses to economic environment and policies, including child support and visitation rights, custody allocation, alimony payments, welfare programs targeting single mothers (like AFDC — TANF in the US), divorce laws and childcare policies like a paid parental leave in general.

In this paper I develop and estimate a structural model that is capable of explaining the empirical patterns about the shotgun marriages and delivers realistic answers about reasons shotgun marriages are created, factors driving their divorce risk, and overall capability of policies to control them. In particular, I argue that in the most shotgun marriages the marginal agent affecting marriage decision is husband, therefore the practice of shotgun marriage can be viewed as a response to limited child support enforcement and limited commitment of unmarried fathers to their children in general. Quantitatively, I show that although the existing schedule of the child support in the US is welfare-improving over no schedule it all, it is likely to incentivize people to enter the shotgun marriages if establishing paternity is disproportionally easier for married parents. [Predictions with numbers TBD]

Understanding shotgun marriages helps to understand the trends in family and related inequality determination. For instance, around half of the kids in the US are currently born by unmarried mothers, nevertheless, most of the couples having kids together are married --- unmarried cohabitation with children is still not a common practice, unlike in many places over the world.\footnote{According to Payne, 2013, only about 3\% of children live in unions with unmarried cohabitating parents, as opposed to 21\% living with single mothers.} Many couples use pregnancies or births as triggers for the transition from unmarried cohabitation to marriage.  While supporting two-parent families is viewed as an anti-poverty measure, incentivizing shotgun marriages is certainly not the best thing to do. Large divorce risk facilitates the chances of the children from these marriages to end up with single mothers, which is considered to be one of the strongest predictors of poor early and adult life outcomes (Chetty, Hendren 2018 or Ginther, Pollak, 2004). Divorce itself is commonly associated with female poverty (Holden, Smock 2001). Additionally, even if an unstable couple does not end up divorcing, the larger divorce risk shapes all their decisions (Voena, 2015) and possibly hurts both the partners and the children.
%More generally, this contributes toward understanding the trends in out-of-wedlock births. I

Finally, many non-negligible interactions arise here: if a woman has a child from someone who is not the right fit for marriage it may be beneficial for both her and the child to wait some time and to find a new partner, yet both uncertainty about the future matches and large unexpected childcare costs drive her away from this decision. Therefore, in addition to obvious marriage stability determinants, I show that the marriage market, costs of childcare, and borrowing constraints to be important determinants of the selection. This justifies the usage of a structural model rather than inferring some parts of causal relationships from the data.

\newcommand{\fn}{\footnote{The dataset I use in establishing the empirical results and estimating the model is the American Community Survey (ACS) since 2009. Its benefits include large sample size and uniform coverage of the US population with a wide range of household and individual characteristics reported. For supplementary robustness exercises, I also use Survey of Income and Program Participation (SIPP). It has more detailed coverage of a few variables that I am interested in, although their small sample sizes do not allow to condition well on some variables like marriage duration.}}

The contribution of the paper is threefold. First, I am the first to document empirically large incidence and significant underperformance of shotgun marriages in the modern US data.\fn I compare two groups of women: ``kids first'' and ``marriage first''. Existing studies, like Alessina, Guiliano 2005, and Tannenbaum 2020, mainly focus on the creation of shotgun marriages and treat them in a narrow sense, focusing only on women who marry while being pregnant. I expand this sense of shotgun marriage to a simpler definition: I define kids-first women as those who had their first birth before or at the year of their first marriage. I argue that these women are reasonably close in characteristics and outcomes to those who had a ``true'' shotgun marriage, and this definition does not require knowing the precise timing of the marriage and fertility which facilitates using large population datasets. In contrast, marriage-first women are those who had their first birth at least at the year following their first marriage.

\begin{figure}[h!]
\begin{center}
\includegraphics[scale=1.0]{../slides-may-20/div_5y_by_dt.pdf}
\caption{Share of divorced in 5 years by relative timing of marriage and fertility.\label{fig1}}
\end{center}
\end{figure}

The main empirical result is that the kids-first women divorce more often: for instance, 18\% of kids-first and 10\% of marriage-first are divorced 5 years after marriage. Figure \ref{fig1} provides an idea of these divorce rates by marriage timing as well as the distribution of the timings. This difference in divorces is robust to controlling for the composition of the groups including many conventional determinants of the divorce rates, and this suggests that unobservable experiences of unplanned pregnancies and related unobservable selection on the quality of marriage play a major role here. In addition, I show that kids first college graduates divorce around 2.8 times more often than marriage first, and for those with high school or less this ratio is just around 1.3. This education gradient suggests that different economic conditions for college and non-college people play the role and provides additional information about possible origins of the difference to discipline the structural model. Finally, I show that shotgun marriages are not specific to young ages. While teenage pregnancies are studied extensively from Card, Wise 1978 to Ashcraft et al. 2013 and Rosenbaum 2019, the shotgun marriage may be viewed as a manifestation of similar phenomenon later in life.

The second contribution is an estimation of a lifecycle model of marriage, divorce, and fertility choice. The seminal papers are Mazzocco 2007, and Voena 2015, notable recent examples include Low et al 2018, Shephard 2019, Blasutto 2020. I am the first to study fertility decisions in this context and their interaction with decisions of marriage. In the model people are heterogeneous with respect to age, labor productivity and savings. In a spirit of random search, they randomly meet potential matches and form couples, which are heterogeneous additionally with respect to relationship quality and relative decision power of spouses, the latter being determined by symmetric Nash Bargaining.  Spouses cannot commit not to divorce, so their decisions are subject to the participation constraints that are determined dynamically based on their option to divorce and possibly find new partners. This exploits developments of the limited commitment literature (Kocherlakota 1996, Marcet and Marimon 2019). Married childless couples endogenously decide on their fertility timing, children bring utility and require both money and time.

A main driving force in the model is random fertility shocks happening to the partners before they make their marriage decision. When a potential couple meets, with some chance they have an unplanned child. If this happens, the couple's bargaining is affected: if the partners agree they become a (kids-first) couple with a child, otherwise the woman risks to become a single mother or to face a costly pregnancy abortion, and the man loses access to the child. Relative to the regular situation, in which upon disagreement partners just wait for their new matches this generates different selection into a marriage based on fertility timing, and the model fundamentals determine its strength and implications. In particular, as people generally like the arrival of children and being a single mother or aborting is not a desirable option, the model predicts that if the unplanned pregnancy happened some couples with inferior match quality agree to marry each other, and this later generates the higher risk of divorce for them.

I estimate the model parameters using a simulated method of moments, matching the established ACS evidence to the quantities in the simulated data. The parameters I estimate are the preference for children, the features of household technology, and the transition probabilities. The moments I match are the percentage of divorced women by different durations of marriage in kids-first and marriage-first group, the share of never married and divorced with and without kids by age, the share of couples having children by the duration of their marriage and few others. To account for the education gradient flexibly I perform two distinct estimations on two subgroups -- college graduates and high school graduates. In both subgroups the model fits very well, with few exceptions for the high school graduates where the fit is still reasonable. The estimates indicate substantial differences in the probability of unplanned pregnancy between the high school and the college groups, while for both of them it is a substantial risk of women with a monetary equivalent of 4---6 yearly incomes. [TBA: interpretation of estimates]

This also contributes to the broader literature studying the general value of marriage, which has been discussed since the work of Becker, 1981. Modern studies emphasize roles of risk-sharing, as in Lise and Yamada 2019, the general comparative advantage of being a couple as in Chiappori 1997, and shared production of public good as in Greenwood et al. 2016. While having all these mechanisms in place, my work here primarily focuses on the last one: children are the crucial part of the value that is created within a couple and therefore are first-order issues in how couples are formed. One fundamental question that is possible to address is how large is the value of marriage per se relative to the value of the possibility of the couple to have children together, and exploiting how people choose to enter inferior quality marriages for the opportunity to raise their child together seems directly relevant.

Third, I contribute to the policy literature that studies the consequences of family regulations like child support, alimony rights, custody allocation, as well as broader things like divorce, property division, and welfare policies. This has received considerable attention recently by two separate lines of literature: Tannenbaum 2020 exploits how the child support obligations affect the formation of new shotgun marriages, and Forester 2020 and Brown, Flinn 2011 explore the impact of the regulations on existing couples. My approach is complementary to both of them: I consider both formation and dissolution together, and, most importantly, show that these two sides interact non-trivially, therefore should be studied together. Some women would choose not to enter marriages if they can just collect child support and raise their children as single mothers, and this selection effect would increase the stability of remaining shotgun marriages. However, women in existing marriages would have more power and possibly would like to leave their partners, and this may create the opposite direction effect on the divorce rates.

[Preview of the policy results here]

A seminal piece of economic research regarding shotgun marriage is Akerlof, Yellen, and Katz 1996. They interpret them as commitment technology to access premarital sex and argue that contraception and abortion technologies lead to the disappearance of them. Further, welfare expansion (Neal, 2004) and changes in the supply of marriageable men (Chiappori, Oreffice, 2008) were argued to influence women's decision to enter marriages following pregnancies. Empirically, Alessina, Guiliano 2005 have shown that easier divorce increased the number of shotgun marriages created, and Tannenbaum 2020 and Rossin-Slater 2017 have shown that stricter child support enforcement reduces pregnancy-related marriages. All these findings are consistent with the story that shotgun marriage is a device to obtain the required commitment and resources for childbearing.

Section 2 documents the shotgun marriages in the data, Section 3 outlines the theoretical model, Section 4 discusses the model solution, estimation and fit, Section 5 talks about the mechanisms qualitatively through the lens of the model, Section 6 presents quantitative policy exercises and Section 7 concludes.

\section{Empirical Patterns}
%
%\begin{enumerate}
%\item ACS table
%\item Graphs by dT
%\item raw and comparable sample
%\item observables cannot explain
%\item SIPP table
%\end{enumerate}

Before presenting a model I summarize the main results about the incidence and presence of shotgun marriage. This section does not aim to establish causality, and yet I attempt to argue that there are no obvious explanation in the data for the reasons relative timing matters. First, I show the main result of marriage performance for kids-first and marriage-first couples. Second, I show that it holds on large subsamples of the data: whether I pick older or younger, or more or less educated I still get the persistent difference, although it magnitude varies. Relatedly, I show that controlling for many variables influencing the share of divorced does not invalidate the result. Third, using a supplementary sample from SIPP data I argue that the fact the in the main dataset marriage and fertility history are not recovered perfectly is not likely to drive the conclusions. Finally, I argue that with my definitions majority of kids-first couples actually have their own children as opposed to being a result of single mothers finding a new partner, though the fact of having their own children does not seem to play a major role.

I mainly use American Community Survey (ACS) samples of 2009--2017 for the main empirical exercise and the model estimation. Its advantage is large sample sizes with very precise household data, bearing enough information to recover the parts of marriage and fertility history I am interested in and to deliver very fine partitions of the data, like focusing only on people surveyed a certain number of year after their marriage. Its main disadvantage is incompleteness of the marriage and fertility histories.

To classify the couples by fertility timing, I consider the following simple measure:
\begin{equation}
\Delta T = \text{Year of the first marriage} -  \text{Year of the first birth}
\end{equation}
assuming both events happened by the time the person is observed. Based on the $\Delta T$ I call kids-first (KF) women those who have $\Delta T \leq 0$ and marriage-first (MF) those who have $\Delta T > 0$. Figure ... presents a motivating graph: the histogram shows distribution of women with given $\Delta T$ between $-5$ and $5$, and the scatter plot shows percentage of those divorced five years after marriage with this value of $\Delta T$.

To form the sample for the analysis I pick 2009--2016 ACS, pick women aged 21--40 at the moment of survey, exclude those with marital statuses ``spouse absent'', ``widowed'' and ``separated''. This is what I refer to as general population of women. Within this population I focus on females who are either married or divorced now, have children present in a household and are married no more than once. Finally, within those I focus on an 11-year window between the years of marriage and fertility $-5 \leq \Delta T \leq 5$.  The year of the first birth is computed based on the age of the eldest child, so this assumes that the eldest child still resides with the woman.\footnote{Some cases have the age difference between the mother and the child is above 14, for them the year was treated as missing.} ACS person-specific weights are used in all calculations.

The described restrictions are aimed to address the imperfections of the ACS data, which is a large cross-sectional household survey. Focusing on women allows to capture divorces more accurately, as children staying with mothers is a default custody allocation in the US. Restricting the age allows to focus on women who are likely to reside with their children. Picking only married once is a limitation of the data, as only the year of the most recent marriage is recorded. I address this selection issue in subsection .... Finally, picking a window around $\Delta T = 0$ allows to mitigate family arrangements which are likely to have step children, as well as families where the older child has moved out. Appendix ... shows that result still holds without this restriction, although observations with $\Delta T$ outside the window decreases the magnitude of the differences I discuss.

%Lag between pregnancy and childbearing is important yet not crucial for the classification. The main group of interest are people with $\Delta T = 0$. Assuming that average pregnancy lasts a little less than nine months, even with random timing less then one-quarter of these couples will be married before their pregnancy. One still can imagine a couple who married in January, got pregnant in March and gave a birth in December, but this requires very precise timing of decisions and, moreover, marrying no later than March, which is uncommon in particular due to cold season in most of the US. By similar reasoning, some share of couples with $\Delta T = 1$ did actually marry while being pregnant, and yet those people are less constrained in the time.

%A substantial share of couples has $\Delta T < 0$, meaning out-of-wedlock childbirth preceding the marriage. In most of the empirical part I pool this group with those with $\Delta T = 0$.

The top part Table \ref{share_table_0} illustrate the differences in the divorce between kids-first and marriage-first groups in general, as well as the relative proportion of the kids-first group. To measure the divorce, I focus on shares of divorced within those ever married. I first use a simple cross-sectional measure, comparing raw percentages of divorced in each of the two groups. This measure does not account for duration properly: given $\Delta T$ some couples may be married longer than others at the moment of the survey, therefore I also present the difference conditional on particular durations. The table presents the results conditional on being 5 or 10  years after the marriage, the large sample size allows to do this. Surprisingly, this conditioning does not change the difference substantially. Further, the bottom part of the table illustrates the heterogeneity by partitioning people on education groups and comparing women with earlier and later births.

\begin{table}[h]
\caption{Share of divorced among kids-first and marriage-first, ACS\label{share_table_0}}
\begin{tabular}{l r r r r }
\hline
& \multicolumn{2}{c}{share of divorced if ... }&  \\
&  marriage-first & kids-first & (share of kids-first) &  \\\hline
\multicolumn{5}{l}{\textbf{All sample}} \\\hline
\textit{Cross-sectional share of divorced} &  9.9 & 18.1 & (26.1) \\
\textit{Divorced 5 years after marriage} &  5.1 & 14.3  & (24.6) \\
\textit{Divorced 10 year after marriage} & 10.4 & 22.8 & (24.4) \\\hline
\multicolumn{5}{l}{\textbf{Cross-sectional share of divorced for subsamples}} \\\hline
\textit{High school only} &  12.8 & 17.3 & (37.0) \\
\textit{Some college} & 14.3 & 21.0 & (31.3) \\
\textit{College or more} &   5.3 & 14.8 & (11.9) \\\hline
\textit{First birth before 25} & 15.5 & 19.9 & (41.8) \\
\textit{First birth at 25 or later} &  6.2 & 10.9 & (10.7)  \\\hline
 \multicolumn{5}{p{0.9\linewidth}}{\footnotesize \textit{Notes.} This is American Community Survey data, 2009--2017. The numbers are percentages. Two left columns show percentage of divorced conditional on being in a marriage-first or kids-first group, respectively. The right column shows relative percentage of kids-first in those who belong to either group. Kids-first refers to women who have their first child before or at the year of marriage, marriage first to those who have their first child at least at the following year. Everything is conditional on being married once, see the text for precise definitions. }\\\hline\hline
\end{tabular}
\end{table}

Few patterns are worth noting. First, people in kids-first group systematically have higher divorce rates. Second, this difference is more pronounced for college graduates, despite the smaller yet sizable share of the kids-first group for them: with comparable absolute difference marriage-first group is relatively much more stable. Third, although very related with the previous one, the difference is more visible for women who give a birth later rather than earlier. Fourth, heterogeneity in shares of kids-first and shares of divorced by subgroups suggests that composition is an important factor.

As the divorced rates are heterogeneous within the population, the ratio of shares of divorced is more informative than the difference. In particular, the table suggests that the divorce rate for kids-first high school graduates is 1.4 times higher than for the marriage-first, and this ratio is 2.8 for college graduates.

Table \ref{share_table_1} uses a flexible linear regression to control for the composition. The specification is
\[\text{Divorced}_i =\Delta\cdot \text{KF}_i + \gamma \cdot X_i + \varepsilon_i\]
where $X_i$ represents possible controls. Raw difference in the divorce rate corresponds to $\hat\Delta$ when $X_i$ contains only constant.

I employ three sets of controls: individual characteristics include dummy variables age and education interacted, as well as race, fixed effects of state, metropolitan area status and survey year. Duration controls include dummies for all interactions of age with age of the first marriage and age of the first birth. Finally, income controls are a 3rd degree polynomial of log income on a subsample of women who are employed, have non-missing income data, work at least 10 hours per week and report labor earnings more than \$3000 a year. To supplement this result I present a matching estimator based on a propensity score, that is an implied probability to be in a kids-first group predicted by the same set of controls (excluding duration controls, as the predict the treatment perfectly by definition).

These regression results are subject to an important limitation: as kids-first and marriage-first couples differ in their marriage and birth timing, the common support is violated --- it is not possible to pick people from two groups with identical timing. This motivates exclusion of duration when predicting propensity score, and this also generally hurts the statistical interpretation of all the differences in divorce, nevertheless, its large magnitude and robustness does suggest that this is not solely an artifact of the data structure.

\begin{table}[h]
\caption{Difference in share of divorced, regression and matchin\label{share_table_1}}
\begin{tabular}{l r r r r }
\multicolumn{5}{c}{\textit{Regression equation:} $\text{Divorced}_i = \Delta \cdot \text{kids-first}_i + \text{Controls}_i + \varepsilon_i$} \\\hline
\hline
& \multicolumn{2}{c}{Estimates} &  & \\
&  Difference ($\Delta$)  & (standard error) & Ratio $= \frac{\text{div if KF}}{\text{div if MF}}$ &  \\\hline
\multicolumn{5}{l}{\textbf{All sample, regression}} \\\hline
\textit{No controls (raw difference) }& 8.2 & (0.1) & 1.83 \\
\textit{Demographic controls }& 5.7 & (0.1) & 1.58 \\
\textit{Duration controls} &  5.8 & (0.2) & 1.59 \\
\textit{Demographic + duration} &  4.8 & (0.2) & 1.48 \\
\textit{Demographic + duration + income} & 4.1 & (0.3) & 1.41 \\\hline
\multicolumn{5}{l}{\textbf{All sample, propensity score matching}} \\\hline
\textit{Demographic controls} & 5.0 & (0.3) & 1.50 \\
\textit{Demographic controls + income} & 4.9 & (0.2) & 1.49 \\\hline
\multicolumn{5}{p{\linewidth}}{\footnotesize \textit{Notes.} This is American Community Survey data, 2009--2017. The numbers are regression estimates in percents.  $\Delta$ corresponds to the difference in share of divorced between kids first and marriage first. Ratio is the ratio of the percentages of divorced between kids-first and marriage first, in a context of regression it is defined as $1 + \frac{\hat{\Delta}}{\text{div if MF}}$. Regression with income controls is for subsample of those whose income data is non-missing and of good quality, see the text for precise definitions. Propensity score matching is done using default Stata \texttt{psmatch2} options, interactions of the controls were not included due to computational reasons.}\\\hline\hline
\end{tabular}
\end{table}

This evidence suggests that although composition plays a role, regardless of observable characteristics kids-first women are divorced more often than marriage first. Marriage duration and demographics only explains less than half of these differences.

The following subsections and Appendix ... provide few extra checks and additional insights. Shortly, using more detailed and more complete data does not change the patterns I discuss here.

\subsection{Full Marital History (SIPP) \label{sipp-results}}
This section provides supplementary evidence using Survey of Income and Program Participation data, Wave 2014. The main distinct feature is concrete recording of the year of the first marriage and the year of the first birth for every woman. This allows to relax two sampling restrictions: conditioning on women ever married instead of married once (counting remarried) and correctly counting women with non-residential children (including women older than 40). In the dimensions excluding those two things the methodology is identical, including the use of $\pm 5$ year window for $\Delta T$.

\begin{table}[h]
\caption{Share of divorced among kids-first and marriage-first, SIPP\label{share_table_0}}
\begin{center}
\begin{tabular}{l r r r r }
\hline
& \multicolumn{2}{c}{share of divorced if ... }&  \\
&  marriage-first & kids-first & (share of kids-first) &  \\\hline
\multicolumn{5}{l}{\textbf{All sample, cross-sectional shares}} \\\hline
\textit{Ever divorced} & 34.4 & 48.0 & (21.0) \\
\textit{Divorced if married once} &  12.4 & 19.3 & \\
\textit{Remarried} &  22.0 & 28.7 &  \\
\multicolumn{5}{l}{\textbf{Cross-sectional share of ever divorced for subsamples}} \\\hline
\textit{High school only} &  37.5 & 46.3 & (26.4) \\
\textit{Some college} & 41.6 & 52.0 & (22.0) \\
\textit{College or more} &   24.6 & 45.8 & (12.7)\\\hline
\textit{First birth before 25} & 45.1 & 53.1 & (28.2) \\
\textit{First birth at 25 or later} &  20.7 & 23.4 & (9.4)  \\\hline
\multicolumn{5}{p{0.8\linewidth}}{\footnotesize \textit{Notes.} This is Survey of Income and Program Participation data, cross-section of Wave 1, 2014. The numbers are percentages. Two left columns show percentage of divorced conditional on being in a marriage-first or kids-first group, respectively. The right column shows relative percentage of kids-first in those who belong to either group. Kids-first refers to women who have their first child before or at the year of marriage, marriage first to those who have their first child at least at the following year. }\\\hline\hline
\end{tabular}
\end{center}
\end{table}

\subsection{Finer Partitions of the Data}
This part supplements the main result by digging dipper into the composition of the kids-first group. Shortly, regardless of who is considered, the qualitative result that kids-first people have higher divorce rates hold, although quantitatively it may differ quite a lot.

\subsubsection{Partition by $\Delta T$ (ACS)}
In the main exercise people with $\Delta T \leq 0$ were pooled together. This parts aims to see how people with $\Delta T = 0$ (having fertility and marriage at the same year) and $\Delta T < 0$ (fertility before marriage) are different. Table \ref{dt_table} shows the first set of results, doing the comparison with and without controls. It shows that regardless of controlling strategy women who have kids at the very year of marriage have somewhat higher divorce rates than those who married at the following years, although all of them divorce more than marriage-first women.

\begin{table}[h]
\caption{Difference in share of divorced, by different $\Delta T$\label{dt_table}}
\begin{tabular}{l r r r }
\multicolumn{4}{c}{\textit{Regression equation:} $\text{Divorced}_i = \Delta_0 \cdot \I[\Delta T = 0] + \Delta_{<} \cdot \I[\Delta T < 0]  + \text{Controls}_i + \varepsilon_i$} \\\hline
\hline
& \multicolumn{2}{c}{Estimates} &   \\
& Kids-first, at ($\Delta_0$)  & Kids-first, before ($\Delta_<$)  \\\hline
\textit{No controls (raw difference) }&  8.6 &  8.0  \\
\textit{Demographic controls }& 6.8 &  5.2 \\ 
\textit{Duration controls} &  6.2 &  5.1 \\ 
\textit{Demographic + duration} &  5.2 &  3.9 \\
\textit{Demographic + duration + income} & 5.0 &  2.3  \\\hline
\multicolumn{4}{p{\linewidth}}{\footnotesize \textit{Notes.} ACS, 2009--2017. The numbers are regression estimates in percents.  The estimates indicate the difference in share of divorced relative to the marriage-first women}\\\hline\hline
\end{tabular}
\end{table}

Finally, I show the graphs with perecentage of divorced for each $\Delta T$ on Figure \ref{fig2}, which is an extended version of Figure \ref{fig1}. Namely, I residualize the share of divorced using the same set of controls as in the tables before (Duration + Demographic), and normalize the share at $\Delta T = 0$ to zero. [I have the version with ``5 years after marriage'' instead of ``cross-section'', it does look a bit better but seems repetitive.]

The graph shows that $\Delta T = 0$ generally has the highest share of divorced, either with or without controlling. The largest difference is between people with $\Delta T = 0$ and $\Delta T \in \{2,3,4\}$, and it drives most of the result, where for other points the difference is less striking or even reversed, although their weight (indicated by height of the bars) is lower, and therefore they do not overturn the main result.

\begin{figure}[h!]
\begin{center}
\includegraphics[scale=1.0]{dt_jul30.eps}
\caption{Share of divorced in 5 years by relative timing of marriage and fertility.\label{fig2}}
\end{center}
\end{figure}

\subsubsection{Multi-Partner Women (SIPP)}
Although the use of $\pm 5$ years window is dedicated to reduce the concern of people having step children, it does not eliminate it completely. Within the kids first group, one may desire to consider women marrying the father of her first child and women marrying someone else than the child father separately. An important issue of most of the surveys is that relation of the male to the child is clear if the male is present in the household, but once the couple divorces and the male is not present then only few sources provide information on retrospective relationship. This issue is not possible to overcome directly in ACS, however it is possible to handle using SIPP.

For a subsample of women having more than one child SIPP asks the question about having children from multiple partners.Therefore I introduce a partition on three groups. Marriage-first women are those having $\Delta T > 0$ as before. Kids-first-own are those with $\Delta T \leq 0$ who (1) do not report to have multi-partner fertility and (2) have had children born both before and after their first marriage. Kids-first-other are those who do not meet these criteria. In the data, more than three quarters of this group do report multipartner fertility. Due to the nature of the multipartner fertility questions I also exclude women who are married more than once --- most of them are likely to be missclassified to the ``other'' group. To repeat the strategy from the ACS part, I also restrict attention to women 40 or younger: as many people remarry later, the attrition will introduce bias considering only married once.

Table \ref{fine_table} shows the distribution of the females 21--40 and older conditional on having more than one child and being married once by the three groups, as well as share of divorced in each of them. Three important conclusions are (1) majority of the kids-first group seem to have their own children, (2) the percentage of divorced is higher for both subgroups of kids-first, (3) people with multipartner fertility have higher divorce rates, but their share is relatively smaller.

\begin{table}[h!]
\caption{Finer partition of kids-first, SIPP\label{fine_table}}
\begin{center}
\begin{tabular}{l r r}
\hline
& share in sample & share of divorced  \\\hline
\multicolumn{3}{l}{\textbf{Women 21--40, 1 marriage, 2+ children}} \\\hline
\textit{Marriage-first} & 72.1 & 10.5 \\
\textit{Kids-first-own} &  16.4 & 14.6 \\
\textit{Kids-first-other} &  11.4 & 27.2 \\\hline
\multicolumn{3}{p{0.6\linewidth}}{ \footnotesize \textit{Notes.} This is Survey of Income and Program Participation data, cross-section of Wave 1, 2014. The numbers are percentages. Kids-first-own is a subset of kids-first for which it is verifiable that woman married a father of the first child, kids-first-other includes all other cases.} \\\hline
\end{tabular}
\end{center}
\end{table}

\newpage
\appendix

\section{Additional ACS Evidence}
\subsection{Flow-Based Divorce Rates}
Demographic statistics (see ... for an example) typically defines divorce rates as number of divorces by $1{,}000$ of women, either all or ever married, that happen within a particular year. This measure is different from mine: I focus on a cumulative measure of share divorced, and the divorce rate is a flow measure of the new divorced. Its advantage though that is is more robust to composition.

Table ... shows the divorce rates: number of divorces in KF group (divorces per 1{,}000 KF women), in MF group (divorces per 1{,}000 MF women), as well as KF-divorces and MF-divorces per 1{,}000 women in general population. Switching the measure does not provide particularly different insights.

\subsection{Changing the $\Delta T$ window}
Here I show the results excluding restriction $-5\leq \Delta T \leq 5$. Many people with $\Delta T > 5$ are divorced, which perhaps indicate that they might have divorced before having the child, and this can bias the judgement of the marital stability, however qualitatively the pattern still holds. See ....

\subsection{Step or Own Children in ACS}
For existing marriages ACS sometimes allows to capture the relation of children to the father. The requirement for this is that the male is the household head, and the exact criterion for it is not clear, however in two-person households ...\% have male heads. Focusing on them, it is easy to recover the percentage of stepchildren in surviving marriages by $\Delta T$. As we can see from the left panel of Table ..., as $\Delta T$ approaching zero, the share of households with stepchildren decreases, and it never exceeds ...\%.

Focusing on surviving marriages is restrictive, as it is possible that marriages with step children live shorter and drive the difference in divorce rate. If this is the case, then one would expect that the observable duration of a surviving marriage with step children will be shorter. Therefore the right panel of Table ... shows the average duration of an observable marriage. It suggests the differences in duration are almost absent, and therefore it is not likely that the marriages with step children have drastically different divorce patterns, which is another validation of the results from SIPP.
\end{document}

### 2020-10-02 jmp-draft/new-3.tex
Current stage

\documentclass[12pt,letter]{article}
 \linespread{1.25}
\usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
\usepackage{amsmath}
\usepackage{pdflscape}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{multicol}
\usepackage{microtype}
\usepackage{euscript}
\usepackage{epsfig}
\usepackage{verbatim}
\usepackage{epstopdf}
\usepackage{mathrsfs}
\usepackage{tikz}
\newcommand{\hypo}{\mathcal{H}}            
\bibliographystyle{ieeetr}

\usepackage[flushleft]{threeparttable}

\usetikzlibrary{shapes,arrows}
\usetikzlibrary{positioning}
\tikzstyle{block} = [rectangle, draw, rounded corners]
\tikzstyle{line} = [draw, -latex']

%\usepackage[cp1251]{inputenc}
\usepackage[english]{babel} 
\DeclareMathOperator{\rank}{rank}
\newcommand*{\hm}[1]{#1\nobreak\discretionary{}
            {\hbox{$\mathsurround=0pt #1$}}{}}

            \def\onepc{$^{\ast\ast}$} \def\fivepc{$^{\ast}$}
\def\tenpc{$^{\dag}$}
\def\legend{\multicolumn{4}{l}{\footnotesize{Significance levels
:\hspace{1em} $\dag$ : 10\% \hspace{1em}
$\ast$ : 5\% \hspace{1em} $\ast\ast$ : 1\% \normalsize}}}

\newcommand{\bs}[1]{\boldsymbol{#1}}  
\newcommand{\bsA}{\boldsymbol{A}}

%\setstretch{1}                         
\flushbottom                            
\righthyphenmin=2                      
\pagestyle{plain}                       
%\settimeformat{hhmmsstime}  
\widowpenalty=300                   
\clubpenalty=3000                     
\setlength{\parindent}{0em}           
\setlength{\topsep}{0pt}              
\usepackage[pdftex,unicode,colorlinks=true,urlcolor=blue]{hyperref}
\usepackage{bbm}
\renewcommand{\emptyset}{\varnothing}

\setlength{\parskip}{0.5\baselineskip plus2pt minus2pt}

\newcommand{\e}{\varepsilon}
\DeclareMathOperator*{\Argmax}{\mathrm{Argmax}}
\DeclareMathOperator*{\Argmin}{\mathrm{Argmin}}
\DeclareMathOperator*{\argmax}{\mathrm{arg\,max}}
\DeclareMathOperator*{\argmin}{\mathrm{argmin}}

\newcommand{\blp}{\mathrm{BLP}}
\DeclareMathOperator*{\plim}{\mathrm{plim}}
\DeclareMathOperator{\Max}{\mathrm{Max}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\Y}{\mathcal{Y}}
\newcommand{\Z}{\mathcal{Z}}
\renewcommand{\geqslant}{\geq}
\renewcommand{\leqslant}{\leq}
\newcommand{\p}{\bs p}
\newcommand{\y}{\bs y}
\def\dd#1#2{\frac{\partial#1}{\partial#2}}

\renewcommand{\emptyset}{\varnothing}

\DeclareMathOperator{\tr}{\mathrm{tr}}

\newcommand{\bb}{\bs \beta}
\newcommand{\X}{\bs X}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\PP}{\mathbb{P}}
\DeclareMathOperator{\V}{\mathbb{V}}
\DeclareMathOperator{\CM}{\mathbb{C}}
\renewcommand{\C}{\CM}
\DeclareMathOperator{\var}{\mathrm{var}}
\DeclareMathOperator{\cov}{\mathrm{cov}}
\DeclareMathOperator{\corr}{\mathrm{corr}}
\DeclareMathOperator{\MSE}{\mathrm{MSE}}
\DeclareMathOperator{\Bias}{\mathrm{Bias}}
\renewcommand{\P}{\PP}
\newcommand{\dsim}{\stackrel{d}{\sim}}
\newcommand{\hn}{\mathcal{H}_0}
\newcommand{\ha}{\mathcal{H}_a}
\newcommand{\thetab}{\bs \theta}
\newcommand{\pv}{\text{P-value}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\MLE}{\scriptscriptstyle MLE}
\newcommand{\LR}{\mathrm{LR}}
\newcommand{\I}{\mathbb{I}}
\newcommand{\sumin}{\sum\limits_{i=1}^n}
\newcommand{\sumti}{\sum\limits_{t=0}^\infty}
\newcommand{\hbeta}{\hat{\beta}}
\newcommand{\halpha}{\hat{\alpha}}
\newcommand{\hsigma}{\hat{\sigma}}
\newcommand{\hvar}{\widehat{\var}}
\newcommand{\hcov}{\widehat{\cov}}
\newcommand{\Q}{\mathbb{Q}}

\newcommand{\pconv}{\xrightarrow{ \ p \ }}
\newcommand{\dconv}{\xrightarrow{ \ d \ }}
\newcommand{\asconv}{\xrightarrow{ \ a.s. \ }}
\newcommand{\msconv}{\xrightarrow{ \ m.s. \ }}

\newcommand{\pic}[4][h!]{\begin{figure}[#1]

\begin{center}\includegraphics[width=#2cm]{#3}\caption{#4\label{#3}}\end{center}
\end{figure}}

%outtex
\def\onepc{$^{\ast\ast}$} \def\fivepc{$^{\ast}$}
\def\tenpc{$^{\dag}$}
\def\legend{\multicolumn{4}{l}{\footnotesize{Significance levels
:\hspace{1em} $\dag$ : 10\% \hspace{1em}
$\ast$ : 5\% \hspace{1em} $\ast\ast$ : 1\% \normalsize}}}
%end outtex

%\bibliographystyle{ieeetr}

\newcommand{\laseq}{\stackrel{\lambda\text{-a.e.}}{=}}
\renewcommand{\d}{\underline}
\renewcommand{\u}{\overline}
\newcommand{\td}{\underline{\theta}}
\newcommand{\tu}{\overline{\theta}}
%\renewcommand{\theenumi}{\alph{enumi}}
%\renewcommand{\labelenumi}{(\theenumi)}
%\renewcommand{\theenumii}{\roman{enumii}}
%\renewcommand{\labelenumii}{\theenumii.}
%\renewcommand{\theenumiii}{\arabic{enumiii}}
%\renewcommand{\labelenumiii}{\theenumiii.}
%\renewcommand{\epsilon}{\varepsilon}
\newcommand{\hneq}{\stackrel{\hn}{=}}
\newcommand{\deq}{\stackrel{d}{=}}

\clubpenalty=10000
\widowpenalty=10000
\begin{document}

\title{The Economics of Shotgun Marriage}
\author{Egor Kozlov\thanks{Economics Department, Northwestern University, egorkozlov2020@u.northwestern.edu. I thank my advisors Matthias Doepke, Martí Mestieri, and Alessandra Voena, for a fantastic amount of help and support as part of my dissertation committee. I thank my colleagues Jane Olmstead-Rumsey, Kristina Manysheva, Bence Bardoczy, Stephanie Johnson, and Gabriela Cugat, as well as all the participants of Macro Lunch at Northwestern, Midwest Macro Meetings, and H2D2 Research Day for lots of helpful comments. I owe special thanks to Fabio Blasutto for tons of help with both technical and substantial parts during his stay at Northwestern and after.}\\[1.5cm] Job Market Paper \\ \href{http://egorkozlov.link/jmp-kozlov.pdf}{Latest version available here}}
\maketitle

\begin{abstract}
Many couples marry either just before or soon after they have their first child. I show that married couples who have the first child before or in the year of marriage (kids-first) divorce around twice more often than those having their first kids in the year following their marriage or later (marriage-first). Various well-known determinants of divorce do not explain this difference. I show that it is consistent with a simple setup where people choose whether to marry based on their potential relationship quality. Unplanned pregnancies can affect their decisions as women face a risk of raising the child alone. I build and estimate a lifecycle model capable of quantifying this selection pattern. Through the model's lens, around forty percent of kids-first couples marry \textit{because} they have a child. However, a natural response of forcing fathers to pay child support has a very mild impact on these couples' marriage and divorce decisions. In contrast, improving people's ability to control their fertility provides the highest return. For existing marriages, delaying or limiting fertility is a response to the risk of divorce.
\end{abstract}

\newpage
\section{Introduction}
Marriage is associated with a range of desirable outcomes. Controlling for age and education, married couples are wealthier on a per-capita basis than never married or divorced. Marriage provides spousal insurance and efficient cooperation within the household. The benefits are especially apparent when it comes to children: married two-parent families are considered the best in terms of early and adult-life outcomes of children relative to never married or divorced parents (Chetty, Hendren, 2018\nocite{chetty2018impacts}; Ginther, Pollak, 2004\nocite{ginther2004family}). It is common, especially among conservative voices, to refer to marriage as ``America's strongest anti-poverty weapon.''\footnote{See congressman Ted Budd \href{https://www.dailysignal.com/2020/02/24/marriage-is-the-ticket-out-of-poverty/}{here}, among others.}

Is it a good idea to make people marry more? Generally, this depends on what we believe are the \emph{causal} benefits of marriage. For sure, married people do better, but is this related to the fact they are married? A simple marginalist answer is negative: if people choose whether to marry, and marriages differ in quality, than making people marry more means allowing for lower quality marriages with the lowest benefit from them to be created. The competitive market may just be not the right setup to think about this. Still, marriages are not created equal, and some marginal partnerships can be very similar to singlehood.

In this paper I argue that the contrast between marriage and singlehood leaves out an important distinction. Some marriages follow the birth of the first child; others take place before the couple has children. Formally, I define ``kids first'' marriages as marriages for which the first child was born before or at the year of the marriage and ``marriage first'' as those marriages for which the childbirth happened at least in the following year. The former group includes what is traditionally referred to as a ``shotgun marriage'' --- a marriage that occurs when a bride is pregnant. Like shotgun marriages, the kids-first marriages suggest that an unintended pregnancy contributed to the decision to get married. And as I show later, the kids-first couples are marginal in terms of responses to the environment: their creation and dissolution are sensitive to the economic situation and policies.

I document that among women 21--40 in the US, 14\% of kids-first unions end in divorce within five years, as opposed to 5\% for marriage-first. Most of this difference cannot be explained by various ways of controlling for age, demographics, education, and income. This pattern is robust to different choices of time horizons and measures of divorce. Perhaps more surprising, the difference is the largest for the educated people, yet the share of the kids-first marriages is lower for them. Namely, kids-first college graduates divorce more than three times more often than the marriage-first, as opposed to less than 1.5 for those who finished only a high school. A similar pattern applies if I separate early and late births: although women giving the first birth before 25 are more likely to be in a kids-first marriage, the underperformance of such marriages is relatively milder than for those with later births.

% Need for the model.
Even though the data pattern is evident, the lack of exogenous variation in marriage timing does not explain why people experience such different outcomes and the channels generating this difference. In a more narrow sense, if we believe that the kids-first marriages have worse unobservable quality causing them to be less stable, studying policy and welfare questions require understanding the magnitudes of these quality differences. I develop a quantitative model that can, essentially, map the differences in the divorce rates into marriage quality differences.

My structural approach helps to answer several questions. First, what explains the difference in performance between the kids-first and other marriages? Second, what part of the value of marriage is generated by having children together? Third, what is the role of child support and related policies in the creation and dissolution of the shotgun marriages and in shaping the number of single mothers in the economy in general?

The model is based on a dynamic limited commitment framework (Kocherlakota 1996\nocite{kocherlakota1996implications}, Ligon, Thomas, Worrall, 2000\nocite{ligon2000mutual}, Marcet, Marimon 2019\nocite{marcet2019recursive}),  in which I endogenize marriage, divorce, and fertility choice. People are heterogeneous with respect to age, labor productivity, and savings. In a random search spirit, they randomly meet potential matches and form couples, which also differ in the relationship quality and relative decision power of spouses. When a couple meets, the decision power is determined by symmetric Nash Bargaining. The relationship quality --- commonly referred to as love shock --- is the crucial factor affecting people's decisions to marry. Spouses cannot commit not to divorce, so their choices are subject to the participation constraints determined dynamically based on their option to divorce and possibly find new partners. Married childless couples endogenously decide on their fertility timing; children bring utility and require both money and time.

A main driving force for the marriage timing in the model is random fertility shocks happening to the partners before making their marriage decision. When a potential couple meets, with some chance they have an unplanned child. If this happens, the couple's bargaining is affected: if the partners agree, they become a (kids-first) couple with a child; otherwise, the woman risks becoming a single mother or facing a costly pregnancy abortion, and the man loses access to the child. Relative to the regular situation, in which upon disagreement partners just wait for their new matches, this generates different selection into a marriage based on fertility timing, and the model fundamentals determine its strength and implications. In particular, the model predicts that if the unplanned pregnancy happened, some couples with inferior match quality agree to marry each other, which later generates a higher risk of divorce. The intuition for this comes from two things: people generally like the arrival of children, and being a single mother or aborting is not a desirable option.

I estimate the model parameters using a simulated method of moments, matching the established data evidence to the quantities in the simulated data. The parameters I estimate are the preference for children, household technology features, and the transition probabilities. I match several sets of moments. The most important thing is the percentage of divorced women by different durations of marriage in the kids-first and the marriage-first group. I also include the share of never-married and divorced with and without kids by age, the percentage of couples having children by the duration of their marriage, and several other things. To account for the education gradient flexibly, I perform two separate estimations on two subgroups --- college graduates and high school graduates. The model fits very well in both subgroups, with few exceptions for the high school graduates where the fit is still reasonable.

Qualitatively, the estimates indicate a significant uninsurable risk of unplanned pregnancies that people face. There are substantial differences in the probability of unplanned pregnancy between the high school and the college groups. For both of them, the risk has the monetary equivalent of several years of labor earnings. The risk is relatively larger for college. Moreover, the estimates indicate that the threshold for relationship quality that is acceptable for partners to agree to marry each other falls following the shock, leading to the creation of marriages that would not happen without an unplanned pregnancy. These marriages almost mechanically have the highest risk of divorce. Finally, following an unexpected pregnancy, the woman's outside option drops: on average, she does not want to raise the kids on her own, unless she is highly productive or has considerable savings. This drop decreases her decision power inside marriage, leading to potentially different outcomes even for couples with relatively high marriage quality.

The model provides a way to look at post-marital policies like the child support. Some women would choose not to enter marriages if they can just collect child support and raise their children as single mothers. In addition, women in existing marriages can be more tempted to leave. This can cause a response such that some marriages would not be created, and many other selection-related results. Therefore, the welfare can be affected either way. The model argues that more child support is generally welfare improving, and to some extent it is welfare improving for both men and women, so it is far from being just a redistribution. On the other hand, its effects on creation and dissolution of shotgun marriages itself is minimal, although it affects the share of single mothers by making couples divorce more and singles to abort pregnancies less often.

% Gram
More generally, the model can touch whether keeping people married is good, even though marriage is an endogenous decision. One obvious way to create more marriages is to make divorce easier! It creates more marriages and induces large welfare gains. However, the model also predicts a sizable decrease in in-marriage fertility. Shortly, those couples that stick together because there is an easy way to exit a marriage rarely have kids. Although people win from easier divorce, this does not look like a desirable outcome from the conservative view.

On the other hand, the model suggests that reducing the pressure for people to marry following an unplanned pregnancy and giving people better control over their fertility is beneficial in welfare for both parents and children. These two measures combined reduce the count of single mothers in the economy. The former is more important for the high school graduates and the latter --- for the college graduates.

Quantitatively, more than half of college and a bit less of high-school kids-first couples marry because they had a pregnancy. Each of three factors - imperfect fertility control, social stigma against unmarried parenting, and difficult remarriage after a birth - explain one-third to one-half of excess divorces of the kids-first couples, and their interaction is non-negligible. Eliminating existing child support has minor effects on the creation of shotgun marriages, and the largest margin that it affects is the percentage of abortions. The welfare benefits of child support can go both ways. In some cases, elimination of it benefits both parents and children, together with decreasing divorce rates and single mothers' share. However, expanding child support may also be beneficial in some welfare dimensions.

My paper contributes to the literature in several ways. First, I am the first to document a strong and robust underperformance of kids-first marriages in modern US data. Second, I am the first to estimate a lifecycle model of marriage and divorce featuring a fertility choice, which is central in the model. Third, I am the first to study the interaction of creation and dissolution of shotgun marriages from public policy prospectives, which previously have been considered in isolation while being importantly related as two sources of formation of single-parent families.

I am not the first to study shotgun marriages in the literature. A seminal piece of research is Akerlof, Yellen, Katz, 1996\nocite{akerlof1996analysis}. They interpret them as commitment technology to access premarital sex and argue that contraception and abortion technologies lead to this practice's disappearance. Further, welfare expansion (Neal, 2004\nocite{neal2004relationship}) and the stock of potential partners (Chiappori, Oreffice, 2008\nocite{chiappori2008birth}) were argued to influence women's decision to enter marriages following pregnancies. However, these models are mostly of illustrative purposes. In contrast to this, my model is a quantitative study capable of delivering policy counterfactuals and welfare implications.

Empirically, Alesina, Giuliano, 2005\nocite{alesina2006divorce} have shown that easier divorce increased the number of shotgun marriages created, and Tannenbaum, 2020\nocite{tannenbaum2020effect} and Rossin-Slater, 2017\nocite{rossin2017signing} have shown that stricter child support enforcement reduces pregnancy-related marriages. These findings are consistent with the story that shotgun marriage is a device to obtain the required commitment and resources for childbearing. However, they focus on the angle of the creation of shotgun marriages rather than on understanding their dynamics. Two related papers --- Forester, 2020\nocite{foerster2020untying} and Brown, Flinn, 2011\nocite{brown2011family} --- explore the impact of child support and alimony payments on the dynamics of divorce using models, yet they isolate themselves from marriage creation. My work combines both of these lines of literature: I study both creation and dissolution, potentially having non-trivial interaction through the selection of marriage quality.

From the model perspective, my work is within a growing branch of literature of lifecycle models of marriage and divorce, pioneered by Mazzocco, 2007\nocite{mazzocco2007household}, and Voena, 2015\nocite{voena2015yours}. Notable recent examples include Low et al., 2018\nocite{low2018marriage}, Shephard, 2019\nocite{shephard2019marriage}, Blasutto, 2020\nocite{blasutto2020cohabitation}. The context of fertility choice within lifecycle models without divorce was studied by Sommer, 2015\nocite{sommer2016fertility}, and Ejrnæs, Jørgensen, 2020\nocite{ejrnaes2020family}, among others. Finally, the model broadly contributes to the literature understanding the value of marriage, which has been discussed since the work of Becker, 1981\nocite{becker1981altruism}. Modern studies emphasize roles of risk-sharing, as in Lise, Yamada, 2019\nocite{lise2019household}, the general comparative advantage of being a couple as in Chiappori, 1997\nocite{chiappori1997introducing}, and shared production of public good as in Greenwood et al., 2016\nocite{greenwood2016technology}. My work here primarily focuses on the last one: children are the crucial part of the value created within a couple and therefore are first-order issues in how couples form.
% Mention on numbers of births unintended. If you ask people, so many births are unintended.
% Welfare impilcation of making divorce higher in qualitative results.

The rest of the paper is organized in the following way. First, I document the behavior of kids-first marriages empirically in Section 2. I discuss and present a theoretical model in Section 3. Section 4 presents the details of the solution and estimation. Section 5 presents the the model fit and the estimation results. Section 6 provides a series of results aiming to understand the causes and implications of the differences in divorce rates. Section 7 presents and discusses marriage promotion policy counterfactuals. Section 8 concludes.

\section{Empirical Patterns\label{empirical-section}}

Before presenting a model, I summarize the main results of the incidence and presence of shotgun marriage.  This section does not aim to establish causality, and yet I attempt to argue that there is no obvious explanation in the data for the reasons why relative timing matters. First, using ACS as the primary data source, I show the main result of marriage performance for kids-first and marriage-first couples. Second, I confirm that it holds on large subsamples of the data. Whether I pick older or younger or more or less educated, I still get a persistent difference, although its magnitude varies. Relatedly, I show that controlling for many variables influencing the share of divorced does not invalidate the result. Third, using a supplementary sample from SIPP data, I argue that the fact the in the primary dataset marriage and fertility history cannot be recovered perfectly is not likely to drive the conclusions. Finally, I demonstrate that with my definitions majority of kids-first couples have their own children as opposed to being a result of single mothers finding a new partner, though the fact of having the own children does not seem to play a huge role.

I mainly use the American Community Survey (ACS) samples of 2009--2017 for the main empirical exercise and the model estimation. Its advantage is large sample sizes with very precise household data: it can deliver very detailed partitions of the data, like focusing only on people surveyed a certain number of years after their marriage. Its primary disadvantage is the incompleteness of the marriage and fertility histories, which I address later.

To classify the couples by fertility timing, I consider the following simple measure:
\begin{equation}
\Delta T = \text{Year of the first birth} -  \text{Year of the first marriage}
\end{equation}
assuming both events happened by the time the person is observed. Based on the $\Delta T$ I call kids-first (KF) women those who have $\Delta T \leq 0$ and marriage-first (MF) those who have $\Delta T > 0$. %Figure ... presents a motivating graph: the histogram shows distribution of women with given $\Delta T$ between $-5$ and $5$, and the scatter plot shows percentage of those divorced five years after marriage with this value of $\Delta T$.

To form the sample for the analysis, I pick 2009--2016 ACS, pick women aged 21--40 at the moment of the survey, exclude those with marital statuses ``spouse absent'', ``widowed'' and ``separated''. The resulting selection is what I refer to as a general population of women. Within this population, I focus on females who are either married or divorced now, have children present in a household, and are married no more than once. Finally, within those, I focus on an 11-year window between the years of marriage and fertility $-5 \leq \Delta T \leq 5$. The first birth year is computed based on the age of the eldest child, so this assumes that the eldest child still resides with the woman.\footnote{Some cases have the age difference between the mother and the child is above 14, for them the year was treated as missing.} ACS person-specific weights are used in all calculations.

The described restrictions aim to address the imperfections of the ACS data, which is a large-scale cross-sectional household survey. Focusing on women allows capturing divorces more accurately, as children staying with mothers is a default custody allocation in the US. Restricting the age allows focusing on women who are likely to reside with their children. Picking married once is a limitation of the data, as only the year of the most recent marriage is recorded. In Subsection \ref{sipp-results} I argue that these restrictions are not crucial for the empirical result. Finally, picking a window around $\Delta T = 0$ allows mitigating family arrangements that are likely to have stepchildren and families where the older child has moved out. Appendix \ref{changing-window} shows that result still holds without this restriction, although observations with $\Delta T$ outside the window decrease the magnitude of the differences I discuss.

%Lag between pregnancy and childbearing is important yet not crucial for the classification. The main group of interest are people with $\Delta T = 0$. Assuming that average pregnancy lasts a little less than nine months, even with random timing less then one-quarter of these couples will be married before their pregnancy. One still can imagine a couple who married in January, got pregnant in March and gave a birth in December, but this requires very precise timing of decisions and, moreover, marrying no later than March, which is uncommon in particular due to cold season in most of the US. By similar reasoning, some share of couples with $\Delta T = 1$ did actually marry while being pregnant, and yet those people are less constrained in the time.

%A substantial share of couples has $\Delta T < 0$, meaning out-of-wedlock childbirth preceding the marriage. In most of the empirical part I pool this group with those with $\Delta T = 0$.

The top part of Table \ref{share_table_0} illustrates the differences in the divorce between kids-first and marriage-first groups in general, as well as the relative proportion of the kids-first group. To measure the divorce, I focus on shares of divorced women among those ever married. I first use a simple cross-sectional measure, comparing the raw percentages of divorced people in each of the two groups. This measure does not account for duration properly: given $\Delta T$, some couples may be married longer than others at the moment of the survey. Therefore I also present the difference conditional on particular durations. The table shows the results conditional on being 5 or 10  years after the marriage; the large sample size allows doing this. Surprisingly, this conditioning does not change the difference substantially. Further, the bottom part of the table illustrates the heterogeneity by partitioning people on education groups and comparing women with earlier and later births.

\begin{table}[h]
\caption{Share of divorced among kids-first and marriage-first, ACS\label{share_table_0}}
\begin{tabular}{l r r r r }
\hline
& \multicolumn{2}{c}{share of divorced if ... }&  \\
&  marriage-first & kids-first & (share of kids-first) &  \\\hline
\multicolumn{5}{l}{\textbf{All sample}} \\\hline
\textit{Cross-sectional share of divorced} &  9.9 & 18.1 & (26.1) \\
\textit{Divorced 5 years after marriage} &  5.1 & 14.3  & (24.6) \\
\textit{Divorced 10 year after marriage} & 10.4 & 22.8 & (24.4) \\\hline
\multicolumn{5}{l}{\textbf{Cross-sectional share of divorced for subsamples}} \\\hline
\textit{High school only} &  12.8 & 17.3 & (37.0) \\
\textit{Some college} & 14.3 & 21.0 & (31.3) \\
\textit{College or more} &   5.3 & 14.8 & (11.9) \\\hline
\textit{First birth before 25} & 15.5 & 19.9 & (41.8) \\
\textit{First birth at 25 or later} &  6.2 & 10.9 & (10.7)  \\\hline
 \multicolumn{5}{p{0.9\linewidth}}{\footnotesize \textit{Notes.} This is American Community Survey data, 2009--2017. The numbers are percentages. Two left columns show the percentage of divorced conditional on being in a marriage-first or kids-first group, respectively. The right column shows the relative proportion of kids-first in those who belong to either group. Kids-first refers to women who have their first child before or at the year of marriage, marriage first to those who have their first child at least in the following year. Everything is conditional on being married once and having children at home; see the text for precise definitions. }\\\hline\hline
\end{tabular}
\end{table}

Few patterns are worth noting. First, people in the kids-first group systematically have higher divorce rates. Second, this difference is more pronounced for college graduates, despite the smaller yet sizable share of the kids-first group for them: with comparable absolute difference marriage-first group is relatively much more stable. Third, although very related to the previous one, the difference is more visible for women who give birth later rather than earlier. Fourth, heterogeneity in shares of kids-first and shares of divorced by subgroups suggests that composition is an essential factor.

As the divorced rates are heterogeneous within the population, the ratio of shares of divorced is more informative than the difference. In particular, the table suggests that the divorce rate for kids-first high school graduates is 1.4 times higher than for the marriage-first, and this ratio is 2.8 for college graduates.

Table \ref{share_table_1} uses a flexible linear regression to control for the composition. The specification is
\[\text{Divorced}_i =\Delta\cdot \text{KF}_i + \gamma \cdot X_i + \varepsilon_i\]
where $X_i$ represents possible controls. Raw difference in the divorce rate corresponds to $\hat\Delta$ when $X_i$ contains only constant.

I employ three sets of controls: individual characteristics include dummy variables age and education interacted, as well as race, fixed effects of state, belonging to a metropolitan area and survey year. Duration controls include dummies for all interactions of age with age of the first marriage and age of the first birth. Finally, income controls are a 3rd degree polynomial of log income on a subsample of women who are employed, have non-missing income data, work at least 10 hours per week, and report labor earnings more than \$3000 a year. To supplement this result, I present a matching estimator based on a propensity score, which is an implied probability of being in the kids-first group predicted by the same controls (excluding duration controls, as predicting the treatment perfectly by definition).

These regression results are subject to an important limitation: as kids-first and marriage-first couples differ in their marriage and birth timing, the common support is violated --- it is not possible to pick people from two groups with identical timing. This motivates exclusion of duration when predicting propensity score, and this also generally hurts the statistical interpretation of all the differences in divorce. Nevertheless, its large magnitude and robustness do suggest that this is not solely an artifact of the data structure.

\begin{table}[h]
\caption{Difference in share of divorced, regression and matching\label{share_table_1}}
\begin{tabular}{l r r r r}
\multicolumn{5}{c}{\textit{Regression equation:} $\text{Divorced}_i = \Delta \cdot \text{kids-first}_i + \text{Controls}_i + \varepsilon_i$} \\\hline
\hline
& \multicolumn{2}{c}{Estimates} &  & \\
&  Difference ($\Delta$)  & (standard error) & Ratio $= \frac{\text{div if KF}}{\text{div if MF}}$ &  \\\hline
\multicolumn{5}{l}{\textbf{All sample, regression}} \\\hline
\textit{No controls (raw difference) }& 8.2 & (0.1) & 1.83 \\
\textit{Demographic controls }& 5.7 & (0.1) & 1.58 \\
\textit{Duration controls} &  5.8 & (0.2) & 1.59 \\
\textit{Demographic + duration} &  4.8 & (0.2) & 1.48 \\
\textit{Demographic + duration + income} & 4.1 & (0.3) & 1.41 \\\hline
\multicolumn{5}{l}{\textbf{All sample, propensity score matching}} \\\hline
\textit{Demographic controls} & 5.0 & (0.3) & 1.50 \\
\textit{Demographic controls + income} & 4.9 & (0.2) & 1.49 \\\hline
\multicolumn{5}{p{\linewidth}}{\footnotesize \textit{Notes.} This is American Community Survey data, 2009--2017. The numbers are regression estimates in percents.  $\Delta$ corresponds to the difference in share of divorced between kids first and marriage first. Ratio is the ratio of the percentages of divorced between kids-first and marriage first, in a context of regression it is defined as $1 + \frac{\hat{\Delta}}{\text{div if MF}}$. Regression with income controls is for subsample of those whose income data is non-missing and of good quality, see the text for precise definitions. Propensity score matching is done using default Stata \texttt{psmatch2} options, interactions of the controls were not included due to computational reasons.}\\\hline\hline
\end{tabular}
\end{table}

This evidence suggests that although composition plays a role, regardless of observable characteristics kids-first women are divorced more often than marriage first. Marriage duration and demographics only explains less than half of these differences.

The following subsections and Appendix \ref{app-additional-evidence} provide few extra checks and additional insights. Shortly, using more detailed and more complete data does not change the patterns I discuss here.

\subsection{Full Marital History (SIPP) \label{sipp-results}}
This section provides further evidence using Survey of Income and Program Participation data, Wave 2014. The main distinguishing feature is the concrete recording of the year of the first marriage and the age of the first birth for every woman. This allows for relaxing two sampling restrictions: conditioning on women ever married instead of married once (counting remarried) and correctly counting women with non-residential children (including women older than 40). In the dimensions excluding those two things, the methodology is identical, including the use of $\pm 5$ year window for $\Delta T$.

\begin{table}[h]
\caption{Share of divorced among kids-first and marriage-first, SIPP\label{share_table_0}}
\begin{center}
\begin{tabular}{l r r r r }
\hline
& \multicolumn{2}{c}{share of divorced if ... }&  \\
&  marriage-first & kids-first & (share of kids-first) &  \\\hline
\multicolumn{5}{l}{\textbf{All sample, cross-sectional shares}} \\\hline
\textit{Ever divorced} & 34.4 & 48.0 & (21.0) \\
\textit{Divorced if married once} &  12.4 & 19.3 & \\
\textit{Remarried} &  22.0 & 28.7 &  \\
\multicolumn{5}{l}{\textbf{Cross-sectional share of ever divorced for subsamples}} \\\hline
\textit{High school only} &  37.5 & 46.3 & (26.4) \\
\textit{Some college} & 41.6 & 52.0 & (22.0) \\
\textit{College or more} &   24.6 & 45.8 & (12.7)\\\hline
\textit{First birth before 25} & 45.1 & 53.1 & (28.2) \\
\textit{First birth at 25 or later} &  20.7 & 23.4 & (9.4)  \\\hline
\multicolumn{5}{p{0.8\linewidth}}{\footnotesize \textit{Notes.} This is Survey of Income and Program Participation data, the cross-section of Wave 1, 2014. The numbers are percentages. Two left columns show the percentage of divorced conditional on being in a marriage-first or kids-first group, respectively. The right column shows the relative proportion of kids-first in those who belong to either group.  Kids-first refers to women who have their first child before or at the year of marriage, marriage first to those who have their first child at least in the following year.  }\\\hline\hline
\end{tabular}
\end{center}
\end{table}

\subsection{Finer Partitions of the Data}
This part supplements the main result by digging deeper into the composition of the kids-first group. Shortly, regardless of who is considered, the qualitative findings that kids-first people have higher divorce rates hold, although quantitatively, it may differ quite a lot.

\subsubsection{Partition by $\Delta T$ (ACS)}
In the main exercise people with $\Delta T \leq 0$ were pooled together. This part aims to see how people with $\Delta T = 0$ (fertility at the marriage year) and $\Delta T < 0$ (fertility before marriage) are different. Table \ref{dt_table} shows the first set of results, redoing the comparison with and without controls. It shows that regardless of controlling strategy, women who have kids at the very year of marriage have somewhat higher divorce rates than those who married in the following years, although all of them divorce more than marriage-first women.

\begin{table}[h]
\caption{Difference in share of divorced, by different $\Delta T$\label{dt_table}}
\begin{tabular}{l r r r }
\multicolumn{4}{c}{\textit{Regression equation:} $\text{Divorced}_i = \Delta_0 \cdot \I[\Delta T = 0] + \Delta_{<} \cdot \I[\Delta T < 0]  + \text{Controls}_i + \varepsilon_i$} \\\hline
\hline
& \multicolumn{2}{c}{Estimates} &   \\
& Kids-first, at ($\Delta_0$)  & Kids-first, before ($\Delta_<$)  \\\hline
\textit{No controls (raw difference) }&  8.6 &  8.0  \\
\textit{Demographic controls }& 6.8 &  5.2 \\ 
\textit{Duration controls} &  6.2 &  5.1 \\ 
\textit{Demographic + duration} &  5.2 &  3.9 \\
\textit{Demographic + duration + income} & 5.0 &  2.3  \\\hline
\multicolumn{4}{p{\linewidth}}{\footnotesize \textit{Notes.} ACS, 2009--2017. The numbers are regression estimates in percents.  The estimates indicate the difference in share of divorced relative to the marriage-first women}\\\hline\hline
\end{tabular}
\end{table}

Finally, I show the graphs with the percentages of divorced for each $\Delta T$ on Figure \ref{fig2}. The graph shows that $\Delta T = 0$ generally has the highest share of divorced. The largest difference is between people with $\Delta T = 0$ and $\Delta T \in \{2,3,4\}$, and it drives most of the result, where for other points the difference is less striking or even reversed, although their weight (indicated by height of the bars) is lower, and therefore they do not overturn the main result.

\begin{figure}[h!]
\begin{center}
\includegraphics[scale=1.0]{../pics/div_5y_by_dt.pdf}
\caption{Share of divorced in 5 years by relative timing of marriage and fertility.\label{fig2}}
\end{center}
\end{figure}

\subsubsection{Multi-Partner Women (SIPP)}
Although the use of $\pm 5$ years window is dedicated to reducing the concern of people having stepchildren, it does not eliminate it entirely. Within the kids-first group, one may desire to consider women marrying the father of her first child and women marrying someone else than the child father separately. An essential issue of most of the surveys is that relation of the male to the child is clear if the male is present in the household, but once the couple divorces and the male is not present, only a few sources provide information on the retrospective relationship. This issue is impossible to overcome directly in ACS; however, it is possible to handle using SIPP.

For a subsample of women having more than one child, SIPP asks the question about having children from multiple partners. Therefore I introduce a partition on three groups. Marriage-first women are those having $\Delta T > 0$ as before. Kids-first-own are those with $\Delta T \leq 0$ who (1) do not report to have multi-partner fertility, and (2) have had children born both before and after their first marriage. Kids-first-other are those who do not meet these criteria. In the data, more than three-quarters of this group do report multipartner fertility. Because of the nature of the multipartner fertility questions, I also exclude women who are married more than once: most of them are likely to be misclassified to the ``other'' group. To repeat the strategy from the ACS part, I also restrict attention to women 40 or younger: as many people remarry later, the attrition will introduce bias considering solely women married once.

Table \ref{fine_table} shows the distribution of the females 21--40 and older conditional on having more than one child and being married once by the three groups, as well as share of divorced in each of them. Three important conclusions are (1) the majority of the kids-first group seem to have their own children, (2) the percentage of divorced is higher for both subgroups of kids-first, (3) people with multipartner fertility have higher divorce rates, but their share is relatively smaller.

\begin{table}[h!]
\caption{Finer partition of kids-first, SIPP\label{fine_table}}
\begin{center}
\begin{tabular}{l r r}
\hline
& share in sample & share of divorced  \\\hline
\multicolumn{3}{l}{\textbf{Women 21--40, 1 marriage, 2+ children}} \\\hline
\textit{Marriage-first} & 72.1 & 10.5 \\
\textit{Kids-first-own} &  16.4 & 14.6 \\
\textit{Kids-first-other} &  11.4 & 27.2 \\\hline
\multicolumn{3}{p{0.6\linewidth}}{ \footnotesize \textit{Notes.} This is Survey of Income and Program Participation data, cross-section of Wave 1, 2014. The numbers are percentages. Kids-first-own is a subset of kids-first for which it is verifiable that woman married a father of the first child, kids-first-other includes all other cases.} \\\hline
\end{tabular}
\end{center}
\end{table}

\section{Model}
There is a number of well-established features for the models of individual agents' lifecycle that reproduce desirable properties of the microdata (see ..., ... among others), the dynamics of couples is explored relatively less. Moreover, endogenous transitions between singleness and marriage and back require these options to be comparable, and the model of collective household with endogenous decision weight is intended to achieve this.

The transitions are based on total values of being in a particular state, so some features of the model (like one-dimensional income process and no state for child's human capital) are simplistic relative to the modern literature and are modeled in a somewhat reduced form. I see three features of the model as crucial. First, the marriage market has to be rich to capture how spouses are selected against particular partners. Second, intrahousehold utility should capture the value of marriage and the possible presence of couples with and without children, so the value of fertility is well-understood. Third, the remarriage market has to be realistic: it should both be related to the primary marriage market and reflect the difficulties in finding partners after a divorce.

The economy consists of women and men, who both start their life single and childless at time $t = 1$. It corresponds to the age of $21$ for women and the age of $23$ for men. Single and childless people without kids work full-time. In each period, with some probability $p^{\text{meet}}_t$ singles meet a random partner, coming from a fixed distribution of singles of the opposite gender.

People are heterogeneous with respect to their age $t$, labor productivity, $z$, and savings (assets) $a$. Couples have two additional dimensions: a common marriage quality $\psi$, that is enjoyed by both partners, and decision weight $\theta$, which reflects the relative decision power of the woman in the couple (Pareto weight). Finally, singles and couples can have up to one child, which stays with them forever, brings value and childbearing costs. If a couple breaks up, child custody is always assigned to the mother.

Therefore, in total, there are five types of agents: a single woman $(f)$, a single man $(m)$, a single woman with a child (single mother henceforth) $(fk)$, a couple without children $(c)$, a couple with a child $(ck)$. Single fathers can be defined analogously, but are outside of the baseline consideration.

The marriage market is central to the creation of shotgun marriages. It is modeled through random shocks to the matches: when two potential partners meet, with some probability, an unplanned pregnancy happens and interferes with their decisions to marry each other. If the pregnancy does not happen, upon disagreement two partners keep being single and meeting other people. If the pregnancy does happen, then upon disagreement woman either stays a single mother or carries costly abortion. Abortion access is imperfect: with probability $1-p^a$ woman gets no access to abortion (note that this also captures a share of women who got their babies born before the marriage decision), and with probability $p^a$ she can choose to have an abortion with utility costs $\phi_a$. It is assumed that women do not know the realization of the abortion costs at the moment of marriage decision. This assumption both justifies friction in decision making (marriage decisions are not immediate, and abortion decisions have to be made quickly) and slightly economizes on computations. If potential partners get a pregnancy shock, the child support may also be paid; this is discussed later in \ref{divorcevalues}. Figure \ref{diagram} summarizes the discussed transitions.

The rest of this section summarizes the decision problems and the agents' value functions, starting from couples to singles. Then it discusses the described marriage market formally, finally it covers divorce and other essential features.

\begin{figure}
\begin{center}
\begin{tikzpicture}[every text node part/.style={align=center}, scale=0.75]
   % Place nodes
   \node [block] (1) {\large \textbf{People}  \\[-0.5ex]  \large  \textbf{Randomly Meet}};
   \node [block, below left = 1.5 cm and 0.0 cm of 1] (2) {\textit{Nash} \\[-0.5ex] \textit{Bargaining}};   
   \node [block, below right = 1.5 cm and 0.0 cm of 1] (3) {\textit{Nash} \\[-0.5ex] \textit{Bargaining}};   
   \node [block, below = 1.5 cm of 2] (5) {\textbf{Kids First} \\[-0.5ex] \textbf{Couple} ($ck$)};
   \node [block, below = 0.8 cm of 3] (7) {No Kids \\[-0.5ex] Couple ($c$)};
   \node [block, below = 0.8 cm of 7] (8) {\textbf{Mariage First} \\[-0.5ex] \textbf{Couple} ($ck$)};
   \draw[->] (1) to [bend left = 20] node [left] {\scriptsize Unplanned \\[-0.75ex] \scriptsize pregnancy} (2);
   \draw[->] (1) to [bend right = 20] node [right] {\scriptsize No \\[-0.75ex] \scriptsize pregnancy} (3);
   \draw[->] (2) to node [left] {\scriptsize Agree} node [right] {\scriptsize \textbf{Shotgun} \\[-0.75ex]  \scriptsize \textbf{marriage}} (5);
   \draw[->] (3) to node [right] {\scriptsize Agree} node [left]  {\scriptsize \textbf{Regular} \\[-0.75ex]  \scriptsize \textbf{marriage}} (7);
   \draw[->] (7) to node [right] {\scriptsize Kids by choice} (8);   
    \node [block, above left   = 0.575 cm and -0.75 cm of 2] (4) {\,Single male \\[-0.5ex] ($m,s$)\,\,};
    \node [block, above left   = 1.8 cm and -0.75 cm of 2] (4-sm) {\,Single mom \\[-0.5ex]  ($f,sk$)};
    %\node [block, left = 0.2 cm of 4-sm] (4-nsm) {\,Single female \\[-0.5ex]  ($f,sk$)};
	\node [block, above right   = 0.575 cm and -0.75 cm of 3] (6) {\,\,Single male\,\,\\[-0.5ex] ($m,s$)};
    \node [block, above right   = 1.8 cm and -0.75 cm of 3] (6-f) {Single female\\[-0.5ex] ($f,s$)\,};   
    \draw[->] (2.180) to [bend left = 30] node [left] {\scriptsize Disagree} (4.270);
    \draw[->,thick,dotted] (4-sm.20) to [bend left = 10] node [above right=-0.2cm and 0.7cm] {\scriptsize Costly abortion \\[-0.75ex] \scriptsize (if possible)} (6-f.50);
    \draw[->] (3.0) to [bend right = 30] node [right] {\scriptsize Disagree} (6.270);
     \draw[->] (6.180) to [bend left = 30] node [above] {\scriptsize Back}  (1.-12);
   \draw[->] (6-f.180) to [bend right = 30] node [above] {\scriptsize Back} (1.12);
   \draw[->,thick,dashed] (4-sm.90) to [bend left = 30] node [above] {\scriptsize Worse marriage prospectives}  (1.168);
   \draw[->] (4.0) to [bend right = 30] node [above] {\scriptsize Back} (1.192);

\end{tikzpicture}
\caption{Transitions around the marriage market.\label{diagram}}
\end{center}
\end{figure}

\subsection{Couple and Child}
Let $V^{f,ck}_t$ and $V^{m,ck}_t$ represent the values of male and female of being in a couple with a child conditional on staying together in period $t$, and $V^{f,dk}_t$, $V^{m,dk}_t$ represent the value of divorce of such a couple. This section describes their recursive definition, starting from value $0$ in the last period.

The couple enters the period with savings $a$, bringing gross return $R\cdot a$, productivities $z^m$ and $z^f$, love shock $\psi$ and decision weights $\theta' = (\theta^f,\theta^m)$. Productivities, together with age, determine female and male wages $W^f_t(z^f)$, $W^m_t(z^m)$. The couple gets utility from spouses' consumption and the child's consumption; part of the latter can be produced at home by reducing the female labor supply. The love shock represents an additive utility surplus of being together.

Conditional on staying together in the period $t$, the choice of the couple are spouses' consumption $c^f$, $c^m$, expenditures on the child $x$, female labor supply $l^f$ and savings $a'$. I summarize the ``exogenous'' state by $\omega = (z^m,z^f,\psi)$. After making the decisions, the couple transitions to the next period $t+1$ and gets a draw of $\omega'$. Given $a'$ and $\omega'$ it chooses whether to divorce, which is represented by a binary indicator $d_{t+1}(a',\omega')$.  If it does not, it enters the next period with the decision weights $\theta'=\theta'(a',\omega',\theta)$, evolution of which is determined by renegotiation; otherwise the couple divorces.

The couple's decisions are defined by the following optimization problem and the continuation values:
\begin{align}(\tilde a',\tilde c^f,\tilde c^m,\tilde x,\tilde l^f)  = \argmax\limits_{a',c^f,c^m,x,l^f} \left\{\theta^f \cdot u(c^f) + \theta^m\cdot u(c^m) + \psi + \phi(q) + \beta\cdot \E_{\omega'|(l^f,\omega)} \mathcal{V}^{ck}_{t+1}\right\}
 \end{align}
\vspace{-2em}
\begin{align*}
\text{s.t. \ } & a' + c + x = R\cdot a  + W^m_t(z^f) + l_f \cdot W^f_t(z^m) & \text{(evolution of joint assets)},\\
                    & c =C(c^f,c^m) & \text{(increasing returns in consumption)},\\
                    & q = f(x,l^f) \geq \underline{q} & \text{(required costs of childcare)},
                  %  &  z^{f\prime} = z^f + \varepsilon^{z,f}_t - \delta(l^f) & \text{ (evolution of female productivity)}\\
				% &  z^{m\prime} = z^m + \varepsilon^{z,m}_t &  \text{ (evolution of male productivity)}\\
                   % & \psi' = \psi + \varepsilon^{\psi}_t & \text{(evolution of marriage surplus),}
\end{align*}
\vspace{-2em}
\begin{align*}
\text{where \ } & \mathcal{V}^{ck}_{t+1} =   (1-d_{t+1})\cdot \left[ \theta^f \cdot V^{f,ck}_{t+1}(a',\omega',\theta') +  \theta^m \cdot V^{m,ck}_{t+1}(a',\omega',\theta')\right]  & \text{(no divorce),} \\
& \hspace{2cm}+ d_{t+1}\cdot \left[ \theta^f \cdot V^{f,dk}_{t+1}(a',\omega',\theta') +  \theta^m \cdot V^{m,dk}_{t+1}(a',\omega',\theta')\right] & \text{(divorce).}
\end{align*}
Note that the decisions are homogeneous of degree $0$ with respect to $\theta$, which allows assuming $\theta^f + \theta^m = 1$. The corresponding individual continuation values are:
\begin{align}
V^{f,ck}_{t}(a,\omega,\theta)  = u(\tilde{c}^f) & + \psi + \phi(\tilde{q}) + \\\nonumber &\beta \cdot \E_{\omega'|(l^f,\omega)} \left\{ (1-d_{t+1}) \cdot V^{f,ck}_{t+1}(a',\omega',\theta')  + d_{t+1} \cdot V^{f,dk}_{t+1}(a',\omega',\theta') \right\},\\
V^{m,ck}_{t}(a,\omega,\theta)  = u(\tilde{c}^m) & + \psi + \phi(\tilde{q}) + \\\nonumber &\beta \cdot \E_{\omega'|(l^f,\omega)} \left\{ (1-d_{t+1}) \cdot  V^{m,ck}_{t+1}(a',\omega',\theta')  + d_{t+1} \cdot V^{m,dk}_{t+1}(a',\omega',\theta') \right\}.
\end{align}
The divorce decision and the law of motion for $\theta$ are defined in Section \ref{divorcedecision}. The divorce values $V^{\cdot,dk}$, and the child support assignment are given in Section \ref{divorcevalues}.

The transition for female productivity includes possible skill depreciation, which depends on the female labor supply's value $l^f$. Therefore the expectation operator is $\E_{\omega'|(l^f,\omega)}$. The transition laws are discussed in \ref{transition-laws}.

\subsection{Childless Couple}

Childless married couples have both partners working full-time. They enjoy the match quality stock $\psi$, and choose their consumption and savings. A childless couple may try to conceive a child. This decision precedes all consumption-savings decisions and happens after realizing the shocks and renegotiation, conditional on staying together at this period. The conception decision at the beginning of the following period is denoted by $k_{t+1}$. The decision problem is otherwise similar to those of a couple with a child:
\begin{align}(\tilde a',\tilde c^f,\tilde c^m,\tilde x,\tilde l^f)  = \argmax\limits_{a',c^f,c^m,x,l^f} \left\{\theta^f \cdot u(c^f) + \theta^m\cdot u(c^m) + \psi + \phi(q) + \beta\cdot \E_{\omega'|\omega} \mathcal{V}^{ck}_{t+1}\right\}
 \end{align}
\vspace{-2em}
\begin{align*}
\text{s.t. \ } & a' + c = R\cdot a  + W^m_t(z^f) + W^f_t(z^m) & \text{(evolution of joint assets)},\\
                    & c =C(c^f,c^m) & \text{(increasing returns in consumption)},
\end{align*}
\vspace{-2em}
\begin{align*}
\text{where \ } &   \mathcal{V}^{c}_{t+1} = \\
&  \hspace{-1cm} (1-d_{t+1})\cdot (1-k_{t+1})\cdot  \left[ \theta^f \cdot V^{f,c}_{t+1}(a',\omega',\theta') +  \theta^m \cdot V^{m,c}_{t+1}(a',\omega',\theta')\right]  & \text{(no divorce, no birth),} \\
&\hspace{-1cm} (1-d_{t+1})\cdot k_{t+1}\cdot  \left[ \theta^f \cdot V^{f,c\tilde{k}}_{t+1}(a',\omega',\theta') +  \theta^m \cdot V^{m,c\tilde{k}}_{t+1}(a',\omega',\theta')\right]  & \text{(no divorce, try birth),} \\
& + d_{t+1}\cdot  \left[ \theta^f \cdot V^{f,d}_{t+1}(a',\omega',\theta') +  \theta^m \cdot V^{m,d}_{t+1}(a',\omega',\theta')\right] & \text{(divorce).}
\end{align*}

I assume childbearing itself to be costless, but its success is random and uncertain. $p_t^{\text{success}}$ is assumed to be age-specific. Therefore, for a fertile couple, the value of trying to conceive a child is
\begin{equation}
V^{j,c\tilde{k}}_{t+1}(a',\omega',\theta') = p_t^{\text{success}}\cdot V^{j,ck}_{t+1}(a',\omega',\theta') + (1-p_t^{\text{success}})\cdot V^{j,c}_{t+1}(a',\omega',\theta'), \ \ j\in\{f,m\}
\end{equation}

The decision to conceive is based on a weighted value function \emph{after} the renegotiation (weighted with the next period's weights). Define the couple's weighted values as
\[V^{fm,c\tilde{k}}(a',\omega',\theta') = \theta^{f\prime} \cdot V^{f,c\tilde{k}}_{t+1}(a',\omega',\theta') + \theta^{m\prime} \cdot V^{m,c\tilde{k}}_{t+1}(a',\omega',\theta'),\]
\[V^{fm,c}(a',\omega',\theta') = \theta^{f\prime} \cdot V^{f,c}_{t+1}(a',\omega',\theta') + \theta^{m\prime} \cdot V^{m,c}_{t+1}(a',\omega',\theta'),\]
and the decision to (try to) give a birth is then
\begin{equation}k_{t+1}(a',\omega',\theta') = \I\left[ V^{fm,c\tilde{k}}(a',\omega',\theta')  \geq V^{fm,c}(a',\omega',\theta')\right].\end{equation}

There are two important remarks here. First, the value function $V^{fm,\cdot}$ is different from the weighted value function used in the couple's problem in the definition of $\mathcal{V}_{t+1}$ because of the timing of the decisions. This weighting is an artifact of the timing, and it makes decisions consistent with the broader limited commitment framework, dynamic complications of which are discussed in Marcet, Marimon, 2019. From a practice view, this formulation preserves homotheticity. However, it prevents writing the weighted terms as one value function, as they depend on a mixture of the current and expected future decision weights.\footnote{For a straightforward example, imagine that almost all bargaining weight belongs to the wife in a couple in the current period. Let $(\theta^f,\theta^m) = (1,\varepsilon)$ for some small $\varepsilon$. Suppose also that starting from the next period with certainty and forever $(\theta^f,\theta^m) = (\varepsilon,1)$. This situation has very low expected value for the wife, and it needs to be assessed accordingly, with decision weights reflecting the current state. Otherwise, the husband's future benefits are counted in the value function that now mostly reflects the utility of the wife.}

\subsection{Single Women and Men\label{singles-main}}

Each period single people may meet one potential partner with probability $p^{\text{meet}}$. The partners differ with assets and productivities $(a,z)$, and each match additionally receives a draw of initial marriage quality $\psi$, on top of this, with some chances men meet single mothers. The details of the partners market are described in \ref{marriagemarket}, right now we just denote the distribution of characteristics of the potential future matches by $\Gamma^M = \Gamma_{t+1}(a^M,z^M,\psi^M)$, and corresponding marriage decision $m^M = m_{t+1}(a^M,z^M,\psi^M)$ and resulting bargaining weight $\theta^M = \theta_{t+1}(a^M,z^M,\psi^M)$. So let $d\Gamma^M$ denotes the density of the matches with specified characteristics, and $\omega^M$ denotes future couple's exogenous state.

As I discuss above, there is a chance of $p^{\text{preg}}$ that an unplanned pregnancy happens when a potential couple meets. Given a match, the resulting marriage decision $m$ and the resulting bargaining weight $\theta$ may depend on pregnancy status, so I denote them $m^{m,p}$, $\theta^{m,p}$ when pregnancy happens and $m^{m,np}$, $\theta^{m,np}$ when it does not.

If unplanned pregnancy does not happen and woman agrees to marry, she enters a childless couple, that can later have a child and become a marriage-first couple. In unplanned pregnancy does happen and woman agrees, she enters a couple with a child, that is classified as a kids-first couple.

In case of unplanned pregnancy both partners face social stigma $\phi_s$, that is expressed as a disutility of refusing to marry following an unplanned pregnancy.

Additionally, if woman refuses to marry she faces uncertain access to costly abortion: with a chance of $p^a$ she may choose to be a single mother or to abort the pregnancy with utility costs $\phi_a$, a chance of $1-p^a$ she stays a single mother. At the moment of marriage decision the realization of the abortion access is unknown.

Finally, the woman who refuse to enter the marriage following a pregnancy and kept the pregnancy may be awarded a child support. This happens with probability $p^{cs,n}$. Realization of the child support access happens after the pregnancy access and is uncertain at the moment of abortion decision, and the value of the child support is specific to the characteristics of the partner and the woman, as discussed in \ref{divorcevalues}.

Therefore woman who rejects a proposal gets the following expected value:
\begin{equation}\label{sm-value}V^{f,sk?}_t(a,z^f) = p^{a}\cdot \max\left\{\E_{cs} V^{f,sk}(a,z^f), V^{f,s}_t(a,z^f) - \phi_a\right\} + (1-p^{a})\cdot V^{f,sk}_t(a,z^f).\end{equation}
where
\begin{equation} \E_{cs} V^{f,sk}(a,z^f) = p^{cs,n}\cdot V^{f,sk}_t(a,z^f+CS^m) + (1-p^{cs,n}) \cdot V^{f,sk}_t(a,z^f).
\end{equation}
The child support is match-specific, so this abuses the notation slightly.

Apart from the meeting of partners and pregnancies, the problem of the singles is a standard consumption-savings problem. Starting with single females:
\begin{align} & V_t^{f,s}(a,z^f)  = \max\limits_{a',c^f} \Bigg\{ \ \ u(c^f) +  \\\nonumber &  \beta\cdot \E_{z^{f\prime}|z^f} \Bigg( (1-p^{\text{meet}}_t)\cdot V^{f,s}(a',z^{f\prime}) \,+ & \text{(no partner met)}\\\nonumber
 & p^{\text{meet}}_t\cdot (1-p^{\text{preg}}_t) \cdot \int \Big[m^{M,np}\cdot V^{f,c}_{t+1}(a'+a^M,\omega^M,\theta^{M,np}) + & \text{(met, no pregnancy, agree)}\\\nonumber
 &\hspace{4cm}(1-m^{M,np})\cdot V^{f,s}_{t+1}(a',z^{f\prime}) \Big]d\Gamma^M + & \text{(disagree)} \\\nonumber
 & p^{\text{meet}}_t\cdot p^{\text{preg}}_t \cdot \int \Big[m^{M,p}\cdot V^{f,ck}_{t+1}(a'+a^M,\omega^M,\theta^{M,p})+  & \text{(met, shotgun marriage)}\\\nonumber
 &\hspace{2cm} (1-m^{M,p})\cdot \left\{V^{f,sk?}_{t+1}(a',z^{f\prime}) - \phi_s\right\} \Big]d\Gamma^M\Bigg) \Bigg\} & \text{(disagree, social stigma)}\\\nonumber
 \text{s.t. \ } & a' + c^f = R\cdot a  + W^f_t(z^f) & \text{(evolution of the assets)}.\\\nonumber
\end{align}

The single males face less consequences of unplanned pregnancies, as they do not have to make an abortion decision. Yet, they may be required to pay the child support and suffer from the same social stigma in case of refusal to enter a shotgun marriage. This depends on the female's match-specific decision to keep the pregnancy, so the value of refusal in this case:
\begin{equation}\label{man-cs}\E_{csm} V^{m,s}_{t}(a,z^{m}) = \begin{cases}p^{cs,n} \cdot V^{m,s}_{t}(a,z^{m}-CS^m)  + (1-p^{cs,n}) V^{m,s}_{t}(a,z^{m}), & \text{kept pregnancy}\\
  V^{m,s}_{t}(a,z^{m}), & \text{abortion}\end{cases}\end{equation}

Finally, the pool of potential partners for them is different, as there are chances they meet a single mother. In case this happens and they agree to marry, they face a one-time utility costs $\phi_r$, that is capturing disutility of having a step child. The decision problem for the males, defining $V^{m,s}_t(a,z^m)$ is described in the Appendix \ref{single-male-appendix}.

\subsection{Single Mothers}

Single mothers have sole custody of their children and pay the costs of them fully out of their labor income (that could have been adjusted in a past to account for the child support received). This can partially be done by a reduction in labor supply $l^f$.

They also have a chance to meet a single male, who has a utility loss $\phi_d$ from potentially having stepchildren, as shown in \ref{singlemaleproblem}. I refer to these losses as a remarriage penalty in the further parts.

The optimization problem and the value function definition therefore is
\begin{align} & V_t^{f,sk}(a,z^f)  = \max\limits_{a',c^f,x,l^f} \Bigg\{ \ \ u(c^f) + \phi(q) + \\\nonumber &  \beta\cdot \E_{z^{f\prime}|(l^f,z^f)} \Bigg( (1-p^{\text{meet}}_t)\cdot V^{f,s}(a',z^{f\prime}) \,+ & \text{(no partner met)}\\\nonumber
 & p^{\text{meet}}_t  \cdot \int \Big[m^{M,sm}\cdot V^{f,ck}_{t+1}(a'+a^M,\omega^M,\theta^{M,np}) + & \text{(met, agree)}\\\nonumber
 &\hspace{4cm}(1-m^{M,sm})\cdot V^{f,sk}_{t+1}(a',z^{f\prime}) \Big]d\Gamma^M\Bigg) \Bigg\}  & \text{(disagree)}\\\nonumber
 \text{s.t. \ } & a' + c^f + x = R\cdot a  + l_f\cdot W^f_t(z^f) & \text{(evolution of the assets)},\\\nonumber
                    & q = f(x,l^f) \geq \underline{q} & \text{(required costs of childcare)}.
\end{align}

\subsection{Shocks and Income Trends\label{transition-laws}}
I assume that shocks for productivity terms $z^f$, $z^m$ and love shock $\psi$ follow random walk process. For male productivity and love shocks the laws of motions are:
\begin{equation}\label{z-male-eq}
z^m_{t+1} = z^m_t + \varepsilon^{z,m}_{t+1}, \ \ \varepsilon^{z,m}_{t+1}\sim \mathcal{N}(0,\sigma^2_{z,m}),
\end{equation}
\begin{equation}\label{psi-eq}
\psi_{t+1} = \psi_t + \varepsilon^{\psi}_{t+1}, \ \ \varepsilon^{z,m}_{t+1}\sim \mathcal{N}(0,\sigma^2_{\psi}).
\end{equation}

The random walk structure with no mean reversion allows to interpret the productivity shocks as components of permanent income, therefore both female skills depreciation and the income reallocation through child support can be interpreted as shifts in $z$. The law of motion for the female productivity is
\begin{equation}\label{zoutlf}z^f_{t+1} = z^f_t + \varepsilon^{z,f}_{t+1} - \delta(l^f), \ \ \varepsilon^{z,f}_{t+1}\sim \mathcal{N}(0,\sigma^2_{z,f}), \ \ \delta(1) = 0.\end{equation}
The model assumes that women without children always have labor supply of 1, and women with children (both single and in couples) can reduce their labor supply as an input to the home production of the child's consumption $q$.

Finally, the labor earnings are determined by $z$ and age trends using the following expressions:
\begin{equation}\label{trends-eq}W^f_t(z^f) = \exp( \text{Trend}^f_t + z^f), \ \ W^m_t(z^m) = \exp( \text{Trend}^m_t + z^m).\end{equation}

\subsection{Marriage Market\label{marriagemarket}}
As all the decisions are homogeneous of degree 0 with respect to $(\theta^f,\theta^m)$, to lighten the notation since this part I assume that $\theta^f = \theta$ and $\theta^m = 1-\theta$, so in this part $\theta$ is a scalar. This also assumes that values of being in a couple, $V_t^{f,ck}$, $V_t^{m,ck}$ are monotonic with respect to $\theta$.\footnote{Numerical solutions may not exactly obey this monotonicity, but the modification of the conditions is straightforward. One important requirement to such a modification is that if, for example, female participation constraint binds, female bargaining power cannot decrease.}

Following Voena, 2015, I assume that men always meet women who are two years younger, and vice versa, which corresponds to the modal age difference in the US. The rest of the marriage market is as follows.

I assume that productivity of singles of each gender follows some exogenous (non-parametric) distribution over age, $F_t(z)$. Conditional on $z$, the distribution of log-assets is assumed to be truncated log-normal, namely, I assume that a latent variable is
\begin{equation}
\label{assets-partner}
\log a^*_t \sim \mathcal{N}(\mu^a_t(z),(\sigma^{a}_t(z))^2),
\end{equation}
and the realized actual value of assets is $a = a^*_t \cdot \I(a^*_t \geq \bar{a})$. This defines a distribution of available singles $F_t(z,a)$.\footnote{Regardless of characteristics, each individual draws potential partners from this distribution, so the mating is not assortative exogenously. Endogenously, however, people tend to agree more if the partner has similar characteristics, so there is endogenously generated assortativeness.}

After meeting a potential partner, the couple draws a realization of their match quality $\psi$ from a distribution $\mathcal{N}(\mu_{\psi,0},\sigma^2_{\psi,0})$.  The realization of this math quality is independent on the qualities of the partner. The match quality together with the potential partner's characteristics define distributions $\Gamma^M$ that are used in the definitions of the value function.

For women, for instance, this is the distribution $\Gamma^f_t(a^m,z^m,\psi)$, that implies that, if they agree to marry, the couple will have $a = a^m + a^f$ and $\omega = (z^f,z^m,\psi)$. The agreement decision and the resulting bargaining weights are determined using the symmetric Nash Bargaining, there are three types of matches: regular match of childless singles without an unplanned pregnancy $(np)$, match with an unplanned pregnancy $(p)$ and males meeting single mothers $(sm)$.

To lighten the notation, in this section we assume that the potential couple's characteristics $(a^f,a^m,z^f,z^m,\psi)$ are fixed, and therefore the value functions depend solely on $\theta$. The correct represetnation of the matching decision for the match type $q$ will be, however, $m_t^q(a^f,a^m,z^f,z^m,\psi)$, and I shorten this to just $m^q_t$.

\textbf{Case 1: regular match ($np$).} This is the case when two childless singles meet each other and no unplanned pregnancy happens. Upon agreement, the partners become a childless couple and receive the values $V_t^{f,c}(\theta)$ and $V_t^{m,c}(\theta)$. If disagree, they remain singles and receive $V_t^{f,s}$, $V_t^{m,s}$. Therefore the marriage decision is given by
\[m^{np}_t = \I(\Theta^{np}_t \neq \varnothing), \ \text{where} \ \Theta^{np}_t = \left\{ \theta \ : \ V_t^{f,c}(\theta) \geq V_t^{f,s}, \ V_t^{m,c}(\theta) \geq V_t^{m,s} \right\},\]
and the initial bargaining weight is defined as
\begin{equation}\label{bargaining-problem-regular}\theta^{np}_t = \argmax\limits_{\theta\in \Theta^{np}_t} \left[ V_t^{f,c}(\theta) - V_t^{f,s} \right] \times \left[ \vphantom{ V_t^{f,c}(\theta)}V_t^{m,c}(\theta) - V_t^{m,s} \right].\end{equation}

\textbf{Case 2: unplanned pregnancy ($p$).} This is the case when two childless singles meet each other and an unplanned pregnancy happens. There are hree important changes from the last case. First, both partners suffer social stigma $\phi_s$, altering their disagreement options. Second, women has a risk of becoming a single mother, and her expected value is described by \ref{sm-value}. Third, men pay child support and women receive child support, so men's value is also altered with $\E_{csm}$ operator as in \ref{man-cs}.\footnote{The timing of this bargaining is a little bit subtle. I assume that the woman cannot commit on whether to keep the pregnancy in case of the man's refusal to marry, and has to re-optimize the decision in this subgame.} The marriage decision is hence given by
\[m^{p}_t = \I(\Theta^{p}_t \neq \varnothing), \ \text{where} \ \Theta^{p}_t = \left\{ \theta \ : \ V_t^{f,ck}(\theta) \geq V_t^{f,sk?} - \phi_s, \ V_t^{m,ck}(\theta) \geq \E_{csm} V_t^{m,s} - \phi_s \right\},\]
and the initial bargaining weight is defined as
\begin{equation}\label{bargaining-problem-upp}\theta^{p}_t = \argmax\limits_{\theta\in \Theta^{p}_t} \left[ V_t^{f,ck}(\theta) - (V_t^{f,sk?} - \phi_s)  \right] \times \left[ \vphantom{ V_t^{f,ck}(\theta)}V_t^{m,ck}(\theta) - (\E_{csm} V_t^{m,s} - \phi_s) \right].\end{equation}

\textbf{Case 3: single mothers match ($sm$).} This is the case when single male meets a woman who already has a child from previous relationships. It is somewhat analogous to the previous case: if the couple agrees to marry, they become couple with a child (kids-first, for the record). However, a conceptual issue is that people generally do not like having stepchildren, and to compensate for this, I introduce a utility penalty to a potential husband $\phi_r$, which is paid as a one-time cost in case the male agrees to enter such a marriage. Additionally, as no new children are born, there is no child support assigned in this case. Therefore
\[m^{sm}_t = \I(\Theta^{sm}_t \neq \varnothing), \ \text{where} \ \Theta^{sm}_t = \left\{ \theta \ : \ V_t^{f,ck}(\theta) \geq V_t^{f,sk}, \ V_t^{m,ck}(\theta) - \phi_r \geq V_t^{m,s}   \right\},\]
and
\begin{equation}\theta^{sm}_t = \argmax\limits_{\theta\in \Theta^{sm}_t} \left[ V_t^{f,ck}(\theta) - V_t^{f,sk}  \right] \times \left[ \vphantom{ V_t^{f,ck}(\theta)} (V_t^{m,ck}(\theta) - \phi_r) - V_t^{m,s} \right].\end{equation}

\subsection{Renegotiation and Divorce Decisions\label{divorcedecision}}

The divorce decision happens before decisions on consumption, savings and labor supply, value of exogenous state $\omega = (z^f,z^m,\psi)$ is realized and couples has savings $a$ from the previous period.  At this moment individuals know their values of staying together and value of divorce. For couples with children, these are $V_t^{f,ck}$, $V_t^{m,ck}$ and $V_t^{f,dk}$ and $V_t^{m,dk}$ respectively.

Given $a$ and $\omega$ I define two sets, that represent values of $\theta$ that are satisfactory for each partner:
\[\Theta^f_t(a,\omega) = \left\{\tilde \theta \ : \ V_t^{f,ck}(a,\omega,\tilde\theta) \geq V_t^{f,dk}(a,\omega)\right\},\]
\[\Theta^m_t(a,\omega) = \left\{\tilde \theta \ : \ V_t^{m,ck}(a,\omega,\tilde\theta) \geq V_t^{m,dk}(a,\omega)\right\}.\]
and their union:
\[\Theta(a,\omega) = \Theta^f_t(a,\omega) \cap \Theta^m_t(a,\omega).\]

The divorce therefore is defined:
\begin{equation}
d_t(a,\omega) = \begin{cases} 1, \ \ \Theta(a,\omega) = \varnothing,\\
 0, \ \ \Theta(a,\omega) \neq \varnothing.
 \end{cases}
\end{equation}

In case of $d_t = 0$, the evolution of $\theta$ is defined through this set. Namely, given $\theta$, and the next period's set $\Theta_{t+1}(a',\omega')$, if $\theta \in \Theta_{t+1}(a',\omega')$ the $\theta$ is kept at the same level, otherwise, it adjusts according to which participation constraint binds to ensure the lowest deviation.%, namely define $\theta^* = \argmin\limits_{t \in \Theta_{t+1}} |t - \theta|$.

To be precise, for a non-empty set one can define $\theta^f_t(a,\omega) = \inf \Theta_t(a,\omega)$ to be the minimal bargaining power such that both participation constraints are satisfied. This corresponds to a point $\tilde{\theta}$ for which female participartion constraint exactly binds $V_t^{f,ck}(a,\omega,\tilde\theta) = V_t^{f,dk}(a,\omega)$ and $V_t^{m,ck}(a,\omega,\tilde\theta) \geq V_t^{m,dk}(a,\omega)$. In a symmetric manner I define $\theta^m_t(a,\omega) = \sup \Theta_t(a,\omega)$. Then, continuing to assume monotonicity,  
\[\theta \in \Theta_t \Leftrightarrow \theta^f_t \leq \theta \leq \theta^m_t,\]
and therefore the transitions for $\theta$ are defined through these two functions:
\begin{equation}\theta'(\theta,a',\omega') = \begin{cases} \theta, & \text{if \ }  \theta^{f\prime}(a',\omega') \leq \theta \leq \theta^{m\prime}(a',\omega'), \ \Theta'(a',\omega') \neq \varnothing,\\
\theta^{f\prime}, & \text{if \ }  \theta < \theta^{f\prime}(a',\omega'),\ \Theta'(a',\omega') \neq \varnothing,\\
\theta^{m\prime}, & \text{if \ }    \theta_t > \theta^{m'}(a',\omega'),\ \Theta'(a',\omega') \neq \varnothing,\\
\varnothing, & \text{if \ }  \Theta'(a',\omega') = \varnothing.
\end{cases}\end{equation}

This definition ensures the minimal deviation from the full commitment allocation, in which the bargaining weights are fixed. This maps into the classic limited commitment framework as in Marcet, Marimon, 2019: a Lagrange multiplier on a binding participation constraint corresponds to an increase in $\theta$ after rescaling, and the rescaling is possible because the decisions are homogeneous of degree zero in the bargaining weights.

%The divorce happens if there does not exists a pair of decision weights $\theta = (\theta^f,\theta^m)$ such that bo

\subsection{Divorce Values and Child Support\label{divorcevalues}}

When couple divorces, the assets are split evenly. This follows Voena, 2015, which provides a broader discussion of this assumption. Shortly, it is justified by the fact that most of the state either follow equal division of marital property or impose on average a division that is approximately equal, and the fact that people accumulate most of their wealth in marriage as they marry being relatively young. I did not find this to be particularly important for the qualitative results, although the property division obviously affects the divorce rates.

Instead, I assume that upon divorce both spouses incur fixed utility costs $\phi^d$. Therefore the divorce values are based on couple's state, and for the couples without children are given by 
\begin{equation}
V^{f,d}_t(a,\omega) = V^{f,s}_t(0.5\cdot a,z^f) - \phi_d, \ \ \ V^{m,d}_t(a,\omega) = V^{m,d}_t(0.5\cdot a,z^m) - \phi_d.
\end{equation}

If couple with a child divorces, the child custody is always given to mother, and the father does not get any utility from the absent child. The father, however, can be forced to pay child support. The permanent nature of productivity shocks $z$ allows to model child support as reallocation in productivities. The child support is assumed to be related solely to the male's labor income.

The child support is awarded following a divorce with probability $p^{cs,d}$, which is different between married and unmarried couples. If the child support is awarded, the evolution of productivity is $\tilde{z}^f_{t} = z^f_{t} + CS^m(z^m_t)$ and  $z^m_{t} = z^m_t - CS^f(z^f_t,z^m_t) $ for women and men respectively. Whether the child support is awarded is not known at the moment of divorce or marriage decisions. This realization does not keeps track of child's age, assuming the child support is paid indefinitely long, however, the fixed chances to avoid child support balance this. Productive males who lose the most because of indefinitely long payment period are also the most protected as they are less likely to pay relative to the case where establishing child support were up to decision of women.

To determine the exact change in $z$, assume that child support is proportion $\kappa$ of male's labor income, then the productivity loss for male $CS^m(z^m)$ sovles \[\exp(z^m_t  - CS^m +  T^m_t) = (1-\kappa)\cdot \exp(z^m_t  +  T^m_t),\] and the productivity gain of female $CS^f(z^f_t,z^m_t)$ solves \[\exp(z^f_t + CS^f +  T^f_t) = \exp(z^f_t  +  T^f_t) + \kappa \cdot \exp(z^m_t  +  T^m_t).\]

Given the values for the child support, the (expected) values of the divorce for a couple with a child are
\begin{align}
V^{f,dk}_t(a,\omega) = p^a\cdot V^{f,sk}_t(0.5\cdot a,z^f + CS^f) + (1-p^a)\cdot V^{f,sk}_t(0.5\cdot a,z^f)  - \phi_d, \\
V^{m,dk}_t(a,\omega) = p^a\cdot V^{m,s}_t(0.5\cdot a,z^m - CS^m) +  (1-p^a)\cdot V^{m,s}_t(0.5\cdot a,z^m) - \phi_d.
\end{align}

In the actual solution the resulting transition of $z$ is approximated by two-point binary distribution over the nearest grid values.

\subsection{Functional Forms}
The model assumes a set of standard CRRA utility and production functions.

Consumption's utility is standard:
\begin{equation}u(c) = \frac{c^{1-\sigma}}{1-\sigma}.\end{equation} When two spouses live in the couple, they can spend less money to achieve consumption levels $c^f$, $c^m$ then alone because of household's returns to scale. I assume that total couple's consumption expenditures are
\begin{equation}c = C(c^f,c^m) \equiv \left[ (c^f)^{1+\rho_c} + (c^m)^{1+\rho_c}\right]^{\frac1{1+\rho_c}}.\end{equation}

In a spirit of Chiappori, Mazzocco, 2017\nocite{chiappori2017static}, couple's intrahousehold problem can be simplified to picking just a total consumption expenditures. To see this, one may consider the problem
\[U(c) = \max_{c^f,c^m} \left\{ \theta^f\cdot \frac{(c^f)^{1-\sigma}}{1-\sigma} + \theta^m\cdot \frac{(c^m)^{1-\sigma}}{1-\sigma}, \ \ \text{ s.t. } \ \ c = \left[ (c^f)^{1+\rho_c} + (c^m)^{1+\rho_c}\right]^{\frac1{1+\rho_c}}\right\}.\]
After doing some algebra, it can be shown that there are functions $A(\theta)$, $k_f(\theta)$ and $k_m(\theta)$ such that
\begin{equation*}U(c) = A(\theta)\cdot \frac{c^{1-\sigma}}{1-\sigma}, \ \ c^f = k^f(\theta)\cdot c, \ \ c^m = k^m(\theta)\cdot c,\end{equation*}
and therefore utility of the couple can simply be represented as a function of total expenditures and bargaining weights, which simplifies the optimization considerably.

When the children are present, parents produce child's consumption $q$ as a public good. Both parent's money $x$ and female time $l^f$ contribute to the its production, so \begin{equation}q = f(x,l^f) \equiv \left[ x^{\lambda} + \kappa\cdot (1-l^f)^{\lambda}\right]^{\frac1\lambda}\end{equation} This function is built in such a way that if female works full-time ($l^f = 1$), $q$ just represents the monetary expenditures $x$. This also helps to interpret the subsistence constraint, as we require $q \geq \bar{q}$.

The utility of having a child consists of fixed utility benefit and part that is dependent on the child's consumption, so
\begin{equation}\label{child-q}\phi(q) = \phi_0 + \alpha\cdot \frac{q^{1-\phi_1}}{1-\phi_1}.\end{equation}

For the skills depreciation functions, the model assumes discrete choice of labor supply, which only has an extensive margin: whether or not to participate in the labor market. Participating corresponds to $l^f = 1$ and $\delta(1) = 0$, and non-participating is $l^f = \underline{l}$ and $\delta(1) = \underline{\delta}$.

\subsection{Additional Details}
All agents retire at the age of $T^R$, where they get a fixed retirement income of $b$, unconditional on their past productivity. I found this to be of minor importance for the results, as all the transitions I focus only happen early in life.

By computational reasons, the value of love shock actually received by couple is clipped to fit to be within $[\underline{\psi},\overline{\psi}]$, which are set to be 2.5 of \emph{initial} standard deviation. This can be intepreted utility contribution of $\psi$ being $u^{\psi} = \min\{\max\{ \underline{\psi}, \psi \},\overline{\psi}\}$. As standard deviation is large and $\psi$ is assumed to be a random walk, this prevents gains from marriage to increase indefinitely over the lifetime without altering the stochastic process. This way agents with two very high $\psi$ are still different: the one with the larger value has smaller probability to fall below a certain number.

I introduce non-linear taxation, that depends on couple's or single's total labor earnings and number of children, following the approximations obtained by Guner et al, 2012\nocite{guner2012taxation}. The approximations are based on social security data and capture not solely the tax schedule, but also the average utilization of the major welfare programs, including EITC and TANF.

\section{Quantification Approach}

The model has a great degree of flexibility, which is achieved by a large number of parameters. Some of them can be estimated outside the model and fixed externally, however, there is a number of crucial preference and transition parameters that cannot be inferred directly as they affect many outcomes simultaneously, for them an estimation procedure is needed.

To solve the model, all state variables and transition probabilities are discretized, and the value function iteration starting from the very last period is used. An important complication is that there is no guarantee that value functions are convex or even monotonic, although the latter can be proved in simple cases. The discreteness of marriage, divorce, fertility and labor supply decisions and the collective nature of the couple's decisions make modern Euler-equation based methods (like Iskhakov et al, 2015\nocite{iskhakov2015estimating}) difficult to apply. Adding taste shocks as a smoothing device can potentially be done, but it is not enough to obtain tractable Euler equations as the envelope theorem is not applicable to the individual values of couple's decisions. Therefore the preferred approach is maximal discretization, and the consumption-savings problems are solved by brute force global search. The details of the method are described in Appendix \ref{solution-appendix}.

In total, the model takes around three minutes to be solved and simulated on a Macbook Pro, 2015. As the routines for optimization, marriage and renegotiation are very parallelizable, it can be efficiently solved on GPUs. Running time is then 30--70 seconds, depending on the GPU model.

\subsection{Strategy}
To provide more focused evidence and also to facilitate understanding the education differences in the shotgun marriage patterns, I perform estimation on two distinct subsamples: college graduates and high school graduates. As Table \ref{share_table_0} suggests, the college graduates have substantially larger differences between kids-first and marriage-first couples in terms of divorce, and this difference is mostly expressed in the lower divorce rates for the marriage-first group. This makes comparing two separate instances of the estimated model particularly promising.

This approach is similar to Low et al, 2018, who study welfare utilization and consider only the lower-education subsample. Similar to them, yet slightly unrealistically I assume that college graduates always marry college graduates and high school graduates always marry high school graduates.

I mainly use ACS data as my key empirical results are obtained from there, however, for the purpose of external estimation of things that are not observed in ACS I utilitze SIPP data. Finally, few estimation targets are taken from literature.

\subsection{Ex-Ante Fixed Parameters}
The agents being at age 21 for females (23 for males), retire at 65 (67) with $0.4$ of median income as a pension benefits (on top of their savings) and die deterministically at 75 (77), leaving no bequest.

Following the standards in the lifecycle literature I take $\beta = 0.96$ and $R = 1/\beta$ (see ... for references).

For the labor supply side, the drift $\delta(\underline{l})$ is set to $0.09$ for college graduates and $0.06$ for high school graduates, so woman loses roughly 9\% or 6\% of her income per 1 year out of labor force. These numbers replicate structural estimates of Adda, Dustmann, Stevens, 2017\nocite{adda2017}, despite them being on German Data. Blasutto, 2020 gets similar numbers through structural estimation of a model similar to this on PSID. Additionally, those out of labor force are assumed to have $\underline{l} = 0.2$ instead of zero, so they still get one-fifth of their regular income. This assumption is mostly for computational reasons, but also reflects heterogeneous potential for home production.

For the preference side, I assume intertemporal elasticity of substitution in consumption to be a standard number $\sigma = 1.5$, and I assume that this elasticity is the same for the child's consumption (equation \ref{child-q}), so $\phi_1 = \sigma = 1.5$. For the returns to scale within a couple I fix $\rho_c = 0.23$ following Voena, 2015, which indicated mildly increasing returns to scale. In the child's quality production function $f(x,l^f)$ the substitutability between money and time is assumed to be high and is set to be $\lambda = 0.7$, as in Sommer, 2015.

For the child support enforcement, I assume that the value of the child support assigned corresponds to 20\% of male's income, that is considered to be an modal value in the US (see ...). For the enforcement strengths, I assume that $p^{a,{nm}} = 0.284$ of never married women and $p^{a,{nm}} = 0.461$ of divorced women get child support, these numbers correspond to the statistical share of eligible mothers receiving the child support in 20??, see ....

For the success in a child's conception I use the following success probabilities: 86\% at 21--24, 78\% at 25--29, 63\% at 30--34 and 52\% at 35--40 (see Rosental, 2002\nocite{rosenthal2002fertility}). These numbers reflect chances of women to get pregnant within a year if using no birth control. More modern Kennes, Knowles, 2012\nocite{kennes2012} use somewhat higher chances, but they model infertility and have taste shocks that may prevernt childbearing, so I pick the lower values to balance this. I assume that infertility is deterministic at starts at 41.

At the beginning of the life $2\%$ of college graduate women and $20\%$ of high-school graduate women have kids. I assume that by the age of 21 nobody is married, which is counterfactual, but does not determine the results. Everyone starts with zero assets, initial distributions of productivities are based on the data, as described below.

The scale of all income-like numbers is thousands of $2017$ USD. I restrict assets choices such that no one can have more that $\$ 1m$ in savings, this constraint is binding for a small share of individuals.

\subsection{Externally Estimated Parameters}

Three crucial objects to estimate from the data are the variance of income shocks, the distribution of assets and productivities singles and the income trends. The variances require observing individual over time and the distribution requires the assets data, none of which is possible using the ACS data. Therefore for them I use SIPP data, 2014 Waves 1--3, that cover three years. To be more in line with the main data, the income trends are still estimated using the ACS, as SIPP estimates are less smooth, and the described SIPP technique does not utilize dynamic structure in estimation of the trends.

\subsubsection{SIPP: Income Shocks and Distribution of Singles}
For SIPP, I impose the following restrictions on the data, replicating the approach that is done in Section \ref{empirical-section} for ACS. I treat as working those who work at least 10 hours per week. I drop people with hourly income less than one-half of the federal minimum wage and with hourly income above $95\%$ quantile of the distribution, I also drop those whose income went up or down more than four times in two consecutive years.

The estimation approach follows Meghir, Pistaferi, 2004\nocite{meghir2004income}. Endogenous choice of female labor supply is an important concern in estimating the individual productivity shocks, therefore a Heckman correction is used.

At the first stage the labor force participation regression is estimated, using presence of the mortgage as an excluded instrument (following Blundell et al, 2016\nocite{blundell2016consumption}, it is supposed to increase labor force participation regardless of productivity). At the second stage, the log wage is regressed to this Heckman correction, together with a set of dummy variables for age, state and year, and the residuals (corrected) are used to obtain the values of productivity $z$ for those in labor force. At the third stage, based on these residuals a simple GMM time-series procedure is used to split the variation in these residuals onto the permanent income components and measurement error. The details of the estimation are given in Appendix ....

Then, I use the same residuals to estimate distribution of singles. It is done in two steps. First, I estimate the distribution of $z$ non-parametrically: given the grid values at each age and $z$ that are observed in the data I can assign weights according to proportion of singles with $z$ that are closest to a certain grid point. This gives me probabilities of each productivity grid point at each age.

Then, I estimate a heteroskedastic Tobit regression of log-assets conditional on $z$, treated people with assets less than $\bar{a}$, that includes zero, as censored observations. I prefer parametric approach as assets data are noisy, and treating them non-parametrically produces undesirable patterns.  The regression specification is
\begin{equation}\label{tobit-partner}
\log a_t^* = \alpha_0 + \alpha_1\cdot t + \alpha_2 \cdot z + \alpha_3 \cdot z\cdot t + \sigma_t \cdot \varepsilon_t, \ \ \sigma_t = e^{\beta_0 + \beta_1\cdot t}, \ \ a_t = a_t^* \cdot \I(\log a_t^* \geq \bar{a}),
\end{equation}
and I take $\bar{a} = 0$ in the estimation.\footnote{Income and assets are measured in thousands of dollars, so this treats all people with net worth below \$1000 as having zero assets.} The model is estimated via standard MLE.

For males, the standard errors of productivity shocks and the distribution of singles are estimated similarly, but without the Heckman correction. For the distribution of singles, women with and without kids were split, and the share of single mothers was estimated non-parametrically (as just percentages in the data).

As SIPP data are monthly, yearly per-hour earnings were assembled out of monthly earnings and hours, and average net worth during a year was used for assets.

\subsubsection{ACS: Income Trends}
The income trends are estimated on ACS separately, where the data is treated in the same way as in Section \ref{empirical-section}. To get rid of the intensive margin of labor supply, the per-hour earnings are converted to yearly assuming that everyone works 2080 hours per year.

The trend is assumed to be non-parametric. The earnings are regressed on a set of dummy variables for each age, 21--50 for females and 23--50 for males, past this point it is assumed to be flat. The regression additionally controls for race, state, survey year and metropolitan area status. I ignore the female labor force participation margin. Theoretically, ignoring this selection causes my estimated to overestimate the real trend, on the other hand, as there is skills depreciation from being out of labor force, the agents in simulated model will be on average below the trend, and I view these two factors as compensating each other. Appendix ... provides the regression specification.

\subsection{Internally Estimated Parameters}
Things to be estimated in the model are:
\begin{enumerate}
\item Preference parameters: additive utility from having a child $\phi_0$, weight of child's consumption in utility $\alpha$, productivity of female labor at home $\kappa$, subsistence constraint for the child's consumption $\bar{q}$
\item Characteristics of unobserved match quality $\psi$: initial distribution parameters $\mu_{\psi,0}$ and $\sigma_{\psi,0}$, the variance of match quality shocks $\sigma_{\psi}$.
\item Reduced-form preference shifters: disutility of divorce $\phi_d$, disutility of marrying a single mother for male $\phi_r$, disutility of not entering a shotgun marriage (social stigma) $\phi_s$.
\item Abortion access and costs: probability to get abortion access $p^{a}$ and utility costs of abortion $\phi_a$.
\item Probabilities to meet a partner $p^{\text{meet}}_t$ and to have an unplanned pregnancy $p^{\text{preg}}_t$.  I parametrize them with values at three age points: $(p^{\text{meet}}_{21},p^{\text{meet}}_{30},p^{\text{meet}}_{40})$ and $(p^{\text{preg}}_{21},p^{\text{preg}}_{28},p^{\text{preg}}_{35})$ and interpolate with quadratic function in between these points.
\end{enumerate}

In total this gives 12 choice parameters and 6 transition probabilities. I define the resulting parameter vector to be
\begin{equation}\label{theta-eq}
\Theta = \left[ \phi_0,\alpha,\kappa,\bar{q},\mu_{\psi,0},\sigma_{\psi,0},\phi_d,\phi_r,\phi_s,p^{a},\phi_a,p^{\text{meet}}_{21},p^{\text{meet}}_{30},p^{\text{meet}}_{40},p^{\text{preg}}_{21},p^{\text{preg}}_{28},p^{\text{preg}}_{35}\right],
\end{equation}
which is 18-dimensional.

To discipline these parameters, I use overidentifying set of moment restrictions derived from the ACS. They include timings of divorce and fertility over duration of marriage, and percentage married, divorced and single with and without kids by age, with few other auxiliary things for pinning down preferences.

I use three external targets that are not available in ACS: share of child-related expenditures in labor earnings from Lino, 2001\nocite{lino2001expenditures}, percentage of unplanned pregnancies aborted from [...] and ratio of number of abortions at 30s over number of abortions at 20s from [...].

To tie the numbers together, I also introduced additional overidentifying restriction imposing certain marital sorting. For married women at 30 with income above median I compute what percentage of their husbands have income above median. In the model this is computed as $\P(z^m>0|z^f>0)$ among married couples. I refer to this as an income sorting measure.

Table \ref{targets-table} describes all the targets used and gives a rough idea of identification directions.  The total number of targets used is 95. The following section describes identification in more details.

\begin{table}[h]
\begin{center}
\caption{\label{targets-table}Estimation targets and parameters related to them.}
{\small
\begin{tabular}{p{0.5\linewidth} p{0.4\linewidth}}
\textbf{Target} &\textbf{Informative about} \\\hline
\% divorced $x$ years after marriage if kids-first, \ \ \ \ \ \ \% divorced $x$ years after marriage if marriage-first, $x \in \{1,...,10\}$ & Match quality shocks $\sigma_\psi$, divorce costs $\phi_d$, social stigma $\phi_s$  \\\hline
\% of kids-first and \% of marriage-first in female population at age $a$; \% of kids-first in population of married once \& with kids at age $a$; $a \in \{23,...,35\}$ & Unplanned pregnancy probabilities $p^{\text{preg}}_t$, meeting probabilities $p^{\text{meet}}_t$, fertility parameters $(\alpha,\phi_0,\bar q)$\\\hline
\% never married  without kids at age $a$, $a \in \{23,...,35\}$ & Meeting probabilities $p^{\text{meet}}_t$, match quality initial distribution $(\mu_{\psi,0},\sigma_{\psi,0})$ \\\hline
\% never married with kids at age $a$, \% divorced with kids at age $a$, \% divorced without kids at age $a$, $a \in \{28,...,35\}$ & Meeting probabilities, match quality initial distribution $(\mu_{\psi,0},\sigma_{\psi,0})$, match quality shocks $\sigma_\psi$, step children penalty $\phi_r$ \\\hline
\% of women with more than one marriage at 40 & Meeting probabilities $p^{\text{meet}}_t$, step children penalty $\phi_r$ \\\hline
\% of couples with kids $x$ years after marriage, $x \in \{4,...,8\}$ (if still married) & Fertility parameters $(\alpha,\phi_0,\bar{q})$ \\\hline
\% in labor force if with kids and married at 30 & Labor productivity at home $\kappa$ \\\hline
\% of child expenditures in total earnings${}^*$ & Fertility parameters $(\alpha,q)$ \\\hline
\% unplanned pregnancies aborted${}^*$, ratio of abortions at 30s to abortions at 20s${}^*$ & Abortion access and costs $(p^a,\phi_a)$\\\hline
\% husbands with income above median for women with income above median (marital sorting) & Extra identification for match quality distribution $(\sigma_\psi,\sigma_{\psi,0},\mu_{\psi,0})$ \\\hline\hline
\multicolumn{2}{p{0.9\linewidth}}{\emph{Notes:} ${}^*$ denotes external targets, the sources are described in text. All other targets are computed on ACS following the methodology of Section \ref{empirical-section}.}\\
\hline
\end{tabular}
}
\end{center}
\end{table}

\subsection{Identification}
Each parameter affects multiple targets simultaneously, and this is the reason a proper estimation procedure as opposed to simple calibration is used. Nevertheless, for each parameter there is an idea about what set of targets is the most informative about it.

Meeting probabilities $p^{\text{meet}}$ are disciplined by share of never-married people over age and the share of kids-first and marriage-first couples. Unplanned pregnancy probabilities $p^{\text{preg}}_t$ are pinned down with share of kids-first couples, especially relative, share of never married women with kids and, less importantly, by percentage of abortions.

The size of match quality shock $\sigma_{\psi}$ is identified by steepness of divorce profile by marriage duration, looking especially at the marriage first group as the least vulnerable to divorce. The overall size of the divorce rates is informative about divorce cots $\phi_d$, and the difference between kids-first and marriage-first is easily affected by the size of social stigma $\phi_s$, among other factors that I discuss later.

The initial distribution of match quality is determined by percentage of never married people over age with and without kids. The latter group is particularly important as single in the model they certainly met at least one partner. Together with the share of women with more than one marriage at 40 these numbers are also informative about the meeting probabilities. The income sorting measure provides additional information, as it essentially picks the importance of income relative to the match quality in the partner selection process.

Steepness of age profile of fertility, represented by share of the marriage-first people in general population, is informative about the role of child's consumption in utility, $\alpha$. It is also affected by the expenditure share of couples, but the latter is determined more directly with the subsistence constraint parameter $\bar{q}$, although the separate identification of two parameters is a little subtle. The constraint $\bar{q}$ is also affected by the share of kids-first couples, as it is the most direct factor making raising children alone undesirable.

Overall level of fertility from the share of marriage-first couples and percentage of couples with kids by years after marriage are informative about overall benefits from having kids $\phi_0$. Additionally, the size of $\phi_0$ versus the scale of love shock is identified by comparing the share of divorced people with and without kids.

Productivity of female labor at home $\kappa$ is almost solely determined by the level of female labor supply. Similarly, the share of unplanned pregnancies aborted determines abortion access $p^a$, and the age profile of abortions is informative about their utility costs $\phi_a$.

Utility loss for having stepchildren $\phi_r$ is determined by percentage of divorced and never married women with kids, especially relative to those without kids, as the dynamics of their relative proportion is informative about changes in the remarriage chances.

\subsection{Simulated Method of Moments}

For every combination of parameters $\Theta$, I solve the model, obtain decision rules, simulate a sample of agents and track their decisions. Based on these decisions, I get simulated data, from which I compute the targets, replicating the way they computed in the data as close as possible. Based on the targets in the data and in the model, I compute the weighted GMM-style distance, and this number represents the overall fit of the model. I use optimization routine to select parameters $\Theta$ maximizing the overall fit of the model, i.e. minimizing this distance.

Formally, let the vector of the data targets be $t$ and the simulated values of the stargets from the model be $m(\Theta)$. The simulated method of moments distance function is a scalar
\[d(\Theta) = \left[ m(\Theta) - t\right]' \cdot W \cdot  \left[ m(\Theta) - t\right].\]

The weighting matrix corresponds to standard errors of the targets in the data, inverse and squared. In a canonical GMM framework the correlation of the targets should also be computed, however, following ... I ignore the covariances and assume the matrix is diagonal, prioritizing finite sample properties of the estimates over theoretical statistical efficiency.

I use standard errors obtained from ACS, except for external moments (marked with stars in Table ...). For the external moments I arbitrarily set standard error of $0.001$, which corresponds to standard error of the ACS numbers of similar magnitude. Finally, as this is central in my model, I artificially multiply the standard errors of shares of people divorced by the years after marriage in KF and MF groups by 1/4 to prioritize fitting these targets.

The model simulates lifecycle of $15{,}000$ women. I did not find random sampling important for the targets I use as most of them are conditional on age or particular marriage duration, so for the sake of faster computations I treat women in each period as separate independent observations. To mitigate the concerns, I compute the moments three times with different sequences of the random draws and average among them.

For the optimization procedure I follow Arnoud et al, 2019\nocite{arnoud2019benchmarking}. There are many discrete decisions and there is noise coming from simulations, that makes the distance function $d(\Theta)$ non-smooth, although most of the approximated underlying decisions are expected to respond smoothly on changes in parameters.

Therefore a two-step optimization procedure TikTak is used: it first performs global search and then runs many local optimizers. First, I take $20{,}000$ candidate values for $\Theta$ from quasi-random Sobol sequence for potential parameters values (rescaled to fit some reasonably wide bounds). Among them, I pick $250$ points with the lowest value of the distance to a sequence $\{\Theta^{1i}\}_{i=1}^{250}$.

Then I run sequence of local optimization procedures. At the first stage I take the best of these 250 points as a starting point. At the stage $i = \{2,...,250\}$, the starting point is a weighted average of the best point found at stage $i-1$ and the $\Theta^{1i}$, where the weight of the best point is growing with $i$.\footnote{I use $w_b = \max\{0.1,\min\{(i/250)^{0.5}\},0.995\}$ following the original paper.}

For the local optimizers, I use non-gradient regression-based nonlinear least squares algorithm DFO-LS, that is developed in Cartis et al, 2019\nocite{dfols}. Other potential choice is Nelder--Mead, the gradient-based methods are found to be extremly inefficient because of the stochastic noise in the function.

This algorithm is easily parallelizable, if in the weighted average instead of picking the point from stage $i-1$ the optimizer just uses the best point found so far. As the time it takes to evaluate the function at one point on GPU is approximately 1 minute, running running 15--20 concurrent optimization routines simultaneously (with each evaluating function 120--180 times) delivers reasonably stable and replicable results within less than 24 hours.

\iffalse
\subsection{Non-Parametric Probability Refinement}
In the main estimation exercise I parametrize probabilities of meeting and pregnancy by two second degree polynomials. As an additional step, after using the method describe above and obtaining a solution with a satisfactory fit, I implement the following estimation procedure aimed to mitigate the effects of this assumption. The parameter vector $\Theta$, defined in \ref{theta-eq}, can be split onto
\[\Theta = [ \Gamma, P],\]
where $P$ are the pregnancy and meeting probabilities and $\Gamma$ are choice parameters. Initial estimates from the TikTak algorithm imply $\Theta_0 = [\Gamma_0,P_0]$, where $P_0$ are given by the values of the quadratic.

Given the fixed decision rules $d(\Gamma_0,P_0)$ during the model simulations I can pick sequence of new transition probabilities $P_1$ that match share of single agents and share of kids-first couples by age exactly. Then the point $\Theta_{01} = [\Gamma_0,P_1]$ is used a starting point for a new estimation of decision parameters $\Gamma$, that allows to get $\Theta_1 = [\Gamma_1,P_1]$. Given $\Theta_1$, new probabilities $P_2$ matching the data can be found, $\Theta_{12}   = [\Gamma_1,P_2]$ used as a new starting point, and the process repeats. When the difference in probabilities $|P_{n} - P_{n-1}|$ is small enough, the estimates converge. The convergence is not guaranteed, but is achieved pretty quickly in the actual problems.

To get the intuition about how data identify the meeting probabilities, imagine the following simplistic examples. Suppose the decision rules are such that one-half of all partners are accepted, and suppose $100\%$ of population are single at $22$ and $80\%$ of population are single at $23$. Then, $20\%$ of singles transition to non-singles, and this implies that $40\%$ of people meet potential partners.

Similar logic applies to the unplanned pregnancy probabilities. If $20\%$ of pregnancies end in shotgun marriages, and number of people having shotgun marriage goes from $0$ at $22$ to $1\%$ at $23$, then $5\%$ of people got unplanned pregnancies, so $p^{\text{meet}}_{22}\cdot p^{\text{preg}}_{22} = 0.05$. If $p^{\text{meet}} = 0.4$ as above, this implies $p^{\text{preg}} = 0.125$.

In the real model, the acceptance probabilities are heterogeneous and depend on the current distribution of characteristics, additionally, the value of $p^{\text{preg}}$ affects the acceptance probabilities for the general population. However, running a simple optimization procedure targeting new shares of singles and kids-first couples allows to get the values matching them exactly fairly quickly. Due to the distribution at each stage being random, there is some randomness in the probability estimates too, but running them several times and averaging provides an adequate result.

Past 35 years, where most of the transitions are over and share of singles is small, the estimation technique just assumes the fixed meeting and pregnancy probabilities at the average level of 33--35.

\fi

\section{Estimates and Fit}

In this section I present the estimates in the model on two subsamples: high school graduates and college graduates. For the purpose of further comparison, I prioritize the fit for the college graduates, as they have more evident patterns related to the shotgun marriages. I estimate the full set of the model parameters on a subsample of college graduates. Then I fix the preference parameters $\bar{\Theta} = (\sigma_{\psi},\sigma_{\psi,0},\mu_{\psi.0},\phi_0,\alpha,\kappa)$ at the level of these estimates, and re-estimate the remaining set of parameters on the data for high school. As the preference parameters directly drive fertility timing by marriage duration, share of child-related expenditures and the labor force participation, I do not use these targets in the high school estimation.

Table \ref{par-estimates} presents parameter estimates for low and high-education samples. Table \ref{model-fit-table} presents the model fit and references the fit graphs representing the model fit.  Additionally, Table \ref{ext-estimates} presents external parameter estimates for income process and distribution of singles.

The fit for the college graduates is exceptionally good, especially given the large number of targets. There only important issue is fertility timing: due to the absence of taste shocks, many couples are close to indifference between when to have children, and in particular, a number of couples, especially elder ones close to the fertility limits, try to conceive quickly after they are married. This is an artifact of the model rather than desirable property. Uncertain conception probability mitigates this slightly, but the most obvious solution is just to remove the first three years after marriage from the targets, which I did.

Since the high school model is estimated on restricted parameter set, it has generally worse fit, but still matches the main targets of the divorce rates of kids-first and marriage-first couples very decently. The share of kids-first and marriage-first couples, on the other hand, has significant discrepancy, though the overall pattern is similar. I see two reasons for this. First, a significant share of singles marry and have children earlier in life, and in the model the earliest moment married couple can have children is 22. The estimator solves this by inflating the meeting probabilities in early life, but this creates an undesirable pattern ($u$-shaped parabola) for the rest of the life. %This is possible to solve with more flexible parametrization, but this is not the main focus of mine yet.

\def\f#1{\multicolumn{2}{l}{\hspace{0.75cm}Figure #1}}
\begin{table}[h]
\begin{center}
\caption{Model Fit\label{model-fit-table}}
\begin{tabular}{l r r r r}\hline\hline
& \multicolumn{2}{c}{\textbf{College}} & \multicolumn{2}{c}{\textbf{High School}} \\\hline
& \textbf{Model} & \textbf{Data} & \textbf{Model} & \textbf{Data} \\\hline
{\footnotesize {\% divorced 1--10 years after marriage if KF and MF} } & \f{\ref{fit-col-1}} &  \f{\ref{fit-hs-1}} \\
{\footnotesize {\% KF and MF at 23--35, in population and relative} }& \f{\ref{fit-col-1}} &  \f{\ref{fit-hs-1}} \\
{\footnotesize \% with kids 4--8 years after marriage } & \f{\ref{fit-col-1}} &  \f{\ref{fit-hs-1}${}^*$}\\\hline
{\footnotesize \% divorced with and without kids at 28--35 } & \f{\ref{fit-col-2}} &  \f{\ref{fit-hs-2}}\\
{\footnotesize \% never married with kids at 28--35 } & \f{\ref{fit-col-2}} &  \f{\ref{fit-hs-2}}\\
{\footnotesize \% never married without kids at 23--35 } & \f{\ref{fit-col-2}} &  \f{\ref{fit-hs-2}}\\\hline
{\footnotesize  \% with more than one marriage at 40} & 6.8 & 11.9 & 9.6 & 17.1 \\
{\footnotesize \% in labor force at 30 if married and kids} & 74.2 & 74.0 & 53.6 & 54.7${}^*$ \\
{\footnotesize mean \% of child expenditures in total earnings} & 40.6 & 40.0 & 39.3 & 40.0${}^*$ \\
{\footnotesize \% unplanned pregnancies aborted} & 39.9 & 40.0 & 37.3 & 40.0 \\
{\footnotesize abortions in 30s as \% of abortions in 20s} & 38.5 & 38.5 & 39.5 & 38.5 \\
{\footnotesize \% husbands above median if wife above median} & ... & ... & ... & ...\\
\hline
\multicolumn{5}{p{0.85\linewidth}}{\footnotesize \textbf{Notes:} KF refers to the kids-first women, MF refers to the marriage-first women. 23--35 refers to all ages from 23 to 35 and so on. ${}^*$ shows the numbers not targeted in the high school calibration.}\\\hline\hline
\end{tabular}

\end{center}
\end{table}

%One important issue for the high school group is that many people with lower education marry and have kids early, but the lifecycle in the model starts at 21. This can be seen by very high assigned probabilities in early life. Overcoming these issues is not hard, but introducing differences in the time bounds makes comparisons subtle.

%Appednix ... presents the unrestricted estimation, which has slightly better but also potentially less comparable results.

\begin{table}
\begin{center}
{\small
\caption{Estimated parameters.\label{par-estimates}}
\begin{tabular}{l p{0.35\linewidth} l l}\hline
\textbf{Parameters}  & \textbf{Description} &  \textbf{College} & \textbf{High School (R)}  \\\hline
$(\phi_0,\alpha)$ & Fertility preferences & $(0.30,0.45)$ & $(0.30^*,0.45^*)$ \\
$q$ & Required childcare costs & $34.39$ & $1.23$ \\
$\sigma_\psi$ & Match quality shock size & $0.67$ & $0.67^*$ \\
$(\mu_{\psi,0},\sigma_{\psi,0})$ & Match quality initial distribution & $(1.28,-1.44)$ & $(1.28^*,-1.44^*)$ \\
$(p^{\text{meet}}_{21},p^{\text{meet}}_{30},p^{\text{meet}}_{40})$ & Meeting probabilities (quadratic) & $(0.33,0.86,0.50)$ & $(0.81,0.30,1.00)$ \\
$(p^{\text{preg}}_{21},p^{\text{meet}}_{28},p^{\text{meet}}_{35})$ & Unplanned preg. probabilities & $(0.016,0.035,0.033)$ & $(0.15,0.06,0.50)$ \\
$(p^a,\phi_a)$ & Abortion access and costs & $(0.86,6.3)$ & $(1.00,0.02)$ \\
$\phi_d$ & Utility costs of divorce & $14.9$ & $10.1$ \\
$\phi_s$ & Shotgun marriage stigma & $5.3$ & $6.3$ \\
$\phi_r$ & Utility loss from stepchildren & $20.9$ & $20.6$ \\
\hline
\multicolumn{4}{p{0.95\linewidth}}{\emph{Notes:} ``High School'' calibration restricts the preference parameters market with ${}^*$ to be the same as for college graduates, and ignores several dimensions of fit.}\\\hline
\end{tabular}
}
\end{center}
\end{table}

\begin{table}
\begin{center}
{\small
\caption{Extrental Estimates.\label{ext-estimates}}
\begin{tabular}{c c}\hline
\textbf{College} & \textbf{High School} \\\hline
\multicolumn{2}{c}{\textit{Income shocks and initial standard deviation (SIPP, equations \ref{z-male-eq} and \ref{zoutlf}):}}\\\hline

\begin{tabular}{l l l}
& $\sigma_{z}$ & $\sigma_{z,0}$ \\
Female & 0.20 & 0.43 \\
Male & 0.16 & 0.42 \\
\end{tabular}

&

\begin{tabular}{l l l}
& $\sigma_{z}$ & $\sigma_{z,0}$ \\
Female & 0.18 & 0.18 \\
Male & 0.17 & 0.23 \\
\end{tabular}

\\\hline
\multicolumn{2}{c}{\textit{Income trends (ACS, equation \ref{trends-eq}):}}\\\hline
\\
\multicolumn{1}{c}{
\includegraphics[scale=0.5]{../pics/earnings_trend_col.pdf}
}
&
\multicolumn{1}{c}{
\includegraphics[scale=0.5]{../pics/earnings_trend_hs.pdf}
}
\\\hline
\multicolumn{2}{c}{\textit{Marriageable singles' log-assets distribution (SIPP, equation \ref{tobit-partner}):}}\\
\hline
\begin{tabular}{l l}
& $(\alpha_0,\alpha_1,\alpha_2,\alpha_3;\beta_0,\beta_1)$ \\
Female & $(0.50, 0.22, 0.89,0.09; 1.15,-0.014)$ \\
Male & $(1.07,0.20,1.31,0.06; 1.06,-0.011)$ \\
\end{tabular}
&
\begin{tabular}{l l}
& $(\alpha_0,\alpha_1,\alpha_2,\alpha_3;\beta_0,\beta_1)$ \\
Female & $(0.53,0.065,1.72,0.005;0.81,0.005)$ \\
Male & $(-0.57,0.16,2.48,- 0.059;1.17,-0.011)$ \\
\end{tabular}
\\\hline\hline
\end{tabular}
}
\end{center}
\end{table}

\section{Why Marriages Differ?}
What in the model leads the kids-first couples divorce more often? There are several answers that I explore throughout this section, but the first and the most direct force is the pregnancy shocks. When an unplanned baby appears, the value of disagreement for the woman drops because of multiple factors in the model. The main factors determining responses to the shock are the social stigma and the risk to become a single mother, associated with increased expenditures and worsened marital perspectives. This alters the bargaining of the new couple, shifting the people's options towards agreement. As a result, some people marry precisely because they had a child, these people can be referred to as compliers.

\subsection{Agreement Thresholds}
From the prospective of the distribution of couples, the unplanned pregnancy lowers the agreement threshold: minimal level of match quality required to agree to enter the relationship. In this section I discuss the dynamics of this threshold starting from its definition.

The right metrics to think about bargaining problems like \ref{bargaining-problem-regular} is the egalitarian bargaining surplus, defined as
\begin{equation}M = \max_{\theta} \min\{V^{f,c}(\theta) - V^{f,s}, V^{m,c}(\theta) - V^{m,s}\}.\end{equation}
The expression under the maximum operator represents the gain of the party who is less willing to agree. Unlike the Nash bargaining surplus, this object is well-defined and smooth even in the event of disagreement. The $\theta$ maximizing this surplus is different from the Nash bargaining solution, but is close to it when $M$ is around zero. The couple agrees to marry if $M\geq 0$ and disagrees if $M<0$. A very analogous representation applies to the divorce case.

Considering couples with different initial draw of $\psi$ I define $M(\psi)$. The point $\psi_m$ such that $M(\psi_m) = 0$ represents the marriage threshold: the minimal level of relationship quality such that the couple agrees to marry.\footnote{The simulated model is solved on discrete grid, therefore the exact value can be obtained by simulations.} For established couples and their bargaining problem, a similar point $\psi_d$ such that $M(\psi_d) = 0$ represents the divorce threshold.

The interaction of two thresholds, $\psi_m$ and $\psi_d$, is informative about marriage creation and dissolution. The lifecycle of marriage can be imagined this way: the married couple gets an initial value of $\psi > \psi_m$. After that, the value of $\psi$ changes over time, and the couple divorces if it ever happens that $\psi < \psi_d$. $\psi_d$ itself moves with couple's shocks and optimal responses to them.

The effect of the marriage threshold is direct: fall in it generates more couples with low $\psi$, indicating the risky marriages, other things equal. The fact that shotgun marriage couples have lower $\psi$ is not enough for them to divorce more: this reasoning assumes they have the same divorce threshold than the regular couples. If we ignore the assets and productivity heterogeneity, this is true: after having kids, the couples in the model are identical. However, as couples who have kids in marriage choose this based on their characteristics, the ``average'' divorce threshold for them can be different from the shotgun marriage couples for which the childbearing is more random. [In simulated model, however, this difference in divorce thresholds was not systematic. TBA]

The left panel of Figure \ref{upp-tholds-graph} plots the agreement thresholds $\psi_m$ for college females 25 years old with no assets by their productivity $z^f$, meeting male with median income and zero assets. The bottom line represents the marriage threshold $\psi_m$ for the case of unplanned pregnancy, and the top shows the threshold for the regular case. The line in the middle shows the same agreement threshold net for a hypothetical unplanned pregnancy match not affected by social stigma $\phi_s$ (considering bargaining problem \ref{bargaining-problem-upp} and setting $\phi_s = 0$, while keeping the value functions intact).

The right of Figure \ref{upp-tholds-graph} panel serves for additional illustration. It shows literal value functions corresponding to each of the scenarios on the lest. It illustrates that social stigma is a major symmetric factor that affects disagreement values of males and females. Net of the social stigma, disagreement value of female still falls quite a lot, and disagreement value of male falls a little, which is mostly due to exposure to the child support. Unplanned pregnancy also changes the values of agreement: females are strictly worse of in the case of pregnancy, and males prefer having child immediately in marriage if their bargaining weight is sufficiently high.

The graphs reveal three things. First, indeed, in the estimated model people who had an unplanned pregnancy agree to marry for the lower values of $\psi$. Second,  social stigma explains the most, but not all of the drop in the agreement threshold. Third, there is mild upward trend in productivity: more productive women are more picky towards their partners, though the magnitude of this seems small relative to the effects of the pregnancy. Fourth, unplanned pregnancy is generally a negative event for females and have uncertain value for males.

\begin{figure}[h]
\caption{Agreement thresholds and impact of unplanned pregnancy\label{upp-tholds-graph}}
\begin{center}
\begin{tabular}{c c}
\includegraphics[scale=0.5]{../pics/tholds_upp.pdf} & \includegraphics[scale=0.5]{../pics/change_upp.pdf}
\end{tabular}
\end{center}
\end{figure}

\subsection{Distinguishing Mechanisms}
For the further understanding and quantification of the results I consider several counterfactual scenarios, where I solve and simulate the model with parameter changes. I summarize the changes in the following experiments:
\begin{itemize}
\item \textbf{Baseline (BL)}. Estimated model, for comparison.
\item \textbf{No stigma (NS)}. Setting $\phi_s = 0$ to remove the pressure to enter shotgun marriage.
\item \textbf{Fertility control (FC)}. Removing abortion costs $\phi_a$ and granting full abortion access $p^a = 1$, so females can perfectly choose whether to be a single mother.
\item \textbf{No remarriage penalty (NR).} Setting $\phi_r = 0$ so women with children have ex-ante the same marital prospectives as without children.
%\item \textbf{No skills depreciation (SD).} Setting $\delta(1) = 0$ so women do not lose skills while taking care of the children.
\item \textbf{Free divorce (FD).} No divorce costs.
\item \textbf{No divorce (ND).} Infinite divorce costs.
%\item \textbf{No pay gap (PG).} Setting wage process, variance and singles' distribution of women identical to men.
\end{itemize}

These scenarios are purely counterfactual, and are aimed to asses various potential determinants of differences between kids-first and marriage-first couples. Through the experiments, I look at the following metrics:
\begin{itemize}
\item Share of divorced women 10 years after marriage is kids first and marriage first.
\item Share of kids-first women at 30 (among those married once and having kids)
\item Percentage of unplanned pregnancies aborted.
\item Percentage of single mothers among all mothers at 35.
\end{itemize}

The top of Table \ref{diff-sources} shows the reaction of the model to the countefactual experiments in the calibration for the college graduates. Focusing on the differences between the divorce share, there are two most important contributors: social stigma against not marrying in case of pregnancy and fertility control --- the ability to terminate the pregnancy in case it happens (or to prevent it from happening). Social stigma alone explains around one-third of the difference, and the fertility control explains a bit less. Remarriage penalty is also an important contributor, though it goes in a tricky direction: together with decreasing divorces for kids-first it increases them for marriage-first couples, and this allows the groups to become more similar. This suggests that remarriage prospectives are also an important determinant of the differences in divorce.

The bottom part provides similar comparison for the high school graduates. Interestingly, for them things seem to behave slightly different: fertility control has minor impact, and social stigma behaves unpredictable: it lowers the divorce rate in both groups. This is consistent with the idea that non-college people face a lot of unplanned pregnancies, and not being pushed towards them makes them more picky in case they receive a good match. Eliminating remarriage penalty, however, works almost exactly as needed, making two groups more similar to each other.

To summarize, social stigma and limited costly abortions make kids-first marriages perform worse, and remarriage penalty makes both kinds of childbearing marriages more stable. Interaction of this factors causes the differences in both education groups.

Apart from divorce, the important angle is the consequences of unplanned pregnancies, that can be either a kids-first marriage, single motherhood or abortion. The results here work in a predictable direction: removing social stigma or adding full fertility control decrease proportion of marriages that are kids-first. The fertility control obviously increases the percentage of abortions, interestingly, it does not become 100\%: some college graduate women choose to be single mothers.

\begin{table}[h]
\caption{Counterfactual experiments: understanding the sources of the differences\label{diff-sources}}
\begin{center}
\begin{tabular}{l r r r r r r}\hline\hline
\multicolumn{7}{c}{\textbf{College graduates}}\\\hline
\textbf{Experiment} & \textbf{BL} & \textbf{NS} & \textbf{FC} & \textbf{NR} & \textbf{FD} & \textbf{ND} \\
\textit{divorced in 10 years, kids-first} & 21.1 & 14.5 & 16.7 & 13.9 & 56.6 & 0.0 \\
\textit{divorced in 10 years, marriage-first} & 6.2 & 5.8 & 5.9 & 11.9 & 1.9 & 0.0 \\
\textit{kids-first at 30} & 11.5 & 8.1 & 8.5 & 12.6 & 36.4 & 7.6 \\
\textit{unplanned pregnancies aborted} & 39.9 & 48.4 & 76.6 & 1.0 & 35.6 & 41.8 \\
\textit{single mothers at 35} & 13.2 & 13.8 & 6.9 & 14.9 & 19.6 & 13.6 \\\hline
\multicolumn{7}{c}{\textbf{High school graduates}}\\\hline
\textit{divorced in 10 years, kids-first} & 21.7 & 13.7 & 21.6 & 22.0 & 37.8 & 0.0 \\
\textit{divorced in 10 years, marriage-first} & 15.9 & 2.0 & 15.8 & 20.2 & 7.3 & 0.0 \\
\textit{kids-first at 30} & 36.5 & 49.5 & 36.5 & 47.7 & 63.3 & 35.7 \\
\textit{unplanned pregnancies aborted} & 37.3 & 87.3 & 37.7 & 0.2 & 34.0 & 10.4 \\
\textit{single mothers at 35} & 42.8 & 49.2 & 42.7 & 32.3 & 50.3 & 44.4 \\\hline
\multicolumn{7}{l}{{\footnotesize \textit{Notes:} all numbers are percentages.}}\\\hline
\multicolumn{7}{p{0.8\linewidth}}{{\footnotesize \textit{Names:} \textbf{BL} -- baseline, \textbf{NS} -- no social stigma, \textbf{FC} -- perfect fertility control, \textbf{NR} -- no remarriage penalty for single mothers, 
%\textbf{SD} -- no skills depreciation when out of labor force,
 \textbf{FD} -- costless divorce, \textbf{ND} -- no divorce at all
 %\textbf{PG} -- no gender pay gap.
 }}\\\hline
\end{tabular}
\end{center}
\end{table}

\subsection{Causal Effects\label{causal-section}}

The model is capable of delivering causal answers, although the definition of causality in family economics questions is ambiguous. Lundberg et al, 2016\nocite{lundberg2016family} provides an excellent discussion. Indeed, unlike in classic randomized experiments, ``causal effect of marriage'' does not immediately correspond to a well-defined counterfactuals: one can compare people marrying right now to postponing marriage, or to marrying a different partner, or to never marrying at all. A good thing about the estimated model is that it allows to do most of these comparisons, once we define which one is to look at.

One obvious angle in a context of the model is to consider a causal effect of receiving an unplanned pregnancy shock, i.e. meeting a partner and getting pregnant at a certain moment of time as opposed not not getting a shock at this particular moment and facing all the regular uncertainties later. To do this, I consider the following experiment. I generate a population of single women from the exact distribution used in the model simulation. Then, I alter their sequence of shocks such that all this population meets the first partner at the age of $25$ and either gets pregnant  or does not get pregnant.

There are multiple ways to compare the two scenarios of getting and not getting pregnant. I pick two of them that I find reasonable. First, a direct way is to consider the outcomes for the population as a whole, keeping in mind that getting a partner draw does not always convert to marriage, and getting pregnant does not always imply childbearing. This gives an idea of a causal effect of an unplanned pregnancy.

A more focused approach is to compare people who agreed to enter the kids-first marriage to their alternative selves --- people with the same sequences of random shocks in the scenario where no pregnancy happened.  This can be viewed as a causal effect of a kids-first marriage.

Table \ref{causal-table} shows the outcomes of unplanned pregnancy in terms of divorce, both unconditional (column Offered) and conditional (column All KF) on entering a kids-first marriage. Further, it provides the decomposition, separating those who married because they had kids, would have married anyway and would have never married each other.

The table gives an idea of the causal content of the observations in the empirical part of the paper. Entering a kids-first marriage  increases chance of having a divorce in the future about 1.5 times. This, however, is mostly driven by the compliers: those who got pregnant but refused to marry are \textit{less} likely to be divorced in the future, and those who would have married anyway face little impact of the pregnancy.

The share of compliers in all kids-first exceeds a half for college graduates and is around forty percent for high school graduates. This suggests that very large share of kids-first people marry \emph{because} they had a child, which is confirmed by the large shift in agreement thresholds in the previous parts.

The overall impact of unplanned pregnancy in terms of divorce is more subtle: statistically, for college graduates, less people who have unplanned pregnancy get a divorce, but this happens precisely because many of them never get married, as indicated by the percentage of married.\footnote{One possible alternative comparison is comparing the rates of ever divorced if ever married, but this makes this groups to similar to the KF.} For high school graduates this effect of non-marriage is weaker, and therefore unplanned pregnancy itself increases the chances of divorce.

Overall, entering a marriage following an unplanned pregnancy increases divorce chances substantially. Receiving unplanned pregnancy unconditional on its outcome decreases chances of being married in the future and greatly increases chances to be a single mother. Causal effects of shotgun marriage on divorce is of 14 percentage points out of the base of 32\% for college graduates and 13 percentage points out of the base of 41\% for high school graduates.

%Finally, a little less obvious outcome of singleness is presented in the table. One may believe that divorce itself is not crucial: high divorce rates may be indicative of finding a better partner in the future. From prospective of the model, however, having an unplanned pregnancy at 25 increases chances of being a single mother in general population more than three times, and this pattern is true in all groups regardless of reaction to the shock.

\begin{table}[h]
\caption{Causal effects of unplanned pregnancy and their decomposition.\label{causal-table}}
\begin{center}
\begin{tabular}{l r r r r r}
\hline\hline
& \textbf{Offered} & \textbf{All KF} & \textbf{Compliers} & \textbf{Always} & \textbf{Never} \\\hline
\textit{Marry if pregnant} & $+,-$ & $+$ & $+$ & $+$ & $-$ \\
\textit{Marry if not pregnant} &  $+,-$ & $+,-$ &  $-$  & $+$ & $-$ \\ \hline
\multicolumn{6}{c}{\textbf{College graduates}}\\\hline
\textit{Share in population}& 100 & 33.9 & 18.3 & 15.6 & 66.1\\\hline
\multicolumn{6}{c}{\textit{Ever divorced by the age of 50}} \\
\textit{...if pregnant at 25} & 27.8 & 45.4 & 53.9 & 35.4 & 18.8\\
\textit{...if not pregnant at 25} & 29.7 & 31.7 & 27.3 & 36.9 & 28.6\\\hline
\multicolumn{6}{c}{\textit{Married to any partner at 50}} \\
\textit{...if pregnant at 25} & 57.2 & 62.9 & 56.5 & 70.4 & 54.2\\
\textit{...if not pregnant at 25} & 70.0 & 72.0 & 68.7 & 75.9 & 68.9\\\hline
\multicolumn{6}{c}{\textit{Single mother at 50}} \\
\textit{...if pregnant at 25} & 38.6 & 37.1 & 43.5 & 29.6 & 39.4\\
\textit{...if not pregnant at 25} & 21.3 & 21.1 & 21.8 & 20.3 & 21.4\\\hline
\multicolumn{6}{c}{\textbf{High school graduates}}\\\hline
\textit{Share in population}& 100 & 40.7 & 16.4 & 24.3 & 59.3\\\hline
\multicolumn{6}{c}{\textit{Ever divorced by the age of 50}} \\
\textit{...if pregnant at 25} & 41.3 & 54.1 & 66.2 & 45.9 & 32.5\\
\textit{...if not pregnant at 25} & 37.1 & 40.7 & 33.2 & 45.7 & 34.7\\\hline
\multicolumn{6}{c}{\textit{Married to any partner at 50}} \\
\textit{...if pregnant at 25} & 62.0 & 63.6 & 58.2 & 67.2 & 60.9\\
\textit{...if not pregnant at 25} & 64.8 & 68.8 & 63.2 & 72.7 & 62.1\\\hline
\multicolumn{6}{c}{\textit{Single mother at 50}} \\
\textit{...if pregnant at 25} & 35.0 & 36.4 & 41.8 & 32.8 & 34.0\\
\textit{...if not pregnant at 25} & 30.3 & 26.8 & 31.9 & 23.3 & 32.7\\
\hline
\multicolumn{6}{p{0.85\linewidth}}{\footnotesize \textit{Notes:} this compares two identical populations, in which everyone met their first potential partners at 25 and either gets or does not get an unplanned pregnancy. \textbf{All KF} refers to subpopulation of those who agreed to enter the marriage following the pregnancy. Among them, \textbf{Compliers} are those who disagreed in case of no pregnancy and \textbf{Always} are those who agreed under both scenarios. Group \textbf{Never} represents those who disagreed both with and without the pregnancy. This is the calibration for the college graduates, the numbers are percentages.}\\\hline
\end{tabular}
\end{center}
\end{table}

\subsection{Aggregate Effects}
The model allows to quantify the aggregate effects of unplanned pregnancies on the behavior and outcomes. The counterfactuals above talk about what happens if people receive the shock, which is a small share of those who receives marital offers. Less then half of these people actually enter the kids-first marriages.

However, the risks of unplanned pregnancies affect everyone, and not only those who actually enters the kids-first marriages. People change their behavior in response to the chance of pregnancy. In particular, we may expect that if some matches are risky, people will be less picky in general: skipping a potential partner means more exposure to the risk. Figure \ref{anticipation-graphs} provides the evidence of this: if people anticipate an unplanned pregnancy (even if it does not come), they got married earlier and have their children faster. This is especially true for high school graduates.

As general population is larger than just the kids-first couples, the anticipation effects are expected to be more relevant for general population than just the effects of kids-first marriages. Table \ref{upp-table} confirms this. Column ``NE'' (never experienced) refers to the baseline model where in simulations no unplanned pregnancies happen, but in the optimal decisions people still anticipate them. Roughly this also corresponds to simulation results of those who never had an unplanned pregnancy. Column ``NA'' (never anticipated) refers to the model where probability of unplanned pregnancy is zero and all agents know this.

The table shows that elimination of unplanned pregnancies creates some delay in marriage and prevents some of earlier divorces, though the overall effect of them on divorce is around one percentage point for college and around for percentage points for high school agents. It also gives an idea about how pregnancies affect percentage of single mothers: eliminating them drops this number by around one-third for high school and more than half for college. To some extent it is predictable, on the other hand, it gives an idea is large share of single mothers is still generated by divorce of existing couples. The table reports the count of single mothers in total population, the effect of count of them among mothers is somewhat offset by the decline in fertility as seen on the graphs. Note that the model is simulated such that there is some initial count of single mothers in the start of the lifecycle, without it the effects are more drastic.

\begin{table}
\begin{center}
\caption{Effects of unplanned pregnancies and their anticipation.\label{upp-table}}
\begin{tabular}{l | r r r | r r r}\hline
& \multicolumn{3}{c|}{\textbf{High School}} & \multicolumn{3}{c}{\textbf{College}} \\\hline
& \textbf{BL} & \textbf{NE} & \textbf{NA} & \textbf{BL} & \textbf{NE} & \textbf{NA} \\\hline
\textit{\% ever married at 25}  & 39.5 & 37.6 & 36.2 & 25.9 & 25.4 & 23.5 \\
\textit{\% ever married at 35}  & 68.7 & 67.2 & 58.8 & 78.9 & 79.9 & 76.1 \\\hline
\textit{\% ever divorced at 30}  & 8.0 & 7.1 & 6.8 & 4.4 & 4.1 & 3.5\\
\textit{\% ever divorced at 50}  & 35.3 & 35.5 & 31.9 & 32.3 & 32.7 & 31.4\\\hline
\textit{\% single mothers at 35}  & 29.8 & 23.5 & 21.3 & 8.4 & 3.9 & 3.4 \\\hline
\multicolumn{7}{c}{\footnotesize \textbf{Welfare compared to BL (1000s of 2017 USD in savings)}}\\\hline
\textit{median male at 23} & &  &  500 &  & &   222 \\
\textit{median female at 21} &  &  &   432 &  &  &   345 \\
\hline
\multicolumn{7}{p{0.675\linewidth}}{\footnotesize \emph{Notes:} \textbf{BL} is the baseline estimation, \textbf{NE} is when unplanned pregnancies are anticipated, but never experienced,  \textbf{NA} is when unplanned pregnancies are neither anticipated nor experienced. Welfare computing details are given in Appendix \ref{welfare-details}. ``$500$'' is the upper bound for welfare change. }\\\hline\hline
\end{tabular}
\end{center}
\end{table}

\begin{figure}[p]
\caption{Anticipation of unplanned pregnancies.\label{anticipation-graphs}}
\begin{center}
\begin{tabular}{c c}
\includegraphics[scale=0.5]{../pics/anticipation/col/evermar.pdf} & \includegraphics[scale=0.5]{../pics/anticipation/col/kids_by_duration.pdf} \\
 \includegraphics[scale=0.5]{../pics/anticipation/hs/evermar.pdf} & \includegraphics[scale=0.5]{../pics/anticipation/hs/kids_by_duration.pdf}\\
 \multicolumn{2}{p{0.9\linewidth}}{\footnotesize \emph{Notes:} \textbf{never experienced} refers to the calibration where unplanned pregnancies do not happen, but agents still anticipate them. \textbf{never anticipated} refers to the calibration where unplanned pregnancies do not happen and everyone knows this.}
\end{tabular}
\end{center}
\end{figure}

% Marriage promotion versus birth control
% Doubling stigma is a metaphor for the marriage promotion. Other ways of reaching it is monetary incentives and child support obligations. Finally, just cheaper divorce promotes marriage. Nothing of it really reaches any good outcomes.
% Better birth control is a uniform good.

%
%\subsection{Consequences for Aggregate Distribution}

\subsection{Origins of The Education Gradient}
There are two things that make high school graduates different in the model: proportion of kids-first and marriage-first marriages and the more similar chances of divorce between kids-first and marriage-first groups. As the model can match the patterns in both groups reasonable well, analyzing what generates differences in patterns can provide a valueable insight.

I perform the following exercise: I take the estimated model for college graduates, and then change its parameters to the values of the high school graduates. The changes are performed in isolation, so everything except the changed parameter is kept at the level for college graduates.The blocks of parameters I use are income process characteristics, meeting probabilities, pregnancy probabilities and the estimated disutilities for divorce, remarriage, skipping a shotgun marriage and abortion.

Table \ref{educ-diffs} presents the results. I present four outcomes: population shares of kids-first and marriage-first couples and their observable divorce rates. It reveals, that no single factor can explain all the differences. Income, meeting and pregnancy probabilities seem to be responsible for the differences in proportion of kids-first couples. Note that pregnancy probabilities alone are not enough: they do increase the share of kids-first people, but have the opposite effects of share of marriage-first couples.

As to the divorce, the key component of the difference is more divorces for marriage-first couples. Larger pregnancy probability and smaller estimated costs of divorce for high school graduates provides most of desirable effects. The latter is mechanical, but the former is more surprising and is potentially related to the anticipation effects and fertility delays. Interestingly, the subsistence constraint also plays some role: it is virtually absent for high school graduates, and ``untying'' it for the college graduates makes them divorce more.

The kids-first and marriage-first groups for college and high school people are different in composition and sizes. Table \ref{causal-table}, however, gives an idea that \emph{causal} effects of shotgun marriages of divorce rates are similar. The shotgun marriage is still a choice conditional on receiving an unplanned pregnancy, and as the choice is made differently comparison of this numbers is subtle. Unconditional effects of pregnancies on divorce are different, but both do not have huge magnitude. Finally, the effects of pregnancy on future marital status are larger for college graduates, and this is well-explained with the differences in the remarriage penalty.

\begin{table}[p]
\begin{center}
\caption{Understanding differences between education groups\label{educ-diffs}}
\begin{tabular}{l r r | r r r r r r r r}\hline\hline
 \multicolumn{11}{c}{\textbf{Starting from college graduates estimates}}\\\hline
 & \multicolumn{2}{c|}{Estimates} & \multicolumn{8}{c}{College with parameters from HS}\\\hline
 & \textbf{Col} & \textbf{HS} & \textbf{I} & \textbf{M} & \textbf{P} & \textbf{D} & \textbf{R} & \textbf{S} & \textbf{A} & \textbf{Q} \\\hline
\multicolumn{11}{c}{\footnotesize percentage divorced 10 years after marriage}\\\hline
{\footnotesize ...if kids first} & 21.1 & 21.7 & 10.5 & 24.0 & 18.6 & 30.1 & 20.5 & 22.5 & 16.7 & 20.7 \\
{\footnotesize ...if marriage first} & 6.2 & 15.9 & 3.0 & 4.9 & 10.9 & 7.5 & 6.5 & 6.3 & 5.9 & 7.0 \\\hline
%{\footnotesize {...ratio KF/MF}} & {5.2} & {1.6} & \textbf{3.0} & 28.5 & \textbf{3.0} & 16.8 & \textbf{3.8} & 5.3 & 5.3 & \textbf{4.8} \\\hline
\multicolumn{11}{c}{\footnotesize share of kids-first and marriage-first in population}\\\hline
{\footnotesize ...kids-first, at 35} & 5.2 & 17.8 & 9.0 & 3.5 & 15.0 & 5.8 & 5.2 & 5.8 & 3.9 & 5.5 \\
{\footnotesize ...marriage-first, at 35} & 50.7 & 27.7 & 22.3 & 35.2 & 55.8 & 41.4 & 51.6 & 51.0 & 51.3 & 53.3\\\hline
%\multicolumn{11}{c}{\textbf{Starting from high school graduates estimates}}\\\hline
% & \multicolumn{2}{c|}{Estimates} & \multicolumn{8}{c}{HS with parameters from college}\\\hline
% & \textbf{HS} & \textbf{Col} & \textbf{I} & \textbf{M} & \textbf{P} & \textbf{D} & \textbf{R} & \textbf{S} & \textbf{A} & \textbf{Q} \\\hline
%\multicolumn{11}{c}{\footnotesize percentage divorced 10 years after marriage}\\\hline
%{\footnotesize ...if kids first} & 12.2 & 10.8 & 23.9 & \textbf{9.0} & \textbf{6.7} & \textbf{4.4} & \textbf{11.0} & 21.6 & \textbf{11.9} & \textbf{10.7} \\
%{\footnotesize ...if marriage first} & 7.6 & 2.1 & \textbf{5.5} & \textbf{3.6} & \textbf{0.8} & \textbf{2.0} & \textbf{2.3} & 9.6 & 8.0 & \textbf{6.7} \\\hline
%{\footnotesize {...ratio KF/MF}} & 1.6 & 5.2 & 4.3 & 2.5 & 8.1 & 2.2 & 4.9 & 2.2 & 1.5 & 1.6 \\\hline
%\multicolumn{11}{c}{\footnotesize relative share of kids-first marriages}\\\hline
%{\footnotesize ...at 35} & 46.2 & 10.5 & \textbf{22.0} & 57.9 & \textbf{40.4} & 47.2 & 52.4 & 48.3 & 46.9 & 48.2 \\\hline
\multicolumn{11}{p{0.9\linewidth}}{\textit{Notes}: \textbf{I} --- income process: trends, variances and the assets-income distribution of available singles, \textbf{M} --- meeting probabilities of potential partners, \textbf{P} --- probabilities to get an unplanned pregnancy following a partner's meeting,
\textbf{D} --- divorce costs $\phi_d$, \textbf{R} --- remarriage penalty for single mothers $\phi_r$, \textbf{S} --- social stigma towards entering a shotgun marriage $\phi_s$, \textbf{A} --- abortion costs and access probabilities $(p^a,\phi_a)$, \textbf{Q} --- required childcare expenditures $\bar{q}$.}\\
%\multicolumn{11}{p{0.875\linewidth}}{Numbers in \textbf{bold} show the changes in the desirable direction.}\\
\hline
\end{tabular}
\end{center}
\end{table}

\subsection{Childbearing and Divorce Chances\label{divorce-fertility-general}}

How does divorce risk affect childbearing, and how does childbearing affects divorce risks? Statistically, marriage-first women have lower divorce rate than general population of ``not-kids-first'' (i.e. marriage-first or childless). Can the model deliver any causal predictions here? Yes, in both ways.

\paragraph{Kids $\Rightarrow$ less divorce.} First, other things equal, childbearing decrease divorce chances, at least for those who tried to conceive the child. As the conception is random in the model, it provides a natural comparison: we can take the agents who conceived the child with their counterfactual selves in the case they never had a successful conception.

The experiment is analogous to the one in Section \ref{causal-section}, but without unplanned pregnancies. I take two identical populations in which everyone meets their partner at 25, then in one population the conception probabilities are (unexpectedly) altered to be zero (never conceive) and in another one they are set to one (always conceive if they want to).

Figure \ref{conception-divorce} shows the percentage of divorced at 40 among those who agreed to marry at 25. On horizontal access I plot the year the first conception decision is made. Solid line represents the percentage of ever divorced if the conception is successful, the dashed line shows this percentage if the conception (and all subsequent ones) fail. On average, successful conception decreases chances of divorce by around one-third.

\paragraph{More risks of divorce $\Rightarrow$ fewer kids.} Among couples, those who have kids together have on average higher relationship quality. Figure \ref{risks-fertility} shows the average match quality by the fertility status and age for college graduate females. This suggests that couples facing more risks of divorce would prefer delaying their fertility. This is consistent with Alesina, Giuliano, 2005 finding about the fertility decline with introduction of unilateral divorce.

In the context of kids-first and marriage-first division, the marriage-first couples who already had their children have on average larger marriage quality than the childless. This additionally contributes to the differences.

\begin{figure}
\begin{center}
\includegraphics[scale=0.7]{../pics/conception_divorce.pdf}  
\caption{How fertility changes divorce risks.\label{conception-divorce}}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\begin{tabular}{c}
\includegraphics[scale=0.5]{../pics/risks_fertility.pdf} 
\end{tabular}
\caption{How divorce risks affects fertility.\label{risks-fertility}}
\end{center}
\end{figure}

\section{Can Policies Promote Marriage?}
Through the lens of the model, marriage quality is a single most important determinant of marriage decisions. Any policy that aims to affect people marriage decision has the most immediate effect of changing marital decisions for some people who are on the margin of whether to marry each other.

The model allows to infer how large is the share of the marginal marriages. Some people marry because they had an unplanned pregnancy, and this essentially means that they are below the threshold for a regular marriage. The data does not allow to identify how many people are this: at best we can see how many of the kids-first couples divorce. The structural model, essentially, allows to convert the ``excess'' divorce rate to the share of people who are on the margin. %If the distribution of shock is identified from general marriage and divorce rates (as in Low et al, [...]), and the divorce rates of ``compliers'' and ``always-takers'' given by the model parameters, than the model recovers the share of the compl

In principle, the distribution of the shock and hence the agreement thresholds can be inferred from just aggregate divorce and marriage rates, as in Low et al, 2015. This paper, in some sense, gives additional ``overidentifying'' restrictions to this inference: the distribution should be such that the share of ``compliers'' is consistent with the differences in divorce rates. On the one hand, this provides additional validation to the previous results, on the other, it endoegenizes more things so it facilitates understanding of additional facts about how fertility affects the decisions.

The first experiment is directly pushing people who had an unplanned pregnancy into marriage. The model provides social stigma parameter as a metaphor for this. I show that pushing people into marriage creates some temporary effects --- people who are pushed may not divorce immediately and stay together for some time. However, these effects are really temporary --- it pushes some young people to short marriages, but they break up by adult ages.

Then, I consider strengthening the child support as popular and more realistic way to treat people who have kids and are on the margin of breaking up. Theoretically, more responsibility of fathers can make people marry more, as walking away is more costly. However, in practice this effect is negligible from aggregate prospective. What is non-negligible is effect on two margins: women do less abortions if they decide not to marry the potential partner, and more couple with kids break up as post-marriage prospectives are now better for women. This all pushes the count of single mothers, both general and among mothers, up.

Finally, I consider more ways of promoting marriage. Direct monetary incentives, like tax brackets for married couples, have pretty small effect on aggregate even if the benefits are large. This suggests that relative importance of relationship quality --- love --- is substantially higher than of money. A non-obvious but guaranteed way of promoting marriages in the model is making divorces easier. Indeed, it makes many people on the margin wanting to ``try'' marriage, and especially those who already have pressure of unplanned pregnancy, as argued by Alessina, Giuliano, 2005. However, on the aggregate level many implications of it are far from desirable: it implies large drop in fertility, which even offsets additional marriage creation.

\subsection{Is Social Stigma Efficient?}

Is it good for the society to push people who had kids to marry? Blind statistical view suggests that having two-parent family benefits at least the child, although more economic prospective suggests that this is not sustainable regulation. Believing in a simple collective environment, if parents caring about the child choose to be apart, they would be worse off staying together. However, the limited commitment setting and search frictions do not guarantee the first-best allocation, and the fact that the child's utility is not modeled properly makes the answers uncertain in the model.

The model includes $\phi_s$ utility parameter, that in some sense precisely does this. Is social stigma that forces people to marry in case of unplanned pregnancy a negative thing? As seen in Table \ref{causal-table}, large share of the compliers divorce. On the other hand, another large share of compliers stays together in couples. Relative to the situation with no pressure to marry, a large share of these couples would never be formed. So the direction is a priori uncertain.

More practically, what direction is the relation between social stigma and share of single mothers? On the one hand, more stigma decreases the number of people who does not marry following pregnancy, hence it should decrease the number of never married single mothers. On the other hand, more stigma causes creation of more risky marriages, therefore it should increase the number of divorced single mothers. In the world without abortions the second effect is smaller than the first if at least some women stay in their shotgun marriages. In presence of abortions this can go either way and the model is capable of making an assessment.

To get the answers, I change the size of social stigma parameter. See Table \ref{stigma-table} for the numbers providing two experiments: doubling the stigma or reducing it twice.

The table provides several dimensions of results. Among the obvious results, pushing more people into marriage in case of pregnancy sindeed makes more people married, though this effect is modest due to the limited scope of the pregnancies. Additionally, as already discussed, this lowers already compromised marriage performance of existing shotgun marriages, increasing the divorce rate for kids-first couples. Then, as more people are pushed towards marriage, less women abort unplanned pregnancies. Finally, as some people have to experience the social stigma, increasing it decreases overall welfare.

There are less obvious results generated by interaction of the channels. First, pushing more people into marriage increases count of single mothers, especially for the less educated population. Second, additional pressure to enter shotgun marriages hurt men relatively much more. Third, and perhaps the most important result, is about the child's well-being. One may think that if parents stay together at least temporary the overall well-being of the child is improved.  The experiment reveals that it is false: temporary marriages do lead to more long-term singleness. If women can avoid dealing with inefficient matches and start searching for a satisfactory partner quickly it brings both them and children large long-term return.

\begin{table}[h]
\begin{center}
\caption{Experiment: more pushing into shotgun marriage.\label{stigma-table}}
\begin{tabular}{l | r r r | r r r}\hline
& \multicolumn{3}{c|}{\textbf{High School}} & \multicolumn{3}{c}{\textbf{College}} \\\hline
\multicolumn{1}{r|}{\textbf{Parameter $\phi_s$:}}& \textbf{$\times 0.5$} & \textbf{Base} & \textbf{$\times 2$} & \textbf{$\times 0.5$} & \textbf{Base} & \textbf{$\times 2$} \\\hline
\multicolumn{7}{c}{\emph{\% of single mothers in population}} \\\hline
\textit{...at 35}  & 24.5 & 29.8 & 33.5 & 8.4 & 8.4 & 8.5\\\hline
\multicolumn{7}{c}{\emph{\% ever married}} \\\hline
\textit{...at 25}  & 37.2 & 39.6 & 45.2 & 26.0 & 26.6 & 27.6\\
\textit{...at 35}  & 65.8 & 68.8 & 75.2 & 77.9 & 78.9 & 81.2 \\\hline
\multicolumn{7}{c}{\emph{\% divorced if kids-first}} \\\hline
\textit{...10 years after}  & 16.9 & 21.4 & 34.4 & 17.3 & 20.6 & 28.0\\\hline
\multicolumn{7}{c}{\emph{\% unplanned pregnancies aborted}} \\\hline
\textit{...total}  & 66.3 & 37.4 & 14.6 & 42.7 & 39.9 & 30.3\\\hline
\multicolumn{7}{c}{\footnotesize \textbf{Welfare compared to Base, 1000s USD}}\\\hline
\textit{median male at 23}  & $304$ & --- & $-288$ & $77$ & --- & $-132$ \\
\textit{median female at 21} & $105$ & --- & $-97$ & $34$ & --- & $-63$\\\hline
\multicolumn{7}{c}{\footnotesize \textbf{Child consumption equivalent, 1000s USD}}\\\hline
\textit{ever born, median}  & 23.9 & 21.3 & 19.2 & 59.9 & 59.7 & 59.0\\
\textit{born at 30, median}  & 35.1 & 29.4 & 25.9 & 61.3 & 61.2 & 60.6\\\hline
\multicolumn{7}{p{0.7\linewidth}}{\footnotesize \emph{Notes: this experiment changes the magnitude of social stigma parameter $\phi_s$. Details on how welfare comparison and child consumption equivalent defined  are given in \ref{welfare-details}.}}\\\hline\hline
\end{tabular}
\end{center}
\end{table}

%[Doubling the stigma is mildly weflare-improving for the least productive women, who seem to generally benefit from the shotgun marriage.]

\subsection{Child Support and Shotgun Marriages}
The model provides natural ground to study child support. To recall, in the baseline calibration the child support is assigned with probability $p^a_{nm} = 0.284$ in case of a couple with an unplanned pregnancy who disagrees to marry and keeps the child and $p^{a}_{div} = 0.461$ in case of a married couple with a child divorcing, and this corresponds to the average compliance rate in the US at 2017, from .... The child support itself is set to be 20\% of male's labor earnings and is modeled as a shift in permanent productivity $z$. More details are given in \ref{divorcevalues}.

The two experiments are eliminating the child support system at all (setting $p^{a} = 0$, experiment ``No CS'') and making the enforcement perfect (setting $p^{a} = 1$, ``Full CS''). Table \ref{child-support-table} presents the comparison.

\begin{table}
\begin{center}
\caption{Impact of child support \label{child-support-table}}
\begin{tabular}{l r r r | r r r}\hline
& \multicolumn{3}{c|}{\textbf{College}} & \multicolumn{3}{c}{\textbf{High School}}\\\hline
\multicolumn{1}{r}{\textbf{Child support regime:}} &\textbf{No} & \textbf{Base}& \textbf{Full}  &\textbf{No} & \textbf{Base}& \textbf{Full}\\\hline
\textit{divorced in 10 years if KF} & 21.3 & \textbf{21.1} & 21.3 & 20.8 & \textbf{21.7} & 22.1 \\
\textit{divorced in 10 years if MF} & 5.9 & \textbf{6.2} & 6.6 & 14.8 & \textbf{15.9} & 17.0 \\
\textit{KF women in population at 30} & 3.7 & \textbf{3.7} & 3.8 & 12.0 & \textbf{12.2} & 12.4 \\
\textit{MF women in population at 30} & 27.7 & \textbf{28.5} & 29.7 & 20.7 & \textbf{21.2} & 22.4 \\
\textit{single mothers among mothers at 35} & 13.1 & \textbf{13.2} & 14.5 & 41.8 & \textbf{42.8} & 44.0 \\
\textit{unplanned pregnancies aborted} & 40.4 & \textbf{39.9} & 33.6 & 40.9 & \textbf{37.3} & 32.7 \\\hline
\multicolumn{7}{c}{\textbf{Welfare compared to Base (1000s of 2017 USD in savings)}} \\\hline
\textit{median man at 23} & $-2.6$ &--- & $-0.1$ & $10.6$ & --- & $-16.3$\\
\textit{median woman at 21} & $-6.3$ & --- & $11.1$ & $-10.0$ & --- & $20.3$ \\\hline
\multicolumn{7}{c}{\textbf{Child Consumption Equivalent, 1000s USD}} \\\hline
\textit{child ever born, median} & $59.7$ & $59.7$ & $59.1$ & $21.9$ & $21.3$ & $20.4$\\
\textit{child born at 30, median} & $61.3$ & $61.2$ & $60.7$ & $31.2$ & $29.4$ & $27.3$\\\hline
\hline
\multicolumn{7}{p{0.9\linewidth}}{\footnotesize \textit{Notes:} \textbf{No} refers to the scenario with absent child support, \textbf{Base} is the one with limited enforcement used in the estimated model, \textbf{Full} is fully enforced child support. Details on how welfare comparison and child consumption equivalent defined  are given in \ref{welfare-details}.}\\\hline
\hline
\end{tabular}
\end{center}
\end{table}

There are several crucial conclusions from the table. First, the child support has relatively modest effect on creation and dissolution of the kids-first marriages, and, if anything, incentivizes people to divorce. Second, the largest margin the child support changes is fertility: more married couples have kids and more women choose to not to abort their unplanned pregnancies. These factors together facilitate induce the economy to have more single mothers when child support gets stronger.

On the welfare side, more child support is generally welfare improving for parents: men lose less than woman gain. But, more surprising, the removal of the child support can hurt men: less responsibility in (unlikely) case of divorce can are offset with fertility decline. This cannot be conclusive, but suggests that there is some optimal level of enforcement that can be different from zero or one.

Finally, child support in the model makes children worse off. This happens almost precisely because the long-run count of single mothers is larger, and the effects of more resources is offset by effect of more singleness. Of course, this measure is crude as in the model the child's well-being is not fully internalized by both parents, nevertheless, the general dominance of non-monetary factors in marriage/divorce decisions produces this result.\footnote{Similar welfare implications were discussed in Brown, Flinn, 2012 and Forester, 2020, but their channels are somewhat different.}

\subsection{Reducing The Pressure Promotes Efficient Marriages}

Two previous results are, to a large extent, negative. This section exploits two things that deliver positive results in more details: removing the social stigma and removing the costs of abortions. Table \ref{promoting-table} presents the results.

Both factors move the economy in desirable direction, unambiguously increasing utilities of both parents and children. Their relative importance vary: for the college graduates, fixing imperfections of fertility control provides relatively larger improvement, for high school graduates, fertility control impact is pretty small and social stigma plays the largest role. Additionally, it fixes the excess divorce rates of kids-first marriages, making two groups more similar (especially for the college graduates).

The main implication of these differences is that the most sensitive margin among the college graduates are women entering the risky marriages because of undesirable consequences of single motherhood, on the other hand. The social stigma pressure is more substantial for the high school graduates, as being single mothers is generally more favorable for them. The idea of single motherhood as a rational choice is similar to Neal, 2004, Chiappori, Oreffice, 2008 and many others.

Note that, on the other hand, option with full fertility control and no social stigma does not make two groups of kids-first and marriage-first couples completely equal: mistimed fertility is still a factor. The results described above suggest that couples with more risky marriages have larger probability to become childless, therefore the marriage-first group --- those chosen to have kids together --- are by construction more risky that the group of people with generally good marriages who received the fertility shock randomly. The selectivity effect implied by the model is pretty large.

\begin{table}[h]
\begin{center}
\caption{Fertility control and removing the stigma perform the best.\label{promoting-table}}
\begin{tabular}{l | r r r r | r r  rr}\hline
& \multicolumn{4}{c|}{\textbf{High School}} & \multicolumn{4}{c}{\textbf{College}} \\\hline
\multicolumn{1}{r|}{\textbf{Experiment}}& \textbf{Base} & \textbf{FC} & \textbf{NS}  & \textbf{Both} & \textbf{BL} & \textbf{FC} & \textbf{NS}  &  \textbf{Both} \\\hline
\multicolumn{9}{c}{\emph{\% of single mothers in population}} \\\hline
\textit{...at 35}  & 29.8 & 29.7 & 21.8 & 21.7 & 8.4 & 4.1 & 8.5 & 3.6\\\hline
\multicolumn{9}{c}{\emph{\% ever married}} \\\hline
\textit{...at 25}  & 39.6 & 39.6 & 35.6 & 35.6 & 26.6 & 25.6 & 25.5 & 24.1 \\
\textit{...at 35}  & 68.8 & 68.8 & 58.3 & 58.3 & 78.9 & 78.8 & 76.3 & 75.9 \\\hline
\multicolumn{9}{c}{\emph{\% divorced 10 years after marriage if...}} \\\hline
\textit{...kids-first}  & 21.4 & 21.4 & 13.3 & 13.3 & 20.6 & 17.2 & 15.0 & 11.5\\
\textit{...marriage-first}  & 16.0 & 16.0 & 2.3 & 2.2 & 6.2 & 6.0 & 5.8 & 5.5\\\hline
\multicolumn{9}{c}{\emph{\% unplanned pregnancies aborted}} \\\hline
\textit{...total}  & 37.4 & 37.7 & 87.5 & 87.7 & 39.9 & 77.0 & 48.0 & 87.1\\\hline
\multicolumn{9}{c}{\footnotesize \textbf{Welfare compared to Base, 1000s USD}}\\\hline
\textit{median male at 23}  & --- & $1.2$ & $500$ & $500$ & --- & $29$ & $176$ & $221$ \\
\textit{median female at 21} & --- & $0.4$ & $435$ & $443$ & --- & $131$ & $92$ & $341$\\\hline
\multicolumn{9}{c}{\footnotesize \textbf{Child consumption equivalent, 1000s USD}}\\\hline
\textit{ever born, median}  & 21.3 & 21.3 & 25.4 & 25.4 & 59.7 & 62.1 & 60.2 & 63.1\\
\textit{born at 30, median}  & 29.4 & 29.4 & 45.8 & 46.1 & 61.2 & 63.4 & 61.6 & 63.5\\\hline
\multicolumn{9}{p{0.85\linewidth}}{\footnotesize \emph{Notes:} \textbf{FC} shows an experiment when women can choose to terminate any unwanted pregnancy at no costs. \textbf{NS} removes the social stigma setting $\phi_s = 0$. \textbf{Both} does both experiments together. Welfare details are given in Appendix \ref{welfare-details}, 500 is an upper bound for welfare changes.}\\\hline\hline
\end{tabular}
\end{center}
\end{table}

\subsection{Liberalization of Divorce [TBA]}

In this experiment I consider counterfactual about changing the divorce costs and its impact on marital selection, shotgun marriages and welfare.

To some extent, the reduction of divorce costs replicates the scenario of introduction of unilateral divorce, studied in great details in context of similar models by Voena, 2015, Reynoso, 2019\nocite{reynoso2018impact}, Blasutto, Kozlov, 2020 (forthcoming), however, from the decision making prospective just reducing the costs is very different to change in how decisions are made. %[I have codes for introduction of unilateral divorce as well, but it will take some time to incorporate.]

%Panels of Figure \ref{risks-fertility} show the results of solving the model with $\phi_d' = 0.5\phi_d$.

\subsection{Can Money Help? [TBA]}
The model mechanics does not propose an obvious way to make people marry more, and makes in pretty undesirable generally: the marriages can be incentivized by something that shifts the agreement thersholds, but those who comply with the shift in the threshold and decide to marry have the worst level of marriage quality among the population and the highest risk of divorce.

Nevertheless, some policies can also affect the general risk of divorce, and hence, add incentives to stay married. A natural candidate for this is different monetary incentives. Different welfare like TANF-AFDC programs are targeted to single mothers and are aimed to make their life better financially. In the context of the model this should generally decrease number of marriages.

One can imagine the opposite direction: taxing singles to fund marriage couples. In particular, reducing tax progressivity and various tax cuts affect couples, who are wealthier than singles, more. This section considers general ways of redistribution using money, not limited to proportional taxation. One can imagine that all agents have some non-linear tax function $T^{i}(\text{earnings})$, and this function is type-specific. Even lump-sum child subsidies can be fit to this framework.

\section{Conclusions}
%This paper aims to identify the causes of differences in divorce rates for couples with different fertility timing and 
Marriage is generally a good thing, but making more people married does not do any good. In a sense, allowing people to delay their marriages and be more selective is the uniformly right direction of policies. The model allows touching the implications of the marriage promotion quantitatively. It uses the divorce and marriage rate information to infer what proportion of marriages are on the marginal in the sense of being the most responsive to the environment changes.

An essential qualitative insight is that judging by the differences in divorce rates, around half of the kids-first marriages result from the response to an unplanned pregnancy shock, meaning that many of these couples would not marry if the shock did not happen. Many of these couples divorce in the future. This share of compliers is slightly larger for high school graduates and somewhat smaller for college, but the pattern is not so drastic. A large chunk (that could exceed half) of these couples divorces in the future, and this is what drives the excess divorce rate for the kids-first.

% What if we add learning? Men single women that they have serious interntions by agreeing to marry....
What makes people enter marriages of lower quality? Generally speaking, this is generated by the model's reaction to unplanned pregnancies. When a couple that thinks about whether to marry each other realize they have an unexpected child, their disagreement options change, and it makes it hard for the parents to say no to each other.

The model confirms and quantifies what forces contribute to the change in outside options generating different outcomes of the kids-first marriages. The crucial things among them are social pressure, or stigma, to marry people you have children with, the risk of women raising the child on their own if they do not enter the marriage, the remarriage complications of women with children. Another force for the difference that is not directly related to the outside options is the marriage-first group's selectivity: risky couples choose not to have kids or have them later.

This paper attempts to understand the general direction different effects go and the relative proportion between them, rather than analyze some concrete policies. The whole idea of regulating the marriage market is dubious; however, many policies that are conditional on marital and fertility status may create effects similar to the changes I discuss in the paper.

Still, the model can provide non-trivial policy insights. First, trying to promote marriages with policies is costly in terms of welfare and pretty inefficient. In some sense, the estimations identify that a potential relationship's match quality is substantially more important than any other determinants of marriage decisions. This does not mean that the partner selection is based solely on one unobservable shock, but it does imply that monetary incentives are relatively week.

The largest return in terms of welfare for parents and children is generated by shaping responses to unplanned pregnancies: helping people whose match is far from being efficient to break up and terminate the pregnancy. This even has empirical content: in places with more liberal abortion and birth control access one should expect kids-first couples to be more stable. Checking this is possible with some detailed data, but is a little outside of the current project's scope. Note that ability to terminate pregnancies alone could be not enough: through the lens of the model, even conditional on a choice to keep the pregnancy entering an inefficient match hurts women in the long term.

Finally, the model provides a link between divorce policies and fertility beyond the shotgun marriage angle. Easier divorce is identified to lower fertility strongly, even though it stimulates more marriages. Conditional on already have children, easier divorce is beneficial for a median parent though it slightly hurt the median child: the child's utility drops. [This part TBD]

\newpage
\bibliography{bib}

\begin{figure}
\caption{Model fit: main targets, college graduates\label{fit-col-1}}
\begin{center}
\begin{tabular}{c c}
\includegraphics[scale=0.6]{../pics/col/div_kfmf.pdf} & \includegraphics[scale=0.6]{../pics/col/popshares.pdf} \\
 \includegraphics[scale=0.6]{../pics/col/relshares.pdf} & \includegraphics[scale=0.6]{../pics/col/kids_by_duration.pdf}
 \\\multicolumn{2}{l}{\textit{Notes}: all points are targeted on the first three graphs.}
\end{tabular}
\end{center}
\end{figure}

\begin{figure}
\caption{Model fit: quantities of singles, college graduates\label{fit-col-2}}
\begin{center}
\begin{tabular}{c c}
\includegraphics[scale=0.6]{../pics/col/fit_single_nonmothers_nm.pdf} & \includegraphics[scale=0.6]{../pics/col/fit_single_nonmothers_div.pdf} \\
 \includegraphics[scale=0.6]{../pics/col/fit_single_mothers_nm.pdf} & \includegraphics[scale=0.6]{../pics/col/fit_single_mothers_div.pdf}\\
\end{tabular}
\end{center}
\end{figure}

\begin{figure}
\caption{Model fit: main targets, high school graduates\label{fit-hs-1}}
\begin{center}
\begin{tabular}{c c}
\includegraphics[scale=0.6]{../pics/hs/div_kfmf.pdf} & \includegraphics[scale=0.6]{../pics/hs/popshares.pdf} \\
 \includegraphics[scale=0.6]{../pics/hs/relshares.pdf} & \includegraphics[scale=0.6]{../pics/hs/kids_by_duration.pdf}
 \\\multicolumn{2}{p{\linewidth}}{\textit{Notes}: all points are targeted on the first three graphs. Nothing is targeted on the last graph, as the preference parameters are fixed at the college graduates' levels.}
\end{tabular}
\end{center}
\end{figure}

\begin{figure}
\caption{Model fit: quantities of singles, high school graduates\label{fit-hs-2}}
\begin{center}
\begin{tabular}{c c}
\includegraphics[scale=0.6]{../pics/hs/fit_single_nonmothers_nm.pdf} & \includegraphics[scale=0.6]{../pics/hs/fit_single_nonmothers_div.pdf} \\
 \includegraphics[scale=0.6]{../pics/hs/fit_single_mothers_nm.pdf} & \includegraphics[scale=0.6]{../pics/hs/fit_single_mothers_div.pdf}\\
\end{tabular}
\end{center}
\end{figure}

%
%\begin{figure}
%\caption{Doubling the social stigma, college\label{double-stigma-col}}
%\begin{center}
%\begin{tabular}{c c}
%\includegraphics[scale=0.6]{../pics/double-ss-col/sm_among_mothers.pdf} & \includegraphics[scale=0.6]{../pics/double-ss-col/relshares.pdf} \\
% \includegraphics[scale=0.6]{../pics/double-ss-col/fit_single_mothers_nm.pdf} & \includegraphics[scale=0.6]{../pics/double-ss-col/fit_single_mothers_div.pdf}
%\end{tabular}
%\end{center}
%\end{figure}
%
%\begin{figure}
%\caption{Doubling the social stigma, high school\label{double-stigma-hs}}
%\begin{center}
%\begin{tabular}{c c}
%\includegraphics[scale=0.6]{../pics/double-ss-hs/sm_among_mothers.pdf} & \includegraphics[scale=0.6]{../pics/double-ss-hs/relshares.pdf} \\
% \includegraphics[scale=0.6]{../pics/double-ss-hs/fit_single_mothers_nm.pdf} & \includegraphics[scale=0.6]{../pics/double-ss-hs/fit_single_mothers_div.pdf}
%\end{tabular}
%\end{center}
%\end{figure}

\newpage
\appendix

\section{Additional ACS Evidence\label{app-additional-evidence}}
\subsection{Flow-Based Divorce Rates}
Demographic statistics (see ... for an example) typically defines divorce rates as number of divorces by $1{,}000$ of women, either all or ever married, that happen within a particular year. This measure is different from mine: I focus on a cumulative measure of share divorced, and the divorce rate is a flow measure of the new divorced. Its advantage though that is is more robust to composition.

Table ... shows the divorce rates: number of divorces in KF group (divorces per 1{,}000 KF women), in MF group (divorces per 1{,}000 MF women), as well as KF-divorces and MF-divorces per 1{,}000 women in general population. Switching the measure does not provide particularly different insights.

\subsection{Changing the $\Delta T$ window\label{changing-window}}
Here I show the results excluding restriction $-5\leq \Delta T \leq 5$. Many people with $\Delta T > 5$ are divorced, which perhaps indicate that they might have divorced before having the child, and this can bias the judgement of the marital stability, however qualitatively the pattern still holds. See ....

\subsection{Step or Own Children in ACS}
For existing marriages ACS sometimes allows to capture the relation of children to the father. The requirement for this is that the male is the household head, and the exact criterion for it is not clear, however in two-person households ...\% have male heads. Focusing on them, it is easy to recover the percentage of stepchildren in surviving marriages by $\Delta T$. As we can see from the left panel of Table ..., as $\Delta T$ approaching zero, the share of households with stepchildren decreases, and it never exceeds ...\%.

Focusing on surviving marriages is restrictive, as it is possible that marriages with step children live shorter and drive the difference in divorce rate. If this is the case, then one would expect that the observable duration of a surviving marriage with step children will be shorter. Therefore the right panel of Table ... shows the average duration of an observable marriage. It suggests the differences in duration are almost absent, and therefore it is not likely that the marriages with step children have drastically different divorce patterns, which is another validation of the results from SIPP.

\section{Additional Model Details}
\subsection{Single Males\label{single-male-appendix}}

The decision problem of single males is similar to single females, except for the details discussed in \ref{singles-main}. It has to account for possibility to meet single mothers and corresponding losses upon marriage to them due to stepchildren.

Let $\Gamma^M$ denotes potential woman matches without kids, and $\Gamma^{Mk}$ denote marriageable single mothers. Also, let $p^{\text{sk}}_t$ denote the share of single mothers in population. To ease the notation, let $p^{\text{meet,np}} = (1-p^{\text{sk}}_t)\cdot p^{\text{meet}} \cdot (1-p^{\text{preg}})$, denote probability to meet a childless woman and not to have an unplanned pregnancy, similarly $p^{\text{meet,p}} = (1-p^{\text{sk}}_t)\cdot p^{\text{meet}} \cdot p^{\text{preg}}$ if for an unplanned pregnancy and $p^{\text{meet,sk}} = p^{\text{sk}}_t\cdot p^{\text{meet}}$ is for meeting a single mother.

The problem for single males is similar:
\begin{align}\label{singlemaleproblem} & V_t^{m,s}(a,z^m)  = \max\limits_{a',c^m} \Bigg\{ \ \ u(c^m) +  \\\nonumber &  \beta\cdot \E_{z^{m\prime}|z^m} \Bigg( (1-p^{\text{meet}}_t)\cdot V^{m,s}(a',z^{m\prime}) \,+ & \text{(no partner met)}\\\nonumber
 & p^{\text{meet,np}}_t \cdot \int \Big[m^{M,np}\cdot V^{m,c}_{t+1}(a'+a^M,\omega^M,\theta^{M,np}) + & \text{(met, no pregnancy, agree)}\\\nonumber
 &\hspace{4cm}(1-m^{M,np})\cdot V^{m,s}_{t+1}(a',z^{m\prime}) \Big]d\Gamma^M + & \text{(disagree)} \\\nonumber
 & p^{\text{meet,p}}_t \cdot \int \Big[m^{M,p}\cdot V^{m,ck}_{t+1}(a'+a^M,\omega^M,\theta^{M,p})+  & \text{(met, shotgun marriage)}\\\nonumber
 &\hspace{2cm} (1-m^{M,p})\cdot \left\{\E_{csm}V^{m,s}_{t+1}(a',z^{m\prime}) - \phi_s\right\} \Big]d\Gamma^M + & \text{(disagree, social stigma)}\\\nonumber
 & p^{\text{meet,sm}}_t \cdot \int \Big[m^{M,sm}\cdot \left\{ V^{m,ck}_{t+1}(a'+a^M,\omega^M,\theta^{M,sm}) - \phi_r\right\}+  & \text{(met a single mother)}\\\nonumber
 &\hspace{4cm} (1-m^{M,sm})\cdot  V^{m,s}_{t+1}(a',z^{m\prime}) \Big]d\Gamma^{Mk}\Bigg) \Bigg\} & \text{(disagree)}\\\nonumber
 \text{s.t. \ } & a' + c^m = R\cdot a  + W^m_t(z^m) & \text{(evolution of the assets)}.\\\nonumber
\end{align}

\section{Solution Technique\label{solution-appendix}}

As the model has a finite lifespan and no interaction of people of different ages, apart from a fixed difference in ages (which is just equivalent to renumbering the periods), it can be solved using a standard backward induction, starting from zero value at the last period of the life. Large dimensionality of state complicates the analysis, for each age and agents' type the value function is a 5-dimensional object, where the dimensions are assets, female and male productivity, match quality and (one-dimensional) bargaining power. Productivities and match quality, however, are exogenous objects and are referred to as $\omega$.

The solution assumes discrete grids for assets $a$ (60 points), female productivity $z^f$ (7 points), male productivity $z^m$ (5 points), couple's match quality $\psi$ (15 points) and female bargaining weight $\theta$ (15 points). These grid points represent the nodes at which the value functions are stored, the decisions regarding savings and bargaining power are allowed to be off-grid, and for them I introduce finer grids for savings $s$ (500 points) and decision weights $\theta$ (150 points), where the value functions at these point are computed by the linear interpolation of the coarser grid values.

To discretize the standard random walk process for $z$ and $\psi$ I used a non-stationary [Rouwenhurst ] method. [Fella, ...] show that it has the best performance for the lifecycle model applications. An important feature of the method is the grid that depends on time, as the standard deviation of the random walk variable increases over time. The case of random walk with drift for the skills depreciation when females are out of labor force requires being handled separately, I discuss this in \ref{tauchen-z}.

More details of the solution technique are given throughout the appendix \ref{solution-appendix}. Shortly, as all the variables are discretized all transitions can be represented with Markov process transition matrices, that can be obtained using classic ([Tauchen ...]) approximation techniques.

\subsection{Tauchen Method}
This is what I refer to as Tauchen method. For a random variable $x$ and the $n$-node grid $[y_1,...,y_n]$ probability to get to the node $k$ is a probability that $x$ is closer to $y_k$ than to anywhere else, assuming the grid is ordered, this becomes $p_k = \P\left(\frac{y_{k-1} + y_k}{2} \leq x \leq \frac{y_{k+1} + y_k}{2}\right)$ for $k \in \{1,...,n\}$, assuming $y_0 = -\infty$ and $y_{n+1} = +\infty$.

This idea is not specific to the normal distribution and can be used with an arbitrary CDF.

\subsection{Approximations for Skills Depreciation}

In case of being out of the labor force, the equation \ref{zoutlf} represents a random walk with drift. [Rouwenhurst ] approximation is not tractable for this case. If done, it will also require having different grid nodes for those in and out labor force.

Therefore, these transition matrices are derived separately using a somewhat synthetic approach: I keep the same [Rouwenhurst ] nodes as for those in the labor force and combine them with Tauchen approach for obtaining the transition probabilities. I did not find the result to be particularly sensitive to the choice approximation method, therefore I used the best method whenever I could and the tractable one whenever the approach is not obvious.

\subsection{Partners' Distribution}
Take females, for example. Given the grid for $z^m = \{z^m_1,...,z^m_{n_z}\}$ in this period, I apply Tauchen method to (non-parametrically estimated) distribution of the single male's productivity at this age $F^{z,m}_t(z)$, and get probabilities to draw a partner with each level $\{p_1,...,p_{n_{z,f}} \}$. The assets distribution from equation \ref{assets-partner} can be represented as \[a = \exp\left\{ \mu^a_t(z) + \varepsilon^a_t\cdot \sigma^{a}_t(z)  \right\}\I(\log a^* \geq \bar{a}),\ \  \varepsilon^a_t \sim \mathcal{N}(0,1).\] I apporximate no $\varepsilon^a_t$ with $q$-point distribution.\footnote{I use slightly arbitrary five-point Tauchen approximation of standard normal with the nodes$\{0,\pm 0.75,\pm 1.5\}$. I find these to be more pretty satisfactory as setting more extreme bounds produces too many partners with zero assets.}  For each $z^m_i$ this implies $q$ assets levels $\{a_{i1},...,a_{iq}\}$ and corresponding probabilities  $\{p_{i1},...,p_{iq}\}$. By combining the two discrete distributions, and adding the distribution for initial match quality $\psi$, which is again a discrete-grid approximation of $\mathcal{N}(\mu_{\psi,0},\sigma^2_{\psi,0})$, we get a distribution of potential matches in terms of $a$, $z$ and $\psi$. Essentially, this can be interpreted as singles drawing potential couple's state $\omega = (z^f,z^m,\psi)$ from some discrete distribution and additionally drawing a one of the random $q$ assets positions, so the single agent has $n_{z^f}\times n_{z^m} \times n_{\psi} \times q$ potential discrete transitions.

\section{Details on Welfare Comparisons\label{welfare-details}}
\subsection{Assets Variation}
The welfare comparison is done in a manner familiar to classic microeconomics, but everything is converted to amounts of money in savings for a single person with no savings in the beginning of their life.

Essentially, for changes which make the person better off, the measure I use asks having how much money on savings account in the beginning of the life is equivalent to having the change. This resembles equivalent variation measure. Namely, if the change is preferred, the assets variation is:
\[AV = a^*_t(z) \text{ such that } V_{t}^{f,s,\text{no change}}(a=a^*,z) = V_{t}^{f,s,\text{with change}}(a=0,z).\]
Otherwise, if the change is not preferred, we cannot deduct the money. Therefore, for the negative changes I define the measure similar to compensating variation:
\[AV = -a^*_t(z) \text{ such that } V_{t}^{f,s,\text{with change}}(a=a^*,z) = V_{t}^{f,s,\text{no change}}(a=0,z),\]
meaning how much money on savings account would compensate the person for the change.

Defining these measures for people in couples is complicated, as the value functions depend on the endogenous bargaining power. For the simpler exposure I omit these comparisons.

\subsection{Child's Consumption Equivalent}
A simple measure for well-being of the child is the average child's consumption $Q$ for the first 18 years of the child's life. At the moment the child is born, it is constructed as
\[\bar{Q}_b =\left[ \frac{1 - \beta}{1 - \beta^{18}} \cdot \sum\limits_{s=0}^{17} \beta^s Q_{b+s}^{1-\xi}\right]^{\frac1{1-\xi}}.\]
Since agents can have up to one child, each childbearing woman has the unique value of this measure.

\end{document}

### 2020-10-02 jmp-draft/new.tex
Current stage

\documentclass[12pt,letter]{article}
\usepackage[left=1.5in,right=1.5in,top=1.5in,bottom=1.5in]{geometry}
\usepackage{amsmath}
\usepackage{pdflscape}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{multicol}
\usepackage{microtype}
\usepackage{euscript}
\usepackage{epsfig}
\usepackage{epstopdf}
\usepackage{mathrsfs}
\usepackage{tikz}
\newcommand{\hypo}{\mathcal{H}}            
\bibliographystyle{ieeetr}

\usepackage[flushleft]{threeparttable}

%\usepackage[cp1251]{inputenc}
\usepackage[english]{babel} 
\DeclareMathOperator{\rank}{rank}
\newcommand*{\hm}[1]{#1\nobreak\discretionary{}
            {\hbox{$\mathsurround=0pt #1$}}{}}

            \def\onepc{$^{\ast\ast}$} \def\fivepc{$^{\ast}$}
\def\tenpc{$^{\dag}$}
\def\legend{\multicolumn{4}{l}{\footnotesize{Significance levels
:\hspace{1em} $\dag$ : 10\% \hspace{1em}
$\ast$ : 5\% \hspace{1em} $\ast\ast$ : 1\% \normalsize}}}

\newcommand{\bs}[1]{\boldsymbol{#1}}  
\newcommand{\bsA}{\boldsymbol{A}}

%\setstretch{1}                         
\flushbottom                            
\righthyphenmin=2                      
\pagestyle{plain}                       
%\settimeformat{hhmmsstime}  
\widowpenalty=300                   
\clubpenalty=3000                     
\setlength{\parindent}{0em}           
\setlength{\topsep}{0pt}              
\usepackage[pdftex,unicode,colorlinks=true,urlcolor=blue]{hyperref}
\usepackage{bbm}
\renewcommand{\emptyset}{\varnothing}

\setlength{\parskip}{0.5\baselineskip plus2pt minus2pt}

\newcommand{\e}{\varepsilon}
\DeclareMathOperator*{\Argmax}{\mathrm{Argmax}}
\DeclareMathOperator*{\Argmin}{\mathrm{Argmin}}
\DeclareMathOperator*{\argmax}{\mathrm{arg\,max}}
\DeclareMathOperator*{\argmin}{\mathrm{argmin}}

\newcommand{\blp}{\mathrm{BLP}}
\DeclareMathOperator*{\plim}{\mathrm{plim}}
\DeclareMathOperator{\Max}{\mathrm{Max}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\Y}{\mathcal{Y}}
\newcommand{\Z}{\mathcal{Z}}
\renewcommand{\geqslant}{\geq}
\renewcommand{\leqslant}{\leq}
\newcommand{\p}{\bs p}
\newcommand{\y}{\bs y}
\def\dd#1#2{\frac{\partial#1}{\partial#2}}

\renewcommand{\emptyset}{\varnothing}

\DeclareMathOperator{\tr}{\mathrm{tr}}

\newcommand{\bb}{\bs \beta}
\newcommand{\X}{\bs X}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\PP}{\mathbb{P}}
\DeclareMathOperator{\V}{\mathbb{V}}
\DeclareMathOperator{\CM}{\mathbb{C}}
\renewcommand{\C}{\CM}
\DeclareMathOperator{\var}{\mathrm{var}}
\DeclareMathOperator{\cov}{\mathrm{cov}}
\DeclareMathOperator{\corr}{\mathrm{corr}}
\DeclareMathOperator{\MSE}{\mathrm{MSE}}
\DeclareMathOperator{\Bias}{\mathrm{Bias}}
\renewcommand{\P}{\PP}
\newcommand{\dsim}{\stackrel{d}{\sim}}
\newcommand{\hn}{\mathcal{H}_0}
\newcommand{\ha}{\mathcal{H}_a}
\newcommand{\thetab}{\bs \theta}
\newcommand{\pv}{\text{P-value}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\MLE}{\scriptscriptstyle MLE}
\newcommand{\LR}{\mathrm{LR}}
\newcommand{\I}{\mathbb{I}}
\newcommand{\sumin}{\sum\limits_{i=1}^n}
\newcommand{\sumti}{\sum\limits_{t=0}^\infty}
\newcommand{\hbeta}{\hat{\beta}}
\newcommand{\halpha}{\hat{\alpha}}
\newcommand{\hsigma}{\hat{\sigma}}
\newcommand{\hvar}{\widehat{\var}}
\newcommand{\hcov}{\widehat{\cov}}
\newcommand{\Q}{\mathbb{Q}}

\newcommand{\pconv}{\xrightarrow{ \ p \ }}
\newcommand{\dconv}{\xrightarrow{ \ d \ }}
\newcommand{\asconv}{\xrightarrow{ \ a.s. \ }}
\newcommand{\msconv}{\xrightarrow{ \ m.s. \ }}

\newcommand{\pic}[4][h!]{\begin{figure}[#1]

\begin{center}\includegraphics[width=#2cm]{#3}\caption{#4\label{#3}}\end{center}
\end{figure}}

%outtex
\def\onepc{$^{\ast\ast}$} \def\fivepc{$^{\ast}$}
\def\tenpc{$^{\dag}$}
\def\legend{\multicolumn{4}{l}{\footnotesize{Significance levels
:\hspace{1em} $\dag$ : 10\% \hspace{1em}
$\ast$ : 5\% \hspace{1em} $\ast\ast$ : 1\% \normalsize}}}
%end outtex

%\bibliographystyle{ieeetr}

\newcommand{\laseq}{\stackrel{\lambda\text{-a.e.}}{=}}
\renewcommand{\d}{\underline}
\renewcommand{\u}{\overline}
\newcommand{\td}{\underline{\theta}}
\newcommand{\tu}{\overline{\theta}}
%\renewcommand{\theenumi}{\alph{enumi}}
%\renewcommand{\labelenumi}{(\theenumi)}
%\renewcommand{\theenumii}{\roman{enumii}}
%\renewcommand{\labelenumii}{\theenumii.}
%\renewcommand{\theenumiii}{\arabic{enumiii}}
%\renewcommand{\labelenumiii}{\theenumiii.}
%\renewcommand{\epsilon}{\varepsilon}
\newcommand{\hneq}{\stackrel{\hn}{=}}
\newcommand{\deq}{\stackrel{d}{=}}
%\title{416-2 Final Project\\
%Reproductive Technologies, Aging and Fertility Choice (Proposal)}
%\author{Egor Kozlov}
\begin{document}

\section{Introduction version 2}
The family institutions faced major changes over the last four decades. In most developed countries, marriage rates have fallen, divorce became more widespread, unmarried cohabitation became more popular and single parenting is on the rise. In the US, around half of the kids are currently born by unmarried mothers, nevertheless, most of the couples having kids together are married ---  unmarried cohabitation with children is still not a common practice\footnote{According to Payne, 2013, only about 3\% of children live in unions with unmarried cohabitating parents, as opposed to 21\% living with single mothers.}, unlike in many places over the world. Both cultural norms and economic environment are expected to contribute to this difference.

In particular, many couples marry shortly after they have a baby.  Historically this practice is referred to as a shotgun marriage. This does not solely refer back to more conservative times: according to Gibson-Davis et al (2016), around 10\% of women at 2012 in North Carolina are pregnant at the moment of their marriage, and as I show later, a comparable share of couples marry already having their biological kids. In total, roughly one-quarter of women of age 21--40 in the US today had their first birth before or at the year of their first marriage.

One may expect that as fertility is relatively easy to control now, the life of the couples who faced a shotgun marriage would not be very different from those who married prior to their fertility decisions. In this paper I am going to argue that this is not the case: I show that marriages that happen after the pregnancies are substantially less stable in terms of divorce rates --- less than one-quarter of the population account for more than one-third of all divorces; and it is not likely to be explained by available observables. This points out that these marriages are different in unobservable relationship quality, and the fertility timing relative to the moment of marriage plays an important role in people's decisions to marry.

The main focus of the paper is understanding the sources of the differences in marital stability between those who have kids before (``kids-first'' couples) and after (``marriage-first'' couples) their marriage and the implications of these differences. From a positive economic perspective, this facilitates understanding the value of marriage and trends in it. For instance, the work of Chetty and Hendren (2018) suggests that high shares of single mothers in neighborhoods are a very strong predictor of low intergenerational mobility and reproduction of inequality. Numerous studies have documented how being born out-of-wedlock, having single or divorced parents or a blended family is associated with worsened human capital accumulation and poor early and adult-life outcomes for the kids (see Kearney, Levine 2017 or Ginther, Pollak, 2004).

Supporting a traditional two-parents married family is often considered as an anti-poverty measure, however not all marriages are created equal, and treating all marriages equally may understate inequality: growing up in an unstable marriage can be a predictor of ending up with a single mother or in a different household, as well as a factor of risk by its own. The families who experienced shotgun marriages are often overlooked by the current literature focusing on child development.

In addition to the child development angle, being divorced, especially with childcare burden, is considered to have a negative impact on women in terms of labor and marriage market outcomes and future fertility. [add references?]

From a normative perspective, it is essential for the right design of the family-related regulations: child support and visitation rights, custody allocation, alimony payments, welfare programs targeting single mothers like (like AFDC --- TANF in the US), divorce laws and childcare policies like a paid parental leave in general. I focus in particular on child support: many empirical studies, including recent Tannenbaum (2019) and Rossin-Slater (2017) document the impact of family regulations on the formation of new marriages. To my knowledge, the effects of these regulations on divorce rates of existing couples are not documented well. However being a single parent after a short-lived marriage or being a single parent while never being married are similar experiences. Moreover, as child support is easier to enforce for those couples who were married, its impact through existing marriages is potentially bigger. However, purely empirical understanding of impacts of the regulation on divorced is tricky, because marriage, divorce, and fertility are interacting endogenous choices. This justifies the use of a quantitative structural model as a tool.

In the empirical part\footnote{The dataset I use in establishing the empirical results and estimating the model is the American Community Survey (ACS) since 2009. Its benefits include large sample size and uniform coverage of the US population with a wide range of household and individual characteristics reported. For supplementary robustness exercises, I also use Survey of Income and Program Participation (SIPP) and National Survey of Family and Growth (NSFG). They have more detailed coverage of a few variables that I am interested in, although their small sample sizes do not allow to condition well on some variables like marriage duration.} I compare two groups of people: those who have their children at least at the next year after their marriage (marriage-first) and those who had their children before or at the year of their first marriage (kids-first). kids-first group is considered as having a shotgun marriage in a broad sense, I refine this definition later.

In the main data exercise I show that the kids-first group have a substantially higher share of divorced than marriage-first (18\% vs 10\%), and this finding is robust to various controlling and partitioning strategies, including picking people only with a particular duration of the marriage, considering separately early and late births, regression adjustment with many controls including income and matching techniques. Figure ... gives a preview of the main empirical result, showing the percentage of women divorced after 5 years of marriage by the relative marriage timing. One important data limitation is that ACS records only the date of the most recent marriage, so these numbers exclude remarried people, which I address later.

Following that I document a remarkable education gradient: there is a substantial difference between these numbers for college graduates: 15\% of kids-first and 5\% or marriage-first women are divorced, and relatively modest for women with no college (17\% vs 14\%). The size of the kids-first group itself is different by education though: only one-tenth of college graduates are kids-first as opposed to one-third for the high school graduates, yet this represents quite a substantial share of the total population.

Finally, using supplementary data from SIPP that has a full marital history I verify that exclusion of remarried people does not affect the patterns I show. Then using NSFH data I further split kids-first onto those who had a true shotgun marriage (married while being pregnant), those who married after the birth of their child and women who married someone else than the biological father of the children. I show that these groups all have comparable shares of divorced, with the one for true shotgun marriage being the highest, and all these shares are higher than for marriage-first. Few other robustness checks are presented and the result still holds --- in all population groups kids-first people divorce more and conditioning on observables does not remove this difference.

To understand this I introduce a structural lifecycle model endogenizing marriage, fertility, and divorce, together with savings and extensive margin of labor supply. The empirical results tell very little about the causes of the differences, and identifying them from the data is tricky as the relative timing of marriage is a choice. 
Marriage is considered as a contract where partners cannot commit not to break up, based on the literature on limited commitment such as Kocherlakota (1996), with dynamic nature as in Marcet and Marimon (2019). The growing branch of literature applies this framework to the collective household models, so my model is related to the work of Voena (2015), Low et al (2018), Mazzocco (2007), or Shephard (2019). The model allows to capture the patterns of selection into marriage and fertility and to separate the effect of unplanned pregnancy on divorce rate with the impact of the composition of the groups. Marriages differ with respect to unobservable marriage quality, which can be interpreted as match-specific utility surplus, and marriage terms, that represent the relative bargaining power of the partners.

The driving force of the selection into the kids-first and the marriage-first groups is fertility shocks: people meet each other infrequently within their lifecycle, and when a potential couple meets, they can have an unplanned pregnancy, and this shifts their bargaining position in the decision of whether to marry. Namely, when a woman refuses to marry following the pregnancy she has to take care of the child herself and faces worsened marriage market opportunities as her new potential partners, if any, may not like having stepchildren. Those who agree to marry following the pregnancy become kids-first couples, and those who did not have pregnancy when they meet can perfectly choose to have kids in the later periods, and when they do so they become a marriage-first couple.

The main conceptual result can be understood through the simple bargaining framework. In the model, given reasonable parametrization, when at the couple's meeting an unplanned pregnancy arrives, a woman's disagreement option becomes worse as being a single mother is hard, and man's agreement option improves, as everyone generally likes kids. Interaction of these two things allows marriages with lower quality to be created, and some couples marry precisely because they have a baby, i.e. comply with the shock. These unions are the marginal marriages with the highest risk of divorce, and they drive most of the difference in the divorce rates between groups. As a result, being kids-first means lower marriage quality in general. Other factors, like distress from mistiming the pregnancy of couples with high-quality, may also contribute to the increase of divorce rates. The magnitude of all these effects depends on preference parameters and household technologies, and therefore quantities in the data are informative about them.

I estimate the model parameters using a simulated method of moments, matching the established ACS evidence to the quantities in the simulated data. The parameters I estimate are the preference for children, the features of household technology, and the transition probabilities. The moments I match are the percentage of divorced women by different durations of marriage in kids-first and marriage-first group, the share of never married and divorced with and without kids by age, the share of couples having children by the duration of their marriage and few others. To account for the education gradient flexibly I perform two distinct estimations on two subgroups --- college graduates and high school graduates. In both subgroups the model fits very well, with few exceptions for the high school graduates where the fit is still reasonable.

The estimates indicate substantial differences in the probability of unplanned pregnancy between the high school and the college groups, while for both of them it is a substantial risk of women with a monetary equivalent of 4--6 yearly incomes. [To be added: few more things on interpretation of the estimates + causal effects implied by the model]

The model is capable of delivering counterfactuals that provide a better understanding of the sources of the differences. [TBD social stigma] [TBD few more counterfactuals]

Finally, the model can be used to study policy design. In the benchmark estimation I match the existing child support schedule with the compliance rates obtained from .... Elimination of child support does [...], and enforcing 100\% compliance does [...]. Additionally, I experiment with re-allocation of custody and changing the property division upon divorce [... ...].

A seminal piece of research regarding shotgun marriage is Akerlof, Yellen, and Katz (1996). Using a simple game theory reasoning they argued that the practice of shotgun marriage essentially disappeared when technologies of contraception and abortion became widespread. First, they argue that before such technologies to have premarital sex men had to commit to marriage in case of unplanned pregnancies, and shotgun marriage was an execution of such commitments. Second, with the contraception technologies the need and the practice of these commitments disappeared, so the demand for premarital sex went up, and the share of women who failed to adopt these new technologies contributed to the rise of single mothers. Therefore, both new technologies and changes in norms are responsible for the recent family trends.

The result that more sex and failure to adopt the contraception technologies are the cause of the rise in single motherhood was debated in the subsequent literature. Neal (2004) have argued that the fact that most of the rise is concentrated among economically disadvantaged population groups is not in favor of the social norms explaining the rise in single parenting, suggesting instead that major improvement of the labor market for women and also of welfare benefits like AFDC played a role. Additionally, Neal (2004) notes that the trends in adoption rates do not suggest that out-of-wedlock children are less wanted now than 40 years ago. Finally, Chiappori, Oreffice (2008) point out that a decrease in the supply of marriageable men among black people perfectly rationalizes the finding of more women deciding to raise children on their own.

The other part of the Akerlof, Yellen, and Katz (1996) conjecture --- disappearance of shotgun marriage as commitment practice, is less obvious to asses directly.
First, the data reject the complete disappearance of a shotgun marriage, it still accounts for a substantial share of marriages and many transitions from cohabitation to marriage. Second, even in the frameworks without commitment, including mine or Chiappori, Oreffice (2008) agents find it beneficial to marry in case of unplanned pregnancy rather than feel obliged to.
Third, apart from non-adoption of contraception technologies, modern couples dynamics, including unstable unmarried cohabitations (as shown by ... or many others) does not completely exclude the possibility of women to have unplanned children, and even in full control of fertility there is some probability of unplanned pregnancy resulting in birth, and at least some part of these pregnancies is unwanted.
Finally, the unmarried cohabitations themselves may be interpreted as a sign of commitment, and social pressure to marry the partner after pregnancy or birth in such unions is a way the society may enforce this. [This can be seen from NSFG data about transitions to marriage from cohabitation]

Few papers have studied shotgun marriages empirically. Alesina, Giuliano (2006) have shown that easier dissolution of marriage caused by the introduction of unilateral divorce laws increased the creation of shotgun marriages, implicitly arguing that women enter more risky unions if they are easier to exit. A recent empirical paper is Tannenbaum (2019), studying an expansion of child support laws in 1977--1992. Focusing exactly on the creation of shotgun marriages, the study finds that the expansion of child support reduced both marriage and abortion rates following a nonmarital pregnancy, leading more women choosing to be single mothers after such an event. Another related contribution is Chiappori, Weiss (2006), who has argued that expansion of the child support can be rationalized by higher remarriage rate and men wanting to provide more consumption to their children in new households. This channel, however, may not work as intended for the mothers who receive welfare, as the child support they receive is typically offset by the reduction of their welfare payments, see Lerman, Sorenson (2003).

%Combining of theory with empirical variations have been 
This work contributes to the broader literature about understanding the general value of marriage, which has been discussed since the work of Becker (1981). Modern studies emphasize roles of risk-sharing, as in Lise and Yamada (2019), the general comparative advantage of being a couple as in Chiappori (1997) and shared production of public good as in Greenwood et al. (2016). While having all these mechanisms in place, my work here primarily focuses on the last one: children are the crucial part of the value that is created within a couple and therefore are first-order issues in how couples are formed. One fundamental question that is possible to address is how large is the value of marriage per se relative to the value of the possibility of the couple to have children together, and exploiting how people choose to enter inferior quality marriages for the opportunity to raise their child together seems directly relevant.

The idea of exploring outcomes of people who married after they got pregnant is similar in spirit to the literature analyzing teenage pregnancy. From pioneering work of Card and Wise (1978) to modern Ashcraft et al. (2013) and Rosenbaum (2019) people were using different data strategies to understand how early births affect outcomes of women, where more recent works suggest that observable impact is driven majorly be self-selection. My work focuses on a distinct angle of the similar issue: although a large chunk of shotgun marriages happens in early life, they do not appear to be the driver of the results: excluding early marriages or controlling for mother’s age at the first birth and the first marriage does not mitigate differences in marital performance
-- shotgun marriages that happen later in life have very similar consequences to the early ones. Finally, my work does not aim to recover the results from the data alone: instead of dealing with the selection empirically, I attempt to model it, though focusing mostly on more narrow issues of marriage performance and couple formation.

The contribution of this paper is threefold. First, I am the first to document the extent of shotgun marriage in the US in different population groups and its robust underperformance in terms of divorce rates. Second, the model I present is the first lifecycle model that includes both marriage and divorce based on limited commitment and endogenous fertility choice, allowing to study their interaction. Third, I am the first to study the impact of child support policies on both single and married couples using a structural model: empirical studies that I mention above focus on the creation of new marriages and existing structural studies Brown et al (2015) and Forester (2020) focus on established couples.

The rest of the paper is structured in the following order. [... ... ...]

%
%\section{Introduction}
%
%
%%Rossin-Slater (2017) provides an excellent review of the issues of maternity leave policies that I also cover here. As a benchmark, most of the world women are guaranteed to have at least 3--4 month of maternity leave and around two-thirds of their usual labor income during the leave. The US with no paid maternity leave policies mandated on the federal level is a large exception. According to the review, there is solid empirical evidence that introduction of paid maternity leave has positive impact on labor market outcomes of women, mainly caused by job continuity, yet long leaves can be even harmful. There is some evidence of maternity leave improving well-being of children, and very limited evidence of the leave affecting the employers. 
%%
%%Many studies document the impact of family regulations like child support laws, visitation rights and custody allocations on the formation of new couples, focusing on how in case of a pregnancy parental responsibility affects the choice of couples to form a union, well-being of children and abortion choices.  The robust finding is that making fathers pay for their children reduce incentives of mothers both to marry and to abort the pregnancy, therefore child support laws increase the number of single mothers. With few exceptions like Brown et al (2015), these studies do not consider existing couples, including childless couples, and how their fertility and divorce rates are affected. As I can show here, this regulations can potentially generate a lot of response for existing couples, with generally more commitment of fathers causing both more fertility and less divorce, with heterogeneous size of these effects among different population groups. 
%%
%%Existing couples are relevant for child support laws by many reasons. First, despite recent moderation in divorce rates, those rates are high, especially for less educated people, so these measures are relevant for very large group of people. Second, the child support in particular relies of establishing paternity of the children, which can be more complicated for never married mothers, and this means that compliance with child support regulations is expected to be generally higher for those who were married in some points. This is confirmed by the following: ... \% of never married mothers with children 1--18 receive child support and \% of divorced mothers receive it. Third, even divorced parents are often considered less vulnerable group, poverty rate of divorced mothers is ... compared to ... in general population, and this poverty is concentrated later in life than for never married women.
%%

\end{document}
\section{Quantification}
\begin{enumerate}
\item ACS table
\item Graphs by dT
\item raw and comparable sample
\item observables cannot explain
\item SIPP table
\end{enumerate}

I use American Community Survey (ACS) samples of 2009--2015 for the main empirical exercise and the model estimation. Its advantage is large sample sizes with very precise household data, bearing enough information to recover the parts of marriage and fertility history I am interested in and to deliver very fine partitions of the data, like focusing only on people surveyed a certain number of year after their marriage. Its main disadvantage is incompleteness of the marriage and fertility histories. To address them I use few supplementary datasets representing the same population but providing more detailed coverage, which I discuss further.

The driver of decisions I discuss is whether couple is expecting (or already has) a baby when making their marriage decision. To classify the couples by timing, I consider the following simple measure:
\begin{equation}
\Delta T = \text{Year of the first marriage} -  \text{Year of the first birth}
\end{equation}
assuming both events happened by the time the person is observed. Due to the nature of household survey data, these variables are easier available for females than males, I henceforth focus on women in most of the analysis. Figure ... presents a motivating graph: the histogram shows distribution of women with given $\Delta T$ between $-5$ and $5$, and the scatter plot shows percentage of those divorced five years after marriage with this value of $\Delta T$. The further parts of this section elaborate on the choices and restrictions of the analysis.

Based on the $\Delta T$ I call kids-first (KF) women those who have $\Delta T \in \{-5,...,-1,0\}$ and marriage-first (MF) those who have $\Delta T \in \{1,...,10\}$. Most of the comparisons are done within this groups. Together with those never married and childless, the dataset completely excludes women with marital statuses ``spouse absent'', ``widowed'' and ``separated'', which I have no intent to capture.

Lag between pregnancy and childbearing is important yet not crucial for the classification. The main group of interest are people with $\Delta T = 0$. Assuming that average pregnancy lasts a little less than nine months, even with random timing less then one-quarter of these couples will be married before their pregnancy. One still can imagine a couple who married in January, got pregnant in March and gave a birth in December, but this requires very precise timing of decisions and, moreover, marrying no later than March, which is uncommon in particular due to cold season in most of the US. By similar reasoning, some share of couples with $\Delta T = 1$ did actually marry while being pregnant, and yet those people are less constrained in the time.

Finally, a substantial share of couples has $\Delta T < 0$, meaning out-of-wedlock childbirth preceding the marriage. In most of the empirical I pool this group with those who had a true shotgun marriage (married while being pregnant). A possible concern is that some share of these couples would have different marriage mechanics and be a result of single mother remarrying someone else then the child's father. To mitigate this, I restrict my attention to cases where $\Delta T \geq -5$. With this restriction the couples with $\Delta T = 0$ and $\Delta T \in [-5,-1]$ have reasonably similar observable characteristics, including the percentage of divorced. In subsection ... I show that majority of couples with $\Delta T \leq 0$ do report to have biological children. Lastly, even if this induces a bias for interpretation of the empirical part, the remarriage process is fully captured by the structural model I use.

There are two important limitations of the ACS data. First, only the date of the most recent marriage is reported. Therefore, whenever the women need to be classified within the ACS, I restrict my attention to only those married once. This may bias the judgement of percentage of divorced if people in kids-first and marriage-first groups remarry with different rates.  Second, instead of women-specific year of the first birth the ages of (her own) children in the household are reported. To mitigate this I focus only on relatively young women 21--40, who are more likely to reside with their eldest child. By subtracting the age of the eldest child from the survey year I can get a reasonable estimate of what year the first birth happened. Thinking about the concrete effect of these imprecisions on the results is tricky, yet they can be fixed with a better fertility history. Further in the section I show that the data without these constraints display very similar results.

To quantify the divorce rate, or, more precisely, share of divorced, I construct two kids of measures. First, the data allows to pick people who were surveyed $y$ years after their marriage, and percentage of divorced among them is a reasonable estimate of cumulative marriage dissolution probability as they have the same marriage duration. The drawback of this is that it picks a particular subset of people. Therefore, I also report a simple cross-sectional measure, that is a percentage of divorced people in a group. It can be interpreted as a weighed average of duration-specific divorce rates, and if distribution of the durations is not systematically different in two groups it can capture aggregate divorce rates differences well.

Table ... shows difference in the divorce rates in ACS by groups. Column \% KF shows percentage of kids-first in population of both kids-first and marriage-first couples. I present the share of divorced for the whole cross-section, and then to test the boundaries of the result I consider few smaller samples: to obtain comparable divorce rates I use people five and ten years after their marriage, and to illustrate heterogeneity I show partition by education and the exclusion of young-age births.

The table illustrates few patterns. First, people in kids-first group systematically have higher divorce rates. Second, this difference is more pronounced for college graduates, despite the smaller yet sizable share of the kids-first group for them: with comparable absolute difference marriage-first group is relatively much more stable. Third, although very related with the previous one, the difference is more visible for women who give a birth later rather than earlier. Fourth, there is substantial mixing of the groups in cross-sectional data, making composition an important factor.

To address the composition differences and also give a sense of standard errors, I run the same comparisons as simple regressions: $\text{Divorced}_i =\Delta\cdot \text{KF}_i + \gamma \cdot X_i + \varepsilon_i$. on a sample of kids-first or marriage-first, where $X_i$ represents possible controls. Raw difference in divorce rate corresponds to $\hat\Delta$ when $X_i$ contains only constant. Ratio corresponds to $\hat r = 1 + \frac{\hat{\Delta}}{\text{Avg Divorced if MF}}$. I present two regressions with controls. The first regression includes fixed effects variables for .... The second regression additionally controls for a third degree polynomial in log per-hour income and is done on subsample of women who is employed, work more than 10 hours per week and more than 48 weeks for the last year and report labor earnings at least \$3000 per year.

Table ... presents the regression results together with standard errors. The main conclusion is that correction for observables, even very non-parametric and flexible, cannot explain any significant part of the difference: net of all the confounding effects kids-first couples still divorce significantly more.

The following subsection and Appendix ... provide few extra checks and additional insights. Shortly, using more detailed and more complete data does not change any of the patterns I discuss here. 
\subsection{Additional Empirical Results}
\begin{enumerate}
\item Full marital history on SIPP
\item Fine partition on NSFG
\item Appendix: Historic Prospective on Census data
\item Appendix: Duration model on NSFH
\item Appendix: household characteristics in KF and MF couples
\item Appendix: some extra SIPP stuff
\end{enumerate}

\section{Theoretical Framework}

\section{Mechanisms and Discussion}

\section{Estimation and Fit}

\section{Policy Experiments}

\section{Results and Discussion}

\section{Conclusions}

\end{document}

### 2020-10-02 main/draft.tex
Current stage

@@
 \usepackage{mathrsfs}
+\usepackage{natbib}
 \usepackage{tikz}
@@
 \usetikzlibrary{positioning}
+
 \tikzstyle{block} = [rectangle, draw, rounded corners]
@@
 \newcommand{\hypo}{\mathcal{H}}
-\bibliographystyle{ieeetr}
 
@@
 \newcommand{\bsA}{\boldsymbol{A}}
-
 %\setstretch{1}                         
@@
 \setlength{\topsep}{0pt}              
-\usepackage[pdftex,unicode,colorlinks=true,urlcolor=blue]{hyperref}
+\usepackage[pdftex,unicode,colorlinks=false,urlcolor=blue,citecolor=black,hidelinks]{hyperref}
 \usepackage{bbm}
@@
 I am grateful to Alessandra Voena for fantastic amount of help and support.
-I thank my colleagues Jane Olmstead-Rumsey, Kristina Manysheva, Bence Bardoczy, Stephanie Johnson and Gabriela Cugat, as well as all the participants of Macro Lunch at Northwestern for lots of helpful comments.}\\
+I thank my colleagues Jane Olmstead-Rumsey, Kristina Manysheva, Bence Bardoczy, Stephanie Johnson and Gabriela Cugat, as well as all the participants of Macro Lunch at Northwestern for lots of helpful comments.
+I owe special thanks to Fabio Blasutto for tons of help with both technical and substantial parts during his stay at Northwestern.}\\
 {\small Northwestern University}}
-\date{August, 2019}
+\date{First Draft: May, 2019 \\[1em] This Draft: February, 2020}
 \begin{document}
+\bibliographystyle{apa}
 %\begin{center}\textbf{Economics 416-1} \\ \emph{By Egor Kozlov}\end{center}
@@
 Many couples marry shortly after woman gets pregnant, which is often referred to as a shotgun marriage.
-I show that in American Community Survey data couples who experienced a shotgun marriage, i.e.
+I show that in the US household data couples who experienced a shotgun marriage, i.e.
 had their first child before or at the year of their marriage, divorce more often in the future.
-%This feature is robust in different subpopulations, being relatively more pronounced for wealthier people, although the share of such marriages is lower among them.
 I argue that this difference in divorce rates can be explained by endogenous selection into marriage: women who experienced unplanned pregnancy agree to enter marriages of lower quality.
@@
 Then I show that these patterns in selection can be rationalized by the structure of childcare costs, where kids require both money and time, and disproportional exposure of women to these costs.
-I estimate a structural model endogenizing marriage selection and use it to argue that causal impact of unplanned fertility shock,
-%differences in marriage quality,
-as opposed to different composition of the groups, drives the major part of the observable difference in divorce chances.
+I estimate a structural model endogenizing marriage selection and use it to argue that causal impact of unplanned fertility shock, as opposed to different composition of the groups, drives the major part of the observable difference in divorce chances.
 Qualitatively, this suggests that the main reason for poor performance of shotgun marriages is large, unexpected and heterogeneous childcare costs.
@@
 First, I document and quantify the persistent difference in divorce rates of couples who experienced shotgun marriages and how it is pronounced for different population groups.
-Second, based on these findings I estimate a structural model that is capable of identifying importance of different channels that drive this difference.
-%In this paper I argue that this pattern is rationalized by partners agreeing to enter marriages of lower quality as a response to unplanned pregnancy shock.
-
-%[Paragraph 2: general importance? Child development]
-Understanding the sources of a variation in divorce rates is not just a household behavior question.
+Second, based on these findings I estimate a structural model that matches the observable variation in the data and is capable of identifying importance of different channels that drive this difference.
+
+Combining the empirical variation with the model provides better understanding of people's marriage decisions and general value of marriage, which is being discussed since the the work of \cite{becker1981}.
+Modern studies emphasize roles of risk-sharing, as in \cite{liseyamada}, general comparative advantage of being a couple as in \cite{chiappori1997} and shared production of public good as in \cite{greenwood}.
+Having all these mechanisms in place the current work primarily focuses on the last one: children are the crucial part of the value that is created within a couple and therefore are first order issue in how couples are formed.
+Using the data alone provides pretty limited insights for this problem, and this justifies using a structural model capable of replicating the observed variation.
+One fundamental question that is possible to address is how large is the value of marriage per se relative to the value of possibility of couple to have children together, and exploiting how people choose to enter inferior quality marriages for the opportunity to raise their child together seems directly relevant.
+
+The idea of exploring outcomes of people who married after they got pregnant is similar in spirit to the literature analyzing teenage pregnancy.
+From pioneering work of \cite{cardwise} to modern \cite{ashcraft} and \cite{rosenbaum} people were using different data strategies to understand how early birth affect outcomes of women, where more recent works suggest that observable impact is driven majorly be self-selection.
+My work focuses on a distinct angle of the similar issue: in fact, although large chunk of shotgun marriages happens in early life, they do not appear to be the driver of the results: excluding early marriages or controlling for mother's age at the first birth and at the first marriage does not mitigate differences in marital performance --- shotgun marriages that happen later in life have very similar consequences to the early ones.
+Finally, my work does not aim to recover the results from the data alone: instead of dealing with the selection empirically I attempt to model it, though focusing mostly on a more narrow issues of marriage performance and couple formation.
+
+Understanding the sources of a variation in marriage performance is not just a household behavior question.
 For example, divorces have major impact on how people raise their children.
@@
 Growing up with a single mother or in a blended family (i.e.
-with step-parents) if conventionally considered\footnote{See Kearney, Levine 2017\nocite{kearney} for an extensive review of these issues} as a predictor of poor early and adult-life outcomes of the children, and those who have two married biological parents are typically not considered to be at risk.
+with step-parents) is conventionally considered as a predictor of poor early and adult-life outcomes of the children that is reviewed, for instance, by \cite{kearney}.
+Moreover, work of \cite{chetty} suggests that high shares of single mothers in neighborhoods are very strong predictor of low intergenerational mobility and reproduction of inequality.
+In these studies those who have two married biological parents are typically not considered to be at risk.
 However, not all marriages are created equal, and treating all marriages equally may understate inequality: growing up in unstable marriage can be a predictor of ending up with a single mother or in a different household, as well as factor of risk by its own.
@@
 These comparisons, however, contain lots of selection issues and unobservable heterogeneity, and to deal with it I set up a structural model that is capable of understanding of anatomy of the result.
-The model is based on a growing branch of literature in dynamic household bargaining, and endogenizes savings, marriage and divorce decisions, fertility decisions and fertility shocks, together with the structure of childcare costs.
+The model considers marriage as a contract where partners cannot commit not to break up, based on literature on limited commitment such as \cite{kocherlakota}, with dynamic nature as in \cite{marcetmarimon}.
+The growing branch of literature applies this framework to the collctive household models, so my model is related to such as \cite{voena-1}, \cite{mazzocco} or \cite{shephard}.
+In the model I endogenize savings, marriage and divorce decisions, fertility decisions and fertility shocks, together with the structure of childcare costs.
 The last element is crucial in generating heterogeneous response in terms of income or education: as time of wealthier parents is more expensive, unexpected childcare costs create relatively more distress for them.
-The model is estimated to match the key patterns of the data coming from American Community Survey, as well as supplementary sources, including ATUS and CEX for disciplining childcare costs part, and the structure I use fits the data well.
+The model is estimated using simulated method of moments to match the key patterns of the data coming from American Community Survey, as well as supplementary sources, including ATUS and CEX for disciplining childcare costs part, and the structure I use fits the data well.
 
@@
 Importance and even presence of this channel, however, depend on parameters of individual preferences and childcare costs that I estimate structurally.
+However, as a generic feature, after a pregnancy shock happen, the key reason that women agree to stick to the poor-quality matches is splitting the unexpected costs, and the key reason for men is the access to their children.
+This suggests that subsidizing childcare or enforcing visitation rights would decrease number of such marriages at the cost of increasing the number of single parents.
 
 %[Paragraph 9: model --- what I get from it, mechanics and anatomy]
-The key conclusion of the estimation part is that couples who experienced an unplanned pregnancy shock indeed agree to enter the marriages of the lower quality, and this is the major driver of the number that we observe in the data.
+The estimation results confirm lower unobserved marriage quality of those who experienced an unplanned pregnancy shock and suggest this to be the major driver of the number that we observe in the data.
 The model is capable of delivering various decompositions that split observable cross-sectional difference in divorce rates onto effects of age, marriage length, income, wealth and marriage quality.
@@
 Second, the general impact of pressure of becoming a single mother on marriage decisions is substantial, and eliminating unexpected pregnancies happening before the marriage decision takes place lowers amount of people married by 25 by [...].
-[Last, the model predicts decline in probabilities of unexpected pregnancies since 1980: the model matching fertility / marriage / divorce profiles from Census of 1980 predicts larger chances of unplanned pregnancies among all income and age groups.]
+%Third, subsample analysis suggest that the most important driver of stronger impact of shotgun marriages on college graduates relative to non-college graduates can be rationalized by larger weight of child quality in their utility.
+%[Last, the model predicts decline in probabilities of unexpected pregnancies since 1980: the model matching fertility / marriage / divorce profiles from Census of 1980 predicts larger chances of unplanned pregnancies among all income and age groups.]
 
@@
 %[Paragraph 12: what I learned about channels]
-The main causal insight is that being a single mother is a major threat affecting how people make their marriage decisions, and large and unequally distributed costs of raising young children can be a driver in the creation of poor quality marriages.
-This channel can be of major importance for studying welfare policies subsidizing childcare, young families and single mothers, as well as for enforcing alimony payments and child support in general.
+The main causal insight is that being a single mother is a major threat affecting how people make their marriage decisions, and large and unequally distributed costs of raising children can be a driver in the creation of poor quality marriages.
+This channel can be of major importance for studying welfare policies subsidizing childcare, young families and single mothers, as well as for enforcing alimony payments, child support and visitation rights in general.
 This also emphasizes the link between contraception and abortion access and marriages, although this impact is limited as even the wealthiest couples do experience sizable share of shotgun marriages, though this may be cultural rather than economic issue.
@@
 \section*{Literature}
+%\bibliographystyle{achicago}
 \bibliography{bib}

### 2020-10-02 poster-march-20/posterfancy3.tex
Current stage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% baposter Landscape Poster
% LaTeX Template
% Version 1.0 (11/06/13)
%
% baposter Class Created by:
% Brian Amberg (baposter@brian-amberg.de)
%
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[landscape,a0paper,fontscale=0.285]{baposter} % Adjust the font scale/size here

\usepackage{graphicx} % Required for including images
\graphicspath{{figures/}} % Directory in which figures are stored

\usepackage{amsmath} % For typesetting math
\usepackage{amssymb} % Adds new symbols to be used in math mode
\usepackage{geometry}
\usepackage{booktabs} % Top and bottom rules for tables
\usepackage{enumitem} % Used to reduce itemize/enumerate spacing
%\usepackage{palatino} % Use the Palatino font
\usepackage{graphicx}
\usepackage{subfig}
\usepackage{caption}
\usepackage{booktabs}
\usepackage[font=small,labelfont=bf]{caption} % Required for specifying captions to tables and figures

%COLORS
\definecolor{bordercol}{RGB}{40,40,40} % Border color of content boxes
\definecolor{headercol1}{RGB}{186,215,230} % Background color for the header in the content boxes (left side)
\definecolor{headercol2}{RGB}{80,80,80} % Background color for the header in the content boxes (right side)
\definecolor{headerfontcol}{RGB}{0,0,0} % Text color for the header text in the content boxes
\definecolor{boxcolor}{RGB}{186,215,230} % Background color for the content in the content boxes
\definecolor{purple}{RGB}{91, 59, 140}

\definecolor{bleunuit}{RGB}{0,38,84}
\definecolor{cyan}{RGB}{96,175,221}
\definecolor{bleugris}{RGB}{155,175,196}
\usepackage{multicol} % Required for multiple columns
\setlength{\columnsep}{0.5em} % Slightly increase the space between columns
\setlength{\columnseprule}{0mm} % No horizontal rule between columns
\usepackage{float}
\usepackage{tikz} % Required for flow chart
\usetikzlibrary{shapes,arrows}
\usetikzlibrary{positioning}
\tikzstyle{block} = [rectangle, draw, rounded corners]
\tikzstyle{line} = [draw, -latex']

\newcommand{\compresslist}{ % Define a command to reduce spacing within itemize/enumerate environments, this is used right after \begin{itemize} or \begin{enumerate}
\setlength{\itemsep}{1pt}
\setlength{\parskip}{0pt}
\setlength{\parsep}{0pt}
}

%Lenght Between figure and caption
\setlength{\abovecaptionskip}{15pt plus 0pt minus 2pt} 
\setlength{\tabcolsep}{2.5pt}
%Header Height
\setlength{\headsep}{5mm}
\setlength{\topskip}{50mm}
%\setlength{\oddsidemargin}{0mm}
%\setlength{\textwidth}{184.3mm}
%\setlength{\columnsep}{6.3mm}
%\setlength{\marginparsep}{0mm}
%\setlength{\marginparwidth}{0mm}
%\setlength{\textheight}{265.2mm}
\definecolor{lightblue}{rgb}{0.145,0.6666,1} % Defines the color used for content box headers
\begin{document}

	\tikzstyle{block} = [draw, fill=white, rectangle, 
	minimum height=2.5em, minimum width=6em]
	\tikzstyle{sum} = [draw, fill=white, circle, node distance=2cm]
	\tikzstyle{input} = [coordinate]
	\tikzstyle{output} = [coordinate]
	\tikzstyle{pinstyle} = [pin edge={to-,thin,black}]
	%Here I put the title, first slide	
%	\setpostertemplate{itemize items}[circle]
%	\setbeamertemplate{enumerate items}[circle]

\begin{poster}
{
headerborder=closed, % Adds a border around the header of content boxes
colspacing=0.1em, % Column spacing
bgColorOne=white, % Background color for the gradient on the left side of the poster
bgColorTwo=white, % Background color for the gradient on the right side of the poster
borderColor=lightgray, % Border color
headerColorOne=purple, % Background color for the header in the content boxes (left side)
headerColorTwo=purple, % Background color for the header in the content boxes (right side)
headerFontColor=white, % Text color for the header text in the content boxes
boxColorOne=white, % Background color of the content boxes
textborder=roundedleft, % Format of the border around content boxes, can be: none, bars, coils, triangles, rectangle, rounded, roundedsmall, roundedright or faded
eyecatcher=false, % Set to false for ignoring the left logo in the title and move the title left
headerheight=0.1\textheight, % Height of the header
headershape=roundedright, % Specify the rounded corner in the content box headers, can be: rectangle, small-rounded, roundedright, roundedleft or rounded
headerfont=\large\bf\textsc, % Large, bold and sans serif font in the headers of content boxes
%textfont={\setlength{\parindent}{1.5em}}, % Uncomment for paragraph indentation
headerheight=2.2cm,
columns=4,
linewidth=0.1pt % Width of the border lines around content boxes
}
%----------------------------------------------------------------------------------------
%	TITLE SECTION 
%----------------------------------------------------------------------------------------

{{\color{bleunuit}\LARGE\bf\textsc{The Economics of Shotgun Marriage}\vspace{0.3em}}} % Poster title
{{\color{bleunuit}\LARGE{Egor Kozlov  \hspace{2pt}---\hspace{2pt} Northwestern University, \textbf{egorkozlov2020@u.northwestern.edu}}}} % Author names and institution
%----------------------------------------------------------------------------------------
%	Motivation
%----------------------------------------------------------------------------------------

\headerbox{1. Shotgun Marriages Are Widespread and Do Not End Well}{name=motivation,column=0,row=-0.015,span=2}{
	\begin{minipage}{0.65\linewidth}
	\footnotesize
	\begin{itemize}
\item Many people marry shortly after a baby appears
\begin{itemize}
\item Shotgun marriage = first child before or at the year of marriage
\item $\sim10\%$ marry while being pregnant
\item Even more marry after their first birth
\item Not just teenage pregnancies, many among 25+ 
\item I view these pregnancies as unplanned yet not exogenous
\end{itemize}
\item People who had a shotgun marriage divorce more in the data 
\begin{itemize}
\item A lot more for college grads, a bit more for high school grads
\end{itemize}
\item Understanding why helps to understand the value of marriage
\item Questions:
\begin{itemize}
\item What is the aggregate role of unplanned pregnancies?
\item What causes people to enter such marriages?
\item Why they perform worse?
\end{itemize}
\end{itemize}
	\end{minipage}
	\hfill
	\begin{minipage}{0.35\linewidth}	
			\begin{center}
			\includegraphics[scale=0.3]{graph_dt_poster.eps}
			 \includegraphics[scale=0.3]{share_kj_poster.eps}
			 \end{center}
    \end{minipage}
   }

\headerbox{2. Model Is Required Here}{name=rq,column=2,row=-0.015,bottomaligned=motivation,span=1}{
	\vspace*{-0.4mm}
{\footnotesize
\begin{itemize}
\item Quantify empirical difference of marriage performance by the timing of childbirth:
\vspace{-1mm}
\begin{itemize}
\item Main sample: ACS 2009--2016, women  21--40 with one marriage (limited marriage history)
\vspace{-1mm}
\item Robustness (full history): SIPP, NSFG, NSFH
%, cross-sectional regressions and duration models
\end{itemize}
\item Estimate a lifecycle model matching this difference
\vspace{-2mm}
\begin{itemize}
\item limited commitment: marriage and divorce
\item savings and heterogeneous productivity
\item fertility choice and shocks
\end{itemize}
\vspace{-2mm}
\item Use the model to study:
\vspace{-1mm}
\begin{itemize}
\item aggregate impact of unplanned childbirths 
\item importance of fertility for marriage
\item difference between education groups
\item policy counterfactuals
\end{itemize}
\end{itemize}
}
}

%----------------------------------------------------------------------------------------
%	Methodology
%---------------------------------------------------------------------------------------
\headerbox{3. Bad Performance Data Facts}{name=meth,column=3,bottomaligned=motivation,row=-0.015,span=1}{
	\vspace*{-0.4mm}
	\footnotesize
	\[\Delta T = \text{Age 1st birth} - \text{Age 1st marriage}.\]
	\begin{itemize}
	\item Kids First (KF): $\Delta T \in \{-5,...,0\}$
	\item Marriage First (MF): $\Delta T \in \{1,...,10\}$
	\end{itemize}
	\vspace{-1em}
	\begin{center}
	\begin{tabular}{|l|c|c|c|c||c|}\hline
	\multicolumn{6}{|c|}{ \textit{ACS: women, married once and have kids}, 21--40}\\\hline
	& \multicolumn{2}{|c|}{\% divorced} & ratio & ratio & \% of  \\ 
	& \textbf{MF} & \textbf{KF} &  (raw) &   (reg)  & KF\\\hline
	Everyone  & 10.1 & 18.0 & \textbf{1.8} & 1.6 & 22.8 \\
	High School & 13.9 &  17.2 & \textbf{1.2} & 1.6 &  33.3 \\
	College & 5.3  & 14.7 & \textbf{2.8}  & 1.7 &  9.5\\
	First birth 25+ & 7.2 & 10.8 & \textbf{1.5} & 1.6 & 8.6 \\\hline
	\multicolumn{6}{|p{0.8\linewidth}|}{\scriptsize These are cross-sectional shares of divorced. ``Divroced'' is being divorced at the moment of the survey. ``Reg'' refers to regression-adjusted ratio with controls: age, age at 1st marriage, age at 1st birth, education, race, state,  ACS year, number of children}\\\hline
	\end{tabular}
	\end{center}
	\vspace{-0.5em}
	\textbf{Main conclusion:} \textit{women who experience a shotgun marriage experience more divorces in all subsamples.}
}

%----------------------------------------------------------------------------------------
%	Data
%---------------------------------------------------------------------------------------
\headerbox{4. The Results Are Stable}{name=data,column=0,row=0.355,span=1}{
\footnotesize
Very similar conclusions are obtained by
	\begin{itemize}
	\item ACS (large samples but incomplete history):
	\vspace{-0.5em}
	\begin{itemize}
	\item Controlling for income (if full-time working)\vspace{-0.25em}
	\item Comparing only marriages of the same duration\vspace{-0.25em}
	\end{itemize}
	\vspace{-0.5em}
	\item SIPP: smaller sample with better marital history (not just married once)\vspace{-0.25em}
	\item NSFH and NSFG: allow to run duration analysis\vspace{-0.5em}
	\begin{itemize}
	\item rel hazards of divorce $\approx$ cross-sectional ratios\vspace{-0.25em}
	\item $\Delta T$ in months, exclusion of step children\vspace{-0.25em}
	\end{itemize}
	\end{itemize}

Additional insights:
	\begin{itemize}
	\item Marriages with $\Delta T = 0$ have the worst performance
	\item Divorces are concentrated around child's age 5--10
	\item Children in KF couples repeat grades more often
	\vspace{2em}
	\end{itemize}
	}
%
%\headerbox{4. Model: Summary}{name=data,column=0,row=0.355,span=1}{
%\footnotesize
%\begin{itemize}
%\item Key features:
%\vspace{-0.5em}
%\begin{itemize}
%\item Males $m$, females $f$; singles $s$, couples $c$
%\item Assets ($a$) and productivity ($z^f$, $z^m$), ``love'' $\psi$
%\item Endogenous marriage and divorce, $p^{\text{meet}}(\text{Age})$
%\item Unplanned pregnancy can happen \textbf{when potential partners meet}: $p^{\text{unplanned}}(\text{Age})$
%\item \textbf{Married couples}: optimal fertility (1 child)
%\end{itemize}
%\item Unplanned pregnancy triggers marriage because of:
%\begin{itemize}
%\item Large childcare expenditures need to be shared
%\item Marriage prospects of single mothers are worse
%\end{itemize}
%\textbf{Result:} preg shock $\Rightarrow$ couples with lower $\psi$ marry
%\end{itemize}
%}

\headerbox{5. Model: Limited Commitment + Optimal Fertility In Couple + Pregnancy Shocks When Meet}{name=theory,column=1,row=0.355,span=3}{
\vspace{-0.25cm}
\hspace{-0.75cm}
\begin{tabular}{c c c c}
\begin{minipage}{0.3\linewidth}
\footnotesize
\begin{itemize}
\item Key features:
\vspace{-0.75em}
\begin{itemize}
\item Males $m$, females $f$; singles $s$, couples $c$\vspace{-0.25em}
\item Heterogeneous earnings $W$ and savings $a$\vspace{-0.25em}
\item Couple-specific $\psi$ --- relationship quality\vspace{-0.25em}
\item Random but assortative mating\vspace{-0.25em}
\item Endogenous marriage and divorce\vspace{-0.25em}
\item Unplanned pregnancy can happen \textbf{when potential partners meet}\vspace{-0.25em}
\item \textbf{Married couples} have optimal fertility\vspace{-0.25em}
\end{itemize}
\item Unplanned pregnancy $\Rightarrow$ low-quality marriage:
\begin{itemize}
\item Childcare expenditures need to be shared
\item Marriage prospects of single mothers drop
\end{itemize}
\textbf{Result:} preg shock $\Rightarrow$ couples with lower $\psi$ marry
\end{itemize}
\end{minipage}
&
\hspace{-0.25cm}
\begin{minipage}{0.27\linewidth}
\vspace{0.0cm}
\begin{center}
\includegraphics[scale=0.35]{scheme.png}
\end{center}
\vspace{-2em}
{\scriptsize \hspace{0.4cm} Marry if $\exists \theta$ s.t.:\vspace{-0.25em}
\begin{itemize}
\item { $\scriptstyle \text{No shock: } V^{f,c}(\theta) \geq V^{f,s} \text{ and } V^{m,c}(\theta) \geq V^{m,s}$}\vspace{-0.4em}
\item { $\scriptstyle \text{Shock: } V^{f,ck}(\theta) \geq V^{f,sk} \text{ and } V^{m,ck}(\theta) \geq V^{m,s}$}\vspace{-0.4em}
\item {$\scriptstyle \text{Single mom} \, = \, \text{Shock, but } V^{m,ck}(\theta) \geq V^{m,s} - \phi^r$}
\end{itemize}
}
\end{minipage}

&
\hspace{-2em}
\begin{minipage}{0.225\linewidth}
{\footnotesize
\vspace{0.3em}
\hspace{2em}\textbf{Value Functions:}
\vspace{-0.5em}
\[\hspace{-2em}m \text{ --- marriage }, \ \ \ d  \text{ --- divorce}\]
\vspace{-2em}
\begin{align*} \scriptsize
& V^{f,s}_t =  u(c_f) + \beta\mathbb{E}_t \big[(1-p^{\text{meet}}_t)\cdot V^{f,s}_{t+1}  \\ + & p^{\text{meet}}_t \big\{ p^{\text{preg}}_t (m_{t+1} V^{f,ck}_{t+1} + (1-m_{t+1})V^{f,sk}_{t+1})  
\\  +  &  (1-p^{\text{preg}}_t) (m_{t+1} V^{f,c}_{t+1} + (1-m_{t+1})V^{f,s}_{t+1})  \big\} \big].
\end{align*}
\vspace{-0.5cm}
\begin{align*} \scriptsize
& V^c_t =  \theta^f \cdot u(c_f) + \theta^m \cdot u(c_m) + \psi \\
+ & \beta\mathbb{E}_t \big[\max\{V^{ck}_{t+1},V^{c}_{t+1}\}\cdot (1-d_{t+1}) 
\\ +  & d_{t+1} \cdot(\theta^f\cdot V^{f,s}_{t+1} + \theta^m\cdot V^{m,s}_{t+1})\big].
\end{align*}
\vspace{-0.5cm}
\begin{align*} \scriptsize & V^{c,k}_t =  \theta^f \cdot u(c_f) + \theta^m \cdot u(c_m) + \psi  \\
+ & u^{\text{kid}}(Q) +  \beta\mathbb{E}_t \big[V^{ck}_{t+1}\cdot (1-d_{t+1}) 
\\ +  & d_{t+1} \cdot(\theta^f\cdot V^{f,sk}_{t+1} + \theta^m\cdot V^{m,s}_{t+1})\big].
\end{align*}
%\begin{itemize}
%\item $\theta$ stays constant if p.c. don't bind
%\item $\theta$ may be renegotiated
%\item If no agreeable $\theta$ $\Rightarrow$ divorce
%\item Low $\psi$ $\Rightarrow$ more likely to divorce
%\end{itemize}
%\hspace{2em}\textbf{Upon Divorce}:
%\begin{itemize}
%\item Only moms pay childcare
%\item Fathers lose access to their kids
%\item Single moms remarry $\Rightarrow$ utility loss of $\phi^r$ for their new partner
%\end{itemize}
}
\end{minipage}
&
\hspace{-2em}
\begin{minipage}{0.225\linewidth}
\vspace{-0.5cm}
{\footnotesize
%\textbf{Notation:} $c$ --- consumption, $\theta$ --- Pareto weight, $\psi$ --- common love shock, $x$ --- child expenditures, $l$ --- labor supply, $t$ --- age.
\begin{itemize}
%\item Child's consumption: $Q = F(x,1-l^f)$
\item S.t. participation constraints:
\vspace{-0.5em}
\[\scriptstyle V^{f,c\bullet}_t \geq V^{f,\text{divorce}}_t, \ V^{m,c\bullet}_t \geq V^{m,\text{divorce}}_t,\]
\vspace{-2em}
\item Violated $\Rightarrow$ renegotiation or divorce
\item Childcare ($x$ --- money, $l^f$ --- labor)
\vspace{-1.0em}
\[\scriptstyle Q = f(x,1-l^f), \ \ Q \geq \bar{Q} \ (\text{subsistence})\]
\vspace{-1.5em}
\item Budget ($\rho>1$ --- returns to scale): 
\vspace{-0.75em}
\[\scriptstyle (c_f^{\rho} + c_m^{\rho})^{\frac1{\rho}} + a' + x = R\cdot a + l^f \cdot W^f_t + W^m_t\]
\[\scriptstyle W^i_t= \exp\left\{z_i + \text{Trend}_{i,t}\right\}\]
\vspace{-1.75em}
\item Shocks: $\scriptstyle z_m' = z_m + \varepsilon^{z}_m$, $\scriptstyle \psi' = \psi + \varepsilon^{\psi}$ 
\item Female productivity can depreciate:
\vspace{-0.75em}
\[\scriptstyle z_f' = z_f -  \text{{\boldmath $\delta (l^f)$}} + \varepsilon^{z}_f.\]
\item Female stays with the child and do not get any child support if divorced
\end{itemize}
}
\end{minipage}
\end{tabular}
}

%----------------------------------------------------------------------------------------
%	Estimation
%---------------------------------------------------------------------------------------
\headerbox{6. Estimation: MSM}{name=estimation,column=0,span=1,row=0.662}{
\footnotesize
	 \begin{itemize}
	 \item Two samples: college and high school graduates
	 \item Income process and trend: match ACS
	 \item I fix externally:\vspace{-0.5em}
	 \begin{itemize}
	 \item Couple's returns to scale $\rho$ (Voena, 2015)\vspace{-0.25em}
	 \item Skill depreciation (Adda et al, 2017)\vspace{-0.25em}
	 \item Substitutability between $l$ and $x$ (Sommer, 2016)\vspace{-0.25em}
	 \item Standard $\beta = R^{-1} = 0.98$ \vspace{-0.25em}
	 \end{itemize}
	 \item Method of Simulated Moments picks parameters for:\vspace{-0.5em}
	 \begin{itemize}
	 \item Transition probabilities: $p^{\text{preg}}_t$, $p^{\text{meet}}_t$\vspace{-0.25em}
	 \item Choice parameters: $u^{\text{kid}}(Q)$, $Q(x,1-l_f)$, $\bar{Q}$\vspace{-0.25em}
	 \item Love shock distribution: $\sigma(\varepsilon^\psi)$, $\sigma(\psi_0)$.\vspace{-0.25em}
	 \item Drop in marriage prospectives after a child $\phi^r$\vspace{-0.5em}
	 \end{itemize}
	 \item In estimation, I keep preference and technology parameters as in college graduates for both samples
	 \item KF--MF difference is mostly driven by probability to meet a partner $p^{\text{meet}}_t$, childcare requirement $\bar{Q}$	and $\phi^r$
	 \end{itemize}
	 \vspace{0.75em}
}

%----------------------------------------------------------------------------------------
%	Fit
%---------------------------------------------------------------------------------------
\headerbox{7. All Dimensions Fit Well}{name=fit,column=1,span=2,aligned=estimation}{
\footnotesize

\begin{tabular}{ c c}
\begin{minipage}{0.45\linewidth}
Main things I target are:\vspace{-0.25em}
 \begin{itemize}
 \item \% of never married, divorced and childless by age\vspace{-0.25em}
 \item Fertility in couples by years after marriage\vspace{-0.25em}
 \item \% of KF (among married \& with kids)\vspace{-0.25em}
 \item \% of divorced KF and MF (cross-sectional)\vspace{-0.25em}
 \item Labor supply w/ kids and childcare expenditures\vspace{-0.25em}
 \item \% of single mothers (mar and never mar) at 30
 \end{itemize}
 \end{minipage}
&
	\begin{tabular}{ l r r}\hline
	\multicolumn{3}{ c }{ College graduates (selected dimensions) }\\
Target & Model & Data \\\hline
\textit{Never mar at (25,30,35,40), \%} & $(77,45,21,9)$ &  $(75,38,21,15)$ \\
 \textit{No kids at (25,30,35), \%}  & $(87,67,46)$ & $(90,60,34)$ \\
 \textit{No kids (1,2,3) years aft mar, \%} & $(70,61,55)$ & $(81,66,51)$ \\
 \textit{Divorced at (25,30,35,40), \%}  & $(5,7,10,9)$ & $(4,7,9,12)$ \\
 \textit{ \% of KF couples at (25,30,35)}  & $(21,10,10)$ & $(21,12,10)$ \\
 \textit{Divorced if (KF, MF), \%}  & $(5,15)$ & $(5,15)$ \\\hline
\end{tabular}
 \end{tabular}
}

\headerbox{8. Why College Is So Different and What Is Overall Effect?}{name=overid,column=1,span=2,below=fit,bottomaligned=estimation}{

	\begin{minipage}{0.5\linewidth}
	\footnotesize

\begin{center}
\begin{tabular}{ l c c c}
\multicolumn{1}{r}{{\textbf{Decomposition}} \ \ \ \ \ \ \ \ \textit{ \% of divorced if...}} & \textit{MF} & \textit{KF} & \textit{Ratio} \\\hline
Baseline: all estimates for college grads & 5.3 & 15.0 &   2.8  \\
College, but income process of HS & 3.9 & 15.3 & 4.0 \\
College, but $p^{\text{preg}}_t$ of HS & 5.5 & 14.4 & 2.6 \\
College, but $p^{\text{meet}}_t$ of HS & \textbf{8.9} & \textbf{12.6}  & \textbf{1.4} \\
College, but $\sigma(\varepsilon^\psi)$, $\sigma(\psi_0)$ of HS & 5.1 & 14.5 & 2.9 \\
College, but $\psi^r$ of HS & 5.0 & 15.1 &  3.0 \\
Comparison: all estimates for high school &  11.8 & 16.6 & 1.4\\
\end{tabular}
\end{center}

	\end{minipage}
	\hfill
	\begin{minipage}{0.5\linewidth}
	\footnotesize
	\begin{tabular}{l c c c c}
	& \multicolumn{2}{c}{\textit{\% married at ...}} & \multicolumn{2}{c}{\textit{\% divorced at ...}} \\
	& \textit{25} & \textit{35} & \textit{25} & \textit{35} \\\hline
{Baseline} & 23 & 79 & 5.1 & 9.8 \\
{No unplanned} & 22 & 81 & 3.2 & 9.5 \\
{No kids at all} & 22 & 81 & 3.5 & 8.3 \\\hline
%{... + no love ($\psi=0$) } & 60 & 99 & 13& 40 \\
%... + no rts ($\rho=1$) & 10 & 32 & 29 & 35 \\
	\end{tabular}
	\vspace{-0.2em}
	\begin{itemize}
	\item Around 12\% of couples marry \textit{because of} the pregnancy
	\item These \textit{compliers} account for most of the divorces
	\item Shotgun marriage $\Rightarrow$ +6 p.p. to be divorced at 40 (col)
	\end{itemize}
\end{minipage}

}

%----------------------------------------------------------------------------------------
%	Experiments
%---------------------------------------------------------------------------------------
\headerbox{9. Take Away}{name=experiment,column=3,span=1,row=0.662}{
\hspace{-1em}
\begin{minipage}{\linewidth}
	{\footnotesize
	\begin{itemize}
	\item \textbf{Unplanned pregnancies have high costs even if people marry}
	\item Childcare costs rationalize all the patterns well
	\item Other factors like social stigma against unmarried parenting may play a role, but are not required
	\item People enter risky shotgun marriages because:
	\vspace{-0.5em}
	\begin{itemize}
	\item Men: limited access to their children otherwise\vspace{-0.25em}
	\item Women: sharing childcare costs\vspace{-0.25em}
	\item Women: decrease in remarriage prospectives\vspace{-0.25em}
	\item Chances to meet new partners decrease with age
	\end{itemize}
	\vspace{-0.25em}
	\item Enforcing child support and visitation rights lead to more single mothers here 
	\item Unplanned pregnancies explain 2/5 of early divorces
	\item Different meeting process explains lots of differences in college vs high school --- marriage markets matter
	\item Model-generated causal impact is similar to the observable cross-sectional difference
	%\item KF is a good predictor of inferior relationship quality
	\end{itemize}
	}

\end{minipage}
}

%----------------------------------------------------------------------------------------
%	Take Away
%---------------------------------------------------------------------------------------
%\headerbox{10. Take Away}{name=theory,column=3,row=0,span=1,below=experiment,bottomaligned=overid}{
%	{\footnotesize
%	\begin{itemize}
%	\item People enter shotgun marriages because:
%	\begin{itemize}
%	\item Men: access to their children
%	\item Women: sharing childcare costs
%	\item Women: decrease in marriage prospectives
%	\end{itemize}
%	\item Unplanned pregnancies explain 2/5 of early divorces
%	\item 
%	\end{itemize}
%	}
%}

\end{poster}

\end{document}

### 2020-10-02 quest-proposal/proposal-2020.tex
Current stage

\documentclass[11pt,letter]{article}
\usepackage[left=0.8in,right=0.8in,top=1in,bottom=1in]{geometry}
\usepackage{amsmath}
\usepackage{pdflscape}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{multicol}
\usepackage{multirow}
\usepackage{microtype}
\usepackage{euscript}
\usepackage{epsfig}
\usepackage{epstopdf}
\usepackage{mathrsfs}
\usepackage{natbib}
\usepackage{tikz}
\usepackage{verbatim}
\usetikzlibrary{shapes,arrows}
\usetikzlibrary{positioning}

\tikzstyle{block} = [rectangle, draw, rounded corners]
\tikzstyle{line} = [draw, -latex']
\newcommand{\hypo}{\mathcal{H}}

\usepackage[flushleft]{threeparttable}

\usepackage[english]{babel} 
\DeclareMathOperator{\rank}{rank}
\newcommand*{\hm}[1]{#1\nobreak\discretionary{}
            {\hbox{$\mathsurround=0pt #1$}}{}}

            \def\onepc{$^{\ast\ast}$} \def\fivepc{$^{\ast}$}
\def\tenpc{$^{\dag}$}
\def\legend{\multicolumn{4}{l}{\footnotesize{Significance levels
:\hspace{1em} $\dag$ : 10\% \hspace{1em}
$\ast$ : 5\% \hspace{1em} $\ast\ast$ : 1\% \normalsize}}}

\newcommand{\bs}[1]{\boldsymbol{#1}}  
\newcommand{\bsA}{\boldsymbol{A}}
%\setstretch{1}                         
\flushbottom                            
\righthyphenmin=2                      
\pagestyle{plain}                       
%\settimeformat{hhmmsstime}  
\widowpenalty=300                   
\clubpenalty=3000                     
\setlength{\parindent}{0em}           
\setlength{\topsep}{0pt}              
\usepackage[pdftex,unicode,colorlinks=true,urlcolor=blue,citecolor=blue]{hyperref}
\usepackage{bbm}
\usepackage{tabularx}
\renewcommand{\emptyset}{\varnothing}

\setlength{\parskip}{0.5\baselineskip plus2pt minus2pt}

\newcommand{\e}{\varepsilon}
\DeclareMathOperator*{\Argmax}{\mathrm{Argmax}}
\DeclareMathOperator*{\Argmin}{\mathrm{Argmin}}
\DeclareMathOperator*{\argmax}{\mathrm{arg\,max}}
\DeclareMathOperator*{\argmin}{\mathrm{argmin}}

\newcommand{\blp}{\mathrm{BLP}}
\DeclareMathOperator*{\plim}{\mathrm{plim}}
\DeclareMathOperator{\Max}{\mathrm{Max}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\Y}{\mathcal{Y}}
\newcommand{\Z}{\mathcal{Z}}
\renewcommand{\geqslant}{\geq}
\renewcommand{\leqslant}{\leq}
\newcommand{\p}{\bs p}
\newcommand{\y}{\bs y}
\def\dd#1#2{\frac{\partial#1}{\partial#2}}

\renewcommand{\emptyset}{\varnothing}

\DeclareMathOperator{\tr}{\mathrm{tr}}

\newcommand{\bb}{\bs \beta}
\newcommand{\X}{\bs X}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\PP}{\mathbb{P}}
\DeclareMathOperator{\V}{\mathbb{V}}
\DeclareMathOperator{\CM}{\mathbb{C}}
\renewcommand{\C}{\CM}
\DeclareMathOperator{\var}{\mathrm{var}}
\DeclareMathOperator{\cov}{\mathrm{cov}}
\DeclareMathOperator{\corr}{\mathrm{corr}}
\DeclareMathOperator{\MSE}{\mathrm{MSE}}
\DeclareMathOperator{\Bias}{\mathrm{Bias}}
\renewcommand{\P}{\PP}
\newcommand{\dsim}{\stackrel{d}{\sim}}
\newcommand{\hn}{\mathcal{H}_0}
\newcommand{\ha}{\mathcal{H}_a}
\newcommand{\thetab}{\bs \theta}
\newcommand{\pv}{\text{P-value}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\MLE}{\scriptscriptstyle MLE}
\newcommand{\LR}{\mathrm{LR}}
\newcommand{\I}{\mathbb{I}}
\newcommand{\sumin}{\sum\limits_{i=1}^n}
\newcommand{\sumti}{\sum\limits_{t=0}^\infty}
\newcommand{\hbeta}{\hat{\beta}}
\newcommand{\halpha}{\hat{\alpha}}
\newcommand{\hsigma}{\hat{\sigma}}
\newcommand{\hvar}{\widehat{\var}}
\newcommand{\hcov}{\widehat{\cov}}
\newcommand{\Q}{\mathbb{Q}}

\begin{document}

\title{The Economics of Shotgun Marriage (Quest Proposal)\footnote{\textbf{egorkozlov2020@u.northwestern.edu}. I am a graduate student at Economics Department entering my year 6. My dissertation committee is: Matthias Doepke (committee chair, \textbf{doepke@northwestern.edu}), Alessandra Voena (\textbf{avoena@uchicago.edu}), Marti Mestieri (\textbf{mmestieri@frbchi.org}).}}
\author{Egor Kozlov}
\date{\today}
\maketitle
\bibliographystyle{apa}

\begin{abstract}
This briefly describes the project I am requesting a Research II Quest allocation for. This project is going to be the leading chapter of my dissertation and my job market paper and I expect to finish it in Fall 2020. I have been using a Research I allocation \verb'p30190' by this moment, but as my model estimation requires lots of experimentation I have a need for extra resources for this summer to make sure I do not lose my priority in a critical moment.
\end{abstract}

\section{Project Summary}

The broad motivation for the study lies within the issues of transformation of the US families and reproduction of inequality. Researchers often compare single parents to two-parent families in terms of future outcomes of the children. I argue that this division leaves out an important distinction: not all two-parent families are alike and some couples marry precisely because they had a baby, and this practice is called a shotgun marriage. Share of these marriages is comparable to the share of single-parent households. In the paper I show that those kids-induced marriages are the least stable in sense of divorce rates and their creation is the most sensitive to changes in the economic environment. Perhaps more surprisingly, the difference in performance between shotgun and regular marriages is more pronounced among more educated and adult couples. To understand the mechanics of formation of the shotgun marriages I set up and estimate a structural lifecycle model with endogenous fertility and limited commitment. Through the lens of the model I can identify factors causing different performance of shotgun marriages in different demographic groups and get credible answers to economic policy questions like child support and visitation rights, divorce laws and welfare programs, accounting for their impact on household formation. These issues are widely understudied: realistic economic models of marriage and divorce are complicated mathematically, involve multidimensional dynamic optimization and require large computational resources.  Several successful papers on these topics were developed in the recent years, including \nocite{voena-1}Voena (2015), \nocite{low-1}Low et al (2018), \nocite{shephard}Shephard (2019), \nocite{reynoso}Reynoso (2018). Yet many aspects of the marriage markets are underexplored.

The economics of the result is generated by endogenous selection into marriage. In the model women who experience unplanned pregnancy agree to enter marriages of lower quality. In addition to the higher divorce rates, the data about how women in shotgun marriages use their time and how often do their children repeat grades suggest the validity of the marriage quality explanation. These features in the marriage selection can be rationalized by the structure of childcare costs, where kids require both money and time, and disproportional exposure of women to these costs, so the women agree to lower quality marriages in a response to a threat of becoming a single mother. Social pressure for partners to marry each other once they have a child also contributes to the result, but is not likely to be the leading explanation. I estimate a structural model endogenizing marriage selection and use it to argue that causal impact of unplanned fertility shock, as opposed to different composition of the group experiencing shotgun marriage, drives the major part of the observable difference in divorce chances. Qualitatively, this suggests that the main reason for poor performance of shotgun marriages is large, unexpected and heterogeneous costs of unplanned pregnancies. Quantitatively, the model suggests sizable effects of risks of unplanned pregnancy on observable divorce rates, especially for young women, and substantial impact of policies enforcing commitment, like child support, on formation of these low-quality marriages.

At the current stage of the project, I have preliminary estimates of the model for two demographic groups: college graduates, who display the highest impact of shotgun marriage in the data, and high school graduates, for whom this impact is relatively modest. In the both groups, the model successfully reproduces the data about many dimensions of marriage, fertility and divorce rates in the US. Those estimates suggest, for instance, that strict enforcement of child support decreases difference in the divorce rates by one-third to one-half, and the monetary valuation of risk of getting into a shotgun marriage is within the magnitude of 4-6 yearly earnings for both college and non-college mothers, even though the incidence of these marriages is lower for college women. Few things, however, are not conclusive yet. The main of them is what is the main cause of the difference between college and non-college couples in terms of divorce: depending on relatively minor details of calibration this can be either attitudes towards childbearing, the characteristics of potential partners or the re-marriage prospectives after a divorce. This part is yet to be done.

In general, the project is approaching its final stage, and this is the reason I need extra computing powers: I want my estimates to be stable, precise and reproducible and to see how deviations in the model assumptions translate to different conclusions. Few things are yet to be done: the model needs state-of-the-art estimates of stochastic processes for labor income of college and non-college individuals and a reasonable approximations of pools of potential partners in terms of income and wealth for men and women at each age. Policy angles are also pretty undeveloped: even when the model is settled, focusing on concrete economic policy questions may require unexpected adjustments to the setup.

\section{Computational Problem}

The mathematical framework for the model is summarized \nocite{mazzocco}Mazzocco (2007). Shortly, partners in a married couple cannot commit not to divorce, so the behavior of the couple is influenced by the value of their options to divorce and finding new partners to remarry. Need to both allocate resources in the most efficient way and simultaneously provide better options than splitting creates a recursive dynamic contracting problem, which has been formalized as a saddle point problem by \nocite{marcetmarimon}Marcet, Marimon (2019). On a discrete grid summarizing couple's wealth, each spouse's income, relationship quality and marriage terms I have to solve a problem of optimal allocation of resources between spouses and children consumption and savings, such that the behavior is efficient and provides to the both spouses more value than splitting up. If this allocation cannot be found, the contract is broken: the couple divorces and the partners re-enter the marriage market looking for new partners.

Mathematically, the model is solved in discrete time using Bellman-like method. Possible characteristics of the couple are approximated on a discrete grid where some variables follow Markov transition process and some variables need to be chosen. For ages from 80 to 21, given the expected value of entering the age $t+1$ with given characteristics I can solve for optimal actions (like savings, amount of work, consumption, child expenditures, giving a birth or divorce), satisfying the participation constraints that I described above (optimization stage). After that, given all transition probabilities I can compute an expected value of entering the period $t$ (integration stage) and proceed backwards. Both stages, essentially, involve brute force global search of the best values on discrete grids, as well as global search for sets of points satisfying certain conditions. These tasks are highly parallelizable and GPUs can be involved. After solving the model for given parameters, I get the decision rules of the agents and simulate the transition of finite number of individuals for ages 21 to 80, starting from some initial conditions. On this simulated data I compute targets, like percentage of married or divorced at each age, and measure variance-weighed (Mahalanobis) distance of these targets to the real data, that allows me to estimate the model fit.

The process of estimation of the model is, essentially, a non-linear least squares problem: I am looking for \[\Theta^* = \argmin\limits_\Theta \sum\limits_i w_i\cdot (T_i^{\text{simulations}}(\Theta) - T_i^{\text{data}})^2,\]
where $T_i$ are the targets and $w_i$ are the data-based weights. Two major challenges for this is expensive function evaluation and the noise of the simulations. For each choice $\Theta$ I need to solve and simulate the model from the beginning, and this takes around 3 minutes on a single core and around 1 minute using GPUs, where the latter number can be improved with investing more time and using better frameworks. Second, the simulated moments are not smooth in $\Theta$ as they involve many discrete decisions, including fertility, marriage and divorce. The simulated data are not smooth by nature, additionally, with a very small movement of the parameters there is a chance that no targets would change because of discreteness. This structure prevents using gradient-based solvers like BFGS. To solve this problem, I adopt TikTak global optimization method suggested by \nocite{benchmarking}Arnoud et al (2019): I first run a massive global stochastic search using $N$ points from Sobol sequence for $\Theta$, then pick $M$ of the best points among them and run series of local derivative-free optimizers iteratively (I use DF-OLS by \nocite{dfols}Cartis et al (2018)), where at each iteration the starting point for a local solver is a combination of the best point so far and one of the $M$ best Sobol points.

\section{Computational Plans and Estimates}

The project is implement on Python, and used NumPy with speeds up from Numba, including Numba-CUDA interface. There is a piece of work in progress that also uses JAX libraries, which potentially can give me much faster execution on GPUs, as the current bottlenecks of the code is slow transfer to and from the GPU device and many limitations of Numba-CUDA. For the data part I occasionally use Stata MP.

Here are the rough time estimates, most of which are in terms of CPU time, as I paused my usage of GPUs dues to the fall in priority. It takes around 3 minutes of one-core and 10-15 Gb of memory to solve the model once. Typically the memory constraint is binding, but with the best attempts I can bring it down to 5--7 Gb per core, which is still pretty high. For the estimation part, $\Theta$ typically contains 10--15 parameters, and the estimation requires \textit{15 jobs with one core and 20 Gb memory on each running around 20--24 hours}, corresponding to solving the model 7-10 thousand times (and translating to 300-400 CPU hours on \verb'genhimem' partition). Each change to the model requires running the estimation again, so I run those things many times and by this moment I probably exceeded 35{,}000 hours limit. Yet as things keep improving I except to do this more during the summer. It is not likely that will exceed 100{,}000 hours of the total usage in the end, and I feel that the main bottleneck is with the amount of memory rather than CPU hours.

\section{Support From Quest}
I may need help with using PETSC with Python, as PETSC includes TAO optimization libraries which I may benefit to use instead of my current algorithm. Previously I also had some issues with making CUDA toolkit for Python work, but I think I resolved them though maybe not in the most elegant way.

\section{Links and Other}

The most recent presentation (May 2020) is available \href{https://drive.google.com/file/d/18xt01V2PYm8XG87fu-pRWxwXoRtPYRHG/view?usp=sharing}{here}.

The current paper draft (last updated February and missing some features) is \href{https://drive.google.com/file/d/1BF--Pq8fCSGkPl8CS0L89G-tQeybVNMH/view?usp=sharing}{here}.

The external presentations for the project included the Fall 2019 Midwest Macroeconomics Meetings (Michigan State University), H2D2 Conference at University of Michigan (March 2020) and Spring 2020 IRES Lunch (UCLouvain, Belgium, postponed to Fall due to Covid-19).

Thank you for your consideration!

\bibliography{bib}

\end{document}

### 2020-10-02 slides-aug-20/slides.tex
Current stage

\documentclass[aspectratio=169]{beamer}
\usepackage[english]{babel}
\usepackage{amsmath,amsfonts}
\usepackage{multicol}

\usepackage{IEEEtrantools}
\usepackage{multirow}
% beamer setup

%\usecolortheme{dove}

\setbeamertemplate{navigation symbols}{}
\setbeamertemplate{itemize items}[ball]

\usepackage{tikz}
\usetikzlibrary{shapes,arrows}
\usetikzlibrary{positioning}
\tikzstyle{block} = [rectangle, draw, rounded corners]
\tikzstyle{line} = [draw, -latex']

\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator*{\argmax}{argmax}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\I}{\mathbb{I}}

\AtBeginSection[]{
  \begin{frame}[plain]
  \addtocounter{framenumber}{-1}
  \vfill
  \centering
  %\begin{beamercolorbox}[sep=8pt,center,shadow=true,rounded=true]{title}
    %\usebeamerfont{title}
    \Huge{\insertsectionhead\par}%
  %\end{beamercolorbox}
  \vfill
  \end{frame}
}

\newcommand{\backupbegin}{
   \newcounter{framenumberappendix}
   \setcounter{framenumberappendix}{\value{framenumber}}
}
\newcommand{\backupend}{
   \addtocounter{framenumberappendix}{-\value{framenumber}}
   \addtocounter{framenumber}{\value{framenumberappendix}} 
}

\title{The Economics of Shotgun Marriage}
\subtitle{and Household Bargaining}
\author{Egor Kozlov}

\institute{
  Department of Economics\\
  Northwestern University}

\setbeamertemplate{footline}[frame number]

%  \usepackage{pgf}
%\logo{\pgfputat{\pgfxy(0,0)}{\pgfbox[right,base]{\footnotesize{\insertframenumber\,/\,\inserttotalframenumber}}}}
%\newcommand{\nologo}{\setbeamertemplate{logo}{}}

\let\olditem\item
\renewcommand{\item}{%
\olditem\vspace{\fill}}

\begin{document}

% REMOVE SLIDE COUNT

\begin{frame}[plain]
\addtocounter{framenumber}{-1}
\date{\scriptsize}
\titlepage
\end{frame}

\begin{frame}
\frametitle{``America's strongest anti-poverty weapon''?}
\begin{itemize}
\item Married two-parent families \ : \ desirable outcomes for parents and children%$\genfrac{}{}{0pt}{1}{\Leftarrow}{\Rightarrow}$
\item Pro-marriage policies: child support, tax brackets, divorce regulations, etc
\item Can economic policies create successful marriages? Love vs money.
\end{itemize}
\end{frame}

\begin{frame}

\frametitle{Kids-First Marriages are Less Stable}
\begin{enumerate}
\item{\textbf{Shotgun marriage}} --- couple marrying when woman gets pregnant
\item{\textbf{Kids-first marriage}} --- child \textit{before or in the year} of marriage
\end{enumerate}
\begin{itemize}
\item Around 1/4 marriages in the US today are kids-first
\item They have much higher divorce rates, especially among college graduates
\item Marginal in economic sense: could be sensitive to policies 
\end{itemize}
\end{frame}

%
%\begin{frame}
%\frametitle{Questions}
%\begin{itemize}
%\item Drivers of inequality:
%\begin{itemize}
%\item Focusing on unmarried young single mothers omits an important transition
%\item Unplanned pregnancies look different between age and education groups
%\item Divorced parents may be worse than never married
%\end{itemize}
%\item Value of marriage:
%\begin{itemize}
%\item Do pregnancies trigger marriages? If so, why?
%\item If shotgun marriage is risky, why people agree?
%\end{itemize}
%\end{itemize}
%
%\end{frame}

\begin{frame}
\frametitle{Roadmap}
\begin{enumerate}
\item Document the shotgun marriage and its outcomes empirically
\item Use these empirical relations to estimate a model capturing:
\begin{itemize}
\item dynamic bargaining and limited commitment in marriage
\item lifecycle, income and savings
\item fertility and child expenditures
\end{itemize}
\item Look at what drives decisions to marry and the differences in divorce
\begin{itemize}
\item what explains the education gradient?
\item can policies fix the difference or reduce the share?
\item what is the value of marriage in general?
\end{itemize}
\end{enumerate}
\end{frame}

\begin{frame}
\frametitle{Summary of Results}
\begin{itemize}
\item 23\% of women had the first child before their first marriage (kids first)
\item They are 1.8 times more likely to be divorced than ``marriage first''
%\item College grads: 2.8 times more likely to be divorced
%\begin{itemize}
%\item though smaller kids first share (just 10\%)
%\end{itemize}
\item My structural model corrects selection and data imperfections
%\item Model results:
%\begin{itemize}
%\item Social stigma + threat of raising the child alone explains most of the difference
%\item Child support $\Rightarrow$ more single mothers, little effect on marriage decisions
%\item Being able to control fertility helps everyone
%\end{itemize}
\item Model answers: Love $>$ Money. In particular:
\begin{itemize}
\item Threat of raising the child alone explains most of the difference
\item Regulating marriages by policies is both hard and inefficient
\item Easier control over fertility fixes shotgun marriages and has bigger return
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Literature}
\begin{itemize}
\item Policy literature:
\begin{itemize}
\item Empirical: Alessina, Guiliano, 2005;  Tannenbaum, 2020; Rossin-Slater, 2017
\item Structural: Forester, 2020; Brown, Flinn, 2011
\item Focus on particular policies, ignore interactions, limited quantitative power
\end{itemize}
\item Lifecycle literature:
\begin{itemize}
\item Dynamic limited commitment for marriage and divorce: Mazzocco, 2007; Voena, 2015; Low et al., 2018; Shephard, 2019; Blasutto, 2020
\item Fertility during lifecycle: Sommer, 2015, Ejrnæs, Jørgensen, 2018
\item Link between divorce and fertility is missing
\end{itemize}
\item Empirics on shotgun marriages (demography and sociology):
\begin{itemize}
\item Gibson-Davis et al (2016), Guzzo (2009),  Nuevo-Chiquero (2014)
\item Mostly descriptive
\end{itemize}
\end{itemize}
%
%\begin{itemize}
%\item Empirical literature documents the scope of shotgun marriage around the world:
%\begin{itemize}
%\item Gibson-Davis et al (2016), Guzzo (2009),  Nuevo-Chiquero (2014)
%\item Mostly descriptive
%\end{itemize}
%\item Dynamic limited commitment models of divorce and marriage:
%\begin{itemize}
%\item Marcet, Marimon (2019), Mazzocco (2007) --- setting up the framework
%\item Voena (2015), Shephard (2019); Lafortune, Low  (2019) --- applied questions
%\item Mostly ignore or simplify interaction with fertility
%\end{itemize}
%\item Unplanned and teenage pregnancy literature:
%\begin{itemize}
%\item Ejrnæs, Jørgensen (2018) --- lifecycle model
%\item Card, Wise (1978) to Ashcraft et al (2013) --- lots of empirical papers
%\item Treat household structure as fixed or exogenous
%\end{itemize}
%\end{itemize}
\end{frame}

\section{Data}
\begin{frame}
\frametitle{Sources: ACS 2009--2017 and others}
\begin{itemize}
\item Main exercises and model estimation are done on American Community Survey
%\item Huge sample size and enough information to show many things 
\item Individual marriage and fertility history are observed partially
\begin{itemize}
\item Only the most recent marriage date --- restrict to ``married once'' when matters
\item Pick age $\in [21,40]$ to have good information on fertility
\end{itemize}
\item Smaller datasets with better histories for robustness and extra evidence:
\begin{itemize}
\item Survey of Income and Program Participation (SIPP)
\item More: NSFG and NSFH
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}[label=kfandmf]
\frametitle{Kids First and Marriage First}
\[\Delta T = \text{Year The First Child Born} - \text{Year The First Marriage Happened}.\]
\begin{itemize}
\item Marriage First \textbf{(MF)}: $\Delta T \in \{1,2,...\}$ (kids at least at the next year)
\item Kids First \textbf{(KF)}: $\Delta T \in \{...,-2,-1,0\}$ (shotgun marriage in a wide sense)
\item Not a full partition%: at 30 only 31\% on the US women have kids and exactly one marriage
\item Most of the comparisons are within this group
\end{itemize}
\hyperlink{extra-restrictions}{\beamerbutton{Extra Restrictions}} 
\end{frame}

\begin{frame}[label=dt_graph]
\frametitle{Kids First Women Are Divorced More Often: Just Look}
\begin{center}
\includegraphics[scale=0.75]{div_5y_by_dt.pdf}
\end{center}
\vspace{-1cm}
\hyperlink{dt_graph_educ}{\beamerbutton{by education}} 
\end{frame}

\begin{frame}
\frametitle{Kids First Women Are Divorced More Often: Quantification}
\framesubtitle{ACS 2009--2017, Women 21--40, married once}
\begin{center}
\begin{tabular}{l r r r r }
\hline
& \multicolumn{2}{c}{share of divorced if ... }&  \\
&  marriage-first & kids-first & (share of kids-first) &  \\\hline
\multicolumn{5}{l}{\textbf{All sample}} \\\hline
\textit{Cross-sectional share of divorced} &  9.9 & 18.1 & (26.1) \\
\textit{Divorced 5 years after marriage} &  5.1 & 14.3  & (24.6) \\
\textit{Divorced 10 years after marriage} & 10.4 & 22.8 & (24.4) \\\hline
\multicolumn{5}{l}{\textbf{Cross-sectional share of divorced for subsamples}} \\\hline
\textit{High school only} &  12.8 & 17.3 & (37.0) \\
\textit{Some college} & 14.3 & 21.0 & (31.3) \\
\textit{College or more} &   5.3 & 14.8 & (11.9) \\\hline
\textit{First birth before 25} & 15.5 & 19.9 & (41.8) \\
\textit{First birth at 25 or later} &  6.2 & 10.9 & (10.7)  \\\hline
\end{tabular}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Kids First Women Are Divorced More Often: Controls}
\begin{center}
%\vspace{-0.5cm}
\begin{tabular}{l r r r }
\multicolumn{4}{c}{\textit{Regression equation:} $\text{Divorced}_i = \Delta \cdot \text{kids-first}_i + \text{Controls}_i + \varepsilon_i$} \\\hline
\hline
& \multicolumn{2}{c}{Estimates}  \\
& Difference ($\Delta$) & (standard error)  \\\hline
\multicolumn{4}{l}{\textbf{All sample, regression}} \\\hline
\textit{No controls (raw difference) }& 8.2 & (0.1)  \\
\textit{Demographic controls }& 5.7 & (0.1)  \\
\textit{Duration controls} &  5.8 & (0.2)  \\
\textit{Demographic + duration} &  4.8 & (0.2)  \\
\textit{Demographic + duration + income} & 4.1 & (0.3)  \\\hline
\multicolumn{4}{l}{\textbf{All sample, propensity score matching}} \\\hline
\textit{Demographic controls} & 5.0 & (0.3)  \\
\textit{Demographic controls + income} & 4.9 & (0.2)  \\\hline
\multicolumn{4}{p{0.825\linewidth}}{\footnotesize \textit{Controls:} demographics: FE for age $\times$ education, race, state, survey year. Duration: FE for age $\times$ age of the first marriage and age $\times$ age of the first birth.  Income: 3rd degree polynomial of log gross income (if observed).}\\\hline\hline
\end{tabular}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Kids First Women Are Divorced More Often: Extra Evidence from SIPP}
\begin{center}
\begin{tabular}{l r r }
\hline
\multicolumn{3}{l}{\textbf{Including remarried people...}} \\\hline
& \multicolumn{2}{c}{share of divorced if ... }  \\
&  marriage-first & kids-first   \\\hline
\textit{Ever divorced} & 34.4 & 48.0  \\
\textit{Divorced if married once} &  12.4 & 19.3 \\
\textit{Remarried} &  22.0 & 28.7 \\\hline
\hline
\multicolumn{3}{l}{\textbf{Separating step and own children (2+ kids only)}} \\\hline
& share in sample & share of divorced   \\\hline
{Marriage-first} & 72.1 & 10.5 \\
{Kids-first-own} &  16.4 & 14.6 \\
{Kids-first-other} &  11.4 & 27.2 \\\hline
\multicolumn{3}{p{0.7\linewidth}}{\footnotesize \textit{Notes.} SIPP data, the cross-section of Wave 1, 2014, women 21+. Numbers are percents.}\\\hline\hline
\end{tabular}
\end{center}
\end{frame}

%\begin{frame}
%\frametitle{Kids First Women Are Divorced More Often: Robustness \& Extra}
%\begin{itemize}
%\item SIPP exercise: smaller sample but full marital history and fertility history
%\begin{itemize}
%\item Similar ratios with larger magnitudes when include remarried people
%\end{itemize}
%\item Duration model: NSFH, historic data from 80s
%\begin{itemize}
%\item Hazard ratio of 1.45, college: 2.23
%\end{itemize}
%\item More details: NSFG, 2015--2017 data
%\begin{itemize}
%\item KF: 30\% true shotgun, 50\% marry after birth, 20\% marry new partner
%\item True shotgun have the highest share of divorced
%\end{itemize}
%\end{itemize}
%\end{frame}
%
%
%\begin{frame}
%\frametitle{Need For The Model}
%\begin{enumerate}
%\item Observables provide limited information. Missing things: relationship quality, potential partners, future divorces
%\item Data are not perfect. Limited histories, unobservable income if out of labor force, step or own children
%\end{enumerate}
%\end{frame}

\section{Model}
\begin{frame}
\frametitle{Summary: Lifecycle, Fertility, Divorce}
\begin{itemize}
\item Lifecycle model with finite horizon
\item People are heterogeneous with respect to age $t$, wages $w$ and savings $a$
\item Couples also differ by relationship quality $\psi$ and spouses' decision power $\theta$
\item Fertility: endogenous decision to have a child (just one) + shocks
\item Limited commitment: decisions are affected by outside options of divorce
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Modeling Shotgun Marriage}
\begin{center}
\vspace{-0.6cm}
\begin{tikzpicture}[every text node part/.style={align=center}, scale=0.75]
   % Place nodes
   \node [block] (1) {\large \textbf{People}  \\[-0.5ex]  \large  \textbf{Randomly Meet}};
   \node [block, below left = 1.5 cm and 0.0 cm of 1] (2) {\textit{Nash} \\[-0.5ex] \textit{Bargaining}};   
   \node [block, below right = 1.5 cm and 0.0 cm of 1] (3) {\textit{Nash} \\[-0.5ex] \textit{Bargaining}};

   \node [block, below = 1.5 cm of 2] (5) {\textbf{Kids First} \\[-0.5ex] \textbf{Couple} ($ck$)};
   \node [block, below = 0.8 cm of 3] (7) {No Kids \\[-0.5ex] Couple ($c$)};
   \node [block, below = 0.8 cm of 7] (8) {\textbf{Mariage First} \\[-0.5ex] \textbf{Couple} ($ck$)};
   \draw[->] (1) to [bend left = 20] node [left] {\scriptsize Unplanned \\[-0.75ex] \scriptsize pregnancy} (2);
   \draw[->] (1) to [bend right = 20] node [right] {\scriptsize No \\[-0.75ex] \scriptsize pregnancy} (3);
   \draw[->] (2) to node [left] {\scriptsize Agree} node [right] {\scriptsize \textbf{Shotgun} \\[-0.75ex]  \scriptsize \textbf{marriage}} (5);
   \draw[->] (3) to node [right] {\scriptsize Agree} node [left]  {\scriptsize \textbf{Regular} \\[-0.75ex]  \scriptsize \textbf{marriage}} (7);
   \draw[->] (7) to node [right] {\scriptsize Kids by choice} (8);   
    \node [block, above left   = 0.8 cm and -0.75 cm of 2] (4) {\,Single male \\[-0.5ex] ($m,s$)\,\,};
    \node [block, above left   = 1.8 cm and -0.75 cm of 2] (4-sm) {\,Single mom \\[-0.5ex]  ($f,sk$)};
    %\node [block, left = 0.2 cm of 4-sm] (4-nsm) {\,Single female \\[-0.5ex]  ($f,sk$)};
	\node [block, above right   = 0.8 cm and -0.75 cm of 3] (6) {\,\,Single male\,\,\\[-0.5ex] ($m,s$)};
    \node [block, above right   = 1.8 cm and -0.75 cm of 3] (6-f) {Single female\\[-0.5ex] ($f,s$)\,};   
    \draw[->] (2.180) to [bend left = 30] node [left] {\scriptsize Disagree} (4.270);
    \draw[->,thick,dotted] (4-sm.20) to [bend left = 10] node [above right=-0.2cm and 0.7cm] {\scriptsize Costly abortion \\[-0.75ex] \scriptsize (if possible)} (6-f.50);
    \draw[->] (3.0) to [bend right = 30] node [right] {\scriptsize Disagree} (6.270);
     \draw[->] (6.180) to [bend left = 30] node [above] {\scriptsize Back}  (1.-12);
   \draw[->] (6-f.180) to [bend right = 30] node [above] {\scriptsize Back} (1.12);
   \draw[->,thick,dashed] (4-sm.90) to [bend left = 30] node [above] {\scriptsize Worse marriage prospectives}  (1.168);
   \draw[->] (4.0) to [bend right = 30] node [above] {\scriptsize Back} (1.192);

\end{tikzpicture}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Value Function: Female, Single}
\framesubtitle{$p^m_t$ --- probability of meeting, $p^p_t$ --- probability of unplanned pregnancy, $m_t$ --- agree to marry}
\begin{align*}
V^{f,s}_t(a,w) = \max\limits_{c,a'} \Bigg\{ \frac{c^{1-\sigma}}{1-\sigma} + \beta \cdot \E_t \Big\{  & 
%& \\ \text{ (did not meet anyone): \ } & 
(1-p^{\text{m}}_t)\cdot V_{t+1}^{f,s}(a',w') \\
\text{ \footnotesize (met and not pregnant): \ }& +  p^{\text{m}}_t\cdot (1- p^{p}_t) \cdot \left[ m^{np}_t \cdot V^{f,c}_{t+1}  + (1-m^{np}_t)\cdot  V^{f,s}_{t+1}\right] \\
\text{ \footnotesize (met and pregnant): \ } & + p^{\text{m}}_t \cdot p^{\text{p}}_t\cdot   \left[ m^p_t \cdot V^{f,ck}_{t+1}  + (1-m^p_t)\cdot  V^{f,\text{preg}}_{t+1}\right] \Big\} \Bigg\}, \\
\text{ \footnotesize (uncertain abortion access): } &  V^{f,\text{preg}}_{t+1} = p^{\text{access}} \cdot \max\{V^{f,sk}_{t+1},V^{f,s}_{t+1} - \phi_a\} \\
   & \hspace{1cm} + (1-p^{\text{access}})\cdot V^{f,sk}_{t+1},\\
\text{ \footnotesize (budget constraint): \ } & c + a' = R\cdot a  + w , \\
\text{ \footnotesize (wage evolution): \ } & \log w = z + \text{Trend}_t, \ \ z' = z + \varepsilon^{f}.
\end{align*}
\end{frame}

\begin{frame}
\frametitle{Value Function: Couple, No Children}
\framesubtitle{$k_t$ --- give a birth, $d_t$ --- get a divorce, $\tilde{V}$ --- values at divorce}
\begin{align*}
V^{c}_t(a,w_f,w_m,\psi,\theta) = \max\limits_{c_f,c_m,a',k} \Bigg\{ & \theta^f \cdot \frac{c_f^{1-\sigma}}{{1-\sigma}}  + \theta^m \cdot \frac{c_m^{1-\sigma}}{1-\sigma}  + \psi + \beta \cdot \E_t \Big\{   \\
\text{ \footnotesize (stay, no birth): \ } & + (1-d_t) \cdot (1-k_t) \cdot  \left[ \theta^f \cdot  V^{f,c}_{t+1} + \theta^m \cdot V^{m,c}_{t+1}  \right] \\
\text{ \footnotesize (stay, give a birth): \ } &  + (1-d_t) \cdot k_t \cdot  \left[ \theta^f \cdot  V^{f,ck}_{t+1} + \theta^m \cdot V^{m,ck}_{t+1}  \right]  \\
\text{ \footnotesize (divorce): \ } & + d_t  \left[ \theta^f \cdot  \tilde{V}^{f,s}_{t+1} + \theta^m \cdot \tilde{V}^{m,s}_{t+1}  \right]  \Big\} \Bigg\}\\
\text{ \footnotesize (budget constraint, rts): \ } &{\normalsize \left[c_f^{1+\rho} + c_m^{1+\rho}\right]^{\frac1{1+\rho}} + a' = R\cdot a + w_f + w_m}\\
\text{ \footnotesize (match quality evolution): \ } &{\normalsize \psi' = \psi + \varepsilon^{\psi}}\\
\text{ \footnotesize \textbf{(participation constraints)}: \ } &{\normalsize V^{f,c\bullet}_{t+1} (\cdots,\theta') \geq \tilde{V}^{f,s}_{t+1}, \ \ V^{m,c\bullet}_{t+1} (\cdots,\theta') \geq \tilde{V}^{m,s}_{t+1}.}\\
\end{align*}
\end{frame}

\begin{frame}
\frametitle{Value Function: Couple and Child}
\framesubtitle{$d_t$ --- get a divorce, $\tilde{V}$ --- values at divorce}
\begin{align*}
V^{ck}_t(a,w_f,w_m,\psi,\theta) = \max\limits_{c_f,c_m,x,l_f,a'} \Bigg\{ & \theta^f \cdot \frac{c_f^{1-\sigma}}{{1-\sigma}}  + \theta^m \cdot \frac{c_m^{1-\sigma}}{1-\sigma}  + \psi + \\ \text{ \footnotesize (child is a public good): \ }  & \phi + \alpha \cdot \frac{Q^{1-\sigma}}{1-\sigma} +  \beta \cdot \E_t \Big\{   \\
\text{ \footnotesize (stay): \ } & + (1-d_t) \cdot  \left[ \theta^f \cdot  V^{f,ck}_{t+1} + \theta^m \cdot V^{m,ck}_{t+1}  \right] \\
\text{ \footnotesize (divorce): \ } & + d_t \cdot \left[ \theta^f \cdot  \tilde V^{f,sk}_{t+1} + \theta^m \cdot \tilde V^{m,s}_{t+1}  \right]  \Big\} \Bigg\}\\
\text{ \footnotesize (budget constraint): \ } &{\normalsize \left[c_f^{1+\rho} + c_m^{1+\rho}\right]^{\frac1{1+\rho}} + x + a' = R\cdot a + w_f\cdot l_f + w_m}\\
\text{ \footnotesize (child consumption flow): \ } &{\normalsize Q = \left[x^\lambda + \kappa \cdot (1-l_f)^\lambda\right]^{\frac1\lambda}, \ \ \ Q\geq \bar{Q}} \text{ \footnotesize  (subsistence constraint) } \\
\text{ \footnotesize (female skills $\downarrow$ if not working): \ } &{\normalsize \log w_f = z_f + \text{Trend}_{f,t}, \ \ z'_f = z_f + \varepsilon^f - \delta(l^f)}\\
\end{align*}
\end{frame}

\begin{frame}
\frametitle{Renegotiation and Divorce}
\framesubtitle{Are generated by the participation constraints}
\begin{itemize}
\item Bargaining weights $\theta$ adjust to satisfy both participation constraints
\[\Delta \theta^i \text{ reflects Lagrange multiplier on $i$'s participation constraint}: V^{i,c\bullet}_{t+1} \geq \tilde{V}^{i,s}_{t+1}\]
\item If no adjustment is possible, the couple divorces
\item Assets are split evenly upon divorce (follows Voena, 2015)
\item Partners have fixed utility costs $\phi_d$ of divorce
\item If couple with a child divorces wife becomes a single mother
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Child Support}
\framesubtitle{Paid forever, but with a chance to avoid}
\begin{itemize}
\item Imperfect enforcement: probability to be awarded is $p^{\text{award}}$
\begin{itemize}
\item Married couples: 46\%, never married couples: 28\% chances
\end{itemize}
\item If enforced:
\begin{itemize}
\item man loses productivity $z$ so earnings fall by 20\%
\item woman gains productivity $z$ matching this value
\end{itemize}
\item As $z' = z + \varepsilon^z$, all changes are permanent
\end{itemize}
\end{frame}

%\begin{frame}
%\frametitle{Abortions / Birth Control}
%\begin{itemize}
%\item When unplanned pregnancy happens, with probability $p^{\text{access}}$ woman can avoid being single mother
%\item If possible, she pays \emph{fixed utility costs} $u^{a}$ for this
%\item Targets: $\sim40\%$ unplanned pregnancies are aborted + age profile of abortions
%\item Some women \textit{choose} to be single mothers even with abortion access
%\end{itemize}
%\end{frame}

\begin{frame}
\frametitle{Marriage}
\framesubtitle{Introducing Social Stigma}
\begin{itemize}
\item Singles meet singles with random $a$ and $z$ (estimated)
\item Potential couple gets a draw of match quality $\psi$
\item If both partners agree to marry, initial $\theta$ is set by symmetric Nash Bargainig
\item More utility shifters (unexplained things):
\begin{itemize}
\item If unplanned pregnancy happens, partners face {\textbf{social stigma}} $\phi_s$ if refuse to marry
\item If male meets a single mother, he lose $\phi_m$ for having step children
\end{itemize}
\end{itemize}
\begin{center}
{\footnotesize
\begin{tabular}{|l||c|c||c|c||c|}\hline
 & \multicolumn{2}{|c||}{Male gets:} & \multicolumn{2}{|c||}{Female gets:} \\\hline
Situation & Agree & Disagree & Agree & Disagree \\\hline
Regular match & $V^{m,c}_t(\theta)$ & $V^{m,s}_t$ & $V^{f,c}_t(\theta)$ & $V^{f,s}_t$ \\
Unplanned pregnancy & $V^{m,ck}_t(\theta)$ & $V^{m,s}_t - \phi_s$ & $V^{f,ck}_t(\theta)$ & $V^{f,s}_t - \phi_s$ \\
Meeting a single mother & $V^{m,ck}_t(\theta) - \phi_m$ & $V^{m,s}_t$ & $V^{f,ck}_t(\theta)$ & $V^{f,s}_t$ \\\hline
\end{tabular}
}
\end{center}
\end{frame}

\section{Mechanics \& Intuition}
\begin{frame}
\frametitle{Unplanned Pregnancy $\Rightarrow$ Worse Match?}
\framesubtitle{Why people enter knowingly risky marriages?}
\begin{itemize}
\item Women enter the shotgun marriages because:
\begin{itemize}
\item Child costs need to be shared (returns to scale)
\item Finding a new partner is trickier once you have a child
\end{itemize}
\item Men enter the shotgun marriage because:
\begin{itemize}
\item They like their kids and lose access to them otherwise
\end{itemize}
\item Even with a good match, mistimed pregnancy is a risk
\item Scope of these effects depend on exact calibration
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Unplanned Pregnancy $\Rightarrow$ Women Lose a Lot, Men Gain a Little}
\framesubtitle{Nash Bargaining Surplus: $NBS(\theta) = (V^{f,\text{agree}}(\theta) - V^{f,\text{disagree}})\times(V^{m,\text{agree}}(\theta) - V^{m,\text{disagree}})$}
\begin{center}
\begin{tabular}{c c}
\hspace{-1cm} \includegraphics[scale=0.5]{change_upp.pdf} &
\hspace{-0.5cm}\includegraphics[scale=0.5]{surp_upp.pdf} 
\end{tabular}

\end{center}

\end{frame}

\begin{frame}
\frametitle{...Hence Less Picky Partners + Lower Female Decision Power}
\begin{center}
\begin{tabular}{c c}
\hspace{-1cm} \includegraphics[scale=0.5]{barg_reg_match.pdf} & \includegraphics[scale=0.5]{barg_upp.pdf}
\end{tabular}
\end{center}
\end{frame}

\section{Fit and Results}

\begin{frame}
\frametitle{Too Much Heterogeneity}
\begin{itemize}
\item College and non-college people have very distinct patterns
\item They cannot be summarized well with one-dimensional productivity
\item Therefore I focus on two subsamples:
\begin{itemize}
\item High education: bachelor degree or more
\item Low education: high school or less
\end{itemize}
\item Model fits well for both groups, but high education results are more settled
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Targets}
\begin{tabular}{l p{0.5\linewidth}}\hline
\multicolumn{1}{l}{Target} & \multicolumn{1}{l}{Informative about} \\\hline
\footnotesize \% with kids 1--10 years after marriage & \footnotesize  Fertility parameters: $\phi$, $\alpha$ \\
\footnotesize \% MF females in population at 23--35 & \footnotesize  $\phi$, $\alpha$ + meeting probabilities \\
\footnotesize \% KF females in population at 23--35 &\footnotesize   meeting + unplanned preg probabilities\\
\footnotesize \% never married with and without kids at 23--35 & \footnotesize  meeting + unplanned preg probabilities + $\phi_r$\\
\footnotesize \% divorced with and without kids at 23--35 & \footnotesize  distribution of $\psi$ + fertility parameters + $\phi_r$ \\
\footnotesize child expenditures share, \% in LF with kids & \footnotesize  household technology: $\kappa$, $\alpha$ and $\bar{Q}$ \\
\footnotesize \% divorced 1--10 years after marriage if KF or MF & \footnotesize  social stigma, $\bar{Q}$ and more ... \\
\footnotesize \% unplanned pregnancies aborted + age profile & \footnotesize  abortion costs and access \\\hline
\end{tabular}
\end{frame}

\begin{frame}
\frametitle{Data}
\framesubtitle{High school in red, college in black}
\begin{center}
\begin{tabular}{c c}
\hspace{-0.5cm} \includegraphics[scale=0.33]{div_kfmf_dc.pdf} & \includegraphics[scale=0.33]{popshares_dc.pdf} \\
\hspace{-0.5cm} \includegraphics[scale=0.33]{evermar_dc.pdf} & \includegraphics[scale=0.33]{everkids_dc.pdf} \\
\end{tabular}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Estimates}
\framesubtitle{Simulated Method of Moments (SMM) with optimization routine from Arnoud et al, 2019; 17 parameters}
\begin{tabular}{l c c c}\hline
& High Education & Low Education & Source \\\hline
\multicolumn{4}{c}{Utility from kids: $\phi + \alpha \cdot \frac{Q^{1-\xi}}{1-\xi}$, \ $Q = [x^{\lambda} + \kappa\cdot(1-l_f)^{\lambda}]^{\frac1{\lambda}}$. $Q\geq \bar{Q}$}\\\hline
$\phi$, $\alpha$, $\kappa$  \footnotesize (utility params) &\footnotesize  $(1.39, 0.36, 0.78)$ &\footnotesize $(1.80,0.45, 1.16)$ & SMM \\
$\bar{Q}$ \footnotesize (subsistence constraint) & 1.7 & 0.70 & SMM\\
\multicolumn{4}{c}{Meeting probability $p^{\text{meet}}$, pregnancy probability $p^{\text{preg}}$: quadratic polynomials}\\\hline
$p^{\text{meet}}$  \footnotesize at (21,30,40) & \footnotesize $(0.11, 0.21, 0.34)$ & \footnotesize  $(0.44, 0.10, 0.65)$ & SMM\\
$p^{\text{preg}}$ \footnotesize  at (21,28,35) & \footnotesize  $(0.01, 0.04, 0.01)$ & \footnotesize  $(0.09, 0.07, -0.05)$ & SMM \\\hline\hline
\multicolumn{4}{c}{Additive utility shifters}\\\hline
$\sigma(\varepsilon^\psi), \sigma(\psi_0)$   \footnotesize (match quality distribution) & $(0.10,0.14)$ & $(0.88,4.8)$ & SMM \\
$\phi_s$ \footnotesize  (social stigma) & 4.21 & 5.31 & SMM\\
$\phi_m$ \footnotesize   (single mom remarriage loss) & 43.3 & 79.6 & SMM \\\hline
$\phi_d$ \footnotesize   (divorce costs) & 5.3 & 1.2 & SMM \\\hline
$(p^{\text{access}}, \phi_a)$ \footnotesize   (abortion parameters) & $(0.22,37.5)$  & $(0.18,5.51) $& SMM \\\hline
\end{tabular}
\end{frame}

\begin{frame}
\frametitle{Things Fit Well --- Part 1}
\framesubtitle{High Education Sample}
\begin{center}

\begin{tabular}{c c}
\hspace{-1cm}\includegraphics[scale=0.5]{evermar.pdf} &\hspace{-0.5cm} \includegraphics[scale=0.5]{popshares.pdf} \\
\end{tabular}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Things Fit Well --- Part 2}
\framesubtitle{High Education Sample}
\begin{center}
\begin{tabular}{c c}
\hspace{-1cm}\includegraphics[scale=0.5]{div_kfmf.pdf}  & \hspace{-0.5cm} \includegraphics[scale=0.5]{everkids.pdf} 
\end{tabular}
\end{center}
\end{frame}

\begin{frame}[label=counterfactuals-abortions]
\frametitle{What Explains The Difference?}
\framesubtitle{Removing Social Stigma (1/4) + Fertility Control (3/4).}
\begin{center}
\begin{tabular}{c c}
\hspace{-1cm}\includegraphics[scale=0.5]{div_kfmf_equalized.pdf}  & \hspace{-0.5cm} \includegraphics[scale=0.5]{relshares_equalized.pdf} 
\end{tabular}
\end{center}
Welfare: men gain $\approx \$35k$, women gain $\approx \$12k$, \hyperlink{welfare-abortions}{\beamerbutton{details}} 
\end{frame}

\begin{frame}[label=counterfactuals-child-support]
\frametitle{Removing Existing Child Support --- Part 1}
\framesubtitle{Less single mothers, less divorces, more abortions}
No child support instead of current (partially enforced 20\%)
\begin{center}
\begin{tabular}{c c}
\hspace{-1cm}\includegraphics[scale=0.5]{div_kfmf_child_support.pdf}  & \hspace{-0.5cm} \includegraphics[scale=0.5]{sm_child_support.pdf} 
\end{tabular}
\end{center}
Just removing the social stigma \hyperlink{counterfactuals-social-stigma}{\beamerbutton{here}}

Welfare: men gain $\approx \$3k$, women lose $\approx \$1k$ \hyperlink{welfare-child-support}{\beamerbutton{details}} 
\end{frame}

\begin{frame}[label=counterfactuals-child-support]
\frametitle{Removing Existing Child Support --- Part 2}
\framesubtitle{Hardly Affects Marriage in Any Way}
Can you notice the differences?
\begin{center}
\begin{tabular}{c c}
\hspace{-1cm}\includegraphics[scale=0.5]{evermar_cs.pdf}  & \hspace{-0.5cm} \includegraphics[scale=0.5]{popshares_cs.pdf} 
\end{tabular}
\end{center}
\end{frame}

\begin{frame}[label=cf]
\frametitle{More Counterfactuals}
\begin{itemize}
\item No divorce costs \hyperlink{no_divorce_costs}{\beamerbutton{pictures}}
\begin{itemize}
\item More divorces, less kids
\item But less divorces conditional on having kids!
\item Welfare: females lose, males gain, gain on average
\end{itemize}
%\item Removing gender pay gap: same trend + variance for men and women \hyperlink{pay_gap}{\beamerbutton{pictures}} 
\item No unplanned pregnancies at all 
\begin{itemize}
\item Remaining KF (step-child-couples) and MF become very similar
\item Men gain $\sim \$25k$, women gain $\sim \$13k$ 
\item Around 1/10 less divorces
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}[label=cf]
\frametitle{Conclusions}
\begin{itemize}
\item Model mechanisms can rationalize creation and differences in divorce
\item Child support is welfare-decreasing and does not get people together
\item Controlling fertility helps everyones
\item \textbf{Love $>>$ Money}
%\item Nothing beats the birth control in terms of welfare
%\item Child support does not get people together and creates more single mothers
%\item Empirically, more control over fertility $\Rightarrow$ less KF, lower differences in performance
\end{itemize}
\end{frame}

\backupbegin

%\begin{frame}[label=remarriage_penalty]
%\frametitle{Removing Remarriage Penalty For Single Mothers}
%\framesubtitle{Divorced in 10 years if KF: 13.5\% $\to$ 12.4\%. Divorced in cross-section if KF: 9.3\% $\to$ 8.9\% + More Kids}
%\begin{center}
%\begin{tabular}{c c}
%\hspace{-1cm} \includegraphics[scale=0.5]{div_kfmf_no_remar.pdf}  &  \includegraphics[scale=0.5]{ever_kids_no_remar.pdf} 
%\end{tabular}
%\end{center}
%\hyperlink{cf}{\beamerbutton{back}} 
%\end{frame}

\begin{frame}[label=no_divorce_costs]
\frametitle{No Divorce Costs}
\framesubtitle{Less kids, more shotgun marriages (relatively), less divorces if have kids}
\begin{center}
\begin{tabular}{c c}
\hspace{-1cm} \includegraphics[scale=0.5]{popshares_nodivcosts.pdf}  &  \includegraphics[scale=0.5]{div_kfmf_nodivcosts.pdf} 
\end{tabular}
\end{center}
\hyperlink{cf}{\beamerbutton{back}} 
\end{frame}

\begin{frame}[label=pay_gap]
\frametitle{Removing Pay Gap}
\framesubtitle{Divorced in 10 years if KF: 13.5\% $\to$ 12.6\%. Divorced in cross-section if KF: 9.3\% $\to$ 8.8\% + Fewer Kids}
\begin{center}
\begin{tabular}{c c}
\hspace{-1cm} \includegraphics[scale=0.5]{eq_pay.pdf}  &  \includegraphics[scale=0.5]{everkids_eq_pay.pdf} 
\end{tabular}
\end{center}
\hyperlink{cf}{\beamerbutton{back}} 
\end{frame}

\begin{frame}[plain,label=extra-restrictions]
\frametitle{Extra Restrictions}
\begin{itemize}
\item I do not classify as KF or MF people with $\Delta T < -5$ or $\Delta T > 5$
\item I exclude small share of people with marital statuses ``separated'', ``widowed'' and ``spouse absent''
\item This does not affect the results much but helps to ensure consistency
\end{itemize}
\hyperlink{kfandmf}{\beamerbutton{back to KF and MF}} 
\end{frame}

\begin{frame}[plain,label=dt_graph_educ]
\frametitle{Kids First Women Are Divorced More Often: By Education}

\begin{tabular}{c c}
\hspace{-0.5cm}\includegraphics[scale=0.5]{div_5y_by_dt_col.pdf} & \hspace{0.5cm}\includegraphics[scale=0.5]{div_5y_by_dt_hs.pdf}
\end{tabular}
\hyperlink{dt_graph}{\beamerbutton{back to the main graph}} 
\end{frame}

\begin{frame}[label=counterfactuals-social-stigma]
\frametitle{Removing Social Stigma}
\framesubtitle{Less single mothers, less divorces, more abortions}
No child support instead of current (partially enforced 20\%)
\begin{center}
\begin{tabular}{c c}
\hspace{-1cm}\includegraphics[scale=0.5]{div_kfmf_ss.pdf}  & \hspace{-0.5cm} \includegraphics[scale=0.5]{sm_among_mothers_ss.pdf} 
\end{tabular}
\end{center}
\hyperlink{counterfactuals-child-support}{\beamerbutton{back}} 
\end{frame}

\begin{frame}[label=welfare-abortions]
\frametitle{What Explains The Difference? Welfare Implications}
\framesubtitle{Removing Social Stigma (1/4) + Fertility Control (3/4)}
\begin{center}
\begin{tabular}{c c}
\hspace{-1cm}\includegraphics[scale=0.5]{welfare_female_equalized.pdf}  & \hspace{-0.5cm} \includegraphics[scale=0.5]{welfare_male_equalized.pdf} 
\end{tabular}
\end{center}
\hyperlink{counterfactuals-abortions}{\beamerbutton{back}} \hyperlink{welfare-methodology}{\beamerbutton{methodology}} 
\end{frame}

\begin{frame}[label=welfare-child-support]
\frametitle{Child Support Removal}
\framesubtitle{More abortions but not much changes...}
\begin{center}
\begin{tabular}{c c}
\hspace{-1cm}\includegraphics[scale=0.5]{welfare_female_child_support.pdf}  & \hspace{-0.5cm} \includegraphics[scale=0.5]{welfare_male_child_support.pdf} 
\end{tabular}
\end{center}
\hyperlink{counterfactuals-child-support}{\beamerbutton{back}} \hyperlink{welfare-methodology}{\beamerbutton{methodology}} 
\end{frame}

\begin{frame}[label=welfare-methodology]
\frametitle{Definitions}
\framesubtitle{Equivalent Variation (EV) Style}
Value of any policy change can be computed in terms of equivalent amount of assets.

If the change is preferred:
\[EV = a^*_t(w) \text{ such that } V_{t}^{f,s,\text{no change}}(a=a^*,w) = V_{t}^{f,s,\text{with change}}(a=0,w)\]
Otherwise:
\[EV = -a^*_t(w) \text{ such that } V_{t}^{f,s,\text{with change}}(a=a^*,w) = V_{t}^{f,s,\text{no change}}(a=0,w)\]

Analogously for $m$, a bit trickier for couples.

\hyperlink{welfare-abortions}{\beamerbutton{back to abortions}} \hyperlink{welfare-child-support}{\beamerbutton{back to child support}}

\end{frame}

\backupend
\end{document}

### 2020-10-02 slides-march-20/slides-new.tex
Current stage

\documentclass{beamer}
\usepackage[english]{babel}
\usepackage{amsmath,amsfonts}
\usepackage{multicol}

\usepackage{IEEEtrantools}
\usepackage{multirow}
% beamer setup

\usetheme{default}
\usecolortheme{seahorse}
%\usecolortheme{dove}

\setbeamertemplate{navigation symbols}{}

\usepackage{fixmath}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows}
\usetikzlibrary{positioning}
\tikzstyle{block} = [rectangle, draw, rounded corners]
\tikzstyle{line} = [draw, -latex']

\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator*{\argmax}{argmax}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\I}{\mathbb{I}}

\AtBeginSection[]{
  \begin{frame}[plain]
  \addtocounter{framenumber}{-1}
  \vfill
  \centering
  %\begin{beamercolorbox}[sep=8pt,center,shadow=true,rounded=true]{title}
    %\usebeamerfont{title}
    \Huge{\insertsectionhead\par}%
  %\end{beamercolorbox}
  \vfill
  \end{frame}
}

\newcommand{\backupbegin}{
   \newcounter{framenumberappendix}
   \setcounter{framenumberappendix}{\value{framenumber}}
}
\newcommand{\backupend}{
   \addtocounter{framenumberappendix}{-\value{framenumber}}
   \addtocounter{framenumber}{\value{framenumberappendix}} 
}

\title{The Economics of Shotgun Marriage}
\subtitle{and Household Bargaining}
\author{Egor Kozlov}

\institute{
  Department of Economics\\
  Northwestern University}

\setbeamertemplate{footline}[frame number]

%  \usepackage{pgf}
%\logo{\pgfputat{\pgfxy(0,0)}{\pgfbox[right,base]{\footnotesize{\insertframenumber\,/\,\inserttotalframenumber}}}}
%\newcommand{\nologo}{\setbeamertemplate{logo}{}}

\begin{document}

% There is a huge social issue of single mothers
% You typically associate it with unplanned teenage pregnancies of low-income and low-education people
% But looking at just single mothers or young couples is an understatement
% Unplanned pregnancy expresses itself even for married people: not all marriages are created equal
% In later ages unplanned pregnancies force couple to marry
% 
% Even for college 
% In more adult age this is pronounced by shotgun marriage
% Chetty stuff: share of single mothers is the best predictor of inequality and so on

\begin{frame}[plain]
\addtocounter{framenumber}{-1}
\date{\scriptsize}
\titlepage
\end{frame}

\begin{frame}
\frametitle{Unplanned Pregnancy $\Rightarrow$ Shotgun Marriage}
\begin{itemize}
\item Many people marry shortly after a baby appears
\item These marriages perform worse in the data 
\begin{itemize}
\item A lot worse for college grads, a bit worse for high school grads
\item $\sim10\%$ marry while being pregnant
\item Not just a teenage thing, peaks around 30
\end{itemize}
\item Understanding why helps to understand the value of marriage
\item Questions:
\begin{itemize}
\item What is the aggregate role of unplanned pregnancies?
\item What causes people to enter such marriages?
\item Why they perform worse?
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}[label=pic0]
\frametitle{Divorced By Timings}
\begin{center}
\includegraphics[width=0.85\linewidth]{share_div_march5.eps}
\end{center}
\end{frame}

\begin{frame}[label=pic0]
\frametitle{Hazards}
\begin{center}
\includegraphics[width=0.85\linewidth]{haz_kf_march5.eps}
\end{center}
\end{frame}

\begin{frame}[label=pic0]
\frametitle{Shares}
\begin{center}
\includegraphics[width=0.85\linewidth]{share_kj_march5.eps}
\end{center}
\end{frame}

\begin{frame}[plain]
\frametitle{What I Do}
\begin{itemize}
\item Quantify empirical difference of marriage performance by the timing of childbirth
\item Estimate a structural model matching this difference
\item Use the model to study:
\begin{itemize}
\item aggregate impact of unplanned childbirths 
\item importance of fertility for marriage
\item difference between education groups
\item policy counterfactuals
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}[plain]
\frametitle{Preview of Results}

\begin{enumerate}

\item Data: 
{\small
\begin{center}
\begin{tabular}{|c|c|c|c|}\hline
 & \multicolumn{2}{|c|}{\textit{Share of divorced among:}} & \\
& Marriage First & Kids First & Ratio \\\hline
College &  5.3\% &  14.7\% & \textbf{2.77} \\
High School &  13.9\% & 17.2\% &  \textbf{1.24}  \\\hline
\end{tabular}
\end{center}
}

\item Quantifying:
\begin{itemize}
\item Unplanned pregnancy changes outside options in family bargaining
\item Effect is present generically, its size identifies amount of distress
\item To estimate, assume optimal fertility timing after marriage
\end{itemize}

\item Exploring:
\begin{itemize}
\item Unplanned pregnancies account for $\sim50\%$ of early divorces
\item Marriage market (meeting probabilities) mostly drive the college -- high school difference
\end{itemize}

\end{enumerate}
\end{frame}

%
%\begin{frame}
%\frametitle{[Distribution of age at shotgun marriage in my def]}
%
%\end{frame}

%
%\begin{frame}
%\frametitle{Main Results}
%\begin{itemize}
%\item Empirical:
%\begin{itemize}
%\item Shotgun marriage $\Rightarrow$ divorced $\sim1.8$ more often
%\item Among college graduates: $\sim3$ times more often
%\item Differences are present after controlling for lots of observables
%\item Present is many data sets: ACS, SIPP, NSFG, NSFH
%\end{itemize}
%\item From the model:
%\begin{itemize}
%\item Disentangling two main mechanisms:
%\begin{itemize}
%\item Shotgun marriage couples are different in characteristics
%\item Unexpected pregnancy creates large unplanned costs
%\end{itemize}
%\item Expensive eraly-life childcare explains a lot
%\item Higher marginal costs of childcare for skilled women
%\item $\sim10\%$ of couples marry because of unplanned pregnancy
%\item These couples are the least stable
%\end{itemize}
%\item Missing pieces (TBD):
%\begin{itemize}
%\item Quantifying implications
%\item Quantifying changes over time
%\end{itemize}
%\end{itemize}
%\end{frame}

\begin{frame}
\frametitle{General Importance}
\begin{itemize}
\item Unintended pregnancies $\Rightarrow$ people to pick less favorable partners, ``inefficient'' marriages from ex-ante prospective
\item Child development:
\begin{itemize}
\item Single mothers or blended families $\Rightarrow$ inferior child's outcomes
\item Unstable marriages may be even worse but are often overlooked
\end{itemize}
\item Major impact on women, in particular on:
\begin{itemize}
\item Labor market: loss of skills
\item Divorces + children worsen re-marriage prospective
\item Total fertility and marriage rates
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Literature}
Shotgun marriages are still a thing:
\begin{itemize}
\item Gibson-Davis et al (2016): 10\% midpregnancy marriages
%\item Akerlof et al (1996)
\end{itemize}
Lifecycle models:
\begin{enumerate}
\item Endogenous bargaining, marriage, divorce \& assets:
\begin{itemize}
\item Voena (2015) --- just divorce, Low et al (2018) --- both
\item Mazzocco (2007) --- seminal paper
\end{itemize}
%\begin{center}
%\textbf{Contribution:} endogenous + exogenous fertility
%\end{center}
\item Fertility:
\begin{itemize}
\item Planned: Sommer (2016)
\item Unplanned: Ejrnæs, Jørgensen (2018)
\end{itemize}
%\begin{center}
%\textbf{Contribution:} bargaining, marriage and divorce
%\end{center}
\item Child investment and marriage/divorce:
\begin{itemize}
\item Lafortune, Low (2018)
\end{itemize}
%\begin{center}
%\textbf{Contribution:} proper lifecycle, fertility
%\end{center}
\end{enumerate}
%
%\textit{Other relevant evidence:}
%\begin{itemize}
%\item Divorce rates by income: Boertien, Härkönen (2014)
%\item Child expenditures by income: Schneider et al (2014)
%\end{itemize}
%\item Computations: Judd et al (2014) (thank Luigi Bocola)

\end{frame}

\begin{frame}
\frametitle{Updates Since November}
\begin{itemize}
\item Brand new codes in Python \& cuda
\item Remarriage of single mothers
\item Skills depreciation because of time to childcare
\item No unplanned pregnancies in marriage: restrictive but cleaner identification
\item Two separate estimations: college grads versus non-college grads
\item Decomposition exercises as a goal
\end{itemize}
\end{frame}

\section{Mechanism}

\begin{frame}
\frametitle{Framework}
\begin{itemize}
\item Bargaining model with outside option of being single. Together if:
\[V^{\text{fem},\text{married}} \geq V^{\text{fem},\text{single}} \ \& \ V^{\text{male},\text{married}} \geq V^{\text{male},\text{single}}\]
\item Make choices and have some margins of adjustment
\item Heterogeneous relationship quality (love shock), the best predictor of divorce
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Unplanned Pregnancy Shock}
\begin{itemize}
\item Married couples can choose their fertility
\item However, sometimes a child arrives by a random shock $\Rightarrow$ distress
\item Size depends on how indifferent are people between different fertility timings
\item College people are more sensitive?
\item What about marriage/divorce?
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Asymmetric Response to Shock}
People stay together if
\[V^{\text{fem},\text{married}} \geq V^{\text{fem},\text{single}} \ \& \ V^{\text{male},\text{married}} \geq V^{\text{male},\text{single}}\]
\vspace{-1em}
\begin{itemize}
\item Bargaining model: if stress is symmetric (all $V\downarrow$ the same way) neither more marriages nor more divorces happens
\item $\Rightarrow$ females lose more, males gain more
\item If marriages differ in quality, more low-quality marriages are created
\end{itemize}

Why do people enter low-quality marriages?
\begin{itemize}
\item Women:
\begin{itemize}
\item splitting of childcare costs
\item worsening of future marital prospectives
\end{itemize}
\item Men:
\begin{itemize}
\item loss of access to their children otherwise
\end{itemize}
\end{itemize}
\end{frame}

\section{Model}

\begin{frame}
\frametitle{Overview}
\begin{itemize}
\item Lifecycle model
\item Assets ($a$) and productivity shocks ($z^f$, $z^m$)
\item Labor supply for females with kids
\item Endogenous marriage and divorce (limited commitment)
\item Reduced form marriage market:
\[p^{\text{meet}}(\text{Age}) \text{ --- meeting a random partner with similar $a$ and $z$}\]
\item After marriage: endogenous fertility choice
\item Before marriage can have an unplanned child
\[p^{\text{unplanned}}(\text{Age}) \text{ --- arrival of unplanned kids}\]
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Modeling Shotgun Marriage}
\begin{center}
\vspace{-0.75cm}
\begin{tikzpicture}[every text node part/.style={align=center}, scale=0.75]
   % Place nodes
   \node [block] (1) {\large \textbf{People}  \\[-0.5ex]  \large  \textbf{Randomly Meet}};
   \node [block, below left = 1.5 cm and 0.0 cm of 1] (2) {\textit{Nash} \\[-0.5ex] \textit{Bargaining}};   
   \node [block, below right = 1.5 cm and 0.0 cm of 1] (3) {\textit{Nash} \\[-0.5ex] \textit{Bargaining}};

   \node [block, below = 1.5 cm of 2] (5) {\textbf{Kids First} \\[-0.5ex] \textbf{Couple $(ck)$}};
   \node [block, below = 0.8 cm of 3] (7) {No Kids \\[-0.5ex] Couple $(c)$};
   \node [block, below = 0.8 cm of 7] (8) {\textbf{Mariage First} \\[-0.5ex] \textbf{Couple  $(ck)$}};
   \draw[->] (1) to [bend left = 20] node [left] {\scriptsize Unplanned \\[-0.75ex] \scriptsize pregnancy} (2);
   \draw[->] (1) to [bend right = 20] node [right] {\scriptsize No \\[-0.75ex] \scriptsize pregnancy} (3);
   \draw[->] (2) to node [left] {\scriptsize Agree} node [right] {\scriptsize \textbf{Shotgun} \\[-0.75ex]  \scriptsize \textbf{marriage}} (5);
   \draw[->] (3) to node [right] {\scriptsize Agree} node [left]  {\scriptsize \textbf{Regular} \\[-0.75ex]  \scriptsize \textbf{marriage}} (7);
   \draw[->] (7) to node [right] {\scriptsize Kids by choice} (8);

\pause

    \node [block, above left   = 1.0 cm and -0.75 cm of 2] (4) {\,\, Single male $(m,s)$\,\,};
    \node [block, above left   = 1.6 cm and -0.75 cm of 2] (4-sm) {Single mother $(f,sk)$};
	\node [block, above right   = 1.0 cm and -0.75 cm of 3] (6) {\,\, Single male $(m,s)$\,\,};
    \node [block, above right   = 1.6 cm and -0.75 cm of 3] (6-f) {\,Single female $(f,s)$\,};   
    \draw[->] (2.180) to [bend left = 30] node [left] {\scriptsize Disagree} (4.270);
    \draw[->] (3.0) to [bend right = 30] node [right] {\scriptsize Disagree} (6.270);
     \draw[->] (6.180) to [bend left = 30] node [above] {\scriptsize Back}  (1.-12);
   \draw[->] (6-f.180) to [bend right = 30] node [above] {\scriptsize Back} (1.12);
   \draw[->,thick,dashed] (4-sm.90) to [bend left = 30] node [above] {\scriptsize Worse marriage prospectives}  (1.168);
   \draw[->] (4.0) to [bend right = 30] node [above] {\scriptsize Back} (1.192);

\end{tikzpicture}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Modeling Shotgun Marriage}
\begin{center}
\vspace{-0.4cm}
\begin{tikzpicture}[every text node part/.style={align=center}, scale=0.75]
   % Place nodes
   \node [block] (1) {\large \textbf{People}  \\[-0.5ex]  \large  \textbf{Randomly Meet}};
   \node [block, below left = 1.5 cm and 0.0 cm of 1] (2) {\textit{Nash} \\[-0.5ex] \textit{Bargaining}};   
   \node [block, below right = 1.5 cm and 0.0 cm of 1] (3) {\textit{Nash} \\[-0.5ex] \textit{Bargaining}};

   \node [block, below = 1.5 cm of 2] (5) {\textbf{Kids First} \\[-0.5ex] \textbf{Couple} ($ck$)};
   \node [block, below = 0.8 cm of 3] (7) {No Kids \\[-0.5ex] Couple ($c$)};
   \node [block, below = 0.8 cm of 7] (8) {\textbf{Mariage First} \\[-0.5ex] \textbf{Couple} ($ck$)};
   \draw[->] (1) to [bend left = 20] node [left] {\scriptsize Unplanned \\[-0.75ex] \scriptsize pregnancy} (2);
   \draw[->] (1) to [bend right = 20] node [right] {\scriptsize No \\[-0.75ex] \scriptsize pregnancy} (3);
   \draw[->] (2) to node [left] {\scriptsize Agree} node [right] {\scriptsize \textbf{Shotgun} \\[-0.75ex]  \scriptsize \textbf{marriage}} (5);
   \draw[->] (3) to node [right] {\scriptsize Agree} node [left]  {\scriptsize \textbf{Regular} \\[-0.75ex]  \scriptsize \textbf{marriage}} (7);
   \draw[->] (7) to node [right] {\scriptsize Kids by choice} (8);   
    \node [block, above left   = 0.8 cm and -0.75 cm of 2] (4) {\,Single male \\[-0.5ex] ($m,s$)\,\,};
    \node [block, above left   = 1.8 cm and -0.75 cm of 2] (4-sm) {\,Single mom \\[-0.5ex]  ($f,sk$)};
	\node [block, above right   = 0.8 cm and -0.75 cm of 3] (6) {\,\,Single male\,\,\\[-0.5ex] ($m,s$)};
    \node [block, above right   = 1.8 cm and -0.75 cm of 3] (6-f) {Single female\\[-0.5ex] ($f,s$)\,};   
    \draw[->] (2.180) to [bend left = 30] node [left] {\scriptsize Disagree} (4.270);
    \draw[->] (3.0) to [bend right = 30] node [right] {\scriptsize Disagree} (6.270);
     \draw[->] (6.180) to [bend left = 30] node [above] {\scriptsize Back}  (1.-12);
   \draw[->] (6-f.180) to [bend right = 30] node [above] {\scriptsize Back} (1.12);
   \draw[->,thick,dashed] (4-sm.90) to [bend left = 30] node [above] {\scriptsize Worse marriage prospectives}  (1.168);
   \draw[->] (4.0) to [bend right = 30] node [above] {\scriptsize Back} (1.192);

\end{tikzpicture}
\end{center}
\end{frame}

\begin{frame}[label=utilities]
\frametitle{Utility Functions}
\begin{itemize}
\item Couple, $\theta$ --- Pareto weight, $\psi$ --- common love shock:
\[U = \theta^f \cdot u(c_f) + \theta^m \cdot u(c_m) + \psi,\]
\item Couple + child, $Q$ --- child quality (public good):
\[U = \theta^f \cdot {u}(c_f) + \theta^m \cdot {u}(c_m) + \psi + u^{\text{child}}(Q),\]
\item Child quality: $Q = F(x,1-l^f)$, $x$ --- money, $l^f$ --- female labor supply
\item Budget constraint ($C$ reflects returns to scale, $a$ --- assets):
 \[C(c_f,c_m) + a' + x = R\cdot a + l^f \cdot W^f_t(z_f) + W^m_t(z_m),\]
\end{itemize}
\hyperlink{value-functions}{\beamerbutton{Value Functions}}
\end{frame}

\begin{frame}
\frametitle{Children In The Model}
\begin{itemize}
\item People like having children (public good):
\[u^{\text{child}} = \phi + \alpha \cdot \frac{Q^{1-\sigma}}{1-\sigma}.\]
\item $Q$ is produced out of money and female time
\item $Q = \left[ x^{\lambda} + \kappa \cdot (1-l^f)^{\lambda}\right]^{\frac1\lambda}$
\item \textit{Subsistence constraint:} $Q \geq \bar{Q}$
\item $l^f\downarrow$ $\Rightarrow$ loss in current income + loss in productivity
\item Productivity process: $z^f_t = z^f_{t-1} + \varepsilon^f_{t}  - $ {\boldmath $f(l)$}.
\item In calibration $l \in \{0.25,0.5,0.75,0.9\}$
\item Corresponding to depreciation: $f(l) \in \{0.1,0.05,0.0,0.0\}$.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Marriage}
\begin{itemize}
\item People (f \& m) meet, get a draw of love shock $\psi$ and agree to marry if exists a Pareto weight $\theta$ such that
\[V^{f,\text{yes}}(\underset{\scriptsize +}{\theta\vphantom{\psi}},\underset{\scriptsize +}{\psi}) \geq V^{f,\text{no}}, \ \ \ \ V^{m,\text{yes}}(\underset{\scriptsize  -}{\theta\vphantom{\psi}},\underset{\scriptsize +}{\psi}) \geq V^{m,\text{no}}.\]
\item \textbf{Mechanism:} unplanned pregnancy 
\begin{enumerate}
\item[(1)] lowers current $V^{f,\text{no}}$: hard to be a single mother
\item[(2)] increases future $V^{m,\text{yes}}$ and $V^{f,\text{yes}}$: good to have children
\end{enumerate} 
$(1) + (2)$ $\Rightarrow$ the couple agrees to marry with lower $\psi$ and $\theta$.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Divorce}
\begin{itemize}
\item Shocks to $\psi$, $z^f$ and $z^m$
\item People cannot commit not to divorce
\item $\theta$ may be renegotiated, stays constant if not
\item Divorce happens if no $\theta$ exist such that
\[V^{f,\text{together}}(\theta,\psi) \geq V^{f,\text{divorce}}, \ \ \ \ V^{m,\text{together}}(\theta,\psi) \geq V^{m,\text{divorce}}.\]
\item Couples with lower $\psi$ or more unequal $z$ divorce more!
\end{itemize}
\end{frame}

%
%\begin{frame}
%\frametitle{Individual Values}
%
%\end{frame}

\begin{frame}
\frametitle{Discrete Decisions}
\begin{itemize}
\item Couple's problem defines $V^{c}_t$ (couple's value) and $(V^{f,c}_t,V^{m,c})$ (individual values).
\item In the last period $V^{c}_t = \theta^f u(c^f) + \theta^m u(c^m)$, $V^{i,c} = u(c^i)$.
\item Collective model (what I use): have a kid if 
\[ \theta^f u(c^{f,k}) + \theta^m u(c^{m,k})  + u^{k} \geq \theta^f u(c^{f,nk}) + \theta^m u(c^{m,nk}).\]
\item In fact this is ad hoc, and there is another version: have a kid if \textbf{both}
\[ u(c^{f,k}) + u^{k} \geq u(c^{f,nk})  \ \text{ and } \ u(c^{m,k}) + u^{k} \geq u(c^{m,nk}).\]
\item \textit{Decision making is not totally defined...}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Discrete Decisions + Rebargaining}
\begin{itemize}
\item Decision also depends on $\theta$
\item $k_\theta = \I(V^{c,k}_\theta \geq V^{c,nk}_\theta)$ (or the 2nd protocol)
\item Set of acceptable $\theta$ are
\[ \Theta_f =  \{ \theta : V^{f,ck}_\theta \cdot k_\theta  + V^{f,nk}_\theta \cdot  (1-k_\theta ) \geq V^{f,\text{divorce}} \}\]
\[ \Theta_m = \{ \theta : V^{m,ck}_\theta  \cdot  k_\theta+ V^{m,nk}_\theta  \cdot  (1-k_\theta ) \geq V^{m,\text{divorce}} \}\]
\[ \Theta_c = \Theta_f \cap \Theta_m.\]
\item Pick new $\theta$ if $\Theta_c$ is non-empty
\item But in this case $\Theta_c$ may even be disjoint
\item In this case use Marcet, Marimon (2019) idea and pick the closest $\theta$ to the original value
\item But things can become quirky...
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Remarriage}
\begin{itemize}
\item If couple with a child divorces, she becomes a single mother
\item Single mothers still meet single males
\item If partners agree to marry they become a regular couple with a child
\item Males have a one-time utility penalty parameter for marrying a single mother. This affects bargaining:
\[NBS = \left[ V^{m,MK} - {\text{\boldmath$\phi_m$}} - V^{m,S}\right] \times  \left[ V^{f,MK} - V^{f,S}\right].\]
\item Rationalizes decrease in marital prospectives for females with children
\item Estimated by matching share of remarried women with vs without children
\end{itemize}
\end{frame}

\section{Fit and Results}
\begin{frame}
\frametitle{Strategy}
\begin{itemize}

\item Two calibrations: college and high school
\item I use Simulated Method of Moments to calibrate $\sim 11$ parameters:
\begin{itemize}
\item Preferences
\item Probabilities of transitions
\item Child quality technology
\end{itemize}
\item Income paths to match ACS (trend and variance)
\item The target moments are taken from ACS cross-section
\item Childcare expenditure share from CEX
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{College vs High School}
\begin{center}
\begin{tabular}{|l|c|c|c|}
& \multicolumn{2}{|c|}{Share of divorced, \%} & \\
& MF & KF & Ratio \\
Baseline: college & 5.4 & 14.0 &   2.6  \\
+ income process of HS & 3.9 & 15.3 & 4.0 \\
+ pregnancy probabilities of HS & 5.5 & 14.4 & 2.6 \\
\textbf{+ meeting probabilities of HS} & \textbf{8.9} & \textbf{12.6}  & \textbf{1.4} \\
+ love shock distribution of HS & 5.1 & 14.5 & 2.9 \\
+ single mother prospective of HS & 5.0 & 15.1 &  3.0 \\
Comparison: high school &  11.8 & 16.6 & 1.4\\
\end{tabular}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Targeted Moments}
\begin{tabular}{ l r r}\hline
Target & Model & Data \\\hline
\footnotesize\textit{Never married by (25,35,40), \% }& $(61,25,19)$ &  $(68,26,19)$ \\
\footnotesize \textit{One marriage by 40, \%}  & $70$ & $65$ \\\hline
\footnotesize \textit{No children by (25,35,40), \%}  & $(65,27,24)$ & $(67,28,26)$ \\
\footnotesize \textit{... by 30 (low-$z$,high-$z$), \%}  & $(52,57)$ & $(44,58)$ \\\hline
\footnotesize \textit{Divorced by (25,35,40), \%}  & $(5,11,16)$ & $(11,16,19)$ \\
\footnotesize \textit{... by 30 (low-$z$,high-$z$), \%}  & $(17,9)$ & $(21,8)$ \\\hline
\footnotesize \textit{Divorced by (30,35,40), \%}  & $(5,11,16)$ & $(11,16,19)$ \\
\footnotesize \textit{... by 40 (KF,MF), \%}  & $(25,15)$ & $(24,13)$ \\\hline
\footnotesize \textit{Money for childcare ($\%$ tot expend)} & $43.4$ & 40.0 \\
\footnotesize \textit{Time spent low-$z$/high-$z$ (ratio)} & 1.11 & 1.19 \\
\footnotesize \textit{Money spent low-$z$/high-$z$ (ratio)} & 2.56  & 2.67 \\\hline

\end{tabular}
\end{frame}

\begin{frame}
\frametitle{Model Allows for Eliminating Selection}

\begin{itemize}
\item Generate $5{,}000$ individuals from unconditional distribution
\item No one meets partners till 25.
\item At 25, two scenarios. All of them:
\begin{itemize}
\item Meet and become pregnant --- KF if agree
\item Meet and do not become pregnant --- MF if agree
\end{itemize}
\item Look at their outcomes over time
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Causal Impact $\approx$ Observable Impact}
\begin{center}
\includegraphics[width=0.85\linewidth]{../main/individual-av.eps}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Lower Child Investment Even without Selection}
\vspace{-2cm}
\begin{center}
\includegraphics[width=1.0\linewidth]{../main/x-time.pdf}
\end{center}
\end{frame}

\begin{frame}
\frametitle{``Compliers'' Drive Everything}
\begin{center}
\includegraphics[width=0.85\linewidth]{../main/ind-decomp.eps}
\end{center}
\end{frame}

\newcommand{\pz}{\phantom{0}}
\begin{frame}
\frametitle{More Impact In Early Life}
\begin{center}
\begin{tabular}{ l c c c}\hline 
& &  \multicolumn{2}{ c }{\scriptsize  No Unplanned Pregnancies}  \\
& \scriptsize Baseline & \scriptsize Outside Marriage & \scriptsize At all \\\hline
\footnotesize Divorced by 25, \% & $\pz4.5$   & $\pz3.2$ & $\pz2.3$ \\
\footnotesize Divorced by 30 (low-$z$), \%  & $17.2$ & $14.6$   & $11.9$ \\
\footnotesize Divorced by 30 (high-$z$), \%  & $\pz9.0$ & $\pz8.2$  &  $\pz7.4$ \\
\footnotesize Divorced by 35, \%  & $10.8$  & $\pz9.9$&   $\pz9.2$  \\
\footnotesize Divorced by 40, \%  & $16.3$ & $15.4$ &  $14.8$ \\\hline
\end{tabular}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Conclusions And Plans}
\begin{itemize}
\item What we observe about childcare costs generates the pattern
\item KF couples have lower $\psi$ and divorce more
\item Unplanned pregnancies account for $1/2$ of early-age divorces
\item Preparing a new iteration now:
\begin{itemize}
\item TBD: partition by income needs to be improved
\item TBD: decompose by different contributors ($z$, $\psi$, age)
\end{itemize}
\item Model allows to account for other stories:
\begin{itemize}
\item Labor market discrimination of unmarried mothers
\item Social stigma against unmarried cohabitation
\end{itemize}
\end{itemize}

\end{frame}

\section{Data Patterns}
\begin{frame}
\frametitle{Pregnancy $\Rightarrow$ Transition Towards Marriage} 
\framesubtitle{NSFH, 2015--2017, weighted frequencies}
\begin{center}
{\Large Before And After Pregnancy}
\begin{tabular}{ l r r r }
 & Single after & Cohabiting after & Married after \\ \hline
\multicolumn{4}{ c }{\textbf{Everyone who gave a birth:}}\\\hline
Single before  & 69.0 & 21.0 & 10.1 \\
Cohabiting before & 4.0 &  74.8 & 21.2 \\
Married before & 0 & 0 & 100 \\\hline
\multicolumn{4}{ c }{\textbf{Ever married only:}}\\\hline
Single before & 56.0 & 22.4 & 21.7 \\
Cohabiting before  & 0.9 & 58.9 & 40.2 \\
Married before & 0 & 0 & 100 \\\hline
\multicolumn{4}{p{0.95\linewidth}}{\small \textit{Note:} this is based on marital status before and after the first reported successful pregnancy.}\\\hline
\hline
\end{tabular}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Definitions: Kids First and Marriage First}
\begin{itemize}
\item Women are units of observation
\item Two groups:
\begin{enumerate}
\item Kids first (\textbf{KF}): youngest child is born before or at the year of marriage. Includes:
\begin{itemize}
\item True shotgun marriages: married when became pregnant
\item Couples who married some time after child's birth 
\item Single mother marrying different partners
\end{itemize}
\item Marriage first (\textbf{MF}): youngest child at least at the next year after marriage
\end{enumerate}
\item Not a full partition
\item Excludes never married
\item Finer partition among KF is available in some datasets
%\item Exclude if more than 5 years before or 10 years after
\end{itemize}
\end{frame}

\begin{frame}[label=disc-sharp]
\frametitle{Sharp Discontinuity in Divorce Rates (Once Again)}
\begin{center}
\includegraphics[width=0.85\linewidth]{diff_dt_h.pdf}
\end{center}
\hyperlink{disc2}{\beamerbutton{controls}}
\end{frame}

\begin{frame}
\frametitle{Composition is Different}
\begin{center}\small
\begin{tabular}{ l c c c c }\cline{2-4}
\multicolumn{1}{c}{} &  \multicolumn{3}{c}{Married once + have kids}\\\cline{2-4}
\multicolumn{1}{c}{}  & Everyone & Kids-first & Marriage-first \\\cline{2-4}\hline
%\textit{Age}     &  33.1     &       31.7     &       33.6 \\
\textit{Age at first marriage}         &   23.3        &    23.6        &    23.2 \\
\textit{Age at first birth}         &     25.4     &       21.7       &     26.5 \\
\textit{Husband's age}         &    35.8      &      34.4   &         36.2  \\
%\textit{Age difference (W$-$H)}        &      2.8      &       3.0      &       2.7  \\\hline
\textit{Eldest child's age}         &    7.8    &        10.0     &        7.1 \\
\textit{Number of children}        &      2.0      &       2.3     &       2.0 \\\hline
\textit{College or + (own), percent}    &         39.5     &       16.9     &       46.3 \\
\textit{College or + (husband), percent}      &       36.4     &       14.2       &     42.7 \\\hline
\textit{In Labor Force, percent}     &        68.6    &        70.0        &    68.2 \\
\textit{Own income, in \$1000}  &         38.4 &          29.1 &         41.1 \\
%\textit{Hours worked}       &     36.5    &        36.8        &    36.4\\
\hline
\textit{White, percent}       &      76.5      &      72.2     &      77.8 \\
\textit{Black, percent}  &        6.7 &            12.2       &       5.0 \\\hline\hline
\multicolumn{4}{p{0.8\linewidth}}{\footnotesize ACS, women, 20--40. Means. Everything is for subsamples when the characteristic is well-defined, for instance, income and hours are for subsample of women who are employed.}\\\hline
\end{tabular}
\end{center}
\end{frame}

\begin{frame}[label=ratios-ACS]
\frametitle{ACS: Shares of Divorced}
\begin{center}
\begin{tabular}{ l  c  c c c | c }\hline
&  \multicolumn{4}{ l|}{\footnotesize{\% of divorced (now), 20--40}} & \\
  & \footnotesize MF & \footnotesize  KF & \footnotesize Ratio & \footnotesize Adjusted &  \scriptsize  \textbf{\% of  KF} \\ \hline
\textit{Everyone} &  \footnotesize 10.1 &  \footnotesize 18.0 & \footnotesize \textbf{1.8} & \footnotesize 1.6 &  \footnotesize 23.2 \\
\textit{35--40} &  \footnotesize 11.3  &  \footnotesize 23.0  & \footnotesize \textbf{2.0} & \footnotesize 1.6 &  \footnotesize 18.0 \\
\textit{High school} &\footnotesize 13.9 & \footnotesize  17.2 &  \footnotesize  \textbf{1.2} & \footnotesize 1.6 & \footnotesize 34.2 \\
\textit{College grads} & \footnotesize 5.3 &  \footnotesize 14.7 &  \footnotesize  \textbf{2.8} & \footnotesize 1.7 &  \footnotesize 9.9 \\\hline
\multicolumn{6}{p{0.8\linewidth}}{\footnotesize ``Adjusted'' refers to corrected for compositions using controls: age, age at 1st marriage, age at 1st birth, education, race, state,  ACS year, number of children}\\\hline\hline
\end{tabular}
\end{center}

\begin{itemize}
\item ACS does not have full marital history, so this is for sample of women who married exactly once 
\item SIPP with full marital history --- same results
\item Income controls --- minor differences, but attrition.
\end{itemize}

\hyperlink{table-ACS}{\beamerbutton{Diff + SE}}
%\hyperlink{pic-ACS}{\beamerbutton{Picture}} 
\hyperlink{table-sipp}{\beamerbutton{SIPP table}} 
\hyperlink{age1b}{\beamerbutton{By age of the first birth}} 
\hyperlink{age1m}{\beamerbutton{By age of the first marriage}}

\end{frame}

\begin{frame}[label=nsfg]
\frametitle{NSFG: The Finest Partition}
Recent data: 2015--2017
\begin{center}
\begin{tabular}{ l  c  c c | c }\hline
\multicolumn{4}{ c |}{Percentage of Divorced, Everyone, 20--40} & \\\hline
 &\footnotesize Percentage & \footnotesize Ratio & \footnotesize Adjusted & \footnotesize \textit{Share}  \\\hline
\footnotesize Marriage First &  24.5 & \textbf{1.00} & 1.00   &  58.8 \\
\footnotesize KF: true shotgun  & 50.7 & \textbf{2.07} & 1.68  & 12.4 \\
\footnotesize KF: before marriage  & 36.8  & \textbf{1.50} & 1.60 & 20.6 \\
\footnotesize KF: new partner & 31.7 & \textbf{1.29} & 1.64 & \phantom{0}8.1  \\\hline
\multicolumn{4}{p{0.7\linewidth}}{\footnotesize ``Adjusted'' refers to corrected for compositions using controls: age, age at 1st marriage, age at 1st birth, education, race, number of children}\\\hline\hline
\end{tabular}
\end{center}
\end{frame}

\begin{frame}[label=hazards]
\frametitle{NSFH: Hazards of Divorce}
Historical Data: average 40 years old at 1987
\begin{center}
\begin{tabular}{ l  c  c c c }\hline
\multicolumn{5}{ c }{Cox Regression: Hazard Ratios (S.E.)}\\\hline
 &\footnotesize Everyone & \footnotesize Everyone & \footnotesize Non-college & \footnotesize  College \\\hline
\footnotesize all KF & \footnotesize 1.45 (0.12) & --- &\footnotesize  1.45 (0.15) &\footnotesize  2.23 (0.81)  \\ 
\footnotesize true shotgun  & --- &\footnotesize  1.34 (0.12) & --- & --- \\
\footnotesize before marriage  & --- & \footnotesize  2.18 (0.28) & --- & ---  \\\hline\hline
\multicolumn{5}{p{0.9\linewidth}}{\footnotesize controls: age, age at 1st marriage, age at 1st birth, education, race, number of children}\\\hline\hline
\end{tabular}
\end{center}
\end{frame}

\begin{frame}
\frametitle{More Evidence}
\begin{enumerate}
\item Divorce rates start being different when the eldest child is 5--10 years old
\item Comparing marriages of the same duration exacerbates the difference
\item NLSY, Cox duration model: 2.3 times more likely to divorce if same-year (thank Fabio Blasutto)
\item Children in KF couples repeat grades more often
\end{enumerate}
\end{frame}

\backupbegin

\appendix
\section{Appendix}

\begin{frame}[label=pic0]
\frametitle{In One Picture (Old \& Extended Version)}
\begin{center}
\includegraphics[width=0.85\linewidth]{diff_dt_h.pdf}
\end{center}
\end{frame}

\begin{frame}[label=table-ACS]
\frametitle{Slicing and Adding Controls}
\begin{center}
\begin{tabular}{ l  c  c c c | c }\hline
&  \multicolumn{4}{ l|}{\footnotesize{\% of divorced (now), 20--40}} & \\
  & \footnotesize MF & \footnotesize  KF & \footnotesize Diff / (s.e.) & \footnotesize Diff w/controls${}^*$ &  \scriptsize  \textbf{\% of  KF} \\ \hline
\textit{Everyone} &  \footnotesize10.1 &  \footnotesize 18.0 & \footnotesize \textbf{\phantom{0}7.9 / (0.1)} & \footnotesize 5.7 / (0.2) &  \footnotesize 23.2 \\
\textit{35--40} &  \footnotesize 11.3  &  \footnotesize 23.0  & \footnotesize \textbf{11.7 / (0.1)} & \footnotesize 6.7 / (0.3) &  \footnotesize 18.0 \\
\textit{High school} &\footnotesize 13.9 & \footnotesize  17.2 &  \footnotesize  \textbf{\phantom{0}3.3 / (0.2) }& \footnotesize 8.4 / (0.3) & \footnotesize 34.2 \\
\textit{College grads} & \footnotesize 5.3 &  \footnotesize 14.7 &  \footnotesize  \textbf{\phantom{0}9.4 / (0.2) }& \footnotesize 3.8 / (0.3) &  \footnotesize 9.9 \\\hline
\multicolumn{6}{p{0.9\linewidth}}{\footnotesize ${}^*$ controls include: age, age at 1st marriage, age at 1st birth, race, state, ACS year, number of children}\\\hline\hline
\end{tabular}
\end{center}

\hyperlink{ratios-ACS}{\beamerbutton{back}} 
\hyperlink{pic-ACS}{\beamerbutton{Picture}} 
\hyperlink{table-sipp}{\beamerbutton{SIPP table}} 
\hyperlink{age1b}{\beamerbutton{By age of the first birth}} 
\hyperlink{age1m}{\beamerbutton{By age of the first marriage}}
\end{frame}

\begin{frame}[label=pic-ACS]
\begin{center}
\includegraphics[width=1.0\linewidth]{../main/shrs.eps}
\end{center}
\hyperlink{table-ACS}{\beamerbutton{back}}
\end{frame}

\begin{frame}[label=table-sipp]
\frametitle{SIPP}
\framesubtitle{Now with full marital history}
\begin{center}
\begin{tabular}{ l  c  c c c | c }\hline
&  \multicolumn{4}{ l|}{\footnotesize{\% of divorced (ever) among women of 20--40}} & \\
  & \footnotesize MF & \footnotesize  KF & \footnotesize Diff / (s.e.) & \footnotesize Diff w/controls${}^*$ &  \scriptsize  \textbf{\% of  KF} \\ \hline
\textit{Everyone} &  \footnotesize 21.1 &  \footnotesize 32.0 & \footnotesize \textbf{10.9 / (0.6)} & \footnotesize 10.4 / (0.8) &\footnotesize 22.8 \\
\textit{35--40} &  \footnotesize 24.5 &  \footnotesize 44.1 & \footnotesize \textbf{19.6 / (1.0)} & \footnotesize 12.4 / (1.4) &\footnotesize 18.9 \\
\textit{High school} &\footnotesize 26.6  &\footnotesize 30.2 &\footnotesize \textbf{\phantom{0}3.6 / (0.9) }& \footnotesize 14.5 / (1.4) &\footnotesize 33.3\\
\textit{College grads} & \footnotesize 11.7  & \footnotesize26.9 &\footnotesize \textbf{15.2 / (1.3)} & \footnotesize 11.5 / (1.4) &\footnotesize 9.5 \\\hline
\multicolumn{6}{p{0.9\linewidth}}{\footnotesize ${}^*$ controls include: age, age at 1st marriage, age at 1st birth, race, state, number of children}\\\hline\hline
\end{tabular}
\end{center}
\begin{itemize}
\item Drawback: smaller sample size 
\end{itemize}
\hyperlink{table-ACS}{\beamerbutton{back}}
\end{frame}

\begin{frame}[label=disc2]
\frametitle{Discontinuity is Still Here}
\begin{center}
\includegraphics[width=0.85\linewidth]{diff-by-dt-con.eps}
\end{center}
\hyperlink{disc-sharp}{\beamerbutton{back}}
\end{frame}

\begin{frame}[label=age1b]
\frametitle{By age of 1st birth}
\begin{center}
\includegraphics[width=0.8\linewidth]{joint-reg-by-age1b.eps}
\end{center}
\hyperlink{age1b-con}{\beamerbutton{controls}}
\hyperlink{table-ACS}{\beamerbutton{back}}
\end{frame}

\begin{frame}[label=age1b-con]
\frametitle{By age of 1st birth + controls}
\begin{center}
\includegraphics[width=0.8\linewidth]{joint-reg-by-age1b-con.eps}
\end{center}
\hyperlink{table-ACS}{\beamerbutton{back}}
\end{frame}

\begin{frame}[label=age1m]
\frametitle{By age of 1st marriage}
\begin{center}
\includegraphics[width=0.8\linewidth]{joint-reg-by-age1m.eps}
\end{center}
\hyperlink{age1m-con}{\beamerbutton{controls}}
\hyperlink{table-ACS}{\beamerbutton{back}}
\end{frame}

\begin{frame}[label=age1m-con]
\frametitle{By age of 1st marriage + controls}
\begin{center}
\includegraphics[width=0.8\linewidth]{joint-reg-by-age1m-con.eps}
\end{center}
\hyperlink{table-ACS}{\beamerbutton{back}}
\end{frame}

\begin{frame}[label=value-functions]
\frametitle{Value Functions --- 1}
\framesubtitle{Single Female}
{\scriptsize
\begin{align}V^{f,s}_t(\omega) = \max\limits_{c} & \bigg\{ u(c) + \beta \E_t \Big[ (1 - p^{\text{meet}}_t)\cdot V^{f,s}_{t+1}(\omega') + \nonumber \\  \nonumber
& \hspace{7em} p^{\text{meet}}_t (1-p^{\text{preg}}_t) \big\{ m^{np} \cdot V^{f,c}_{t+1}(\Omega^c) + (1-m^{np})V^{f,s}_t(\omega')\big\} + \\  \nonumber
& \hspace{10em} p^{\text{meet}}_t p^{\text{preg}}_t \big\{ m^{p} \cdot V^{f,ck}_{t+1}(\Omega^c) + (1-m^{p})V^{f,sk}_{t+1}(\omega')\big\}  \Big]  \bigg\},
\end{align}\vspace{-3em}
\begin{align*}
 \text{s.t. \ }  &  a' = R\cdot a  + W^f_t(z) - c  & \text{ (evolution of assets)}\\
 &  z' = z + \varepsilon^{z,f}_t, \ \ \varepsilon^{z,f}_t \sim \mathcal{N}(0;\sigma_{z,f}^2) &  \text{ (evolution of productivity)}\\
  & \log W^f_t = z_t + \text{Trend}^f_t, \ \ \  \text{Trend}^f_t = a^f_0 + a^f_1\cdot t  +  a^f_2 \cdot t^2 &  \text{ (labor income and trend)}\\
  & \Omega^c = \mathcal{M}^f(a',z') &  \text{ (marriage prospectives)}
\end{align*}
}
\hyperlink{utilities}{\beamerbutton{back}}
\end{frame}

\begin{frame}[label=value-functions]
\frametitle{Value Functions --- 2}
\framesubtitle{Couple, No Children}
{\scriptsize
\begin{align}& \hspace{5em}  V^{c}_t(\Omega) = \max\limits_{c^f,c^m}  \bigg\{ \theta^f \cdot u(c^f) + \theta^m \cdot u(c^m) + \psi + \nonumber \\   \nonumber
 &  \beta \E_t \Big[   (1 - p^{\text{preg}}_t)\cdot \left\{ (1-d^{\text{np}})\cdot \max\left\{ V^{ck}_{t+1},V^{c}_{t+1}\right\} + d^{\text{np}}\cdot [ \theta V_{t+1}^{f,s} + (1-\theta)V_{t+1}^{m,s}]\right\}  +  \\  \nonumber
& \hspace{5em} p^{\text{preg}}_t\cdot \left\{ (1-d^{\text{p}})\cdot V^{ck}_{t+1} + d^{\text{p}}\cdot [ \theta V_{t+1}^{f,sk} + (1-\theta)V_{t+1}^{m,s}]\right\} \Big] \bigg\},
\end{align}\vspace{-2em}
\begin{align*}
\text{s.t. \ }& a' + c = R\cdot a  + W^m_t(z^m) + W^f_t(z^f) & \text{(evolution of joint assets)},\\
				 & c = [(c^f)^{1+\rho_c} + (c^m)^{1+\rho_c}]^{\frac1{1+\rho_c}} & \text{(increasing returns in consumption)},\\
				 &  z^{f\prime} = z^f + \varepsilon^{z,f}_t, \ \ \varepsilon^{z,f}_t \sim \mathcal{N}(0;\sigma_{z,f}^2) &  \text{ (evolution of female productivity)}\\
				 &  z^{m\prime} = z^m + \varepsilon^{z,m}_t, \ \ \varepsilon^{z,m}_t \sim \mathcal{N}(0;\sigma_{z,m}^2) &  \text{ (evolution of male productivity)}\\
                    & \psi' = \psi + \varepsilon^{\psi}_t, \ \ \varepsilon^{\psi}_t \sim \mathcal{N}(0;\sigma_{\psi}^2)  & \text{(evolution of marriage surplus),} \\
                    & (\Omega',\omega^{df},\omega^{dm}) = \mathcal{R}(\theta,a',z^{m\prime},z^{f\prime},\psi') & \text{(renegotiation correspondence)},
\end{align*}
}
\hyperlink{utilities}{\beamerbutton{back}}
\end{frame}

\begin{frame}[label=value-functions]
\frametitle{Value Functions --- 3}
\framesubtitle{Couple + Child}
\hspace{-2em}
{\scriptsize
\begin{align}& \hspace{5em}  V^{ck}_t(\Omega) = \max\limits_{c^f,c^m,q,l_f,x}  \bigg\{ \theta^f\cdot u(c^f) + \theta^m \cdot u(c^m) + u^{\text{kid}}(q) +  \psi + \nonumber \\  \nonumber
 & \hspace{8em} \beta \E_t \Big[   \left\{ (1-d)\cdot   V^{ck}_{t+1}(\Omega') + d\cdot [ \theta^f V_{t+1}^{f,sk}(\omega^{df}) + \theta^m V_{t+1}^{m,s}(\omega^{dm})]\right\} \Big] \bigg\},
\end{align}\vspace{-2em}
\begin{align*}
 & a' + c + x = R\cdot a  + W^m_t(z^m) + l_f \cdot W^f_t(z^f) & \text{(evolution of joint assets)},\\
                    & c = [(c^f)^{1+\rho_c} + (c^m)^{1+\rho_c}]^{\frac1{1+\rho_c}} & \text{(increasing returns in consumption)},\\
                    & q = [ x^{\lambda} + \kappa \cdot (1-l_f)^{\lambda} ]^{\frac1{\lambda}}  & \text{(production function of child quality)},\\
                    & q \geq \underline{q} & \text{(required childcare costs)},\\
                    &  z^{f\prime} = z^f + \varepsilon^{z,f}_t - f(l), \ \ \varepsilon^{z,f}_t \sim \mathcal{N}(0;\sigma_{z,f}^2) &  \text{ (evolution of female productivity)}\\
				 &  z^{m\prime} = z^m + \varepsilon^{z,m}_t, \ \ \varepsilon^{z,m}_t \sim \mathcal{N}(0;\sigma_{z,m}^2) &  \text{ (evolution of male productivity)}\\
                    & \psi' = \psi + \varepsilon^{\psi}_t, \ \ \varepsilon^{\psi}_t \sim \mathcal{N}(0;\sigma_{\psi}^2)  & \text{(evolution of marriage surplus),} \\
                    & (\Omega',\omega^{df},\omega^{dm}) = \mathcal{R}(\theta^f,\theta^m,,a',z^{m\prime},z^{f\prime},\psi',\xi) & \text{(renegotiation correspondence)},
\end{align*}
}
\hyperlink{utilities}{\beamerbutton{back}}
\end{frame}

\begin{frame}[label=value-functions]
\frametitle{Value Functions --- 4}
\framesubtitle{Utilities}
\hspace{-2em}
{\scriptsize
\begin{align*}
 & u(c) = \frac{c^{1-\sigma}}{1-\sigma}  & \text{(individual)},\\
 & U(c^f,c^m) = \theta\cdot \frac{(c^f)^{1-\sigma}}{1-\sigma} + (1-\theta)\cdot \frac{(c^m)^{1-\sigma}}{1-\sigma} \rightsquigarrow A(\theta) \frac{c^{1-\sigma}}{1-\sigma}  & \text{(couple)},                 \\
 & \tilde{u}(c) = \frac{c^{1-\sigma}}{1-\sigma}  + \phi_0 + \frac{Q^{1-\sigma}}{1-\sigma} & \text{(individual + child)},\\
 & \tilde{u}(c) = \theta \frac{(c^f)^{1-\sigma}}{1-\sigma}  +(1- \theta) \frac{(c^m)^{1-\sigma}}{1-\sigma}  + \phi_0 + \frac{Q^{1-\sigma}}{1-\sigma} & \text{(couple + child)}.
\end{align*}
}
\hyperlink{utilities}{\beamerbutton{back}}
\end{frame}

%
%\begin{frame}
%{\scriptsize
%\begin{center}
%\begin{tabular}{ c p{0.15\linewidth} c p{0.15\linewidth}}\hline\hline
%Parameters & Description & Values & Identifying variation \\\hline
%($\sigma_{\psi,0},\sigma_{\psi}$) & { \footnotesize
% Marriage surplus initial distribution \& shocks, standard deviations} &{\small$(0.76,1.73)$ }& {\footnotesize
% \% of divorced at $(25,35,40)$, \% of women with one marriage at 40} \\\hline\hline
%$p^{\text{meet}}_{\{t = 21, 30, 37, 45\}}$ & {\footnotesize Meeting probabilities by age (linearly interpolated in between)} & {\small
% $(0.13,0.10,0.07,0.12)$ }& {\footnotesize \% of never married by age , \% of women with one marriage at 40} \\\hline
%$(\phi_{0,c},\phi_3)$ & {\footnotesize Scale parameters for utility of child's quality} & {\small$(-1.89,2.58)$} & {\footnotesize \% of women with no children at 25, 35, and at 30 by income groups}\\\hline
%$(\phi_1,\phi_2,\theta)$ & {\footnotesize Child quality preference structure and substitutability} & {\small$(0.44,1.07,-1.02)$}  & {\footnotesize Average share of money spent on children, ratio of money and time spent by income groups} \\\hline
%$(\mu_y,\mu_a)$ & {\footnotesize Time structure of childcare by child's age} &  {\small$(0.35,0.01)$} & {\footnotesize Time spent with children 0--5 relative to 6--12}  \\\hline
%$\bar{q}$ & {\footnotesize Childcare quality requirement } & {\small$2.12$} & {\footnotesize Income of never married with kids and divorced with kids relative to general population} \\\hline
%$(p^{\text{out}},\phi_{0,s})$ & {\footnotesize Single mothers: probability of re-entering marriage market and utility shifter} & {\small$(0.10,-3.34)$} & {\footnotesize \% of divorced with kids and never married with kids at $30$} \\\hline\hline
%$p^{\text{preg}}_{0,1,2,3}$  & {\footnotesize Arrival probabilities of unplanned pregnancies by age and $z$} & {\small$(0.67,0.13,-0.50,-0.005)$} & {\footnotesize Share of kids-first at 25, 35; share of kids-first at 30 and 40 by income groups} \\\hline
%\end{tabular}
%\end{center}
%}
%\end{frame}

\backupend
\end{document}

### 2020-10-02 slides-march-20/slides-old.tex
Current stage

--- 2020-10-02 slides-march-20/slides-new.tex
@@
 
-\usepackage{fixmath}
 \usepackage{tikz}
@@
 
-% There is a huge social issue of single mothers
-% You typically associate it with unplanned teenage pregnancies of low-income and low-education people
-% But looking at just single mothers or young couples is an understatement
-% Unplanned pregnancy expresses itself even for married people: not all marriages are created equal
-% In later ages unplanned pregnancies force couple to marry
-% 
-% Even for college 
-% In more adult age this is pronounced by shotgun marriage
-% Chetty stuff: share of single mothers is the best predictor of inequality and so on
-
 \begin{frame}[plain]
@@
 
+\begin{frame}%[label=pic0]
+\frametitle{In One Picture}
+\begin{center}
+\includegraphics[width=0.85\linewidth]{diff-by-dt.eps}
+\end{center}
+%\hyperlink{disc2}{\beamerbutton{dis}}
+\end{frame}
+
 \begin{frame}
-\frametitle{Unplanned Pregnancy $\Rightarrow$ Shotgun Marriage}
+\frametitle{Shotgun Marriages Perform Worse}
 \begin{itemize}
-\item Many people marry shortly after a baby appears
-\item These marriages perform worse in the data 
+\item Couples who have their kids before marriage divorce more:
 \begin{itemize}
-\item A lot worse for college grads, a bit worse for high school grads
-\item $\sim10\%$ marry while being pregnant
-\item Not just a teenage thing, peaks around 30
+\item \textbf{Especially} among college graduates
+\item Selection may not be the primary reason
 \end{itemize}
-\item Understanding why helps to understand the value of marriage
-\item Questions:
+\item This teaches us about heterogeneity of marriages:
 \begin{itemize}
-\item What is the aggregate role of unplanned pregnancies?
-\item What causes people to enter such marriages?
-\item Why they perform worse?
+\item Raising children alone is hard
+\item Unplanned pregnancies $\Rightarrow$ lower quality marriages
+\item Still important despite access to contraception and abortion
+\end{itemize}
+\pause
+\item \textit{Unexpected pregnancy often triggers marriage}
+\item \textit{Unmarried unions with children are rare}
+\begin{itemize}
+\item \textit{2013: 3\% of children live with unmarried parents}
+\item \textit{As opposed to 21\% with single mothers}
 \end{itemize}
@@
 
-\begin{frame}[label=pic0]
-\frametitle{Divorced By Timings}
-\begin{center}
-\includegraphics[width=0.85\linewidth]{share_div_march5.eps}
-\end{center}
+\begin{frame}
+\frametitle{What I Do}
+
+\begin{itemize}
+\item Document the fact empirically. Lots of selection!
+\item Construct and estimate a structural model 
+\item Use it to understand observable differences in divorce rates
+\end{itemize}
+
+Main questions:
+\begin{itemize}
+\item What explains bad performance of shotgun marriage
+\item Why college graduates are affected relatively more
+\item Role of policies (childcare etc)
+\end{itemize}
 \end{frame}
 
-\begin{frame}[label=pic0]
-\frametitle{Hazards}
-\begin{center}
-\includegraphics[width=0.85\linewidth]{haz_kf_march5.eps}
-\end{center}
-\end{frame}
-
-\begin{frame}[label=pic0]
-\frametitle{Shares}
-\begin{center}
-\includegraphics[width=0.85\linewidth]{share_kj_march5.eps}
-\end{center}
-\end{frame}
-
-\begin{frame}[plain]
-\frametitle{What I Do}
+\begin{frame}
+\frametitle{Main Results}
 \begin{itemize}
-\item Quantify empirical difference of marriage performance by the timing of childbirth
-\item Estimate a structural model matching this difference
-\item Use the model to study:
+\item Empirical:
 \begin{itemize}
-\item aggregate impact of unplanned childbirths 
-\item importance of fertility for marriage
-\item difference between education groups
-\item policy counterfactuals
+\item Shotgun marriage $\Rightarrow$ divorced $\sim1.8$ more often
+\item Among college graduates: $\sim3$ times more often
+\item Differences are present after controlling for lots of observables
+\item Present is many data sets: ACS, SIPP, NSFG, NSFH
+\end{itemize}
+\item From the model:
+\begin{itemize}
+\item Disentangling two main mechanisms:
+\begin{itemize}
+\item Shotgun marriage couples are different in characteristics
+\item Unexpected pregnancy creates large unplanned costs
+\end{itemize}
+\item Expensive eraly-life childcare explains a lot
+\item Higher marginal costs of childcare for skilled women
+\item $\sim10\%$ of couples marry because of unplanned pregnancy
+\item These couples are the least stable
+\end{itemize}
+\item Missing pieces (TBD):
+\begin{itemize}
+\item Quantifying implications
+\item Quantifying changes over time
 \end{itemize}
@@
 
-\begin{frame}[plain]
-\frametitle{Preview of Results}
-
-\begin{enumerate}
-
-\item Data: 
-{\small
-\begin{center}
-\begin{tabular}{|c|c|c|c|}\hline
- & \multicolumn{2}{|c|}{\textit{Share of divorced among:}} & \\
-& Marriage First & Kids First & Ratio \\\hline
-College &  5.3\% &  14.7\% & \textbf{2.77} \\
-High School &  13.9\% & 17.2\% &  \textbf{1.24}  \\\hline
-\end{tabular}
-\end{center}
-}
-
-\item Quantifying:
-\begin{itemize}
-\item Unplanned pregnancy changes outside options in family bargaining
-\item Effect is present generically, its size identifies amount of distress
-\item To estimate, assume optimal fertility timing after marriage
-\end{itemize}
-
-\item Exploring:
-\begin{itemize}
-\item Unplanned pregnancies account for $\sim50\%$ of early divorces
-\item Marriage market (meeting probabilities) mostly drive the college -- high school difference
-\end{itemize}
-
-\end{enumerate}
-\end{frame}
-
-%
-%\begin{frame}
-%\frametitle{[Distribution of age at shotgun marriage in my def]}
-%
-%\end{frame}
-
-%
-%\begin{frame}
-%\frametitle{Main Results}
-%\begin{itemize}
-%\item Empirical:
-%\begin{itemize}
-%\item Shotgun marriage $\Rightarrow$ divorced $\sim1.8$ more often
-%\item Among college graduates: $\sim3$ times more often
-%\item Differences are present after controlling for lots of observables
-%\item Present is many data sets: ACS, SIPP, NSFG, NSFH
-%\end{itemize}
-%\item From the model:
-%\begin{itemize}
-%\item Disentangling two main mechanisms:
-%\begin{itemize}
-%\item Shotgun marriage couples are different in characteristics
-%\item Unexpected pregnancy creates large unplanned costs
-%\end{itemize}
-%\item Expensive eraly-life childcare explains a lot
-%\item Higher marginal costs of childcare for skilled women
-%\item $\sim10\%$ of couples marry because of unplanned pregnancy
-%\item These couples are the least stable
-%\end{itemize}
-%\item Missing pieces (TBD):
-%\begin{itemize}
-%\item Quantifying implications
-%\item Quantifying changes over time
-%\end{itemize}
-%\end{itemize}
-%\end{frame}
-
 \begin{frame}
-\frametitle{General Importance}
+\frametitle{Implications}
 \begin{itemize}
@@
 \begin{itemize}
-\item Labor market: loss of skills
+\item Labor market choices and outcomes
 \item Divorces + children worsen re-marriage prospective
@@
 \end{enumerate}
-%
-%\textit{Other relevant evidence:}
-%\begin{itemize}
-%\item Divorce rates by income: Boertien, Härkönen (2014)
-%\item Child expenditures by income: Schneider et al (2014)
-%\end{itemize}
+
+\textit{Other relevant evidence:}
+\begin{itemize}
+\item Divorce rates by income: Boertien, Härkönen (2014)
+\item Child expenditures by income: Schneider et al (2014)
+\end{itemize}
 %\item Computations: Judd et al (2014) (thank Luigi Bocola)
-
-\end{frame}
-
-\begin{frame}
-\frametitle{Updates Since November}
-\begin{itemize}
-\item Brand new codes in Python \& cuda
-\item Remarriage of single mothers
-\item Skills depreciation because of time to childcare
-\item No unplanned pregnancies in marriage: restrictive but cleaner identification
-\item Two separate estimations: college grads versus non-college grads
-\item Decomposition exercises as a goal
-\end{itemize}
-\end{frame}
-
-\section{Mechanism}
-
-\begin{frame}
-\frametitle{Framework}
-\begin{itemize}
-\item Bargaining model with outside option of being single. Together if:
-\[V^{\text{fem},\text{married}} \geq V^{\text{fem},\text{single}} \ \& \ V^{\text{male},\text{married}} \geq V^{\text{male},\text{single}}\]
-\item Make choices and have some margins of adjustment
-\item Heterogeneous relationship quality (love shock), the best predictor of divorce
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{Unplanned Pregnancy Shock}
-\begin{itemize}
-\item Married couples can choose their fertility
-\item However, sometimes a child arrives by a random shock $\Rightarrow$ distress
-\item Size depends on how indifferent are people between different fertility timings
-\item College people are more sensitive?
-\item What about marriage/divorce?
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{Asymmetric Response to Shock}
-People stay together if
-\[V^{\text{fem},\text{married}} \geq V^{\text{fem},\text{single}} \ \& \ V^{\text{male},\text{married}} \geq V^{\text{male},\text{single}}\]
-\vspace{-1em}
-\begin{itemize}
-\item Bargaining model: if stress is symmetric (all $V\downarrow$ the same way) neither more marriages nor more divorces happens
-\item $\Rightarrow$ females lose more, males gain more
-\item If marriages differ in quality, more low-quality marriages are created
-\end{itemize}
-
-Why do people enter low-quality marriages?
-\begin{itemize}
-\item Women:
-\begin{itemize}
-\item splitting of childcare costs
-\item worsening of future marital prospectives
-\end{itemize}
-\item Men:
-\begin{itemize}
-\item loss of access to their children otherwise
-\end{itemize}
-\end{itemize}
-\end{frame}
-
-\section{Model}
-
-\begin{frame}
-\frametitle{Overview}
-\begin{itemize}
-\item Lifecycle model
-\item Assets ($a$) and productivity shocks ($z^f$, $z^m$)
-\item Labor supply for females with kids
-\item Endogenous marriage and divorce (limited commitment)
-\item Reduced form marriage market:
-\[p^{\text{meet}}(\text{Age}) \text{ --- meeting a random partner with similar $a$ and $z$}\]
-\item After marriage: endogenous fertility choice
-\item Before marriage can have an unplanned child
-\[p^{\text{unplanned}}(\text{Age}) \text{ --- arrival of unplanned kids}\]
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{Modeling Shotgun Marriage}
-\begin{center}
-\vspace{-0.75cm}
-\begin{tikzpicture}[every text node part/.style={align=center}, scale=0.75]
-   % Place nodes
-   \node [block] (1) {\large \textbf{People}  \\[-0.5ex]  \large  \textbf{Randomly Meet}};
-   \node [block, below left = 1.5 cm and 0.0 cm of 1] (2) {\textit{Nash} \\[-0.5ex] \textit{Bargaining}};   
-   \node [block, below right = 1.5 cm and 0.0 cm of 1] (3) {\textit{Nash} \\[-0.5ex] \textit{Bargaining}};
-
-   \node [block, below = 1.5 cm of 2] (5) {\textbf{Kids First} \\[-0.5ex] \textbf{Couple $(ck)$}};
-   \node [block, below = 0.8 cm of 3] (7) {No Kids \\[-0.5ex] Couple $(c)$};
-   \node [block, below = 0.8 cm of 7] (8) {\textbf{Mariage First} \\[-0.5ex] \textbf{Couple  $(ck)$}};
-   \draw[->] (1) to [bend left = 20] node [left] {\scriptsize Unplanned \\[-0.75ex] \scriptsize pregnancy} (2);
-   \draw[->] (1) to [bend right = 20] node [right] {\scriptsize No \\[-0.75ex] \scriptsize pregnancy} (3);
-   \draw[->] (2) to node [left] {\scriptsize Agree} node [right] {\scriptsize \textbf{Shotgun} \\[-0.75ex]  \scriptsize \textbf{marriage}} (5);
-   \draw[->] (3) to node [right] {\scriptsize Agree} node [left]  {\scriptsize \textbf{Regular} \\[-0.75ex]  \scriptsize \textbf{marriage}} (7);
-   \draw[->] (7) to node [right] {\scriptsize Kids by choice} (8);
-
-\pause
-
-    \node [block, above left   = 1.0 cm and -0.75 cm of 2] (4) {\,\, Single male $(m,s)$\,\,};
-    \node [block, above left   = 1.6 cm and -0.75 cm of 2] (4-sm) {Single mother $(f,sk)$};
-	\node [block, above right   = 1.0 cm and -0.75 cm of 3] (6) {\,\, Single male $(m,s)$\,\,};
-    \node [block, above right   = 1.6 cm and -0.75 cm of 3] (6-f) {\,Single female $(f,s)$\,};   
-    \draw[->] (2.180) to [bend left = 30] node [left] {\scriptsize Disagree} (4.270);
-    \draw[->] (3.0) to [bend right = 30] node [right] {\scriptsize Disagree} (6.270);
-     \draw[->] (6.180) to [bend left = 30] node [above] {\scriptsize Back}  (1.-12);
-   \draw[->] (6-f.180) to [bend right = 30] node [above] {\scriptsize Back} (1.12);
-   \draw[->,thick,dashed] (4-sm.90) to [bend left = 30] node [above] {\scriptsize Worse marriage prospectives}  (1.168);
-   \draw[->] (4.0) to [bend right = 30] node [above] {\scriptsize Back} (1.192);
-
-\end{tikzpicture}
-\end{center}
-\end{frame}
-
-\begin{frame}
-\frametitle{Modeling Shotgun Marriage}
-\begin{center}
-\vspace{-0.4cm}
-\begin{tikzpicture}[every text node part/.style={align=center}, scale=0.75]
-   % Place nodes
-   \node [block] (1) {\large \textbf{People}  \\[-0.5ex]  \large  \textbf{Randomly Meet}};
-   \node [block, below left = 1.5 cm and 0.0 cm of 1] (2) {\textit{Nash} \\[-0.5ex] \textit{Bargaining}};   
-   \node [block, below right = 1.5 cm and 0.0 cm of 1] (3) {\textit{Nash} \\[-0.5ex] \textit{Bargaining}};
-
-   \node [block, below = 1.5 cm of 2] (5) {\textbf{Kids First} \\[-0.5ex] \textbf{Couple} ($ck$)};
-   \node [block, below = 0.8 cm of 3] (7) {No Kids \\[-0.5ex] Couple ($c$)};
-   \node [block, below = 0.8 cm of 7] (8) {\textbf{Mariage First} \\[-0.5ex] \textbf{Couple} ($ck$)};
-   \draw[->] (1) to [bend left = 20] node [left] {\scriptsize Unplanned \\[-0.75ex] \scriptsize pregnancy} (2);
-   \draw[->] (1) to [bend right = 20] node [right] {\scriptsize No \\[-0.75ex] \scriptsize pregnancy} (3);
-   \draw[->] (2) to node [left] {\scriptsize Agree} node [right] {\scriptsize \textbf{Shotgun} \\[-0.75ex]  \scriptsize \textbf{marriage}} (5);
-   \draw[->] (3) to node [right] {\scriptsize Agree} node [left]  {\scriptsize \textbf{Regular} \\[-0.75ex]  \scriptsize \textbf{marriage}} (7);
-   \draw[->] (7) to node [right] {\scriptsize Kids by choice} (8);   
-    \node [block, above left   = 0.8 cm and -0.75 cm of 2] (4) {\,Single male \\[-0.5ex] ($m,s$)\,\,};
-    \node [block, above left   = 1.8 cm and -0.75 cm of 2] (4-sm) {\,Single mom \\[-0.5ex]  ($f,sk$)};
-	\node [block, above right   = 0.8 cm and -0.75 cm of 3] (6) {\,\,Single male\,\,\\[-0.5ex] ($m,s$)};
-    \node [block, above right   = 1.8 cm and -0.75 cm of 3] (6-f) {Single female\\[-0.5ex] ($f,s$)\,};   
-    \draw[->] (2.180) to [bend left = 30] node [left] {\scriptsize Disagree} (4.270);
-    \draw[->] (3.0) to [bend right = 30] node [right] {\scriptsize Disagree} (6.270);
-     \draw[->] (6.180) to [bend left = 30] node [above] {\scriptsize Back}  (1.-12);
-   \draw[->] (6-f.180) to [bend right = 30] node [above] {\scriptsize Back} (1.12);
-   \draw[->,thick,dashed] (4-sm.90) to [bend left = 30] node [above] {\scriptsize Worse marriage prospectives}  (1.168);
-   \draw[->] (4.0) to [bend right = 30] node [above] {\scriptsize Back} (1.192);
-
-\end{tikzpicture}
-\end{center}
-\end{frame}
-
-\begin{frame}[label=utilities]
-\frametitle{Utility Functions}
-\begin{itemize}
-\item Couple, $\theta$ --- Pareto weight, $\psi$ --- common love shock:
-\[U = \theta^f \cdot u(c_f) + \theta^m \cdot u(c_m) + \psi,\]
-\item Couple + child, $Q$ --- child quality (public good):
-\[U = \theta^f \cdot {u}(c_f) + \theta^m \cdot {u}(c_m) + \psi + u^{\text{child}}(Q),\]
-\item Child quality: $Q = F(x,1-l^f)$, $x$ --- money, $l^f$ --- female labor supply
-\item Budget constraint ($C$ reflects returns to scale, $a$ --- assets):
- \[C(c_f,c_m) + a' + x = R\cdot a + l^f \cdot W^f_t(z_f) + W^m_t(z_m),\]
-\end{itemize}
-\hyperlink{value-functions}{\beamerbutton{Value Functions}}
-\end{frame}
-
-\begin{frame}
-\frametitle{Children In The Model}
-\begin{itemize}
-\item People like having children (public good):
-\[u^{\text{child}} = \phi + \alpha \cdot \frac{Q^{1-\sigma}}{1-\sigma}.\]
-\item $Q$ is produced out of money and female time
-\item $Q = \left[ x^{\lambda} + \kappa \cdot (1-l^f)^{\lambda}\right]^{\frac1\lambda}$
-\item \textit{Subsistence constraint:} $Q \geq \bar{Q}$
-\item $l^f\downarrow$ $\Rightarrow$ loss in current income + loss in productivity
-\item Productivity process: $z^f_t = z^f_{t-1} + \varepsilon^f_{t}  - $ {\boldmath $f(l)$}.
-\item In calibration $l \in \{0.25,0.5,0.75,0.9\}$
-\item Corresponding to depreciation: $f(l) \in \{0.1,0.05,0.0,0.0\}$.
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{Marriage}
-\begin{itemize}
-\item People (f \& m) meet, get a draw of love shock $\psi$ and agree to marry if exists a Pareto weight $\theta$ such that
-\[V^{f,\text{yes}}(\underset{\scriptsize +}{\theta\vphantom{\psi}},\underset{\scriptsize +}{\psi}) \geq V^{f,\text{no}}, \ \ \ \ V^{m,\text{yes}}(\underset{\scriptsize  -}{\theta\vphantom{\psi}},\underset{\scriptsize +}{\psi}) \geq V^{m,\text{no}}.\]
-\item \textbf{Mechanism:} unplanned pregnancy 
-\begin{enumerate}
-\item[(1)] lowers current $V^{f,\text{no}}$: hard to be a single mother
-\item[(2)] increases future $V^{m,\text{yes}}$ and $V^{f,\text{yes}}$: good to have children
-\end{enumerate} 
-$(1) + (2)$ $\Rightarrow$ the couple agrees to marry with lower $\psi$ and $\theta$.
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{Divorce}
-\begin{itemize}
-\item Shocks to $\psi$, $z^f$ and $z^m$
-\item People cannot commit not to divorce
-\item $\theta$ may be renegotiated, stays constant if not
-\item Divorce happens if no $\theta$ exist such that
-\[V^{f,\text{together}}(\theta,\psi) \geq V^{f,\text{divorce}}, \ \ \ \ V^{m,\text{together}}(\theta,\psi) \geq V^{m,\text{divorce}}.\]
-\item Couples with lower $\psi$ or more unequal $z$ divorce more!
-\end{itemize}
-\end{frame}
-
-%
-%\begin{frame}
-%\frametitle{Individual Values}
-%
-%\end{frame}
-
-\begin{frame}
-\frametitle{Discrete Decisions}
-\begin{itemize}
-\item Couple's problem defines $V^{c}_t$ (couple's value) and $(V^{f,c}_t,V^{m,c})$ (individual values).
-\item In the last period $V^{c}_t = \theta^f u(c^f) + \theta^m u(c^m)$, $V^{i,c} = u(c^i)$.
-\item Collective model (what I use): have a kid if 
-\[ \theta^f u(c^{f,k}) + \theta^m u(c^{m,k})  + u^{k} \geq \theta^f u(c^{f,nk}) + \theta^m u(c^{m,nk}).\]
-\item In fact this is ad hoc, and there is another version: have a kid if \textbf{both}
-\[ u(c^{f,k}) + u^{k} \geq u(c^{f,nk})  \ \text{ and } \ u(c^{m,k}) + u^{k} \geq u(c^{m,nk}).\]
-\item \textit{Decision making is not totally defined...}
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{Discrete Decisions + Rebargaining}
-\begin{itemize}
-\item Decision also depends on $\theta$
-\item $k_\theta = \I(V^{c,k}_\theta \geq V^{c,nk}_\theta)$ (or the 2nd protocol)
-\item Set of acceptable $\theta$ are
-\[ \Theta_f =  \{ \theta : V^{f,ck}_\theta \cdot k_\theta  + V^{f,nk}_\theta \cdot  (1-k_\theta ) \geq V^{f,\text{divorce}} \}\]
-\[ \Theta_m = \{ \theta : V^{m,ck}_\theta  \cdot  k_\theta+ V^{m,nk}_\theta  \cdot  (1-k_\theta ) \geq V^{m,\text{divorce}} \}\]
-\[ \Theta_c = \Theta_f \cap \Theta_m.\]
-\item Pick new $\theta$ if $\Theta_c$ is non-empty
-\item But in this case $\Theta_c$ may even be disjoint
-\item In this case use Marcet, Marimon (2019) idea and pick the closest $\theta$ to the original value
-\item But things can become quirky...
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{Remarriage}
-\begin{itemize}
-\item If couple with a child divorces, she becomes a single mother
-\item Single mothers still meet single males
-\item If partners agree to marry they become a regular couple with a child
-\item Males have a one-time utility penalty parameter for marrying a single mother. This affects bargaining:
-\[NBS = \left[ V^{m,MK} - {\text{\boldmath$\phi_m$}} - V^{m,S}\right] \times  \left[ V^{f,MK} - V^{f,S}\right].\]
-\item Rationalizes decrease in marital prospectives for females with children
-\item Estimated by matching share of remarried women with vs without children
-\end{itemize}
-\end{frame}
-
-\section{Fit and Results}
-\begin{frame}
-\frametitle{Strategy}
-\begin{itemize}
-
-\item Two calibrations: college and high school
-\item I use Simulated Method of Moments to calibrate $\sim 11$ parameters:
-\begin{itemize}
-\item Preferences
-\item Probabilities of transitions
-\item Child quality technology
-\end{itemize}
-\item Income paths to match ACS (trend and variance)
-\item The target moments are taken from ACS cross-section
-\item Childcare expenditure share from CEX
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{College vs High School}
-\begin{center}
-\begin{tabular}{|l|c|c|c|}
-& \multicolumn{2}{|c|}{Share of divorced, \%} & \\
-& MF & KF & Ratio \\
-Baseline: college & 5.4 & 14.0 &   2.6  \\
-+ income process of HS & 3.9 & 15.3 & 4.0 \\
-+ pregnancy probabilities of HS & 5.5 & 14.4 & 2.6 \\
-\textbf{+ meeting probabilities of HS} & \textbf{8.9} & \textbf{12.6}  & \textbf{1.4} \\
-+ love shock distribution of HS & 5.1 & 14.5 & 2.9 \\
-+ single mother prospective of HS & 5.0 & 15.1 &  3.0 \\
-Comparison: high school &  11.8 & 16.6 & 1.4\\
-\end{tabular}
-\end{center}
-\end{frame}
-
-\begin{frame}
-\frametitle{Targeted Moments}
-\begin{tabular}{ l r r}\hline
-Target & Model & Data \\\hline
-\footnotesize\textit{Never married by (25,35,40), \% }& $(61,25,19)$ &  $(68,26,19)$ \\
-\footnotesize \textit{One marriage by 40, \%}  & $70$ & $65$ \\\hline
-\footnotesize \textit{No children by (25,35,40), \%}  & $(65,27,24)$ & $(67,28,26)$ \\
-\footnotesize \textit{... by 30 (low-$z$,high-$z$), \%}  & $(52,57)$ & $(44,58)$ \\\hline
-\footnotesize \textit{Divorced by (25,35,40), \%}  & $(5,11,16)$ & $(11,16,19)$ \\
-\footnotesize \textit{... by 30 (low-$z$,high-$z$), \%}  & $(17,9)$ & $(21,8)$ \\\hline
-\footnotesize \textit{Divorced by (30,35,40), \%}  & $(5,11,16)$ & $(11,16,19)$ \\
-\footnotesize \textit{... by 40 (KF,MF), \%}  & $(25,15)$ & $(24,13)$ \\\hline
-\footnotesize \textit{Money for childcare ($\%$ tot expend)} & $43.4$ & 40.0 \\
-\footnotesize \textit{Time spent low-$z$/high-$z$ (ratio)} & 1.11 & 1.19 \\
-\footnotesize \textit{Money spent low-$z$/high-$z$ (ratio)} & 2.56  & 2.67 \\\hline
-
-\end{tabular}
-\end{frame}
-
-\begin{frame}
-\frametitle{Model Allows for Eliminating Selection}
-
-\begin{itemize}
-\item Generate $5{,}000$ individuals from unconditional distribution
-\item No one meets partners till 25.
-\item At 25, two scenarios. All of them:
-\begin{itemize}
-\item Meet and become pregnant --- KF if agree
-\item Meet and do not become pregnant --- MF if agree
-\end{itemize}
-\item Look at their outcomes over time
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{Causal Impact $\approx$ Observable Impact}
-\begin{center}
-\includegraphics[width=0.85\linewidth]{../main/individual-av.eps}
-\end{center}
-\end{frame}
-
-\begin{frame}
-\frametitle{Lower Child Investment Even without Selection}
-\vspace{-2cm}
-\begin{center}
-\includegraphics[width=1.0\linewidth]{../main/x-time.pdf}
-\end{center}
-\end{frame}
-
-\begin{frame}
-\frametitle{``Compliers'' Drive Everything}
-\begin{center}
-\includegraphics[width=0.85\linewidth]{../main/ind-decomp.eps}
-\end{center}
-\end{frame}
-
-\newcommand{\pz}{\phantom{0}}
-\begin{frame}
-\frametitle{More Impact In Early Life}
-\begin{center}
-\begin{tabular}{ l c c c}\hline 
-& &  \multicolumn{2}{ c }{\scriptsize  No Unplanned Pregnancies}  \\
-& \scriptsize Baseline & \scriptsize Outside Marriage & \scriptsize At all \\\hline
-\footnotesize Divorced by 25, \% & $\pz4.5$   & $\pz3.2$ & $\pz2.3$ \\
-\footnotesize Divorced by 30 (low-$z$), \%  & $17.2$ & $14.6$   & $11.9$ \\
-\footnotesize Divorced by 30 (high-$z$), \%  & $\pz9.0$ & $\pz8.2$  &  $\pz7.4$ \\
-\footnotesize Divorced by 35, \%  & $10.8$  & $\pz9.9$&   $\pz9.2$  \\
-\footnotesize Divorced by 40, \%  & $16.3$ & $15.4$ &  $14.8$ \\\hline
-\end{tabular}
-\end{center}
-\end{frame}
-
-\begin{frame}
-\frametitle{Conclusions And Plans}
-\begin{itemize}
-\item What we observe about childcare costs generates the pattern
-\item KF couples have lower $\psi$ and divorce more
-\item Unplanned pregnancies account for $1/2$ of early-age divorces
-\item Preparing a new iteration now:
-\begin{itemize}
-\item TBD: partition by income needs to be improved
-\item TBD: decompose by different contributors ($z$, $\psi$, age)
-\end{itemize}
-\item Model allows to account for other stories:
-\begin{itemize}
-\item Labor market discrimination of unmarried mothers
-\item Social stigma against unmarried cohabitation
-\end{itemize}
-\end{itemize}
 
@@
 
+\section{Model Features}
+
+\begin{frame}
+\frametitle{Overview}
+\begin{itemize}
+\item Lifecycle model
+\item Assets ($a$) and productivity shocks ($z^f$, $z^m$)
+\item Endogenous marriage and divorce (limited commitment)
+\item Reduced form marriage market:
+\[p^{\text{meet}}(\text{Age}) \text{ --- meeting a random partner with similar $a$ and $z$}\]
+\item Both endogenous and unplanned fertility:
+\[p^{\text{unplanned}}(\text{Age},z^f) \text{ --- arrival of unplanned kids}\]
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Modeling Shotgun Marriage}
+\begin{center}
+\vspace{-0.75cm}
+\begin{tikzpicture}[every text node part/.style={align=center}, scale=0.75]
+   % Place nodes
+   \node [block] (1) {\large \textbf{People}  \\[-0.5ex]  \large  \textbf{Randomly Meet}};
+   \node [block, below left = 1.5 cm and 0.0 cm of 1] (2) {\textit{Nash} \\[-0.5ex] \textit{Bargaining}};   
+   \node [block, below right = 1.5 cm and 0.0 cm of 1] (3) {\textit{Nash} \\[-0.5ex] \textit{Bargaining}};
+
+   \node [block, below = 1.5 cm of 2] (5) {\textbf{Kids First} \\[-0.5ex] \textbf{Couple}};
+   \node [block, below = 0.8 cm of 3] (7) {No Kids \\[-0.5ex] Couple};
+   \node [block, below = 0.8 cm of 7] (8) {\textbf{Mariage First} \\[-0.5ex] \textbf{Couple}};
+   \draw[->] (1) to [bend left = 20] node [left] {\scriptsize Unplanned \\[-0.75ex] \scriptsize pregnancy} (2);
+   \draw[->] (1) to [bend right = 20] node [right] {\scriptsize No \\[-0.75ex] \scriptsize pregnancy} (3);
+   \draw[->] (2) to node [left] {\scriptsize Agree} node [right] {\scriptsize \textbf{Shotgun} \\[-0.75ex]  \scriptsize \textbf{marriage}} (5);
+   \draw[->] (3) to node [right] {\scriptsize Agree} node [left]  {\scriptsize \textbf{Regular} \\[-0.75ex]  \scriptsize \textbf{marriage}} (7);
+   \draw[->] (7) to node [right] {\scriptsize Kids by choice \\[-0.75ex] \scriptsize or randomly} (8);
+
+\pause
+
+    \node [block, above left   = 1.0 cm and -0.75 cm of 2] (4) {\,\, Single male\,\,};
+    \node [block, above left   = 1.6 cm and -0.75 cm of 2] (4-sm) {Single mother};
+	\node [block, above right   = 1.0 cm and -0.75 cm of 3] (6) {\,\, Single male\,\,};
+    \node [block, above right   = 1.6 cm and -0.75 cm of 3] (6-f) {\,Single female\,};   
+    \draw[->] (2.180) to [bend left = 30] node [left] {\scriptsize Disagree} (4.270);
+    \draw[->] (3.0) to [bend right = 30] node [right] {\scriptsize Disagree} (6.270);
+     \draw[->] (6.180) to [bend left = 30] node [above] {\scriptsize Back}  (1.-12);
+   \draw[->] (6-f.180) to [bend right = 30] node [above] {\scriptsize Back} (1.12);
+   \draw[->,thick,dashed] (4-sm.90) to [bend left = 30] node [above] {\scriptsize Back with delay}  (1.168);
+   \draw[->] (4.0) to [bend right = 30] node [above] {\scriptsize Back} (1.192);
+
+\end{tikzpicture}
+\end{center}
+\end{frame}
+
+\begin{frame}[label=utilities]
+\frametitle{Utility Functions}
+\begin{itemize}
+\item Couple, $\theta$ --- Pareto weight, $\psi$ --- common love shock:
+\[U = \theta \cdot u(c_f) + (1-\theta)\cdot u(c_m) + \psi,\]
+\item Couple + child, $Q$ --- child quality (public good):
+\[U = \theta \cdot \tilde{u}(c_f,Q) + (1-\theta)\cdot \tilde{u}(c_m,Q) + \psi,\]
+\item Child quality: $Q = F(x,l^f)$, $x$ --- money, $l^f$ --- female time
+\item Budget constraint ($C$ reflects returns to scale, $a$ --- assets):
+ \[C(c_f,c_m) + a' + x = R\cdot a + (1-l^f)\cdot W^f_t(z_f) + W^m_t(z_m),\]
+\end{itemize}
+\hyperlink{value-functions}{\beamerbutton{Value Functions}}
+\end{frame}
+
+\begin{frame}
+\frametitle{Marriage}
+\begin{itemize}
+\item People (f \& m) meet, get a draw of love shock $\psi$ and agree to marry if exists a Pareto weight $\theta$ such that
+\[V^{f,\text{yes}}(\underset{\scriptsize +}{\theta\vphantom{\psi}},\underset{\scriptsize +}{\psi}) \geq V^{f,\text{no}}, \ \ \ \ V^{m,\text{yes}}(\underset{\scriptsize  -}{\theta\vphantom{\psi}},\underset{\scriptsize +}{\psi}) \geq V^{m,\text{no}}.\]
+\item Young children require lots of female time
+\item \textbf{Mechanism:} unplanned pregnancy 
+\begin{enumerate}
+\item[(1)] lowers current $V^{f,\text{no}}$: hard to be a single mother
+\item[(2)] increases future $V^{m,\text{yes}}$ and $V^{f,\text{yes}}$: good to have children
+\end{enumerate} 
+$(1) + (2)$ $\Rightarrow$ the couple agrees to marry with lower $\psi$ and $\theta$.
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Divorce}
+\begin{itemize}
+\item Shocks to $\psi$, $z^f$ and $z^m$
+\item People cannot commit not to divorce
+\item $\theta$ may be renegotiated, stays constant if not
+\item Divorce happens if no $\theta$ exist such that
+\[V^{f,\text{together}}(\theta,\psi) \geq V^{f,\text{divorce}}, \ \ \ \ V^{m,\text{togetehr}}(\theta,\psi) \geq V^{m,\text{divorce}}.\]
+\item Couples with lower $\psi$ or more unequal $z$ divorce more!
+\item As children get older they require less time
+\end{itemize}
+\end{frame}
+
+\section{Fit and Results}
+\begin{frame}
+\frametitle{Strategy}
+\begin{itemize}
+\item I use Simulated Method of Moments to calibrate $\sim 16$ parameters:
+\begin{itemize}
+\item Preferences
+\item Probabilities of transitions
+\item Child quality technology
+\end{itemize}
+\item Income paths to match ACS
+\item The target moments are taken from:
+\begin{itemize}
+\item ACS (raw divorce, fertility and marriage shares by groups)
+\item CEX + ATUS (time and money spent on childcare)
+\end{itemize}
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Targeted Moments}
+\begin{tabular}{ l r r}\hline
+Target & Model & Data \\\hline
+\footnotesize\textit{Never married by (25,35,40), \% }& $(61,25,19)$ &  $(68,26,19)$ \\
+\footnotesize \textit{One marriage by 40, \%}  & $70$ & $65$ \\\hline
+\footnotesize \textit{No children by (25,35,40), \%}  & $(65,27,24)$ & $(67,28,26)$ \\
+\footnotesize \textit{... by 30 (low-$z$,high-$z$), \%}  & $(52,57)$ & $(44,58)$ \\\hline
+\footnotesize \textit{Divorced by (25,35,40), \%}  & $(5,11,16)$ & $(11,16,19)$ \\
+\footnotesize \textit{... by 30 (low-$z$,high-$z$), \%}  & $(17,9)$ & $(21,8)$ \\\hline
+\footnotesize \textit{Divorced by (30,35,40), \%}  & $(5,11,16)$ & $(11,16,19)$ \\
+\footnotesize \textit{... by 40 (KF,MF), \%}  & $(25,15)$ & $(24,13)$ \\\hline
+\footnotesize \textit{Money for childcare ($\%$ tot expend)} & $43.4$ & 40.0 \\
+\footnotesize \textit{Time spent low-$z$/high-$z$ (ratio)} & 1.11 & 1.19 \\
+\footnotesize \textit{Money spent low-$z$/high-$z$ (ratio)} & 2.56  & 2.67 \\\hline
+
+\end{tabular}
+\end{frame}
+
+\begin{frame}
+\frametitle{Model Allows for Eliminating Selection}
+
+\begin{itemize}
+\item Generate $5{,}000$ individuals from unconditional distribution
+\item No one meets partners till 25.
+\item At 25, two scenarios. All of them:
+\begin{itemize}
+\item Meet and become pregnant --- KF if agree
+\item Meet and do not become pregnant --- MF if agree
+\end{itemize}
+\item Look at their outcomes over time
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Causal Impact $\approx$ Observable Impact}
+\begin{center}
+\includegraphics[width=0.85\linewidth]{../main/individual-av.eps}
+\end{center}
+\end{frame}
+
+\begin{frame}
+\frametitle{Lower Child Investment Even without Selection2}
+\vspace{-2cm}
+\begin{center}
+\includegraphics[width=1.0\linewidth]{../main/x-time.pdf}
+\end{center}
+\end{frame}
+
+\begin{frame}
+\frametitle{``Compliers'' Drive Everything}
+\begin{center}
+\includegraphics[width=0.85\linewidth]{../main/ind-decomp.eps}
+\end{center}
+\end{frame}
+
+\newcommand{\pz}{\phantom{0}}
+\begin{frame}
+\frametitle{More Impact In Early Life}
+\begin{center}
+\begin{tabular}{ l c c c}\hline 
+& &  \multicolumn{2}{ c }{\scriptsize  No Unplanned Pregnancies}  \\
+& \scriptsize Baseline & \scriptsize Outside Marriage & \scriptsize At all \\\hline
+\footnotesize Divorced by 25, \% & $\pz4.5$   & $\pz3.2$ & $\pz2.3$ \\
+\footnotesize Divorced by 30 (low-$z$), \%  & $17.2$ & $14.6$   & $11.9$ \\
+\footnotesize Divorced by 30 (high-$z$), \%  & $\pz9.0$ & $\pz8.2$  &  $\pz7.4$ \\
+\footnotesize Divorced by 35, \%  & $10.8$  & $\pz9.9$&   $\pz9.2$  \\
+\footnotesize Divorced by 40, \%  & $16.3$ & $15.4$ &  $14.8$ \\\hline
+\end{tabular}
+\end{center}
+\end{frame}
+
+\begin{frame}
+\frametitle{Conclusions And Plans}
+\begin{itemize}
+\item What we observe about childcare costs generates the pattern
+\item KF couples have lower $\psi$ and divorce more
+\item Unplanned pregnancies account for $1/2$ of early-age divorces
+\item Preparing a new iteration now:
+\begin{itemize}
+\item TBD: partition by income needs to be improved
+\item TBD: decompose by different contributors ($z$, $\psi$, age)
+\end{itemize}
+\item Model allows to account for other stories:
+\begin{itemize}
+\item Labor market discrimination of unmarried mothers
+\item Social stigma against unmarried cohabitation
+\item ...
+\end{itemize}
+\end{itemize}
+
+\end{frame}
+
 \backupbegin
@@
 \section{Appendix}
-
-\begin{frame}[label=pic0]
-\frametitle{In One Picture (Old \& Extended Version)}
-\begin{center}
-\includegraphics[width=0.85\linewidth]{diff_dt_h.pdf}
-\end{center}
-\end{frame}
 
@@
 {\scriptsize
-\begin{align}& \hspace{5em}  V^{c}_t(\Omega) = \max\limits_{c^f,c^m}  \bigg\{ \theta^f \cdot u(c^f) + \theta^m \cdot u(c^m) + \psi + \nonumber \\   \nonumber
+\begin{align}& \hspace{5em}  V^{c}_t(\Omega) = \max\limits_{c^f,c^m}  \bigg\{ \theta\cdot u(c^f) + (1-\theta)\cdot u(c^m) + \psi + \nonumber \\   \nonumber
  &  \beta \E_t \Big[   (1 - p^{\text{preg}}_t)\cdot \left\{ (1-d^{\text{np}})\cdot \max\left\{ V^{ck}_{t+1},V^{c}_{t+1}\right\} + d^{\text{np}}\cdot [ \theta V_{t+1}^{f,s} + (1-\theta)V_{t+1}^{m,s}]\right\}  +  \\  \nonumber
@@
 {\scriptsize
-\begin{align}& \hspace{5em}  V^{ck}_t(\Omega) = \max\limits_{c^f,c^m,q,l_f,x}  \bigg\{ \theta^f\cdot u(c^f) + \theta^m \cdot u(c^m) + u^{\text{kid}}(q) +  \psi + \nonumber \\  \nonumber
- & \hspace{8em} \beta \E_t \Big[   \left\{ (1-d)\cdot   V^{ck}_{t+1}(\Omega') + d\cdot [ \theta^f V_{t+1}^{f,sk}(\omega^{df}) + \theta^m V_{t+1}^{m,s}(\omega^{dm})]\right\} \Big] \bigg\},
+\begin{align}& \hspace{5em}  V^{ck}_t(\Omega) = \max\limits_{c^f,c^m,q,l_f,x}  \bigg\{ \theta\cdot u(c^f,q) + (1-\theta)\cdot u(c^m,q) + \psi + \nonumber \\  \nonumber
+ & \hspace{8em} \beta \E_t \Big[   \left\{ (1-d)\cdot   V^{ck}_{t+1}(\Omega') + d\cdot [ \theta V_{t+1}^{f,sk}(\omega^{df}) + (1-\theta)V_{t+1}^{m,s}(\omega^{dm})]\right\} \Big] \bigg\},
 \end{align}\vspace{-2em}
 \begin{align*}
- & a' + c + x = R\cdot a  + W^m_t(z^m) + l_f \cdot W^f_t(z^f) & \text{(evolution of joint assets)},\\
+ & a' + c + x = R\cdot a  + W^m_t(z^m) + (1-l_f)\cdot W^f_t(z^f) & \text{(evolution of joint assets)},\\
                     & c = [(c^f)^{1+\rho_c} + (c^m)^{1+\rho_c}]^{\frac1{1+\rho_c}} & \text{(increasing returns in consumption)},\\
-                    & q = [ x^{\lambda} + \kappa \cdot (1-l_f)^{\lambda} ]^{\frac1{\lambda}}  & \text{(production function of child quality)},\\
+                    & q = [\mu_\xi\cdot l_f^{\theta_q} + (1-\mu_\xi)\cdot x^{\theta_q}]^{\frac1{\theta_q}} & \text{(production function of child quality)},\\
                     & q \geq \underline{q} & \text{(required childcare costs)},\\
-                    &  z^{f\prime} = z^f + \varepsilon^{z,f}_t - f(l), \ \ \varepsilon^{z,f}_t \sim \mathcal{N}(0;\sigma_{z,f}^2) &  \text{ (evolution of female productivity)}\\
+                    &  z^{f\prime} = z^f + \varepsilon^{z,f}_t, \ \ \varepsilon^{z,f}_t \sim \mathcal{N}(0;\sigma_{z,f}^2) &  \text{ (evolution of female productivity)}\\
 				 &  z^{m\prime} = z^m + \varepsilon^{z,m}_t, \ \ \varepsilon^{z,m}_t \sim \mathcal{N}(0;\sigma_{z,m}^2) &  \text{ (evolution of male productivity)}\\
                     & \psi' = \psi + \varepsilon^{\psi}_t, \ \ \varepsilon^{\psi}_t \sim \mathcal{N}(0;\sigma_{\psi}^2)  & \text{(evolution of marriage surplus),} \\
-                    & (\Omega',\omega^{df},\omega^{dm}) = \mathcal{R}(\theta^f,\theta^m,,a',z^{m\prime},z^{f\prime},\psi',\xi) & \text{(renegotiation correspondence)},
+                   &  P(\xi' = s | \xi = y) = p^s, \ \ P(\xi' = s | \xi = s) = 1 & \text{(evolution of child's age)},\\
+                    & (\Omega',\omega^{df},\omega^{dm}) = \mathcal{R}(\theta,a',z^{m\prime},z^{f\prime},\psi',\xi) & \text{(renegotiation correspondence)},
 \end{align*}
@@
  & U(c^f,c^m) = \theta\cdot \frac{(c^f)^{1-\sigma}}{1-\sigma} + (1-\theta)\cdot \frac{(c^m)^{1-\sigma}}{1-\sigma} \rightsquigarrow A(\theta) \frac{c^{1-\sigma}}{1-\sigma}  & \text{(couple)},                 \\
- & \tilde{u}(c) = \frac{c^{1-\sigma}}{1-\sigma}  + \phi_0 + \frac{Q^{1-\sigma}}{1-\sigma} & \text{(individual + child)},\\
- & \tilde{u}(c) = \theta \frac{(c^f)^{1-\sigma}}{1-\sigma}  +(1- \theta) \frac{(c^m)^{1-\sigma}}{1-\sigma}  + \phi_0 + \frac{Q^{1-\sigma}}{1-\sigma} & \text{(couple + child)}.
+ & \tilde{u}(c,Q) = \frac{\left[c^{\phi_1} (\phi_2 + Q)^{1-\phi_1} \right]^{1-\sigma}}{1-\sigma} + \phi_0 & \text{(individual + child)},\\
+ & \tilde{U}(c^f,c^m,Q) \rightsquigarrow \tilde{A}(\theta)  \frac{\left[c^{\phi_1} (\phi_2 + Q)^{1-\phi_1} \right]^{1-\sigma}}{1-\sigma} + \phi_0 & \text{(couple + child)}.
 \end{align*}

### 2020-10-02 slides-march-20/slides.tex
Current stage

\documentclass{beamer}
\usepackage[english]{babel}
\usepackage{amsmath,amsfonts}
\usepackage{multicol}

\usepackage{IEEEtrantools}
\usepackage{multirow}
% beamer setup

\usetheme{default}
\usecolortheme{seahorse}
%\usecolortheme{dove}

\setbeamertemplate{navigation symbols}{}

\usepackage{tikz}
\usetikzlibrary{shapes,arrows}
\usetikzlibrary{positioning}
\tikzstyle{block} = [rectangle, draw, rounded corners]
\tikzstyle{line} = [draw, -latex']

\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator*{\argmax}{argmax}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\I}{\mathbb{I}}

\AtBeginSection[]{
  \begin{frame}[plain]
  \addtocounter{framenumber}{-1}
  \vfill
  \centering
  %\begin{beamercolorbox}[sep=8pt,center,shadow=true,rounded=true]{title}
    %\usebeamerfont{title}
    \Huge{\insertsectionhead\par}%
  %\end{beamercolorbox}
  \vfill
  \end{frame}
}

\newcommand{\backupbegin}{
   \newcounter{framenumberappendix}
   \setcounter{framenumberappendix}{\value{framenumber}}
}
\newcommand{\backupend}{
   \addtocounter{framenumberappendix}{-\value{framenumber}}
   \addtocounter{framenumber}{\value{framenumberappendix}} 
}

\title{The Economics of Shotgun Marriage}
\subtitle{and Household Bargaining}
\author{Egor Kozlov}

\institute{
  Department of Economics\\
  Northwestern University}

\setbeamertemplate{footline}[frame number]

%  \usepackage{pgf}
%\logo{\pgfputat{\pgfxy(0,0)}{\pgfbox[right,base]{\footnotesize{\insertframenumber\,/\,\inserttotalframenumber}}}}
%\newcommand{\nologo}{\setbeamertemplate{logo}{}}

\begin{document}

\begin{frame}[plain]
\addtocounter{framenumber}{-1}
\date{\scriptsize}
\titlepage
\end{frame}

\begin{frame}[plain]
\frametitle{Question}
\begin{itemize}

\item Why people marry? What part of couple's value come from having children together?

\item People get pregnant and choose to marry each other. Their marriage performs worse yet they choose to enter it. Rational reasons?

\end{itemize}
\end{frame}

\begin{frame}[plain]
\frametitle{What I Do}
\begin{itemize}

\item Quantify difference of marriage performance by the timing of childbirth

\item Use a structural model matching these difference to understand sources of it

\item Quantify importance of fertility for marriage, aggregate impact of unplanned childbirths and policy counterfactuals

\end{itemize}
\end{frame}

\begin{frame}[plain]
\frametitle{Preview of Results}

\begin{enumerate}

\item Data: 
\begin{itemize}
\item Kids before marriage $\Rightarrow$ divorce more often.
\item College graduates: $\sim$ 3 times. Non-college: $\sim$ 1.2 times.
\item Robust yet not causal
\end{itemize}

\item Quantifying:
\begin{itemize}
\item Unplanned pregnancy changes outside options in family bargaining
\item Effect is present generically, its size identifies amount of distress
\end{itemize}

\item Exploring:
\begin{itemize}
\item 
\end{itemize}

\end{enumerate}
\end{frame}

\section{Identification}

\begin{frame}
\frametitle{Marriage}
Distribution of partners:
\begin{itemize}
\item $z^{\text{part}} = \mu_z +z^{\text{own}} + \sigma^{z,p} \cdot \epsilon_{z,p}$

target: regression of spouse's log-wage to own log-wage
\item $\log a^{\text{part}} = \mu_a + \log a^{\text{own}} + \sigma^{a,p} \cdot \epsilon_{a,p}$

target: $\mu_a = 0.2$ suggested by ..., small fixed $\sigma^{a,p} = 0.1$
\end{itemize}

Meeting process:
\begin{itemize}
\item $\psi_0 \sim \mathcal{N}(0,\sigma^2_{\psi,0})$. $\sigma^2_{\psi,0}$ estimated
\item $p^{\text{meet}} = g_0 + g_1\cdot \text{Age}$, $g_1<0$. $g_0$, $g_1$ estimated
\item Identifying variation: share of never married by age, share of re-married at 
\end{itemize}
\end{frame}

\begin{itemize}
\item Identified 
\end{itemize}

\begin{frame}
\frametitle{Updates Since November}
\begin{itemize}
\item Brand new codes in Python
\item Remarriage of single mothers
\item Skills depreciation because of time to childcare
\item No unplanned pregnancies in marriage: restrictive but cleaner identification
\item Two separate estimations: college grads versus non-college grads
\item Decomposition exercises as a goal
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Summary of Data Part}
\begin{itemize}
\item Kids First: those who had kids before or at the year of marriage
\item Marriage First: those who had kids at least at the next year
\item Sharp difference in divorce rates, robust and large
\item Much larger for college graduates
\end{itemize}
\begin{center}
\begin{tabular}{|c|c|c|c|}
& Kids First & Marriage First & Ratio \\
College & & & \\
High School & & & \\\hline
\end{tabular}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Generic Feature?}
\begin{itemize}
\item Suppose that people:
\begin{itemize}
\item Can choose fertility 
\item But sometimgs kids arrive randomly
\end{itemize}
\item Then, negative impact of unplanned pregnancy is generic
\item If people choose not to have a kid in this period they are worse off b/c of arrival of the shock
\item How large is the effect depends on how indifferent are people between different fertility timings
\item We don't know how bad people feel
\item But this makes sense from prospective of college--non-college partition
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Generic Feature?}
\begin{itemize}
\item Bargaining model: if stress is symmetric things don't happen
\item Stress has to be asymmetric: females suffer more
\item College grads suffer more if they care more about fertility timing
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Theory}
\begin{itemize}
\item Why do people enter low quality marriage when kids happen?
\begin{itemize}
\item Women: 
\begin{itemize}
\item splitting childcare costs
\item worsening of marital prospectives otherwise
\end{itemize}
\item Men:
\begin{itemize}
\item access to their children
\end{itemize}
\end{itemize}
\item Why are college people hurt more by unplanned pregnancy?
\begin{itemize}
\item They are more sensitive to fertility timing
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Model-Free Evidence}
\begin{itemize}
\item \textit{If people are free to choose their fertility, arrival of a child unexpectedly makes them worse off}
\item Couple is together if
\[V^{fM} \geq V^{fS} \& V^{mM} \geq V^{mS}.\]
\item People worse off $\Rightarrow$ more divorce? Not necessary, as baby enters both $V^{\bullet M}$ and $V^{\bullet S}$. 
\item If values decrease asymmetrically, this can happen
\item Size of the asymmetry defines the size of the effect
\item Perfect child support enforcement = ?
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Planned Fertility}
I estimate $(\phi,\alpha,\kappa,\bar{Q})$

\[u_{\text{child}} = \phi + \alpha \cdot \frac{Q^{1-\sigma}}{1-\sigma},\]

\[Q = \left[x^{\lambda} + \kappa \cdot (1-l_f)^{\lambda}\right]^{\frac1{\lambda}}, \ \ Q\geq \bar{Q},\]

\[\text{budget constraint}: \ \ C + x = R\cdot a + W^m_t + W^f_t \cdot l_f.\]

Identification:
\begin{itemize}
\item $\phi$ --- total fertility level
\item $\alpha$ --- age structure (older $=$ more money)
\item $(\bar{Q},\kappa)$ are tricky:
\begin{itemize}
\item Share of earnings spent on $x$ 
\item \textit{Fertility by spouse's income}
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Unplanned Fertility \& Meetings}
$p^{\text{preg}} = p_0 + p_1\cdot \text{Age}$.
Identification:
\begin{itemize}
\item Share of kids first couples at 25, 30, 35 (stock): $p_0$
\item \% who married and had kids exactly at age 25, 30, 25: $p_1$
\end{itemize}
Concern: the data suggests $p_1>0$.
\end{frame}

\begin{frame}
\frametitle{Income Process}
\framesubtitle{$i \in \{m,f\}$}
\[W^i_t = \exp(z^i_t + T^i_t)\]

\begin{enumerate}
\item Trend: $ T^i_t = a^i_0 + a^i_1\cdot t +  a^i_2\cdot t^2 + a^i_2\cdot t^3.$
\begin{itemize}
\item Estimate based on cross-section per-hour wage in ACS with state and year dummies
\item Normalize wage at 25 to 1
\end{itemize}
\item Shocks: $z^i_t = z^m_{t-1} + \varepsilon^i_t$; $z^m_0 \sim \mathcal{N}(0;\sigma^2_{zi0})$; $\varepsilon^m_t \sim \mathcal{N}(0;\sigma^2_{zi})$.
\begin{itemize}
\item Pick $\sigma^2_{zi0}$ and  $\sigma^2_{zi}$ to match cross-sectional variance in log earnings at 25 and 30.
\item Outliers matter a lot: remove top and bottom $5\%$ of per-hour earnings
\item Male: $\sigma_{zm0} = 0.44$, $\sigma_{zm} = 0.033$.
\item Female: $\sigma_{zf0} = 0.41$, $\sigma_{zm} = 0.042$. \textbf{Rescale} to match the targets.
\item 3\%--4\% wage variation seems reasonable...
\item Skill depreciation
\end{itemize}
\end{enumerate}
\end{frame}

\begin{frame}
\frametitle{Labor Supply Margin}
\begin{itemize}
\item Females can reduce labor supply and use as input to child quality process
\item $Q = \left[x^{\lambda} + \kappa\cdot (1-l)^{\lambda}\right]^{\frac1\lambda}$
\item Labor supply affects current income:
\[c + x  = W^m + W^f\cdot l + R a.\]
\item Reduction in labor supply hurts female's productivity:
\[z^f_t = z^f_{t-1} + \varepsilon^f_{t} - f(l)\cdot \mu,\]
\item Currently $l$ is discrete, $l \in \{0.25,0.5,0.75,0.9\}$, corresponding $f(l) \in \{0.1,0.05,0,0\}$. $0.1$ roughly matches Adda et al (...)
\item $\kappa$ is calibrated such that average married female's labor supply is $0.75$.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Remarriage}
\begin{itemize}
\item If woman with kids divorces, it becomes a single mother
\item Single mothers meet single males
\item If partners agree to marry they become a regular couple with a child
\item Males have a one-time utility penalty parameter for marrying a single mother. This affects bargaining:
\[NBS = \left[ V^{m,MK} - \phi_m - V^{m,S}\right] \times  \left[ V^{f,MK} - V^{f,S}\right].\]
\item Rationalizes decrease in marital prospectives for females with children
\item Estimated by matching share of remarried women with vs without children
\end{itemize}
\end{frame}

\section{Results}

\begin{frame}
\frametitle{Causal Consequences of Unplanned Pregnancy}
Graphs: divorce probabilities, labor supply / child's expenditures, earnings loss
\end{frame}

\begin{frame}
\frametitle{Decomposing College versus Non-college}
\begin{center}
\begin{tabular}{|c|c|c|c|}
& MF & KF & Ratio \\
Baseline: college & \\
+ income process of HS & \\
+ meeting probabilities of HS & \\
+ pregnancy probabilities of HS & \\
+ love shock distribution of HS & \\
+ single mother prospective of HS & \\
Comparison: high school & \\
\end{tabular}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Understanding Value of Marriage}
\begin{tabular}{|c|c|c|c|c|c|}
Married by 25 & \\
Married by 30 & \\
Married by 40 & \\
Divorced by 25 &\\
Divorced by 30 &\\
DIvorced by 40 &\\
\multicolumn{4}{|c|}{Model Partss:}\\
Unplanned fertility & $+$ & $-$ & $-$ & $-$ & $-$ \\
Fertility & $+$ &  $+$ & $-$ & $-$  & $-$ \\
Love & $+$ & $+$ & $+$ & $-$ & $-$ \\
Returns to scale & $+$ & $+$ & $+$ & $+$ & $-$ \\ 
Risk-sharing & $+$ & $+$ & $+$ & $+$ & $+$ \\
\end{tabular}
\end{frame}

\begin{frame}
\frametitle{Hey}
\[\lim\limits_{n\to\infty} |f_n(\theta)| = 0 \text{a.s.}, \ \ \text{ but } \lim\limits_{n\to\infty}\sup_\theta |f_n(\theta)| \neq 0,\]

\[|Y - bX|'_b = -X\cdot 1 \cdot I(Y - bX>0) +X \cdot 1 \cdot I(Y - bX<0)  = \]
\[-X\cdot 1 \cdot I(Y - bX>0) + X\cdot 1 (1 - I(Y - bX>0))\]

\[(|Y - bX|^5)'_b = (Y-bX)^4 \cdot |Y - bX|'_b\]

\[ \E( |Y - bX|'_b ) = 0 \Rightarrow \E( -X\cdot 1 \cdot I(Y - bX>0) +X \cdot 1 \cdot I(Y - bX<0) ) = 0\]
\[ \E(X\cdot I(A) ) = \E(X|A)\cdot P(A).\]

\end{frame}

\begin{frame}
\frametitle{Consistency Theorem}
\[Q_n(\theta) \text{ (sums) } \to Q(\theta)  \text{ (expectations) }\]
show that
\[Q'(\theta) = 0 \text{ at } \theta = \theta_0\]
\end{frame}

\end{document}

### 2020-10-02 slides-may-20/slides.tex
Current stage

--- 2020-10-02 slides-aug-20/slides.tex
@@
 
-% REMOVE SLIDE COUNT
-
 \begin{frame}[plain]
@@
 \begin{frame}
-\frametitle{``America's strongest anti-poverty weapon''?}
+\frametitle{Two-Parent Family Is A Choice}
+
+\textbf{Shotgun marriage} --- marriage that happens soon after a woman gets pregnant.
+% many, much less stable, worse among the educated. pt is economics how they get formed, how do the
 \begin{itemize}
-\item Married two-parent families \ : \ desirable outcomes for parents and children%$\genfrac{}{}{0pt}{1}{\Leftarrow}{\Rightarrow}$
-\item Pro-marriage policies: child support, tax brackets, divorce regulations, etc
-\item Can economic policies create successful marriages? Love vs money.
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-
-\frametitle{Kids-First Marriages are Less Stable}
-\begin{enumerate}
-\item{\textbf{Shotgun marriage}} --- couple marrying when woman gets pregnant
-\item{\textbf{Kids-first marriage}} --- child \textit{before or in the year} of marriage
-\end{enumerate}
-\begin{itemize}
-\item Around 1/4 marriages in the US today are kids-first
-\item They have much higher divorce rates, especially among college graduates
-\item Marginal in economic sense: could be sensitive to policies 
+\item Around 1/4 couples in the US have kids \textbf{before} or \textbf{at the year} of their marriage
+\item These marriages are significantly less stable, especially among \textbf{educated people} % marginal marriage
+\item Both social norms and economics are expected to drive this
+%\item Especially less stable among \textbf{more educated and adult}%(college: 6\% single mothers, 4\% in shotgun marriages)
+\item Overall, they contribute to creation of single mothers and blended families
+% college grads: 4% in KF marriages and 4% are single mothers
+% high school grads: 11% in KF marriages and 20% are single mothers
 \end{itemize}
@@
 \begin{itemize}
-\item dynamic bargaining and limited commitment in marriage
+\item limited commitment in marriage
 \item lifecycle, income and savings
@@
 \end{itemize}
-\item Look at what drives decisions to marry and the differences in divorce
+\item Look at what drives the difference in marriage performance
 \begin{itemize}
 \item what explains the education gradient?
-\item can policies fix the difference or reduce the share?
+\item should policies bother to fix the difference?
 \item what is the value of marriage in general?
@@
 \item My structural model corrects selection and data imperfections
-%\item Model results:
-%\begin{itemize}
-%\item Social stigma + threat of raising the child alone explains most of the difference
-%\item Child support $\Rightarrow$ more single mothers, little effect on marriage decisions
-%\item Being able to control fertility helps everyone
-%\end{itemize}
-\item Model answers: Love $>$ Money. In particular:
+\item Model answers:
 \begin{itemize}
-\item Threat of raising the child alone explains most of the difference
-\item Regulating marriages by policies is both hard and inefficient
-\item Easier control over fertility fixes shotgun marriages and has bigger return
+%\item Causal content: kids first people divorce X times more
+\item Social stigma explains 40\%--50\% of the divorce differences for college people
+\item Expensive childcare explains the rest
+\item Stigma matters less for high school graduates
 \end{itemize}
@@
 \begin{itemize}
-\item Policy literature:
-\begin{itemize}
-\item Empirical: Alessina, Guiliano, 2005;  Tannenbaum, 2020; Rossin-Slater, 2017
-\item Structural: Forester, 2020; Brown, Flinn, 2011
-\item Focus on particular policies, ignore interactions, limited quantitative power
-\end{itemize}
-\item Lifecycle literature:
-\begin{itemize}
-\item Dynamic limited commitment for marriage and divorce: Mazzocco, 2007; Voena, 2015; Low et al., 2018; Shephard, 2019; Blasutto, 2020
-\item Fertility during lifecycle: Sommer, 2015, Ejrnæs, Jørgensen, 2018
-\item Link between divorce and fertility is missing
-\end{itemize}
-\item Empirics on shotgun marriages (demography and sociology):
+\item Empirical literature documents the scope of shotgun marriage around the world:
 \begin{itemize}
@@
 \end{itemize}
+\item Dynamic limited commitment models of divorce and marriage:
+\begin{itemize}
+\item Marcet, Marimon (2019), Mazzocco (2007) --- setting up the framework
+\item Voena (2015), Shephard (2019); Lafortune, Low  (2019) --- applied questions
+\item Mostly ignore or simplify interaction with fertility
 \end{itemize}
-%
-%\begin{itemize}
-%\item Empirical literature documents the scope of shotgun marriage around the world:
-%\begin{itemize}
-%\item Gibson-Davis et al (2016), Guzzo (2009),  Nuevo-Chiquero (2014)
-%\item Mostly descriptive
-%\end{itemize}
-%\item Dynamic limited commitment models of divorce and marriage:
-%\begin{itemize}
-%\item Marcet, Marimon (2019), Mazzocco (2007) --- setting up the framework
-%\item Voena (2015), Shephard (2019); Lafortune, Low  (2019) --- applied questions
-%\item Mostly ignore or simplify interaction with fertility
-%\end{itemize}
-%\item Unplanned and teenage pregnancy literature:
-%\begin{itemize}
-%\item Ejrnæs, Jørgensen (2018) --- lifecycle model
-%\item Card, Wise (1978) to Ashcraft et al (2013) --- lots of empirical papers
-%\item Treat household structure as fixed or exogenous
-%\end{itemize}
-%\end{itemize}
+\item Unplanned and teenage pregnancy literature:
+\begin{itemize}
+\item Ejrnæs, Jørgensen (2018) --- lifecycle model
+\item Card, Wise (1978) to Ashcraft et al (2013) --- lots of empirical papers
+\item Treat household structure as fixed or exogenous
+\end{itemize}
+\end{itemize}
 \end{frame}
@@
 \begin{frame}
-\frametitle{Sources: ACS 2009--2017 and others}
+\frametitle{Sources: ACS 2009--2015 and others}
 \begin{itemize}
 \item Main exercises and model estimation are done on American Community Survey
-%\item Huge sample size and enough information to show many things 
+\item Huge sample size and enough information to show many things 
 \item Individual marriage and fertility history are observed partially
@@
 \begin{itemize}
-\item Survey of Income and Program Participation (SIPP)
-\item More: NSFG and NSFH
+\item SIPP, NSFG, NSFH --- all have similar patterns
 \end{itemize}
@@
 \begin{frame}[label=dt_graph]
-\frametitle{Kids First Women Are Divorced More Often: Just Look}
+\frametitle{Kids First Women Are Divorced More Often: Key Graph}
 \begin{center}
@@
 \frametitle{Kids First Women Are Divorced More Often: Quantification}
-\framesubtitle{ACS 2009--2017, Women 21--40, married once}
 \begin{center}
-\begin{tabular}{l r r r r }
-\hline
-& \multicolumn{2}{c}{share of divorced if ... }&  \\
-&  marriage-first & kids-first & (share of kids-first) &  \\\hline
-\multicolumn{5}{l}{\textbf{All sample}} \\\hline
-\textit{Cross-sectional share of divorced} &  9.9 & 18.1 & (26.1) \\
-\textit{Divorced 5 years after marriage} &  5.1 & 14.3  & (24.6) \\
-\textit{Divorced 10 years after marriage} & 10.4 & 22.8 & (24.4) \\\hline
-\multicolumn{5}{l}{\textbf{Cross-sectional share of divorced for subsamples}} \\\hline
-\textit{High school only} &  12.8 & 17.3 & (37.0) \\
-\textit{Some college} & 14.3 & 21.0 & (31.3) \\
-\textit{College or more} &   5.3 & 14.8 & (11.9) \\\hline
-\textit{First birth before 25} & 15.5 & 19.9 & (41.8) \\
-\textit{First birth at 25 or later} &  6.2 & 10.9 & (10.7)  \\\hline
+\begin{tabular}{l c c c c c}\hline
+& &  \multicolumn{2}{c}{\small \% divorced if ...} \\
+& \% KF & MF & KF & \% diff &  Ratio \\\hline
+\multicolumn{6}{c}{\textit{Cross-Section: All Females 21--40 married once}}\\
+All females & 23.2 &  10.2 & 18.1 & 7.9 & 1.77  \\
+College & 10.0 & 5.4 & 14.8 & 9.4 & 2.76 \\
+High school & 34.3 & 13.9 & 17.3 & 3.3 & 1.24 \\\hline
+\multicolumn{6}{c}{\textit{Controlling strategies}}\\\hline
+%\multicolumn{2}{l}{College 5 y after marriage} & 5.1 & 14.3 & 9.2 & 2.79 \\
+\multicolumn{4}{l}{Regression with controls}   & 7.0 & 1.69 \\
+\multicolumn{4}{l}{Regression with controls + income}   & 6.1 & 1.60 \\\hline
+\multicolumn{6}{p{0.7\linewidth}}{\footnotesize Regression: $\text{Divorced}_i = \Delta \cdot \text{KF}_i + X_i\gamma + \varepsilon_i$ (if $\text{KF}_i = 1$ or $\text{MF}_i = 1$)}\\
+\multicolumn{6}{p{0.7\linewidth}}{\footnotesize Controls are two-way FE for interactions of age with education, time since marriage, time since birth, and one-way FE for geography, race, survey year. 
+Income is 3rd degree polynomial of log per-hour reported income for working $\geq 10$ hours per week and earning $\geq \$3000$ per year.
+}\\
+\multicolumn{6}{p{0.7\linewidth}}{\footnotesize Data: American Community Survey 2009--2016.}\\\hline
 \end{tabular}
@@
 \begin{frame}
-\frametitle{Kids First Women Are Divorced More Often: Controls}
-\begin{center}
-%\vspace{-0.5cm}
-\begin{tabular}{l r r r }
-\multicolumn{4}{c}{\textit{Regression equation:} $\text{Divorced}_i = \Delta \cdot \text{kids-first}_i + \text{Controls}_i + \varepsilon_i$} \\\hline
-\hline
-& \multicolumn{2}{c}{Estimates}  \\
-& Difference ($\Delta$) & (standard error)  \\\hline
-\multicolumn{4}{l}{\textbf{All sample, regression}} \\\hline
-\textit{No controls (raw difference) }& 8.2 & (0.1)  \\
-\textit{Demographic controls }& 5.7 & (0.1)  \\
-\textit{Duration controls} &  5.8 & (0.2)  \\
-\textit{Demographic + duration} &  4.8 & (0.2)  \\
-\textit{Demographic + duration + income} & 4.1 & (0.3)  \\\hline
-\multicolumn{4}{l}{\textbf{All sample, propensity score matching}} \\\hline
-\textit{Demographic controls} & 5.0 & (0.3)  \\
-\textit{Demographic controls + income} & 4.9 & (0.2)  \\\hline
-\multicolumn{4}{p{0.825\linewidth}}{\footnotesize \textit{Controls:} demographics: FE for age $\times$ education, race, state, survey year.
-Duration: FE for age $\times$ age of the first marriage and age $\times$ age of the first birth.
-Income: 3rd degree polynomial of log gross income (if observed).}\\\hline\hline
-\end{tabular}
-\end{center}
+\frametitle{Kids First Women Are Divorced More Often: Robustness \& Extra}
+\begin{itemize}
+\item SIPP exercise: smaller sample but full marital history and fertility history
+\begin{itemize}
+\item Similar ratios with larger magnitudes when include remarried people
+\end{itemize}
+\item Duration model: NSFH, historic data from 80s
+\begin{itemize}
+\item Hazard ratio of 1.45, college: 2.23
+\end{itemize}
+\item More details: NSFG, 2015--2017 data
+\begin{itemize}
+\item KF: 30\% true shotgun, 50\% marry after birth, 20\% marry new partner
+\item True shotgun have the highest share of divorced
+\end{itemize}
+\end{itemize}
 \end{frame}
-
-\begin{frame}
-\frametitle{Kids First Women Are Divorced More Often: Extra Evidence from SIPP}
-\begin{center}
-\begin{tabular}{l r r }
-\hline
-\multicolumn{3}{l}{\textbf{Including remarried people...}} \\\hline
-& \multicolumn{2}{c}{share of divorced if ... }  \\
-&  marriage-first & kids-first   \\\hline
-\textit{Ever divorced} & 34.4 & 48.0  \\
-\textit{Divorced if married once} &  12.4 & 19.3 \\
-\textit{Remarried} &  22.0 & 28.7 \\\hline
-\hline
-\multicolumn{3}{l}{\textbf{Separating step and own children (2+ kids only)}} \\\hline
-& share in sample & share of divorced   \\\hline
-{Marriage-first} & 72.1 & 10.5 \\
-{Kids-first-own} &  16.4 & 14.6 \\
-{Kids-first-other} &  11.4 & 27.2 \\\hline
-\multicolumn{3}{p{0.7\linewidth}}{\footnotesize \textit{Notes.} SIPP data, the cross-section of Wave 1, 2014, women 21+. Numbers are percents.}\\\hline\hline
-\end{tabular}
-\end{center}
-\end{frame}
-
-%\begin{frame}
-%\frametitle{Kids First Women Are Divorced More Often: Robustness \& Extra}
-%\begin{itemize}
-%\item SIPP exercise: smaller sample but full marital history and fertility history
-%\begin{itemize}
-%\item Similar ratios with larger magnitudes when include remarried people
-%\end{itemize}
-%\item Duration model: NSFH, historic data from 80s
-%\begin{itemize}
-%\item Hazard ratio of 1.45, college: 2.23
-%\end{itemize}
-%\item More details: NSFG, 2015--2017 data
-%\begin{itemize}
-%\item KF: 30\% true shotgun, 50\% marry after birth, 20\% marry new partner
-%\item True shotgun have the highest share of divorced
-%\end{itemize}
-%\end{itemize}
-%\end{frame}
 %
@@
 \begin{center}
-\vspace{-0.6cm}
+\vspace{-0.2cm}
 \begin{tikzpicture}[every text node part/.style={align=center}, scale=0.75]
@@
     \node [block, above left   = 1.8 cm and -0.75 cm of 2] (4-sm) {\,Single mom \\[-0.5ex]  ($f,sk$)};
-    %\node [block, left = 0.2 cm of 4-sm] (4-nsm) {\,Single female \\[-0.5ex]  ($f,sk$)};
 	\node [block, above right   = 0.8 cm and -0.75 cm of 3] (6) {\,\,Single male\,\,\\[-0.5ex] ($m,s$)};
@@
     \draw[->] (2.180) to [bend left = 30] node [left] {\scriptsize Disagree} (4.270);
-    \draw[->,thick,dotted] (4-sm.20) to [bend left = 10] node [above right=-0.2cm and 0.7cm] {\scriptsize Costly abortion \\[-0.75ex] \scriptsize (if possible)} (6-f.50);
     \draw[->] (3.0) to [bend right = 30] node [right] {\scriptsize Disagree} (6.270);
@@
 \text{ \footnotesize (met and not pregnant): \ }& +  p^{\text{m}}_t\cdot (1- p^{p}_t) \cdot \left[ m^{np}_t \cdot V^{f,c}_{t+1}  + (1-m^{np}_t)\cdot  V^{f,s}_{t+1}\right] \\
-\text{ \footnotesize (met and pregnant): \ } & + p^{\text{m}}_t \cdot p^{\text{p}}_t\cdot   \left[ m^p_t \cdot V^{f,ck}_{t+1}  + (1-m^p_t)\cdot  V^{f,\text{preg}}_{t+1}\right] \Big\} \Bigg\}, \\
-\text{ \footnotesize (uncertain abortion access): } &  V^{f,\text{preg}}_{t+1} = p^{\text{access}} \cdot \max\{V^{f,sk}_{t+1},V^{f,s}_{t+1} - \phi_a\} \\
-   & \hspace{1cm} + (1-p^{\text{access}})\cdot V^{f,sk}_{t+1},\\
+\text{ \footnotesize (met and pregnant): \ } & + p^{\text{m}}_t \cdot p^{\text{p}}_t\cdot   \left[ m^p_t \cdot V^{f,ck}_{t+1}  + (1-m^p_t)\cdot  V^{f,sk}_{t+1}\right] \Big\} \Bigg\} \\
 \text{ \footnotesize (budget constraint): \ } & c + a' = R\cdot a  + w , \\
@@
 \text{ \footnotesize (divorce): \ } & + d_t  \left[ \theta^f \cdot  \tilde{V}^{f,s}_{t+1} + \theta^m \cdot \tilde{V}^{m,s}_{t+1}  \right]  \Big\} \Bigg\}\\
-\text{ \footnotesize (budget constraint, rts): \ } &{\normalsize \left[c_f^{1+\rho} + c_m^{1+\rho}\right]^{\frac1{1+\rho}} + a' = R\cdot a + w_f + w_m}\\
+\text{ \footnotesize (budget constraint): \ } &{\normalsize \left[c_f^{1+\rho} + c_m^{1+\rho}\right]^{\frac1{1+\rho}} + a' = R\cdot a + w_f + w_m}\\
 \text{ \footnotesize (match quality evolution): \ } &{\normalsize \psi' = \psi + \varepsilon^{\psi}}\\
@@
 \text{ \footnotesize (stay): \ } & + (1-d_t) \cdot  \left[ \theta^f \cdot  V^{f,ck}_{t+1} + \theta^m \cdot V^{m,ck}_{t+1}  \right] \\
-\text{ \footnotesize (divorce): \ } & + d_t \cdot \left[ \theta^f \cdot  \tilde V^{f,sk}_{t+1} + \theta^m \cdot \tilde V^{m,s}_{t+1}  \right]  \Big\} \Bigg\}\\
+\text{ \footnotesize (divorce): \ } & + d_t  \left[ \theta^f \cdot  \tilde V^{f,sk}_{t+1} + \theta^m \cdot \tilde V^{m,s}_{t+1}  \right]  \Big\} \Bigg\}\\
 \text{ \footnotesize (budget constraint): \ } &{\normalsize \left[c_f^{1+\rho} + c_m^{1+\rho}\right]^{\frac1{1+\rho}} + x + a' = R\cdot a + w_f\cdot l_f + w_m}\\
-\text{ \footnotesize (child consumption flow): \ } &{\normalsize Q = \left[x^\lambda + \kappa \cdot (1-l_f)^\lambda\right]^{\frac1\lambda}, \ \ \ Q\geq \bar{Q}} \text{ \footnotesize  (subsistence constraint) } \\
-\text{ \footnotesize (female skills $\downarrow$ if not working): \ } &{\normalsize \log w_f = z_f + \text{Trend}_{f,t}, \ \ z'_f = z_f + \varepsilon^f - \delta(l^f)}\\
+\text{ \footnotesize (child quality flow): \ } &{\normalsize Q = \left[x^\lambda + \kappa \cdot (1-l_f)^\lambda\right]^{\frac1\lambda}, \ \ \ Q\geq \bar{Q}} \text{ \footnotesize  (subsistence constraint) } \\
+\text{ \footnotesize (female skills depreciate): \ } &{\normalsize \log w_f = z_f + \text{Trend}_{f,t}, \ \ z'_f = z_f + \varepsilon^f - \delta(l^f)}\\
 \end{align*}
@@
 \begin{itemize}
-\item Bargaining weights $\theta$ adjust to satisfy both participation constraints
-\[\Delta \theta^i \text{ reflects Lagrange multiplier on $i$'s participation constraint}: V^{i,c\bullet}_{t+1} \geq \tilde{V}^{i,s}_{t+1}\]
+\item Bargaining weights $\theta$ adjust to satisfy both participation constraints.
 \item If no adjustment is possible, the couple divorces
-\item Assets are split evenly upon divorce (follows Voena, 2015)
-\item Partners have fixed utility costs $\phi_d$ of divorce
+\item Assets are split evenly upon divorce
 \item If couple with a child divorces wife becomes a single mother
+\item Child support: adjustment in permanent income component $z$
 \end{itemize}
@@
 \begin{frame}
-\frametitle{Child Support}
-\framesubtitle{Paid forever, but with a chance to avoid}
+\frametitle{Marriage}
 \begin{itemize}
-\item Imperfect enforcement: probability to be awarded is $p^{\text{award}}$
-\begin{itemize}
-\item Married couples: 46\%, never married couples: 28\% chances
-\end{itemize}
-\item If enforced:
-\begin{itemize}
-\item man loses productivity $z$ so earnings fall by 20\%
-\item woman gains productivity $z$ matching this value
-\end{itemize}
-\item As $z' = z + \varepsilon^z$, all changes are permanent
-\end{itemize}
-\end{frame}
-
-%\begin{frame}
-%\frametitle{Abortions / Birth Control}
-%\begin{itemize}
-%\item When unplanned pregnancy happens, with probability $p^{\text{access}}$ woman can avoid being single mother
-%\item If possible, she pays \emph{fixed utility costs} $u^{a}$ for this
-%\item Targets: $\sim40\%$ unplanned pregnancies are aborted + age profile of abortions
-%\item Some women \textit{choose} to be single mothers even with abortion access
-%\end{itemize}
-%\end{frame}
-
-\begin{frame}
-\frametitle{Marriage}
-\framesubtitle{Introducing Social Stigma}
-\begin{itemize}
-\item Singles meet singles with random $a$ and $z$ (estimated)
+\item Singles meet singles with similar $a$ and $z$ 
 \item Potential couple gets a draw of match quality $\psi$
 \item If both partners agree to marry, initial $\theta$ is set by symmetric Nash Bargainig
-\item More utility shifters (unexplained things):
+\item Two utility shifters:
 \begin{itemize}
-\item If unplanned pregnancy happens, partners face {\textbf{social stigma}} $\phi_s$ if refuse to marry
+\item If unplanned pregnancy happens, partners face social stigma $\phi_s$ if refuse to marry
 \item If male meets a single mother, he lose $\phi_m$ for having step children
@@
 
-\section{Mechanics \& Intuition}
+\section{Mechanics}
 \begin{frame}
 \frametitle{Unplanned Pregnancy $\Rightarrow$ Worse Match?}
-\framesubtitle{Why people enter knowingly risky marriages?}
+\framesubtitle{Model Implications}
+
 \begin{itemize}
@@
 \begin{center}
-\begin{tabular}{c c}
-\hspace{-1cm} \includegraphics[scale=0.5]{change_upp.pdf} &
-\hspace{-0.5cm}\includegraphics[scale=0.5]{surp_upp.pdf} 
-\end{tabular}
-
+\includegraphics[scale=0.7]{change_upp.pdf}
 \end{center}
@@
 \end{center}
+Social stigma $\approx$ draw $\psi$ from different distribution if pregnant.  %More productive women suffer more.
 \end{frame}
@@
 \end{itemize}
-\item Model fits well for both groups, but high education results are more settled
+\item \textbf{Today:} estimates for high education (large impact of unplanned pregnancy)% + preliminary comparison with low education
 \end{itemize}
@@
 \footnotesize child expenditures share, \% in LF with kids & \footnotesize  household technology: $\kappa$, $\alpha$ and $\bar{Q}$ \\
-\footnotesize \% divorced 1--10 years after marriage if KF or MF & \footnotesize  social stigma, $\bar{Q}$ and more ... \\
-\footnotesize \% unplanned pregnancies aborted + age profile & \footnotesize  abortion costs and access \\\hline
+\footnotesize \% divorced 1--10 years after marriage if KF or MF & \footnotesize  social stigma, $\bar{Q}$ and more ... \\\hline
 \end{tabular}
@@
 \frametitle{Estimates}
-\framesubtitle{Simulated Method of Moments (SMM) with optimization routine from Arnoud et al, 2019; 17 parameters}
+\framesubtitle{Simulated Method of Moments (SMM) with optimization routine from Arnoud et al, 2019; 14 parameters}
 \begin{tabular}{l c c c}\hline
@@
 \multicolumn{4}{c}{Utility from kids: $\phi + \alpha \cdot \frac{Q^{1-\xi}}{1-\xi}$, \ $Q = [x^{\lambda} + \kappa\cdot(1-l_f)^{\lambda}]^{\frac1{\lambda}}$. $Q\geq \bar{Q}$}\\\hline
-$\phi$, $\alpha$, $\kappa$  \footnotesize (utility params) &\footnotesize  $(1.39, 0.36, 0.78)$ &\footnotesize $(1.80,0.45, 1.16)$ & SMM \\
-$\bar{Q}$ \footnotesize (subsistence constraint) & 1.7 & 0.70 & SMM\\
+$\phi$, $\alpha$, $\kappa$  \footnotesize (utility params) &\footnotesize  $(0.96, 0.38, 0.90)$ &\footnotesize $(1.89,0.46, 0.85)$ & SMM \\
+$\bar{Q}$ \footnotesize (subsistence constraint) & 0.87 & 0.30 & SMM\\
+$\xi$ \footnotesize (power)  & 1.5 & 1.5 & As $\sigma$ \\
+$\lambda$ \footnotesize (substitutability)  & 0.7 & 0.7 & \footnotesize Sommer (2016) \\\hline\hline %Attanasio and Weber (1995) \\\hline
+%\multicolumn{4}{p{0.9\linewidth}}{\footnotesize \textbf{Identification:} hazards of new children over age, \% with kids by years after marriage}\\\hline\hline
 \multicolumn{4}{c}{Meeting probability $p^{\text{meet}}$, pregnancy probability $p^{\text{preg}}$: quadratic polynomials}\\\hline
-$p^{\text{meet}}$  \footnotesize at (21,30,40) & \footnotesize $(0.11, 0.21, 0.34)$ & \footnotesize  $(0.44, 0.10, 0.65)$ & SMM\\
-$p^{\text{preg}}$ \footnotesize  at (21,28,35) & \footnotesize  $(0.01, 0.04, 0.01)$ & \footnotesize  $(0.09, 0.07, -0.05)$ & SMM \\\hline\hline
+$p^{\text{meet}}$  \footnotesize at (21,28,35) & \footnotesize $(0.23, 0.67, 0.72)$ & \footnotesize  $(0.49, 0.17, 0.49)$& SMM\\
+$p^{\text{preg}}$ \footnotesize  at (21,28,35) & \footnotesize  $(0.06, 0.01, 0.06)$ & \footnotesize  $(0.40, 0.08, 0.52)$ & SMM \\\hline\hline
 \multicolumn{4}{c}{Additive utility shifters}\\\hline
-$\sigma(\varepsilon^\psi), \sigma(\psi_0)$   \footnotesize (match quality distribution) & $(0.10,0.14)$ & $(0.88,4.8)$ & SMM \\
-$\phi_s$ \footnotesize  (social stigma) & 4.21 & 5.31 & SMM\\
-$\phi_m$ \footnotesize   (single mom remarriage loss) & 43.3 & 79.6 & SMM \\\hline
-$\phi_d$ \footnotesize   (divorce costs) & 5.3 & 1.2 & SMM \\\hline
-$(p^{\text{access}}, \phi_a)$ \footnotesize   (abortion parameters) & $(0.22,37.5)$  & $(0.18,5.51) $& SMM \\\hline
+$\sigma(\psi_0)$  \footnotesize (match quality initial) & 2.70 & 2.58 & SMM \\
+$\sigma(\varepsilon^\psi)$  \footnotesize (match quality shock) & 0.38 & 0.48 & SMM \\
+$\phi_s$ \footnotesize  (social stigma) & 0.68 & 1.11 & SMM\\
+$\phi_m$ \footnotesize   (single mom remarriage loss) & 14.2 & 20.3 & SMM \\\hline
 \end{tabular}
@@
 
-\begin{frame}[label=counterfactuals-abortions]
-\frametitle{What Explains The Difference?}
-\framesubtitle{Removing Social Stigma (1/4) + Fertility Control (3/4).}
+%
+%\begin{frame}
+%\frametitle{Fertility Timing Is Hard To Get}
+%\framesubtitle{Need Large Taste Shocks To Get The Results}
+%\begin{center}
+%\includegraphics[scale=0.75]{kids_yaftmar.pdf} 
+%\end{center}
+%\end{frame}
+%
+
+\begin{frame}
+\frametitle{Unplanned Pregnancies Are Large Risk}
+\framesubtitle{College women}
+\[\text{Definition:} \ \  EV = a^*_t(w) \text{ such that } V_{t}^{f,s,\text{with upp}}(a=a^*,w) = V_{t}^{f,s,\text{no upp}}(a=0,w)\]
 \begin{center}
-\begin{tabular}{c c}
-\hspace{-1cm}\includegraphics[scale=0.5]{div_kfmf_equalized.pdf}  & \hspace{-0.5cm} \includegraphics[scale=0.5]{relshares_equalized.pdf} 
-\end{tabular}
-\end{center}
-Welfare: men gain $\approx \$35k$, women gain $\approx \$12k$, \hyperlink{welfare-abortions}{\beamerbutton{details}} 
-\end{frame}
-
-\begin{frame}[label=counterfactuals-child-support]
-\frametitle{Removing Existing Child Support --- Part 1}
-\framesubtitle{Less single mothers, less divorces, more abortions}
-No child support instead of current (partially enforced 20\%)
-\begin{center}
-\begin{tabular}{c c}
-\hspace{-1cm}\includegraphics[scale=0.5]{div_kfmf_child_support.pdf}  & \hspace{-0.5cm} \includegraphics[scale=0.5]{sm_child_support.pdf} 
-\end{tabular}
-\end{center}
-Just removing the social stigma \hyperlink{counterfactuals-social-stigma}{\beamerbutton{here}}
-
-Welfare: men gain $\approx \$3k$, women lose $\approx \$1k$ \hyperlink{welfare-child-support}{\beamerbutton{details}} 
-\end{frame}
-
-\begin{frame}[label=counterfactuals-child-support]
-\frametitle{Removing Existing Child Support --- Part 2}
-\framesubtitle{Hardly Affects Marriage in Any Way}
-Can you notice the differences?
-\begin{center}
-\begin{tabular}{c c}
-\hspace{-1cm}\includegraphics[scale=0.5]{evermar_cs.pdf}  & \hspace{-0.5cm} \includegraphics[scale=0.5]{popshares_cs.pdf} 
-\end{tabular}
+\includegraphics[scale=0.6]{upp_welfare.pdf} 
 \end{center}
@@
 \begin{frame}[label=cf]
-\frametitle{More Counterfactuals}
+\frametitle{Counterfactual Exercises}
 \begin{itemize}
-\item No divorce costs \hyperlink{no_divorce_costs}{\beamerbutton{pictures}}
+\item Enforcing 20\% child support (divorced fathers pay) \hyperlink{child_support}{\beamerbutton{pictures}} 
 \begin{itemize}
-\item More divorces, less kids
-\item But less divorces conditional on having kids!
-\item Welfare: females lose, males gain, gain on average
+\item Divorced in 10 years if KF: 13.5\% $\to$ 10.7\%.
+\item More kids + fewer divorces
 \end{itemize}
-%\item Removing gender pay gap: same trend + variance for men and women \hyperlink{pay_gap}{\beamerbutton{pictures}} 
-\item No unplanned pregnancies at all 
+\item Removing social stigma related to shotgun marriage: setting $\phi_s = 0$ \hyperlink{noss}{\beamerbutton{pictures}} 
 \begin{itemize}
-\item Remaining KF (step-child-couples) and MF become very similar
-\item Men gain $\sim \$25k$, women gain $\sim \$13k$ 
-\item Around 1/10 less divorces
+\item Divorced in 10 years if KF: 13.5\% $\to$ 9.9\%
+\end{itemize}
+\item Removing remarriage penalty: setting $\phi_r = 0$ \hyperlink{remarriage_penalty}{\beamerbutton{pictures}} 
+\begin{itemize}
+\item Divorced in 10 years if KF: 13.5\% $\to$ 12.4\%
+\item More kids
+\end{itemize}
+\item Removing gender pay gap: same trend + variance for men and women \hyperlink{pay_gap}{\beamerbutton{pictures}} 
+\begin{itemize}
+\item Divorced in 10 years if KF: 13.5\% $\to$ 12.6\%
+\item Fewer kids somehow
 \end{itemize}
@@
 \begin{itemize}
-\item Model mechanisms can rationalize creation and differences in divorce
-\item Child support is welfare-decreasing and does not get people together
-\item Controlling fertility helps everyones
-\item \textbf{Love $>>$ Money}
-%\item Nothing beats the birth control in terms of welfare
-%\item Child support does not get people together and creates more single mothers
-%\item Empirically, more control over fertility $\Rightarrow$ less KF, lower differences in performance
+\item Risk of unplanned pregnancies explains the pattern well
+\item Men generally win from unexpected kids when commitment is limited, enforcing commitment helps
+\item Social stigma explains less than half of the difference
 \end{itemize}
@@
 
-%\begin{frame}[label=remarriage_penalty]
-%\frametitle{Removing Remarriage Penalty For Single Mothers}
-%\framesubtitle{Divorced in 10 years if KF: 13.5\% $\to$ 12.4\%. Divorced in cross-section if KF: 9.3\% $\to$ 8.9\% + More Kids}
-%\begin{center}
-%\begin{tabular}{c c}
-%\hspace{-1cm} \includegraphics[scale=0.5]{div_kfmf_no_remar.pdf}  &  \includegraphics[scale=0.5]{ever_kids_no_remar.pdf} 
-%\end{tabular}
-%\end{center}
-%\hyperlink{cf}{\beamerbutton{back}} 
-%\end{frame}
-
-\begin{frame}[label=no_divorce_costs]
-\frametitle{No Divorce Costs}
-\framesubtitle{Less kids, more shotgun marriages (relatively), less divorces if have kids}
+\begin{frame}[label=child_support]
+\frametitle{Enforcing Child Support}
+\framesubtitle{Divorced in 10 years if KF: 13.5\% $\to$ 10.7\%. Divorced in cross-section if KF: 9.3\% $\to$ 6.7\%. More kids!}
 \begin{center}
 \begin{tabular}{c c}
-\hspace{-1cm} \includegraphics[scale=0.5]{popshares_nodivcosts.pdf}  &  \includegraphics[scale=0.5]{div_kfmf_nodivcosts.pdf} 
+\includegraphics[scale=0.45]{div_kfmf_cs.pdf} & \includegraphics[scale=0.45]{kids_cs.pdf} 
+\end{tabular}
+\end{center}
+Mechanics: divorce $\Rightarrow$ $z^m$ decreases so $w^m$ $\downarrow$ 20\%, $z^f$ increases accordingly.
+
+\hyperlink{cf}{\beamerbutton{back}} 
+\end{frame}
+
+\begin{frame}[label=noss]
+\frametitle{Removing Social Stigma}
+\framesubtitle{Divorced in 10 years if KF: 13.5\% $\to$ 9.9\%. Divorced in cross-section if KF: 9.3\% $\to$ 7.7\%}
+\begin{center}
+\includegraphics[scale=0.66]{removing_ss.pdf} 
+\end{center}
+\hyperlink{cf}{\beamerbutton{back}} 
+\end{frame}
+
+\begin{frame}[label=remarriage_penalty]
+\frametitle{Removing Remarriage Penalty For Single Mothers}
+\framesubtitle{Divorced in 10 years if KF: 13.5\% $\to$ 12.4\%. Divorced in cross-section if KF: 9.3\% $\to$ 8.9\% + More Kids}
+\begin{center}
+\begin{tabular}{c c}
+\hspace{-1cm} \includegraphics[scale=0.5]{div_kfmf_no_remar.pdf}  &  \includegraphics[scale=0.5]{ever_kids_no_remar.pdf} 
 \end{tabular}
@@
 \begin{itemize}
-\item I do not classify as KF or MF people with $\Delta T < -5$ or $\Delta T > 5$
+\item I do not classify as KF or MF people with $\Delta T < -5$ or $\Delta T > 10$
 \item I exclude small share of people with marital statuses ``separated'', ``widowed'' and ``spouse absent''
@@
 
-\begin{frame}[label=counterfactuals-social-stigma]
-\frametitle{Removing Social Stigma}
-\framesubtitle{Less single mothers, less divorces, more abortions}
-No child support instead of current (partially enforced 20\%)
-\begin{center}
-\begin{tabular}{c c}
-\hspace{-1cm}\includegraphics[scale=0.5]{div_kfmf_ss.pdf}  & \hspace{-0.5cm} \includegraphics[scale=0.5]{sm_among_mothers_ss.pdf} 
-\end{tabular}
-\end{center}
-\hyperlink{counterfactuals-child-support}{\beamerbutton{back}} 
-\end{frame}
-
-\begin{frame}[label=welfare-abortions]
-\frametitle{What Explains The Difference? Welfare Implications}
-\framesubtitle{Removing Social Stigma (1/4) + Fertility Control (3/4)}
-\begin{center}
-\begin{tabular}{c c}
-\hspace{-1cm}\includegraphics[scale=0.5]{welfare_female_equalized.pdf}  & \hspace{-0.5cm} \includegraphics[scale=0.5]{welfare_male_equalized.pdf} 
-\end{tabular}
-\end{center}
-\hyperlink{counterfactuals-abortions}{\beamerbutton{back}} \hyperlink{welfare-methodology}{\beamerbutton{methodology}} 
-\end{frame}
-
-\begin{frame}[label=welfare-child-support]
-\frametitle{Child Support Removal}
-\framesubtitle{More abortions but not much changes...}
-\begin{center}
-\begin{tabular}{c c}
-\hspace{-1cm}\includegraphics[scale=0.5]{welfare_female_child_support.pdf}  & \hspace{-0.5cm} \includegraphics[scale=0.5]{welfare_male_child_support.pdf} 
-\end{tabular}
-\end{center}
-\hyperlink{counterfactuals-child-support}{\beamerbutton{back}} \hyperlink{welfare-methodology}{\beamerbutton{methodology}} 
-\end{frame}
-
-\begin{frame}[label=welfare-methodology]
-\frametitle{Definitions}
-\framesubtitle{Equivalent Variation (EV) Style}
-Value of any policy change can be computed in terms of equivalent amount of assets.
-
-If the change is preferred:
-\[EV = a^*_t(w) \text{ such that } V_{t}^{f,s,\text{no change}}(a=a^*,w) = V_{t}^{f,s,\text{with change}}(a=0,w)\]
-Otherwise:
-\[EV = -a^*_t(w) \text{ such that } V_{t}^{f,s,\text{with change}}(a=a^*,w) = V_{t}^{f,s,\text{no change}}(a=0,w)\]
-
-Analogously for $m$, a bit trickier for couples.
-
-\hyperlink{welfare-abortions}{\beamerbutton{back to abortions}} \hyperlink{welfare-child-support}{\beamerbutton{back to child support}}
-
-\end{frame}
-
 \backupend

### 2020-10-02 slides-oct-20/slides.tex
Current stage

--- 2020-10-02 slides-aug-20/slides.tex
@@
 
-\setbeamertemplate{footline}[frame number]
+%\setbeamertemplate{footline}[frame number]
+
+\setbeamertemplate{footline}{% 
+  \hfill% 
+  \usebeamercolor[fg]{page number in head/foot}% 
+  \usebeamerfont{page number in head/foot}% 
+  \insertframenumber%
+  %\,/\,\inserttotalframenumber
+  \kern1em\vskip2pt% 
+}
 
@@
 \begin{frame}
-\frametitle{``America's strongest anti-poverty weapon''?}
+\frametitle{Can Economic Policies Create Successful Marriages?}
 \begin{itemize}
-\item Married two-parent families \ : \ desirable outcomes for parents and children%$\genfrac{}{}{0pt}{1}{\Leftarrow}{\Rightarrow}$
+\item Reproduction of poverty is a motivation
+\item Married two-parent families \ $\genfrac{}{}{0pt}{1}{\Leftarrow \text{ or}}{\Rightarrow \text{  ?}}$ \ desirable outcomes for parents and children%
 \item Pro-marriage policies: child support, tax brackets, divorce regulations, etc
-\item Can economic policies create successful marriages? Love vs money.
+\item I propose a measurable dimension of marriage heterogeneity
 \end{itemize}
@@
 
-\frametitle{Kids-First Marriages are Less Stable}
+\frametitle{Kids-First Marriages are Closer to The Threshold}
 \begin{enumerate}
-\item{\textbf{Shotgun marriage}} --- couple marrying when woman gets pregnant
-\item{\textbf{Kids-first marriage}} --- child \textit{before or in the year} of marriage
+\item{\textbf{Kids-first couples}} --- child \textit{before or in the year} of marriage
+\item{\textbf{Marriage-first couples}} --- child \textit{at the following year or later}
 \end{enumerate}
 \begin{itemize}
-\item Around 1/4 marriages in the US today are kids-first
+\item Around 1/4 of childbearing marriages are kids-first
 \item They have much higher divorce rates, especially among college graduates
-\item Marginal in economic sense: could be sensitive to policies 
+\item Natural distinction to study marriage quality
+%\item Marginal in economic sense: sensitive to policies 
 \end{itemize}
@@
 \begin{enumerate}
-\item Document the shotgun marriage and its outcomes empirically
+\item Document the kids-first marriage and its outcomes empirically
 \item Use these empirical relations to estimate a model capturing:
@@
 \end{itemize}
-\item Look at what drives decisions to marry and the differences in divorce
+\item Use it to correct the data imperfections and answer the questions:
 \begin{itemize}
-\item what explains the education gradient?
-\item can policies fix the difference or reduce the share?
-\item what is the value of marriage in general?
+\item Why do the kids-first couples divorce more?
+\item Can policies fix the difference or reduce the share?
+\item How to promote \emph{good} marriages?
 \end{itemize}
@@
 \begin{frame}
+\frametitle{Conceptual Framework}
+\begin{itemize}
+\item Marriages differ in match quality $\psi$ (amount of love)
+\item Potential spouses observe their $\psi$ and marry if $\psi > \psi_m$ (\textit{agreement threshold})
+\item Match quality evolves, and couple divorces if later $\psi < \psi_d$ (\textit{divorce threshold})
+\item Lower $\psi$ implies higher risk of divorce
+\item Structural model recovers \textit{distribution} of $\psi$ from the risks of divorce
+\end{itemize}
+\end{frame}
+
+\begin{frame}
 \frametitle{Summary of Results}
+\begin{enumerate}
+\item Differences in divorce:
 \begin{itemize}
-\item 23\% of women had the first child before their first marriage (kids first)
-\item They are 1.8 times more likely to be divorced than ``marriage first''
+\item Kids before marriage lower the agreement threshold, divorce chances increase by 1/3
+\item 40--50\% of the kids-first marriages happen \emph{because} of kids
+\item Risks of being a single mother and social stigma are the main factors
+\end{itemize}
+\item Implications for policies:
+\begin{itemize}
+\item Pushing people to marry more is welfare-decreasing for everyone
+\item Making divorce harder has some winners, but people lose on average
+\item Child support has limited effects on marriage formation, but increases fertility, stimulates divorce and may hurt children
+\end{itemize}
+%\item 23\% of women had the first child before their first marriage (kids first)
+%\item They are 1.8 times more likely to be divorced than ``marriage first''
 %\item College grads: 2.8 times more likely to be divorced
@@
 %\end{itemize}
-\item My structural model corrects selection and data imperfections
+%\item My structural model corrects selection and data imperfections
 %\item Model results:
@@
 %\end{itemize}
-\item Model answers: Love $>$ Money. In particular:
-\begin{itemize}
-\item Threat of raising the child alone explains most of the difference
-\item Regulating marriages by policies is both hard and inefficient
-\item Easier control over fertility fixes shotgun marriages and has bigger return
-\end{itemize}
-\end{itemize}
+%\item Model answers: Love $>$ Money. In particular:
+%\begin{itemize}
+%\item Threat of raising the child alone explains most of the difference
+%\item Regulating marriages by policies is both hard and inefficient
+%\item Easier control over fertility fixes shotgun marriages and has bigger return
+%\end{itemize}
+\end{enumerate}
 \end{frame}
@@
 \begin{itemize}
+\item Empirics on shotgun marriages (demography and sociology):
+\begin{itemize}
+\item {\small  \color{gray}Gibson-Davis et al (2016), Guzzo (2009),  Nuevo-Chiquero (2014)}
+\item I can recover the causal content of these numbers
+\end{itemize}
+\item Lifecycle models:
+\begin{itemize}
+\item Dynamic limited commitment for marriage and divorce: {\small  \color{gray}Mazzocco, 2007; Voena, 2015; Low et al., 2018; Shephard, 2019; Blasutto, 2020}
+\item Fertility during lifecycle: {\small  \color{gray}Sommer, 2015, Ejrnæs, Jørgensen, 2018}
+\item I am the first to estimate a marriage/divorce model with fertility choice
+\end{itemize}
 \item Policy literature:
 \begin{itemize}
-\item Empirical: Alessina, Guiliano, 2005;  Tannenbaum, 2020; Rossin-Slater, 2017
-\item Structural: Forester, 2020; Brown, Flinn, 2011
-\item Focus on particular policies, ignore interactions, limited quantitative power
-\end{itemize}
-\item Lifecycle literature:
-\begin{itemize}
-\item Dynamic limited commitment for marriage and divorce: Mazzocco, 2007; Voena, 2015; Low et al., 2018; Shephard, 2019; Blasutto, 2020
-\item Fertility during lifecycle: Sommer, 2015, Ejrnæs, Jørgensen, 2018
-\item Link between divorce and fertility is missing
-\end{itemize}
-\item Empirics on shotgun marriages (demography and sociology):
-\begin{itemize}
-\item Gibson-Davis et al (2016), Guzzo (2009),  Nuevo-Chiquero (2014)
-\item Mostly descriptive
+\item Empirical: {\small \color{gray}Alessina, Guiliano, 2005;  Tannenbaum, 2020; Rossin-Slater, 2017}
+\item Structural: {\small  \color{gray}Forester, 2020; Brown, Flinn, 2011; Kennes, Knowles, 2020}
+\item I am the first to consider creation and dissolution of shotgun marriages structurally
 \end{itemize}
 \end{itemize}
-%
-%\begin{itemize}
-%\item Empirical literature documents the scope of shotgun marriage around the world:
-%\begin{itemize}
-%\item Gibson-Davis et al (2016), Guzzo (2009),  Nuevo-Chiquero (2014)
-%\item Mostly descriptive
-%\end{itemize}
-%\item Dynamic limited commitment models of divorce and marriage:
-%\begin{itemize}
-%\item Marcet, Marimon (2019), Mazzocco (2007) --- setting up the framework
-%\item Voena (2015), Shephard (2019); Lafortune, Low  (2019) --- applied questions
-%\item Mostly ignore or simplify interaction with fertility
-%\end{itemize}
-%\item Unplanned and teenage pregnancy literature:
-%\begin{itemize}
-%\item Ejrnæs, Jørgensen (2018) --- lifecycle model
-%\item Card, Wise (1978) to Ashcraft et al (2013) --- lots of empirical papers
-%\item Treat household structure as fixed or exogenous
-%\end{itemize}
-%\end{itemize}
 \end{frame}
 
-\section{Data}
+\section{Empirical Motivation}
 \begin{frame}
@@
 
-\begin{frame}
-\frametitle{Kids First Women Are Divorced More Often: Quantification}
-\framesubtitle{ACS 2009--2017, Women 21--40, married once}
-\begin{center}
-\begin{tabular}{l r r r r }
-\hline
-& \multicolumn{2}{c}{share of divorced if ... }&  \\
-&  marriage-first & kids-first & (share of kids-first) &  \\\hline
-\multicolumn{5}{l}{\textbf{All sample}} \\\hline
-\textit{Cross-sectional share of divorced} &  9.9 & 18.1 & (26.1) \\
-\textit{Divorced 5 years after marriage} &  5.1 & 14.3  & (24.6) \\
-\textit{Divorced 10 years after marriage} & 10.4 & 22.8 & (24.4) \\\hline
-\multicolumn{5}{l}{\textbf{Cross-sectional share of divorced for subsamples}} \\\hline
-\textit{High school only} &  12.8 & 17.3 & (37.0) \\
-\textit{Some college} & 14.3 & 21.0 & (31.3) \\
-\textit{College or more} &   5.3 & 14.8 & (11.9) \\\hline
-\textit{First birth before 25} & 15.5 & 19.9 & (41.8) \\
-\textit{First birth at 25 or later} &  6.2 & 10.9 & (10.7)  \\\hline
+\begin{frame}[label=dt_more]
+\frametitle{Kids First Women Are Divorced More Often: More Evidence}
+\begin{itemize}
+\item The divorce pattern is present in {all subgroups}
+\item The pattern is \textit{sharper} for {college graduates} or those with later births
+\item Controlling for many confounders explains no more than a half
+\item Step-children drive relatively small part of the result
+\end{itemize}
+\vspace{1cm}
+\begin{tabular}{c c c c}
+More results: & \hyperlink{emp-table}{\beamerbutton{subgroups}} & \hyperlink{emp-controls}{\beamerbutton{controls}} & \hyperlink{emp-step}{\beamerbutton{step-children}} 
 \end{tabular}
-\end{center}
-\end{frame}
 
-\begin{frame}
-\frametitle{Kids First Women Are Divorced More Often: Controls}
-\begin{center}
-%\vspace{-0.5cm}
-\begin{tabular}{l r r r }
-\multicolumn{4}{c}{\textit{Regression equation:} $\text{Divorced}_i = \Delta \cdot \text{kids-first}_i + \text{Controls}_i + \varepsilon_i$} \\\hline
-\hline
-& \multicolumn{2}{c}{Estimates}  \\
-& Difference ($\Delta$) & (standard error)  \\\hline
-\multicolumn{4}{l}{\textbf{All sample, regression}} \\\hline
-\textit{No controls (raw difference) }& 8.2 & (0.1)  \\
-\textit{Demographic controls }& 5.7 & (0.1)  \\
-\textit{Duration controls} &  5.8 & (0.2)  \\
-\textit{Demographic + duration} &  4.8 & (0.2)  \\
-\textit{Demographic + duration + income} & 4.1 & (0.3)  \\\hline
-\multicolumn{4}{l}{\textbf{All sample, propensity score matching}} \\\hline
-\textit{Demographic controls} & 5.0 & (0.3)  \\
-\textit{Demographic controls + income} & 4.9 & (0.2)  \\\hline
-\multicolumn{4}{p{0.825\linewidth}}{\footnotesize \textit{Controls:} demographics: FE for age $\times$ education, race, state, survey year.
-Duration: FE for age $\times$ age of the first marriage and age $\times$ age of the first birth.
-Income: 3rd degree polynomial of log gross income (if observed).}\\\hline\hline
-\end{tabular}
-\end{center}
-\end{frame}
-
-\begin{frame}
-\frametitle{Kids First Women Are Divorced More Often: Extra Evidence from SIPP}
-\begin{center}
-\begin{tabular}{l r r }
-\hline
-\multicolumn{3}{l}{\textbf{Including remarried people...}} \\\hline
-& \multicolumn{2}{c}{share of divorced if ... }  \\
-&  marriage-first & kids-first   \\\hline
-\textit{Ever divorced} & 34.4 & 48.0  \\
-\textit{Divorced if married once} &  12.4 & 19.3 \\
-\textit{Remarried} &  22.0 & 28.7 \\\hline
-\hline
-\multicolumn{3}{l}{\textbf{Separating step and own children (2+ kids only)}} \\\hline
-& share in sample & share of divorced   \\\hline
-{Marriage-first} & 72.1 & 10.5 \\
-{Kids-first-own} &  16.4 & 14.6 \\
-{Kids-first-other} &  11.4 & 27.2 \\\hline
-\multicolumn{3}{p{0.7\linewidth}}{\footnotesize \textit{Notes.} SIPP data, the cross-section of Wave 1, 2014, women 21+. Numbers are percents.}\\\hline\hline
-\end{tabular}
-\end{center}
 \end{frame}
@@
 \begin{align*}
-V^{f,s}_t(a,w) = \max\limits_{c,a'} \Bigg\{ \frac{c^{1-\sigma}}{1-\sigma} + \beta \cdot \E_t \Big\{  & 
-%& \\ \text{ (did not meet anyone): \ } & 
-(1-p^{\text{m}}_t)\cdot V_{t+1}^{f,s}(a',w') \\
-\text{ \footnotesize (met and not pregnant): \ }& +  p^{\text{m}}_t\cdot (1- p^{p}_t) \cdot \left[ m^{np}_t \cdot V^{f,c}_{t+1}  + (1-m^{np}_t)\cdot  V^{f,s}_{t+1}\right] \\
-\text{ \footnotesize (met and pregnant): \ } & + p^{\text{m}}_t \cdot p^{\text{p}}_t\cdot   \left[ m^p_t \cdot V^{f,ck}_{t+1}  + (1-m^p_t)\cdot  V^{f,\text{preg}}_{t+1}\right] \Big\} \Bigg\}, \\
-\text{ \footnotesize (uncertain abortion access): } &  V^{f,\text{preg}}_{t+1} = p^{\text{access}} \cdot \max\{V^{f,sk}_{t+1},V^{f,s}_{t+1} - \phi_a\} \\
-   & \hspace{1cm} + (1-p^{\text{access}})\cdot V^{f,sk}_{t+1},\\
+V^{f,s}_t(a,w) = & \max\limits_{c,a'} \Bigg\{ \frac{c^{1-\sigma}}{1-\sigma} + \beta \cdot \E_t  \text{\textbf{Transitions}}_{t+1} \Bigg\}  \\
 \text{ \footnotesize (budget constraint): \ } & c + a' = R\cdot a  + w , \\
@@
 \end{align*}
+\textbf{Transitions}:
+\begin{itemize}
+\item \footnotesize No partner met: stay single, $V^{f,s}_{t+1}$, random with probability $1-p^{\text{meet}}_t$
+\item \footnotesize Met, no unplanned pregnancy, random with probability $p^{\text{meet}}_t\cdot (1-p^{\text{preg}}_t)$: 
+\begin{itemize}
+\item \footnotesize Agree to marry: enter couple without a child, $V^{f,c}_{t+1}$
+\item\footnotesize  Disagree, stay single: $V^{f,s}_{t+1}$
+\end{itemize}
+\item \footnotesize Met, no unplanned pregnancy, random with probability $p^{\text{meet}}_t\cdot p^{\text{preg}}_t$: 
+\begin{itemize}
+\item \footnotesize Agree to marry: enter couple with a child, $V^{f,ck}_{t+1}$
+\item \footnotesize Disagree, \textit{pick} $\max\{V^{f,s}_{t+1} - \phi_a, V^{f,sk}_{t+1}\}$ \textit{if abortion is possible ($p^a$), otherwise} $V^{f,sk}_{t+1}$
+\end{itemize}
+\end{itemize}
 \end{frame}
@@
 \begin{align*}
-V^{c}_t(a,w_f,w_m,\psi,\theta) = \max\limits_{c_f,c_m,a',k} \Bigg\{ & \theta^f \cdot \frac{c_f^{1-\sigma}}{{1-\sigma}}  + \theta^m \cdot \frac{c_m^{1-\sigma}}{1-\sigma}  + \psi + \beta \cdot \E_t \Big\{   \\
-\text{ \footnotesize (stay, no birth): \ } & + (1-d_t) \cdot (1-k_t) \cdot  \left[ \theta^f \cdot  V^{f,c}_{t+1} + \theta^m \cdot V^{m,c}_{t+1}  \right] \\
-\text{ \footnotesize (stay, give a birth): \ } &  + (1-d_t) \cdot k_t \cdot  \left[ \theta^f \cdot  V^{f,ck}_{t+1} + \theta^m \cdot V^{m,ck}_{t+1}  \right]  \\
-\text{ \footnotesize (divorce): \ } & + d_t  \left[ \theta^f \cdot  \tilde{V}^{f,s}_{t+1} + \theta^m \cdot \tilde{V}^{m,s}_{t+1}  \right]  \Big\} \Bigg\}\\
+V^{c}_t(a,w_f,w_m,\psi,\theta) = \max\limits_{c_f,c_m,a',k} \Bigg\{ & \theta^f \cdot \frac{c_f^{1-\sigma}}{{1-\sigma}}  + \theta^m \cdot \frac{c_m^{1-\sigma}}{1-\sigma}  + \psi + \beta \cdot \E_t \text{\textbf{Transitions}}_{t+1} \Bigg\}   \\
 \text{ \footnotesize (budget constraint, rts): \ } &{\normalsize \left[c_f^{1+\rho} + c_m^{1+\rho}\right]^{\frac1{1+\rho}} + a' = R\cdot a + w_f + w_m}\\
 \text{ \footnotesize (match quality evolution): \ } &{\normalsize \psi' = \psi + \varepsilon^{\psi}}\\
-\text{ \footnotesize \textbf{(participation constraints)}: \ } &{\normalsize V^{f,c\bullet}_{t+1} (\cdots,\theta') \geq \tilde{V}^{f,s}_{t+1}, \ \ V^{m,c\bullet}_{t+1} (\cdots,\theta') \geq \tilde{V}^{m,s}_{t+1}.}\\
+\text{ \footnotesize \textbf{(participation constraints)}: \ } &{\normalsize V^{f,c\bullet}_{t+1} (\cdots,\theta') \geq \tilde{V}^{f,s}_{t+1}, \ \ V^{m,c\bullet}_{t+1} (\cdots,\theta') \geq \tilde{V}^{m,s}_{t+1}.}
 \end{align*}
+\textbf{Transitions}:
+\begin{itemize}
+\item  \footnotesize Stay together, no birth, choice: $\theta^f \cdot V^{f,c}_{t+1} + \theta^m \cdot V^{m,c}_{t+1}$
+\item  \footnotesize Stay together, try to give a birth, choice:
+\begin{itemize}
+\item  \footnotesize  \textit{If successful} ($p^{\text{birth}}_t$):  $\theta^f\cdot V^{f,ck}_{t+1} + \theta^m\cdot V^{m,ck}_{t+1}$, \textit{otherwise} $\theta^f\cdot V^{f,c}_{t+1} + \theta^m\cdot V^{m,c}_{t+1}$
+\end{itemize}
+\item  \footnotesize  Divorce, choice: $\theta^f \cdot \tilde V^{f,s}_{t+1} + \theta^m \cdot \tilde V^{m,s}_{t+1}$ (property division + divorce costs inside)
+\end{itemize}
 \end{frame}
@@
 \framesubtitle{$d_t$ --- get a divorce, $\tilde{V}$ --- values at divorce}
+\vspace{-1cm}
 \begin{align*}
-V^{ck}_t(a,w_f,w_m,\psi,\theta) = \max\limits_{c_f,c_m,x,l_f,a'} \Bigg\{ & \theta^f \cdot \frac{c_f^{1-\sigma}}{{1-\sigma}}  + \theta^m \cdot \frac{c_m^{1-\sigma}}{1-\sigma}  + \psi + \\ \text{ \footnotesize (child is a public good): \ }  & \phi + \alpha \cdot \frac{Q^{1-\sigma}}{1-\sigma} +  \beta \cdot \E_t \Big\{   \\
-\text{ \footnotesize (stay): \ } & + (1-d_t) \cdot  \left[ \theta^f \cdot  V^{f,ck}_{t+1} + \theta^m \cdot V^{m,ck}_{t+1}  \right] \\
-\text{ \footnotesize (divorce): \ } & + d_t \cdot \left[ \theta^f \cdot  \tilde V^{f,sk}_{t+1} + \theta^m \cdot \tilde V^{m,s}_{t+1}  \right]  \Big\} \Bigg\}\\
+V^{ck}_t(a,w_f,w_m,\psi,\theta) = \max\limits_{c_f,c_m,x,l_f,a'} \Bigg\{ & \theta^f \cdot \frac{c_f^{1-\sigma}}{{1-\sigma}}  + \theta^m \cdot \frac{c_m^{1-\sigma}}{1-\sigma}  + \psi + \\ \text{ \footnotesize (child is a public good): \ }  & \phi + \alpha \cdot \frac{Q^{1-\sigma}}{1-\sigma} +  \beta \cdot \E_t \text{\textbf{Transitions}}_{t+1}  \Bigg\} \\
 \text{ \footnotesize (budget constraint): \ } &{\normalsize \left[c_f^{1+\rho} + c_m^{1+\rho}\right]^{\frac1{1+\rho}} + x + a' = R\cdot a + w_f\cdot l_f + w_m}\\
-\text{ \footnotesize (child consumption flow): \ } &{\normalsize Q = \left[x^\lambda + \kappa \cdot (1-l_f)^\lambda\right]^{\frac1\lambda}, \ \ \ Q\geq \bar{Q}} \text{ \footnotesize  (subsistence constraint) } \\
-\text{ \footnotesize (female skills $\downarrow$ if not working): \ } &{\normalsize \log w_f = z_f + \text{Trend}_{f,t}, \ \ z'_f = z_f + \varepsilon^f - \delta(l^f)}\\
+\text{ \footnotesize (child consumption flow): \ } &{\normalsize Q = \left[x^\lambda + \kappa \cdot (1-l_f)^\lambda\right]^{\frac1\lambda}}, \\
+\text{ \footnotesize  (subsistence constraint): } & Q\geq \bar{Q} \\
+\text{ \footnotesize (female skills $\downarrow$ if not working): \ } &{\normalsize \log w_f = z_f + \text{Trend}_{f,t}, \ \ z'_f = z_f + \varepsilon^f - \delta(l^f)}
 \end{align*}
+\textbf{Transitions}:
+\begin{itemize}
+\item  \footnotesize Stay together, choice: $\theta^f \cdot V^{f,ck}_{t+1} + \theta^m \cdot V^{m,ck}_{t+1}$
+\item  \footnotesize Divorce, choice: $\theta^f \cdot \tilde V^{f,sk}_{t+1} + \theta^m \cdot \tilde V^{m,s}_{t+1}$ (property division + divorce costs + child support inside)
+\end{itemize}
 \end{frame}
@@
 
+\begin{frame}[plain,label=emp-table]
+\frametitle{Kids First Women Are Divorced More Often: Quantification}
+\framesubtitle{ACS 2009--2017, Women 21--40, married once}
+\begin{center}
+\begin{tabular}{l r r r r }
+\hline
+& \multicolumn{2}{c}{share of divorced if ... }&  \\
+&  marriage-first & kids-first & (share of kids-first) &  \\\hline
+\multicolumn{5}{l}{\textbf{All sample}} \\\hline
+\textit{Cross-sectional share of divorced} &  9.9 & 18.1 & (26.1) \\
+\textit{Divorced 5 years after marriage} &  5.1 & 14.3  & (24.6) \\
+\textit{Divorced 10 years after marriage} & 10.4 & 22.8 & (24.4) \\\hline
+\multicolumn{5}{l}{\textbf{Cross-sectional share of divorced for subsamples}} \\\hline
+\textit{High school only} &  12.8 & 17.3 & (37.0) \\
+\textit{Some college} & 14.3 & 21.0 & (31.3) \\
+\textit{College or more} &   5.3 & 14.8 & (11.9) \\\hline
+\textit{First birth before 25} & 15.5 & 19.9 & (41.8) \\
+\textit{First birth at 25 or later} &  6.2 & 10.9 & (10.7)  \\\hline
+\end{tabular}
+\end{center}
+\hyperlink{dt_more}{\beamerbutton{back}} 
+\end{frame}
+
+\begin{frame}[plain,label=emp-controls]
+\frametitle{Kids First Women Are Divorced More Often: Controls}
+\begin{center}
+%\vspace{-0.5cm}
+\begin{tabular}{l r r r }
+\multicolumn{4}{c}{\textit{Regression equation:} $\text{Divorced}_i = \Delta \cdot \text{kids-first}_i + \text{Controls}_i + \varepsilon_i$} \\\hline
+\hline
+& \multicolumn{2}{c}{Estimates}  \\
+& Difference ($\Delta$) & (standard error)  \\\hline
+\multicolumn{4}{l}{\textbf{All sample, regression}} \\\hline
+\textit{No controls (raw difference) }& 8.2 & (0.1)  \\
+\textit{Demographic controls }& 5.7 & (0.1)  \\
+\textit{Duration controls} &  5.8 & (0.2)  \\
+\textit{Demographic + duration} &  4.8 & (0.2)  \\
+\textit{Demographic + duration + income} & 4.1 & (0.3)  \\\hline
+\multicolumn{4}{l}{\textbf{All sample, propensity score matching}} \\\hline
+\textit{Demographic controls} & 5.0 & (0.3)  \\
+\textit{Demographic controls + income} & 4.9 & (0.2)  \\\hline
+\multicolumn{4}{p{0.825\linewidth}}{\footnotesize \textit{Controls:} demographics: FE for age $\times$ education, race, state, survey year.
+Duration: FE for age $\times$ age of the first marriage and age $\times$ age of the first birth.
+Income: 3rd degree polynomial of log gross income (if observed).}\\\hline\hline
+\end{tabular}
+\end{center}
+\vspace{-0.5cm}
+\hyperlink{dt_more}{\beamerbutton{back}} 
+\end{frame}
+
+\begin{frame}[plain,label=emp-step]
+\frametitle{Kids First Women Are Divorced More Often: Extra Evidence from SIPP}
+\begin{center}
+\begin{tabular}{l r r }
+\hline
+\multicolumn{3}{l}{\textbf{Including remarried people...}} \\\hline
+& \multicolumn{2}{c}{share of divorced if ... }  \\
+&  marriage-first & kids-first   \\\hline
+\textit{Ever divorced} & 34.4 & 48.0  \\
+\textit{Divorced if married once} &  12.4 & 19.3 \\
+\textit{Remarried} &  22.0 & 28.7 \\\hline
+\hline
+\multicolumn{3}{l}{\textbf{Separating step and own children (2+ kids only)}} \\\hline
+& share in sample & share of divorced   \\\hline
+{Marriage-first} & 72.1 & 10.5 \\
+{Kids-first-own} &  16.4 & 14.6 \\
+{Kids-first-other} &  11.4 & 27.2 \\\hline
+\multicolumn{3}{p{0.7\linewidth}}{\footnotesize \textit{Notes.} SIPP data, the cross-section of Wave 1, 2014, women 21+. Numbers are percents.}\\\hline\hline
+\end{tabular}
+\end{center}
+\hyperlink{dt_more}{\beamerbutton{back}} 
+\end{frame}
+
 \begin{frame}[plain,label=dt_graph_educ]

### 2020-10-02 slides-very-short/slides-short.tex
Current stage

--- 2020-10-02 slides-aug-20/slides.tex
@@
 
+\usepackage{xcolor}
+
 \begin{document}
-
-% REMOVE SLIDE COUNT
 
@@
 \begin{frame}
-\frametitle{``America's strongest anti-poverty weapon''?}
+\frametitle{Kids First or Marriage First}
 \begin{itemize}
-\item Married two-parent families \ : \ desirable outcomes for parents and children%$\genfrac{}{}{0pt}{1}{\Leftarrow}{\Rightarrow}$
-\item Pro-marriage policies: child support, tax brackets, divorce regulations, etc
-\item Can economic policies create successful marriages? Love vs money.
+\item Around 1/4 of childbearing marriages in the US are \textbf{kids-first}:
+\begin{itemize}
+\item they have their first child \textit{before or in the year} of their marriage
+\end{itemize}
+\item These marriages have much higher divorce rates but are still widespread
+\item What economic reasons make people do this? 
+\item Unplanned pregnancies are a potential cause
 \end{itemize}
@@
 \begin{frame}
-
-\frametitle{Kids-First Marriages are Less Stable}
-\begin{enumerate}
-\item{\textbf{Shotgun marriage}} --- couple marrying when woman gets pregnant
-\item{\textbf{Kids-first marriage}} --- child \textit{before or in the year} of marriage
-\end{enumerate}
+\frametitle{``America's strongest anti-poverty weapon''?}
 \begin{itemize}
-\item Around 1/4 marriages in the US today are kids-first
-\item They have much higher divorce rates, especially among college graduates
-\item Marginal in economic sense: could be sensitive to policies 
+\item Marriage is a good thing (statistically)
+%\item Conservatives blame millennials for not following ``the sequence''
+\item Is making people marry more and stay married longer welfare improving?
+\item The answer depends on understanding the economic \emph{value} of marriage
+\item Simple competitive market setup says no (but may not fit here)
 \end{itemize}
@@
 
-%
-%\begin{frame}
-%\frametitle{Questions}
-%\begin{itemize}
-%\item Drivers of inequality:
-%\begin{itemize}
-%\item Focusing on unmarried young single mothers omits an important transition
-%\item Unplanned pregnancies look different between age and education groups
-%\item Divorced parents may be worse than never married
-%\end{itemize}
-%\item Value of marriage:
-%\begin{itemize}
-%\item Do pregnancies trigger marriages? If so, why?
-%\item If shotgun marriage is risky, why people agree?
-%\end{itemize}
-%\end{itemize}
-%
-%\end{frame}
-
 \begin{frame}
-\frametitle{Roadmap}
+\frametitle{Questions}
+\begin{itemize}
+\item Narrow: \begin{enumerate}
+\item Why marriage / childbearing timing matters for divorce? Is it a causal relation?
+\item How policies shape the creation and dissolution of the marriages? Should they?
+\end{enumerate}
+\item Broad:
 \begin{enumerate}
-\item Document the shotgun marriage and its outcomes empirically
-\item Use these empirical relations to estimate a model capturing:
-\begin{itemize}
-\item dynamic bargaining and limited commitment in marriage
-\item lifecycle, income and savings
-\item fertility and child expenditures
+\item How marriage works? What makes people enter ``risky'' marriages?
+\end{enumerate}
 \end{itemize}
-\item Look at what drives decisions to marry and the differences in divorce
-\begin{itemize}
-\item what explains the education gradient?
-\item can policies fix the difference or reduce the share?
-\item what is the value of marriage in general?
-\end{itemize}
-\end{enumerate}
 \end{frame}
@@
 \begin{frame}
-\frametitle{Summary of Results}
+\frametitle{What I Do}
+\begin{enumerate}
+\item Describe empirically the behavior of kids-first and marriage-first couples
+\item Use this empirical evidence to estimate a structural lifecycle model
 \begin{itemize}
-\item 23\% of women had the first child before their first marriage (kids first)
-\item They are 1.8 times more likely to be divorced than ``marriage first''
-%\item College grads: 2.8 times more likely to be divorced
-%\begin{itemize}
-%\item though smaller kids first share (just 10\%)
-%\end{itemize}
-\item My structural model corrects selection and data imperfections
-%\item Model results:
-%\begin{itemize}
-%\item Social stigma + threat of raising the child alone explains most of the difference
-%\item Child support $\Rightarrow$ more single mothers, little effect on marriage decisions
-%\item Being able to control fertility helps everyone
-%\end{itemize}
-\item Model answers: Love $>$ Money. In particular:
+\item Singles meet potential partners and sometimes get unplanned pregnancies
+\item Agents are heterogeneous in wages and savings, marriages differ in quality
+\item People marry endogenously and cannot commit not to divorce
+\end{itemize}
+\item Use the model to talk about causality and policy
 \begin{itemize}
-\item Threat of raising the child alone explains most of the difference
-\item Regulating marriages by policies is both hard and inefficient
-\item Easier control over fertility fixes shotgun marriages and has bigger return
+\item Endogenous response to unplanned pregnancy explains the data well
 \end{itemize}
-\end{itemize}
+\end{enumerate}
 \end{frame}
@@
 \begin{itemize}
-\item Empirical: Alessina, Guiliano, 2005;  Tannenbaum, 2020; Rossin-Slater, 2017
-\item Structural: Forester, 2020; Brown, Flinn, 2011
-\item Focus on particular policies, ignore interactions, limited quantitative power
+\item Empirical: {\small \color{gray}Alessina, Guiliano, 2005;  Tannenbaum, 2020; Rossin-Slater, 2017}
+\item Structural: {\small  \color{gray}Forester, 2020; Brown, Flinn, 2011; Kennes, Knowles, 2020}
+\item I am the first to consider creation and dissolution of shotgun marriages structurally
 \end{itemize}
@@
 \begin{itemize}
-\item Dynamic limited commitment for marriage and divorce: Mazzocco, 2007; Voena, 2015; Low et al., 2018; Shephard, 2019; Blasutto, 2020
-\item Fertility during lifecycle: Sommer, 2015, Ejrnæs, Jørgensen, 2018
-\item Link between divorce and fertility is missing
+\item Dynamic limited commitment for marriage and divorce: {\small  \color{gray}Mazzocco, 2007; Voena, 2015; Low et al., 2018; Shephard, 2019; Blasutto, 2020}
+\item Fertility during lifecycle: {\small  \color{gray}Sommer, 2015, Ejrnæs, Jørgensen, 2018}
+\item I am the first to estimate a marriage/divorce model with fertility choice
 \end{itemize}
@@
 \begin{itemize}
-\item Gibson-Davis et al (2016), Guzzo (2009),  Nuevo-Chiquero (2014)
-\item Mostly descriptive
-\end{itemize}
-\end{itemize}
-%
-%\begin{itemize}
-%\item Empirical literature documents the scope of shotgun marriage around the world:
-%\begin{itemize}
-%\item Gibson-Davis et al (2016), Guzzo (2009),  Nuevo-Chiquero (2014)
-%\item Mostly descriptive
-%\end{itemize}
-%\item Dynamic limited commitment models of divorce and marriage:
-%\begin{itemize}
-%\item Marcet, Marimon (2019), Mazzocco (2007) --- setting up the framework
-%\item Voena (2015), Shephard (2019); Lafortune, Low  (2019) --- applied questions
-%\item Mostly ignore or simplify interaction with fertility
-%\end{itemize}
-%\item Unplanned and teenage pregnancy literature:
-%\begin{itemize}
-%\item Ejrnæs, Jørgensen (2018) --- lifecycle model
-%\item Card, Wise (1978) to Ashcraft et al (2013) --- lots of empirical papers
-%\item Treat household structure as fixed or exogenous
-%\end{itemize}
-%\end{itemize}
-\end{frame}
-
-\section{Data}
-\begin{frame}
-\frametitle{Sources: ACS 2009--2017 and others}
-\begin{itemize}
-\item Main exercises and model estimation are done on American Community Survey
-%\item Huge sample size and enough information to show many things 
-\item Individual marriage and fertility history are observed partially
-\begin{itemize}
-\item Only the most recent marriage date --- restrict to ``married once'' when matters
-\item Pick age $\in [21,40]$ to have good information on fertility
-\end{itemize}
-\item Smaller datasets with better histories for robustness and extra evidence:
-\begin{itemize}
-\item Survey of Income and Program Participation (SIPP)
-\item More: NSFG and NSFH
+\item {\small  \color{gray}Gibson-Davis et al (2016), Guzzo (2009),  Nuevo-Chiquero (2014)}
+\item I can recover the causal content of these numbers
 \end{itemize}
@@
 
-\begin{frame}[label=kfandmf]
-\frametitle{Kids First and Marriage First}
-\[\Delta T = \text{Year The First Child Born} - \text{Year The First Marriage Happened}.\]
+\begin{frame}
+\frametitle{Statistical Pattern: Kids-First $\Rightarrow$ Divorce More}
+\begin{center}
+\includegraphics[scale=0.5]{../pics/div_5y_by_dt.pdf}
+\end{center}
 \begin{itemize}
-\item Marriage First \textbf{(MF)}: $\Delta T \in \{1,2,...\}$ (kids at least at the next year)
-\item Kids First \textbf{(KF)}: $\Delta T \in \{...,-2,-1,0\}$ (shotgun marriage in a wide sense)
-\item Not a full partition%: at 30 only 31\% on the US women have kids and exactly one marriage
-\item Most of the comparisons are within this group
-\end{itemize}
-\hyperlink{extra-restrictions}{\beamerbutton{Extra Restrictions}} 
-\end{frame}
-
-\begin{frame}[label=dt_graph]
-\frametitle{Kids First Women Are Divorced More Often: Just Look}
-\begin{center}
-\includegraphics[scale=0.75]{div_5y_by_dt.pdf}
-\end{center}
-\vspace{-1cm}
-\hyperlink{dt_graph_educ}{\beamerbutton{by education}} 
-\end{frame}
-
-\begin{frame}
-\frametitle{Kids First Women Are Divorced More Often: Quantification}
-\framesubtitle{ACS 2009--2017, Women 21--40, married once}
-\begin{center}
-\begin{tabular}{l r r r r }
-\hline
-& \multicolumn{2}{c}{share of divorced if ... }&  \\
-&  marriage-first & kids-first & (share of kids-first) &  \\\hline
-\multicolumn{5}{l}{\textbf{All sample}} \\\hline
-\textit{Cross-sectional share of divorced} &  9.9 & 18.1 & (26.1) \\
-\textit{Divorced 5 years after marriage} &  5.1 & 14.3  & (24.6) \\
-\textit{Divorced 10 years after marriage} & 10.4 & 22.8 & (24.4) \\\hline
-\multicolumn{5}{l}{\textbf{Cross-sectional share of divorced for subsamples}} \\\hline
-\textit{High school only} &  12.8 & 17.3 & (37.0) \\
-\textit{Some college} & 14.3 & 21.0 & (31.3) \\
-\textit{College or more} &   5.3 & 14.8 & (11.9) \\\hline
-\textit{First birth before 25} & 15.5 & 19.9 & (41.8) \\
-\textit{First birth at 25 or later} &  6.2 & 10.9 & (10.7)  \\\hline
-\end{tabular}
-\end{center}
-\end{frame}
-
-\begin{frame}
-\frametitle{Kids First Women Are Divorced More Often: Controls}
-\begin{center}
-%\vspace{-0.5cm}
-\begin{tabular}{l r r r }
-\multicolumn{4}{c}{\textit{Regression equation:} $\text{Divorced}_i = \Delta \cdot \text{kids-first}_i + \text{Controls}_i + \varepsilon_i$} \\\hline
-\hline
-& \multicolumn{2}{c}{Estimates}  \\
-& Difference ($\Delta$) & (standard error)  \\\hline
-\multicolumn{4}{l}{\textbf{All sample, regression}} \\\hline
-\textit{No controls (raw difference) }& 8.2 & (0.1)  \\
-\textit{Demographic controls }& 5.7 & (0.1)  \\
-\textit{Duration controls} &  5.8 & (0.2)  \\
-\textit{Demographic + duration} &  4.8 & (0.2)  \\
-\textit{Demographic + duration + income} & 4.1 & (0.3)  \\\hline
-\multicolumn{4}{l}{\textbf{All sample, propensity score matching}} \\\hline
-\textit{Demographic controls} & 5.0 & (0.3)  \\
-\textit{Demographic controls + income} & 4.9 & (0.2)  \\\hline
-\multicolumn{4}{p{0.825\linewidth}}{\footnotesize \textit{Controls:} demographics: FE for age $\times$ education, race, state, survey year.
-Duration: FE for age $\times$ age of the first marriage and age $\times$ age of the first birth.
-Income: 3rd degree polynomial of log gross income (if observed).}\\\hline\hline
-\end{tabular}
-\end{center}
-\end{frame}
-
-\begin{frame}
-\frametitle{Kids First Women Are Divorced More Often: Extra Evidence from SIPP}
-\begin{center}
-\begin{tabular}{l r r }
-\hline
-\multicolumn{3}{l}{\textbf{Including remarried people...}} \\\hline
-& \multicolumn{2}{c}{share of divorced if ... }  \\
-&  marriage-first & kids-first   \\\hline
-\textit{Ever divorced} & 34.4 & 48.0  \\
-\textit{Divorced if married once} &  12.4 & 19.3 \\
-\textit{Remarried} &  22.0 & 28.7 \\\hline
-\hline
-\multicolumn{3}{l}{\textbf{Separating step and own children (2+ kids only)}} \\\hline
-& share in sample & share of divorced   \\\hline
-{Marriage-first} & 72.1 & 10.5 \\
-{Kids-first-own} &  16.4 & 14.6 \\
-{Kids-first-other} &  11.4 & 27.2 \\\hline
-\multicolumn{3}{p{0.7\linewidth}}{\footnotesize \textit{Notes.} SIPP data, the cross-section of Wave 1, 2014, women 21+. Numbers are percents.}\\\hline\hline
-\end{tabular}
-\end{center}
-\end{frame}
-
-%\begin{frame}
-%\frametitle{Kids First Women Are Divorced More Often: Robustness \& Extra}
-%\begin{itemize}
-%\item SIPP exercise: smaller sample but full marital history and fertility history
-%\begin{itemize}
-%\item Similar ratios with larger magnitudes when include remarried people
-%\end{itemize}
-%\item Duration model: NSFH, historic data from 80s
-%\begin{itemize}
-%\item Hazard ratio of 1.45, college: 2.23
-%\end{itemize}
-%\item More details: NSFG, 2015--2017 data
-%\begin{itemize}
-%\item KF: 30\% true shotgun, 50\% marry after birth, 20\% marry new partner
-%\item True shotgun have the highest share of divorced
-%\end{itemize}
-%\end{itemize}
-%\end{frame}
-%
-%
-%\begin{frame}
-%\frametitle{Need For The Model}
-%\begin{enumerate}
-%\item Observables provide limited information. Missing things: relationship quality, potential partners, future divorces
-%\item Data are not perfect. Limited histories, unobservable income if out of labor force, step or own children
-%\end{enumerate}
-%\end{frame}
-
-\section{Model}
-\begin{frame}
-\frametitle{Summary: Lifecycle, Fertility, Divorce}
-\begin{itemize}
-\item Lifecycle model with finite horizon
-\item People are heterogeneous with respect to age $t$, wages $w$ and savings $a$
-\item Couples also differ by relationship quality $\psi$ and spouses' decision power $\theta$
-\item Fertility: endogenous decision to have a child (just one) + shocks
-\item Limited commitment: decisions are affected by outside options of divorce
+\item Controlling for many observable dimensions do not mitigate the difference
+\item In subgroups, the pattern is sharper for college graduates
+\item No exogenous variation, so the model is needed
 \end{itemize}
@@
 \begin{frame}
-\frametitle{Modeling Shotgun Marriage}
-\begin{center}
-\vspace{-0.6cm}
-\begin{tikzpicture}[every text node part/.style={align=center}, scale=0.75]
-   % Place nodes
-   \node [block] (1) {\large \textbf{People}  \\[-0.5ex]  \large  \textbf{Randomly Meet}};
-   \node [block, below left = 1.5 cm and 0.0 cm of 1] (2) {\textit{Nash} \\[-0.5ex] \textit{Bargaining}};   
-   \node [block, below right = 1.5 cm and 0.0 cm of 1] (3) {\textit{Nash} \\[-0.5ex] \textit{Bargaining}};
-
-   \node [block, below = 1.5 cm of 2] (5) {\textbf{Kids First} \\[-0.5ex] \textbf{Couple} ($ck$)};
-   \node [block, below = 0.8 cm of 3] (7) {No Kids \\[-0.5ex] Couple ($c$)};
-   \node [block, below = 0.8 cm of 7] (8) {\textbf{Mariage First} \\[-0.5ex] \textbf{Couple} ($ck$)};
-   \draw[->] (1) to [bend left = 20] node [left] {\scriptsize Unplanned \\[-0.75ex] \scriptsize pregnancy} (2);
-   \draw[->] (1) to [bend right = 20] node [right] {\scriptsize No \\[-0.75ex] \scriptsize pregnancy} (3);
-   \draw[->] (2) to node [left] {\scriptsize Agree} node [right] {\scriptsize \textbf{Shotgun} \\[-0.75ex]  \scriptsize \textbf{marriage}} (5);
-   \draw[->] (3) to node [right] {\scriptsize Agree} node [left]  {\scriptsize \textbf{Regular} \\[-0.75ex]  \scriptsize \textbf{marriage}} (7);
-   \draw[->] (7) to node [right] {\scriptsize Kids by choice} (8);   
-    \node [block, above left   = 0.8 cm and -0.75 cm of 2] (4) {\,Single male \\[-0.5ex] ($m,s$)\,\,};
-    \node [block, above left   = 1.8 cm and -0.75 cm of 2] (4-sm) {\,Single mom \\[-0.5ex]  ($f,sk$)};
-    %\node [block, left = 0.2 cm of 4-sm] (4-nsm) {\,Single female \\[-0.5ex]  ($f,sk$)};
-	\node [block, above right   = 0.8 cm and -0.75 cm of 3] (6) {\,\,Single male\,\,\\[-0.5ex] ($m,s$)};
-    \node [block, above right   = 1.8 cm and -0.75 cm of 3] (6-f) {Single female\\[-0.5ex] ($f,s$)\,};   
-    \draw[->] (2.180) to [bend left = 30] node [left] {\scriptsize Disagree} (4.270);
-    \draw[->,thick,dotted] (4-sm.20) to [bend left = 10] node [above right=-0.2cm and 0.7cm] {\scriptsize Costly abortion \\[-0.75ex] \scriptsize (if possible)} (6-f.50);
-    \draw[->] (3.0) to [bend right = 30] node [right] {\scriptsize Disagree} (6.270);
-     \draw[->] (6.180) to [bend left = 30] node [above] {\scriptsize Back}  (1.-12);
-   \draw[->] (6-f.180) to [bend right = 30] node [above] {\scriptsize Back} (1.12);
-   \draw[->,thick,dashed] (4-sm.90) to [bend left = 30] node [above] {\scriptsize Worse marriage prospectives}  (1.168);
-   \draw[->] (4.0) to [bend right = 30] node [above] {\scriptsize Back} (1.192);
-
-\end{tikzpicture}
-\end{center}
+\frametitle{Conclusions}
+\begin{enumerate}
+\item Value of the marriage:
+\begin{itemize}
+\item Causality: kids-first marriage increases chances of future divorce by $\frac13$ to $\frac12$
+\item Why do kids-first divorce more? Imperfect control over fertility, social stigma, disutility from stepchildren are the leading causes.
+\item Selection: couples choose to have kids only if their risk of divorce is low enough
+\item \textit{Risks} of unplanned pregnancies shape people's behavior quite a lot
+\end{itemize}
+\item Policy insights:
+\begin{itemize}
+\item Child support is inefficient in regulating marriage decisions for kids-first
+\item Child support creates more single mothers but is still welfare improving
+\item Pushing people towards marriage generally works, but only temporary
+\end{itemize}
+\end{enumerate}
 \end{frame}
@@
 \begin{frame}
-\frametitle{Value Function: Female, Single}
-\framesubtitle{$p^m_t$ --- probability of meeting, $p^p_t$ --- probability of unplanned pregnancy, $m_t$ --- agree to marry}
-\begin{align*}
-V^{f,s}_t(a,w) = \max\limits_{c,a'} \Bigg\{ \frac{c^{1-\sigma}}{1-\sigma} + \beta \cdot \E_t \Big\{  & 
-%& \\ \text{ (did not meet anyone): \ } & 
-(1-p^{\text{m}}_t)\cdot V_{t+1}^{f,s}(a',w') \\
-\text{ \footnotesize (met and not pregnant): \ }& +  p^{\text{m}}_t\cdot (1- p^{p}_t) \cdot \left[ m^{np}_t \cdot V^{f,c}_{t+1}  + (1-m^{np}_t)\cdot  V^{f,s}_{t+1}\right] \\
-\text{ \footnotesize (met and pregnant): \ } & + p^{\text{m}}_t \cdot p^{\text{p}}_t\cdot   \left[ m^p_t \cdot V^{f,ck}_{t+1}  + (1-m^p_t)\cdot  V^{f,\text{preg}}_{t+1}\right] \Big\} \Bigg\}, \\
-\text{ \footnotesize (uncertain abortion access): } &  V^{f,\text{preg}}_{t+1} = p^{\text{access}} \cdot \max\{V^{f,sk}_{t+1},V^{f,s}_{t+1} - \phi_a\} \\
-   & \hspace{1cm} + (1-p^{\text{access}})\cdot V^{f,sk}_{t+1},\\
-\text{ \footnotesize (budget constraint): \ } & c + a' = R\cdot a  + w , \\
-\text{ \footnotesize (wage evolution): \ } & \log w = z + \text{Trend}_t, \ \ z' = z + \varepsilon^{f}.
-\end{align*}
-\end{frame}
-
-\begin{frame}
-\frametitle{Value Function: Couple, No Children}
-\framesubtitle{$k_t$ --- give a birth, $d_t$ --- get a divorce, $\tilde{V}$ --- values at divorce}
-\begin{align*}
-V^{c}_t(a,w_f,w_m,\psi,\theta) = \max\limits_{c_f,c_m,a',k} \Bigg\{ & \theta^f \cdot \frac{c_f^{1-\sigma}}{{1-\sigma}}  + \theta^m \cdot \frac{c_m^{1-\sigma}}{1-\sigma}  + \psi + \beta \cdot \E_t \Big\{   \\
-\text{ \footnotesize (stay, no birth): \ } & + (1-d_t) \cdot (1-k_t) \cdot  \left[ \theta^f \cdot  V^{f,c}_{t+1} + \theta^m \cdot V^{m,c}_{t+1}  \right] \\
-\text{ \footnotesize (stay, give a birth): \ } &  + (1-d_t) \cdot k_t \cdot  \left[ \theta^f \cdot  V^{f,ck}_{t+1} + \theta^m \cdot V^{m,ck}_{t+1}  \right]  \\
-\text{ \footnotesize (divorce): \ } & + d_t  \left[ \theta^f \cdot  \tilde{V}^{f,s}_{t+1} + \theta^m \cdot \tilde{V}^{m,s}_{t+1}  \right]  \Big\} \Bigg\}\\
-\text{ \footnotesize (budget constraint, rts): \ } &{\normalsize \left[c_f^{1+\rho} + c_m^{1+\rho}\right]^{\frac1{1+\rho}} + a' = R\cdot a + w_f + w_m}\\
-\text{ \footnotesize (match quality evolution): \ } &{\normalsize \psi' = \psi + \varepsilon^{\psi}}\\
-\text{ \footnotesize \textbf{(participation constraints)}: \ } &{\normalsize V^{f,c\bullet}_{t+1} (\cdots,\theta') \geq \tilde{V}^{f,s}_{t+1}, \ \ V^{m,c\bullet}_{t+1} (\cdots,\theta') \geq \tilde{V}^{m,s}_{t+1}.}\\
-\end{align*}
-\end{frame}
-
-\begin{frame}
-\frametitle{Value Function: Couple and Child}
-\framesubtitle{$d_t$ --- get a divorce, $\tilde{V}$ --- values at divorce}
-\begin{align*}
-V^{ck}_t(a,w_f,w_m,\psi,\theta) = \max\limits_{c_f,c_m,x,l_f,a'} \Bigg\{ & \theta^f \cdot \frac{c_f^{1-\sigma}}{{1-\sigma}}  + \theta^m \cdot \frac{c_m^{1-\sigma}}{1-\sigma}  + \psi + \\ \text{ \footnotesize (child is a public good): \ }  & \phi + \alpha \cdot \frac{Q^{1-\sigma}}{1-\sigma} +  \beta \cdot \E_t \Big\{   \\
-\text{ \footnotesize (stay): \ } & + (1-d_t) \cdot  \left[ \theta^f \cdot  V^{f,ck}_{t+1} + \theta^m \cdot V^{m,ck}_{t+1}  \right] \\
-\text{ \footnotesize (divorce): \ } & + d_t \cdot \left[ \theta^f \cdot  \tilde V^{f,sk}_{t+1} + \theta^m \cdot \tilde V^{m,s}_{t+1}  \right]  \Big\} \Bigg\}\\
-\text{ \footnotesize (budget constraint): \ } &{\normalsize \left[c_f^{1+\rho} + c_m^{1+\rho}\right]^{\frac1{1+\rho}} + x + a' = R\cdot a + w_f\cdot l_f + w_m}\\
-\text{ \footnotesize (child consumption flow): \ } &{\normalsize Q = \left[x^\lambda + \kappa \cdot (1-l_f)^\lambda\right]^{\frac1\lambda}, \ \ \ Q\geq \bar{Q}} \text{ \footnotesize  (subsistence constraint) } \\
-\text{ \footnotesize (female skills $\downarrow$ if not working): \ } &{\normalsize \log w_f = z_f + \text{Trend}_{f,t}, \ \ z'_f = z_f + \varepsilon^f - \delta(l^f)}\\
-\end{align*}
-\end{frame}
-
-\begin{frame}
-\frametitle{Renegotiation and Divorce}
-\framesubtitle{Are generated by the participation constraints}
+\frametitle{Key Takeaways}
 \begin{itemize}
-\item Bargaining weights $\theta$ adjust to satisfy both participation constraints
-\[\Delta \theta^i \text{ reflects Lagrange multiplier on $i$'s participation constraint}: V^{i,c\bullet}_{t+1} \geq \tilde{V}^{i,s}_{t+1}\]
-\item If no adjustment is possible, the couple divorces
-\item Assets are split evenly upon divorce (follows Voena, 2015)
-\item Partners have fixed utility costs $\phi_d$ of divorce
-\item If couple with a child divorces wife becomes a single mother
+\item Kids-first marriages are adult analogues of teenage pregnancies
+\item Reaction to unplanned pregnancies rationalizes the divorce pattern
+\item Managing this reaction is difficult, but its causes can be controlled
 \end{itemize}
@@
 
-\begin{frame}
-\frametitle{Child Support}
-\framesubtitle{Paid forever, but with a chance to avoid}
-\begin{itemize}
-\item Imperfect enforcement: probability to be awarded is $p^{\text{award}}$
-\begin{itemize}
-\item Married couples: 46\%, never married couples: 28\% chances
-\end{itemize}
-\item If enforced:
-\begin{itemize}
-\item man loses productivity $z$ so earnings fall by 20\%
-\item woman gains productivity $z$ matching this value
-\end{itemize}
-\item As $z' = z + \varepsilon^z$, all changes are permanent
-\end{itemize}
-\end{frame}
-
-%\begin{frame}
-%\frametitle{Abortions / Birth Control}
-%\begin{itemize}
-%\item When unplanned pregnancy happens, with probability $p^{\text{access}}$ woman can avoid being single mother
-%\item If possible, she pays \emph{fixed utility costs} $u^{a}$ for this
-%\item Targets: $\sim40\%$ unplanned pregnancies are aborted + age profile of abortions
-%\item Some women \textit{choose} to be single mothers even with abortion access
-%\end{itemize}
-%\end{frame}
-
-\begin{frame}
-\frametitle{Marriage}
-\framesubtitle{Introducing Social Stigma}
-\begin{itemize}
-\item Singles meet singles with random $a$ and $z$ (estimated)
-\item Potential couple gets a draw of match quality $\psi$
-\item If both partners agree to marry, initial $\theta$ is set by symmetric Nash Bargainig
-\item More utility shifters (unexplained things):
-\begin{itemize}
-\item If unplanned pregnancy happens, partners face {\textbf{social stigma}} $\phi_s$ if refuse to marry
-\item If male meets a single mother, he lose $\phi_m$ for having step children
-\end{itemize}
-\end{itemize}
-\begin{center}
-{\footnotesize
-\begin{tabular}{|l||c|c||c|c||c|}\hline
- & \multicolumn{2}{|c||}{Male gets:} & \multicolumn{2}{|c||}{Female gets:} \\\hline
-Situation & Agree & Disagree & Agree & Disagree \\\hline
-Regular match & $V^{m,c}_t(\theta)$ & $V^{m,s}_t$ & $V^{f,c}_t(\theta)$ & $V^{f,s}_t$ \\
-Unplanned pregnancy & $V^{m,ck}_t(\theta)$ & $V^{m,s}_t - \phi_s$ & $V^{f,ck}_t(\theta)$ & $V^{f,s}_t - \phi_s$ \\
-Meeting a single mother & $V^{m,ck}_t(\theta) - \phi_m$ & $V^{m,s}_t$ & $V^{f,ck}_t(\theta)$ & $V^{f,s}_t$ \\\hline
-\end{tabular}
-}
-\end{center}
-\end{frame}
-
-\section{Mechanics \& Intuition}
-\begin{frame}
-\frametitle{Unplanned Pregnancy $\Rightarrow$ Worse Match?}
-\framesubtitle{Why people enter knowingly risky marriages?}
-\begin{itemize}
-\item Women enter the shotgun marriages because:
-\begin{itemize}
-\item Child costs need to be shared (returns to scale)
-\item Finding a new partner is trickier once you have a child
-\end{itemize}
-\item Men enter the shotgun marriage because:
-\begin{itemize}
-\item They like their kids and lose access to them otherwise
-\end{itemize}
-\item Even with a good match, mistimed pregnancy is a risk
-\item Scope of these effects depend on exact calibration
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{Unplanned Pregnancy $\Rightarrow$ Women Lose a Lot, Men Gain a Little}
-\framesubtitle{Nash Bargaining Surplus: $NBS(\theta) = (V^{f,\text{agree}}(\theta) - V^{f,\text{disagree}})\times(V^{m,\text{agree}}(\theta) - V^{m,\text{disagree}})$}
-\begin{center}
-\begin{tabular}{c c}
-\hspace{-1cm} \includegraphics[scale=0.5]{change_upp.pdf} &
-\hspace{-0.5cm}\includegraphics[scale=0.5]{surp_upp.pdf} 
-\end{tabular}
-
-\end{center}
-
-\end{frame}
-
-\begin{frame}
-\frametitle{...Hence Less Picky Partners + Lower Female Decision Power}
-\begin{center}
-\begin{tabular}{c c}
-\hspace{-1cm} \includegraphics[scale=0.5]{barg_reg_match.pdf} & \includegraphics[scale=0.5]{barg_upp.pdf}
-\end{tabular}
-\end{center}
-\end{frame}
-
-\section{Fit and Results}
-
-\begin{frame}
-\frametitle{Too Much Heterogeneity}
-\begin{itemize}
-\item College and non-college people have very distinct patterns
-\item They cannot be summarized well with one-dimensional productivity
-\item Therefore I focus on two subsamples:
-\begin{itemize}
-\item High education: bachelor degree or more
-\item Low education: high school or less
-\end{itemize}
-\item Model fits well for both groups, but high education results are more settled
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{Targets}
-\begin{tabular}{l p{0.5\linewidth}}\hline
-\multicolumn{1}{l}{Target} & \multicolumn{1}{l}{Informative about} \\\hline
-\footnotesize \% with kids 1--10 years after marriage & \footnotesize  Fertility parameters: $\phi$, $\alpha$ \\
-\footnotesize \% MF females in population at 23--35 & \footnotesize  $\phi$, $\alpha$ + meeting probabilities \\
-\footnotesize \% KF females in population at 23--35 &\footnotesize   meeting + unplanned preg probabilities\\
-\footnotesize \% never married with and without kids at 23--35 & \footnotesize  meeting + unplanned preg probabilities + $\phi_r$\\
-\footnotesize \% divorced with and without kids at 23--35 & \footnotesize  distribution of $\psi$ + fertility parameters + $\phi_r$ \\
-\footnotesize child expenditures share, \% in LF with kids & \footnotesize  household technology: $\kappa$, $\alpha$ and $\bar{Q}$ \\
-\footnotesize \% divorced 1--10 years after marriage if KF or MF & \footnotesize  social stigma, $\bar{Q}$ and more ... \\
-\footnotesize \% unplanned pregnancies aborted + age profile & \footnotesize  abortion costs and access \\\hline
-\end{tabular}
-\end{frame}
-
-\begin{frame}
-\frametitle{Data}
-\framesubtitle{High school in red, college in black}
-\begin{center}
-\begin{tabular}{c c}
-\hspace{-0.5cm} \includegraphics[scale=0.33]{div_kfmf_dc.pdf} & \includegraphics[scale=0.33]{popshares_dc.pdf} \\
-\hspace{-0.5cm} \includegraphics[scale=0.33]{evermar_dc.pdf} & \includegraphics[scale=0.33]{everkids_dc.pdf} \\
-\end{tabular}
-\end{center}
-\end{frame}
-
-\begin{frame}
-\frametitle{Estimates}
-\framesubtitle{Simulated Method of Moments (SMM) with optimization routine from Arnoud et al, 2019; 17 parameters}
-\begin{tabular}{l c c c}\hline
-& High Education & Low Education & Source \\\hline
-\multicolumn{4}{c}{Utility from kids: $\phi + \alpha \cdot \frac{Q^{1-\xi}}{1-\xi}$, \ $Q = [x^{\lambda} + \kappa\cdot(1-l_f)^{\lambda}]^{\frac1{\lambda}}$. $Q\geq \bar{Q}$}\\\hline
-$\phi$, $\alpha$, $\kappa$  \footnotesize (utility params) &\footnotesize  $(1.39, 0.36, 0.78)$ &\footnotesize $(1.80,0.45, 1.16)$ & SMM \\
-$\bar{Q}$ \footnotesize (subsistence constraint) & 1.7 & 0.70 & SMM\\
-\multicolumn{4}{c}{Meeting probability $p^{\text{meet}}$, pregnancy probability $p^{\text{preg}}$: quadratic polynomials}\\\hline
-$p^{\text{meet}}$  \footnotesize at (21,30,40) & \footnotesize $(0.11, 0.21, 0.34)$ & \footnotesize  $(0.44, 0.10, 0.65)$ & SMM\\
-$p^{\text{preg}}$ \footnotesize  at (21,28,35) & \footnotesize  $(0.01, 0.04, 0.01)$ & \footnotesize  $(0.09, 0.07, -0.05)$ & SMM \\\hline\hline
-\multicolumn{4}{c}{Additive utility shifters}\\\hline
-$\sigma(\varepsilon^\psi), \sigma(\psi_0)$   \footnotesize (match quality distribution) & $(0.10,0.14)$ & $(0.88,4.8)$ & SMM \\
-$\phi_s$ \footnotesize  (social stigma) & 4.21 & 5.31 & SMM\\
-$\phi_m$ \footnotesize   (single mom remarriage loss) & 43.3 & 79.6 & SMM \\\hline
-$\phi_d$ \footnotesize   (divorce costs) & 5.3 & 1.2 & SMM \\\hline
-$(p^{\text{access}}, \phi_a)$ \footnotesize   (abortion parameters) & $(0.22,37.5)$  & $(0.18,5.51) $& SMM \\\hline
-\end{tabular}
-\end{frame}
-
-\begin{frame}
-\frametitle{Things Fit Well --- Part 1}
-\framesubtitle{High Education Sample}
-\begin{center}
-
-\begin{tabular}{c c}
-\hspace{-1cm}\includegraphics[scale=0.5]{evermar.pdf} &\hspace{-0.5cm} \includegraphics[scale=0.5]{popshares.pdf} \\
-\end{tabular}
-\end{center}
-\end{frame}
-
-\begin{frame}
-\frametitle{Things Fit Well --- Part 2}
-\framesubtitle{High Education Sample}
-\begin{center}
-\begin{tabular}{c c}
-\hspace{-1cm}\includegraphics[scale=0.5]{div_kfmf.pdf}  & \hspace{-0.5cm} \includegraphics[scale=0.5]{everkids.pdf} 
-\end{tabular}
-\end{center}
-\end{frame}
-
-\begin{frame}[label=counterfactuals-abortions]
-\frametitle{What Explains The Difference?}
-\framesubtitle{Removing Social Stigma (1/4) + Fertility Control (3/4).}
-\begin{center}
-\begin{tabular}{c c}
-\hspace{-1cm}\includegraphics[scale=0.5]{div_kfmf_equalized.pdf}  & \hspace{-0.5cm} \includegraphics[scale=0.5]{relshares_equalized.pdf} 
-\end{tabular}
-\end{center}
-Welfare: men gain $\approx \$35k$, women gain $\approx \$12k$, \hyperlink{welfare-abortions}{\beamerbutton{details}} 
-\end{frame}
-
-\begin{frame}[label=counterfactuals-child-support]
-\frametitle{Removing Existing Child Support --- Part 1}
-\framesubtitle{Less single mothers, less divorces, more abortions}
-No child support instead of current (partially enforced 20\%)
-\begin{center}
-\begin{tabular}{c c}
-\hspace{-1cm}\includegraphics[scale=0.5]{div_kfmf_child_support.pdf}  & \hspace{-0.5cm} \includegraphics[scale=0.5]{sm_child_support.pdf} 
-\end{tabular}
-\end{center}
-Just removing the social stigma \hyperlink{counterfactuals-social-stigma}{\beamerbutton{here}}
-
-Welfare: men gain $\approx \$3k$, women lose $\approx \$1k$ \hyperlink{welfare-child-support}{\beamerbutton{details}} 
-\end{frame}
-
-\begin{frame}[label=counterfactuals-child-support]
-\frametitle{Removing Existing Child Support --- Part 2}
-\framesubtitle{Hardly Affects Marriage in Any Way}
-Can you notice the differences?
-\begin{center}
-\begin{tabular}{c c}
-\hspace{-1cm}\includegraphics[scale=0.5]{evermar_cs.pdf}  & \hspace{-0.5cm} \includegraphics[scale=0.5]{popshares_cs.pdf} 
-\end{tabular}
-\end{center}
-\end{frame}
-
-\begin{frame}[label=cf]
-\frametitle{More Counterfactuals}
-\begin{itemize}
-\item No divorce costs \hyperlink{no_divorce_costs}{\beamerbutton{pictures}}
-\begin{itemize}
-\item More divorces, less kids
-\item But less divorces conditional on having kids!
-\item Welfare: females lose, males gain, gain on average
-\end{itemize}
-%\item Removing gender pay gap: same trend + variance for men and women \hyperlink{pay_gap}{\beamerbutton{pictures}} 
-\item No unplanned pregnancies at all 
-\begin{itemize}
-\item Remaining KF (step-child-couples) and MF become very similar
-\item Men gain $\sim \$25k$, women gain $\sim \$13k$ 
-\item Around 1/10 less divorces
-\end{itemize}
-\end{itemize}
-\end{frame}
-
-\begin{frame}[label=cf]
-\frametitle{Conclusions}
-\begin{itemize}
-\item Model mechanisms can rationalize creation and differences in divorce
-\item Child support is welfare-decreasing and does not get people together
-\item Controlling fertility helps everyones
-\item \textbf{Love $>>$ Money}
-%\item Nothing beats the birth control in terms of welfare
-%\item Child support does not get people together and creates more single mothers
-%\item Empirically, more control over fertility $\Rightarrow$ less KF, lower differences in performance
-\end{itemize}
-\end{frame}
-
-\backupbegin
-
-%\begin{frame}[label=remarriage_penalty]
-%\frametitle{Removing Remarriage Penalty For Single Mothers}
-%\framesubtitle{Divorced in 10 years if KF: 13.5\% $\to$ 12.4\%. Divorced in cross-section if KF: 9.3\% $\to$ 8.9\% + More Kids}
-%\begin{center}
-%\begin{tabular}{c c}
-%\hspace{-1cm} \includegraphics[scale=0.5]{div_kfmf_no_remar.pdf}  &  \includegraphics[scale=0.5]{ever_kids_no_remar.pdf} 
-%\end{tabular}
-%\end{center}
-%\hyperlink{cf}{\beamerbutton{back}} 
-%\end{frame}
-
-\begin{frame}[label=no_divorce_costs]
-\frametitle{No Divorce Costs}
-\framesubtitle{Less kids, more shotgun marriages (relatively), less divorces if have kids}
-\begin{center}
-\begin{tabular}{c c}
-\hspace{-1cm} \includegraphics[scale=0.5]{popshares_nodivcosts.pdf}  &  \includegraphics[scale=0.5]{div_kfmf_nodivcosts.pdf} 
-\end{tabular}
-\end{center}
-\hyperlink{cf}{\beamerbutton{back}} 
-\end{frame}
-
-\begin{frame}[label=pay_gap]
-\frametitle{Removing Pay Gap}
-\framesubtitle{Divorced in 10 years if KF: 13.5\% $\to$ 12.6\%. Divorced in cross-section if KF: 9.3\% $\to$ 8.8\% + Fewer Kids}
-\begin{center}
-\begin{tabular}{c c}
-\hspace{-1cm} \includegraphics[scale=0.5]{eq_pay.pdf}  &  \includegraphics[scale=0.5]{everkids_eq_pay.pdf} 
-\end{tabular}
-\end{center}
-\hyperlink{cf}{\beamerbutton{back}} 
-\end{frame}
-
-\begin{frame}[plain,label=extra-restrictions]
-\frametitle{Extra Restrictions}
-\begin{itemize}
-\item I do not classify as KF or MF people with $\Delta T < -5$ or $\Delta T > 5$
-\item I exclude small share of people with marital statuses ``separated'', ``widowed'' and ``spouse absent''
-\item This does not affect the results much but helps to ensure consistency
-\end{itemize}
-\hyperlink{kfandmf}{\beamerbutton{back to KF and MF}} 
-\end{frame}
-
-\begin{frame}[plain,label=dt_graph_educ]
-\frametitle{Kids First Women Are Divorced More Often: By Education}
-
-\begin{tabular}{c c}
-\hspace{-0.5cm}\includegraphics[scale=0.5]{div_5y_by_dt_col.pdf} & \hspace{0.5cm}\includegraphics[scale=0.5]{div_5y_by_dt_hs.pdf}
-\end{tabular}
-\hyperlink{dt_graph}{\beamerbutton{back to the main graph}} 
-\end{frame}
-
-\begin{frame}[label=counterfactuals-social-stigma]
-\frametitle{Removing Social Stigma}
-\framesubtitle{Less single mothers, less divorces, more abortions}
-No child support instead of current (partially enforced 20\%)
-\begin{center}
-\begin{tabular}{c c}
-\hspace{-1cm}\includegraphics[scale=0.5]{div_kfmf_ss.pdf}  & \hspace{-0.5cm} \includegraphics[scale=0.5]{sm_among_mothers_ss.pdf} 
-\end{tabular}
-\end{center}
-\hyperlink{counterfactuals-child-support}{\beamerbutton{back}} 
-\end{frame}
-
-\begin{frame}[label=welfare-abortions]
-\frametitle{What Explains The Difference? Welfare Implications}
-\framesubtitle{Removing Social Stigma (1/4) + Fertility Control (3/4)}
-\begin{center}
-\begin{tabular}{c c}
-\hspace{-1cm}\includegraphics[scale=0.5]{welfare_female_equalized.pdf}  & \hspace{-0.5cm} \includegraphics[scale=0.5]{welfare_male_equalized.pdf} 
-\end{tabular}
-\end{center}
-\hyperlink{counterfactuals-abortions}{\beamerbutton{back}} \hyperlink{welfare-methodology}{\beamerbutton{methodology}} 
-\end{frame}
-
-\begin{frame}[label=welfare-child-support]
-\frametitle{Child Support Removal}
-\framesubtitle{More abortions but not much changes...}
-\begin{center}
-\begin{tabular}{c c}
-\hspace{-1cm}\includegraphics[scale=0.5]{welfare_female_child_support.pdf}  & \hspace{-0.5cm} \includegraphics[scale=0.5]{welfare_male_child_support.pdf} 
-\end{tabular}
-\end{center}
-\hyperlink{counterfactuals-child-support}{\beamerbutton{back}} \hyperlink{welfare-methodology}{\beamerbutton{methodology}} 
-\end{frame}
-
-\begin{frame}[label=welfare-methodology]
-\frametitle{Definitions}
-\framesubtitle{Equivalent Variation (EV) Style}
-Value of any policy change can be computed in terms of equivalent amount of assets.
-
-If the change is preferred:
-\[EV = a^*_t(w) \text{ such that } V_{t}^{f,s,\text{no change}}(a=a^*,w) = V_{t}^{f,s,\text{with change}}(a=0,w)\]
-Otherwise:
-\[EV = -a^*_t(w) \text{ such that } V_{t}^{f,s,\text{with change}}(a=a^*,w) = V_{t}^{f,s,\text{no change}}(a=0,w)\]
-
-Analogously for $m$, a bit trickier for couples.
-
-\hyperlink{welfare-abortions}{\beamerbutton{back to abortions}} \hyperlink{welfare-child-support}{\beamerbutton{back to child support}}
-
-\end{frame}
-
-\backupend
 \end{document}

### 2020-10-02 taste-shocks-slides/taste-shocks.tex
Current stage

\documentclass[aspectratio=169]{beamer}
\usepackage[english]{babel}
\usepackage{amsmath,amsfonts}
\usepackage{multicol}

\usepackage{IEEEtrantools}
\usepackage{multirow}
% beamer setup

%\usecolortheme{dove}

\setbeamertemplate{navigation symbols}{}
\setbeamertemplate{itemize items}[ball]

\usepackage{tikz}
\usetikzlibrary{shapes,arrows}
\usetikzlibrary{positioning}
\tikzstyle{block} = [rectangle, draw, rounded corners]
\tikzstyle{line} = [draw, -latex']

\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator*{\argmax}{argmax}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\I}{\mathbb{I}}

\AtBeginSection[]{
  \begin{frame}[plain]
  \addtocounter{framenumber}{-1}
  \vfill
  \centering
  %\begin{beamercolorbox}[sep=8pt,center,shadow=true,rounded=true]{title}
    %\usebeamerfont{title}
    \Huge{\insertsectionhead\par}%
  %\end{beamercolorbox}
  \vfill
  \end{frame}
}

\newcommand{\backupbegin}{
   \newcounter{framenumberappendix}
   \setcounter{framenumberappendix}{\value{framenumber}}
}
\newcommand{\backupend}{
   \addtocounter{framenumberappendix}{-\value{framenumber}}
   \addtocounter{framenumber}{\value{framenumberappendix}} 
}

\title{Bargaining with Taste Shocks}
\author{Egor Kozlov}

\institute{
  Department of Economics\\
  Northwestern University}

\setbeamertemplate{footline}[frame number]

%  \usepackage{pgf}
%\logo{\pgfputat{\pgfxy(0,0)}{\pgfbox[right,base]{\footnotesize{\insertframenumber\,/\,\inserttotalframenumber}}}}
%\newcommand{\nologo}{\setbeamertemplate{logo}{}}

\let\olditem\item
\renewcommand{\item}{%
\olditem\vspace{\fill}}

\begin{document}

\begin{frame}[plain]
\addtocounter{framenumber}{-1}
\date{\scriptsize}
\titlepage
\end{frame}

\begin{frame}
\frametitle{Goal}
\begin{itemize}
\item This is a technical note
\item I try add taste shocks to the limited commitment model so the value functions are more smooth
\item Non-smoothness is large problem: more points are needed to achieve accuracy, policy functions (like savings) are discontinuous
\item Smoothing discrete choices is cool and modern (machine learning guys confirm)
\item DC-EGM (Iskhakov et al) approach is promising
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Existing Approach}
Standard renegotiation problem (discretized):
\begin{itemize}
\item \textit{Discrete} set $\Theta$ of possible bargaining powers, one-dimensional, $\Theta \subset [0,1]$
\item $\Theta = \{\theta_1,...\theta_n\}$
\item Couple starts with some $\theta_k$ and values of singleness $V^{fs}$, $V^{ms}$ and values values $\{V^f_1,...,V^f_n\}$, $\{V^m_1,...,V^m_n\}$ on $\Theta$. Here $V_i^{h} = V^h(\theta_i)$ for $h \in \{f,m\}$.
\item Then nested choice:
\begin{itemize}
\item Divorce if $\not\exists i \ : \ V^f_i - V^{fs}  \geq 0\ \& \ V^m_i - V^{ms} \geq 0$
\item Otherwise: 
\begin{itemize}
\item keep current $\theta_{sq}$ if $V^f_{sq} - V^{fs} \geq 0\ \& \ V^m_{sq} - V^{ms} \geq 0$
\item otherwise renegotiate to $\theta_{r^*}$: $\Theta_r = \{\theta_r \in \Theta: V^f_r - V^{fs} \geq 0\ \& \ V^m_r - V^{ms} \geq 0\}$, pick $\theta_{r^*}$ from $\Theta_r$ that is the closest to initial $\theta_{sq}$
\end{itemize}
%\item Equivalent metrics: $\Rightarrow \max\limits_{i} \left[ \min\{ V^f_i - V^{fs}, V^m_i - V^{ms} \} \right] < 0$.
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Reformulation}
Define the Rawlsian surplus $M_i = \min \{ V^f_i - V^{fs}, V^m_i - V^{ms} \}$

\textbf{Egalitarian bargaining solution (EBS)}: $\max\limits_i \min \{ V^f_i - V^{fs}, V^m_i - V^{ms} \} = M_{ebs}$

The choice problem then is:
\begin{itemize}
\item Divorce if $\not\exists i \ : \ V^f_i - V^{fs}  \geq 0\ \& \ V^m_i - V^{ms} \geq 0 \ \ \ \Leftrightarrow \ \ \ \boxed{ \max\limits_i M_i = M_{\text{EBS}} < 0 }$
\item Otherwise: 
\begin{itemize}
\item keep current $\theta_{sq}$ if $V^f_{sq} - V^{fs} \geq 0\ \& \ V^m_{sq} - V^{ms} \geq 0 \ \ \  \Leftrightarrow \ \ \ \boxed{ M_{sq} \geq 0 }$
\item otherwise renegotiate to $\theta_{r^*}$: $\Theta_R = \{\theta_r \in \Theta \ : \ V^f_r - V^{fs} \geq 0\ \& \ V^m_r - V^{ms} \geq 0\}$, pick $\theta_{r^*}$ from $\Theta_R$ that is the closest to initial $\theta_{sq}$. Equivalent representation: 
\[ \boxed{ \Theta_R = \{\theta_r \in \Theta \ : M_r \geq 0 \}. \ \ \theta_{r^*} \text{ is the closest value to $\theta_{sq}$ from $\Theta_R$ } }\]
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Single-Crossing Assumption in Renegotiation}
\begin{itemize}
\item \textbf{Assumption:} $V^f(\theta)$ and $V^m(\theta)$ are monotonic in $\theta$
\item \textbf{Then:} $M(\theta) = \min \{ V^f(\theta) - V^{fs}, V^m(\theta) - V^{ms} \}$ is single-crossing.
\item \textbf{Then:} If at $\theta_{sq}$ $M_{sq} < 0$ and $M_{ebs} > 0$, then $\theta_{sq} < \theta_{r^*}  < \theta_{ebs}$ (WLOG $\theta_{sq} < \theta_{\text{EBS}}$)
\item In real calculations, monotonicity can be violated (slightly). I hope that smoothing fixes this.
\item Two cases: $S^h = V^h(\theta) - V^{hs}$, $h \in \{f,m\}$
\end{itemize}

\begin{center}
\begin{tabular}{c c}
\includegraphics[scale=0.11]{m_good.jpg} & \includegraphics[scale=0.11]{m_bad.jpg} 
\end{tabular}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Introducing Taste Shocks}
\begin{itemize}
\item Let $\xi$ is a shock to both partners' outside options. Participation constraints:
\[V^{f}(\theta) \geq V^{fs} + \xi, \ \ V^{m}(\theta) \geq V^{ms} + \xi.\]
\[S^f(\theta) = V^{f}(\theta) - V^{fs}, \ \ S^m = V^{m}(\theta) - V^{ms}, \ \ M(\theta) = \min\{S^f(\theta),S^m(\theta)\}.\]
Couple's decisions given $\xi$ are:
\begin{itemize}
\item Divorce if $\forall i \ \ M_i < \xi \ \ \Rightarrow \boxed{\xi > M_{ebs}}$
\item Stay together when having $\theta_k$ if $\boxed{\xi < M_k}$. Note that $M_k \leq M_{ebs}$.
\item Renegotiate current $\theta$ if $\xi \in (M_{sq}, M_{ebs})$. In this case, \textit{assuming single-crossing}, there exist two monotonic sequences from values of $\theta \in \Theta$:
\[ \theta_{sq} < \theta_{r_1} < ... < \theta_{r_h} < \theta_{ebs} \text{\ \ and \ \ } M_{sq} < M_{r_1} < ... < M_{r_h} < M_{ebs}.\]
$\theta_{r_k}$ is picked if $M_{r_{k-1}} < \xi < M_{r_k}$.
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Taste Shocks Create Distribution Over $\{\theta_{sq},\theta_1,\theta_2,\theta_3,\theta_{ebs},\text{divorce}\}$}
\begin{center}
\includegraphics[scale=0.25]{taste_shocks.jpg}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Renegotiation: Summary}
\begin{itemize}
\item With the taste shocks: 
\begin{itemize}
\item $P(\text{divorce}) = P(\xi > M_{ebs})$, $P(\text{keep sq: }\theta = \theta_{sq}) = P(\xi < M_{sq})$
\item $P(\text{renegotiate: }\theta = \theta_{r_j}) = P(M_{r_{j-1}} < \xi < M_{r_j})$, $P(\text{other $\theta$'s}) = 0$
\end{itemize}
\item When single crossing is violated, all prettiness goes away. Assume it is not...
\item Let $\theta_0 = \theta_{sq}$ and $\theta_{h+1} = \theta_{ebs}$, $M_{h+1} = M_{ebs}$ and $M_0 = M_{sq}$ 
\item Expected value function with this transition is:
\begin{align*}& \E_\xi \left[ \I(\xi < M_{sq}) V(\theta_{sq}) + \sum\limits_{j=1}^{h+1} \I(M_{j-1} < \xi < M_j) V(\theta_j) + \I(\xi>M_{ebs})(V^{\text{div}} + \xi) \right] = \\
& \ \ \Lambda_\xi (M_{sq}) V(\theta_{sq}) + (\Lambda_\xi(M_{j}) - \Lambda_{\xi}(M_{j-1})) V_{\theta_j} + (1-\Lambda(M_{ebs}))\cdot [V^{\text{div}} + \E(\xi|\xi>M_{ebs})].
 \end{align*}
Here $\Lambda_\xi(t) = \left[1 + \exp\left( t / \sigma_{\xi} \right)\right]^{-1}$. As $\sigma_\xi \to 0$ this convegres to $\I(t < 0)$.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Adding Marriage is Simple}
\begin{itemize}
\item All the setup calls for egalitarian bargaining instead of Nash Bargaining Solution
\item Unlike $\theta_{nbs}$ that requires positive surplus, $\theta_{ebs}$ is always well-defined
\item When couple meet they pick $\theta = \theta_{ebs}$ and get common taste shock $-\zeta$
\item Agreement happens if $\min\{V^{f}_{ebs} - V^{fs} - \zeta, V^m_{ebs} - V^{ms} - \zeta \} > 0 \Leftrightarrow M_{ebs} > \zeta$
\item Expected value function (for the female) is then
\begin{align*}&\E_{\zeta} \left[ \I(\zeta < M_{ebs})(V^f - \zeta) + \I(\zeta > M_{ebs}) V^{fs} \right] = \\
& \ \ \Lambda_\zeta(M_{ebs}) \cdot \left[ V^f - \E(\zeta|\zeta<M_{ebs})\right] + \left[1-\Lambda_\zeta(M_{ebs})\right]\cdot V^{fs}
\end{align*}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Conclusions}
\begin{itemize}
\item This smoothing can fix numeric problems and is relatively easy to implement.
\item Potentially it can allow to use Euler equation based methods (EGM) in the limited commitment models and save large chunk of time by avoiding expensive optimization. This requires thinking but is already doable.
\item Even without Euler equations it should make value function iteration more stable
\item I should be writing
\end{itemize}
\end{frame}

\end{document}
